### 2021-03-31 submission/manuscript.Rmd
Begin drafting manuscript

---
title: ''
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace{35mm}

# An osmotic laxative renders mice susceptible to prolonged *Clostridioides difficile* colonization and hinders clearance

\vspace{35mm}

Sarah Tomkovich^1, Ana Taylor, Jacob Kingg, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI, USA

2\. The Unit for Laboratory Animal Medicine, University off Michigan, Ann Arbor, MI, USA

\newpage
\linenumbers

## Abstract
(Modify depending on target journal) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630 spores. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-infection (dpi). To examine how PEG treatment impacts clearance, we administered PEG for 1 day to clindamycin-treated, *C. difficile*-challenged mice either immediately following challenge or 3 dpi. We collected longitudinal stool samples to examine *C. difficile* levels in the stool via anaerobic culture and profiled the microbiota by 16S rRNA sequencing. PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpi. Additionally, 5-day PEG treated mice remained susceptible to CDI 10-days post treatment. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpi. Although 5-day PEG-treated mice exhibited prolonged *C. difficile* colonization, we saw no difference in histological inflammation between PEG- and clindamycin-treated mice. Additionally, administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpi in mice that received PEG immediately after challenge and 15 dpi in mice that received PEG 3 dpi. When we examined microbiota composition across our different treatment groups, we found decreased richness in the PEG-treated mice that exhibited prolonged *C. difficile* colonization. Importantly, there were increased Bacteroides and Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in most of the PEG-treated mice with prolonged *C. difficile* colonization. Our findings suggest the osmotic laxative PEG 3350 alters the mouse microbiota and disrupts colonization resistance to *C. difficile*, as well as clearance in mice with a CDI. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.

\newpage

## Introduction

## Results and Discussion

## Conclusions

## Acknowledgements
We thank members of the Schloss lab for feedback on planning the experiments and data presentation. We also thank Andrew Henry for help with media preparation and bacterial culture. We also thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help in maintaining the Schloss lab's anaerobic chamber. This work was supported by the National Institutes of Health (U01AI124255). ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240 from the National Center for Advancing Translational Sciences).

## Materials and Methods

\newpage

\includegraphics{figure_1.pdf}
**Figure 1. Chronic PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.** A. Setup of the experimental timeline for subset of experiments with 5-day PEG treated mice. B. *C. difficile* CFU/gram stool measured over time (N = `(insert variable name)`-`(insert variable name)` mice per timepoint) via serial dilutions. The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mcie came off of the PEG treatment) or early deaths. Lines represent the median for each source and circles represent individual mouse samples.

\newpage

## References

### 2021-03-31 code/1_day_PEG_16S.R
Update 1 Day PCoA gif

@@ -244,7 +244,7 @@ pcoa_animated <- plot_pcoa(pcoa_1_day_PEG)+
   labs(x = paste("PCoA 1 (", axis1, "%)", sep = ""), #Annotations for each axis from loadings file
        y = paste("PCoA 2 (", axis2,"%)", sep = ""))+
   labs(title = 'Day: {frame_time}') + #Adds time variable to title
-  transition_time(day)+  #Day variable used to cycle through time on animation
+  transition_time(as.integer(day))+  #Day variable used to cycle through time on animation
   shadow_mark() #Shows previous timepoints
 
 # Implement better frames per second for animation

### 2021-04-04 code/utilities.R
Add function to plot genus over time

@@ -794,6 +794,40 @@ hm_1_genus <- function(sample_df, specify_genus, timepoints){
           text = element_text(size = 16)) # Change font size for entire plot
 }
 
+#Function to plot a genus over time
+#genus_plot = genus to plot in quotes. Ex: "Bacteroides"
+#sample_df = subset dataframe of just stool or tissue samples
+genus_over_time <- function(genus_plot, sample_df){
+  specify_genus_name <- sample_df %>% 
+    filter(genus == genus_plot) %>% 
+    pull(genus)
+  genus_median <- sample_df %>% 
+    filter(genus == genus_plot) %>% 
+    group_by(group, day) %>% 
+    summarize(median=(median(agg_rel_abund + 1/2000))) %>% 
+    ungroup
+  genus_mice <- sample_df %>% 
+    filter(genus == genus_plot) %>% 
+    mutate(agg_rel_abund = agg_rel_abund + 1/2000) %>%
+    select(day, agg_rel_abund, genus, group)
+  genus_time <- ggplot(NULL)+
+    geom_point(genus_mice, mapping = aes(x=day, y=agg_rel_abund, color=group), size  = 1.5, position = position_dodge(width = 0.6))+
+    geom_line(genus_median, mapping = aes(x=day, y=median, group = group, color=group), size = 1, show.legend = FALSE)+
+    scale_colour_manual(name=NULL,
+                        values=color_scheme,
+                        breaks=color_groups,
+                        labels=color_labels)+
+    geom_hline(yintercept=1/1000, color="gray")+
+    labs(title=specify_genus_name,
+         x="Day",
+         y="Relative abundance (%)") +
+    scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+
+    theme_classic()+
+    theme(plot.title=element_text(hjust = 0.5, face = "italic"),
+          panel.grid.minor.x = element_line(size = 0.4, color = "grey"),  # Add gray lines to clearly separate symbols by days)
+          text = element_text(size = 18)) # Change font size for entire plot
+}
+
 #Function to plot an otu_over_time
 #otu_plot = otu to plot in quotes. Ex: "Peptostreptococcaceae (OTU 12)"
 #sample_df = subset dataframe of just stool or tissue samples

### 2021-04-04 code/5_days_PEG_16S.R
Examine genera over time, remove commented out code at top changing day to integer

@@ -11,11 +11,6 @@ color_labels <- c( "Clind.", "5-day PEG 3350", "5-day PEG 3350 + Clind.", "5-day
 shape_scheme <- c(1, 19)
 shape_infected <- c("no", "yes")
 
-#metadata <- metadata %>%
-#  mutate(day = as.integer(day))  #Day variable (transformed to integer to get rid of decimals on PCoA animation
-#agg_otu_data <- agg_otu_data %>% 
-#  mutate(day = as.integer(day))  #Day variable (transformed to integer to get rid of decimals on PCoA animation
-
 #Statistical Analysis----
 set.seed(19760620) #Same seed used for mothur analysis
 
@@ -452,7 +447,7 @@ save_plot(filename = "results/figures/5_days_PEG_genus_heatmap_tissues.png", hm_
 
 #List of genera that overlap between top genera for stool and tissue samples----
 top_genera <- intersect_all(top_10_sig_genus, top_sig_genus_tissues)
-#5 genera: Bacteroides, Clostridales Unclassified, Ruminococcaceae Unclassified, 
+#5 genera: Bacteroides, Clostridiales Unclassified, Ruminococcaceae Unclassified, 
 #Peptostreptococcaceae Unclassified, Acetatifactor
 
 #OTU analysis----
@@ -747,3 +742,93 @@ peg_impacted_genera_plot <- pairwise_genus_stools %>%
         strip.background = element_blank(),
         legend.position = "none") 
 save_plot(filename = paste0("results/figures/5_days_PEG_genera_impacted_by_PEG.png"), peg_impacted_genera_plot, base_height = 9, base_width = 8)
+
+#Examine genera of interest over time----
+genus_stools_t <- genus_stools %>% 
+  mutate(day = as.integer(day))  #Day variable transformed to integer for over time plots
+genus_tissues_t <- genus_tissues %>% 
+  mutate(day = as.integer(day))  #Day variable transformed to integer for over time plots  
+
+#Examine Bacteroides over time
+bacteroides_stools <- genus_over_time("Bacteroides", genus_stools_t)+
+  scale_x_continuous(breaks = c(-15, -10, -5, -4, -2, -1:10, 15, 20, 30),
+                     limits = c(-16,31),
[…]

### 2021-04-04 code/5_days_PEG_16S.R
Reorder list of genera in heatmaps to match order of list (genera that were significant overr the most timepoints)

@@ -415,7 +415,10 @@ top_10_sig_genus <- kw_genus_stools %>%
 hm_stool_days <- diversity_stools %>% distinct(day) %>% pull(day) #Have more timepoints than we tested (some days we only have sequenced samples for 1 group)
 facet_labels <- color_labels #Create descriptive labels for facets
 names(facet_labels) <- c("C", "WM", "WMC", "WMR") #values that correspond to group, which is the variable we're faceting by
-hm_stool <- hm_plot_genus(genus_stools, top_10_sig_genus, hm_stool_days)+
+hm_stool <- hm_plot_genus(genus_stools %>% 
+                            #Reorder list of genera to match number of days they were significantly different, Bacteroides was the top
+                            mutate(genus = fct_relevel(genus, rev(top_10_sig_genus))),
+                          top_10_sig_genus, hm_stool_days)+
   scale_x_discrete(breaks = c(-15, -10, -5, -4, -2, -1:10, 15, 20, 30), labels = c(-15, -10, -5, -4, -2, -1:10, 15, 20, 30)) 
 save_plot(filename = "results/figures/5_days_PEG_genus_heatmap_stools.png", hm_stool, base_height = 14, base_width = 15)
 
@@ -441,7 +444,10 @@ sig_genus_tissue_top_p <- sig_genus_tissues %>%
 top_sig_genus_tissues <- c(sig_genus_tissues_multi_day, sig_genus_tissue_top_p)
 #Create heatmap of significant genera for tissue samples----
 hm_tissue_days <- diversity_tissues %>% distinct(day) %>% pull(day)
-hm_tissues <- hm_plot_genus(genus_tissues, top_sig_genus_tissues, hm_tissue_days)+
+hm_tissues <- hm_plot_genus(genus_tissues%>% 
+                              #Reorder list of genera to match number of days they were significantly different, Bacteroides was the top
+                              mutate(genus = fct_relevel(genus, rev(top_sig_genus_tissues))), 
+                            top_sig_genus_tissues, hm_tissue_days)+
   scale_x_discrete(breaks = c(4, 6, 20, 30), labels = c(4, 6, 20, 30)) 
 save_plot(filename = "results/figures/5_days_PEG_genus_heatmap_tissues.png", hm_tissues, base_height = 14, base_width = 15)
 

### 2021-04-04 code/utilities.R
Add function to create lineplot of genera of interest from day -1 to 10

@@ -557,7 +557,7 @@ plot_genus_dx <- function(sample_df, genera, timepoint){
                         values=color_scheme,
                         breaks=color_groups,
                         labels=color_labels)+
-    geom_hline(yintercept=1/1000, color="gray")+
+    geom_hline(yintercept=1/1000, color="gray")+ #Represents limit of detection
     stat_summary(fun = 'median',
                  fun.max = function(x) quantile(x, 0.75),
                  fun.min = function(x) quantile(x, 0.25),
@@ -610,7 +610,7 @@ hm_plot_otus <- function(sample_df, otus, timepoints){
 #Function to create a heatmap plot the relative abundances of a list of Genera over time, faceted by group----
 #Arguments: 
 #sample_df = subset dataframe of samples to be plotted
-#Genera = list of genera to plot
+#genera = list of genera to plot
 #timepoints = days of the experiment to plot
 hm_plot_genus <- function(sample_df, genera, timepoints){
   sample_df %>%
@@ -894,4 +894,36 @@ otu_gi_distrib <- function(otu_plot, sample_df, timepoint, group_name){
           text = element_text(size = 18)) # Change font size for entire plot
 }
 
-
+#Function to create faceted line plots of relative abundances of genera of interest over time
+#sample_df = subset dataframe of samples to be plotted
+#genera = list of genera to plot
+#timepoints = days of the experiment to plot
+line_plot_genus <- function(sample_df, genera, timepoints){
+  sample_df %>% 
+    filter(genus %in% genera) %>% #Select only genera of interest
+    mutate(genus = fct_relevel(genus, genera)) %>% #Reorder genera to match order of genera of interest
+    mutate(group = fct_relevel(group, rev(color_groups))) %>% #Specify the order of the groups
+    filter(day %in% timepoints) %>% #Select only timepoints of interest
+    mutate(day = as.integer(day)) %>% 
+    group_by(group, genus, day) %>% 
+    summarize(median=median(agg_rel_abund + 1/2000),`.groups` = "drop") %>%  #Add small value (1/2Xsubssampling parameter) so that there are no infinite values with log transformation
+    ggplot()+
+    geom_line(aes(x = day, y=median, color=group))+
+    scale_colour_manual(name=NULL,
+                        values=color_scheme,
+                        breaks=color_groups,
+                        labels=color_labels)+
[…]

### 2021-04-04 code/5_days_PEG_16S.R
Add line plots for tissue and stool samples

@@ -422,6 +422,15 @@ hm_stool <- hm_plot_genus(genus_stools %>%
   scale_x_discrete(breaks = c(-15, -10, -5, -4, -2, -1:10, 15, 20, 30), labels = c(-15, -10, -5, -4, -2, -1:10, 15, 20, 30)) 
 save_plot(filename = "results/figures/5_days_PEG_genus_heatmap_stools.png", hm_stool, base_height = 14, base_width = 15)
 
+#Lineplots of the top 6 significant genera in stool samples----
+lp_stool_days <- diversity_stools %>% distinct(day) %>% 
+  filter(!day %in% c("-15", "-10", "-5", "-4", "-2", "15", "20", "30")) %>% #Focus on day -1 through 10 timepoints
+  pull(day)
+facet_labels <- top_10_sig_genus[1:6] #Pick just the top 6
+names(facet_labels) <- top_10_sig_genus[1:6] #Pick just the top 6
+lp_stool <- line_plot_genus(genus_stools, top_10_sig_genus[1:6], lp_stool_days)
+save_plot(filename = "results/figures/5_days_PEG_genus_lineplot_stools.png", lp_stool, base_height = 5, base_width = 8)
+
 #List of the 10 significant genera in tissue samples that are significant over the most timepoints
 sig_genus_tissues<- kw_genus_tissues %>% 
   filter(p.value.adj < 0.05) %>% #Select only significant p-values
@@ -450,6 +459,16 @@ hm_tissues <- hm_plot_genus(genus_tissues%>%
                             top_sig_genus_tissues, hm_tissue_days)+
   scale_x_discrete(breaks = c(4, 6, 20, 30), labels = c(4, 6, 20, 30)) 
 save_plot(filename = "results/figures/5_days_PEG_genus_heatmap_tissues.png", hm_tissues, base_height = 14, base_width = 15)
+
+#Lineplots of the top 6 significant genera in tissue samples----
+lp_tissue_days <- diversity_tissues %>% distinct(day) %>% 
+  filter(day %in% c("4", "6", "20", "30")) %>% #Focus on day -1 through 10 timepoints
+  pull(day)
+facet_labels <- top_sig_genus_tissues[1:6] #Pick just the top 6
+names(facet_labels) <- top_sig_genus_tissues[1:6] #Pick just the top 6
+lp_tissue <- line_plot_genus(genus_tissues, top_sig_genus_tissues[1:6], lp_tissue_days)+
[…]

### 2021-04-04 code/utilities.R
Add argument to lineplot function to specify linetype to differentiate infected vs mock challenged mice

@@ -898,7 +898,8 @@ otu_gi_distrib <- function(otu_plot, sample_df, timepoint, group_name){
 #sample_df = subset dataframe of samples to be plotted
 #genera = list of genera to plot
 #timepoints = days of the experiment to plot
-line_plot_genus <- function(sample_df, genera, timepoints){
+#specify_linetype = "solid" for C. diff challenged. "dashed" for mice that were mock challenged
+line_plot_genus <- function(sample_df, genera, timepoints, specify_linetype){
   sample_df %>% 
     filter(genus %in% genera) %>% #Select only genera of interest
     mutate(genus = fct_relevel(genus, genera)) %>% #Reorder genera to match order of genera of interest
@@ -908,7 +909,7 @@ line_plot_genus <- function(sample_df, genera, timepoints){
     group_by(group, genus, day) %>% 
     summarize(median=median(agg_rel_abund + 1/2000),`.groups` = "drop") %>%  #Add small value (1/2Xsubssampling parameter) so that there are no infinite values with log transformation
     ggplot()+
-    geom_line(aes(x = day, y=median, color=group))+
+    geom_line(aes(x = day, y=median, color=group), linetype = specify_linetype)+
     scale_colour_manual(name=NULL,
                         values=color_scheme,
                         breaks=color_groups,

### 2021-04-04 code/5_days_PEG_16S.R
Add lineplots for mock infected mice

@@ -428,7 +428,7 @@ lp_stool_days <- diversity_stools %>% distinct(day) %>%
   pull(day)
 facet_labels <- top_10_sig_genus[1:6] #Pick just the top 6
 names(facet_labels) <- top_10_sig_genus[1:6] #Pick just the top 6
-lp_stool <- line_plot_genus(genus_stools, top_10_sig_genus[1:6], lp_stool_days)
+lp_stool <- line_plot_genus(genus_stools, top_10_sig_genus[1:6], lp_stool_days, "solid")
 save_plot(filename = "results/figures/5_days_PEG_genus_lineplot_stools.png", lp_stool, base_height = 5, base_width = 8)
 
 #List of the 10 significant genera in tissue samples that are significant over the most timepoints
@@ -466,7 +466,7 @@ lp_tissue_days <- diversity_tissues %>% distinct(day) %>%
   pull(day)
 facet_labels <- top_sig_genus_tissues[1:6] #Pick just the top 6
 names(facet_labels) <- top_sig_genus_tissues[1:6] #Pick just the top 6
-lp_tissue <- line_plot_genus(genus_tissues, top_sig_genus_tissues[1:6], lp_tissue_days)+
+lp_tissue <- line_plot_genus(genus_tissues, top_sig_genus_tissues[1:6], lp_tissue_days, "solid")+
   scale_x_continuous(limits = c(3.5,30.5), breaks = c(4, 6, 20, 30), labels = c(4, 6,20, 30))#Change scale since we have less timepoints for tissues
 save_plot(filename = "results/figures/5_days_PEG_genus_lineplot_tissues.png", lp_tissue, base_height = 5, base_width = 8)
 
@@ -857,3 +857,33 @@ akkermansia_tissues <- genus_over_time("Akkermansia", genus_tissues_t)+
                      minor_breaks = c(3.5, 4.5, 5.5, 6.5, 19.5, 20.5, 29.5, 30.5)) +
   theme(legend.position = "bottom")
 save_plot(filename = "results/figures/5_days_PEG_genus_akkermansia_tissues.png", akkermansia_tissues, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
+
+#Plots of mock-infected mice----
+#Use same color scheme but change symbols to open circles and lines to dashes
+#Define color scheme to match 5_days_PEG plots----
+color_scheme <- c("#238b45", "#88419d") #Adapted from http://colorbrewer2.org/#type=sequential&scheme=BuPu&n=4
[…]

### 2021-04-04 submission/manuscript.Rmd
Add example code chunk while demoing R markdown for Jacob
Add figure titles from outline

@@
 \newpage
+```{r sample_size for figure legends}
+
+test <- 4
+
+```
 
@@
 *C.
-difficile* CFU/gram stool measured over time (N = `(insert variable name)`-`(insert variable name)` mice per timepoint) via serial dilutions.
+difficile* CFU/gram stool measured over time (N = `r test`-`(insert variable name)` mice per timepoint) via serial dilutions.
 The black line represents the limit of detection for the first serial dilution.
@@
 Lines represent the median for each source and circles represent individual mouse samples.
+\newpage
 
+\includegraphics{figure_2.pdf}
+**Figure 2. Chronic PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. 
+\newpage
+
+\includegraphics{figure_3.pdf}
+**Figure 3. Chronic PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.** A. 
+\newpage
+
+\includegraphics{figure_4.pdf}
+**Figure 4. Acute PEG 3350 treatment renders mice susceptible to transient *C. difficile* colonization.** A. 
+\newpage
+
+\includegraphics{figure_5.pdf}
+**Figure 5. Acute PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A. 
+\newpage
+
+\includegraphics{figure_6.pdf}
+**Figure 6. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.** A. 
+\newpage
+
+\includegraphics{figure_7.pdf}
+**Figure 7. Schematic summarizing findings.** A. 
+\newpage
+
+\includegraphics{figure_S1.pdf}
+**Figure S1. Chronic PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.** A. 
+\newpage
+
+\includegraphics{figure_S2.pdf}
+**Figure S2. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A. 
 \newpage

### 2021-04-04 code/5_days_PEG_16S.R
Fix filename

@@ -549,7 +549,7 @@ porph_tissues <- otu_over_time("Porphyromonadaceae (OTU 8)", otu_tissues)+
                      minor_breaks = c(3.5, 4.5, 5.5, 6.5, 19.5, 20.5, 29.5, 30.5)) +
   theme(legend.position = "bottom")
 
-save_plot(filename = "results/figures/5_days_otu_porphyromonadaceae8_porph_tissues.png", porph_tissues, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
+save_plot(filename = "results/figures/5_days_PEG_otu_porphyromonadaceae8_porph_tissues.png", porph_tissues, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
 #Examine changes that happen in WMR group "5-day PEG 3350 + 10-day recovery" 
 #post C. difficile challenge (day 1 versus day 15)----

### 2021-04-04 code/5_days_PEG_histology_scores.R
Make histology plots more uniform by using the same save function, update filenames

@@ -164,8 +164,7 @@ pairwise_day6_plot <-d6_plot_pairwise_stats %>%
 
 #Plot of day 6 histology scores----
 d6_plot <- plot_histology(d6_histology, pairwise_day6_plot)
-save_plot("results/figures/histo_scores_d6.png", d6_plot) #Use save_plot instead of ggsave because it works better with cowplot
-
+save_plot("results/figures/5_days_PEG_histo_scores_d6.png", d6_plot) #Use save_plot instead of ggsave because it works better with cowplot
 
 #Plot of day 4 (plus day 0 for 1 group) histology scores----
 #Define color scheme for d4_histology----
@@ -181,5 +180,5 @@ pairwise_day4_plot <- d4_plot_pairwise_stats %>%
 
 #Plot of day 4 histology scores----
 d4_plot <- plot_histology(d4_histology, pairwise_day4_plot)
-ggsave("results/figures/histo_scores_d4.png", d4_plot)
+save_plot("results/figures/5_days_PEG_histo_scores_d4.png", d4_plot) #Use save_plot instead of ggsave because it works better with cowplot
 

### 2021-04-04 submission/manuscript.Rmd
Change chronic/acute to 5-day/1-day

@@
 **Figure 1.
-Chronic PEG treatment prolongs susceptibility and mice become persistently colonized with *C.
+5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C.
 difficile*.** A.
@@
 \includegraphics{figure_2.pdf}
-**Figure 2. Chronic PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. 
+**Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. 
 \newpage
@@
 \includegraphics{figure_3.pdf}
-**Figure 3. Chronic PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.** A. 
+**Figure 3. 5-day PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.** A. 
 \newpage
@@
 \includegraphics{figure_4.pdf}
-**Figure 4. Acute PEG 3350 treatment renders mice susceptible to transient *C. difficile* colonization.** A. 
+**Figure 4. 1-day PEG treatment renders mice susceptible to transient *C. difficile* colonization.** A. 
 \newpage
@@
 \includegraphics{figure_5.pdf}
-**Figure 5. Acute PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A. 
+**Figure 5. 1-day PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A. 
 \newpage
@@
 \includegraphics{figure_S1.pdf}
-**Figure S1. Chronic PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.** A. 
+**Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.** A. 
 \newpage

### 2021-04-05 code/1_day_PEG_16S.R
Make alternative version of lineplot with function in code/utilites.R

@@ -1,7 +1,6 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 source("code/16S_common_files.R") #Reads in mothur output files
 
-
 #Define color scheme to match 1_Day_PEG Weight/CFU Plots----
 color_scheme <- c("#238b45", "#88419d", "#225ea8") #Adapted from http://colorbrewer2.org/#type=sequential&scheme=BuPu&n=4
 color_groups <- c( "C", "M1", "1RM1")
@@ -640,5 +639,17 @@ line_plot_baselinetoD1_5_genera <- agg_genus_data_subset %>%
         text = element_text(size = 16)) # Change font size for entire plot
 save_plot(filename = "results/figures/1_Day_PEG_genus_5_baselinetoD1_lineplot.png", line_plot_baselinetoD1_5_genera, base_height = 5, base_width = 15)
 
+#Lineplot v2 with function from code/utilities.R #
+#Chose 1 more genera to make it an even 6
+sig_genus_top_list <- sig_genus_pairs[!(sig_genus_pairs %in% sig_genus_pairs_D7)] #14 genera
+sig_genus_remove_names <- c("Ruminococcus", "Unclassified", "Alistipes", "Clostridium XlVb", "Enterorhabdus", "Pseudoflavonifractor", 
+                            "Anaeroplasma","Firmicutes Unclassified", "Bacteroidales Unclassified",
+                            "Acetatifactor", "Akkermansia", "Butyricicoccus")
+sig_genus_top_list <- sig_genus_top_list[!(sig_genus_top_list %in% sig_genus_remove_names)]
 
+
+facet_labels <- sig_genus_top_list
+names(facet_labels) <- sig_genus_top_list
+line_plot_baselinetoD1_5_genera <- line_plot_genus(agg_genus_data_subset, sig_genus_top_list, hm_days, "solid")
+save_plot(filename = "results/figures/1_Day_PEG_genus_5_baselinetoD1_lineplot.png", line_plot_baselinetoD1_5_genera, base_height = 5, base_width = 15)
 

### 2021-04-05 code/1_day_PEG_cfu_weight.R
Add vertical lines to cfu plot

@@ -144,10 +144,10 @@ cfudata <- cfudata %>%
   mutate(day = fct_relevel(day, "0", "1", "2","3", "4", "5", "6", "7", "8", "9", "10")) %>%
   mutate(group = fct_relevel(group, "C", "M1", "1RM1"))
 cfu <- plot_cfu_data(cfudata) + 
+  scale_x_discrete(breaks = c(1:10))+
+  geom_vline(xintercept = c((1:11) - 0.5 ), color = "grey") + # Add gray lines to clearly separate OTUs
   theme(legend.position = "none",
-        text = element_text(size = 14))
-  scale_x_discrete(breaks = c(1:10))
-
+        text = element_text(size = 16))
 save_plot(filename = "results/figures/1_day_PEG_cfu.png", cfu, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
 #Plot of weight change data----

### 2021-04-05 code/1_day_PEG_16S.R
Update x axis scale for lineplot

@@ -108,7 +108,6 @@ richness_tissues <- plot_richness_overtime(diversity_data_subset) +
   theme(legend.position = "bottom")
 save_plot(filename = "results/figures/1_Day_PEG_richness.png", richness_tissues, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
-
 #Pairwise-Comparison of diversity data----
 #Function for pairwise Wilcoxon rank sum test for shannon diversity
 pairwise_shannon <- function(diversity_subset, timepoints){
@@ -650,6 +649,7 @@ sig_genus_top_list <- sig_genus_top_list[!(sig_genus_top_list %in% sig_genus_rem
 
 facet_labels <- sig_genus_top_list
 names(facet_labels) <- sig_genus_top_list
-line_plot_baselinetoD1_5_genera <- line_plot_genus(agg_genus_data_subset, sig_genus_top_list, hm_days, "solid")
+line_plot_baselinetoD1_5_genera <- line_plot_genus(agg_genus_data_subset, sig_genus_top_list, hm_days, "solid")+
+  scale_x_discrete(limits = c("baseline", "1", "2", "5", "7"), breaks = c("baseline", "1", "2", "5", "7"), labels = c("baseline", "1", "2", "5", "7")) +
 save_plot(filename = "results/figures/1_Day_PEG_genus_5_baselinetoD1_lineplot.png", line_plot_baselinetoD1_5_genera, base_height = 5, base_width = 15)
 

### 2021-04-05 code/figure_3.R
Increase figure size, update to include revised plots, add figure to submission folderr

@@ -8,6 +8,6 @@ e <- ggdraw() + draw_image("results/figures/1_Day_PEG_genus_5_baselinetoD1_linep
 top_panel <- plot_grid(a, b, c, d, labels = c("A", "B", "C", "D"), label_size = 12, ncol = 2, rel_widths = c(.60, 1, 1,.60), rel_heights = c(1, 1))
 bottom_panel <- plot_grid(e, labels = c("E"), label_size = 12)
 plot_grid(top_panel, bottom_panel, nrow = 2, rel_heights = c(1.1, .8)) +
-  ggsave("results/figures/figure_3.pdf", width = 5.25, height = 5.0)
-  ggsave("submission/figure_3", width = 5.25, height = 5.0)
+  ggsave("results/figures/figure_3.pdf", width = 6.25, height = 5.5)+
+  ggsave("submission/figure_3.pdf", width = 6.25, height = 5.5)
 

### 2021-04-05 code/figure_3.R
Renumber figures

@@ -1,13 +1,0 @@
-source("code/utilities.R")
-
-a <- ggdraw() + draw_image("results/figures/1_Day_PEG_Schematic.png", scale = 1.2, x = .07)
-b <- ggdraw() + draw_image("results/figures/1_day_PEG_cfu.png", scale = 1.00)
-c <- ggdraw() + draw_image("results/figures/1_Day_PEG_PCoA.png", scale = 1.25)
-d <- ggdraw() + draw_image("results/figures/1_Day_PEG_shannon.png")
-e <- ggdraw() + draw_image("results/figures/1_Day_PEG_genus_5_baselinetoD1_lineplot.png")
-top_panel <- plot_grid(a, b, c, d, labels = c("A", "B", "C", "D"), label_size = 12, ncol = 2, rel_widths = c(.60, 1, 1,.60), rel_heights = c(1, 1))
-bottom_panel <- plot_grid(e, labels = c("E"), label_size = 12)
-plot_grid(top_panel, bottom_panel, nrow = 2, rel_heights = c(1.1, .8)) +
-  ggsave("results/figures/figure_3.pdf", width = 6.25, height = 5.5)+
-  ggsave("submission/figure_3.pdf", width = 6.25, height = 5.5)
-

### 2021-04-05 code/figure_4.R

@@ -1,48 +1,13 @@
-source("code/utilities.R") #Loads libraries, reads in metadata, functions
-#One figure with schematic, CFU over time, and shannon over time (for stools only)? Check dimensions of Vendor paper for long wide version. plot grid for three panels that match up, and do another one? Vendor paper fig 1 can choose which panel goes were. 
-#One figure with important 16S highlights: still pcoa, heatmap? Or choose a select couple days and do top genera/OTUs? 
-#heat map main diffs between groups, keep it at genera level that don't recover in all PEG groups that remain colonized or have recovered by day 15
-#Should I do a whole another fig for the tissue differences? Or should I combine with the 16S fig above?
-#OTU12 over time probably in supplements bc it isnt as reliable
-#Vendor paper Fig 6 makes fig legend a lot smaller
-#Correlate c diff with microbiome changes
-#Tropini 2018 says taxa that got effected by laxative treatments (bacteriodes, S24-7)
+source("code/utilities.R")
 
-PEG_treatment_legend <- ggplot() +
-  theme_void() +
-  scale_x_continuous(limits = c(-10, 10))+
-  scale_y_continuous(limits = c(-1.5, 1.5))+
-  annotate("rect", xmin = -10, xmax = -9, ymin = -.75, ymax = .75, fill = "#88419d", alpha = .2)+ #shade to indicate PEG treatment in Clind + 1-day PEG group
-  annotate("rect", xmin = -1.5, xmax = -.5, ymin = 0, ymax = .75, fill = "#f768a1", alpha = .2)+ #shade to indicate PEG treatment in Clind + 3-day recovery + 1-day PEG + FMT/PBS
-  annotate("rect", xmin = -1.5, xmax = -.5, ymin = -.75, ymax = 0, fill = "#225ea8", alpha = .2)+
-  annotate("text", x = -5.25, y = 0, label = "PEG treatment in Clind + 1-day PEG group", size = 3.78)+
-  annotate("text", x = 5.3, y = 0, label = "PEG treatment in Clind + 3-day recovery + 1-day PEG (+ FMT/PBS)", size = 3.78)+
-  xlab(NULL) 
[…]

### 2021-04-05 code/figure_5.R

@@ -1,0 +1,48 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+#One figure with schematic, CFU over time, and shannon over time (for stools only)? Check dimensions of Vendor paper for long wide version. plot grid for three panels that match up, and do another one? Vendor paper fig 1 can choose which panel goes were. 
+#One figure with important 16S highlights: still pcoa, heatmap? Or choose a select couple days and do top genera/OTUs? 
+#heat map main diffs between groups, keep it at genera level that don't recover in all PEG groups that remain colonized or have recovered by day 15
+#Should I do a whole another fig for the tissue differences? Or should I combine with the 16S fig above?
+#OTU12 over time probably in supplements bc it isnt as reliable
+#Vendor paper Fig 6 makes fig legend a lot smaller
+#Correlate c diff with microbiome changes
+#Tropini 2018 says taxa that got effected by laxative treatments (bacteriodes, S24-7)
+
+PEG_treatment_legend <- ggplot() +
+  theme_void() +
+  scale_x_continuous(limits = c(-10, 10))+
+  scale_y_continuous(limits = c(-1.5, 1.5))+
+  annotate("rect", xmin = -10, xmax = -9, ymin = -.75, ymax = .75, fill = "#88419d", alpha = .2)+ #shade to indicate PEG treatment in Clind + 1-day PEG group
+  annotate("rect", xmin = -1.5, xmax = -.5, ymin = 0, ymax = .75, fill = "#f768a1", alpha = .2)+ #shade to indicate PEG treatment in Clind + 3-day recovery + 1-day PEG + FMT/PBS
+  annotate("rect", xmin = -1.5, xmax = -.5, ymin = -.75, ymax = 0, fill = "#225ea8", alpha = .2)+
+  annotate("text", x = -5.25, y = 0, label = "PEG treatment in Clind + 1-day PEG group", size = 3.78)+
+  annotate("text", x = 5.3, y = 0, label = "PEG treatment in Clind + 3-day recovery + 1-day PEG (+ FMT/PBS)", size = 3.78)+
+  xlab(NULL) 
+save_plot("results/figures/post_CDI_PEG_treatment_legend.png", PEG_treatment_legend, base_height = 1, base_width = 9)
+ 
[…]

### 2021-04-05 submission/manuscript.Rmd

@@
 B.
+Weight change from baseline weight in groups after treatment with PEG and/or clindamycin, followed by *C.
+difficile* challenge.
+C.
 *C.
@@
 The black line represents the limit of detection for the first serial dilution.
-CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mcie came off of the PEG treatment) or early deaths.
+CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths.
 Lines represent the median for each source and circles represent individual mouse samples.
+Asterisks indicate timepoints where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple timepoints.
 \newpage

### 2021-04-05 submission/manuscript.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\vspace{35mm}

\hypertarget{an-osmotic-laxative-renders-mice-susceptible-to-prolonged-clostridioides-difficile-colonization-and-hinders-clearance}{%
\section{\texorpdfstring{An osmotic laxative renders mice susceptible to
prolonged \emph{Clostridioides difficile} colonization and hinders
clearance}{An osmotic laxative renders mice susceptible to prolonged Clostridioides difficile colonization and hinders clearance}}\label{an-osmotic-laxative-renders-mice-susceptible-to-prolonged-clostridioides-difficile-colonization-and-hinders-clearance}}

\vspace{35mm}

Sarah Tomkovich\^{}1, Ana Taylor, Jacob Kingg, Joanna Colovas, Lucas
Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid
L. Bergin\^{}2, Patrick D. Schloss\textsuperscript{1\(\dagger\)}

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI, USA

2. The Unit for Laboratory Animal Medicine, University off Michigan, Ann
Arbor, MI, USA

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

(Modify depending on target journal) Antibiotics are a major risk factor
for \emph{Clostridioides difficile} infections (CDIs) because of their
impact on the intestinal microbiome. However, non-antibiotic medications
such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350,
also alter the microbiota, but whether PEG impacts CDI susceptibility
and clearance is unclear. To examine how PEG impacts susceptibility, we
treated C57Bl/6 mice with 5-day and 1-day doses of 15\% PEG in the
drinking water and then challenged the mice with \emph{C. difficile} 630
spores. We used clindamycin-treated mice as a control because they
consistently clear \emph{C. difficile} within 10 days post-infection
(dpi). To examine how PEG treatment impacts clearance, we administered
PEG for 1 day to clindamycin-treated, \emph{C. difficile}-challenged
mice either immediately following challenge or 3 dpi. We collected
longitudinal stool samples to examine \emph{C. difficile} levels in the
stool via anaerobic culture and profiled the microbiota by 16S rRNA
sequencing. PEG treatment alone was sufficient to render mice
susceptible to CDI and 5-day PEG-treated mice remain colonized for up to
30 dpi. Additionally, 5-day PEG treated mice remained susceptible to CDI
10-days post treatment. In contrast, 1-day PEG treated mice were
transiently colonized, clearing \emph{C. difficile} within 7 dpi.
Although 5-day PEG-treated mice exhibited prolonged \emph{C. difficile}
colonization, we saw no difference in histological inflammation between
PEG- and clindamycin-treated mice. Additionally, administering PEG to
mice after \emph{C. difficile} challenge prolonged colonization up to 30
dpi in mice that received PEG immediately after challenge and 15 dpi in
mice that received PEG 3 dpi. When we examined microbiota composition
across our different treatment groups, we found decreased richness in
the PEG-treated mice that exhibited prolonged \emph{C. difficile}
colonization. Importantly, there were increased Bacteroides and
Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in
most of the PEG-treated mice with prolonged \emph{C. difficile}
colonization. Our findings suggest the osmotic laxative PEG 3350 alters
the mouse microbiota and disrupts colonization resistance to \emph{C.
difficile}, as well as clearance in mice with a CDI. Considering that
most hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact human
microbiota colonization resistance.

\newpage

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\hypertarget{results-and-discussion}{%
\subsection{Results and Discussion}\label{results-and-discussion}}

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

We thank members of the Schloss lab for feedback on planning the
experiments and data presentation. We also thank Andrew Henry for help
with media preparation and bacterial culture. We also thank the Unit for
Laboratory Animal Medicine at the University of Michigan for maintaining
our mouse colony and providing the institutional support for our mouse
experiments. Finally, we thank Kwi Kim, Austin Campbell, and Kimberly
Vendrov for their help in maintaining the Schloss lab's anaerobic
chamber. This work was supported by the National Institutes of Health
(U01AI124255). ST was supported by the Michigan Institute for Clincial
and Health Research Postdoctoral Translation Scholars Program
(UL1TR002240 from the National Center for Advancing Translational
Sciences).

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\newpage

\includegraphics{figure_1.pdf} \textbf{Figure 1. Chronic PEG treatment
prolongs susceptibility and mice become persistently colonized with
\emph{C. difficile}.} A. Setup of the experimental timeline for subset
of experiments with 5-day PEG treated mice. B. \emph{C. difficile}
CFU/gram stool measured over time (N =
4-\texttt{(insert\ variable\ name)} mice per timepoint) via serial
dilutions. The black line represents the limit of detection for the
first serial dilution. CFU quantification data was not available for
each mouse due to stool sampling difficulties (particularly the day the
mcie came off of the PEG treatment) or early deaths. Lines represent the
median for each source and circles represent individual mouse samples.

\newpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\end{document}

### 2021-04-05 code/ml_results.R
Edit script after merging in genus level results

@@ -47,13 +47,13 @@ performance_genus <- genus_results %>%
         axis.text.x = element_text(angle = 45, hjust = 1), #Angle axis labels
         strip.background = element_blank()) +#Make Strip backgrounds blank
   guides(color=guide_legend(nrow = 2))+   #Legend in 2 rows so it doesn't get cut off 
-  ggsave("results/figures/ml_performance_otu.png", height = 5, width = 8)
+  ggsave("results/figures/ml_performance_genus.png", height = 5, width = 8)
 
 #Examine permutation importance results for random forest---- 
 # random forest  had the highest median AUC
 #Function to read in feature importance from random forest model----
-#Will also get the corresponding otu name from taxonomy file
-#And format otu name to work with glue package for italicizing labels
+#Will also get the corresponding genus name from taxonomy file
+#And format genus name to work with glue package for italicizing labels
 #file_path = path to the file name
 read_feat_imp <- function(file_path){
   feat_imp <- read_csv(file_path)
@@ -93,9 +93,9 @@ top_20 <- function(df){
 rf_top_feat <- top_20(rf_feat) %>% pull(genus)
 
 
-#Function to filter to top OTUs for each pairwise comparison & plot results----
+#Function to filter to top genera for each pairwise comparison & plot results----
 #df = dataframes of feature importances for all seeds
-#top_otus = dataframes of top otus
+#top_feat = dataframes of top features
 #comp_name = name of comparison to title the plot (in quotes)
 plot_feat_imp <- function(df, top_feat){
   df %>% 
@@ -121,7 +121,7 @@ plot_feat_imp <- function(df, top_feat){
           legend.position = "none")   
 }
 
-#Plot feature importances for the top OTUs for each comparison----
+#Plot feature importances for the top genera for each comparison----
 rf_feat_5dpi <- plot_feat_imp(rf_feat, rf_top_feat)+
   ggsave("results/figures/ml_top_features_genus.png", height = 5, width = 8)
 
@@ -135,7 +135,7 @@ shape_labels <- c("Clind.", "1-day", "5-day", "Post-CDI")
 
 interp_genera_d5_top_10 <- rf_top_feat[1:10]
 #Plot the top 10 features that were important to Day 5 model and 
-#using facet_wrap, highlight the OTUs that correlate with colonization
[…]

### 2021-04-05 code/figure_6.R
Create summmary of machine learning results with 5dpi input data at the genus level

@@ -3,9 +3,10 @@ source("code/utilities.R") #Loads libraries, reads in metadata, functions
 #Figure with Schematic, CFU over time, and shannon over time
 a <- ggdraw() + draw_image("results/figures/ml_performance_genus.png")
 b <- ggdraw() + draw_image("results/figures/ml_top_features_genus.png")
-top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow=1, rel_widths = c(1,2))
+top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow=1, rel_widths = c(1,1.5), rel_heights = c(2, 1))
 c <- ggdraw() + draw_image("results/figures/ml_d5_top10_genus.png")#Note scale could be removed
 
-plot_grid(top_panel, c, labels = c("", "C"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_6.pdf", width=8, height=8)+
-  ggsave("submission/figure_6.pdf", width=8, height=8)
+plot_grid(top_panel, c, labels = c("", "C"), label_size = 12, ncol=1, rel_widths = c(1, 3), rel_heights = c(1, 1.5))+
+  ggsave("results/figures/figure_6.pdf", width=8, height=7)+
+  ggsave("submission/figure_6.pdf", width=8, height=7)
+

### 2021-04-05 code/figure_6.R
Remove previous description

@@ -1,6 +1,5 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-#Figure with Schematic, CFU over time, and shannon over time
 a <- ggdraw() + draw_image("results/figures/ml_performance_genus.png")
 b <- ggdraw() + draw_image("results/figures/ml_top_features_genus.png")
 top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow=1, rel_widths = c(1,1.5), rel_heights = c(2, 1))

### 2021-04-05 code/figure_7.R
Add draft of summary figure

@@ -1,0 +1,8 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+
+a <- ggdraw() + draw_image("results/figures/summary_schematic.png")
+
+plot_grid(a, labels = NULL, label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_7.pdf", width=6.875, height=4.5)+
+  ggsave("submission/figure_7.pdf", width=6.875, height=4.5)
+

### 2021-04-05 code/figure_S1.R
Intial drafts of supplemental figures

@@ -1,0 +1,7 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_heatmap_stools_WMR.png")
+
+plot_grid(a, labels = c("A"), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_S1.pdf", width=6.875, height=7)+
+  ggsave("submission/figure_S1.pdf", width=6.875, height=7)

### 2021-04-05 code/figure_S2.R

@@ -1,0 +1,7 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+
+a <- ggdraw() + draw_image("results/figures/otus_potential_muribaculum_blast_results.png")
+
+plot_grid(a, labels = c("A"), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_S2.pdf", width=6.875, height=6)+
+  ggsave("submission/figure_S2.pdf", width=6.875, height=6)

### 2021-04-05 code/figure_2.R
Add preliminary figures

@@ -1,0 +1,15 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_stool_PCoA.png", scale = 1)
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools.png", scale = 1.1)
+c <- ggdraw() + draw_image("results/figures/5_days_PEG_genera_impacted_by_PEG.png")
+d <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_stools.png", scale =0.95)
+
+left_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol = 1, rel_widths = c(1.5, 1))
+right_panel <- plot_grid(c, labels = c("C"), label_size = 12, nrow = 1)
+top_panel <- plot_grid(left_panel, right_panel, nrow = 1)
+bottom_panel <- plot_grid(d, labels = c("D"), label_size = 12, ncol = 1)
+plot_grid(top_panel, bottom_panel, nrow = 2) +
+  ggsave("results/figures/figure_2.pdf", width = 6.875, height = 9)+
+  ggsave("submission/figure_2.pdf", width = 6.875, height = 9)
+

### 2021-04-05 code/figure_3.R

@@ -1,0 +1,15 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_histo_scores_d4.png")
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_histo_scores_d6.png")
+c <- ggdraw() + draw_image("results/figures/5_days_PEG_tissues_PCoA.png")
+d <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues.png")
+e <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_tissues.png")
+
+left_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol = 1, rel_widths = c(1.5, 1))
+right_panel <- plot_grid(c, labels = c("C"), label_size = 12, nrow = 1)
+top_panel <- plot_grid(left_panel, right_panel, nrow = 1)
+bottom_panel <- plot_grid(d, e, labels = c("D", "E"), label_size = 12, nrow = 2)
+plot_grid(top_panel, bottom_panel, nrow = 2, rel_widths = c(.5, 3)) +
+  ggsave("results/figures/figure_3.pdf", width = 6.875, height = 9)+
+  ggsave("submission/figure_3.pdf", width = 6.875, height = 9)

### 2021-04-05 code/figure_5.R
Add figure 5 to submission folder

@@ -28,7 +28,8 @@ d <- ggdraw() + draw_image("results/figures/post_CDI_PEG_richness_overtime_stool
 e <- ggdraw() + draw_image("results/figures/post_CDI_PEG_treatment_legend.png")
 fig <- image_graph(width = 400, height = 400, res = 96)
 plot_grid(a, e, b, c, d, labels = c("A", "", "B", "C", "D"), label_size = 12, ncol=1, rel_heights = c(.8,.2, 1,1,1))+
-  ggsave("results/figures/figure_5.pdf", width=4.6, height=7.5)
+  ggsave("results/figures/figure_5.pdf", width=4.6, height=7.5)+
+  ggsave("submission/figure_5.pdf", width=4.6, height=7.5)
   
 #Figure showing pcoa and genera heat map over time
 d <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png")
@@ -45,4 +46,5 @@ g <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend_vert.png")
 pcoa1 <- plot_grid(f,g, label_size = 12, nrow = 1, rel_heights = c(1,.8), rel_widths = c(1,.8))
 
 plot_grid(pcoa1, heatmap, labels = c("A", "B"), label_size = 12, nrow = 2)+
-  ggsave("results/figures/figure_5_16S.pdf", width = 6, height = 6.7)
+  ggsave("results/figures/figure_5_16S.pdf", width = 6, height = 6.7)+
+  ggsave("submission/figure_5_16S.pdf", width = 6, height = 6.7)

### 2021-04-05 submission/manuscript.Rmd
Add outline and figures to manuscript draft

@@
 ## Abstract
-(Modify depending on target journal) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome.
+(Modify depending on target journal, currently abstract submitted to World Microbe Forum) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome.
 However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear.
@@
 
-## Results and Discussion
+* Medications have been shown to alter the intestinal microbiota, but how this may influence susceptibility to enteric infections is less clear.
+
+*  The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 (Miralax/Nulytely/Golytely)  that is also commonly used as bowel preparation for colonoscopies (also one of the possible administration routes used for fecal microbiota transplantation, a therapy for recurrent C.
+difficile infection) has been shown to influence the intestinal microbiota and mucosal barrier.
+
+* Review of previous mouse studies with PEG
+
+* 2 main questions:
+  + Does PEG 3350 prolong colonization and result in more severe C. difficile infections?
+  + How does PEG 3350 influence C. difficile clearance and fecal microbiota transplant (FMT) in mice?
+
+## Results
+
+* Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.
+
+* Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.
+
+* Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.
+
+* Figure 3. 5-day PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.
+
+* Figure 4. 1-day PEG treatment renders mice susceptible to transient *C. difficile* colonization.
+
+* Figure 5. 1-day PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.
+
+* Figure 6. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.
+
+* Figure S2. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization.
+  + Ex. *Muribaculum intestinale*.
+
+* Figure 7. Schematic summarizing findings.
+
+##Discussion
+
+* Summary of major findings
+
+* Discussion of prolonged persistence. C. difficile sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent [a[ers]]
+
+* Discuss why we might not have observed more severe histology in PEG mice relative to clindamycin-treated mice
+  + Antibiotics may also impact mucus layer
+  + Strain of bacteria used
+
+* Protective bacteria missing in PEG-treated mice
+
+* Discuss what these findings might mean for human patients
+  + What's known regarding laxatives and susceptibility to CDIs
+  + Relevance to human FMTs? Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.
 
@@
 \includegraphics{figure_1.pdf}
+
 **Figure 1.
@@
 \includegraphics{figure_5.pdf}
+\includegraphics{figure_5_16S.pdf}
 **Figure 5. 1-day PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A. 

### 2021-04-05 submission/manuscript.tex
Add outline and figures to manuscript draft

@@
 
-(Modify depending on target journal) Antibiotics are a major risk factor
-for \emph{Clostridioides difficile} infections (CDIs) because of their
+(Modify depending on target journal, currently abstract submitted to
+World Microbe Forum) Antibiotics are a major risk factor for
+\emph{Clostridioides difficile} infections (CDIs) because of their
 impact on the intestinal microbiome. However, non-antibiotic medications
@@
 
-\hypertarget{results-and-discussion}{%
-\subsection{Results and Discussion}\label{results-and-discussion}}
+\begin{itemize}
+\item
+  Medications have been shown to alter the intestinal microbiota, but
+  how this may influence susceptibility to enteric infections is less
+  clear.
+\item
+  The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350
+  (Miralax/Nulytely/Golytely) that is also commonly used as bowel
+  preparation for colonoscopies (also one of the possible administration
+  routes used for fecal microbiota transplantation, a therapy for
+  recurrent C. difficile infection) has been shown to influence the
+  intestinal microbiota and mucosal barrier.
+\item
+  Review of previous mouse studies with PEG
+\item
+  2 main questions:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Does PEG 3350 prolong colonization and result in more severe C.
+    difficile infections?
+  \item
+    How does PEG 3350 influence C. difficile clearance and fecal
+    microbiota transplant (FMT) in mice?
+  \end{itemize}
+\end{itemize}
+
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
+
+\begin{itemize}
+\item
+  Figure 1. 5-day PEG treatment prolongs susceptibility and mice become
+  persistently colonized with \emph{C. difficile}.
+\item
+  Figure 2. 5-day PEG treatment disrupts the stool microbiota for a
+  longer amount of time compared to clindamycin-treated mice.
+\item
+  Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota
+  dynamics post-infection.
+\item
+  Figure 3. 5-day PEG treatment does not result in more severe CDIs,
+  although mucosal microbiota is altered.
+\item
+  Figure 4. 1-day PEG treatment renders mice susceptible to transient
+  \emph{C. difficile} colonization.
+\item
+  Figure 5. 1-day PEG treatment post C. difficile challenge prolongs
+  colonization regardless of whether an FMT is also administered.
+\item
+  Figure 6. Specific microbiota features associated with prolonged
+  \emph{C. difficile} colonization in PEG treated mice.
+\item
+  Figure S2. Specific OTUs associated with clearance that are mostly
+  absent in mice with prolonged \emph{C. difficile} colonization.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Ex. \emph{Muribaculum intestinale}.
+  \end{itemize}
+\item
+  Figure 7. Schematic summarizing findings.
+\end{itemize}
+
+\#\#Discussion
+
+\begin{itemize}
+\item
+  Summary of major findings
+\item
+  Discussion of prolonged persistence. C. difficile sequences detected
+  in tissue samples. Association with mucin-degrading bacteria suggested
+  by recent {[}a{[}ers{]}{]}
+\item
+  Discuss why we might not have observed more severe histology in PEG
+  mice relative to clindamycin-treated mice
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Antibiotics may also impact mucus layer
+  \item
+    Strain of bacteria used
+  \end{itemize}
+\item
+  Protective bacteria missing in PEG-treated mice
+\item
+  Discuss what these findings might mean for human patients
+
+  \begin{itemize}
+  \tightlist
+  \item
+    What's known regarding laxatives and susceptibility to CDIs
+  \item
+    Relevance to human FMTs? Unclear what the best administration route
+    is because there have been no studies designed to evaluate the best
+    administration route for FMTs.
+  \end{itemize}
+\end{itemize}
 
@@
 
-\includegraphics{figure_1.pdf} \textbf{Figure 1. Chronic PEG treatment
+\includegraphics{figure_1.pdf} \textbf{Figure 1. 5-day PEG treatment
 prolongs susceptibility and mice become persistently colonized with
 \emph{C. difficile}.} A. Setup of the experimental timeline for subset
-of experiments with 5-day PEG treated mice. B. \emph{C. difficile}
+of experiments with 5-day PEG treated mice. B. Weight change from
+baseline weight in groups after treatment with PEG and/or clindamycin,
+followed by \emph{C. difficile} challenge. C. \emph{C. difficile}
 CFU/gram stool measured over time (N =
@@
 each mouse due to stool sampling difficulties (particularly the day the
-mcie came off of the PEG treatment) or early deaths. Lines represent the
+mice came off of the PEG treatment) or early deaths. Lines represent the
 median for each source and circles represent individual mouse samples.
+Asterisks indicate timepoints where the weight change or CFU/g was
+significantly different between groups by the Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing multiple timepoints. \newpage
 
+\includegraphics{figure_2.pdf} \textbf{Figure 2. 5-day PEG treatment
+disrupts the stool microbiota for a longer amount of time compared to
+clindamycin-treated mice.} A. \newpage
+
+\includegraphics{figure_3.pdf} \textbf{Figure 3. 5-day PEG treatment
+does not result in more severe CDIs, although mucosal microbiota is
+altered.} A. \newpage
+
+\includegraphics{figure_4.pdf} \textbf{Figure 4. 1-day PEG treatment
+renders mice susceptible to transient \emph{C. difficile} colonization.}
+A. \newpage
+
+\includegraphics{figure_5.pdf} \includegraphics{figure_5_16S.pdf}
+\textbf{Figure 5. 1-day PEG treatment post C. difficile challenge
+prolongs colonization regardless of whether an FMT is also
+administered.} A. \newpage
+
+\includegraphics{figure_6.pdf} \textbf{Figure 6. Specific microbiota
+features associated with prolonged \emph{C. difficile} colonization in
+PEG treated mice.} A. \newpage
+
+\includegraphics{figure_7.pdf} \textbf{Figure 7. Schematic summarizing
+findings.} A. \newpage
+
+\includegraphics{figure_S1.pdf} \textbf{Figure S1. 5-day PEG treatment
+plus 10-day recovery mice microbiota dynamics post-infection.} A.
 \newpage
+
+\includegraphics{figure_S2.pdf} \textbf{Figure S2. Specific OTUs
+associated with clearance that are mostly absent in mice with prolonged
+\emph{C. difficile} colonization. Ex. \emph{Muribaculum intestinale}.}
+A. \newpage
 

### 2021-04-05 code/ml_results.R
Add area plot for cleared vs colonized mice

@@ -1,4 +1,5 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
+library(viridis)
 
 set.seed(19760620) #Same seed used for mothur analysis
 
@@ -135,7 +136,7 @@ shape_labels <- c("Clind.", "1-day", "5-day", "Post-CDI")
 
 interp_genera_d5_top_10 <- rf_top_feat[1:10]
 #Plot the top 10 features that were important to Day 5 model and 
-#using facet_wrap, highlight the Ogenera that correlate with colonization
+#using facet_wrap, highlight the genera that correlate with colonization
 top10_d5_model_taxa <- agg_genus_data %>% 
   filter(day == 5) %>% #Used d5 timepoint for ml input data
   filter(clearance_status_d10 %in% c("colonized", "cleared")) %>% #Remove samples we don't have clearance status d10 data for
@@ -144,7 +145,7 @@ top10_d5_model_taxa <- agg_genus_data %>%
   mutate(genus = fct_relevel(genus, interp_genera_d5_top_10)) %>% 
   mutate(agg_rel_abund = agg_rel_abund + 1/2000) %>% 
   group_by(clearance_status_d10, genus) %>% 
-  mutate(median=(median(agg_rel_abund))) %>% #create a column of median values for each group
+  mutate(median=(median(agg_rel_abund + 1/2000))) %>% #create a column of median values for each group
   ungroup() %>% 
   ggplot(aes(x=clearance_status_d10, y =agg_rel_abund, colour= clearance_status_d10))+
   scale_x_discrete(guide = guide_axis(n.dodge = 2))+
@@ -170,3 +171,54 @@ top10_d5_model_taxa <- agg_genus_data %>%
         axis.text.x = element_blank(),
         legend.position = "bottom")
 save_plot(filename = paste0("results/figures/ml_d5_top10_genus.png"), top10_d5_model_taxa, base_height = 5, base_width = 8)  
+
+#Create area plot of genera that vary beween mice that clear within 10 dpi and mice that have prolonged colonization
+interp_genera_d5_top_10 #List of genera to include in area plot
+#Labels for facet
+facet_labels <- c("cleared", "colonized")
+names(facet_labels) <- c("cleared", "colonized")
+#Create plot
+top10_d5_model_taxa_area_plot <- agg_genus_data %>% 
+  #Solve different baseline timepoints across groups by specifying baseline for each group
[…]

### 2021-04-05 code/ml_results.R
Change area plot to column instead of row

@@ -206,7 +206,7 @@ top10_d5_model_taxa_area_plot <- agg_genus_data %>%
        y="Relative abundance") +
 #To do: Figure out how to get scale to match other plots  
 #  scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100), limits = c(1/10900, 1))+
-  facet_wrap(~clearance_status_d10, labeller = labeller(clearance_status_d10 = facet_labels))+
+  facet_wrap(~clearance_status_d10, labeller = labeller(clearance_status_d10 = facet_labels), ncol = 1)+
   theme_classic()+
   theme(strip.background = element_blank(), #get rid of box around facet_wrap labels
         legend.text = element_text(face = "italic"), #Italicize genus name
@@ -220,5 +220,5 @@ save_plot("results/figures/ml_d5_top10_genus_area_legend.png", legend_area_plot,
 #Save area plot without legend
 top10_d5_model_taxa_area_plot <- top10_d5_model_taxa_area_plot+
   theme(legend.position = "none") #Remove legend
-save_plot(filename = paste0("results/figures/ml_d5_top10_genus_area.png"), top10_d5_model_taxa_area_plot, base_height = 5, base_width = 8)  
+save_plot(filename = paste0("results/figures/ml_d5_top10_genus_area.png"), top10_d5_model_taxa_area_plot, base_height = 5, base_width = 5)  
 

### 2021-04-08 submission/manuscript.Rmd
Rough draft of fig 4 legend, pval and R-sq still to be included

@@
 
-Sarah Tomkovich^1, Ana Taylor, Jacob Kingg, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^
+Sarah Tomkovich^1, Ana Taylor, Jacob King, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^
 
@@
 
+fig_4_pval = 1
+fig_4_rsq = 1
+
 ```
@@
 \includegraphics{figure_4.pdf}
-**Figure 4. 1-day PEG treatment renders mice susceptible to transient *C. difficile* colonization.** A. 
-\newpage
+**Figure 4.
+1-day PEG treatment renders mice susceptible to transient *C.
+difficile* colonization.** A.
+Setup of the experimental timeline for the 1-day PEG treated subset of mice.
+B.
+CFU/gram stool measured over time (N = 6 mice per timepoint) via several dilutions.
+The black dotted line represents the limit of detection for the first serial dilution.
+Asterisks indicate timepoints where the CFU/gram was siginificantly different between groups using the Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple timepoints.
+C.
+Principle Coordinate Analysis plot of the groups over time with the alpha representating the same time scale as in panel D (P = `r fig_4_pval` R^2^ = `r fig_4_rsq`).
+D.
+Shannon Diveristy Index of the groups over time.
+Only days with samples from all groups are shown.
+Samples for some mice were difficult to obtain
+due to the laxative treatment.
+The alpha scale follows accordingly with the timeline.
+E.
+Line plots of relative percent abundance of selected genera over time.
+Only days with samples from all groups shown.
+The gray line represents the limit of detection.
 

### 2021-04-08 submission/manuscript.tex
Demo rmarkdown with Ana
Rough draft of fig 4 legend, pval and R-sq still to be included

@@
 
-Sarah Tomkovich\^{}1, Ana Taylor, Jacob Kingg, Joanna Colovas, Lucas
+Sarah Tomkovich\^{}1, Ana Taylor, Jacob King, Joanna Colovas, Lucas
 Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid
@@
 
-\includegraphics{figure_1.pdf} \textbf{Figure 1. 5-day PEG treatment
-prolongs susceptibility and mice become persistently colonized with
-\emph{C. difficile}.} A. Setup of the experimental timeline for subset
-of experiments with 5-day PEG treated mice. B. Weight change from
-baseline weight in groups after treatment with PEG and/or clindamycin,
-followed by \emph{C. difficile} challenge. C. \emph{C. difficile}
-CFU/gram stool measured over time (N =
+\includegraphics{figure_1.pdf}
+
+\textbf{Figure 1. 5-day PEG treatment prolongs susceptibility and mice
+become persistently colonized with \emph{C. difficile}.} A. Setup of the
+experimental timeline for subset of experiments with 5-day PEG treated
+mice. B. Weight change from baseline weight in groups after treatment
+with PEG and/or clindamycin, followed by \emph{C. difficile} challenge.
+C. \emph{C. difficile} CFU/gram stool measured over time (N =
 4-\texttt{(insert\ variable\ name)} mice per timepoint) via serial
@@
 renders mice susceptible to transient \emph{C. difficile} colonization.}
-A. \newpage
+A. Setup of the experimental timeline for the 1-day PEG treated subset
+of mice. B. CFU/gram stool measured over time (N = 6 mice per timepoint)
+via several dilutions. The black dotted line represents the limit of
+detection for the first serial dilution. Asterisks indicate timepoints
+where the CFU/gram was siginificantly different between groups using the
+Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple
+timepoints. C. Principle Coordinate Analysis plot of the groups over
+time with the alpha representating the same time scale as in panel D (P
+= 1 R\textsuperscript{2} = 1). D. Shannon Diveristy Index of the groups
+over time. Only days with samples from all groups are shown. Samples for
+some mice were difficult to obtain due to the laxative treatment. The
+alpha scale follows accordingly with the timeline. E. Line plots of
+relative percent abundance of selected genera over time. Only days with
+samples from all groups shown. The gray line represents the limit of
+detection.
 

### 2021-04-08 submission/manuscript.Rmd
Add to introduction

--- 2021-04-05 submission/manuscript.Rmd
@@
 
-Sarah Tomkovich^1, Ana Taylor, Jacob Kingg, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^
+Sarah Tomkovich^1^, Ana Taylor^1^, Jacob Kingg, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^
 
@@
 
-2\. The Unit for Laboratory Animal Medicine, University off Michigan, Ann Arbor, MI, USA
+2\. The Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI, USA
 
@@
 
-* Medications have been shown to alter the intestinal microbiota, but how this may influence susceptibility to enteric infections is less clear.
-
-*  The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 (Miralax/Nulytely/Golytely)  that is also commonly used as bowel preparation for colonoscopies (also one of the possible administration routes used for fecal microbiota transplantation, a therapy for recurrent C.
-difficile infection) has been shown to influence the intestinal microbiota and mucosal barrier.
-
-* Review of previous mouse studies with PEG
+Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota.
+However, antibiotics are not the only types of medications that disrupt the microbiota.
+Although, other medications such as proton pump inhibitors have been implicated as risk factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Tomkovich2019].
+Many of the non-antibiotic medications associated with CDIs are known to modulate intestinal motility, which in turn also strongly impacts microbiota composition and function.
+Interestingly our group has previously demonstrated that there are similar microbiota features in stool samples from CDI patients and patients with diarrhea that tested negative for *C.
+difficile*  with both groups having lower alpha diversity than non-diarrheal control samples [@Schubert2015].
+These results led to our hypothesis that bacterial communities from patients experiencing diarrheal controls are susceptible, but have not been exposed to *C.
+difficile* spores.
+To examine if diarrhea alone is sufficient to disrupt microbiota colonization resistance, we turned to osmotic laxatives, which increase intestinal motility and can lead to diarrhea.
+The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies.
+Interestingly, previous studies has shown that treating mice with PEG alone rendered the mice susceptible to *C.
+difficile *infection, altered microbiota composition, reduced acetate and butyrate and altered the mucus barrier [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020].
+The mucus barrier is thought to mediate protection from *C.
+difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C.
+difficile* (Ref).
+PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT).
+For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure.
+Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a case study with 2 pediatric patients where the authors suggested the laxative may help flush *C.
+difficile* spores and toxins from the intestine [@Liacouras1996].
 
@@
 ## Results
+
+**Laxative treatment alone leads to prolonged *C.
+difficile* colonization in mice.** We compared PEG-treated mice to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C.
+difficile*, with mice typically clearing C.
+difficile within 10 days post-infection [@Tomkovich2019; @Tomkovich2020].
+All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig.
+1A).
+After PEG and/or antibiotic treatment all mice were challenged with 10^3^ *C.
+difficile* 630 spores.
 
@@
 
-##Discussion
+## Discussion
 
@@
 
-* Discussion of prolonged persistence. C. difficile sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent [a[ers]]
+* Discussion of prolonged persistence. C. difficile sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent papers.
 

### 2021-04-08 submission/manuscript.Rmd

@@
 
-Sarah Tomkovich^1^, Ana Taylor^1^, Jacob Kingg, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^
+Sarah Tomkovich^1, Ana Taylor, Jacob King, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^$
 

### 2021-04-08 submission/manuscript.tex
Test updated manuscript after merging conflict

@@
 Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid
-L. Bergin\^{}2, Patrick D. Schloss\textsuperscript{1\(\dagger\)}
+L. Bergin\^{}2, Patrick D. Schloss\textsuperscript{1\(\dagger\)}\$
 
@@
 
-2. The Unit for Laboratory Animal Medicine, University off Michigan, Ann
+2. The Unit for Laboratory Animal Medicine, University of Michigan, Ann
 Arbor, MI, USA
@@
 
+Antibiotics are a major risk factor for \emph{Clostridioides difficile}
+infections (CDIs) because of their impact on the microbiota. However,
+antibiotics are not the only types of medications that disrupt the
+microbiota. Although, other medications such as proton pump inhibitors
+have been implicated as risk factors for CDIs through epidemiological
+studies, whether the association is due to their impact on the
+microbiome is still unclear (1). Many of the non-antibiotic medications
+associated with CDIs are known to modulate intestinal motility, which in
+turn also strongly impacts microbiota composition and function.\\
+Interestingly our group has previously demonstrated that there are
+similar microbiota features in stool samples from CDI patients and
+patients with diarrhea that tested negative for \emph{C. difficile} with
+both groups having lower alpha diversity than non-diarrheal control
+samples (2). These results led to our hypothesis that bacterial
+communities from patients experiencing diarrheal controls are
+susceptible, but have not been exposed to \emph{C. difficile} spores. To
+examine if diarrhea alone is sufficient to disrupt microbiota
+colonization resistance, we turned to osmotic laxatives, which increase
+intestinal motility and can lead to diarrhea. The ubiquitous osmotic
+laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely,
+and Golytely and is also commonly used as bowel preparation for
+colonoscopies. Interestingly, previous studies has shown that treating
+mice with PEG alone rendered the mice susceptible to \emph{C. difficile
+}infection, altered microbiota composition, reduced acetate and butyrate
+and altered the mucus barrier (3--6). The mucus barrier is thought to
+mediate protection from \emph{C. difficile} infections by protecting
+intestinal epithelial cells from the toxins produced by \emph{C.
+difficile} (Ref). PEG is also relevant in the context of treating
+recurrent CDIs via fecal microbiota transplant (FMT). For FMTs that are
+delivered via colonoscopy, patients typically undergo bowel preparation
+by taking an osmotic laxative prior to the procedure. Many of the FMT
+studies to date rationalize the use of laxatives (Ref) based on a case
+study with 2 pediatric patients where the authors suggested the laxative
+may help flush \emph{C. difficile} spores and toxins from the intestine
+(7).
+
 \begin{itemize}
-\item
-  Medications have been shown to alter the intestinal microbiota, but
-  how this may influence susceptibility to enteric infections is less
-  clear.
-\item
-  The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350
-  (Miralax/Nulytely/Golytely) that is also commonly used as bowel
-  preparation for colonoscopies (also one of the possible administration
-  routes used for fecal microbiota transplantation, a therapy for
-  recurrent C. difficile infection) has been shown to influence the
-  intestinal microbiota and mucosal barrier.
-\item
-  Review of previous mouse studies with PEG
+\tightlist
 \item
@@
 \subsection{Results}\label{results}}
+
+\textbf{Laxative treatment alone leads to prolonged \emph{C. difficile}
+colonization in mice.} We compared PEG-treated mice to our standard 10
+mg/kg clindamycin treatment, which temporarily renders the mice
+susceptible to \emph{C. difficile}, with mice typically clearing C.
+difficile within 10 days post-infection (1, 8). All PEG-treated mice
+were administered a 15\% PEG solution in the drinking water for 5-days,
+one group was also treated with clindamycin, and one group was allowed
+to recover for 10 days prior to challenge (Fig. 1A). After PEG and/or
+antibiotic treatment all mice were challenged with 10\textsuperscript{3}
+\emph{C. difficile} 630 spores.
 
@@
 
-\#\#Discussion
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 
@@
   in tissue samples. Association with mucin-degrading bacteria suggested
-  by recent {[}a{[}ers{]}{]}
+  by recent papers.
 \item
@@
 \hypertarget{references}{%
-\subsection{References}\label{references}}
+\subsection*{References}\label{references}}
+\addcontentsline{toc}{subsection}{References}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Tomkovich2019}{}%
+1. \textbf{Tomkovich S}, \textbf{Lesniak NA}, \textbf{Li Y},
+\textbf{Bishop L}, \textbf{Fitzgerald MJ}, \textbf{Schloss PD}. 2019.
+The proton pump inhibitor omeprazole does not promote
+\emph{Clostridioides difficile} colonization in a murine model. mSphere
+\textbf{4}.
+doi:\href{https://doi.org/10.1128/msphere.00693-19}{10.1128/msphere.00693-19}.
+
+\leavevmode\hypertarget{ref-Schubert2015}{}%
+2. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+Antibiotic-induced alterations of the murine gut microbiota and
+subsequent effects on colonization resistance against \emph{Clostridium
+difficile}. mBio \textbf{6}.
+doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+
+\leavevmode\hypertarget{ref-Kashyap2013}{}%
+3. \textbf{Kashyap PC}, \textbf{Marcobal A}, \textbf{Ursell LK},
+\textbf{Larauche M}, \textbf{Duboc H}, \textbf{Earle KA},
+\textbf{Sonnenburg ED}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
+\textbf{Million M}, \textbf{Tache Y}, \textbf{Pasricha PJ},
+\textbf{Knight R}, \textbf{Farrugia G}, \textbf{Sonnenburg JL}. 2013.
+Complex interactions among diet, gastrointestinal transit, and gut
+microbiota in humanized mice. Gastroenterology \textbf{144}:967--977.
+doi:\href{https://doi.org/10.1053/j.gastro.2013.01.047}{10.1053/j.gastro.2013.01.047}.
+
+\leavevmode\hypertarget{ref-Ferreyra2014}{}%
+4. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
+\textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
+Gut microbiota-produced succinate promotes c.~difficile infection after
+antibiotic treatment or motility disturbance. Cell Host \& Microbe
+\textbf{16}:770--777.
+doi:\href{https://doi.org/10.1016/j.chom.2014.11.003}{10.1016/j.chom.2014.11.003}.
+
+\leavevmode\hypertarget{ref-Tropini2018}{}%
+5. \textbf{Tropini C}, \textbf{Moss EL}, \textbf{Merrill BD}, \textbf{Ng
+KM}, \textbf{Higginbottom SK}, \textbf{Casavant EP}, \textbf{Gonzalez
+CG}, \textbf{Fremin B}, \textbf{Bouley DM}, \textbf{Elias JE},
+\textbf{Bhatt AS}, \textbf{Huang KC}, \textbf{Sonnenburg JL}. 2018.
+Transient osmotic perturbation causes long-term alteration to the gut
+microbiota. Cell \textbf{173}:1742--1754.e17.
+doi:\href{https://doi.org/10.1016/j.cell.2018.05.008}{10.1016/j.cell.2018.05.008}.
+
+\leavevmode\hypertarget{ref-VanInsberghe2020}{}%
+6. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
+\textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
+Diarrhoeal events can trigger long-term clostridium difficile
+colonization with recurrent blooms. Nature Microbiology
+\textbf{5}:642--650.
+doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.
+
+\leavevmode\hypertarget{ref-Liacouras1996}{}%
+7. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
+irrigation as an adjunct to the treatment of chronic, relapsing
+clostridium difficile colitis. Journal of Clinical Gastroenterology
+\textbf{22}:186--189.
+doi:\href{https://doi.org/10.1097/00004836-199604000-00007}{10.1097/00004836-199604000-00007}.
+
+\leavevmode\hypertarget{ref-Tomkovich2020}{}%
+8. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
+antibiotic perturbation influence clostridioides difficile clearance in
+mice. mSphere \textbf{5}.
+doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
 

### 2021-04-08 submission/manuscript.Rmd
Add R-sq values to fig 4 legend and update readme with here package

@@
 library(knitr)
+library(here)
 
@@
 
-fig_4_pval = 1
-fig_4_rsq = 1
-
 ```
@@
 
+``` {r figure_4_permnova}
+one_day_permnova_results <- read_tsv(here("data/process/1_day_PEG_permanova.tsv"))
+one_day_top_2_contrib <- one_day_permnova_results %>% arrange(desc(R2)) %>%
+  top_n(3, R2)
+one_day_top_contrib_name = (one_day_top_2_contrib[2,1])
+one_day_sec_contrib_name = (one_day_top_2_contrib[3,1])
+top_R2 = one_day_top_2_contrib[2,6]
+sec_R2 = one_day_top_2_contrib[3,6]
+```
 \includegraphics{figure_4.pdf}
@@
 C.
-Principle Coordinate Analysis plot of the groups over time with the alpha representating the same time scale as in panel D (P = `r fig_4_pval` R^2^ = `r fig_4_rsq`).
+Principle Coordinate Analysis plot of the groups over time with the alpha representating the same time scale as in panel D (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`).
 D.

### 2021-04-08 submission/manuscript.tex
Add R-sq values to fig 4 legend and update readme with here package

@@
 timepoints. C. Principle Coordinate Analysis plot of the groups over
-time with the alpha representating the same time scale as in panel D (P
-= 1 R\textsuperscript{2} = 1). D. Shannon Diveristy Index of the groups
-over time. Only days with samples from all groups are shown. Samples for
-some mice were difficult to obtain due to the laxative treatment. The
-alpha scale follows accordingly with the timeline. E. Line plots of
-relative percent abundance of selected genera over time. Only days with
-samples from all groups shown. The gray line represents the limit of
-detection.
+time with the alpha representating the same time scale as in panel D
+(day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} = 0.19).
+D. Shannon Diveristy Index of the groups over time. Only days with
+samples from all groups are shown. Samples for some mice were difficult
+to obtain due to the laxative treatment. The alpha scale follows
+accordingly with the timeline. E. Line plots of relative percent
+abundance of selected genera over time. Only days with samples from all
+groups shown. The gray line represents the limit of detection.
 

### 2021-04-09 submission/manuscript.tex
Fix titlepage superscripts

@@
 
-Sarah Tomkovich\^{}1, Ana Taylor, Jacob King, Joanna Colovas, Lucas
-Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid
-L. Bergin\^{}2, Patrick D. Schloss\textsuperscript{1\(\dagger\)}\$
+Sarah Tomkovich\textsuperscript{1}, Ana Taylor\textsuperscript{1}, Jacob
+King\textsuperscript{1}, Joanna Colovas\textsuperscript{1}, Lucas
+Bishop\textsuperscript{1}, Kathryn McBride\textsuperscript{1}, Sonya
+Royzenblat\textsuperscript{1}, Nicholas A. Lesniak\textsuperscript{1},
+Ingrid L. Bergin\textsuperscript{2}, Patrick D.
+Schloss\textsuperscript{1\(\dagger\)}
 

### 2021-04-09 submission/manuscript.Rmd
Fix titlepage superscripts
Additions to the introduction

@@
 
-Sarah Tomkovich^1, Ana Taylor, Jacob King, Joanna Colovas, Lucas Bishop, Kathryn McBride, Sonya Royzenblat, Nicholas A. Lesniak, Ingrid L. Bergin^2, Patrick D. Schloss^1$\dagger$^$
+Sarah Tomkovich^1^, Ana Taylor^1^, Jacob King^1^, Joanna Colovas^1^, Lucas Bishop^1^, Kathryn McBride^1^, Sonya Royzenblat^1^, Nicholas A.
+Lesniak^1^, Ingrid L.
+Bergin^2^, Patrick D.
+Schloss^1$\dagger$^
 
@@
 
-Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota.
-However, antibiotics are not the only types of medications that disrupt the microbiota.
-Although, other medications such as proton pump inhibitors have been implicated as risk factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Tomkovich2019].
-Many of the non-antibiotic medications associated with CDIs are known to modulate intestinal motility, which in turn also strongly impacts microbiota composition and function.
-Interestingly our group has previously demonstrated that there are similar microbiota features in stool samples from CDI patients and patients with diarrhea that tested negative for *C.
-difficile*  with both groups having lower alpha diversity than non-diarrheal control samples [@Schubert2015].
-These results led to our hypothesis that bacterial communities from patients experiencing diarrheal controls are susceptible, but have not been exposed to *C.
+Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota [@Britton2014].
+However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020].
+Although, other medications have been implicated as risk factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].
+Many of the non-antibiotic medications associated with CDIs are known to modulate intestinal motility, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020].
+
+Stool consistency often serves as an approximation of intestinal motility.
+Our group has shown that when *C.
+difficile* negative controls are separated into two groups based on stool consistency, there are microbiota features such as alpha diversity that overlap between samples from CDI patients and control patients with diarrhea [@Schubert2015].
+These results led to our hypothesis that bacterial communities from patients experiencing diarrhea are susceptible, but may not have been exposed to *C.
 difficile* spores.
-To examine if diarrhea alone is sufficient to disrupt microbiota colonization resistance, we turned to osmotic laxatives, which increase intestinal motility and can lead to diarrhea.
+
+Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupts the intestinal microbiota in humans [@Nagata2019].
 The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies.
-Interestingly, previous studies has shown that treating mice with PEG alone rendered the mice susceptible to *C.
-difficile *infection, altered microbiota composition, reduced acetate and butyrate and altered the mucus barrier [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020].
+Interestingly, previous studies have shown that treating mice with PEG alone rendered the mice susceptible to *C.
+difficile* infection, altered microbiota composition, reduced acetate and butyrate and altered the mucus barrier [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020].
 The mucus barrier is thought to mediate protection from *C.
@@
 difficile* (Ref).
-PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT).
+However, whether laxative results in more severe CDIs in mice is unclear.
+
+Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization.
 For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure.
-Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a case study with 2 pediatric patients where the authors suggested the laxative may help flush *C.
+Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested the laxative may help flush *C.
 difficile* spores and toxins from the intestine [@Liacouras1996].
 
-* 2 main questions:
-  + Does PEG 3350 prolong colonization and result in more severe C. difficile infections?
-  + How does PEG 3350 influence C. difficile clearance and fecal microbiota transplant (FMT) in mice?
+In the past, our group has used C57BL6 mice to characterize how antibiotics disrupt the microbiota and influence *C.
+difficile* susceptibility and clearance [ref].
+Although, two groups have now shown PEG treatment alone renders mice susceptible to *C.
+difficile*, these studies have raised additional questions that should be addressed given their relevance to CDIs.
+Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in sustained *C.
+difficile* colonization, and if PEG treatment post-CDI can promote *C.
+difficile* clearance.
 

### 2021-04-09 submission/manuscript.tex
Additions to the introduction

@@
 Antibiotics are a major risk factor for \emph{Clostridioides difficile}
-infections (CDIs) because of their impact on the microbiota. However,
-antibiotics are not the only types of medications that disrupt the
-microbiota. Although, other medications such as proton pump inhibitors
-have been implicated as risk factors for CDIs through epidemiological
-studies, whether the association is due to their impact on the
-microbiome is still unclear (1). Many of the non-antibiotic medications
-associated with CDIs are known to modulate intestinal motility, which in
-turn also strongly impacts microbiota composition and function.\\
-Interestingly our group has previously demonstrated that there are
-similar microbiota features in stool samples from CDI patients and
-patients with diarrhea that tested negative for \emph{C. difficile} with
-both groups having lower alpha diversity than non-diarrheal control
-samples (2). These results led to our hypothesis that bacterial
-communities from patients experiencing diarrheal controls are
-susceptible, but have not been exposed to \emph{C. difficile} spores. To
-examine if diarrhea alone is sufficient to disrupt microbiota
-colonization resistance, we turned to osmotic laxatives, which increase
-intestinal motility and can lead to diarrhea. The ubiquitous osmotic
-laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely,
-and Golytely and is also commonly used as bowel preparation for
-colonoscopies. Interestingly, previous studies has shown that treating
-mice with PEG alone rendered the mice susceptible to \emph{C. difficile
-}infection, altered microbiota composition, reduced acetate and butyrate
-and altered the mucus barrier (3--6). The mucus barrier is thought to
-mediate protection from \emph{C. difficile} infections by protecting
-intestinal epithelial cells from the toxins produced by \emph{C.
-difficile} (Ref). PEG is also relevant in the context of treating
-recurrent CDIs via fecal microbiota transplant (FMT). For FMTs that are
-delivered via colonoscopy, patients typically undergo bowel preparation
-by taking an osmotic laxative prior to the procedure. Many of the FMT
-studies to date rationalize the use of laxatives (Ref) based on a case
-study with 2 pediatric patients where the authors suggested the laxative
-may help flush \emph{C. difficile} spores and toxins from the intestine
-(7).
+infections (CDIs) because of their impact on the microbiota (1).
+However, antibiotics are not the only types of medications that disrupt
+the microbiota (2--4). Although, other medications have been implicated
+as risk factors for CDIs through epidemiological studies, whether the
+association is due to their impact on the microbiome is still unclear
+(5--9). Many of the non-antibiotic medications associated with CDIs are
+known to modulate intestinal motility, which in turn also strongly
+impacts microbiota composition and function (10, 11).
 
-\begin{itemize}
-\tightlist
-\item
-  2 main questions:
+Stool consistency often serves as an approximation of intestinal
+motility. Our group has shown that when \emph{C. difficile} negative
+controls are separated into two groups based on stool consistency, there
+are microbiota features such as alpha diversity that overlap between
+samples from CDI patients and control patients with diarrhea (12). These
+results led to our hypothesis that bacterial communities from patients
+experiencing diarrhea are susceptible, but may not have been exposed to
+\emph{C. difficile} spores.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Does PEG 3350 prolong colonization and result in more severe C.
-    difficile infections?
-  \item
-    How does PEG 3350 influence C. difficile clearance and fecal
-    microbiota transplant (FMT) in mice?
-  \end{itemize}
-\end{itemize}
+Osmotic laxatives can lead to diarrhea depending on the administered
+dose and temporarily disrupts the intestinal microbiota in humans (13).
+The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
+in Miralax, Nulytely, and Golytely and is also commonly used as bowel
+preparation for colonoscopies. Interestingly, previous studies have
+shown that treating mice with PEG alone rendered the mice susceptible to
+\emph{C. difficile} infection, altered microbiota composition, reduced
+acetate and butyrate and altered the mucus barrier (14--17). The mucus
+barrier is thought to mediate protection from \emph{C. difficile}
+infections by protecting intestinal epithelial cells from the toxins
+produced by \emph{C. difficile} (Ref). However, whether laxative results
+in more severe CDIs in mice is unclear.
+
+Beyond susceptibility, PEG is also relevant in the context of treating
+recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
+microbiota is administered to the patient to restore colonization. For
+FMTs that are delivered via colonoscopy, patients typically undergo
+bowel preparation by taking an osmotic laxative prior to the procedure.
+Many of the FMT studies to date rationalize the use of laxatives (Ref)
+based on a 1996 case study with 2 pediatric patients where the authors
+suggested the laxative may help flush \emph{C. difficile} spores and
+toxins from the intestine (18).
+
+In the past, our group has used C57BL6 mice to characterize how
+antibiotics disrupt the microbiota and influence \emph{C. difficile}
+susceptibility and clearance {[}ref{]}. Although, two groups have now
+shown PEG treatment alone renders mice susceptible to \emph{C.
+difficile}, these studies have raised additional questions that should
+be addressed given their relevance to CDIs. Here, we used our C57BL/6
+clindamycin model as a control group to characterize how long
+PEG-treated mice remain susceptible, whether PEG treatment results in
+sustained \emph{C. difficile} colonization, and if PEG treatment
+post-CDI can promote \emph{C. difficile} clearance.
 
@@
 susceptible to \emph{C. difficile}, with mice typically clearing C.
-difficile within 10 days post-infection (1, 8). All PEG-treated mice
+difficile within 10 days post-infection (9, 19). All PEG-treated mice
 were administered a 15\% PEG solution in the drinking water for 5-days,
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Britton2014}{}%
+1. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
+microbiota in resistance to colonization by clostridium difficile.
+Gastroenterology \textbf{146}:1547--1553.
+doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
+
+\leavevmode\hypertarget{ref-Maier2018}{}%
+2. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
+\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
+\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
+\textbf{Mori H}, \textbf{Patil KR}, \textbf{Bork P}, \textbf{Typas A}.
+2018. Extensive impact of non-antibiotic drugs on human gut bacteria.
+Nature \textbf{555}:623--628.
+doi:\href{https://doi.org/10.1038/nature25979}{10.1038/nature25979}.
+
+\leavevmode\hypertarget{ref-LeBastard2017}{}%
+3. \textbf{Bastard QL}, \textbf{Al-Ghalith GA}, \textbf{Grégoire M},
+\textbf{Chapelet G}, \textbf{Javaudin F}, \textbf{Dailly E},
+\textbf{Batard E}, \textbf{Knights D}, \textbf{Montassier E}. 2017.
+Systematic review: Human gut dysbiosis induced by non-antibiotic
+prescription medications. Alimentary Pharmacology \& Therapeutics
+\textbf{47}:332--345.
+doi:\href{https://doi.org/10.1111/apt.14451}{10.1111/apt.14451}.
+
+\leavevmode\hypertarget{ref-VichVila2020}{}%
+4. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
+T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
+\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
+\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
+\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
+composition and metabolic function of the gut microbiota. Nature
+Communications \textbf{11}.
+doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
+
+\leavevmode\hypertarget{ref-Oh2018}{}%
+5. \textbf{Oh J}, \textbf{Makar M}, \textbf{Fusco C}, \textbf{McCaffrey
+R}, \textbf{Rao K}, \textbf{Ryan EE}, \textbf{Washer L}, \textbf{West
+LR}, \textbf{Young VB}, \textbf{Guttag J}, \textbf{Hooper DC},
+\textbf{Shenoy ES}, \textbf{Wiens J}. 2018. A generalizable, data-driven
+approach to predict daily risk ofClostridium difficileInfection at two
+large academic health centers. Infection Control \& Hospital
+Epidemiology \textbf{39}:425--433.
+doi:\href{https://doi.org/10.1017/ice.2018.16}{10.1017/ice.2018.16}.
+
+\leavevmode\hypertarget{ref-Mora2012}{}%
+6. \textbf{Mora AL}, \textbf{Salazar M}, \textbf{Pablo-Caeiro J},
+\textbf{Frost CP}, \textbf{Yadav Y}, \textbf{DuPont HL}, \textbf{Garey
+KW}. 2012. Moderate to high use of opioid analgesics are associated with
+an increased risk of clostridium difficile infection. The American
+Journal of the Medical Sciences \textbf{343}:277--280.
+doi:\href{https://doi.org/10.1097/maj.0b013e31822f42eb}{10.1097/maj.0b013e31822f42eb}.
+
+\leavevmode\hypertarget{ref-Nehra2018}{}%
+7. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
+\textbf{Nehra V}. 2018. Proton pump inhibitors: Review of emerging
+concerns. Mayo Clinic Proceedings \textbf{93}:240--246.
+doi:\href{https://doi.org/10.1016/j.mayocp.2017.10.022}{10.1016/j.mayocp.2017.10.022}.
+
+\leavevmode\hypertarget{ref-Krishna2013}{}%
+8. \textbf{Krishna SG}, \textbf{Zhao W}, \textbf{Apewokin SK},
+\textbf{Krishna K}, \textbf{Chepyala P}, \textbf{Anaissie EJ}. 2013.
+Risk factors, preemptive therapy, and antiperistaltic agents
+forClostridium difficileinfection in cancer patients. Transplant
+Infectious Disease n/a--n/a.
+doi:\href{https://doi.org/10.1111/tid.12112}{10.1111/tid.12112}.
+
 \leavevmode\hypertarget{ref-Tomkovich2019}{}%
-1. \textbf{Tomkovich S}, \textbf{Lesniak NA}, \textbf{Li Y},
+9. \textbf{Tomkovich S}, \textbf{Lesniak NA}, \textbf{Li Y},
 \textbf{Bishop L}, \textbf{Fitzgerald MJ}, \textbf{Schloss PD}. 2019.
@@
 
+\leavevmode\hypertarget{ref-Vandeputte2015}{}%
+10. \textbf{Vandeputte D}, \textbf{Falony G}, \textbf{Vieira-Silva S},
+\textbf{Tito RY}, \textbf{Joossens M}, \textbf{Raes J}. 2015. Stool
+consistency is strongly associated with gut microbiota richness and
+composition, enterotypes and bacterial growth rates. Gut
+\textbf{65}:57--62.
+doi:\href{https://doi.org/10.1136/gutjnl-2015-309618}{10.1136/gutjnl-2015-309618}.
+
+\leavevmode\hypertarget{ref-VujkovicCvijin2020}{}%
+11. \textbf{Vujkovic-Cvijin I}, \textbf{Sklar J}, \textbf{Jiang L},
+\textbf{Natarajan L}, \textbf{Knight R}, \textbf{Belkaid Y}. 2020. Host
+variables confound gut microbiota studies of human disease. Nature
+\textbf{587}:448--454.
+doi:\href{https://doi.org/10.1038/s41586-020-2881-9}{10.1038/s41586-020-2881-9}.
+
 \leavevmode\hypertarget{ref-Schubert2015}{}%
-2. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+12. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
 Antibiotic-induced alterations of the murine gut microbiota and
@@
 
+\leavevmode\hypertarget{ref-Nagata2019}{}%
+13. \textbf{Nagata N}, \textbf{Tohya M}, \textbf{Fukuda S}, \textbf{Suda
+W}, \textbf{Nishijima S}, \textbf{Takeuchi F}, \textbf{Ohsugi M},
+\textbf{Tsujimoto T}, \textbf{Nakamura T}, \textbf{Shimomura A},
+\textbf{Yanagisawa N}, \textbf{Hisada Y}, \textbf{Watanabe K},
+\textbf{Imbe K}, \textbf{Akiyama J}, \textbf{Mizokami M},
+\textbf{Miyoshi-Akiyama T}, \textbf{Uemura N}, \textbf{Hattori M}. 2019.
+Effects of bowel preparation on the human gut microbiome and metabolome.
+Scientific Reports \textbf{9}.
+doi:\href{https://doi.org/10.1038/s41598-019-40182-9}{10.1038/s41598-019-40182-9}.
+
 \leavevmode\hypertarget{ref-Kashyap2013}{}%
-3. \textbf{Kashyap PC}, \textbf{Marcobal A}, \textbf{Ursell LK},
+14. \textbf{Kashyap PC}, \textbf{Marcobal A}, \textbf{Ursell LK},
 \textbf{Larauche M}, \textbf{Duboc H}, \textbf{Earle KA},
@@
 \leavevmode\hypertarget{ref-Ferreyra2014}{}%
-4. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
+15. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
 \textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
@@
 \leavevmode\hypertarget{ref-Tropini2018}{}%
-5. \textbf{Tropini C}, \textbf{Moss EL}, \textbf{Merrill BD}, \textbf{Ng
-KM}, \textbf{Higginbottom SK}, \textbf{Casavant EP}, \textbf{Gonzalez
-CG}, \textbf{Fremin B}, \textbf{Bouley DM}, \textbf{Elias JE},
-\textbf{Bhatt AS}, \textbf{Huang KC}, \textbf{Sonnenburg JL}. 2018.
-Transient osmotic perturbation causes long-term alteration to the gut
-microbiota. Cell \textbf{173}:1742--1754.e17.
+16. \textbf{Tropini C}, \textbf{Moss EL}, \textbf{Merrill BD},
+\textbf{Ng KM}, \textbf{Higginbottom SK}, \textbf{Casavant EP},
+\textbf{Gonzalez CG}, \textbf{Fremin B}, \textbf{Bouley DM},
+\textbf{Elias JE}, \textbf{Bhatt AS}, \textbf{Huang KC},
+\textbf{Sonnenburg JL}. 2018. Transient osmotic perturbation causes
+long-term alteration to the gut microbiota. Cell
+\textbf{173}:1742--1754.e17.
 doi:\href{https://doi.org/10.1016/j.cell.2018.05.008}{10.1016/j.cell.2018.05.008}.
@@
 \leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-6. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
+17. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
 \textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
@@
 \leavevmode\hypertarget{ref-Liacouras1996}{}%
-7. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
+18. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
 irrigation as an adjunct to the treatment of chronic, relapsing
@@
 \leavevmode\hypertarget{ref-Tomkovich2020}{}%
-8. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+19. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
 \textbf{Schloss PD}. 2020. The initial gut microbiota and response to

### 2021-04-12 code/figure_5.R
Update draft figure with latest 16S plots

@@ -31,20 +31,13 @@ plot_grid(a, e, b, c, d, labels = c("A", "", "B", "C", "D"), label_size = 12, nc
   ggsave("results/figures/figure_5.pdf", width=4.6, height=7.5)+
   ggsave("submission/figure_5.pdf", width=4.6, height=7.5)
   
-#Figure showing pcoa and genera heat map over time
-d <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png")
-e <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend.png")
-pcoa <- plot_grid(d, e, label_size = 12, ncol = 1, rel_widths = c(1,1), rel_heights = c(5,1))
-heatmap <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_heatmap_facet.png")
-
-plot_grid(pcoa, heatmap, labels = c("A", "B"), label_size = 10, ncol = 2, rel_heights = c(1,10), rel_widths = c(.5, 1))+
-  ggsave("results/figures/figure_5_16S.pdf", width = 6, height = 2.04)
-
 #Vert version of Fig 4 16S
-f <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png")
+f <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png", scale = 1.2)
 g <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend_vert.png")
-pcoa1 <- plot_grid(f,g, label_size = 12, nrow = 1, rel_heights = c(1,.8), rel_widths = c(1,.8))
+h <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_fmt.png")
+lineplot <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_stools.png")
+pcoa1 <- plot_grid(f,g, abels = c("A", ""), label_size = 12, nrow = 1, rel_heights = c(1,.5), rel_widths = c(1,.5))
 
-plot_grid(pcoa1, heatmap, labels = c("A", "B"), label_size = 12, nrow = 2)+
-  ggsave("results/figures/figure_5_16S.pdf", width = 6, height = 6.7)+
-  ggsave("submission/figure_5_16S.pdf", width = 6, height = 6.7)
[…]

### 2021-04-12 code/5_days_PEG_16S.R
Remove alpha legend from PCoA plots

@@ -170,7 +170,8 @@ axis2 <- pcoa_axes_5_day_PEG_stool %>% filter(axis == 2) %>% pull(loading) %>% r
 
 pcoa_subset_plot_stool <- plot_pcoa(pcoa_5_day_PEG_stool)+
   labs(x = paste("PCoA 1 (", axis1, "%)", sep = ""), #Annotations for each axis from loadings file
-       y = paste("PCoA 2 (", axis2,"%)", sep = ""))
+       y = paste("PCoA 2 (", axis2,"%)", sep = ""))+
+  theme(legend.position = "none")
 save_plot(filename = paste0("results/figures/5_days_PEG_stool_PCoA.png"), pcoa_subset_plot_stool, base_height = 5, base_width = 5)
 
 ##Animation of PCoA plot: Stool Subset--
@@ -217,7 +218,8 @@ pcoa_subset_plot_tissue <- plot_pcoa(pcoa_5_day_PEG_tissues)+
        y = paste("PCoA 2 (", axis2,"%)", sep = ""))+
   scale_alpha_continuous(range = c(.3, 1),
                          breaks= c(4, 6, 20, 30),
-                         labels=c(4, 6, 20, 30))
+                         labels=c(4, 6, 20, 30))+
+  theme(legend.position = NULL)
 save_plot(filename = paste0("results/figures/5_days_PEG_tissues_PCoA.png"), pcoa_subset_plot_tissue, base_height = 5, base_width = 5)
 
 #PCoA faceted over time

### 2021-04-12 code/figure_6.R
Add area plot and make colors in panels B & C match

@@ -3,9 +3,12 @@ source("code/utilities.R") #Loads libraries, reads in metadata, functions
 a <- ggdraw() + draw_image("results/figures/ml_performance_genus.png")
 b <- ggdraw() + draw_image("results/figures/ml_top_features_genus.png")
 top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow=1, rel_widths = c(1,1.5), rel_heights = c(2, 1))
-c <- ggdraw() + draw_image("results/figures/ml_top10_d5_genus.png")#Note scale could be removed
 
-plot_grid(top_panel, c, labels = c("", "C"), label_size = 12, ncol=1, rel_widths = c(1, 3), rel_heights = c(1, 1.5))+
+c1 <- ggdraw() + draw_image("results/figures/ml_abund_5_genus_area.png")#Note scale could be removed
+c2 <-   ggdraw() + draw_image("results/figures/ml_abund_5_genus_area_legend.png")#Note scale could be removed
+c <- plot_grid(c1, c2, labels = NULL, label_size = 12, ncol=1, rel_widths = c(1,.3), rel_heights = c(1, .3))
+
+plot_grid(top_panel, c, labels = c("", "C"), label_size = 12, ncol=1, rel_widths = c(1, 2), rel_heights = c(1, 1.5))+
   ggsave("results/figures/figure_6.pdf", width=8, height=7)+
   ggsave("submission/figure_6.pdf", width=8, height=7)
 

### 2021-04-12 code/ml_results.R

@@ -1,5 +1,6 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 library(viridis)
+library(scales)
 
 set.seed(19760620) #Same seed used for mothur analysis
 
@@ -100,7 +101,7 @@ rf_top_feat <- top_20(rf_feat) %>% pull(genus)
 plot_feat_imp <- function(df, top_feat){
   df %>% 
     filter(genus %in% top_feat) %>% 
-    ggplot(aes(fct_reorder(genus, -perf_metric_diff, .desc = TRUE), perf_metric_diff))+
+    ggplot(aes(fct_reorder(genus, -perf_metric_diff, .desc = TRUE), perf_metric_diff, color = genus))+
     stat_summary(fun = 'median', 
                  fun.max = function(x) quantile(x, 0.75), 
                  fun.min = function(x) quantile(x, 0.25),
@@ -122,6 +123,21 @@ plot_feat_imp <- function(df, top_feat){
 }
 
 #Plot feature importances for the top genera for each comparison----
+#Figure out colors for top 5 most abundant genera based on viridis scale
+show_col(viridis_pal()(5))
+color_scheme_df <- rf_feat %>% 
+  distinct(genus) %>% #Limit to unique genera
+  filter(genus %in% rf_top_feat) %>% 
+  #Assign viridis colors to top 5 most abundant genera
+  mutate(color = case_when(genus == "Porphyromonadaceae Unclassified" ~ "#440154FF",
+                           genus == "Akkermansia" ~ "#3B528BFF",
+                           genus == "Enterobacteriaceae Unclassified" ~ "#21908CFF",
+                           genus == "Lachnospiraceae Unclassified" ~ "#5DC863FF",
+                           genus == "Bacteroides" ~ "#FDE725FF",
+                           TRUE ~ "black")) #Rest of colors should be black
+color_scheme <- color_scheme_df %>% pull(color)
+color_bact <- color_scheme_df %>% pull(genus)
+legend_bact <- color_scheme_df %>% pull(genus)
 rf_feat_5dpi <- plot_feat_imp(rf_feat, rf_top_feat)+
   ggsave("results/figures/ml_top_features_genus.png", height = 5, width = 8)
 

### 2021-04-13 submission/manuscript.Rmd
Update manuscript with latest figures
Add to results section

@@
 
-**Laxative treatment alone leads to prolonged *C.
+**5-day laxative treatment leads to prolonged *C.
 difficile* colonization in mice.** We compared PEG-treated mice to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C.
-difficile*, with mice typically clearing C.
-difficile within 10 days post-infection [@Tomkovich2019; @Tomkovich2020].
+difficile*, with mice typically clearing *C.
+difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020].
 All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig.
 1A).
-After PEG and/or antibiotic treatment all mice were challenged with 10^3^ *C.
+PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment (Fig.
+1B).
+After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C.
 difficile* 630 spores.
+All treatments rendered mice susceptible to *C.
+difficile* colonization (Fig.
+1C), however PEG-treated mice remained colonized at a high level through 30 days post-infection.
+In contrast, the clindamycin-treated mice that cleared *C.
+difficile* within 10 days post-infection.
+Surprisingly, mice were still susceptible to *C.
+difficile* infection after 10-days of recovery from PEG treatment although *C.
+difficile* was not detectable in most of the group in the initial 5 days post-infection (Fig.
+1C).
+From 9 days post-infection onward, the median *C.
+difficile* stabilized for the 5-day PEG plus 10-day recovery group of mice and remained high through 30 days post-infection (Fig.
+1C).
+Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C.
+difficile* colonization and PEG-treated mice remained susceptible for up to 10 days post PEG treatment.
+**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Given 
+**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given
 
@@
 
-* Figure 5. 1-day PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.
+* Figure 5. 1-day PEG treatment post *C. difficile* challenge prolongs colonization regardless of whether an FMT is also administered.
 
@@
 \includegraphics{figure_5.pdf}
+\newpage
 \includegraphics{figure_5_16S.pdf}

### 2021-04-13 submission/manuscript.tex
Update manuscript with latest figures
Add to results section

@@
 
-\textbf{Laxative treatment alone leads to prolonged \emph{C. difficile}
+\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
 colonization in mice.} We compared PEG-treated mice to our standard 10
 mg/kg clindamycin treatment, which temporarily renders the mice
-susceptible to \emph{C. difficile}, with mice typically clearing C.
-difficile within 10 days post-infection (9, 19). All PEG-treated mice
-were administered a 15\% PEG solution in the drinking water for 5-days,
-one group was also treated with clindamycin, and one group was allowed
-to recover for 10 days prior to challenge (Fig. 1A). After PEG and/or
-antibiotic treatment all mice were challenged with 10\textsuperscript{3}
-\emph{C. difficile} 630 spores.
+susceptible to \emph{C. difficile}, with mice typically clearing
+\emph{C. difficile} within 10 days post-infection (9, 19). All
+PEG-treated mice were administered a 15\% PEG solution in the drinking
+water for 5-days, one group was also treated with clindamycin, and one
+group was allowed to recover for 10 days prior to challenge (Fig. 1A).
+PEG treatment resulted in weight loss in all 3 groups of PEG-treated
+mice, with the greatest change in weight observed on the fifth day of
+PEG treatment (Fig. 1B). After either PEG, clindamycin, or PEG and
+clindamycin treatment all mice were challenged with
+10\textsuperscript{3} \emph{C. difficile} 630 spores. All treatments
+rendered mice susceptible to \emph{C. difficile} colonization (Fig. 1C),
+however PEG-treated mice remained colonized at a high level through 30
+days post-infection. In contrast, the clindamycin-treated mice that
+cleared \emph{C. difficile} within 10 days post-infection. Surprisingly,
+mice were still susceptible to \emph{C. difficile} infection after
+10-days of recovery from PEG treatment although \emph{C. difficile} was
+not detectable in most of the group in the initial 5 days post-infection
+(Fig. 1C). From 9 days post-infection onward, the median \emph{C.
+difficile} stabilized for the 5-day PEG plus 10-day recovery group of
+mice and remained high through 30 days post-infection (Fig. 1C). Thus,
+osmotic laxative treatment alone was sufficient to render mice
+susceptible to prolonged \emph{C. difficile} colonization and
+PEG-treated mice remained susceptible for up to 10 days post PEG
+treatment. \textbf{5-day laxative treatment differentially disrupts the
+fecal microbiota compared to clindamycin treatment.} Given \textbf{5-day
+laxative treatment does not promote more severe CDIs despite altering
+the mucosal microbiota.} Given
 
@@
 \item
-  Figure 5. 1-day PEG treatment post C. difficile challenge prolongs
-  colonization regardless of whether an FMT is also administered.
+  Figure 5. 1-day PEG treatment post \emph{C. difficile} challenge
+  prolongs colonization regardless of whether an FMT is also
+  administered.
 \item
@@
 
-\includegraphics{figure_5.pdf} \includegraphics{figure_5_16S.pdf}
-\textbf{Figure 5. 1-day PEG treatment post C. difficile challenge
-prolongs colonization regardless of whether an FMT is also
-administered.} A. \newpage
+\includegraphics{figure_5.pdf} \newpage
+\includegraphics{figure_5_16S.pdf} \textbf{Figure 5. 1-day PEG treatment
+post C. difficile challenge prolongs colonization regardless of whether
+an FMT is also administered.} A. \newpage
 

### 2021-04-13 submission/manuscript.Rmd
Edit intro based on writing group feedback

@@
 
-Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota [@Britton2014].
+Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because they disrupt microbiota colonization resistance [@Britton2014].
 However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020].
-Although, other medications have been implicated as risk factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].
-Many of the non-antibiotic medications associated with CDIs are known to modulate intestinal motility, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020].
+Although, other medications (proton pump inhibitors, osmotic laxatives, antimotility agents, and opioids) have been implicated as risk or protective factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].
 
+Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020].
 Stool consistency often serves as an approximation of intestinal motility.
 Our group has shown that when *C.
-difficile* negative controls are separated into two groups based on stool consistency, there are microbiota features such as alpha diversity that overlap between samples from CDI patients and control patients with diarrhea [@Schubert2015].
-These results led to our hypothesis that bacterial communities from patients experiencing diarrhea are susceptible, but may not have been exposed to *C.
-difficile* spores.
+difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C.
+difficile* negative with non-diarrheal consistency [@Schubert2015].
+These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.
 
-Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupts the intestinal microbiota in humans [@Nagata2019].
+Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019].
 The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies.
-Interestingly, previous studies have shown that treating mice with PEG alone rendered the mice susceptible to *C.
-difficile* infection, altered microbiota composition, reduced acetate and butyrate and altered the mucus barrier [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020].
+Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C.
+difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020].
 The mucus barrier is thought to mediate protection from *C.
@@
 difficile* (Ref).
-However, whether laxative results in more severe CDIs in mice is unclear.
+However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C.
+difficile* after challenge is unclear.
 
@@
 For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure.
-Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested the laxative may help flush *C.
+Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C.
 difficile* spores and toxins from the intestine [@Liacouras1996].
 
-In the past, our group has used C57BL6 mice to characterize how antibiotics disrupt the microbiota and influence *C.
+In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C.
 difficile* susceptibility and clearance [ref].
 Although, two groups have now shown PEG treatment alone renders mice susceptible to *C.
-difficile*, these studies have raised additional questions that should be addressed given their relevance to CDIs.
-Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in sustained *C.
+difficile*, these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C.
+difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs.
+Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in more severe CDI and sustained *C.
 difficile* colonization, and if PEG treatment post-CDI can promote *C.

### 2021-04-13 submission/manuscript.tex
Edit intro based on writing group feedback

@@
 Antibiotics are a major risk factor for \emph{Clostridioides difficile}
-infections (CDIs) because of their impact on the microbiota (1).
-However, antibiotics are not the only types of medications that disrupt
-the microbiota (2--4). Although, other medications have been implicated
-as risk factors for CDIs through epidemiological studies, whether the
-association is due to their impact on the microbiome is still unclear
-(5--9). Many of the non-antibiotic medications associated with CDIs are
-known to modulate intestinal motility, which in turn also strongly
-impacts microbiota composition and function (10, 11).
+infections (CDIs) because they disrupt microbiota colonization
+resistance (1). However, antibiotics are not the only types of
+medications that disrupt the microbiota (2--4). Although, other
+medications (proton pump inhibitors, osmotic laxatives, antimotility
+agents, and opioids) have been implicated as risk or protective factors
+for CDIs through epidemiological studies, whether the association is due
+to their impact on the microbiome is still unclear (5--9).
 
-Stool consistency often serves as an approximation of intestinal
-motility. Our group has shown that when \emph{C. difficile} negative
-controls are separated into two groups based on stool consistency, there
-are microbiota features such as alpha diversity that overlap between
-samples from CDI patients and control patients with diarrhea (12). These
-results led to our hypothesis that bacterial communities from patients
-experiencing diarrhea are susceptible, but may not have been exposed to
-\emph{C. difficile} spores.
+Many of the non-antibiotic medications associated with CDIs are known to
+modulate gastrointestinal motility leading to either increased or
+decreased colonic transit time, which in turn also strongly impacts
+microbiota composition and function (10, 11). Stool consistency often
+serves as an approximation of intestinal motility. Our group has shown
+that when \emph{C. difficile} negative controls are separated into two
+groups based on stool consistency, there are shared microbiota features
+such as lower alpha diversity in samples from CDI patients and control
+patients with diarrhea compared to control samples that were \emph{C.
+difficile} negative with non-diarrheal consistency (12). These results
+led to a hypothesis that bacterial communities from patients
+experiencing diarrhea are susceptible to developing CDIs.
 
 Osmotic laxatives can lead to diarrhea depending on the administered
-dose and temporarily disrupts the intestinal microbiota in humans (13).
-The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
-in Miralax, Nulytely, and Golytely and is also commonly used as bowel
+dose and temporarily disrupt the human intestinal microbiota (13). The
+ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
+Miralax, Nulytely, and Golytely and is also commonly used as bowel
 preparation for colonoscopies. Interestingly, previous studies have
-shown that treating mice with PEG alone rendered the mice susceptible to
-\emph{C. difficile} infection, altered microbiota composition, reduced
-acetate and butyrate and altered the mucus barrier (14--17). The mucus
-barrier is thought to mediate protection from \emph{C. difficile}
-infections by protecting intestinal epithelial cells from the toxins
-produced by \emph{C. difficile} (Ref). However, whether laxative results
-in more severe CDIs in mice is unclear.
+shown that treating mice with PEG alone altered microbiota composition,
+reduced acetate and butyrate production, altered the mucus barrier, and
+rendered the mice susceptible to \emph{C. difficile} infection (14--17).
+The mucus barrier is thought to mediate protection from \emph{C.
+difficile} infections by protecting intestinal epithelial cells from the
+toxins produced by \emph{C. difficile} (Ref). However, whether laxative
+results in more severe CDIs in mice and how long mice remain colonized
+with \emph{C. difficile} after challenge is unclear.
 
@@
 based on a 1996 case study with 2 pediatric patients where the authors
-suggested the laxative may help flush \emph{C. difficile} spores and
-toxins from the intestine (18).
+suggested in the discussion that the laxative may help flush \emph{C.
+difficile} spores and toxins from the intestine (18).
 
 In the past, our group has used C57BL6 mice to characterize how
-antibiotics disrupt the microbiota and influence \emph{C. difficile}
-susceptibility and clearance {[}ref{]}. Although, two groups have now
-shown PEG treatment alone renders mice susceptible to \emph{C.
-difficile}, these studies have raised additional questions that should
-be addressed given their relevance to CDIs. Here, we used our C57BL/6
-clindamycin model as a control group to characterize how long
-PEG-treated mice remain susceptible, whether PEG treatment results in
-sustained \emph{C. difficile} colonization, and if PEG treatment
-post-CDI can promote \emph{C. difficile} clearance.
+antibiotics including clindamycin disrupt the microbiota and influence
+\emph{C. difficile} susceptibility and clearance {[}ref{]}. Although,
+two groups have now shown PEG treatment alone renders mice susceptible
+to \emph{C. difficile}, these studies have raised additional questions
+regarding the dynamics and severity of infection as well as the role of
+laxative treatment in \emph{C. difficile} clearance that should be
+addressed to better inform how we think about laxatives in the context
+of CDIs. Here, we used our C57BL/6 clindamycin model as a control group
+to characterize how long PEG-treated mice remain susceptible, whether
+PEG treatment results in more severe CDI and sustained \emph{C.
+difficile} colonization, and if PEG treatment post-CDI can promote
+\emph{C. difficile} clearance.
 

### 2021-04-13 code/5_days_PEG_cfu_weight.R
Add weight data through day 3- to figure, add schematic edits

@@ -194,6 +194,8 @@ label <- kw_label(weight_kruskal_wallis_adjust %>%
 
 weight_subset <- weight_subset %>% 
   mutate(day = as.integer(day))
+weightdata <- weightdata %>% 
+  mutate(day = as.integer(day))
 
 #Plot of weight data for days -15 through 10 of the experiment:
 weight_subset_plot <- plot_weight(weight_subset) +
@@ -202,18 +204,26 @@ weight_subset_plot <- plot_weight(weight_subset) +
                      minor_breaks = c(-15.5:10.5)) 
 save_plot(filename = "results/figures/5_days_PEG_weight_subset.png", weight_subset_plot, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
-
-#Plot of weight data for all days of the experiment:
-#Note don't need to redo statistical annotations since there were no significant differences past 1 day post-infection
-weight_plot <- plot_weight(weightdata) +
-  scale_x_continuous(breaks = c(-15, -10, -5, 0, 5, 10, 15, 20, 25, 30),
-                     limits = c(-16, 31),
-                     minor_breaks = c(-15.5:10.5, 11.5, 12.5, 14.5, 15.5, 19.5, 20.5, 24.5, 25.5, 29.5, 30.5)) #only show grey lines around days on days with points)
-
 #Plots with just the median lines for each group
 v2_weight_subset <- plot_weight_medians(weight_subset) +
   scale_x_continuous(breaks = c(-15, -10, -5, 0, 5, 10),
                      limits = c(-16, 11),
                      minor_breaks = c(-15.5:10.5)) #only show grey lines separating days on days with statistically sig points)
 save_plot(filename = "results/figures/5_days_PEGv2_weight_subset.png", v2_weight_subset, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
+
+#Plot of weight data for all days of the experiment:
+#Note don't need to redo statistical annotations since there were no significant differences past 10 day post-infection
+#Statistical annotation labels based on adjusted kruskal-wallis p-values for first 10 days of experiment:
+x_annotation <- weight_kruskal_wallis_adjust %>%
+  filter(p.value.adj <= 0.05) %>%
+  pull(day)
+x_annotation == sig_weight_days #All the days where weight change significantly varied across groups of mice occured within the first 10 days post-infection
[…]

### 2021-04-13 code/figure_1.R

@@ -2,7 +2,7 @@ source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
 #Figure with Schematic, CFU over time, and shannon over time
 a <- ggdraw() + draw_image("results/figures/5_days_PEG_exp_schematic.png")
-b <- ggdraw() + draw_image("results/figures/5_days_PEG_weight_subset.png")
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_weight.png")
 c <- ggdraw() + draw_image("results/figures/5_days_PEG_cfu.png")#Note scale could be removed
 
 plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1)+

### 2021-04-13 code/5_days_PEG_cfu_weight.R
Fix issues with cfu plot code, update figure 1

@@ -136,14 +136,17 @@ weight_plot_format_stats <- weight_stats_pairwise %>%
   bind_rows()
 
 #Plots of CFU and weight data----
+#Transform day column variable from character to integer variable
+cfudata <- cfudata %>% 
+  mutate(day = as.integer(day))
 
 #Dataframe of cfu data for just the initial 10 days of the experiment
 cfudata_10dsubset <- cfudata %>%
-  mutate(day = as.integer(day)) %>% #transform day to factor variable
+  mutate(day = as.integer(day)) %>% #transform day to integer variable
   filter(day < 12) #only include data through day 10
 
 cfu_kruskal_wallis_adjust <- cfu_kruskal_wallis_adjust %>% 
-  mutate(day = as.integer(day)) #transform day to factor variable
+  mutate(day = as.integer(day)) #transform day to integer variable
   
 #Statistical annotation labels based on adjusted kruskal-wallis p-values for first 10 days of experiment:
 x_annotation <- cfu_kruskal_wallis_adjust %>%
@@ -155,19 +158,23 @@ label <- kw_label(cfu_kruskal_wallis_adjust %>% filter(day < 12))
  #Only include results through day 10 for this plot
 
 #Plot cfu for just the inital 10days
-cfu_10d <- plot_cfu_data(cfudata_10dsubset) +
+cfu_10d <- plot_cfu_data(cfudata_10dsubset %>% 
+                           filter(day %in% c("-1", "0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))) + #Filter to just include timepoints that will be plotted
       scale_x_continuous(breaks = c(0:10),
                          limits = c(-1, 11),
                          minor_breaks = c(-.5:10.5))
 save_plot(filename = "results/figures/5_days_PEG_cfu_10d.png", cfu_10d, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
 #Statistical annotation labels based on adjusted kruskal-wallis p-values for all timepoints:
-x_annotation <- sig_cfu_days
+x_annotation <- cfu_kruskal_wallis_adjust %>%
+  filter(p.value.adj <= 0.05) %>%
+  pull(day)
 y_position <- max(cfudata$avg_cfu)
 label <- kw_label(cfu_kruskal_wallis_adjust)
 
 #Plot of cfu data for all days of the experiment
-cfu <- plot_cfu_data(cfudata) +
[…]

### 2021-04-13 code/5_days_PEG_cfu_weight.R
Remove legend from panel C

@@ -177,7 +177,8 @@ cfu <- plot_cfu_data(cfudata %>%
                        filter(day %in% c("-1", "0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "15", "20", "25", "30"))) +
   scale_x_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30),
                      limits = c(-1, 31),
-                     minor_breaks = c(-.5:10.5, 11.5, 12.5, 14.5, 15.5, 19.5, 20.5, 24.5, 25.5, 29.5, 30.5)) #only show grey lines separating days on days with statistically sig points
+                     minor_breaks = c(-.5:10.5, 11.5, 12.5, 14.5, 15.5, 19.5, 20.5, 24.5, 25.5, 29.5, 30.5))+ #only show grey lines separating days on days with statistically sig points
+  theme(legend.position = "none")
 save_plot(filename = "results/figures/5_days_PEG_cfu.png", cfu, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
 #Weight change plot----

### 2021-04-13 code/utilities.R
Change alpha variable to day for cfu and weight over time functions to match alpha diversity over time plots

@@ -373,7 +373,7 @@ plot_cfu_data <- function(df){
     summarize(median_avg_cfu = median(avg_cfu, na.rm = TRUE))
   #Plot cfu for just the inital 10days
   cfu_plot <- ggplot(NULL) +
-    geom_point(df, mapping = aes(x = day, y = avg_cfu, color= group, fill = group), alpha = .2, size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
+    geom_point(df, mapping = aes(x = day, y = avg_cfu, color= group, fill = group, alpha = day), size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
     geom_line(median_summary, mapping = aes(x = day, y = median_avg_cfu, group = group, color = group), alpha = 0.6, size = 1.5) +
     scale_colour_manual(name=NULL,
                         values=color_scheme,
@@ -399,7 +399,7 @@ plot_weight <- function(df){
     group_by(group, day) %>%
     summarize(median_weight_change = median(weight_change, na.rm = TRUE))
   ggplot(NULL) +
-    geom_point(df, mapping = aes(x = day, y = weight_change, color= group, fill = group), alpha = .2, size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
+    geom_point(df, mapping = aes(x = day, y = weight_change, color= group, fill = group, alpha = day), size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
     geom_line(median_summary, mapping = aes(x = day, y = median_weight_change, color = group), alpha = 0.6, size = 1.5) +
     scale_colour_manual(name=NULL,
                         values=color_scheme,

### 2021-04-14 submission/manuscript.Rmd
Add M&I PDA writing group to acknowledgements
Add results text

@@
 difficile* colonization and PEG-treated mice remained susceptible for up to 10 days post PEG treatment.
-**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Given 
-**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given
 
-* Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.
+``` {r 5-day PEG microbiota}
+f_d_permanova <- read_tsv(here("data/process/5_days_PEG_permanova_stools.tsv"))
+f_d_top_2_contrib <- f_d_permanova %>% arrange(desc(R2)) %>%
+  filter(!effects == "Total") %>% #Remove total since it is not one of the variables
+  top_n(2, R2) %>% #Select top 2 variables based on R2
+  select(effects, R2, `Pr(>F)`) #Select relevant column names
+
+f_d_top_contrib= (f_d_top_2_contrib[1,1])
+f_d_sec_contrib = (f_d_top_2_contrib[2,1])
+f_d_top_R2_sum = f_d_top_2_contrib %>% 
+  select(R2) %>% sum
+f_d_top_p <- f_d_top_2_contrib %>% 
+  select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text
+f_d_top_R2 = f_d_top_2_contrib[1,2]
+f_d_sec_R2 = f_d_top_2_contrib[2,2]
+```
+**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C.
+difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to their differential effects on the microbiota.
+We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups.
+We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X).
+The impact of cage effects has been well documented due to copraphagy and is expected.
+
+**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given previous work has shown that PEG treatment disrupts the mucus layer, we decided to examine the impact of the laxative on the mucosal microbiota and infection severity by evaluating histopathology.
 
@@
 We thank members of the Schloss lab for feedback on planning the experiments and data presentation.
-We also thank Andrew Henry for help with media preparation and bacterial culture.
+We thank Andrew Henry for help with media preparation and bacterial culture and the Microbiology and Immunology department's postdoc association writing group members for their feedback on manuscript drafts.
 We also thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments.

### 2021-04-14 submission/manuscript.tex
Add results text

@@
 PEG-treated mice remained susceptible for up to 10 days post PEG
-treatment. \textbf{5-day laxative treatment differentially disrupts the
-fecal microbiota compared to clindamycin treatment.} Given \textbf{5-day
-laxative treatment does not promote more severe CDIs despite altering
-the mucosal microbiota.} Given
+treatment.
+
+\textbf{5-day laxative treatment differentially disrupts the fecal
+microbiota compared to clindamycin treatment.} Since laxatives and
+clindamycin have previously been shown to disrupt the murine microbiota
+(ref), we hypothesized the different \emph{C. difficile} colonization
+dynamics between mice treated with laxatives versus clindamycin were due
+to their differential effects on the microbiota. We profiled the stool
+microbiota over time by sequencing the V4 region of the 16S rRNA gene to
+compare changes across treatment groups. We found time and treatment
+group explained half of the observed variation between fecal communities
+(PERMANOVA combined R\textsuperscript{2} = 0.50, \emph{P} \textless{}
+0.001).
+
+\textbf{5-day laxative treatment does not promote more severe CDIs
+despite altering the mucosal microbiota.} Given previous work has shown
+that PEG treatment disrupts the mucus layer, we decided to examine the
+impact of the laxative on the mucosal microbiota and infection severity
+by evaluating histopathology.
 
 \begin{itemize}
-\item
-  Figure 1. 5-day PEG treatment prolongs susceptibility and mice become
-  persistently colonized with \emph{C. difficile}.
 \item
@@
 We thank members of the Schloss lab for feedback on planning the
-experiments and data presentation. We also thank Andrew Henry for help
-with media preparation and bacterial culture. We also thank the Unit for
+experiments and data presentation. We thank Andrew Henry for help with
+media preparation and bacterial culture and the Microbiology and
+Immunology department's postdoc association writing group members for
+their feedback on manuscript drafts. We also thank the Unit for
 Laboratory Animal Medicine at the University of Michigan for maintaining

### 2021-04-14 submission/manuscript.Rmd
Fix typos

@@
 The black dotted line represents the limit of detection for the first serial dilution.
-Asterisks indicate timepoints where the CFU/gram was siginificantly different between groups using the Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple timepoints.
+Asterisks indicate timepoints where the CFU/gram was significantly different between groups using the Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple timepoints.
 C.
@@
 D.
-Shannon Diveristy Index of the groups over time.
+Shannon diversity Index of the groups over time.
 Only days with samples from all groups are shown.

### 2021-04-15 code/1_day_PEG_16S.R
Change baseline to B and rerun one day subset

@@ -8,14 +8,14 @@ color_labels <- c("Clind.", "1-day PEG", "1-day PEG + 1-day recovery")
 
 #Subset alpha diversity data (16S_common_files) to analyze one day PEG subset mice----
 diversity_data_subset <- one_day_PEG_subset(diversity_data) %>%
-  mutate(day = case_when(group == "C" & day == "-15" ~ "baseline",
-                         group == "M1" & day == "-11" ~ "baseline",
-                         group == "1RM1" & day == "-2" ~ "baseline",
+  mutate(day = case_when(group == "C" & day == "-15" ~ "B",
+                         group == "M1" & day == "-11" ~ "B",
+                         group == "1RM1" & day == "-2" ~ "B",
                          TRUE ~ day))#Replace day -2 and day -1 with baseline to represent the baseline timepoint
 
  diversity_data_subset <- diversity_data_subset %>%
-   filter(day %in% c("baseline", 0, 1, 2, 4, 5, 7)) %>%
-   mutate(day = fct_relevel(day, "baseline", "0", "1" , "2" , "4", "5" , "7")) #Specify the order of the days for plotting overtime
+   filter(day %in% c("B", 0, 1, 2, 3, 4, 5, 6, 7)) %>%
+   mutate(day = fct_relevel(day, "B", "0", "1" , "2", "3", "4", "5", "6", "7")) #Specify the order of the days for plotting overtime
  
 #Get exp days sequenced in both pretreatment and other days
 exp_days_seq <- unique(diversity_data_subset %>% pull(day)) 
@@ -28,11 +28,11 @@ set.seed(19760620) #Same seed used for mothur analysis
 
 #Function to test at the otu level:
 agg_otu_data_subset <- one_day_PEG_subset(agg_otu_data) %>%
-  mutate(day = case_when(group == "C" & day == "-15" ~ "baseline",
-                             group == "M1" & day == "-11" ~ "baseline",
-                             group == "1RM1" & day == "-2" ~ "baseline",
+  mutate(day = case_when(group == "C" & day == "-15" ~ "B",
+                             group == "M1" & day == "-11" ~ "B",
+                             group == "1RM1" & day == "-2" ~ "B",
[…]

### 2021-04-15 code/1_day_PEG_16S.R
Fix one day shannon x axis and update fig 4 with uniform timepoints

@@ -91,9 +91,10 @@ x_annotation <- sig_shannon_days
 y_position <- max(diversity_data_subset$shannon)+0.15
 label <- kw_label(kw_shannon)
 #Plot
+fig_days <- c("B", 0:10)
 shannon_stools <- plot_shannon_overtime(diversity_data_subset) +
-  scale_x_discrete(breaks = c("B", 0:7)) +
-  geom_vline(xintercept = c((1:7) - 0.5 ), color = "grey") + # Add gray lines to clearly separate OTUs
+  scale_x_discrete(limits = fig_days, breaks = fig_days, labels = fig_days) +
+  geom_vline(xintercept = c((1:13) - 0.5 ), color = "grey") + # Add gray lines to clearly separate OTUs
   theme(legend.position = "none")
 save_plot(filename = "results/figures/1_Day_PEG_shannon.png", shannon_stools, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
@@ -647,9 +648,9 @@ sig_genus_top_list_v2 <- sig_genus_top_list_v2[!(sig_genus_top_list_v2 %in% sig_
 
 facet_labels <- sig_genus_top_list
 names(facet_labels) <- sig_genus_top_list
-lineplot_days <- c("B", 0:10)
-line_plot_baselinetoD1_5_genera <- line_plot_genus(agg_genus_data_subset, sig_genus_top_list_v2, lineplot_days, "solid") +
-  scale_x_discrete(limits = lineplot_days, breaks = lineplot_days, labels = lineplot_days)
+fig_days <- c("B", 0:10)
+line_plot_baselinetoD1_5_genera <- line_plot_genus(agg_genus_data_subset, sig_genus_top_list_v2, fig_days, "solid") +
+  scale_x_discrete(limits = fig_days, breaks = fig_days, labels = fig_days)
 save_plot(filename = "results/figures/1_Day_PEG_genus_6_baselinetoD1_lineplot.png", line_plot_baselinetoD1_5_genera, base_height = 5, base_width = 15)
 
 #Plot significant genera not still significant on Day 7 on lineplot using V2 above for comp
@@ -657,8 +658,8 @@ sig_genus_top_list <- sig_genus_pairs[!(sig_genus_pairs %in% sig_genus_pairs_D7)
 facet_labels <- sig_genus_top_list
 names(facet_labels) <- sig_genus_top_list
 
[…]

### 2021-04-15 submission/manuscript.Rmd
Add mice per timepoint range to Fig 4 legend and update Fig 4 in manuscript

@@
 library(here)
+library(readxl)
 
@@
 knitr::knit_hooks$set(inline=inline_hook)
+
+metadata <- read_excel(here("data/process/metadata.xlsx"), col_types = c("text", "numeric", "text", "text", "numeric", "text", "numeric", "text", "text", "text", "text", "numeric", "numeric")) #specify column types
+
 ```
@@
 sec_R2 = one_day_top_2_contrib[3,6]
+
+one_day_samples_per_day <- metadata %>%
+  filter(group == "C" & exp_num %in% c("M6")| #Only use C mice from this experiments. Allocated groups to figures based on paper outline.
+             group == "1RM1" & exp_num %in% c("M6R")| #Had to differentiate experiment 6 from 6R in the metadata to create unique_mouse_id that wouldn't overlap for the M1 & 1RM1 mice that are both labeled with mouse_ids that are #s1-6
+             group == "M1" & exp_num %in% c("M6")) %>%
+  group_by(day) %>%
+  count() 
+one_day_sample_min <- min(one_day_samples_per_day$n)
+one_day_sample_max <- max(one_day_samples_per_day$n)
+
 ```
@@
 B.
-CFU/gram stool measured over time (N = 6 mice per timepoint) via several dilutions.
+CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per timepoint) via several dilutions.
 The black dotted line represents the limit of detection for the first serial dilution.
@@
 Only days with samples from all groups are shown.
-Samples for some mice were difficult to obtain
-due to the laxative treatment.
-The alpha scale follows accordingly with the timeline.
 E.

### 2021-04-15 submission/manuscript.tex
Add mice per timepoint range to Fig 4 legend and update Fig 4 in manuscript

@@
 (PERMANOVA combined R\textsuperscript{2} = 0.50, \emph{P} \textless{}
-0.001).
+0.001) with the remaining variation explained by interactions between
+treatment group and other experimental variables including time, cage
+effects, and sequencing preparation (Data Set S1, Sheet X). The impact
+of cage effects has been well documented due to copraphagy and is
+expected.
 
@@
 A. Setup of the experimental timeline for the 1-day PEG treated subset
-of mice. B. CFU/gram stool measured over time (N = 6 mice per timepoint)
-via several dilutions. The black dotted line represents the limit of
-detection for the first serial dilution. Asterisks indicate timepoints
-where the CFU/gram was siginificantly different between groups using the
-Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple
-timepoints. C. Principle Coordinate Analysis plot of the groups over
-time with the alpha representating the same time scale as in panel D
-(day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} = 0.19).
-D. Shannon Diveristy Index of the groups over time. Only days with
-samples from all groups are shown. Samples for some mice were difficult
-to obtain due to the laxative treatment. The alpha scale follows
-accordingly with the timeline. E. Line plots of relative percent
-abundance of selected genera over time. Only days with samples from all
-groups shown. The gray line represents the limit of detection.
+of mice. B. CFU/gram stool measured over time (N = 12-18 mice per
+timepoint) via several dilutions. The black dotted line represents the
+limit of detection for the first serial dilution. Asterisks indicate
+timepoints where the CFU/gram was significantly different between groups
+using the Kruskall-Wallis test with a Benjamini-Hochberg correction for
+multiple timepoints. C. Principle Coordinate Analysis plot of the groups
+over time with the alpha representating the same time scale as in panel
+D (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} =
+0.19). D. Shannon diversity Index of the groups over time. Only days
+with samples from all groups are shown. E. Line plots of relative
+percent abundance of selected genera over time. Only days with samples
+from all groups shown. The gray line represents the limit of detection.
 

### 2021-04-15 code/figure_4.R

@@ -4,7 +4,7 @@ a <- ggdraw() + draw_image("results/figures/1_Day_PEG_Schematic.png", scale = 1.
 b <- ggdraw() + draw_image("results/figures/1_day_PEG_cfu.png", scale = 1.00)
 c <- ggdraw() + draw_image("results/figures/1_Day_PEG_PCoA.png", scale = 1.25)
 d <- ggdraw() + draw_image("results/figures/1_Day_PEG_shannon.png")
-e <- ggdraw() + draw_image("results/figures/1_Day_PEG_genus_6_baselinetoD1_lineplot.png")
+e <- ggdraw() + draw_image("results/figures/1_Day_PEG_genus_6_v2_baselinetoD1_lineplot.png")
 top_panel <- plot_grid(a, b, c, d, labels = c("A", "B", "C", "D"), label_size = 12, ncol = 2, rel_widths = c(.60, 1, 1,.60), rel_heights = c(1, 1))
 bottom_panel <- plot_grid(e, labels = c("E"), label_size = 12)
 plot_grid(top_panel, bottom_panel, nrow = 2, rel_heights = c(1.1, .8)) +

### 2021-04-15 code/post_CDI_PEG_16S.R
Make Fig 5 edits (alpha, remove 3350, replcace hm w/ lp)

@@ -129,9 +129,7 @@ shannon_post_cdi_peg_overtime_stool <- diversity_stools %>%
                      limits = c(-2,31), #removes day -15 here
                      minor_breaks = c(-1.5:10.5, 14.5, 15.5, 19.5, 20.5, 24.5, 25.5, 29.5, 30.5)) +
   #theme(legend.position = "none")+ #Removing legend to save separately
-  guides(color = guide_legend(ncol = 1),
-         alpha = guide_legend(ncol = 1),
-         group = FALSE)+
+  guides(color = guide_legend(ncol = 1))+
   theme(legend.title = element_blank())+
   annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = Inf, fill = "#88419d", alpha = .15)+ #shade to indicate PEG treatment in Clind + 1-day PEG group
   annotate("rect", xmin = 3, xmax = 4, ymin = 0, ymax = 2, fill = "#225ea8", alpha = .15)+ #shade to indicate PEG treatment in Clind + 3-day recovery + 1-day PEG + FMT/PBS
@@ -669,7 +667,7 @@ genus_day5_stats %>%
   filter(p.adj < .05) %>% 
   pull(genus) #[1] "Porphyromonadaceae Unclassified"
 
-#Select Genera to show what's different across all groups
+#Select Genera to show what's different across all groups---------
 pairwise_genus_rank <- genus_day5_stats %>% filter(p.adj < .05) %>% 
   count(genus) %>% arrange(desc(n)) %>% #Rank genera according to # of groups with sig diff
   top_n(6) %>% #select top 6 for over time plots
@@ -677,6 +675,7 @@ pairwise_genus_rank <- genus_day5_stats %>% filter(p.adj < .05) %>%
 kw_genus_stools %>% #Check to see if there were any genera that were sig in the kw test that were not sig in the pairwise comparisons
   filter(day == 5,p.value.adj < .05,
          !(genus %in% pairwise_genus_rank)) %>% pull(genus) #Only [1] "Turicibacter"
+
 #Plot alt line plots faceted by genus
 agg_genus_data_subset_hm <- agg_genus_data_subset %>% filter(!(day %in% c(4, 20, 25))) #drop day 20 and 25 (only data for one group)
 exp_groups <- c("RM", "FRM", "CWM", "C") #Arrange this way to match pcoa legend
@@ -718,7 +717,8 @@ top_10_sig_genus <- kw_genus_stools %>%
[…]

### 2021-04-15 code/figure_5.R
Fix typo in Fig 5 render

@@ -36,7 +36,7 @@ f <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png", scale 
 g <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend_vert.png")
 h <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_fmt.png")
 lineplot <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_stools.png")
-pcoa1 <- plot_grid(f,g, abels = c("A", ""), label_size = 12, nrow = 1, rel_heights = c(1,.5), rel_widths = c(1,.5))
+pcoa1 <- plot_grid(f,g, labels = c("A", ""), label_size = 12, nrow = 1, rel_heights = c(1,.5), rel_widths = c(1,.5))
 
 plot_grid(pcoa1, h, lineplot, labels = c("A", "B", "C"), label_size = 12, nrow = 3)+
   ggsave("results/figures/figure_5_16S.pdf", width = 4, height = 6.7)+

### 2021-04-15 code/post_CDI_PEG_16S.R
Clean up post cdi peg code, remove pre/post treat code

@@ -667,49 +667,7 @@ genus_day5_stats %>%
   filter(p.adj < .05) %>% 
   pull(genus) #[1] "Porphyromonadaceae Unclassified"
 
-#Select Genera to show what's different across all groups---------
-pairwise_genus_rank <- genus_day5_stats %>% filter(p.adj < .05) %>% 
-  count(genus) %>% arrange(desc(n)) %>% #Rank genera according to # of groups with sig diff
-  top_n(6) %>% #select top 6 for over time plots
-  pull(genus)
-kw_genus_stools %>% #Check to see if there were any genera that were sig in the kw test that were not sig in the pairwise comparisons
-  filter(day == 5,p.value.adj < .05,
-         !(genus %in% pairwise_genus_rank)) %>% pull(genus) #Only [1] "Turicibacter"
-
-#Plot alt line plots faceted by genus
-agg_genus_data_subset_hm <- agg_genus_data_subset %>% filter(!(day %in% c(4, 20, 25))) #drop day 20 and 25 (only data for one group)
-exp_groups <- c("RM", "FRM", "CWM", "C") #Arrange this way to match pcoa legend
-exp_group_labels <- c("Clind. + 3-day recovery + 1-day PEG + PBS","Clind. + 3-day recovery + 1-day PEG + FMT", "Clind. + 1-day PEG", "Clind.")
-line_plot_stool_days <- diversity_stools %>% distinct(day) %>% 
-  filter(!(day %in% c(-15, 30))) %>% pull(day)
-genus_line_plot_facet <- agg_genus_data_subset_hm %>% 
-  mutate(group = fct_relevel(group, exp_groups)) %>% #Specify the order of the groups
-  filter(genus %in% pairwise_genus_rank) %>%
-  filter(day %in% line_plot_stool_days) %>% 
-  mutate(day = as.numeric(day)) %>% 
-  group_by(group, genus, day) %>% 
-  summarize(median=median(agg_rel_abund + 1/2000),`.groups` = "drop") %>%  #Add small value (1/2Xsubssampling parameter) so that there are no infinite values with log transformation
-  ggplot()+
-  geom_line(aes(x = day, y=median, color=group))+
-  scale_colour_manual(name=NULL,
-                      values=color_scheme,
-                      breaks=color_groups,
-                      labels=color_labels)+
[…]

### 2021-04-15 submission/manuscript.Rmd
Edit legend for figure 1

@@
 
-test <- 4
+#Use same code from five_day_PEG_subset() in code/utilities.R to determine number of samples we had C. difficile CFU counts for per day
+five_day_mice <- metadata %>% filter(group == "C" & exp_num %in% c("M3","M4", "M5", "M8")| #Only use C mice from these experiments. Allocated groups to figures based on paper outline.
+             group == "WM" & exp_num %in% c("M3","M4", "M5", "M8")|
+             group == "WMC" & exp_num %in% c("M3","M4")|
+             group == "WMR" & exp_num %in% c("M5","M6"))
+five_d_cfu <- five_day_mice %>% 
+  filter(sample_type == "stool") %>% #Select only stool samples
+  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 8, 9, 10, 15, 20, 25, 30)) %>%  #select timepoints on plot
+  filter(!is.na(avg_cfu)) %>% #Remove rows with NA values
+  group_by(day) %>% 
+  count()
+#Figure out min & max number of samples we had cfu counts for
+five_d_cfu_n_min <- min(five_d_cfu$n)
+five_d_cfu_n_max <- max(five_d_cfu$n)
 
@@
 difficile*.** A.
-Setup of the experimental timeline for subset of experiments with 5-day PEG treated mice.
+Setup of the experimental timeline for experiments with 5-day PEG treated mice.
+Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
+15% PEG 3350 was administered in the drinking water for five days.
 B.
@@
 *C.
-difficile* CFU/gram stool measured over time (N = `r test`-`(insert variable name)` mice per timepoint) via serial dilutions.
+difficile* CFU/gram stool measured over time (N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per timepoint) via serial dilutions.
 The black line represents the limit of detection for the first serial dilution.
 CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths.
-Lines represent the median for each source and circles represent individual mouse samples.
+For B-C, lines represent the median for each treatment group and circles represent samples from individual mice.
 Asterisks indicate timepoints where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple timepoints.
@@
 
+#Use same code from one_day_PEG_subset() in code/utilities.R to determine number of samples per day
 one_day_samples_per_day <- metadata %>%
@@
   count() 
+#To Do: this code is not specific for CFU values or the timepoints shown on panel B
 one_day_sample_min <- min(one_day_samples_per_day$n)

### 2021-04-15 submission/manuscript.tex
Edit legend for figure 1

@@
 become persistently colonized with \emph{C. difficile}.} A. Setup of the
-experimental timeline for subset of experiments with 5-day PEG treated
-mice. B. Weight change from baseline weight in groups after treatment
-with PEG and/or clindamycin, followed by \emph{C. difficile} challenge.
-C. \emph{C. difficile} CFU/gram stool measured over time (N =
-4-\texttt{(insert\ variable\ name)} mice per timepoint) via serial
-dilutions. The black line represents the limit of detection for the
-first serial dilution. CFU quantification data was not available for
-each mouse due to stool sampling difficulties (particularly the day the
-mice came off of the PEG treatment) or early deaths. Lines represent the
-median for each source and circles represent individual mouse samples.
-Asterisks indicate timepoints where the weight change or CFU/g was
-significantly different between groups by the Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing multiple timepoints. \newpage
+experimental timeline for experiments with 5-day PEG treated mice.
+Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
+15\% PEG 3350 was administered in the drinking water for five days. B.
+Weight change from baseline weight in groups after treatment with PEG
+and/or clindamycin, followed by \emph{C. difficile} challenge. C.
+\emph{C. difficile} CFU/gram stool measured over time (N = 16-59 mice
+per timepoint) via serial dilutions. The black line represents the limit
+of detection for the first serial dilution. CFU quantification data was
+not available for each mouse due to stool sampling difficulties
+(particularly the day the mice came off of the PEG treatment) or early
+deaths. For B-C, lines represent the median for each treatment group and
+circles represent samples from individual mice. Asterisks indicate
+timepoints where the weight change or CFU/g was significantly different
+between groups by the Kruskal-Wallis test with Benjamini-Hochberg
+correction for testing multiple timepoints. \newpage
 

### 2021-04-15 code/permanova_stats.R
Remove previous Permanova script

@@ -1,103 +1,0 @@
-source("code/utilities.R") #Loads libraries, reads in metadata, functions
-source("code/16S_common_files.R") #Loads 16S output files for all sequenced samples
-
-set.seed(19760620) #Same seed used for mothur analysis
-
-#Remove variables not needed for this analysis
-rm(agg_otu_data, diversity_data)
-
-metadata <- metadata %>% 
-  mutate(day = factor(day, levels = c(unique(as.factor(day)), "-15", "-11", "-10", "-5", "-4", "-2", "-1", "0"))) #Transform day variable into factor variable
-
-#PERMANOVA of 5-days PEG subset----
-#Stools
-five_d_stools <- read_dist("data/process/5_day_PEG/stools/peg3350.opti_mcc.braycurtis.0.03.lt.ave.dist")
-five_d_stools_variables <- tibble(unique_label = attr(five_d_stools, "Labels")) %>%
-  left_join(metadata, by = "unique_label")
-five_d_stools_adonis <- adonis(five_d_stools~(group/(exp_num*unique_cage_no*ext_plate*miseq_run))*day, strata = five_d_stools_variables$unique_mouse_id, data = five_d_stools_variables, permutations = 1000, parallel = 32)
-five_d_stools_adonis_table <- as_tibble(rownames_to_column(five_d_stools_adonis$aov.tab, var = "effects")) %>% 
-  write_tsv("data/process/5_day_PEG_permanova_stools.tsv")
-#Tissues
-five_d_tissues <- read_dist("data/process/5_day_PEG/tissues/peg3350.opti_mcc.braycurtis.0.03.lt.ave.dist")
-five_d_tissues_variables <- tibble(unique_label = attr(five_d_tissues, "Labels")) %>%
-  left_join(metadata, by = "unique_label")
-#PERMANOVA of all relevant variables (exclude Miseq run since they were the same for all samples in this subset)
-five_d_tissues_adonis <- adonis(five_d_tissues~(group/(sample_type*exp_num*unique_cage_no*ext_plate))*day, strata = five_d_tissues_variables$unique_mouse_id, data = five_d_tissues_variables, permutations = 1000, parallel = 32)
-five_d_tissues_adonis
-#Write PERMANOVA results to tsv
-five_d_tissues_adonis_table <- as_tibble(rownames_to_column(five_d_tissues_adonis$aov.tab, var = "effects")) %>% 
[…]

### 2021-04-15 submission/manuscript.Rmd
Add to results section

@@
 **5-day laxative treatment leads to prolonged *C.
-difficile* colonization in mice.** We compared PEG-treated mice to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C.
+difficile* colonization in mice.** We compared mice treated with the osmotic laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C.
 difficile*, with mice typically clearing *C.
@@
 1A).
-PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment (Fig.
+PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice regained most of the weight five days after treatment (Fig.
 1B).
@@
 difficile* colonization (Fig.
-1C), however PEG-treated mice remained colonized at a high level through 30 days post-infection.
+1C), however PEG-treated mice remained colonized at a high level through thirty days post-infection.
 In contrast, the clindamycin-treated mice that cleared *C.
-difficile* within 10 days post-infection.
-Surprisingly, mice were still susceptible to *C.
-difficile* infection after 10-days of recovery from PEG treatment although *C.
-difficile* was not detectable in most of the group in the initial 5 days post-infection (Fig.
+difficile* within ten days post-infection.
+Surprisingly, PEG-treated mice were still susceptible to *C.
+difficile* infection after 10-days of recovery from treatment although *C.
+difficile* was not detectable in most of the group in the initial five days post-infection (Fig.
 1C).
@@
 Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C.
-difficile* colonization and PEG-treated mice remained susceptible for up to 10 days post PEG treatment.
+difficile* colonization and PEG-treated mice remained susceptible for up to ten days post-treatment.
 
@@
 f_d_sec_R2 = f_d_top_2_contrib[2,2]
+
+#Number of experiments represented in 5-day subset
+#Use same code from five_day_PEG_subset() in code/utilities.R to determine number of samples we had C. difficile CFU counts for per day
+five_day_mice <- metadata %>% filter(group == "C" & exp_num %in% c("M3","M4", "M5", "M8")| #Only use C mice from these experiments. Allocated groups to figures based on paper outline.
+             group == "WM" & exp_num %in% c("M3","M4", "M5", "M8")|
+             group == "WMC" & exp_num %in% c("M3","M4")|
+             group == "WMR" & exp_num %in% c("M5","M6"))
+#Total number of experiments in 5-day subset
+five_day_total_exp <- five_day_mice %>% 
+  filter(!duplicated(exp_num)) %>% #Remove duplicate experiment numbers
+  tally() #Total number of unique experiments
+
 ```
 **5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C.
-difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to their differential effects on the microbiota.
+difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to the two drugs having differential effects on the microbiota.
 We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups.
 We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X).
-The impact of cage effects has been well documented due to copraphagy and is expected.
+Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
+Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
+When we examine alpha diversity by looking at Shannon diversity index over time, we see that although both clindamycin and PEG treatments decrease diversity, the diversity index is less in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection.
 
@@
 Asterisks indicate timepoints where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple timepoints.
+The data presented are from a total of `r five_day_total_exp` separate experiments.
 \newpage
@@
 
-``` {r figure_4_permnova}
+``` {r figure_4_permanova}
 one_day_permnova_results <- read_tsv(here("data/process/1_day_PEG_permanova.tsv"))

### 2021-04-15 submission/manuscript.tex
Add to results section

@@
 \textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
-colonization in mice.} We compared PEG-treated mice to our standard 10
-mg/kg clindamycin treatment, which temporarily renders the mice
-susceptible to \emph{C. difficile}, with mice typically clearing
-\emph{C. difficile} within 10 days post-infection (9, 19). All
-PEG-treated mice were administered a 15\% PEG solution in the drinking
-water for 5-days, one group was also treated with clindamycin, and one
-group was allowed to recover for 10 days prior to challenge (Fig. 1A).
-PEG treatment resulted in weight loss in all 3 groups of PEG-treated
-mice, with the greatest change in weight observed on the fifth day of
-PEG treatment (Fig. 1B). After either PEG, clindamycin, or PEG and
-clindamycin treatment all mice were challenged with
-10\textsuperscript{3} \emph{C. difficile} 630 spores. All treatments
-rendered mice susceptible to \emph{C. difficile} colonization (Fig. 1C),
-however PEG-treated mice remained colonized at a high level through 30
-days post-infection. In contrast, the clindamycin-treated mice that
-cleared \emph{C. difficile} within 10 days post-infection. Surprisingly,
-mice were still susceptible to \emph{C. difficile} infection after
-10-days of recovery from PEG treatment although \emph{C. difficile} was
-not detectable in most of the group in the initial 5 days post-infection
-(Fig. 1C). From 9 days post-infection onward, the median \emph{C.
-difficile} stabilized for the 5-day PEG plus 10-day recovery group of
-mice and remained high through 30 days post-infection (Fig. 1C). Thus,
-osmotic laxative treatment alone was sufficient to render mice
-susceptible to prolonged \emph{C. difficile} colonization and
-PEG-treated mice remained susceptible for up to 10 days post PEG
-treatment.
+colonization in mice.} We compared mice treated with the osmotic
+laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
+temporarily renders the mice susceptible to \emph{C. difficile}, with
+mice typically clearing \emph{C. difficile} within 10 days
+post-infection (9, 19). All PEG-treated mice were administered a 15\%
+PEG solution in the drinking water for 5-days, one group was also
+treated with clindamycin, and one group was allowed to recover for 10
+days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
+in all 3 groups of PEG-treated mice, with the greatest change in weight
+observed on the fifth day of PEG treatment and the mice regained most of
+the weight five days after treatment (Fig. 1B). After either PEG,
+clindamycin, or PEG and clindamycin treatment all mice were challenged
+with 10\textsuperscript{3} \emph{C. difficile} 630 spores. All
+treatments rendered mice susceptible to \emph{C. difficile} colonization
+(Fig. 1C), however PEG-treated mice remained colonized at a high level
+through thirty days post-infection. In contrast, the clindamycin-treated
+mice that cleared \emph{C. difficile} within ten days post-infection.
+Surprisingly, PEG-treated mice were still susceptible to \emph{C.
+difficile} infection after 10-days of recovery from treatment although
+\emph{C. difficile} was not detectable in most of the group in the
+initial five days post-infection (Fig. 1C). From 9 days post-infection
+onward, the median \emph{C. difficile} stabilized for the 5-day PEG plus
+10-day recovery group of mice and remained high through 30 days
+post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was
+sufficient to render mice susceptible to prolonged \emph{C. difficile}
+colonization and PEG-treated mice remained susceptible for up to ten
+days post-treatment.
 
@@
 dynamics between mice treated with laxatives versus clindamycin were due
-to their differential effects on the microbiota. We profiled the stool
-microbiota over time by sequencing the V4 region of the 16S rRNA gene to
-compare changes across treatment groups. We found time and treatment
-group explained half of the observed variation between fecal communities
-(PERMANOVA combined R\textsuperscript{2} = 0.50, \emph{P} \textless{}
-0.001) with the remaining variation explained by interactions between
-treatment group and other experimental variables including time, cage
-effects, and sequencing preparation (Data Set S1, Sheet X). The impact
-of cage effects has been well documented due to copraphagy and is
-expected.
+to the two drugs having differential effects on the microbiota. We
+profiled the stool microbiota over time by sequencing the V4 region of
+the 16S rRNA gene to compare changes across treatment groups. We found
+time and treatment group explained half of the observed variation
+between fecal communities (PERMANOVA combined R\textsuperscript{2} =
+0.50, \emph{P} \textless{} 0.001) with the remaining variation explained
+by interactions between treatment group and other experimental variables
+including time, cage effects, and sequencing preparation (Data Set S1,
+Sheet X). Cage effects refer to the well-documented phenomenon that mice
+housed in the same cages have similar microbial communities due to
+copraphagy, we tried to minimize the impact of cage effects on our
+experiment by breaking up cagemates when assigning mice to treatment
+groups and primarily housing only two mice per cage. Importantly,
+although we conducted a total of 5 separate experiments, the experiment
+number and it's interaction with treatment group was not one of the
+variables that significantly explained the observed variation in fecal
+communities (Data Set S1, Sheet X). When we examine alpha diversity by
+looking at Shannon diversity index over time, we see that although both
+clindamycin and PEG treatments decrease diversity, the diversity index
+is less in the groups of mice that received PEG treatment compared to
+those that received clindamycin through thirty days post-infection.
 
@@
 between groups by the Kruskal-Wallis test with Benjamini-Hochberg
-correction for testing multiple timepoints. \newpage
+correction for testing multiple timepoints. The data presented are from
+a total of 5 separate experiments. \newpage
 

### 2021-04-17 submission/manuscript.Rmd
Fig 5 caption draft

@@
 \includegraphics{figure_5.pdf}
+A.
+Experimental timeline for groups receiving PEG treatment post clindamycin gavage.
+B.
+Median and individual CFU/g of C.
+difficile measured over time via culture.
+Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction.
+Limit of detection is 1e2 CFU/g.
+Background shading indicates PEG treatment period for applicable groups.
+Opacity of points reflects time point.
+C.
+Median and individual Shannon Diversity measured over time.
+Asterisks, background shading, and opacity of points consistent with (B).
+D.
+Median and individual richness measured over time.
+Asterisks, background shading, and opacity of points consistent with (B).
+
 \newpage
 \includegraphics{figure_5_16S.pdf}
-**Figure 5. 1-day PEG treatment post C. difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A. 
+**Figure 5.
+1-day PEG treatment post C.
+difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A.
+PCoA of all groups + FMT plotted over time.
+Opacity of points reflects time point.
+B.
+Relative abundance of genera significantly different over multiple time points post FMT or PBS treatment, plotted over time.
+Limit of detection is .1%.
+C.
+Relative abundances of top 6 significant genera ranked by number of days significant, plotted over time.
+Limit of detection is .1%.
+
 \newpage

### 2021-04-19 submission/manuscript.Rmd
Add to results

@@
 
+#PEG impacted genera
+PEG_genera <- read_tsv(here("data/process/5_days_PEG_genus_PEG_paired.tsv")) %>% 
+  filter(p.value.adj < 0.05) %>% #Limit to significant genera
+  tally()
+
 ```
@@
 We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups.
-We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X).
+We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig.
+2A) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X).
 Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
 Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
-When we examine alpha diversity by looking at Shannon diversity index over time, we see that although both clindamycin and PEG treatments decrease diversity, the diversity index is less in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection.
+Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C.
+difficile* colonization in PEG-treated mice (Fig.
+2B).
+Next, we examined  alpha diversity by looking at Shannon diversity index over time, we see that although both clindamycin and PEG treatments decrease diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
+2C).
+We next examined the bacterial genera that shifted after PEG treatment by analyzing what changed in mice treated with PEG alone between baseline and one day post-treatment.
+We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the impact on *Bacteroides* was different compared to clindamycin treated mice (Fig.
+2D).
+Thus, PEG has detrimental impacts on the microbiome that are maintained over time.
+Surprisingly, *C.
+difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge.
+We decided to examine what other bacteria were changing during the post-infection period when *C.
+difficile* changed from undectable to detectable.
+Interestingly, we found ___ bacteria were shifting during the time period when C.
+difficile was becoming detectable in the stools (Fig.
+S1)
 
-**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given previous work has shown that PEG treatment disrupts the mucus layer, we decided to examine the impact of the laxative on the mucosal microbiota and infection severity by evaluating histopathology.
+**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given that previous work had demonstrated that PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
+We examined CDI severity by evaluating cecum and colon histopathology [].
 
-* Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.
+***C. difficile* challenge does not have a synergistic disruptive effect on the microbiota of PEG-treated mice**
+Because *C.
+difficile* itself can have an impact on the microbiota (ref), we also sequenced the tissue and stools of mock-challenged mice.
+Examining the stools of the mock-infected PEG- and clindamycin-treated mice revealed similar bacterial disruptions as the *C.
+difficile* challenged mice (Fig.
+S2A).
+Similarly, there was no difference between the tissues of mock and *C.
+difficile* challenged mice (Fig.
+S2B).
+Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C.
+difficile*.
 
-* Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.
+**1-day laxative treatment results in transient *C. difficile* colonization and minor microbiota disruption**
+  + Explain motivation for experiment, set-up of experiment (Fig. 4A). 
+  + Transient *C. difficile* colonization in 1-day PEG treated mice with all PEG mice clearing by 7dpi (Fig. 4B)
+  + PCoA results (Day has larger R2 than treatment group). PEG mice are closer to baseline communites after 7-day period (Fig. 4C)
+  + Shannon results show alpha diversity is only transiently disrupted (Fig. 4D)
+  + Highlight bacteria that are disrupted by PEG or clindamycin treatment but recover within 7 days (Fig. 4E)
 
-* Figure 3. 5-day PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.
+**Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered**
+  + Reiterate motivation for experiment.
+Saw only a transient microbiota disruption with 1-day PEG treatment alone.
+Also, see if adding FMT helps with clearance, mention groups have been using PEG or Abx/PEG treatments as a tool to engraft mice with human bacteria communities.
+Set up of experiment (Fig.
+5A).
+  + Prolonged *C. difficile* colonization in post-CDI PEG treated mice (Fig. 5B)
+  + FMT appears to partially restore alpha diversity (Shannon, but not richness Fig. 5C-D)
+  + PCoA (necessary?) Could comment on the clustering of Clindamycin and PEG-treated mice that received FMT (Fig. 5E)
+  + Only 2 genera significantly impacted by FMT treatment, likely not as important for clearance (Fig. 5F)
+  + Bacteria that consistently differ between groups over time, associated with prolonged colonization (Fig. G)
 
-* Figure 4. 1-day PEG treatment renders mice susceptible to transient *C. difficile* colonization.
-
-* Figure 5. 1-day PEG treatment post *C. difficile* challenge prolongs colonization regardless of whether an FMT is also administered.
+**Five-day post-infection community data can predict mice that will have prolonged *C. difficile* colonization** After identifying
 
@@
 
-* Figure 7. Schematic summarizing findings.
-
 ## Discussion
 
-* Summary of major findings
+* Summary of major findings (Fig. 7A)
 
@@
 
-* Discuss what these findings might mean for human patients
+* Discuss what these findings might mean for human patients (Fig. 7B)
   + What's known regarding laxatives and susceptibility to CDIs

### 2021-04-19 code/relative_dist_to_baseline.R
Decrease font size in Bray-Curtis plots

@@ -61,7 +61,7 @@ plot_bc_dist <- function(input_df, timepoint){
     theme_classic()+
     lims(y = c(0, 1))+
     theme(legend.position = "none",
-          text = element_text(size = 16),# Change font size for entire plot
+          text = element_text(size = 14),# Change font size for entire plot
           axis.text.x = element_blank(),
           axis.ticks.x = element_blank()) #Remove x axis ticks
 }

### 2021-04-19 submission/manuscript.Rmd
Update results description for 5-day PEG PERMANOVA

@@
 f_d_permanova <- read_tsv(here("data/process/5_days_PEG_permanova_stools.tsv"))
-f_d_top_2_contrib <- f_d_permanova %>% arrange(desc(R2)) %>%
+f_d_sign_contrib <- f_d_permanova %>% arrange(desc(R2)) %>%
   filter(!effects == "Total") %>% #Remove total since it is not one of the variables
-  top_n(2, R2) %>% #Select top 2 variables based on R2
-  select(effects, R2, `Pr(>F)`) #Select relevant column names
+  filter(`Pr(>F)` < 0.05) #Only select significant contributors
 
-f_d_top_contrib= (f_d_top_2_contrib[1,1])
-f_d_sec_contrib = (f_d_top_2_contrib[2,1])
-f_d_top_R2_sum = f_d_top_2_contrib %>% 
+#List of variables that significantly explained variation in stool samples
+f_d_top_effects <- f_d_sign_contrib %>% 
+  pull(effects)
+#Sum of R2 values for variables that significantly explained variation in stool samples
+f_d_top_R2_sum <- f_d_sign_contrib %>% 
   select(R2) %>% sum
-f_d_top_p <- f_d_top_2_contrib %>% 
+#P-value for variables that significantly explained variation in stool samples
+f_d_top_p <- f_d_sign_contrib %>% 
   select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text
-f_d_top_R2 = f_d_top_2_contrib[1,2]
-f_d_sec_R2 = f_d_top_2_contrib[2,2]
 
@@
 We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups.
-We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig.
-2A) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X).
+We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig.
+2A, Data Set S1, Sheet X).
 Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
@@
 2B).
-Next, we examined  alpha diversity by looking at Shannon diversity index over time, we see that although both clindamycin and PEG treatments decrease diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
+Next, we examined  alpha diversity by looking at Shannon diversity index over time, although both clindamycin and PEG treatments decreased diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
 2C).
-We next examined the bacterial genera that shifted after PEG treatment by analyzing what changed in mice treated with PEG alone between baseline and one day post-treatment.
-We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the impact on *Bacteroides* was different compared to clindamycin treated mice (Fig.
+We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment.
+We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as Bacteroides actually decreased in clindamycin treated mice (Fig.
 2D).
-Thus, PEG has detrimental impacts on the microbiome that are maintained over time.
+Finally, we examined the bacteria that differ across treatment groups over multiple timepoints.
+We found...
+(Fig.
+2E).
+Thus, PEG has a significant impact on the fecal microbiota that was maintained over time.
+
 Surprisingly, *C.
@@
 \includegraphics{figure_2.pdf}
-**Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. 
+**Figure 2.
+5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A.
+PCoA of Bray-Curtis distances from stool samples collected throughout the experiment.
 \newpage

### 2021-04-19 submission/manuscript.tex
Update results description for 5-day PEG PERMANOVA

@@
 time and treatment group explained half of the observed variation
-between fecal communities (PERMANOVA combined R\textsuperscript{2} =
-0.50, \emph{P} \textless{} 0.001) with the remaining variation explained
+between fecal communities with most of the remaining variation explained
 by interactions between treatment group and other experimental variables
-including time, cage effects, and sequencing preparation (Data Set S1,
-Sheet X). Cage effects refer to the well-documented phenomenon that mice
-housed in the same cages have similar microbial communities due to
-copraphagy, we tried to minimize the impact of cage effects on our
-experiment by breaking up cagemates when assigning mice to treatment
-groups and primarily housing only two mice per cage. Importantly,
-although we conducted a total of 5 separate experiments, the experiment
-number and it's interaction with treatment group was not one of the
-variables that significantly explained the observed variation in fecal
-communities (Data Set S1, Sheet X). When we examine alpha diversity by
-looking at Shannon diversity index over time, we see that although both
-clindamycin and PEG treatments decrease diversity, the diversity index
-is less in the groups of mice that received PEG treatment compared to
-those that received clindamycin through thirty days post-infection.
+including time, cage effects, and sequencing preparation (PERMANOVA
+combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
+2A, Data Set S1, Sheet X). Cage effects refer to the well-documented
+phenomenon that mice housed in the same cages have similar microbial
+communities due to copraphagy, we tried to minimize the impact of cage
+effects on our experiment by breaking up cagemates when assigning mice
+to treatment groups and primarily housing only two mice per cage.
+Importantly, although we conducted a total of 5 separate experiments,
+the experiment number and it's interaction with treatment group was not
+one of the variables that significantly explained the observed variation
+in fecal communities (Data Set S1, Sheet X). Interestingly, none of the
+treatment groups recovered to their baseline community structure either
+10 or 30 days post-infection suggesting other community features besides
+recovery to baseline were responsible for the prolonged \emph{C.
+difficile} colonization in PEG-treated mice (Fig. 2B). Next, we examined
+alpha diversity by looking at Shannon diversity index over time,
+although both clindamycin and PEG treatments decreased diversity, the
+diversity index was lower in the groups of mice that received PEG
+treatment compared to those that received clindamycin through thirty
+days post-infection (Fig. 2C). We next examined the bacterial genera
+that shifted after PEG treatment by comparing the baseline samples of
+mice treated with only PEG to samples from the same mice one day
+post-treatment. We found 18 OTUs that were altered by PEG, the majority
+of these OTUs decreased after PEG, but \emph{Enterobacteriaceae} and
+\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
+unique to PEG treatment as Bacteroides actually decreased in clindamycin
+treated mice (Fig. 2D). Finally, we examined the bacteria that differ
+across treatment groups over multiple timepoints. We found\ldots{} (Fig.
+2E). Thus, PEG has a significant impact on the fecal microbiota that was
+maintained over time.
+
+Surprisingly, \emph{C. difficile} was not immediately detectable in the
+stools of the PEG-treated mice that were allowed to recover for 10 days
+prior to challenge. We decided to examine what other bacteria were
+changing during the post-infection period when \emph{C. difficile}
+changed from undectable to detectable. Interestingly, we found \_\_\_
+bacteria were shifting during the time period when C. difficile was
+becoming detectable in the stools (Fig. S1)
 
 \textbf{5-day laxative treatment does not promote more severe CDIs
-despite altering the mucosal microbiota.} Given previous work has shown
-that PEG treatment disrupts the mucus layer, we decided to examine the
-impact of the laxative on the mucosal microbiota and infection severity
-by evaluating histopathology.
+despite altering the mucosal microbiota.} Given that previous work had
+demonstrated that PEG treatment disrupts the mucus layer and alters the
+immune response in mice (16), we decided to examine the impact of PEG
+treatment on the mucosal microbiota and CDI severity. We examined CDI
+severity by evaluating cecum and colon histopathology {[}{]}.
+
+\textbf{\emph{C. difficile} challenge does not have a synergistic
+disruptive effect on the microbiota of PEG-treated mice} Because
+\emph{C. difficile} itself can have an impact on the microbiota (ref),
+we also sequenced the tissue and stools of mock-challenged mice.
+Examining the stools of the mock-infected PEG- and clindamycin-treated
+mice revealed similar bacterial disruptions as the \emph{C. difficile}
+challenged mice (Fig. S2A). Similarly, there was no difference between
+the tissues of mock and \emph{C. difficile} challenged mice (Fig. S2B).
+Thus, most of the microbiota alterations we observed in the PEG-treated
+mice were a result of the laxative and not an interaction between the
+laxative and \emph{C. difficile}.
+
+\textbf{1-day laxative treatment results in transient \emph{C.
+difficile} colonization and minor microbiota disruption} + Explain
+motivation for experiment, set-up of experiment (Fig. 4A). + Transient
+\emph{C. difficile} colonization in 1-day PEG treated mice with all PEG
+mice clearing by 7dpi (Fig. 4B) + PCoA results (Day has larger R2 than
+treatment group). PEG mice are closer to baseline communites after 7-day
+period (Fig. 4C) + Shannon results show alpha diversity is only
+transiently disrupted (Fig. 4D) + Highlight bacteria that are disrupted
+by PEG or clindamycin treatment but recover within 7 days (Fig. 4E)
+
+\textbf{Post-CDI laxative treatment disrupts clearance in
+clindamycin-treated mice regardless of whether an FMT is also
+administered} + Reiterate motivation for experiment. Saw only a
+transient microbiota disruption with 1-day PEG treatment alone. Also,
+see if adding FMT helps with clearance, mention groups have been using
+PEG or Abx/PEG treatments as a tool to engraft mice with human bacteria
+communities. Set up of experiment (Fig. 5A). + Prolonged \emph{C.
+difficile} colonization in post-CDI PEG treated mice (Fig. 5B) + FMT
+appears to partially restore alpha diversity (Shannon, but not richness
+Fig. 5C-D) + PCoA (necessary?) Could comment on the clustering of
+Clindamycin and PEG-treated mice that received FMT (Fig. 5E) + Only 2
+genera significantly impacted by FMT treatment, likely not as important
+for clearance (Fig. 5F) + Bacteria that consistently differ between
+groups over time, associated with prolonged colonization (Fig. G)
+
+\textbf{Five-day post-infection community data can predict mice that
+will have prolonged \emph{C. difficile} colonization} After identifying
 
 \begin{itemize}
-\item
-  Figure 2. 5-day PEG treatment disrupts the stool microbiota for a
-  longer amount of time compared to clindamycin-treated mice.
-\item
-  Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota
-  dynamics post-infection.
-\item
-  Figure 3. 5-day PEG treatment does not result in more severe CDIs,
-  although mucosal microbiota is altered.
-\item
-  Figure 4. 1-day PEG treatment renders mice susceptible to transient
-  \emph{C. difficile} colonization.
-\item
-  Figure 5. 1-day PEG treatment post \emph{C. difficile} challenge
-  prolongs colonization regardless of whether an FMT is also
-  administered.
 \item
@@
   \end{itemize}
-\item
-  Figure 7. Schematic summarizing findings.
 \end{itemize}
@@
 \item
-  Summary of major findings
+  Summary of major findings (Fig. 7A)
 \item
@@
 \item
-  Discuss what these findings might mean for human patients
+  Discuss what these findings might mean for human patients (Fig. 7B)
 
@@
 disrupts the stool microbiota for a longer amount of time compared to
-clindamycin-treated mice.} A. \newpage
+clindamycin-treated mice.} A. PCoA of Bray-Curtis distances from stool
+samples collected throughout the experiment. \newpage
 
@@
 
-\includegraphics{figure_5.pdf} \newpage
+\includegraphics{figure_5.pdf} A. Experimental timeline for groups
+receiving PEG treatment post clindamycin gavage. B. Median and
+individual CFU/g of C. difficile measured over time via culture.
+Asterisks indicate time points with significant differences between
+groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction.
+Limit of detection is 1e2 CFU/g. Background shading indicates PEG
+treatment period for applicable groups. Opacity of points reflects time
+point. C. Median and individual Shannon Diversity measured over time.
+Asterisks, background shading, and opacity of points consistent with
+(B). D. Median and individual richness measured over time. Asterisks,
+background shading, and opacity of points consistent with (B).
+
+\newpage
+
 \includegraphics{figure_5_16S.pdf} \textbf{Figure 5. 1-day PEG treatment
 post C. difficile challenge prolongs colonization regardless of whether
-an FMT is also administered.} A. \newpage
+an FMT is also administered.} A. PCoA of all groups + FMT plotted over
+time. Opacity of points reflects time point. B. Relative abundance of
+genera significantly different over multiple time points post FMT or PBS
+treatment, plotted over time. Limit of detection is .1\%. C. Relative
+abundances of top 6 significant genera ranked by number of days
+significant, plotted over time. Limit of detection is .1\%.
+
+\newpage
 

### 2021-04-20 code/5_days_PEG_16S.R
Add diversity analysis of mock challenged mice

@@ -25,9 +25,6 @@ diversity_stools <- subset_stool(diversity_subset) %>%
   mutate(day = as.numeric(day))
 diversity_tissues <- subset_tissue(diversity_subset) %>%
   mutate(day = as.numeric(day))
-#Also create dataframes of diversity data that includes mock challenged mice (WMN and C), separated into stool and tissue samples
-diversity_mock_stools <- subset_stool(add_mocks(diversity_subset, diversity_data))
-diversity_mock_tissues <- subset_tissue(add_mocks(diversity_subset, diversity_data))
 
 #Figure out how many samples we have per group per day for each subset
 num_stool <- count_subset(diversity_stools) #Number of stool samples per group per day
@@ -38,9 +35,7 @@ num_mock_tissue <- count_subset(diversity_mock_tissues) #Number of stool samples
 #Experimental days to analyze with the Kruskal-Wallis test (timepoints with 16S data for at least 3 groups)
 #Baseline (before treatment) for WMR is day -15. For C, WM, and WMC baseline is day -5
 stool_test_days <- c(-5, -1, 0, 1, 2, 3, 4, 5, 6, 10, 30)
-#stool_mock_test_days #Don't test, included in code/all_subsets_16S.R
 tissue_test_days <- c(6, 30) #Only 2 days with samples from at least 3 groups
-#tissue_mock_test_days #Don't test, included in code/all_subsets_16S.R
 
 #Function to perform Kruskal-Wallis test for differences in Shannon diversity index across groups on a particular day with Benjamini Hochberg correction
 #Arguments: 
@@ -58,7 +53,7 @@ kruskal_wallis_shannon <- function(diversity_subset, timepoints, subset_name){
     unnest(c(model, median)) %>% 
     ungroup()  
   diversity_stats_adjust <- diversity_stats %>% 
-    select(day, statistic, p.value, parameter, method, C, WM, WMC, WMR) %>% 
+    select(-data) %>% 
     mutate(p.value.adj=p.adjust(p.value, method="BH")) %>% 
     arrange(p.value.adj) %>% 
     write_tsv(path = paste0("data/process/5_days_PEG_shannon_stats_", subset_name, "_subset.tsv"))
@@ -86,7 +81,7 @@ kruskal_wallis_richness <- function(diversity_subset, timepoints, subset_name){
     unnest(c(model, median)) %>% 
     ungroup()  
   diversity_stats_adjust <- diversity_stats %>% 
[…]

### 2021-04-20 code/figure_S2.R

@@ -1,6 +1,6 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/otus_potential_muribaculum_blast_results.png")
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_mock.png")
 
 plot_grid(a, labels = c("A"), label_size = 12, ncol=1)+
   ggsave("results/figures/figure_S2.pdf", width=6.875, height=6)+

### 2021-04-20 code/subset_analysis.R
Narrow down mock PCoAs to just 4 groups of interest: C, CN, WM, WMN

@@ -83,7 +83,7 @@ pcoa_5_day_PEG <- pcoa_data %>%
 #Concatenate output and add - between each sample.
 five_day_PEG_unique_labels <- mothur_format(pcoa_5_day_PEG)
 
-#Split 5 day PEG into stool (+FMT) and tissue subsets:
+#Split 5 day PEG into stool and tissue subsets:
 five_day_PEG_stool_labels <- mothur_format(pcoa_5_day_PEG %>% filter(sample_type == "stool") %>% 
                                              filter(!group %in% c("WMN", "CN"))
                                            )
@@ -92,10 +92,14 @@ five_day_PEG_tissue_labels <-   mothur_format(pcoa_5_day_PEG %>% filter(!sample_
                                                 filter(!group %in% c("WMN", "CN"))
                                               )
 
-#Create subsets without mock samples
-five_day_PEG_stool_mock_labels <- mothur_format(pcoa_5_day_PEG %>% filter(sample_type == "stool"))
+#Create subsets with mock samples
+five_day_PEG_stool_mock_labels <- mothur_format(pcoa_5_day_PEG %>% filter(sample_type == "stool") %>% 
+                                                  #For these PCoAs we just want to compare clindamycin +/- C. difficile and 5-day PEG +/- C. difficilie
+                                                  filter(!group %in% c("WMC", "WMR")))
 
-five_day_PEG_tissue_mock_labels <- mothur_format(pcoa_5_day_PEG %>% filter(!sample_type == "stool")) 
+five_day_PEG_tissue_mock_labels <- mothur_format(pcoa_5_day_PEG %>% filter(!sample_type == "stool")%>% 
+                                                   #For these PCoAs we just want to compare clindamycin +/- C. difficile and 5-day PEG +/- C. difficilie
+                                                   filter(!group %in% c("WMC", "WMR"))) 
 
 #Post-CDI PEG samples with FMT solution & Post-CDI PEG tissues added----
 pcoa_post_CDI_PEG <- pcoa_data %>% 

### 2021-04-20 code/5_days_PEG_16S.R
Add diversity plots of tissue samples from mock challenged mice

@@ -927,10 +927,10 @@ lp_tissue_mock <- line_plot_genus(genus_mock_tissues %>%
   scale_x_continuous(limits = c(3.5, 30.5), breaks = c(4, 6, 30), labels = c(4, 6, 30))#Change scale since we have less timepoints for tissues
 save_plot(filename = "results/figures/5_days_PEG_genus_lineplot_mock_tissues.png", lp_tissue_mock, base_height = 5, base_width = 8)
 
-#Analysis of mock-infected mice compared to their corresponding group that was challenged with C. difficile
+#Analysis of mock-infected mice compared to their corresponding group that was challenged with C. difficile----
 #Restrict mock analysis to C, WM, CN, and WMN mice. 
 
-#Custom scale for mock plots
+#Custom scale for mock plots----
 color_scheme_m <- c("#238b45", "#88419d", "#238b45", "#88419d")
 color_groups_m <- c("CN", "WMN", "C", "WM")
 color_labels_m <- c("Clind. without infection", "5-day PEG without infection", "Clind.", "5-day PEG")
@@ -1020,6 +1020,32 @@ shannon_mock_stools_plot <- diversity_mock_stools %>%
   scale_x_continuous(breaks = c(-5, -1, 0, 4, 6, 30),
                      limits = c(-5.5,31))
 save_plot(filename = "results/figures/5_days_PEG_shannon_stools_mock.png", shannon_mock_stools_plot, base_height = 4, base_width = 9, base_aspect_ratio = 2)
+#Plot of shannon diversity in mock tissues over time
+shannon_mock_tissues_plot <- diversity_mock_tissues %>% 
+  mutate(day = as.integer(day)) %>% 
+  group_by(group, day) %>%
+  mutate(median_shannon = median(shannon)) %>%
+  ggplot(x = day, y = shannon, colour = group)+
+  geom_line(mapping = aes(x = day, y = median_shannon, group = group, color = group, linetype = infected), alpha = 0.6, size = 1) +
+  scale_colour_manual(name=NULL,
+                      values=color_scheme_m,
+                      breaks=color_groups_m,
+                      labels=color_labels_m) +
+  scale_linetype_manual(name="Infected", #Dashed lines are mock challenged mice
+                        values=linetype_scheme_m,
+                        breaks=linetype_infected,
+                        labels=linetype_infected) +
+  scale_y_continuous(limits = c(0,4.1), expand = c(0, 0))+ #expand argument gets rid of the extra space around the scale
+  theme_classic()+
[…]

### 2021-04-20 code/5_days_PEG_16S.R
Add PCoAs for mock challenge comparisons

@@ -1145,4 +1145,91 @@ richness_mock_tissues_plot <- diversity_mock_tissues %>%
                      limits = c(-0.5,31))
 save_plot(filename = "results/figures/5_days_PEG_richness_tissues_mock.png", richness_mock_tissues_plot, base_height = 4, base_width = 9, base_aspect_ratio = 2)
 
-#PCoA analysis of mock challenged mice
+#PCoA analysis of mock challenged mice----
+#Define shape scheme based on Infected status
+shape_scheme <- c(1, 19)
+shape_infected <- c("no", "yes")
+
+#Function to plot PCoA data for mock-challenged mice comparison
+plot_mock_pcoa <- function(df){
+  ggplot(df, aes(x=axis1, y=axis2, color = group, alpha = day, shape = infected)) +
+    geom_point(size=2) +
+    scale_colour_manual(name=NULL,
+                        values=color_scheme_m,
+                        breaks=color_groups_m,
+                        labels=color_labels_m)+ 
+    scale_shape_manual(name="Infected",
+                       values=shape_scheme,
+                       breaks=shape_infected,
+                       labels=shape_infected) +
+    coord_fixed() +
+    labs(x="PCoA 1",
+         y="PCoA 2",
+         alpha= "Day") +
+    theme_classic()+
+    theme(legend.position = "bottom",
+          text = element_text(size = 16))
+}
+
+pcoa_mock_stool <- read_tsv("data/process/5_day_PEG/stools_mock/peg3350.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes") %>%
+  select(group, axis1, axis2) %>% #Limit to 2 PCoA axes
+  rename("unique_label" = group) %>%
+  left_join(metadata, by= "unique_label") %>% #merge metadata and PCoA data frames
+  mutate(day = as.integer(day)) %>% #Transform day into continuous
+  filter(!is.na(axis1)) #Remove all samples that weren't sequenced or were sequenced and didn't make the subsampling cutoff
+
+pcoa_axes_5_day_PEG_stool <- read_tsv("data/process/5_day_PEG/stools_mock/peg3350.opti_mcc.braycurtis.0.03.lt.ave.pcoa.loadings")
[…]

### 2021-04-20 code/5_days_PEG_16S.R
Update lineplots of bacterial genera in mock mice to include corresponding challenged groups

@@ -898,35 +898,6 @@ akkermansia_tissues <- genus_over_time("Akkermansia", genus_tissues_t)+
   theme(legend.position = "bottom")
 save_plot(filename = "results/figures/5_days_PEG_genus_akkermansia_tissues.png", akkermansia_tissues, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
-#Plots of mock-infected mice----
-#Use same color scheme but change symbols to open circles and lines to dashes
-#Define color scheme to match 5_days_PEG plots----
-color_scheme <- c("#238b45", "#88419d") #Adapted from http://colorbrewer2.org/#type=sequential&scheme=BuPu&n=4
-color_groups <- c("CN", "WMN")
-color_labels <- c( "Clind.", "5-day PEG 3350")
-
-#Lineplots of the top 6 significant genera in mock stool samples----
-lp_stool_days <- c("-5","-1", "0", "4", "6")
-facet_labels <- top_10_sig_genus[1:6] #Pick just the top 6
-names(facet_labels) <- top_10_sig_genus[1:6] #Pick just the top 6
-lp_stool_mock <- line_plot_genus(genus_mock_stools %>% 
-                              filter(group %in% c("CN","WMN")), #Select only mock challenged groups
-                            top_10_sig_genus[1:6], lp_stool_days, "longdash")+
-  scale_x_continuous(limits = c(-5.5,6.5), breaks = c(-5, -1, 0, 4, 6), labels = c(-5, -1, 0, 4, 6))#Change scale
-save_plot(filename = "results/figures/5_days_PEG_genus_lineplot_mock_stools.png", lp_stool_mock, base_height = 5, base_width = 8)
-
-#Lineplots of the top 6 significant genera in mock tissue samples----
-lp_tissue_days <- diversity_tissues %>% distinct(day) %>% 
-  filter(day %in% c("4", "6", "30")) %>% #Focus on day -1 through 10 timepoints
-  pull(day)
-facet_labels <- top_sig_genus_tissues[1:6] #Pick just the top 6
-names(facet_labels) <- top_sig_genus_tissues[1:6] #Pick just the top 6
-lp_tissue_mock <- line_plot_genus(genus_mock_tissues %>% 
-                                    filter(group %in% c("CN","WMN")), #Select only mock challenged groups 
-                                  top_sig_genus_tissues[1:6], lp_tissue_days, "longdash")+
[…]

### 2021-04-20 code/figure_S2.R
Add plots to figure comparing mock-challenged mice

@@ -1,7 +1,19 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_mock.png")
-
-plot_grid(a, labels = c("A"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S2.pdf", width=6.875, height=6)+
-  ggsave("submission/figure_S2.pdf", width=6.875, height=6)
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_stool_PCoA_mock.png")
+a_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_legend.png")
+a <- plot_grid(a, a_legend, ncol = 1, rel_heights = c(1, .2))
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_tissue_PCoA_mock.png")
+b_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_alpha_legend_tissues.png")
+b <- plot_grid(b, b_legend, ncol = 1, rel_heights = c(1, .1))
+top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow=1)
+c <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_mock.png")
+d <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues_mock.png")
+middle_panel <- plot_grid(c, d, labels = c("C", "D"), label_size = 12, ncol=2)
+e <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_stools.png")
+f <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_tissues.png")
+bottom_panel <- plot_grid(e, f, labels = c("E", "F"), label_size = 12, nrow=1)
+                       
+plot_grid(top_panel, middle_panel, bottom_panel, labels = NULL, label_size = 12, ncol=1, rel_heights = c(1.4, 1, 1))+
+  ggsave("results/figures/figure_S2.pdf", width=6.875, height=7)+
+  ggsave("submission/figure_S2.pdf", width=6.875, height=7)

### 2021-04-20 code/figure_S2.R
Relabel panels

@@ -3,16 +3,16 @@ source("code/utilities.R") #Loads libraries, reads in metadata, functions
 a <- ggdraw() + draw_image("results/figures/5_days_PEG_stool_PCoA_mock.png")
 a_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_legend.png")
 a <- plot_grid(a, a_legend, ncol = 1, rel_heights = c(1, .2))
-b <- ggdraw() + draw_image("results/figures/5_days_PEG_tissue_PCoA_mock.png")
-b_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_alpha_legend_tissues.png")
-b <- plot_grid(b, b_legend, ncol = 1, rel_heights = c(1, .1))
-top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow=1)
-c <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_mock.png")
-d <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues_mock.png")
-middle_panel <- plot_grid(c, d, labels = c("C", "D"), label_size = 12, ncol=2)
-e <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_stools.png")
+d <- ggdraw() + draw_image("results/figures/5_days_PEG_tissue_PCoA_mock.png")
+d_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_alpha_legend_tissues.png")
+d <- plot_grid(d, d_legend, ncol = 1, rel_heights = c(1, .1))
+top_panel <- plot_grid(a, d, labels = c("A", "D"), label_size = 12, nrow=1)
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_mock.png")
+e <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues_mock.png")
+middle_panel <- plot_grid(b, e, labels = c("B", "E"), label_size = 12, ncol=2)
+c <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_stools.png")
 f <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_tissues.png")
-bottom_panel <- plot_grid(e, f, labels = c("E", "F"), label_size = 12, nrow=1)
[…]

### 2021-04-20 code/5_days_PEG_16S.R
Remove alpha legend from PCoA

@@ -214,7 +214,7 @@ pcoa_subset_plot_tissue <- plot_pcoa(pcoa_5_day_PEG_tissues)+
   scale_alpha_continuous(range = c(.3, 1),
                          breaks= c(4, 6, 20, 30),
                          labels=c(4, 6, 20, 30))+
-  theme(legend.position = NULL)
+  theme(legend.position = "none")
 save_plot(filename = paste0("results/figures/5_days_PEG_tissues_PCoA.png"), pcoa_subset_plot_tissue, base_height = 5, base_width = 5)
 
 #PCoA faceted over time

### 2021-04-20 code/figure_3.R
Figure edits

@@ -1,15 +1,13 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/5_days_PEG_histo_scores_d4.png")
-b <- ggdraw() + draw_image("results/figures/5_days_PEG_histo_scores_d6.png")
-c <- ggdraw() + draw_image("results/figures/5_days_PEG_tissues_PCoA.png")
-d <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues.png")
-e <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_tissues.png")
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues.png")
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_tissues_PCoA.png")
+c <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_tissues.png")
+middle_panel <- plot_grid(b, c, labels = c("B", "C"), label_size = 12, nrow = 1, rel_widths = c(.6, 1))
+d <- ggdraw() + draw_image("results/figures/5_days_PEG_histo_scores_d4.png")
+e <- ggdraw() + draw_image("results/figures/5_days_PEG_histo_scores_d6.png")
+bottom_panel <- plot_grid(d, e, labels = c("D", "E"), label_size = 12, nrow = 1, rel_heights = c(1.2, 1))
 
-left_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol = 1, rel_widths = c(1.5, 1))
-right_panel <- plot_grid(c, labels = c("C"), label_size = 12, nrow = 1)
-top_panel <- plot_grid(left_panel, right_panel, nrow = 1)
-bottom_panel <- plot_grid(d, e, labels = c("D", "E"), label_size = 12, nrow = 2)
-plot_grid(top_panel, bottom_panel, nrow = 2, rel_widths = c(.5, 3)) +
+plot_grid(a, middle_panel, bottom_panel, labels = c("A", "", ""), nrow = 3) +
   ggsave("results/figures/figure_3.pdf", width = 6.875, height = 9)+
   ggsave("submission/figure_3.pdf", width = 6.875, height = 9)

### 2021-04-20 submission/manuscript.Rmd
Add to results

@@
 
-``` {r 5-day PEG microbiota}
+``` {r 5-day PEG fecal microbiota}
 f_d_permanova <- read_tsv(here("data/process/5_days_PEG_permanova_stools.tsv"))
@@
 
+#Genera that were different between treatment groups
+f_d_group_genera <- read_tsv(here("data/process/5_days_PEG_genus_group_stools.tsv")) %>% 
+  filter(p.value.adj < 0.05)  #Select only significant p-values
+f_d_group_total_genera <- f_d_group_genera %>% 
+  filter(!duplicated(genus))  %>% #Remove genera that appear multiple times
+  tally()
+#Genera that were different between treatment groups over multiple days
+f_d_group_multi_day <- f_d_group_genera %>% 
+  filter(p.value.adj < 0.05) %>% #Select only significant p-values
+  group_by(genus) %>% 
+  tally() %>% 
+  filter(n > 1) %>% #select genera that vary over multiple timepoints = n > 1
+  arrange(desc(n)) %>%  #Rank by how many time each genera shows up 
+  nrow() #24 genera includes Peptostreptococcaceae and Unclassified
+
 ```
@@
 2B).
+
 Next, we examined  alpha diversity by looking at Shannon diversity index over time, although both clindamycin and PEG treatments decreased diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
@@
 We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as Bacteroides actually decreased in clindamycin treated mice (Fig.
-2D).
+2D, Data Set S1, Sheet X).
 Finally, we examined the bacteria that differ across treatment groups over multiple timepoints.
-We found...
-(Fig.
-2E).
-Thus, PEG has a significant impact on the fecal microbiota that was maintained over time.
+We found `r f_d_group_multi_day` were different over multiple timepoints out of the `r f_d_group_total_genera` that were different between treatment groups (Fig.
+2E, Data Set S1, Sheet X).
+Thus, PEG has a significant impact on the fecal microbiota that was maintained over time and distinct from clindamycin treatment.
 
@@
 difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge.
-We decided to examine what other bacteria were changing during the post-infection period when *C.
-difficile* changed from undectable to detectable.
-Interestingly, we found ___ bacteria were shifting during the time period when C.
-difficile was becoming detectable in the stools (Fig.
-S1)
+We decided to examine the bacteria that changed during the post-infection period when the group median *C.
+difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 dayspost-infection (Fig.
+S1A).
+Interestingly, we found *Erysipelotrichaceae*, *Enterobacteriaceae*, and *Akkermansia* were changing during the time period when *C.
+difficile* was becoming detectable in the stools (Fig.
+S1B), although none of the bacteria were significant after multiple hypothesis correction (Data Set S1, Sheet X).
+Although we did not identify a clear signal to explain the delayed appearance of *C.
+difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.
 
-**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given that previous work had demonstrated that PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
-We examined CDI severity by evaluating cecum and colon histopathology [].
+``` {r 5-day PEG tissue microbiota}
+f_d_permanova_t <- read_tsv(here("data/process/5_days_PEG_permanova_tissues.tsv"))
+f_d_sign_contrib_t <- f_d_permanova_t %>% arrange(desc(R2)) %>%
+  filter(!effects == "Total") %>% #Remove total since it is not one of the variables
+  filter(`Pr(>F)` < 0.05) #Only select significant contributors
 
-***C. difficile* challenge does not have a synergistic disruptive effect on the microbiota of PEG-treated mice**
-Because *C.
-difficile* itself can have an impact on the microbiota (ref), we also sequenced the tissue and stools of mock-challenged mice.
+#List of variables that significantly explained variation in stool samples
+f_d_top_effects_t <- f_d_sign_contrib_t %>% 
+  pull(effects)
+#Sum of R2 values for variables that significantly explained variation in stool samples
+f_d_top_R2_sum_t <- f_d_sign_contrib_t %>% 
+  select(R2) %>% sum
+#P-value for variables that significantly explained variation in stool samples
+f_d_top_p_t <- f_d_sign_contrib_t %>% 
+  select(`Pr(>F)`) %>% max # P value will be rounded in text
+
+```
+**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated that PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
+To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon.
+Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treatment (Fig.
+3A).
+Although alpha diversity continued to increase over time based on the communities from PEG-treated mice collected at 20 and 30 days post-infection (Fig.
+3A).
+Group, day, and the interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig.
+3B, Data Set S1, Sheet X).
+*Bacteroides*, *Butyricoccus*, *Clostridiales*, *Clostridium Cluster XIVb*, *Firmicutes*, and *Ruminococcaceae* were consistently different between treatment groups in both the tissue and stool communities (Fig.
+3C, Fig.
+2E, Data Set S1, Sheet X).
+Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C.
+difficile* at 4 days post-infection (Fig.
+3D), the timepoint typically examined in *C.
+difficile* 630 challenged mice (ref).
+We also looked at 6 days post-infection because that was when we started to see a large difference in C*.
+difficile* colonization levels between PEG- and clindamycin-treated mice (Fig.
+1C).
+Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is 12 (Fig.
+3E).
+Thus, although PEG treatment had a profound impact on the mucosal microbiota, the impact of *C.
+difficile* on the cecum and colon was similar between PEG and clindamycin treated mice.
+
+***C.
+difficile* challenge does not have a synergistic disruptive effect on the microbiota of PEG-treated mice** Because *C.
+difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice.
 Examining the stools of the mock-infected PEG- and clindamycin-treated mice revealed similar bacterial disruptions as the *C.
 difficile* challenged mice (Fig.
-S2A).
+S2A-C).
 Similarly, there was no difference between the tissues of mock and *C.
 difficile* challenged mice (Fig.
-S2B).
+S2D-F).
 Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C.

### 2021-04-20 submission/manuscript.tex
Add to results

@@
 recovery to baseline were responsible for the prolonged \emph{C.
-difficile} colonization in PEG-treated mice (Fig. 2B). Next, we examined
-alpha diversity by looking at Shannon diversity index over time,
-although both clindamycin and PEG treatments decreased diversity, the
-diversity index was lower in the groups of mice that received PEG
-treatment compared to those that received clindamycin through thirty
-days post-infection (Fig. 2C). We next examined the bacterial genera
-that shifted after PEG treatment by comparing the baseline samples of
-mice treated with only PEG to samples from the same mice one day
-post-treatment. We found 18 OTUs that were altered by PEG, the majority
-of these OTUs decreased after PEG, but \emph{Enterobacteriaceae} and
-\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
-unique to PEG treatment as Bacteroides actually decreased in clindamycin
-treated mice (Fig. 2D). Finally, we examined the bacteria that differ
-across treatment groups over multiple timepoints. We found\ldots{} (Fig.
-2E). Thus, PEG has a significant impact on the fecal microbiota that was
-maintained over time.
+difficile} colonization in PEG-treated mice (Fig. 2B).
+
+Next, we examined alpha diversity by looking at Shannon diversity index
+over time, although both clindamycin and PEG treatments decreased
+diversity, the diversity index was lower in the groups of mice that
+received PEG treatment compared to those that received clindamycin
+through thirty days post-infection (Fig. 2C). We next examined the
+bacterial genera that shifted after PEG treatment by comparing the
+baseline samples of mice treated with only PEG to samples from the same
+mice one day post-treatment. We found 18 OTUs that were altered by PEG,
+the majority of these OTUs decreased after PEG, but
+\emph{Enterobacteriaceae} and \emph{Bacteroides} increased and the
+increase in \emph{Bacteroides} was unique to PEG treatment as
+Bacteroides actually decreased in clindamycin treated mice (Fig. 2D,
+Data Set S1, Sheet X). Finally, we examined the bacteria that differ
+across treatment groups over multiple timepoints. We found 24 were
+different over multiple timepoints out of the 33 that were different
+between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG has
+a significant impact on the fecal microbiota that was maintained over
+time and distinct from clindamycin treatment.
 
@@
 stools of the PEG-treated mice that were allowed to recover for 10 days
-prior to challenge. We decided to examine what other bacteria were
-changing during the post-infection period when \emph{C. difficile}
-changed from undectable to detectable. Interestingly, we found \_\_\_
-bacteria were shifting during the time period when C. difficile was
-becoming detectable in the stools (Fig. S1)
+prior to challenge. We decided to examine the bacteria that changed
+during the post-infection period when the group median \emph{C.
+difficile} shifted from undetectable at 1 day post-infection to
+detectable in the stool samples with the median stabilizing around 8
+dayspost-infection (Fig. S1A). Interestingly, we found
+\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
+\emph{Akkermansia} were changing during the time period when \emph{C.
+difficile} was becoming detectable in the stools (Fig. S1B), although
+none of the bacteria were significant after multiple hypothesis
+correction (Data Set S1, Sheet X). Although we did not identify a clear
+signal to explain the delayed appearance of \emph{C. difficile} in the
+5-day PEG mice that were allowed to recover for 10 days prior to
+challenge, the delay is striking and could reflect changes in microbial
+activity or metabolites that were not examined in this study.
 
 \textbf{5-day laxative treatment does not promote more severe CDIs
-despite altering the mucosal microbiota.} Given that previous work had
-demonstrated that PEG treatment disrupts the mucus layer and alters the
-immune response in mice (16), we decided to examine the impact of PEG
-treatment on the mucosal microbiota and CDI severity. We examined CDI
-severity by evaluating cecum and colon histopathology {[}{]}.
+despite altering the mucosal microbiota.} Given the findings from a
+previous study that demonstrated that PEG treatment disrupts the mucus
+layer and alters the immune response in mice (16), we decided to examine
+the impact of PEG treatment on the mucosal microbiota and CDI severity.
+To evaluate the mucosal microbiota, we sequenced snips of tissue
+collected from the cecum, proximal colon, and distal colon. Similar to
+what was observed with the stool samples, alpha diversity was lower in
+the PEG-treated mice compared to clindamycin treatment (Fig. 3A).
+Although alpha diversity continued to increase over time based on the
+communities from PEG-treated mice collected at 20 and 30 days
+post-infection (Fig. 3A). Group, day, and the interactions with other
+variables (cage, experiment number, and sample type) explained the
+majority of the variation observed in mucosal communities (PERMANOVA
+combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
+3B, Data Set S1, Sheet X). \emph{Bacteroides}, \emph{Butyricoccus},
+\emph{Clostridiales}, \emph{Clostridium Cluster XIVb},
+\emph{Firmicutes}, and \emph{Ruminococcaceae} were consistently
+different between treatment groups in both the tissue and stool
+communities (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Next, we examined
+CDI severity by evaluating cecum and colon histopathology (20) and found
+there was no difference in cecum and colon scores between clindamycin
+and PEG-treated mice that were challenged with \emph{C. difficile} at 4
+days post-infection (Fig. 3D), the timepoint typically examined in
+\emph{C. difficile} 630 challenged mice (ref). We also looked at 6 days
+post-infection because that was when we started to see a large
+difference in C\emph{. difficile} colonization levels between PEG- and
+clindamycin-treated mice (Fig. 1C). Although, there was a slight
+difference in the colon between PEG and clindamycin-treated mice, there
+was no difference in the cecum and the overall score is still relatively
+low given that the max possible summary score is 12 (Fig. 3E). Thus,
+although PEG treatment had a profound impact on the mucosal microbiota,
+the impact of \emph{C. difficile} on the cecum and colon was similar
+between PEG and clindamycin treated mice.
 
@@
 disruptive effect on the microbiota of PEG-treated mice} Because
-\emph{C. difficile} itself can have an impact on the microbiota (ref),
-we also sequenced the tissue and stools of mock-challenged mice.
-Examining the stools of the mock-infected PEG- and clindamycin-treated
-mice revealed similar bacterial disruptions as the \emph{C. difficile}
-challenged mice (Fig. S2A). Similarly, there was no difference between
-the tissues of mock and \emph{C. difficile} challenged mice (Fig. S2B).
-Thus, most of the microbiota alterations we observed in the PEG-treated
-mice were a result of the laxative and not an interaction between the
-laxative and \emph{C. difficile}.
+\emph{C. difficile} itself can have an impact on the microbiota (21), we
+also sequenced the tissue and stools of mock-challenged clindamycin and
+5-day PEG treated mice. Examining the stools of the mock-infected PEG-
+and clindamycin-treated mice revealed similar bacterial disruptions as
+the \emph{C. difficile} challenged mice (Fig. S2A-C). Similarly, there
+was no difference between the tissues of mock and \emph{C. difficile}
+challenged mice (Fig. S2D-F). Thus, most of the microbiota alterations
+we observed in the PEG-treated mice were a result of the laxative and
+not an interaction between the laxative and \emph{C. difficile}.
 
@@
 
+\leavevmode\hypertarget{ref-Reeves2011}{}%
+20. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
+interplay between microbiome dynamics and pathogen dynamics in a murine
+model of \emph{Clostridium difficile} infection \textbf{2}:145--158.
+doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.
+
+\leavevmode\hypertarget{ref-Jenior2018}{}%
+21. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+\textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
+structure and metabolism of distinct cecal microbiomes during initial
+infection to promote sustained colonization. mSphere \textbf{3}.
+doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
+
 \end{document}

### 2021-04-21 submission/manuscript.Rmd
Add to results section

@@
 
-**Five-day post-infection community data can predict mice that will have prolonged *C. difficile* colonization** After identifying
+``` {r ml results}
+#AUC results
+auc_results <-  read_csv(here("results/performance_results.csv")) %>% 
+  mutate(taxa_level = "genus") %>%  
+  group_by(method) %>% 
+  summarize(median = median(AUC)) %>% 
+  arrange(desc(median))
+#Highest AUC value
+max_auc <- auc_results %>% select(median) %>% max()
+
+```
+**Five-day post-infection community data can predict mice that will have prolonged *C.
+difficile* colonization** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the taxa that were influencing prolonged *C.
+difficile* colonization.
+We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with input bacterial community data from 5 days post-infection to predict whether the mice were colonized with *C.
+difficile* 10 days post-infection.
+We chose 5 days post-infection because that was the earliest timepoint where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after *C.
+difficile* challenge and then administered an FMT or PBS gavage.
+The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were driving performance.
+We selected the top 20 bacteria contributing to our models performance and examined their relative abundance at 5 days post-infection, the timepoint used to predict C.
+difficile colonization status on day 10 (Fig.
+6A).
+Next, we focused on the 5 genera that were > 1 % relative abundance in either the cleared or colonized mice and examined the relative abundance dynamics of these bacteria over time.
+We found _, _, and _ tended to have a higher relative abundance and __, __ had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
+difficile* (Fig.
+6B).
+Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of S24-7, also known as *Muribaculum intestinale* [@Tropini2018].
+We decided to check our *Porphyromonadaceae* OTUs because *Muribaculum intestinale* is known to be classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020], *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
+difficile* colonization, and had a high relative abundance in the communities of mice that cleared *C.
+difficile* within 10 days.
+We identified 4 OTUs that had at least (insert minimum percent identity) to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C.
+difficile* at 10 days post-infection (Fig.
+S3).
 
@@
 
-* Discussion of prolonged persistence. C. difficile sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent papers.
+* Discussion of prolonged persistence. *C. difficile* sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent papers.
 

### 2021-04-21 submission/manuscript.tex
Add to results section

@@
 will have prolonged \emph{C. difficile} colonization} After identifying
+bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG
+treatments, we decided to examine the taxa that were influencing
+prolonged \emph{C. difficile} colonization. We trained 3 types of
+machine learning models (random forest, logistic regression, and support
+vector machine) with input bacterial community data from 5 days
+post-infection to predict whether the mice were colonized with \emph{C.
+difficile} 10 days post-infection. We chose 5 days post-infection
+because that was the earliest timepoint where we would see a treatment
+effect in the mice that were given 1-day PEG treatment three days after
+\emph{C. difficile} challenge and then administered an FMT or PBS
+gavage. The random forest model had the highest performance (AUROC =
+0.90, Data Set S1, Sheet X), so we next performed permutation importance
+to examine the bacteria that were driving performance. We selected the
+top 20 bacteria contributing to our models performance and examined
+their relative abundance at 5 days post-infection, the timepoint used to
+predict C. difficile colonization status on day 10 (Fig. 6A). Next, we
+focused on the 5 genera that were \textgreater{} 1 \% relative abundance
+in either the cleared or colonized mice and examined the relative
+abundance dynamics of these bacteria over time. We found \emph{, }, and
+\_ tended to have a higher relative abundance and \textbf{, } had a
+lower relative abundance in the mice with prolonged colonization
+compared to the mice that cleared \emph{C. difficile} (Fig. 6B).
+Previous work examining the impact of PEG on the murine microbiota found
+that PEG treatment resulted in the permanent loss of S24-7, also known
+as \emph{Muribaculum intestinale} (16). We decided to check our
+\emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
+known to be classified as \emph{Porphyromonadaceae} by the Ribosomal
+Database Project (RDP) database (22), \emph{Porphyromonadaceae} was a
+top feature in the random forest model predicting prolonged \emph{C.
+difficile} colonization, and had a high relative abundance in the
+communities of mice that cleared \emph{C. difficile} within 10 days. We
+identified 4 OTUs that had at least (insert minimum percent identity) to
+\emph{Muribaculum intestinale} and examined their abundance in mice that
+either cleared or were still colonized with \emph{C. difficile} at 10
+days post-infection (Fig. S3).
 
@@
 \item
-  Discussion of prolonged persistence. C. difficile sequences detected
-  in tissue samples. Association with mucin-degrading bacteria suggested
-  by recent papers.
+  Discussion of prolonged persistence. \emph{C. difficile} sequences
+  detected in tissue samples. Association with mucin-degrading bacteria
+  suggested by recent papers.
 \item
@@
 
+\leavevmode\hypertarget{ref-Vornhagen2020}{}%
+22. \textbf{Vornhagen J}, \textbf{Bassis CM}, \textbf{Ramakrishnan S},
+\textbf{Hein R}, \textbf{Mason S}, \textbf{Bergman Y}, \textbf{Sunshine
+N}, \textbf{Fan Y}, \textbf{Timp W}, \textbf{Schatz MC}, \textbf{Young
+VB}, \textbf{Simner PJ}, \textbf{Bachman MA}. 2020. A plasmid locus
+associated with klebsiella clinical infections encodes a
+microbiome-dependent gut fitness factor.
+doi:\href{https://doi.org/10.1101/2020.02.26.963322}{10.1101/2020.02.26.963322}.
+
 \end{document}

### 2021-04-23 code/figure_S3.R
Reorder supplemental figures

@@ -1,9 +1,19 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/ml_abund_otu_muribaculum_lineplot.png")
-b <- ggdraw() + draw_image("results/figures/ml_abund_otu_porphyromonadaceae_lineplot.png")
-c <- ggdraw() + draw_image("results/figures/ml_abund_otu_lachnospiraceae_lineplot.png")
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_stool_PCoA_mock.png")
+a_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_legend.png")
+a <- plot_grid(a, a_legend, ncol = 1, rel_heights = c(1, .2))
+d <- ggdraw() + draw_image("results/figures/5_days_PEG_tissue_PCoA_mock.png")
+d_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_alpha_legend_tissues.png")
+d <- plot_grid(d, d_legend, ncol = 1, rel_heights = c(1, .1))
+top_panel <- plot_grid(a, d, labels = c("A", "D"), label_size = 12, nrow=1)
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_mock.png")
+e <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues_mock.png")
+middle_panel <- plot_grid(b, e, labels = c("B", "E"), label_size = 12, ncol=2)
+c <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_stools.png")
+f <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_tissues.png")
+bottom_panel <- plot_grid(c, f, labels = c("C", "F"), label_size = 12, nrow=1)
 
-plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S3.pdf", width=6.875, height=6.5)+
-  ggsave("submission/figure_S3.pdf", width=6.875, height=6.5)
+plot_grid(top_panel, middle_panel, bottom_panel, labels = NULL, label_size = 12, ncol=1, rel_heights = c(1.4, 1, 1))+
+  ggsave("results/figures/figure_S3.pdf", width=6.875, height=7)+
+  ggsave("submission/figure_S3.pdf", width=6.875, height=7)

### 2021-04-23 code/figure_S4.R

@@ -1,0 +1,9 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+
+a <- ggdraw() + draw_image("results/figures/ml_abund_otu_muribaculum_lineplot.png")
+b <- ggdraw() + draw_image("results/figures/ml_abund_otu_porphyromonadaceae_lineplot.png")
+c <- ggdraw() + draw_image("results/figures/ml_abund_otu_lachnospiraceae_lineplot.png")
+
+plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_S4.pdf", width=6.875, height=6.5)+
+  ggsave("submission/figure_S4.pdf", width=6.875, height=6.5)

### 2021-04-23 code/5_days_PEG_16S.R
Add lineplots of day 6 bacteria that were different across treatment groups across all 3 types of tissues

@@ -571,6 +571,144 @@ lp_cecum <- line_plot_genus(genus_cecum, shared_sig_cecum_genus, lp_tissue_days,
   scale_x_continuous(limits = c(3.5,30.5), breaks = c(4, 6, 20, 30), labels = c(4, 6,20, 30))#Change scale since we have less timepoints for tissues
 save_plot(filename = "results/figures/5_days_PEG_genus_lineplot_cecum.png", lp_cecum, base_height = 5, base_width = 8)
 
+#Lineplots of significant genera across compartments
+#Function to create faceted line plots of relative abundances of genera of interest across tissue compartments
+#sample_df = subset dataframe of samples to be plotted
+#genera = list of genera to plot
+#timepoint = day of the experiment to plot
+line_plot_tissue_comp <- function(sample_df, genera, timepoint){
+  sample_df %>% 
+    filter(genus %in% genera) %>% #Select only genera of interest
+    mutate(genus = fct_relevel(genus, genera)) %>% #Reorder genera to match order of genera of interest
+    mutate(group = fct_relevel(group, rev(color_groups))) %>% #Specify the order of the groups
+    mutate(sample_type = fct_relevel(sample_type, c("cecum", "proximal_colon", "distal_colon"))) %>% #Make sure sample type is in the correct order
+    filter(day == timepoint) %>% #Select only timepoint of interest
+    group_by(group, genus, sample_type) %>% 
+    summarize(median=median(agg_rel_abund + 1/2000),`.groups` = "drop") %>%  #Add small value (1/2Xsubssampling parameter) so that there are no infinite values with log transformation
+    ggplot()+
+    geom_line(aes(x = sample_type, y=median, color=group, group = group), linetype = "solid")+
+    scale_colour_manual(name=NULL,
+                        values=color_scheme,
+                        breaks=color_groups,
+                        labels=color_labels)+
+    scale_x_discrete(breaks = c("cecum", "proximal_colon", "distal_colon"), labels = c("C", "PC", "DC"))+
+    scale_y_continuous(trans = "log10", limits = c(1/10900, 1), breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+
+    geom_hline(yintercept=1/1000, color="gray")+ #Represents limit of detection
+    labs(title=NULL,
+         x="Days Post-Infection",
+         y="Relative abundance (%)")+
[…]

### 2021-04-23 code/figure_S2.R

@@ -1,19 +1,10 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/5_days_PEG_stool_PCoA_mock.png")
-a_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_legend.png")
-a <- plot_grid(a, a_legend, ncol = 1, rel_heights = c(1, .2))
-d <- ggdraw() + draw_image("results/figures/5_days_PEG_tissue_PCoA_mock.png")
-d_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_mock_alpha_legend_tissues.png")
-d <- plot_grid(d, d_legend, ncol = 1, rel_heights = c(1, .1))
-top_panel <- plot_grid(a, d, labels = c("A", "D"), label_size = 12, nrow=1)
-b <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_mock.png")
-e <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_tissues_mock.png")
-middle_panel <- plot_grid(b, e, labels = c("B", "E"), label_size = 12, ncol=2)
-c <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_stools.png")
-f <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_mock_tissues.png")
-bottom_panel <- plot_grid(c, f, labels = c("C", "F"), label_size = 12, nrow=1)
-                       
-plot_grid(top_panel, middle_panel, bottom_panel, labels = NULL, label_size = 12, ncol=1, rel_heights = c(1.4, 1, 1))+
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_cecum_d6_16_genera.png")
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_peptostreptococcacea.png")
+b <- plot_grid(b, NULL, labels = NULL)
+
+plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol=1, rel_heights = c(1.4, 1, 1))+
   ggsave("results/figures/figure_S2.pdf", width=6.875, height=7)+
   ggsave("submission/figure_S2.pdf", width=6.875, height=7)
+

### 2021-04-23 code/5_days_PEG_16S.R
Implement barbell plot for PEG-impacted genera

@@ -1084,6 +1084,35 @@ peg_impacted_genera_bb_plot <- peg_impacted_genera_bb %>%
         legend.position = "none") 
 save_plot(filename = paste0("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell.png"), peg_impacted_genera_bb_plot, base_height = 9, base_width = 8)
 
+#Create alternative barbell plot without need of facets
+peg_impacted_genera_bb_plot <- peg_impacted_genera_bb %>% 
+  ggplot(aes(x= genus, color=group, ymin= B, ymax = P))+
+  geom_linerange(aes(color = group, xend = genus), size = 1,
+                 arrow= arrow(length=unit(0.30,"cm"), ends="last"),
+                 position = position_dodge(width = 1))+
+  geom_point(aes(y = B, color = group), shape = 1, stroke = 1, size = 3,
+             position = position_dodge(width = 1)) + 
+  geom_point(aes(y = P, color = group), shape = 16, size = 3,
+             position = position_dodge(width = 1)) + 
+  scale_colour_manual(name=NULL,
+                      values=color_scheme,
+                      breaks=color_groups,
+                      labels=color_labels)+
+  geom_hline(yintercept=1/1000, color="gray")+
+  labs(title=NULL, 
+       x=NULL,
+       y="Relative abundance (%)")+
+  scale_y_log10(breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100), limits = c(1/1/10000, 1))+
+  coord_flip()+
+  theme_classic()+
+  geom_vline(xintercept = c((1:18) - 0.5 ), color = "grey") + # Add gray lines to clearly separate OTUs
+  theme(plot.title=element_text(hjust=0.5),
+        text = element_text(size = 16),# Change font size for entire plot
+        axis.text.y = element_markdown(face = "italic"), #Make sure genera names are in italics
+        strip.background = element_blank(),
+        panel.spacing = unit(2, "lines"), #Increase spacing between facets
+        legend.position = "none") 
+save_plot(filename = paste0("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell_v2.png"), peg_impacted_genera_bb_plot, base_height = 7, base_width = 6)
 
 #Examine genera of interest over time----
 genus_stools_t <- genus_stools %>% 

### 2021-04-23 code/figure_2.R

@@ -3,7 +3,7 @@ source("code/utilities.R") #Loads libraries, reads in metadata, functions
 a <- ggdraw() + draw_image("results/figures/5_days_PEG_stool_PCoA.png", scale = 1)
 b <- ggdraw() + draw_image("results/figures/5_days_PEG_bc_combined_baseline.png", scale = 1)
 c <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools.png", scale = 1)
-d <- ggdraw() + draw_image("results/figures/5_days_PEG_genera_impacted_by_PEG.png")
+d <- ggdraw() + draw_image("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell_v2.png")
 e <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_stools.png", scale =0.95)
 
 top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow = 1, rel_widths = c(1, 1.5))

### 2021-04-23 code/figure_1.R
Increase figure dimensions

@@ -6,6 +6,6 @@ b <- ggdraw() + draw_image("results/figures/5_days_PEG_weight.png")
 c <- ggdraw() + draw_image("results/figures/5_days_PEG_cfu.png")#Note scale could be removed
 
 plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_1.pdf", width=5, height=6.5)+
-  ggsave("submission/figure_1.pdf", width=5, height=6.5)
+  ggsave("results/figures/figure_1.pdf", width=6.875, height=8.375)+
+  ggsave("submission/figure_1.pdf", width=6.875, height=8.375)
   

### 2021-04-23 code/figure_4.R

@@ -8,6 +8,6 @@ e <- ggdraw() + draw_image("results/figures/1_Day_PEG_genus_6_v2_baselinetoD1_li
 top_panel <- plot_grid(a, b, c, d, labels = c("A", "B", "C", "D"), label_size = 12, ncol = 2, rel_widths = c(.60, 1, 1,.60), rel_heights = c(1, 1))
 bottom_panel <- plot_grid(e, labels = c("E"), label_size = 12)
 plot_grid(top_panel, bottom_panel, nrow = 2, rel_heights = c(1.1, .8)) +
-  ggsave("results/figures/figure_4.pdf", width = 6.25, height = 5.5)+
-  ggsave("submission/figure_4.pdf", width = 6.25, height = 5.5)
+  ggsave("results/figures/figure_4.pdf", width = 6.875, height = 6)+
+  ggsave("submission/figure_4.pdf", width = 6.875, height = 6)
 

### 2021-04-23 code/figure_6.R

@@ -5,6 +5,6 @@ b <- ggdraw() + draw_image("results/figures/ml_top10_d5_genus.png")
 c <- ggdraw() + draw_image("results/figures/ml_abund_5_genus_lineplot.png")
 
 plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1, rel_heights = c(1, 2, 1))+
-  ggsave("results/figures/figure_6.pdf", width=6.2, height=6.5)+
-  ggsave("submission/figure_6.pdf", width=6.2, height=6.5)
+  ggsave("results/figures/figure_6.pdf", width=6.8, height=7.1)+
+  ggsave("submission/figure_6.pdf", width=6.8, height=7.1)
 

### 2021-04-23 code/figure_5.R
Renumber figures, adjust figure dimensions

@@ -28,16 +28,6 @@ d <- ggdraw() + draw_image("results/figures/post_CDI_PEG_richness_overtime_stool
 e <- ggdraw() + draw_image("results/figures/post_CDI_PEG_treatment_legend.png")
 fig <- image_graph(width = 400, height = 400, res = 96)
 plot_grid(a, e, b, c, d, labels = c("A", "", "B", "C", "D"), label_size = 12, ncol=1, rel_heights = c(.8,.2, 1,1,1))+
-  ggsave("results/figures/figure_5.pdf", width=4.6, height=7.5)+
-  ggsave("submission/figure_5.pdf", width=4.6, height=7.5)
+  ggsave("results/figures/figure_5.pdf", width=5.5, height=9)+
+  ggsave("submission/figure_5.pdf", width=5.5, height=9)
   
-#Vert version of Fig 4 16S
-f <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png", scale = 1.2)
-g <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend_vert.png")
-h <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_fmt.png")
-lineplot <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_stools.png")
-pcoa1 <- plot_grid(f,g, labels = c("A", ""), label_size = 12, nrow = 1, rel_heights = c(1,.5), rel_widths = c(1,.5))
-
-plot_grid(pcoa1, h, lineplot, labels = c("A", "B", "C"), label_size = 12, nrow = 3)+
-  ggsave("results/figures/figure_5_16S.pdf", width = 4, height = 6.7)+
-  ggsave("submission/figure_5_16S.pdf", width = 4, height = 6.7)

### 2021-04-23 code/figure_6.R

@@ -1,10 +1,12 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/ml_top_features_genus.png")
-b <- ggdraw() + draw_image("results/figures/ml_top10_d5_genus.png")
-c <- ggdraw() + draw_image("results/figures/ml_abund_5_genus_lineplot.png")
+#Vert version of Fig 4 16S
+a <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png", scale = 1.2)
+b <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend_vert.png")
+c <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_fmt.png")
+lineplot <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_stools.png")
+pcoa1 <- plot_grid(a ,b, labels = c("A", ""), label_size = 12, nrow = 1, rel_heights = c(1,.5), rel_widths = c(1,.5))
 
-plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1, rel_heights = c(1, 2, 1))+
-  ggsave("results/figures/figure_6.pdf", width=6.8, height=7.1)+
-  ggsave("submission/figure_6.pdf", width=6.8, height=7.1)
-
+plot_grid(pcoa1, c, lineplot, labels = c("", "B", "C"), label_size = 12, nrow = 3)+
+  ggsave("results/figures/figure_6.pdf", width = 6.3, height = 9)+
+  ggsave("submission/figure_6.pdf", width = 6.3, height = 9)

### 2021-04-23 code/figure_7.R

@@ -1,8 +1,11 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/summary_schematic.png")
+a <- ggdraw() + draw_image("results/figures/ml_top_features_genus.png")
+b <- ggdraw() + draw_image("results/figures/ml_top10_d5_genus.png")
+c <- ggdraw() + draw_image("results/figures/ml_abund_5_genus_lineplot.png")
 
-plot_grid(a, labels = NULL, label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_7.pdf", width=6.875, height=4.5)+
-  ggsave("submission/figure_7.pdf", width=6.875, height=4.5)
+plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1, rel_heights = c(1, 2, 1))+
+  ggsave("results/figures/figure_7.pdf", width=6.8, height=7.1)+
+  ggsave("submission/figure_7.pdf", width=6.8, height=7.1)
+
 

### 2021-04-23 code/figure_8.R

@@ -1,0 +1,7 @@
+source("code/utilities.R") #Loads libraries, reads in metadata, functions
+
+a <- ggdraw() + draw_image("results/figures/summary_schematic.png")
+
+plot_grid(a, labels = NULL, label_size = 12, ncol=1)+
+  ggsave("results/figures/figure_8.pdf", width=6.875, height=4.5)+
+  ggsave("submission/figure_8.pdf", width=6.875, height=4.5)

### 2021-04-23 submission/manuscript.Rmd
Generate latest draft with updated figures

@@
 **1-day laxative treatment results in transient *C. difficile* colonization and minor microbiota disruption**
+
+* Draft of results
   + Explain motivation for experiment, set-up of experiment (Fig. 4A). 
@@
 **Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered**
+
+* Draft of results
   + Reiterate motivation for experiment.
@@
 S3).
-
-* Figure 6. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.
-
-* Figure S2. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization.
-  + Ex. *Muribaculum intestinale*.
 
@@
 \newpage
-\includegraphics{figure_5_16S.pdf}
-**Figure 5.
+\includegraphics{figure_6.pdf}
+**Figure 6.
 1-day PEG treatment post C.
@@
 
-\includegraphics{figure_6.pdf}
-**Figure 6. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.** A. 
+\includegraphics{figure_7.pdf}
+**Figure 7. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.** A. 
 \newpage
 
-\includegraphics{figure_7.pdf}
-**Figure 7. Schematic summarizing findings.** A. 
+\includegraphics{figure_8.pdf}
+**Figure 8. Schematic summarizing findings.** A. 
 \newpage
@@
 \includegraphics{figure_S2.pdf}
-**Figure S2. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A. 
+**Figure S2. Impact of PEG treatment on the mucosal microbiota is greatest 6 days post-infection**  
+\newpage
+
+\includegraphics{figure_S3.pdf}
+**Figure S3. *C. difficile* challenge does not enhance the disruptive effect of PEG on the microbiota.** A. 
+\newpage
+
+\includegraphics{figure_S4.pdf}
+**Figure S4. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A. 
 \newpage

### 2021-04-23 submission/manuscript.tex
Generate latest draft with updated figures

@@
 \textbf{1-day laxative treatment results in transient \emph{C.
-difficile} colonization and minor microbiota disruption} + Explain
-motivation for experiment, set-up of experiment (Fig. 4A). + Transient
-\emph{C. difficile} colonization in 1-day PEG treated mice with all PEG
-mice clearing by 7dpi (Fig. 4B) + PCoA results (Day has larger R2 than
-treatment group). PEG mice are closer to baseline communites after 7-day
-period (Fig. 4C) + Shannon results show alpha diversity is only
-transiently disrupted (Fig. 4D) + Highlight bacteria that are disrupted
-by PEG or clindamycin treatment but recover within 7 days (Fig. 4E)
+difficile} colonization and minor microbiota disruption}
+
+\begin{itemize}
+\tightlist
+\item
+  Draft of results
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Explain motivation for experiment, set-up of experiment (Fig. 4A).
+  \item
+    Transient \emph{C. difficile} colonization in 1-day PEG treated mice
+    with all PEG mice clearing by 7dpi (Fig. 4B)
+  \item
+    PCoA results (Day has larger R2 than treatment group). PEG mice are
+    closer to baseline communites after 7-day period (Fig. 4C)
+  \item
+    Shannon results show alpha diversity is only transiently disrupted
+    (Fig. 4D)
+  \item
+    Highlight bacteria that are disrupted by PEG or clindamycin
+    treatment but recover within 7 days (Fig. 4E)
+  \end{itemize}
+\end{itemize}
 
@@
 clindamycin-treated mice regardless of whether an FMT is also
-administered} + Reiterate motivation for experiment. Saw only a
-transient microbiota disruption with 1-day PEG treatment alone. Also,
-see if adding FMT helps with clearance, mention groups have been using
-PEG or Abx/PEG treatments as a tool to engraft mice with human bacteria
-communities. Set up of experiment (Fig. 5A). + Prolonged \emph{C.
-difficile} colonization in post-CDI PEG treated mice (Fig. 5B) + FMT
-appears to partially restore alpha diversity (Shannon, but not richness
-Fig. 5C-D) + PCoA (necessary?) Could comment on the clustering of
-Clindamycin and PEG-treated mice that received FMT (Fig. 5E) + Only 2
-genera significantly impacted by FMT treatment, likely not as important
-for clearance (Fig. 5F) + Bacteria that consistently differ between
-groups over time, associated with prolonged colonization (Fig. G)
+administered}
+
+\begin{itemize}
+\tightlist
+\item
+  Draft of results
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Reiterate motivation for experiment. Saw only a transient microbiota
+    disruption with 1-day PEG treatment alone. Also, see if adding FMT
+    helps with clearance, mention groups have been using PEG or Abx/PEG
+    treatments as a tool to engraft mice with human bacteria
+    communities. Set up of experiment (Fig. 5A).
+  \item
+    Prolonged \emph{C. difficile} colonization in post-CDI PEG treated
+    mice (Fig. 5B)
+  \item
+    FMT appears to partially restore alpha diversity (Shannon, but not
+    richness Fig. 5C-D)
+  \item
+    PCoA (necessary?) Could comment on the clustering of Clindamycin and
+    PEG-treated mice that received FMT (Fig. 5E)
+  \item
+    Only 2 genera significantly impacted by FMT treatment, likely not as
+    important for clearance (Fig. 5F)
+  \item
+    Bacteria that consistently differ between groups over time,
+    associated with prolonged colonization (Fig. G)
+  \end{itemize}
+\end{itemize}
 
@@
 days post-infection (Fig. S3).
-
-\begin{itemize}
-\item
-  Figure 6. Specific microbiota features associated with prolonged
-  \emph{C. difficile} colonization in PEG treated mice.
-\item
-  Figure S2. Specific OTUs associated with clearance that are mostly
-  absent in mice with prolonged \emph{C. difficile} colonization.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Ex. \emph{Muribaculum intestinale}.
-  \end{itemize}
-\end{itemize}
 
@@
 
-\includegraphics{figure_5_16S.pdf} \textbf{Figure 5. 1-day PEG treatment
+\includegraphics{figure_6.pdf} \textbf{Figure 6. 1-day PEG treatment
 post C. difficile challenge prolongs colonization regardless of whether
@@
 
-\includegraphics{figure_6.pdf} \textbf{Figure 6. Specific microbiota
+\includegraphics{figure_7.pdf} \textbf{Figure 7. Specific microbiota
 features associated with prolonged \emph{C. difficile} colonization in
@@
 
-\includegraphics{figure_7.pdf} \textbf{Figure 7. Schematic summarizing
+\includegraphics{figure_8.pdf} \textbf{Figure 8. Schematic summarizing
 findings.} A. \newpage
@@
 
-\includegraphics{figure_S2.pdf} \textbf{Figure S2. Specific OTUs
+\includegraphics{figure_S2.pdf} \textbf{Figure S2. Impact of PEG
+treatment on the mucosal microbiota is greatest 6 days post-infection}\\
+\newpage
+
+\includegraphics{figure_S3.pdf} \textbf{Figure S3. \emph{C. difficile}
+challenge does not enhance the disruptive effect of PEG on the
+microbiota.} A. \newpage
+
+\includegraphics{figure_S4.pdf} \textbf{Figure S4. Specific OTUs
 associated with clearance that are mostly absent in mice with prolonged

### 2021-04-24 submission/manuscript.Rmd
Fig 5 caption edits

@@
 
+``` {r figure_5_permanova}
+fig_5_permanova_results = read_tsv("data/process/post_CDI_PEG_permanova_stools.tsv")
+r_sq_ordered_perm = permanova_results %>% arrange(desc(R2)) %>% 
+  select(`effects`, `R2`, `Pr(>F)`) %>%
+  filter(`Pr(>F)` < .05) #filter for sig
+fig_5_permanova_sig_P_val <- r_sq_ordered_perm[,3] %>% sum()
+fig_5_permanova_sig_R2 <- r_sq_ordered_perm[,2] %>% sum() #0.9280951 
+```
+
 \includegraphics{figure_5.pdf}
 A.
-Experimental timeline for groups receiving PEG treatment post clindamycin gavage.
+Setup of experimental timeline for groups receiving PEG treatment post clindamycin gavage.
+All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with C.
+difﬁcile 630 spores on day 0.
+15% PEG 3350 was administered in the drinking water for 1 day.
+Mice feces were collected daily through the end of the experiment (30 days post-infection for Clind.
++ 1-day PEG group, 15 days post-infection for all others).
+CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths.
 B.
-Median and individual CFU/g of C.
-difficile measured over time via culture.
-Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction.
-Limit of detection is 1e2 CFU/g.
+CFU/g of C.
+difficile stool measured over time via serial dilutions.
+C.
+Median and individual Shannon Diversity measured over time.
+D.
+Median and individual richness measured over time.
+B-D.
+Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction for testing multiple times points.
+Black line represents the limit of detection for the first serial dilution.
 Background shading indicates PEG treatment period for applicable groups.
 Opacity of points reflects time point.
-C.
-Median and individual Shannon Diversity measured over time.
-Asterisks, background shading, and opacity of points consistent with (B).
-D.
-Median and individual richness measured over time.
-Asterisks, background shading, and opacity of points consistent with (B).
+The data presented are from a total of *** separate experiments.
 

### 2021-04-24 submission/manuscript.Rmd
Edits and additions to results

--- 2021-04-23 submission/manuscript.Rmd
@@
 output:
+  word_document: default
   pdf_document:
@@
     keep_tex: yes
-  word_document: default
   html_document:
@@
 After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C.
-difficile* 630 spores.
+difficile* 630 spores (Fig.
+1A).
 All treatments rendered mice susceptible to *C.
-difficile* colonization (Fig.
-1C), however PEG-treated mice remained colonized at a high level through thirty days post-infection.
-In contrast, the clindamycin-treated mice that cleared *C.
+difficile* colonization.
+However, PEG-treated mice remained colonized with *C.
+difficile* at high levels through thirty days post-infection (Fig.
+1C).
+In contrast, the clindamycin-treated mice cleared *C.
 difficile* within ten days post-infection.
@@
 From 9 days post-infection onward, the median *C.
-difficile* stabilized for the 5-day PEG plus 10-day recovery group of mice and remained high through 30 days post-infection (Fig.
+difficile* stabilized for the 10-day recovery group of PEG-treated mice and remained high through 30 days post-infection (Fig.
 1C).
@@
 ```
-**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C.
-difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to the two drugs having differential effects on the microbiota.
+**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C.
+difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota.
 We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups.
@@
 2A, Data Set S1, Sheet X).
-Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
+Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy (ref), we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
 Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
@@
 
-Next, we examined  alpha diversity by looking at Shannon diversity index over time, although both clindamycin and PEG treatments decreased diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
+Next, we examined  alpha diversity by looking at Shannon diversity index over time.
+Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
 2C).
 We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment.
-We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as Bacteroides actually decreased in clindamycin treated mice (Fig.
-2D, Data Set S1, Sheet X).
+We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X).
+The majority of the bacterial relative abundances decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as *Bacteroides* actually decreased in clindamycin treated mice (Fig.
+2D).
 Finally, we examined the bacteria that differ across treatment groups over multiple timepoints.
-We found `r f_d_group_multi_day` were different over multiple timepoints out of the `r f_d_group_total_genera` that were different between treatment groups (Fig.
+We found `r f_d_group_multi_day` genera were different over multiple timepoints out of the `r f_d_group_total_genera` that were different between treatment groups (Fig.
 2E, Data Set S1, Sheet X).
@@
 We decided to examine the bacteria that changed during the post-infection period when the group median *C.
-difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 dayspost-infection (Fig.
+difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 days post-infection (Fig.
 S1A).
@@
 
-#List of variables that significantly explained variation in stool samples
+#List of variables that significantly explained variation in tissue samples
 f_d_top_effects_t <- f_d_sign_contrib_t %>% 
   pull(effects)
-#Sum of R2 values for variables that significantly explained variation in stool samples
+#Sum of R2 values for variables that significantly explained variation in tissue samples
 f_d_top_R2_sum_t <- f_d_sign_contrib_t %>% 
   select(R2) %>% sum
-#P-value for variables that significantly explained variation in stool samples
+#P-value for variables that significantly explained variation in tissue samples
 f_d_top_p_t <- f_d_sign_contrib_t %>% 
@@
 Although alpha diversity continued to increase over time based on the communities from PEG-treated mice collected at 20 and 30 days post-infection (Fig.
-3A).
-Group, day, and the interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig.
+3A, Data Set S1, Sheet X).
+Group, day of the experiment, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig.
 3B, Data Set S1, Sheet X).
-*Bacteroides*, *Butyricoccus*, *Clostridiales*, *Clostridium Cluster XIVb*, *Firmicutes*, and *Ruminococcaceae* were consistently different between treatment groups in both the tissue and stool communities (Fig.
+We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with 16 (insert code instead) genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
+S3A, Data Set S1, Sheet X).
+However, at 30 days post-infection only *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and just in the cecum tissues (Fig.
 3C, Fig.
 2E, Data Set S1, Sheet X).
+Interestingly, one of the genera that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera with sequences that match C.
+difficile).
+*Peptostreptococcaceae* was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig.
+S2B).
+Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C.
+difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.
+
 Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C.
@@
 Thus, although PEG treatment had a profound impact on the mucosal microbiota, the impact of *C.
-difficile* on the cecum and colon was similar between PEG and clindamycin treated mice.
+difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.
 
@@
 difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice.
-Examining the stools of the mock-infected PEG- and clindamycin-treated mice revealed similar bacterial disruptions as the *C.
+Examining the stools of the mock-challenged mice revealed similar bacterial disruptions as the *C.
 difficile* challenged mice (Fig.
@@
 difficile*.
+``` {r 1-day PEG stool microbiota}
 
-**1-day laxative treatment results in transient *C. difficile* colonization and minor microbiota disruption**
+#Genera that were different between treatment days
+#Genera that were changed by treatment
+pt_d1 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_PTtoD1.tsv")) %>% 
+  filter(p.value.adj < 0.05) 
+#Genera that were still different at 7 days post-infection  
+pt_d7 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_PTtoD7.tsv")) %>% 
+  filter(p.value.adj < 0.05)
+pt_d7_genera <- pt_d7 %>% pull(genus)
+
+recovered_genera <- pt_d1 %>% 
+  filter(!genus %in% pt_d7_genera) %>% #Remove genera that were still different at day 7
+  nrow() #Tally number of rows
+
+```
+**1-day laxative treatment results in transient *C.
+difficile* colonization and minor microbiota disruption** Next, we decided to examine how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C.
+difficile*.
+We administered mice either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or a clindamycin treatment before challenging them with *C.
+difficile 630* (Fig.
+3A).
+In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C.
+difficile* by 7 days post-infection (Fig.
+3B).
+The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity close to baseline by 7 days post-infection (Fig.
+3C-D).
+We found `r recovered_genera` bacteria were impacted by treatment but recovered close to baseline levels within 7 days post-infection including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig.
+3E, Data Set S1, Sheet X).
+These findings suggest the duration of treatment matters when considering the impact of the laxatives, with shorter treatments resulting in a transient loss of *C.
+difficile* colonization resistance.
+
+**Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered** Since 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
+difficile* spores from the intestine.
+We were also interested in exploring whether PEG might help with engraftment in the context of FMTs.
+To examine the impact of PEG treatment on *C.
+difficile* clearance and FMT treatment, we treated 4 groups of mice with clindamycin and then challenged all mice with *C.
+difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution via gavage (Fig.
+5A).
+In contrast to our hypothesis, all groups of mice that received PEG actually exhibited prolonged *C.
+difficile* colonization (Fig.
+5B).
 
 * Draft of results
-  + Explain motivation for experiment, set-up of experiment (Fig. 4A). 
-  + Transient *C. difficile* colonization in 1-day PEG treated mice with all PEG mice clearing by 7dpi (Fig. 4B)
-  + PCoA results (Day has larger R2 than treatment group). PEG mice are closer to baseline communites after 7-day period (Fig. 4C)
-  + Shannon results show alpha diversity is only transiently disrupted (Fig. 4D)
-  + Highlight bacteria that are disrupted by PEG or clindamycin treatment but recover within 7 days (Fig. 4E)
 
-**Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered**
-
-* Draft of results
-  + Reiterate motivation for experiment.
-Saw only a transient microbiota disruption with 1-day PEG treatment alone.
-Also, see if adding FMT helps with clearance, mention groups have been using PEG or Abx/PEG treatments as a tool to engraft mice with human bacteria communities.
-Set up of experiment (Fig.
-5A).
   + Prolonged *C. difficile* colonization in post-CDI PEG treated mice (Fig. 5B)
   + FMT appears to partially restore alpha diversity (Shannon, but not richness Fig. 5C-D)
-  + PCoA (necessary?) Could comment on the clustering of Clindamycin and PEG-treated mice that received FMT (Fig. 5E)
-  + Only 2 genera significantly impacted by FMT treatment, likely not as important for clearance (Fig. 5F)
-  + Bacteria that consistently differ between groups over time, associated with prolonged colonization (Fig. G)
+  + PCoA (necessary?) Could comment on the clustering of Clindamycin and PEG-treated mice that received FMT (Fig. 6A)
+  + Only 2 genera significantly impacted by FMT treatment, likely not as important for clearance (Fig. 6B)
+  + Bacteria that consistently differ between groups over time, associated with prolonged colonization (Fig. 6C)
 
@@
 The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were driving performance.
-We selected the top 20 bacteria contributing to our models performance and examined their relative abundance at 5 days post-infection, the timepoint used to predict C.
+We selected the top 10 bacteria contributing to our models performance (Fig.
+7A) and examined their relative abundance at 5 days post-infection, the timepoint used to predict C.
 difficile colonization status on day 10 (Fig.
-6A).
+7B).
 Next, we focused on the 5 genera that were > 1 % relative abundance in either the cleared or colonized mice and examined the relative abundance dynamics of these bacteria over time.
@@
 difficile* (Fig.
-6B).
+7C).
 Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of S24-7, also known as *Muribaculum intestinale* [@Tropini2018].
@@
 difficile* at 10 days post-infection (Fig.
-S3).
+S4).
 
@@
 
-* Summary of major findings (Fig. 7A)
+* Summary of major findings (Fig. 8A)
 
@@
 
-* Discuss what these findings might mean for human patients (Fig. 7B)
+* Discuss what these findings might mean for human patients (Fig. 8B)
   + What's known regarding laxatives and susceptibility to CDIs
@@
 \includegraphics{figure_5.pdf}
-A.
+**Figure 5.
+1-day PEG treatment post *C.
+difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** A.
 Experimental timeline for groups receiving PEG treatment post clindamycin gavage.
@@
 **Figure 6.
-1-day PEG treatment post C.
-difficile challenge prolongs colonization regardless of whether an FMT is also administered.** A.
+For 1-day PEG treatment post *C.
+difficile* challenge mice that also receive an FMT only some bacterial genera were restored.** A.
 PCoA of all groups + FMT plotted over time.

### 2021-04-24 submission/manuscript.Rmd

@@
 
+``` {r figure_5_permanova}
+fig_5_permanova_results = read_tsv("data/process/post_CDI_PEG_permanova_stools.tsv")
+r_sq_ordered_perm = permanova_results %>% arrange(desc(R2)) %>%
+  select(`effects`, `R2`, `Pr(>F)`) %>%
+  filter(`Pr(>F)` < .05) #filter for sig
+fig_5_permanova_sig_P_val <- r_sq_ordered_perm[,3] %>% sum()
+fig_5_permanova_sig_R2 <- r_sq_ordered_perm[,2] %>% sum() #0.9280951
+```
+
 \includegraphics{figure_5.pdf}
@@
 1-day PEG treatment post *C.
-difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** A.
-Experimental timeline for groups receiving PEG treatment post clindamycin gavage.
+difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with C.
+difﬁcile 630 spores on day 0.
+15% PEG 3350 was administered in the drinking water for 1 day.
+Mice feces were collected daily through the end of the experiment (30 days post-infection for Clind.
++ 1-day PEG group, 15 days post-infection for all others).
+CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths.
 B.
-Median and individual CFU/g of C.
-difficile measured over time via culture.
-Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction.
-Limit of detection is 1e2 CFU/g.
+CFU/g of C.
+difficile stool measured over time via serial dilutions.
+C.
+Median and individual Shannon Diversity measured over time.
+D.
+Median and individual richness measured over time.
+B-D.
+Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction for testing multiple times points.
+Black line represents the limit of detection for the first serial dilution.
 Background shading indicates PEG treatment period for applicable groups.
 Opacity of points reflects time point.
-C.
-Median and individual Shannon Diversity measured over time.
-Asterisks, background shading, and opacity of points consistent with (B).
-D.
-Median and individual richness measured over time.
-Asterisks, background shading, and opacity of points consistent with (B).
+The data presented are from a total of *** separate experiments.
+\newpage
 
-\newpage
 \includegraphics{figure_6.pdf}

### 2021-04-24 submission/manuscript.Rmd
Corrections after trying to render manuscript

@@
 output:
-  word_document: default
   pdf_document:
@@
     keep_tex: yes
+  word_document: default
   html_document:
@@
 ``` {r figure_5_permanova}
-fig_5_permanova_results = read_tsv("data/process/post_CDI_PEG_permanova_stools.tsv")
-r_sq_ordered_perm = permanova_results %>% arrange(desc(R2)) %>%
+fig_5_permanova_results = read_tsv(here("data/process/post_CDI_PEG_permanova_stools.tsv"))
+r_sq_ordered_perm = fig_5_permanova_results %>% arrange(desc(R2)) %>%
   select(`effects`, `R2`, `Pr(>F)`) %>%
@@
 1-day PEG treatment post *C.
-difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with C.
-difﬁcile 630 spores on day 0.
+difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C.
+difficile* 630 spores on day 0.
 15% PEG 3350 was administered in the drinking water for 1 day.
@@
 B.
-CFU/g of C.
-difficile stool measured over time via serial dilutions.
+CFU/g of *C.
+difficile* stool measured over time via serial dilutions.
 C.

### 2021-04-24 submission/manuscript.tex
Corrections after trying to render manuscript

@@
 clindamycin, or PEG and clindamycin treatment all mice were challenged
-with 10\textsuperscript{3} \emph{C. difficile} 630 spores. All
-treatments rendered mice susceptible to \emph{C. difficile} colonization
-(Fig. 1C), however PEG-treated mice remained colonized at a high level
-through thirty days post-infection. In contrast, the clindamycin-treated
-mice that cleared \emph{C. difficile} within ten days post-infection.
-Surprisingly, PEG-treated mice were still susceptible to \emph{C.
-difficile} infection after 10-days of recovery from treatment although
-\emph{C. difficile} was not detectable in most of the group in the
-initial five days post-infection (Fig. 1C). From 9 days post-infection
-onward, the median \emph{C. difficile} stabilized for the 5-day PEG plus
-10-day recovery group of mice and remained high through 30 days
-post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was
-sufficient to render mice susceptible to prolonged \emph{C. difficile}
-colonization and PEG-treated mice remained susceptible for up to ten
-days post-treatment.
+with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 1A). All
+treatments rendered mice susceptible to \emph{C. difficile}
+colonization. However, PEG-treated mice remained colonized with \emph{C.
+difficile} at high levels through thirty days post-infection (Fig. 1C).
+In contrast, the clindamycin-treated mice cleared \emph{C. difficile}
+within ten days post-infection. Surprisingly, PEG-treated mice were
+still susceptible to \emph{C. difficile} infection after 10-days of
+recovery from treatment although \emph{C. difficile} was not detectable
+in most of the group in the initial five days post-infection (Fig. 1C).
+From 9 days post-infection onward, the median \emph{C. difficile}
+stabilized for the 10-day recovery group of PEG-treated mice and
+remained high through 30 days post-infection (Fig. 1C). Thus, osmotic
+laxative treatment alone was sufficient to render mice susceptible to
+prolonged \emph{C. difficile} colonization and PEG-treated mice remained
+susceptible for up to ten days post-treatment.
 
 \textbf{5-day laxative treatment differentially disrupts the fecal
-microbiota compared to clindamycin treatment.} Since laxatives and
-clindamycin have previously been shown to disrupt the murine microbiota
-(ref), we hypothesized the different \emph{C. difficile} colonization
-dynamics between mice treated with laxatives versus clindamycin were due
-to the two drugs having differential effects on the microbiota. We
-profiled the stool microbiota over time by sequencing the V4 region of
-the 16S rRNA gene to compare changes across treatment groups. We found
-time and treatment group explained half of the observed variation
-between fecal communities with most of the remaining variation explained
-by interactions between treatment group and other experimental variables
-including time, cage effects, and sequencing preparation (PERMANOVA
-combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
-2A, Data Set S1, Sheet X). Cage effects refer to the well-documented
-phenomenon that mice housed in the same cages have similar microbial
-communities due to copraphagy, we tried to minimize the impact of cage
-effects on our experiment by breaking up cagemates when assigning mice
-to treatment groups and primarily housing only two mice per cage.
-Importantly, although we conducted a total of 5 separate experiments,
-the experiment number and it's interaction with treatment group was not
-one of the variables that significantly explained the observed variation
-in fecal communities (Data Set S1, Sheet X). Interestingly, none of the
-treatment groups recovered to their baseline community structure either
-10 or 30 days post-infection suggesting other community features besides
-recovery to baseline were responsible for the prolonged \emph{C.
-difficile} colonization in PEG-treated mice (Fig. 2B).
+microbiota compared to clindamycin treatment.} Since osmotic laxatives
+and clindamycin have previously been shown to disrupt the murine
+microbiota (ref), we hypothesized the different \emph{C. difficile}
+colonization dynamics between mice treated with the osmotic laxative or
+clindamycin were due to the two drugs having differential effects on the
+microbiota. We profiled the stool microbiota over time by sequencing the
+V4 region of the 16S rRNA gene to compare changes across treatment
+groups. We found time and treatment group explained half of the observed
+variation between fecal communities with most of the remaining variation
+explained by interactions between treatment group and other experimental
+variables including time, cage effects, and sequencing preparation
+(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
+0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
+well-documented phenomenon that mice housed in the same cages have
+similar microbial communities due to copraphagy (ref), we tried to
+minimize the impact of cage effects on our experiment by breaking up
+cagemates when assigning mice to treatment groups and primarily housing
+only two mice per cage. Importantly, although we conducted a total of 5
+separate experiments, the experiment number and it's interaction with
+treatment group was not one of the variables that significantly
+explained the observed variation in fecal communities (Data Set S1,
+Sheet X). Interestingly, none of the treatment groups recovered to their
+baseline community structure either 10 or 30 days post-infection
+suggesting other community features besides recovery to baseline were
+responsible for the prolonged \emph{C. difficile} colonization in
+PEG-treated mice (Fig. 2B).
 
 Next, we examined alpha diversity by looking at Shannon diversity index
-over time, although both clindamycin and PEG treatments decreased
-diversity, the diversity index was lower in the groups of mice that
+over time. Although both clindamycin and PEG treatments decreased
+diversity, Shannon diversity was lower in the groups of mice that
 received PEG treatment compared to those that received clindamycin
@@
 baseline samples of mice treated with only PEG to samples from the same
-mice one day post-treatment. We found 18 OTUs that were altered by PEG,
-the majority of these OTUs decreased after PEG, but
-\emph{Enterobacteriaceae} and \emph{Bacteroides} increased and the
-increase in \emph{Bacteroides} was unique to PEG treatment as
-Bacteroides actually decreased in clindamycin treated mice (Fig. 2D,
-Data Set S1, Sheet X). Finally, we examined the bacteria that differ
-across treatment groups over multiple timepoints. We found 24 were
-different over multiple timepoints out of the 33 that were different
-between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG has
-a significant impact on the fecal microbiota that was maintained over
-time and distinct from clindamycin treatment.
+mice one day post-treatment. We found 18 genera that were altered by PEG
+treatment (Data Set S1, Sheet X). The majority of the bacterial relative
+abundances decreased after PEG, but \emph{Enterobacteriaceae} and
+\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
+unique to PEG treatment as \emph{Bacteroides} actually decreased in
+clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria
+that differ across treatment groups over multiple timepoints. We found
+24 genera were different over multiple timepoints out of the 33 that
+were different between treatment groups (Fig. 2E, Data Set S1, Sheet X).
+Thus, PEG has a significant impact on the fecal microbiota that was
+maintained over time and distinct from clindamycin treatment.
 
@@
 detectable in the stool samples with the median stabilizing around 8
-dayspost-infection (Fig. S1A). Interestingly, we found
+days post-infection (Fig. S1A). Interestingly, we found
 \emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
@@
 communities from PEG-treated mice collected at 20 and 30 days
-post-infection (Fig. 3A). Group, day, and the interactions with other
-variables (cage, experiment number, and sample type) explained the
-majority of the variation observed in mucosal communities (PERMANOVA
-combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
-3B, Data Set S1, Sheet X). \emph{Bacteroides}, \emph{Butyricoccus},
-\emph{Clostridiales}, \emph{Clostridium Cluster XIVb},
-\emph{Firmicutes}, and \emph{Ruminococcaceae} were consistently
-different between treatment groups in both the tissue and stool
-communities (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Next, we examined
-CDI severity by evaluating cecum and colon histopathology (20) and found
-there was no difference in cecum and colon scores between clindamycin
-and PEG-treated mice that were challenged with \emph{C. difficile} at 4
-days post-infection (Fig. 3D), the timepoint typically examined in
-\emph{C. difficile} 630 challenged mice (ref). We also looked at 6 days
-post-infection because that was when we started to see a large
-difference in C\emph{. difficile} colonization levels between PEG- and
-clindamycin-treated mice (Fig. 1C). Although, there was a slight
-difference in the colon between PEG and clindamycin-treated mice, there
-was no difference in the cecum and the overall score is still relatively
-low given that the max possible summary score is 12 (Fig. 3E). Thus,
-although PEG treatment had a profound impact on the mucosal microbiota,
-the impact of \emph{C. difficile} on the cecum and colon was similar
-between PEG and clindamycin treated mice.
+post-infection (Fig. 3A, Data Set S1, Sheet X). Group, day of the
+experiment, and their interactions with other variables (cage,
+experiment number, and sample type) explained the majority of the
+variation observed in mucosal communities (PERMANOVA combined
+R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
+Set S1, Sheet X). We saw the greatest difference in the mucosal
+microbiota between treatment groups at 6 days post-infection with 16
+(insert code instead) genera that were significantly different in all
+three of the tissue types we collected (cecum, proximal colon, and
+distal colon; Fig. S3A, Data Set S1, Sheet X). However, at 30 days
+post-infection only \emph{Bacteroides}, \emph{Clostridiales},
+\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
+treatment groups and just in the cecum tissues (Fig. 3C, Fig. 2E, Data
+Set S1, Sheet X). Interestingly, one of the genera that was different
+between treatment groups at 6 days post-infection was
+\emph{Peptostreptococcaceae} (the genera with sequences that match C.
+difficile). \emph{Peptostreptococcaceae} was primarily only present in
+the 5-day PEG treatment group of mice and decreased in the proximal and
+distal colon tissues over time (Fig. S2B). Thus, PEG treatment had a
+significant impact on the mucosal microbiota and we detected \emph{C.
+difficile} sequences in the cecum, proximal colon, and distal colon
+tissue communities.
+
+Next, we examined CDI severity by evaluating cecum and colon
+histopathology (20) and found there was no difference in cecum and colon
+scores between clindamycin and PEG-treated mice that were challenged
+with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the
+timepoint typically examined in \emph{C. difficile} 630 challenged mice
+(ref). We also looked at 6 days post-infection because that was when we
+started to see a large difference in C\emph{. difficile} colonization
+levels between PEG- and clindamycin-treated mice (Fig. 1C). Although,
+there was a slight difference in the colon between PEG and
+clindamycin-treated mice, there was no difference in the cecum and the
+overall score is still relatively low given that the max possible
+summary score is 12 (Fig. 3E). Thus, although PEG treatment had a
+profound impact on the mucosal microbiota, the impact of \emph{C.
+difficile} 630 infection on the cecum and colon was similar between PEG
+and clindamycin treated mice.
 
@@
 also sequenced the tissue and stools of mock-challenged clindamycin and
-5-day PEG treated mice. Examining the stools of the mock-infected PEG-
-and clindamycin-treated mice revealed similar bacterial disruptions as
-the \emph{C. difficile} challenged mice (Fig. S2A-C). Similarly, there
-was no difference between the tissues of mock and \emph{C. difficile}
-challenged mice (Fig. S2D-F). Thus, most of the microbiota alterations
-we observed in the PEG-treated mice were a result of the laxative and
-not an interaction between the laxative and \emph{C. difficile}.
+5-day PEG treated mice. Examining the stools of the mock-challenged mice
+revealed similar bacterial disruptions as the \emph{C. difficile}
+challenged mice (Fig. S2A-C). Similarly, there was no difference between
+the tissues of mock and \emph{C. difficile} challenged mice (Fig.
+S2D-F). Thus, most of the microbiota alterations we observed in the
+PEG-treated mice were a result of the laxative and not an interaction
+between the laxative and \emph{C. difficile}.
 
 \textbf{1-day laxative treatment results in transient \emph{C.
-difficile} colonization and minor microbiota disruption}
+difficile} colonization and minor microbiota disruption} Next, we
+decided to examine how a shorter osmotic laxative perturbation would
+impact the microbiome and susceptibility to \emph{C. difficile}. We
+administered mice either a 1-day PEG treatment, a 1-day PEG treatment
+with a 1-day recovery period, or a clindamycin treatment before
+challenging them with \emph{C. difficile 630} (Fig. 3A). In contrast to
+the 5-day PEG treated mice, the 1-day PEG treated mice were only
+transiently colonized and cleared \emph{C. difficile} by 7 days
+post-infection (Fig. 3B). The stool communities of PEG-treated mice were
+also only transiently disrupted, with Shannon diversity close to
+baseline by 7 days post-infection (Fig. 3C-D). We found 14 bacteria were
+impacted by treatment but recovered close to baseline levels within 7
+days post-infection including \emph{Enterobacteriaceae},
+\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
+\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
+suggest the duration of treatment matters when considering the impact of
+the laxatives, with shorter treatments resulting in a transient loss of
+\emph{C. difficile} colonization resistance.
+
+\textbf{Post-CDI laxative treatment disrupts clearance in
+clindamycin-treated mice regardless of whether an FMT is also
+administered} Since 1-day PEG treatment resulted in a more mild
+microbiota perturbation, we decided to use the 1-day treatment to
+examine the hypothesis that PEG helps to flush \emph{C. difficile}
+spores from the intestine. We were also interested in exploring whether
+PEG might help with engraftment in the context of FMTs. To examine the
+impact of PEG treatment on \emph{C. difficile} clearance and FMT
+treatment, we treated 4 groups of mice with clindamycin and then
+challenged all mice with \emph{C. difficile} before administering the
+following treatments: no additional treatment, 1-day PEG immediately
+after challenge, and 1-day PEG treatment 3 days after challenge followed
+by either administration of an FMT or PBS solution via gavage (Fig. 5A).
+In contrast to our hypothesis, all groups of mice that received PEG
+actually exhibited prolonged \emph{C. difficile} colonization (Fig. 5B).
 
 \begin{itemize}
-\tightlist
 \item
@@
   \tightlist
-  \item
-    Explain motivation for experiment, set-up of experiment (Fig. 4A).
-  \item
-    Transient \emph{C. difficile} colonization in 1-day PEG treated mice
-    with all PEG mice clearing by 7dpi (Fig. 4B)
-  \item
-    PCoA results (Day has larger R2 than treatment group). PEG mice are
-    closer to baseline communites after 7-day period (Fig. 4C)
-  \item
-    Shannon results show alpha diversity is only transiently disrupted
-    (Fig. 4D)
-  \item
-    Highlight bacteria that are disrupted by PEG or clindamycin
-    treatment but recover within 7 days (Fig. 4E)
-  \end{itemize}
-\end{itemize}
-
-\textbf{Post-CDI laxative treatment disrupts clearance in
-clindamycin-treated mice regardless of whether an FMT is also
-administered}
-
-\begin{itemize}
-\tightlist
-\item
-  Draft of results
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Reiterate motivation for experiment. Saw only a transient microbiota
-    disruption with 1-day PEG treatment alone. Also, see if adding FMT
-    helps with clearance, mention groups have been using PEG or Abx/PEG
-    treatments as a tool to engraft mice with human bacteria
-    communities. Set up of experiment (Fig. 5A).
   \item
@@
     PCoA (necessary?) Could comment on the clustering of Clindamycin and
-    PEG-treated mice that received FMT (Fig. 5E)
+    PEG-treated mice that received FMT (Fig. 6A)
   \item
     Only 2 genera significantly impacted by FMT treatment, likely not as
-    important for clearance (Fig. 5F)
+    important for clearance (Fig. 6B)
   \item
     Bacteria that consistently differ between groups over time,
-    associated with prolonged colonization (Fig. G)
+    associated with prolonged colonization (Fig. 6C)
   \end{itemize}
@@
 to examine the bacteria that were driving performance. We selected the
-top 20 bacteria contributing to our models performance and examined
-their relative abundance at 5 days post-infection, the timepoint used to
-predict C. difficile colonization status on day 10 (Fig. 6A). Next, we
-focused on the 5 genera that were \textgreater{} 1 \% relative abundance
-in either the cleared or colonized mice and examined the relative
-abundance dynamics of these bacteria over time. We found \emph{, }, and
-\_ tended to have a higher relative abundance and \textbf{, } had a
-lower relative abundance in the mice with prolonged colonization
-compared to the mice that cleared \emph{C. difficile} (Fig. 6B).
-Previous work examining the impact of PEG on the murine microbiota found
-that PEG treatment resulted in the permanent loss of S24-7, also known
-as \emph{Muribaculum intestinale} (16). We decided to check our
+top 10 bacteria contributing to our models performance (Fig. 7A) and
+examined their relative abundance at 5 days post-infection, the
+timepoint used to predict C. difficile colonization status on day 10
+(Fig. 7B). Next, we focused on the 5 genera that were \textgreater{} 1
+\% relative abundance in either the cleared or colonized mice and
+examined the relative abundance dynamics of these bacteria over time. We
+found \emph{, }, and \_ tended to have a higher relative abundance and
+\textbf{, } had a lower relative abundance in the mice with prolonged
+colonization compared to the mice that cleared \emph{C. difficile} (Fig.
+7C). Previous work examining the impact of PEG on the murine microbiota
+found that PEG treatment resulted in the permanent loss of S24-7, also
+known as \emph{Muribaculum intestinale} (16). We decided to check our
 \emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
@@
 either cleared or were still colonized with \emph{C. difficile} at 10
-days post-infection (Fig. S3).
+days post-infection (Fig. S4).
 
@@
 \item
-  Summary of major findings (Fig. 7A)
+  Summary of major findings (Fig. 8A)
 \item
@@
 \item
-  Discuss what these findings might mean for human patients (Fig. 7B)
+  Discuss what these findings might mean for human patients (Fig. 8B)
 
@@
 
-\includegraphics{figure_5.pdf} A. Experimental timeline for groups
-receiving PEG treatment post clindamycin gavage. B. Median and
-individual CFU/g of C. difficile measured over time via culture.
+\includegraphics{figure_5.pdf} \textbf{Figure 5. 1-day PEG treatment
+post \emph{C. difficile} challenge prolongs colonization regardless of
+whether an FMT is also administered.} All mice were administered 10
+mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
+\emph{C. difficile} 630 spores on day 0. 15\% PEG 3350 was administered
+in the drinking water for 1 day. Mice feces were collected daily through
+the end of the experiment (30 days post-infection for Clind. + 1-day PEG
+group, 15 days post-infection for all others). CFU quantification data
+was not available for each mouse due to stool sampling difficulties
+(particularly the day the mice came off of the PEG treatment) or early
+deaths. B. CFU/g of \emph{C. difficile} stool measured over time via
+serial dilutions. C. Median and individual Shannon Diversity measured
+over time. D. Median and individual richness measured over time. B-D.
 Asterisks indicate time points with significant differences between
-groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction.
-Limit of detection is 1e2 CFU/g. Background shading indicates PEG
-treatment period for applicable groups. Opacity of points reflects time
-point. C. Median and individual Shannon Diversity measured over time.
-Asterisks, background shading, and opacity of points consistent with
-(B). D. Median and individual richness measured over time. Asterisks,
-background shading, and opacity of points consistent with (B).
+groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction
+for testing multiple times points. Black line represents the limit of
+detection for the first serial dilution. Background shading indicates
+PEG treatment period for applicable groups. Opacity of points reflects
+time point. The data presented are from a total of *** separate
+experiments. \newpage
 
-\newpage
-
-\includegraphics{figure_6.pdf} \textbf{Figure 6. 1-day PEG treatment
-post C. difficile challenge prolongs colonization regardless of whether
-an FMT is also administered.} A. PCoA of all groups + FMT plotted over
-time. Opacity of points reflects time point. B. Relative abundance of
-genera significantly different over multiple time points post FMT or PBS
-treatment, plotted over time. Limit of detection is .1\%. C. Relative
-abundances of top 6 significant genera ranked by number of days
-significant, plotted over time. Limit of detection is .1\%.
+\includegraphics{figure_6.pdf} \textbf{Figure 6. For 1-day PEG treatment
+post \emph{C. difficile} challenge mice that also receive an FMT only
+some bacterial genera were restored.} A. PCoA of all groups + FMT
+plotted over time. Opacity of points reflects time point. B. Relative
+abundance of genera significantly different over multiple time points
+post FMT or PBS treatment, plotted over time. Limit of detection is
+.1\%. C. Relative abundances of top 6 significant genera ranked by
+number of days significant, plotted over time. Limit of detection is
+.1\%.
 

### 2021-04-25 submission/manuscript.Rmd
Add references

@@
 difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C.
-difficile* negative with non-diarrheal consistency [@Schubert2015].
+difficile* negative with non-diarrheal consistency [@Schubert2014].
 These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.
@@
 difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C.
-difficile* (Ref).
+difficile* [@Olson2013; @Diebel2014].
 However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C.
@@
 For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure.
-Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C.
+Many of the FMT studies to date rationalize the use of laxatives [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C.
 difficile* spores and toxins from the intestine [@Liacouras1996].
@@
 In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C.
-difficile* susceptibility and clearance [ref].
+difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020].
 Although, two groups have now shown PEG treatment alone renders mice susceptible to *C.
-difficile*, these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C.
+difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C.
 difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs.
@@
 ```
-**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C.
+**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C.
 difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota.
@@
 2A, Data Set S1, Sheet X).
-Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy (ref), we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
+Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015], we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
 Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
@@
 3D), the timepoint typically examined in *C.
-difficile* 630 challenged mice (ref).
+difficile* 630 challenged mice [@Dieterle2020].
 We also looked at 6 days post-infection because that was when we started to see a large difference in C*.
@@
   + What's known regarding laxatives and susceptibility to CDIs
+  + Clinical trial of PEG, results never posted [@Dieterle2018]
   + Relevance to human FMTs? Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.
@@
 ## Materials and Methods
+* Histopathology [@Theriot2011]
 

### 2021-04-25 submission/manuscript.tex
Add references

@@
 difficile} infections by protecting intestinal epithelial cells from the
-toxins produced by \emph{C. difficile} (Ref). However, whether laxative
-results in more severe CDIs in mice and how long mice remain colonized
-with \emph{C. difficile} after challenge is unclear.
+toxins produced by \emph{C. difficile} (18, 19). However, whether
+laxative results in more severe CDIs in mice and how long mice remain
+colonized with \emph{C. difficile} after challenge is unclear.
 
@@
 bowel preparation by taking an osmotic laxative prior to the procedure.
-Many of the FMT studies to date rationalize the use of laxatives (Ref)
-based on a 1996 case study with 2 pediatric patients where the authors
-suggested in the discussion that the laxative may help flush \emph{C.
-difficile} spores and toxins from the intestine (18).
+Many of the FMT studies to date rationalize the use of laxatives
+(20--22) based on a 1996 case study with 2 pediatric patients where the
+authors suggested in the discussion that the laxative may help flush
+\emph{C. difficile} spores and toxins from the intestine (23).
 
@@
 antibiotics including clindamycin disrupt the microbiota and influence
-\emph{C. difficile} susceptibility and clearance {[}ref{]}. Although,
-two groups have now shown PEG treatment alone renders mice susceptible
-to \emph{C. difficile}, these studies have raised additional questions
-regarding the dynamics and severity of infection as well as the role of
-laxative treatment in \emph{C. difficile} clearance that should be
-addressed to better inform how we think about laxatives in the context
-of CDIs. Here, we used our C57BL/6 clindamycin model as a control group
-to characterize how long PEG-treated mice remain susceptible, whether
-PEG treatment results in more severe CDI and sustained \emph{C.
-difficile} colonization, and if PEG treatment post-CDI can promote
-\emph{C. difficile} clearance.
+\emph{C. difficile} susceptibility and clearance (24--26). Although, two
+groups have now shown PEG treatment alone renders mice susceptible to
+\emph{C. difficile} (15, 17), these studies have raised additional
+questions regarding the dynamics and severity of infection as well as
+the role of laxative treatment in \emph{C. difficile} clearance that
+should be addressed to better inform how we think about laxatives in the
+context of CDIs. Here, we used our C57BL/6 clindamycin model as a
+control group to characterize how long PEG-treated mice remain
+susceptible, whether PEG treatment results in more severe CDI and
+sustained \emph{C. difficile} colonization, and if PEG treatment
+post-CDI can promote \emph{C. difficile} clearance.
 
@@
 mice typically clearing \emph{C. difficile} within 10 days
-post-infection (9, 19). All PEG-treated mice were administered a 15\%
+post-infection (9, 26). All PEG-treated mice were administered a 15\%
 PEG solution in the drinking water for 5-days, one group was also
@@
 and clindamycin have previously been shown to disrupt the murine
-microbiota (ref), we hypothesized the different \emph{C. difficile}
+microbiota (14--17), we hypothesized the different \emph{C. difficile}
 colonization dynamics between mice treated with the osmotic laxative or
@@
 well-documented phenomenon that mice housed in the same cages have
-similar microbial communities due to copraphagy (ref), we tried to
+similar microbial communities due to copraphagy (27), we tried to
 minimize the impact of cage effects on our experiment by breaking up
@@
 Next, we examined CDI severity by evaluating cecum and colon
-histopathology (20) and found there was no difference in cecum and colon
+histopathology (28) and found there was no difference in cecum and colon
 scores between clindamycin and PEG-treated mice that were challenged
@@
 timepoint typically examined in \emph{C. difficile} 630 challenged mice
-(ref). We also looked at 6 days post-infection because that was when we
+(29). We also looked at 6 days post-infection because that was when we
 started to see a large difference in C\emph{. difficile} colonization
@@
 disruptive effect on the microbiota of PEG-treated mice} Because
-\emph{C. difficile} itself can have an impact on the microbiota (21), we
+\emph{C. difficile} itself can have an impact on the microbiota (30), we
 also sequenced the tissue and stools of mock-challenged clindamycin and
@@
 known to be classified as \emph{Porphyromonadaceae} by the Ribosomal
-Database Project (RDP) database (22), \emph{Porphyromonadaceae} was a
+Database Project (RDP) database (31), \emph{Porphyromonadaceae} was a
 top feature in the random forest model predicting prolonged \emph{C.
@@
   \item
+    Clinical trial of PEG, results never posted (32)
+  \item
     Relevance to human FMTs? Unclear what the best administration route
@@
 \subsection{Materials and Methods}\label{materials-and-methods}}
+
+\begin{itemize}
+\tightlist
+\item
+  Histopathology (33)
+\end{itemize}
 
@@
 
-\leavevmode\hypertarget{ref-Schubert2015}{}%
-12. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against \emph{Clostridium
-difficile}. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+\leavevmode\hypertarget{ref-Schubert2014}{}%
+12. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
+\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
+distinguish patients with clostridium difficile infection and non-c.
+Difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
+doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
 
@@
 
+\leavevmode\hypertarget{ref-Olson2013}{}%
+18. \textbf{Olson A}, \textbf{Diebel LN}, \textbf{Liberati DM}. 2013.
+Effect of host defenses on clostridium difficile toxininduced intestinal
+barrier injury. Journal of Trauma and Acute Care Surgery
+\textbf{74}:983--990.
+doi:\href{https://doi.org/10.1097/ta.0b013e3182858477}{10.1097/ta.0b013e3182858477}.
+
+\leavevmode\hypertarget{ref-Diebel2014}{}%
+19. \textbf{Diebel LN}, \textbf{Liberati DM}. 2014. Reinforcement of the
+intestinal mucus layer protects against clostridium difficile intestinal
+injury in~vitro. Journal of the American College of Surgeons
+\textbf{219}:460--468.
+doi:\href{https://doi.org/10.1016/j.jamcollsurg.2014.05.005}{10.1016/j.jamcollsurg.2014.05.005}.
+
+\leavevmode\hypertarget{ref-vanNood2013}{}%
+20. \textbf{Nood E van}, \textbf{Vrieze A}, \textbf{Nieuwdorp M},
+\textbf{Fuentes S}, \textbf{Zoetendal EG}, \textbf{Vos WM de},
+\textbf{Visser CE}, \textbf{Kuijper EJ}, \textbf{Bartelsman JFWM},
+\textbf{Tijssen JGP}, \textbf{Speelman P}, \textbf{Dijkgraaf MGW},
+\textbf{Keller JJ}. 2013. Duodenal infusion of donor feces for
+RecurrentClostridium difficile. New England Journal of Medicine
+\textbf{368}:407--415.
+doi:\href{https://doi.org/10.1056/nejmoa1205037}{10.1056/nejmoa1205037}.
+
+\leavevmode\hypertarget{ref-Razik2017}{}%
+21. \textbf{Razik R}, \textbf{Osman M}, \textbf{Lieberman A},
+\textbf{Allegretti JR}, \textbf{Kassam Z}. 2017. Faecal microbiota
+transplantation for clostridium difficile infection: A multicentre study
+of non-responders. Medical Journal of Australia \textbf{207}:159--160.
+doi:\href{https://doi.org/10.5694/mja16.01452}{10.5694/mja16.01452}.
+
+\leavevmode\hypertarget{ref-Postigo2012}{}%
+22. \textbf{Postigo R}, \textbf{Kim JH}. 2012. Colonoscopic versus
+nasogastric fecal transplantation for the treatment of clostridium
+difficile infection: A review and pooled analysis. Infection
+\textbf{40}:643--648.
+doi:\href{https://doi.org/10.1007/s15010-012-0307-9}{10.1007/s15010-012-0307-9}.
+
 \leavevmode\hypertarget{ref-Liacouras1996}{}%
-18. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
+23. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
 irrigation as an adjunct to the treatment of chronic, relapsing
@@
 
+\leavevmode\hypertarget{ref-Schubert2015}{}%
+24. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+Antibiotic-induced alterations of the murine gut microbiota and
+subsequent effects on colonization resistance against \emph{Clostridium
+difficile}. mBio \textbf{6}.
+doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+
+\leavevmode\hypertarget{ref-Jenior2017}{}%
+25. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
+alternative nutrient niches during infection across distinct murine gut
+microbiomes. mSystems \textbf{2}.
+doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
+
 \leavevmode\hypertarget{ref-Tomkovich2020}{}%
-19. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+26. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
 \textbf{Schloss PD}. 2020. The initial gut microbiota and response to
@@
 
+\leavevmode\hypertarget{ref-Nguyen2015}{}%
+27. \textbf{Nguyen TLA}, \textbf{Vieira-Silva S}, \textbf{Liston A},
+\textbf{Raes J}. 2015. How informative is the mouse for human gut
+microbiota research? Disease Models \& Mechanisms \textbf{8}:1--16.
+doi:\href{https://doi.org/10.1242/dmm.017400}{10.1242/dmm.017400}.
+
 \leavevmode\hypertarget{ref-Reeves2011}{}%
-20. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+28. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
 \textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
@@
 
+\leavevmode\hypertarget{ref-Dieterle2020}{}%
+29. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
+\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
+\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
+K}. 2020. Systemic inflammatory mediators are effective biomarkers for
+predicting adverse outcomes in clostridioides difficile infection. mBio
+\textbf{11}.
+doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
+
 \leavevmode\hypertarget{ref-Jenior2018}{}%
-21. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+30. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
@@
 \leavevmode\hypertarget{ref-Vornhagen2020}{}%
-22. \textbf{Vornhagen J}, \textbf{Bassis CM}, \textbf{Ramakrishnan S},
+31. \textbf{Vornhagen J}, \textbf{Bassis CM}, \textbf{Ramakrishnan S},
 \textbf{Hein R}, \textbf{Mason S}, \textbf{Bergman Y}, \textbf{Sunshine
@@
 
+\leavevmode\hypertarget{ref-Dieterle2018}{}%
+32. \textbf{Dieterle MG}, \textbf{Rao K}, \textbf{Young VB}. 2018. Novel
+therapies and preventative strategies for primary and
+recurrentClostridium difficileinfections. Annals of the New York Academy
+of Sciences \textbf{1435}:110--138.
+doi:\href{https://doi.org/10.1111/nyas.13958}{10.1111/nyas.13958}.
+
+\leavevmode\hypertarget{ref-Theriot2011}{}%
+33. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
+\textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
+Cefoperazone-treated mice as an experimental platform to assess
+differential virulence ofClostridium difficilestrains. Gut Microbes
+\textbf{2}:326--334.
+doi:\href{https://doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}.
+
 \end{document}

### 2021-04-25 code/post_CDI_PEG_16S.R
Combine all KW results for differences across groups into 1 file

@@ -542,6 +542,9 @@ for (d in c(-1, 0, 1, 2, 3, 5, 6, 7, 8, 9, 10, 15)){
   kw_genus_stools <- add_row(kw_genus_stools, stats)  #combine all the dataframes together
 }
 
+#Save combined Kruskal-Wallis tests at the genus level
+kw_genus_stools %>% write_tsv("data/process/post_CDI_PEG_genus_group_stools.tsv")
+
 #Shared significant genera across from Day 2, 3, 5, 6, 7, 8, 10, 15
 shared_sig_genus_D2toD15 <- intersect_all(sig_genus_day2, sig_genus_day3, sig_genus_day5, sig_genus_day6, sig_genus_day7, sig_genus_day8, sig_genus_day9, sig_genus_day10, sig_genus_day15)
 view(shared_sig_genus_D2toD15)

### 2021-04-26 submission/manuscript.Rmd
Latest draft of the paper after editing the intro and results

@@
 difficile* [@Olson2013; @Diebel2014].
-However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C.
+However, whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C.
 difficile* after challenge is unclear.
 
-Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization.
+Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization resistance.
 For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure.
@@
 1A).
-PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice regained most of the weight five days after treatment (Fig.
+PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice recovered most of the weight five days after treatment (Fig.
 1B).
@@
 However, PEG-treated mice remained colonized with *C.
-difficile* at high levels through thirty days post-infection (Fig.
+difficile* at a high level through thirty days post-infection (Fig.
 1C).
@@
 Surprisingly, PEG-treated mice were still susceptible to *C.
-difficile* infection after 10-days of recovery from treatment although *C.
+difficile* infection after 10-days of recovery from treatment, although *C.
 difficile* was not detectable in most of the group in the initial five days post-infection (Fig.
 1C).
-From 9 days post-infection onward, the median *C.
-difficile* stabilized for the 10-day recovery group of PEG-treated mice and remained high through 30 days post-infection (Fig.
+From 8 days post-infection onward, the median *C.
+difficile* CFU stabilized for the 10-day recovery group of PEG-treated mice and remained high through 30 days post-infection (Fig.
 1C).
 Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C.
-difficile* colonization and PEG-treated mice remained susceptible for up to ten days post-treatment.
+difficile* colonization and PEG-treated mice remained susceptible up to ten days post-treatment.
 
@@
   tally() %>%
-  filter(n > 1) %>% #select genera that vary over multiple timepoints = n > 1
+  filter(n > 1) %>% #select genera that vary over multiple time points = n > 1
   arrange(desc(n)) %>%  #Rank by how many time each genera shows up
@@
 We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups.
-We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig.
+We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig.
 2A, Data Set S1, Sheet X).
-Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015], we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
+Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015].
+We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
 Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
@@
 
-Next, we examined  alpha diversity by looking at Shannon diversity index over time.
+Next, we examined  alpha diversity by looking at Shannon diversity over time.
 Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
@@
 We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X).
-The majority of the bacterial relative abundances decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as *Bacteroides* actually decreased in clindamycin treated mice (Fig.
+The majority of the bacterial relative abundances decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment, as *Bacteroides* actually decreased in clindamycin treated mice (Fig.
 2D).
-Finally, we examined the bacteria that differ across treatment groups over multiple timepoints.
-We found `r f_d_group_multi_day` genera were different over multiple timepoints out of the `r f_d_group_total_genera` that were different between treatment groups (Fig.
+Finally, we examined the bacteria that differed across treatment groups over multiple time points.
+We found `r f_d_group_multi_day` genera were different over multiple time points out of the `r f_d_group_total_genera` genera that were different between treatment groups (Fig.
 2E, Data Set S1, Sheet X).
-Thus, PEG has a significant impact on the fecal microbiota that was maintained over time and distinct from clindamycin treatment.
+Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.
 
-Surprisingly, *C.
+Interestingly, *C.
 difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge.
@@
 S1B), although none of the bacteria were significant after multiple hypothesis correction (Data Set S1, Sheet X).
-Although we did not identify a clear signal to explain the delayed appearance of *C.
+While we did not identify a clear signal to explain the delayed appearance of *C.
 difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.
@@
 
+#To Do: Insert code for number of genera different in all 3 types of tissue samples
+
 ```
-**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated that PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
+**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
 To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon.
@@
 3A, Data Set S1, Sheet X).
-Group, day of the experiment, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig.
+Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig.
 3B, Data Set S1, Sheet X).
-We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with 16 (insert code instead) genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
+We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with 16 genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
 S3A, Data Set S1, Sheet X).
-However, at 30 days post-infection only *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and just in the cecum tissues (Fig.
+However, at 30 days post-infection only *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig.
 3C, Fig.
 2E, Data Set S1, Sheet X).
-Interestingly, one of the genera that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera with sequences that match C.
-difficile).
+Interestingly, one of the bacteria that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera with an OTU sequence that matches *C.
+difficile*).
 *Peptostreptococcaceae* was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig.
@@
 difficile* at 4 days post-infection (Fig.
-3D), the timepoint typically examined in *C.
+3D), the time point typically examined in *C.
 difficile* 630 challenged mice [@Dieterle2020].
-We also looked at 6 days post-infection because that was when we started to see a large difference in C*.
+We also looked at 6 days post-infection because that was when we started to see a large difference in *C.
 difficile* colonization levels between PEG- and clindamycin-treated mice (Fig.
 1C).
-Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is 12 (Fig.
+Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is twelve (Fig.
 3E).
-Thus, although PEG treatment had a profound impact on the mucosal microbiota, the impact of *C.
+Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C.
 difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.
@@
 difficile*.
-We administered mice either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or a clindamycin treatment before challenging them with *C.
-difficile 630* (Fig.
+We administered either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or clindamycin to mice before challenging them with *C.
+difficile* (Fig.
 3A).
@@
 3B).
-The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity close to baseline by 7 days post-infection (Fig.
+The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering close to baseline by 7 days post-infection (Fig.
 3C-D).
-We found `r recovered_genera` bacteria were impacted by treatment but recovered close to baseline levels within 7 days post-infection including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig.
+We found `r recovered_genera` bacteria were impacted by treatment, but recovered close to baseline levels by 7 days post-infection including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig.
 3E, Data Set S1, Sheet X).
@@
 
+``` {r Post-CDI PEG stool microbiota}
+
+#Genera that were different between treatment groups
+post_group_genera <- read_tsv(here("data/process/post_CDI_PEG_genus_group_stools.tsv")) %>%
+  filter(p.value.adj < 0.05)  #Select only significant p-values
+post_group_total_genera <- post_group_genera %>%
+  filter(!duplicated(genus))  %>% #Remove genera that appear multiple times
+  tally() #26 total genera
+
+#Genera that were different between treatment groups over multiple days
+post_group_multi_day <- post_group_genera %>%
+  filter(p.value.adj < 0.05) %>% #Select only significant p-values
+  group_by(genus) %>%
+  tally() %>%
+  filter(n > 1) %>% #select genera that vary over multiple time points = n > 1
+  arrange(desc(n)) %>%  #Rank by how many time each genera shows up
+  nrow() #24 genera includes Peptostreptococcaceae and Unclassified
+
+```
 **Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered** Since 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
@@
 difficile* clearance and FMT treatment, we treated 4 groups of mice with clindamycin and then challenged all mice with *C.
-difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution via gavage (Fig.
+difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig.
 5A).
-In contrast to our hypothesis, all groups of mice that received PEG actually exhibited prolonged *C.
+Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C.
 difficile* colonization (Fig.
 5B).
-
-* Draft of results
-
-  + Prolonged *C. difficile* colonization in post-CDI PEG treated mice (Fig. 5B)
-  + FMT appears to partially restore alpha diversity (Shannon, but not richness Fig. 5C-D)
-  + PCoA (necessary?) Could comment on the clustering of Clindamycin and PEG-treated mice that received FMT (Fig. 6A)
-  + Only 2 genera significantly impacted by FMT treatment, likely not as important for clearance (Fig. 6B)
-  + Bacteria that consistently differ between groups over time, associated with prolonged colonization (Fig. 6C)
+Next we examined how post-CDI PEG treatment impacted the microbiota.
+Alpha diversity was lower in the PEG-treated mice with the exception of the PEG-treated mice that were administered an FMT (Fig 5C-D).
+The FMT appeared to partially restore Shannon diversity but not richness (Fig.
+5C-D).
+Similarly, we saw some overlap between the communities of mice that received FMT and the mice treated with only clindamycin after 5 days post-infection (Fig.
+6A).
+The increase in Shannon diversity suggests that the FMT did have an impact on the microbiota, despite seeing prolonged *C.
+difficile* colonization in the FMT treated mice.
+However, only *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice received either an FMT or PBS gavage (Fig.
+6B), suggesting the FMT only restored a couple of genera.
+Overall, we found there were `r post_group_multi_day` bacteria that differed over multiple time points between treatment groups including increased *Akkermansia* and decreased *Ruminococcaceae*, *Clostridiales*, *Lachnospiraceae*, and *Oscillibacter* in mice that received PEG after *C.
+difficile* challenge (Fig.
+6C).
+In sum, administering PEG actually prolonged *C.
+difficile* colonization, including in mice that received an FMT, which only restored a couple of bacterial genera.
 
@@
 
+#To Do: Insert minimum Percent identity as variable name once blast results to Muribaculum have been combined into 1 file
+
 ```
-**Five-day post-infection community data can predict mice that will have prolonged *C.
-difficile* colonization** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the taxa that were influencing prolonged *C.
+**Five-day post-infection community data can predict which mice that will have prolonged *C.
+difficile* colonization** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C.
 difficile* colonization.
-We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with input bacterial community data from 5 days post-infection to predict whether the mice were colonized with *C.
+We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with input bacterial community data from 5 days post-infection to predict whether the mice were still colonized with *C.
 difficile* 10 days post-infection.
-We chose 5 days post-infection because that was the earliest timepoint where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after *C.
+We chose 5 days post-infection because that was the earliest time point where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after *C.
 difficile* challenge and then administered an FMT or PBS gavage.
-The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were driving performance.
+The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C.
+difficile *colonization.
 We selected the top 10 bacteria contributing to our models performance (Fig.
-7A) and examined their relative abundance at 5 days post-infection, the timepoint used to predict C.
-difficile colonization status on day 10 (Fig.
+7A) and examined their relative abundance at 5 days post-infection, the time point used to predict *C.
+difficile* colonization status on day 10 (Fig.
 7B).
-Next, we focused on the 5 genera that were > 1 % relative abundance in either the cleared or colonized mice and examined the relative abundance dynamics of these bacteria over time.
-We found _, _, and _ tended to have a higher relative abundance and __, __ had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
+Next, we focused on the 5 genera that had a greater than 1 % relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time.
+We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
 difficile* (Fig.
 7C).
+Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-CDI PEG treated mice.
+
 Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of S24-7, also known as *Muribaculum intestinale* [@Tropini2018].
-We decided to check our *Porphyromonadaceae* OTUs because *Muribaculum intestinale* is known to be classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020], *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
-difficile* colonization, and had a high relative abundance in the communities of mice that cleared *C.
-difficile* within 10 days.
-We identified 4 OTUs that had at least (insert minimum percent identity) to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C.
-difficile* at 10 days post-infection (Fig.
-S4).
+We decided to check our *Porphyromonadaceae* OTUs because *Muribaculum intestinale* is classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020] and *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
+difficile* colonization.
+We identified 4 OTUs that had at least 92% identity to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C.
+difficile* at 10 days post-infection.
+While all of the OTUs, were decreased by PEG and clindamycin treatment, there was some recovery in the mice that cleared (Fig.
+S4A).
+We also examined other *Porphyromonadaceae* and *Lachnospiraceae* OTUs since these were the 2 genera that were important to our classification model and contained multiple OTUs that were different at 5 days post-infection between mice that either cleared or remained colonized with *C.
+difficile* by 10 days post-infection (Data Set S1, sheet X).
+While individual *Porphyromonadaceae* and *Lachnospiraceae* OTUs tended to be more abundant in the mice that clear *C.
+difficile* relative to the mice that exhibit prolonged colonization (Fig.
+S4B-C), there is no single OTU that fits the pattern we observed at the genus level (Fig.
+7C), suggesting multiple *Porphyromonadaceae* and *Lachnospiraceae* OTUs influenced *C.
+difficile* clearance.
+Overall, our results suggest that specific bacterial community differences explain the prolonged *C.
+difficile* colonization we observed in 5-day PEG and post-CDI 1-day PEG treated mice.
 
@@
   filter(sample_type == "stool") %>% #Select only stool samples
-  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 8, 9, 10, 15, 20, 25, 30)) %>%  #select timepoints on plot
+  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 8, 9, 10, 15, 20, 25, 30)) %>%  #select time points on plot
   filter(!is.na(avg_cfu)) %>% #Remove rows with NA values
@@
 *C.
-difficile* CFU/gram stool measured over time (N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per timepoint) via serial dilutions.
+difficile* CFU/gram stool measured over time (N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per time point) via serial dilutions.
 The black line represents the limit of detection for the first serial dilution.
@@
 For B-C, lines represent the median for each treatment group and circles represent samples from individual mice.
-Asterisks indicate timepoints where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple timepoints.
+Asterisks indicate time points where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
 The data presented are from a total of `r five_day_total_exp` separate experiments.
@@
   count()
-#To Do: this code is not specific for CFU values or the timepoints shown on panel B
+#To Do: this code is not specific for CFU values or the time points shown on panel B
 one_day_sample_min <- min(one_day_samples_per_day$n)
@@
 B.
-CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per timepoint) via several dilutions.
+CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per time point) via several dilutions.
 The black dotted line represents the limit of detection for the first serial dilution.
-Asterisks indicate timepoints where the CFU/gram was significantly different between groups using the Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple timepoints.
+Asterisks indicate time points where the CFU/gram was significantly different between groups using the Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple time points.
 C.

### 2021-04-26 submission/manuscript.tex
Latest draft of the paper after editing the intro and results

@@
 toxins produced by \emph{C. difficile} (18, 19). However, whether
-laxative results in more severe CDIs in mice and how long mice remain
-colonized with \emph{C. difficile} after challenge is unclear.
+laxative administration results in more severe CDIs in mice and how long
+mice remain colonized with \emph{C. difficile} after challenge is
+unclear.
 
@@
 recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
-microbiota is administered to the patient to restore colonization. For
-FMTs that are delivered via colonoscopy, patients typically undergo
-bowel preparation by taking an osmotic laxative prior to the procedure.
-Many of the FMT studies to date rationalize the use of laxatives
-(20--22) based on a 1996 case study with 2 pediatric patients where the
-authors suggested in the discussion that the laxative may help flush
-\emph{C. difficile} spores and toxins from the intestine (23).
+microbiota is administered to the patient to restore colonization
+resistance. For FMTs that are delivered via colonoscopy, patients
+typically undergo bowel preparation by taking an osmotic laxative prior
+to the procedure. Many of the FMT studies to date rationalize the use of
+laxatives (20--22) based on a 1996 case study with 2 pediatric patients
+where the authors suggested in the discussion that the laxative may help
+flush \emph{C. difficile} spores and toxins from the intestine (23).
 
@@
 in all 3 groups of PEG-treated mice, with the greatest change in weight
-observed on the fifth day of PEG treatment and the mice regained most of
-the weight five days after treatment (Fig. 1B). After either PEG,
+observed on the fifth day of PEG treatment and the mice recovered most
+of the weight five days after treatment (Fig. 1B). After either PEG,
 clindamycin, or PEG and clindamycin treatment all mice were challenged
@@
 colonization. However, PEG-treated mice remained colonized with \emph{C.
-difficile} at high levels through thirty days post-infection (Fig. 1C).
+difficile} at a high level through thirty days post-infection (Fig. 1C).
 In contrast, the clindamycin-treated mice cleared \emph{C. difficile}
@@
 still susceptible to \emph{C. difficile} infection after 10-days of
-recovery from treatment although \emph{C. difficile} was not detectable
+recovery from treatment, although \emph{C. difficile} was not detectable
 in most of the group in the initial five days post-infection (Fig. 1C).
-From 9 days post-infection onward, the median \emph{C. difficile}
+From 8 days post-infection onward, the median \emph{C. difficile} CFU
 stabilized for the 10-day recovery group of PEG-treated mice and
@@
 prolonged \emph{C. difficile} colonization and PEG-treated mice remained
-susceptible for up to ten days post-treatment.
+susceptible up to ten days post-treatment.
 
@@
 explained by interactions between treatment group and other experimental
-variables including time, cage effects, and sequencing preparation
+variables including time, cage effects, and sequencing preparation plate
 (PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
@@
 well-documented phenomenon that mice housed in the same cages have
-similar microbial communities due to copraphagy (27), we tried to
+similar microbial communities due to copraphagy (27). We tried to
 minimize the impact of cage effects on our experiment by breaking up
@@
 
-Next, we examined alpha diversity by looking at Shannon diversity index
-over time. Although both clindamycin and PEG treatments decreased
-diversity, Shannon diversity was lower in the groups of mice that
-received PEG treatment compared to those that received clindamycin
-through thirty days post-infection (Fig. 2C). We next examined the
-bacterial genera that shifted after PEG treatment by comparing the
-baseline samples of mice treated with only PEG to samples from the same
-mice one day post-treatment. We found 18 genera that were altered by PEG
-treatment (Data Set S1, Sheet X). The majority of the bacterial relative
+Next, we examined alpha diversity by looking at Shannon diversity over
+time. Although both clindamycin and PEG treatments decreased diversity,
+Shannon diversity was lower in the groups of mice that received PEG
+treatment compared to those that received clindamycin through thirty
+days post-infection (Fig. 2C). We next examined the bacterial genera
+that shifted after PEG treatment by comparing the baseline samples of
+mice treated with only PEG to samples from the same mice one day
+post-treatment. We found 18 genera that were altered by PEG treatment
+(Data Set S1, Sheet X). The majority of the bacterial relative
 abundances decreased after PEG, but \emph{Enterobacteriaceae} and
 \emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
-unique to PEG treatment as \emph{Bacteroides} actually decreased in
+unique to PEG treatment, as \emph{Bacteroides} actually decreased in
 clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria
-that differ across treatment groups over multiple timepoints. We found
-24 genera were different over multiple timepoints out of the 33 that
-were different between treatment groups (Fig. 2E, Data Set S1, Sheet X).
-Thus, PEG has a significant impact on the fecal microbiota that was
-maintained over time and distinct from clindamycin treatment.
+that differed across treatment groups over multiple time points. We
+found 24 genera were different over multiple time points out of the 33
+genera that were different between treatment groups (Fig. 2E, Data Set
+S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota
+that was maintained over time and was distinct from clindamycin
+treatment.
 
-Surprisingly, \emph{C. difficile} was not immediately detectable in the
+Interestingly, \emph{C. difficile} was not immediately detectable in the
 stools of the PEG-treated mice that were allowed to recover for 10 days
@@
 none of the bacteria were significant after multiple hypothesis
-correction (Data Set S1, Sheet X). Although we did not identify a clear
+correction (Data Set S1, Sheet X). While we did not identify a clear
 signal to explain the delayed appearance of \emph{C. difficile} in the
@@
 despite altering the mucosal microbiota.} Given the findings from a
-previous study that demonstrated that PEG treatment disrupts the mucus
-layer and alters the immune response in mice (16), we decided to examine
-the impact of PEG treatment on the mucosal microbiota and CDI severity.
-To evaluate the mucosal microbiota, we sequenced snips of tissue
-collected from the cecum, proximal colon, and distal colon. Similar to
-what was observed with the stool samples, alpha diversity was lower in
-the PEG-treated mice compared to clindamycin treatment (Fig. 3A).
-Although alpha diversity continued to increase over time based on the
-communities from PEG-treated mice collected at 20 and 30 days
-post-infection (Fig. 3A, Data Set S1, Sheet X). Group, day of the
-experiment, and their interactions with other variables (cage,
-experiment number, and sample type) explained the majority of the
-variation observed in mucosal communities (PERMANOVA combined
-R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
-Set S1, Sheet X). We saw the greatest difference in the mucosal
-microbiota between treatment groups at 6 days post-infection with 16
-(insert code instead) genera that were significantly different in all
-three of the tissue types we collected (cecum, proximal colon, and
-distal colon; Fig. S3A, Data Set S1, Sheet X). However, at 30 days
-post-infection only \emph{Bacteroides}, \emph{Clostridiales},
-\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
-treatment groups and just in the cecum tissues (Fig. 3C, Fig. 2E, Data
-Set S1, Sheet X). Interestingly, one of the genera that was different
-between treatment groups at 6 days post-infection was
-\emph{Peptostreptococcaceae} (the genera with sequences that match C.
-difficile). \emph{Peptostreptococcaceae} was primarily only present in
-the 5-day PEG treatment group of mice and decreased in the proximal and
-distal colon tissues over time (Fig. S2B). Thus, PEG treatment had a
-significant impact on the mucosal microbiota and we detected \emph{C.
-difficile} sequences in the cecum, proximal colon, and distal colon
-tissue communities.
+previous study that demonstrated PEG treatment disrupts the mucus layer
+and alters the immune response in mice (16), we decided to examine the
+impact of PEG treatment on the mucosal microbiota and CDI severity. To
+evaluate the mucosal microbiota, we sequenced snips of tissue collected
+from the cecum, proximal colon, and distal colon. Similar to what was
+observed with the stool samples, alpha diversity was lower in the
+PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although
+alpha diversity continued to increase over time based on the communities
+from PEG-treated mice collected at 20 and 30 days post-infection (Fig.
+3A, Data Set S1, Sheet X). Group, time point, and their interactions
+with other variables (cage, experiment number, and sample type)
+explained the majority of the variation observed in mucosal communities
+(PERMANOVA combined R\textsuperscript{2} = 0.83, \emph{P} \textless{}
+0.05, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in
+the mucosal microbiota between treatment groups at 6 days post-infection
+with 16 genera that were significantly different in all three of the
+tissue types we collected (cecum, proximal colon, and distal colon; Fig.
+S3A, Data Set S1, Sheet X). However, at 30 days post-infection only
+\emph{Bacteroides}, \emph{Clostridiales}, \emph{Firmicutes}, and
+\emph{Ruminococcaceae} were different between treatment groups and only
+in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X).
+Interestingly, one of the bacteria that was different between treatment
+groups at 6 days post-infection was \emph{Peptostreptococcaceae} (the
+genera with an OTU sequence that matches \emph{C. difficile}).
+\emph{Peptostreptococcaceae} was primarily only present in the 5-day PEG
+treatment group of mice and decreased in the proximal and distal colon
+tissues over time (Fig. S2B). Thus, PEG treatment had a significant
+impact on the mucosal microbiota and we detected \emph{C. difficile}
+sequences in the cecum, proximal colon, and distal colon tissue
+communities.
 
@@
 scores between clindamycin and PEG-treated mice that were challenged
-with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the
-timepoint typically examined in \emph{C. difficile} 630 challenged mice
+with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the time
+point typically examined in \emph{C. difficile} 630 challenged mice
 (29). We also looked at 6 days post-infection because that was when we
-started to see a large difference in C\emph{. difficile} colonization
+started to see a large difference in \emph{C. difficile} colonization
 levels between PEG- and clindamycin-treated mice (Fig. 1C). Although,
@@
 overall score is still relatively low given that the max possible
-summary score is 12 (Fig. 3E). Thus, although PEG treatment had a
-profound impact on the mucosal microbiota, the impact of \emph{C.
+summary score is twelve (Fig. 3E). Therefore, although PEG treatment had
+a disruptive effect on the mucosal microbiota, the impact of \emph{C.
 difficile} 630 infection on the cecum and colon was similar between PEG
@@
 impact the microbiome and susceptibility to \emph{C. difficile}. We
-administered mice either a 1-day PEG treatment, a 1-day PEG treatment
-with a 1-day recovery period, or a clindamycin treatment before
-challenging them with \emph{C. difficile 630} (Fig. 3A). In contrast to
-the 5-day PEG treated mice, the 1-day PEG treated mice were only
-transiently colonized and cleared \emph{C. difficile} by 7 days
-post-infection (Fig. 3B). The stool communities of PEG-treated mice were
-also only transiently disrupted, with Shannon diversity close to
-baseline by 7 days post-infection (Fig. 3C-D). We found 14 bacteria were
-impacted by treatment but recovered close to baseline levels within 7
-days post-infection including \emph{Enterobacteriaceae},
+administered either a 1-day PEG treatment, a 1-day PEG treatment with a
+1-day recovery period, or clindamycin to mice before challenging them
+with \emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
+mice, the 1-day PEG treated mice were only transiently colonized and
+cleared \emph{C. difficile} by 7 days post-infection (Fig. 3B). The
+stool communities of PEG-treated mice were also only transiently
+disrupted, with Shannon diversity recovering close to baseline by 7 days
+post-infection (Fig. 3C-D). We found 14 bacteria were impacted by
+treatment, but recovered close to baseline levels by 7 days
+post-infection including \emph{Enterobacteriaceae},
 \emph{Clostridiales}, \emph{Porpyromonadaceae}, and
@@
 after challenge, and 1-day PEG treatment 3 days after challenge followed
-by either administration of an FMT or PBS solution via gavage (Fig. 5A).
-In contrast to our hypothesis, all groups of mice that received PEG
-actually exhibited prolonged \emph{C. difficile} colonization (Fig. 5B).
+by either administration of an FMT or PBS solution by oral gavage (Fig.
+5A). Contrary to our hypothesis, all groups of mice that received PEG
+exhibited prolonged \emph{C. difficile} colonization (Fig. 5B). Next we
+examined how post-CDI PEG treatment impacted the microbiota. Alpha
+diversity was lower in the PEG-treated mice with the exception of the
+PEG-treated mice that were administered an FMT (Fig 5C-D). The FMT
+appeared to partially restore Shannon diversity but not richness (Fig.
+5C-D). Similarly, we saw some overlap between the communities of mice
+that received FMT and the mice treated with only clindamycin after 5
+days post-infection (Fig. 6A). The increase in Shannon diversity
+suggests that the FMT did have an impact on the microbiota, despite
+seeing prolonged \emph{C. difficile} colonization in the FMT treated
+mice. However, only \emph{Bacteroidales} and \emph{Porphyromonadaceae}
+consistently differed between the mice received either an FMT or PBS
+gavage (Fig. 6B), suggesting the FMT only restored a couple of genera.
+Overall, we found there were 24 bacteria that differed over multiple
+time points between treatment groups including increased
+\emph{Akkermansia} and decreased \emph{Ruminococcaceae},
+\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Oscillibacter}
+in mice that received PEG after \emph{C. difficile} challenge (Fig. 6C).
+In sum, administering PEG actually prolonged \emph{C. difficile}
+colonization, including in mice that received an FMT, which only
+restored a couple of bacterial genera.
 
-\begin{itemize}
-\item
-  Draft of results
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Prolonged \emph{C. difficile} colonization in post-CDI PEG treated
-    mice (Fig. 5B)
-  \item
-    FMT appears to partially restore alpha diversity (Shannon, but not
-    richness Fig. 5C-D)
-  \item
-    PCoA (necessary?) Could comment on the clustering of Clindamycin and
-    PEG-treated mice that received FMT (Fig. 6A)
-  \item
-    Only 2 genera significantly impacted by FMT treatment, likely not as
-    important for clearance (Fig. 6B)
-  \item
-    Bacteria that consistently differ between groups over time,
-    associated with prolonged colonization (Fig. 6C)
-  \end{itemize}
-\end{itemize}
-
-\textbf{Five-day post-infection community data can predict mice that
-will have prolonged \emph{C. difficile} colonization} After identifying
-bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG
-treatments, we decided to examine the taxa that were influencing
+\textbf{Five-day post-infection community data can predict which mice
+that will have prolonged \emph{C. difficile} colonization} After
+identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day
+PEG treatments, we decided to examine the bacteria that influenced
 prolonged \emph{C. difficile} colonization. We trained 3 types of
@@
 vector machine) with input bacterial community data from 5 days
-post-infection to predict whether the mice were colonized with \emph{C.
-difficile} 10 days post-infection. We chose 5 days post-infection
-because that was the earliest timepoint where we would see a treatment
-effect in the mice that were given 1-day PEG treatment three days after
-\emph{C. difficile} challenge and then administered an FMT or PBS
-gavage. The random forest model had the highest performance (AUROC =
-0.90, Data Set S1, Sheet X), so we next performed permutation importance
-to examine the bacteria that were driving performance. We selected the
-top 10 bacteria contributing to our models performance (Fig. 7A) and
-examined their relative abundance at 5 days post-infection, the
-timepoint used to predict C. difficile colonization status on day 10
-(Fig. 7B). Next, we focused on the 5 genera that were \textgreater{} 1
-\% relative abundance in either the cleared or colonized mice and
-examined the relative abundance dynamics of these bacteria over time. We
-found \emph{, }, and \_ tended to have a higher relative abundance and
-\textbf{, } had a lower relative abundance in the mice with prolonged
-colonization compared to the mice that cleared \emph{C. difficile} (Fig.
-7C). Previous work examining the impact of PEG on the murine microbiota
-found that PEG treatment resulted in the permanent loss of S24-7, also
-known as \emph{Muribaculum intestinale} (16). We decided to check our
+post-infection to predict whether the mice were still colonized with
+\emph{C. difficile} 10 days post-infection. We chose 5 days
+post-infection because that was the earliest time point where we would
+see a treatment effect in the mice that were given 1-day PEG treatment
+three days after \emph{C. difficile} challenge and then administered an
+FMT or PBS gavage. The random forest model had the highest performance
+(AUROC = 0.90, Data Set S1, Sheet X), so we next performed permutation
+importance to examine the bacteria that were the top contributors to the
+random forest model predicting prolonged \emph{C. difficile
+}colonization. We selected the top 10 bacteria contributing to our
+models performance (Fig. 7A) and examined their relative abundance at 5
+days post-infection, the time point used to predict \emph{C. difficile}
+colonization status on day 10 (Fig. 7B). Next, we focused on the 5
+genera that had a greater than 1 \% relative abundance in either the
+cleared or colonized mice and examined how the bacteria changed over
+time. We found \emph{Enterobacteriaceae} and \emph{Bacteroides} tended
+to have a higher relative abundance, \emph{Akkermansia} was initially
+decreased and then increased, and \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} had a lower relative abundance in the mice with
+prolonged colonization compared to the mice that cleared \emph{C.
+difficile} (Fig. 7C). Together these results suggest a combination of
+low and high abundance bacterial genera influence the prolonged
+colonization observed in 5-day PEG and post-CDI PEG treated mice.
+
+Previous work examining the impact of PEG on the murine microbiota found
+that PEG treatment resulted in the permanent loss of S24-7, also known
+as \emph{Muribaculum intestinale} (16). We decided to check our
 \emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
-known to be classified as \emph{Porphyromonadaceae} by the Ribosomal
-Database Project (RDP) database (31), \emph{Porphyromonadaceae} was a
-top feature in the random forest model predicting prolonged \emph{C.
-difficile} colonization, and had a high relative abundance in the
-communities of mice that cleared \emph{C. difficile} within 10 days. We
-identified 4 OTUs that had at least (insert minimum percent identity) to
-\emph{Muribaculum intestinale} and examined their abundance in mice that
-either cleared or were still colonized with \emph{C. difficile} at 10
-days post-infection (Fig. S4).
+classified as \emph{Porphyromonadaceae} by the Ribosomal Database
+Project (RDP) database (31) and \emph{Porphyromonadaceae} was a top
+feature in the random forest model predicting prolonged \emph{C.
+difficile} colonization. We identified 4 OTUs that had at least 92\%
+identity to \emph{Muribaculum intestinale} and examined their abundance
+in mice that either cleared or were still colonized with \emph{C.
+difficile} at 10 days post-infection. While all of the OTUs, were
+decreased by PEG and clindamycin treatment, there was some recovery in
+the mice that cleared (Fig. S4A). We also examined other
+\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs since these
+were the 2 genera that were important to our classification model and
+contained multiple OTUs that were different at 5 days post-infection
+between mice that either cleared or remained colonized with \emph{C.
+difficile} by 10 days post-infection (Data Set S1, sheet X). While
+individual \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs
+tended to be more abundant in the mice that clear \emph{C. difficile}
+relative to the mice that exhibit prolonged colonization (Fig. S4B-C),
+there is no single OTU that fits the pattern we observed at the genus
+level (Fig. 7C), suggesting multiple \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} OTUs influenced \emph{C. difficile} clearance.
+Overall, our results suggest that specific bacterial community
+differences explain the prolonged \emph{C. difficile} colonization we
+observed in 5-day PEG and post-CDI 1-day PEG treated mice.
 
@@
 \emph{C. difficile} CFU/gram stool measured over time (N = 16-59 mice
-per timepoint) via serial dilutions. The black line represents the limit
-of detection for the first serial dilution. CFU quantification data was
-not available for each mouse due to stool sampling difficulties
+per time point) via serial dilutions. The black line represents the
+limit of detection for the first serial dilution. CFU quantification
+data was not available for each mouse due to stool sampling difficulties
 (particularly the day the mice came off of the PEG treatment) or early
 deaths. For B-C, lines represent the median for each treatment group and
-circles represent samples from individual mice. Asterisks indicate
-timepoints where the weight change or CFU/g was significantly different
+circles represent samples from individual mice. Asterisks indicate time
+points where the weight change or CFU/g was significantly different
 between groups by the Kruskal-Wallis test with Benjamini-Hochberg
-correction for testing multiple timepoints. The data presented are from
+correction for testing multiple time points. The data presented are from
 a total of 5 separate experiments. \newpage
@@
 A. Setup of the experimental timeline for the 1-day PEG treated subset
-of mice. B. CFU/gram stool measured over time (N = 12-18 mice per
-timepoint) via several dilutions. The black dotted line represents the
-limit of detection for the first serial dilution. Asterisks indicate
-timepoints where the CFU/gram was significantly different between groups
+of mice. B. CFU/gram stool measured over time (N = 12-18 mice per time
+point) via several dilutions. The black dotted line represents the limit
+of detection for the first serial dilution. Asterisks indicate time
+points where the CFU/gram was significantly different between groups
 using the Kruskall-Wallis test with a Benjamini-Hochberg correction for
-multiple timepoints. C. Principle Coordinate Analysis plot of the groups
-over time with the alpha representating the same time scale as in panel
-D (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} =
+multiple time points. C. Principle Coordinate Analysis plot of the
+groups over time with the alpha representating the same time scale as in
+panel D (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} =
 0.19). D. Shannon diversity Index of the groups over time. Only days

### 2021-04-26 submission/manuscript.Rmd
Update filenames

@@
 #Genera that were changed by treatment
-pt_d1 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_PTtoD1.tsv")) %>%
+pt_d1 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_BtoD1.tsv")) %>%
   filter(p.value.adj < 0.05)
 #Genera that were still different at 7 days post-infection  
-pt_d7 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_PTtoD7.tsv")) %>%
+pt_d7 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_BtoD7.tsv")) %>%
   filter(p.value.adj < 0.05)

### 2021-04-28 code/post_CDI_PEG_16S.R
Clean up post cdi 16S script, delete tiss. sobs and shannon

@@ -117,49 +117,6 @@ kruskal_wallis_richness <- function(diversity_subset, timepoints, subset_name){
 kw_richness_stools <- kruskal_wallis_richness(diversity_stools, stool_test_days, "stools")
 sig_richness_days_stools <- pull_sig_days(kw_richness_stools)
 
-#Plot Richness overtime 
-sobs_post_CDI_PEG <- diversity_data_subset %>%
-  filter(group != "FMT") %>% #Remove FMTs
-  group_by(group, day) %>%
-  mutate(median_sobs = median(sobs)) %>%
-  ggplot(x = day, y = sobs, colour =  group) +
-  geom_point(mapping = aes(x = day, y = sobs, color = group, fill = group), alpha = .2, size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
-  geom_line(mapping = aes(x = day, y = median_sobs, color = group), alpha = 0.6, size = 1) +
-  scale_colour_manual(name=NULL,
-                      values=color_scheme,
-                      breaks=color_groups,
-                      labels=color_labels) +
-  scale_x_continuous(breaks = c(-15, -1:10, 15, 20, 25, 30),
-                     limits = c(-15,35), 
-                     minor_breaks = c(-15.5, -14.5, -1.5:10.5, 14.5, 15.5, 19.5, 20.5, 24.5, 25.5, 29.5, 30.5)) +
-  theme_classic()+
-  theme(legend.position = "none")+ #Remove legend
-  labs(x = "Days Post-Infection",
-       y = "Number of Observed OTUs")
-save_plot("results/figures/post_CDI_PEG_richness_overtime.png", sobs_post_CDI_PEG)
-
-#Richness oVertime 10 Day Version
-sobs_post_CDI_PEG_10d <- diversity_data_subset_10d %>%
-  filter(group != "FMT") %>% #Remove FMTs
-  group_by(group, day) %>%
-  mutate(median_sobs = median(sobs)) %>%
-  ggplot(x = day, y = sobs, colour =  group) +
-  geom_point(mapping = aes(x = day, y = sobs, color = group, fill = group), alpha = .2, size = 1.5, show.legend = FALSE, position = position_dodge(width = 0.6)) +
-  geom_line(mapping = aes(x = day, y = median_sobs, color = group), alpha = 0.6, size = 1) +
-  scale_colour_manual(name=NULL,
-                      values=color_scheme,
[…]

### 2021-04-28 submission/manuscript.Rmd
Figure legend edits and additions

@@
 1B).
-After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C.
+After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C.
 difficile* 630 spores (Fig.
@@
 2E, Data Set S1, Sheet X).
-Interestingly, one of the bacteria that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera with an OTU sequence that matches *C.
+Interestingly, one of the bacteria that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera that contains a sequence that matches *C.
 difficile*).
@@
 Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of S24-7, also known as *Muribaculum intestinale* [@Tropini2018].
-We decided to check our *Porphyromonadaceae* OTUs because *Muribaculum intestinale* is classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020] and *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
+We decided to check our *Porphyromonadaceae* operational taxonomic units (OTUs) because *Muribaculum intestinale* is classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020] and *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
 difficile* colonization.
@@
 
+#Figure out which experiments eneded at 4, 6, or 30 days post-infection
+five_day_mice_samples <- five_day_mice %>% 
+  group_by(exp_num, day, sample_type) %>% 
+  tally()
+#Figure out number of stool samples from 4,6, 10, 20 and 30 days post-infection
+five_day_mice_samples <- five_day_mice %>% 
+  filter(sample_type == "stool" & day %in% c(4, 6, 10, 20, 30)) %>% 
+  group_by(day) %>% 
+  tally()
+
 ```
@@
 difficile*.** A.
-Setup of the experimental timeline for experiments with 5-day PEG treated mice.
+Setup of the experimental time line for experiments with 5-day PEG treated mice consisting of 4 treatment groups.
+1.
 Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
+2.
 15% PEG 3350 was administered in the drinking water for five days.
+3.
+5-day PEG plus clindamycin treatment.
+4.
+5-day PEG plus 10-day recovery treatment.
+All treatment groups were then challenged with 10^5^ *C.
+difficile* 630 spores.
+A subset of mice were euthanized on either 4 or 6 days post-infection and tissues were collected for histopathology analysis, the remaining mice were followed through 20 or 30 days post-infection.
 B.
@@
 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A.
-PCoA of Bray-Curtis distances from stool samples collected throughout the experiment.
+Principal Coordinate analysis (PCoA) of Bray-Curtis distances from stool samples collected throughout the experiment.
+For A and C, each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-infection.
+B.
+Bray-Curtis distances of stool samples collected on either day 10 or 30 post-infection relative to the baseline sample collected for each mouse (before any drug treatments were administered).
+C.
+Shannon diversity in stool communities over time.
+The line indicates the median value for each treatment group.
+The colors of the symbols and lines represent the four treatment groups.
+D.
+14 of the `r f_d_group_total_genera` genera affected by PEG treatment  (Data Set S1, sheet X).
+The symbols represent the median relative abundance for a treatment group at either baseline (open circle) or 1-day post treatment (closed circle).
+Data from the 5-day PEG and 5-day PEG plus 10-day recovery groups were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera.
+The clindamycin and 5-day PEG plus clindamycin treatment groups are shown for comparison.
+E.
+6 of the `r f_d_group_multi_day` genera that were significantly different between the four treatment groups over multiple time points.
+Differences between treatment groups were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified genera.
+The gray vertical line (D) and horizontal vertical lines (E) indicate the limit of detection.
 \newpage
@@
 \includegraphics{figure_3.pdf}
-**Figure 3. 5-day PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.** A.
+**Figure 3.
+5-day PEG treatment does not result in more severe CDIs, although mucosal microbiota is altered.** A.
+Shannon diversity in cecum communities over time.
+The line indicates the median value for each treatment group.
+The colors of the symbols and lines represent the four treatment groups.
+A similar pattern was observed with the proximal and distal colon communities (Data Set S1, sheet X-X).
+B.
+PCoA of Bray-Curtis distances from mucosal samples collected throughout the experiment.
+Circles, triangles, and squares indicate cecum, proximal colon, and distal colon communities, respectively.
+For A-B, transparency of the symbol corresponds to the day post-infection that the sample was collected.
+C.
+The median relative abundance of the 4 genera that were significantly different between the cecum communities of different treatment groups on day 6 and day 30  (Data Set S1, sheet X).
+The gray vertical line indicate the limit of detection.
+D-E.
+The histopathology summary scores from cecum and colon H&E stained slides.
+The summary score is the total score based on evaluation of edema, cellular infiltration, and inflammation.
+Each category is given a score ranging from 0-4, thus the maximum possible summary score is 12.
+The tissue for histology was collected at either 4 (D) or 6 (E) days post-infection with the exception that one set of 5-day PEG treated mock-challenged mice were collected on day 0 post-infection (first set of open circles in D).
+Histology data were analyzed with the Kruskal-Wallis test followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
 \newpage
@@
 difficile* colonization.** A.
-Setup of the experimental timeline for the 1-day PEG treated subset of mice.
+Setup of the experimental time line for the 1-day PEG treated  mice consisting of 3 treatment groups.
+1.
+Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
+2.
+15% PEG 3350 was administered in the drinking water for 1 day.
+3.
+1-day PEG plus 1-day recovery.
+The three treatment groups were then challenged with 10^5^ *C.
+difficile* 630 spores.
 B.
-CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per time point) via several dilutions.
+*C.
+difficile* CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per time point) via several dilutions.
 The black dotted line represents the limit of detection for the first serial dilution.
-Asterisks indicate time points where the CFU/gram was significantly different between groups using the Kruskall-Wallis test with a Benjamini-Hochberg correction for multiple time points.
+Asterisks indicate time points where the CFU/gram was significantly different between treatment groups by Kruskall-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
 C.
-Principle Coordinate Analysis plot of the groups over time with the alpha representating the same time scale as in panel D (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`).
+PCoA of Bray-Curtis distances from stool communities collected from the three treatment groups over time (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`).
 D.
-Shannon diversity Index of the groups over time.
-Only days with samples from all groups are shown.
+Shannon diversity in stool communities over time with colored lines representing the median value for each treatment group.
+For B-D, each symbol represents a sample from an individual mouse and symbol transparency corresponds to the day post-infection that the sample was collected.
 E.
-Line plots of relative percent abundance of selected genera over time.
-Only days with samples from all groups shown.
-The gray line represents the limit of detection.
+Median relative abundances per treatment group for 6 out of the `r recovered_genera` genera that were affected by treatment, but recovered close to baseline levels by 7 days post-infection (Fig.
+3E, Data Set S1, Sheet X).
+Stool samples from either baseline and day 1 or baseline and day 7 were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera.
+The gray horizontal line represents the limit of detection.
 
@@
 fig_5_permanova_sig_R2 <- r_sq_ordered_perm[,2] %>% sum() #0.9280951
+
+#Number of experiments represented in [post-CDI PEG subset
+#Use same code from post-CDI_PEG_subset() in code/utilities.R to determine number of samples we had C. difficile CFU counts for per day
+post_CDI_mice <- metadata %>%     
+  filter(group == "C" & exp_num %in% c("M7","M9")| #Only use C mice from these experiments. Allocated groups to figures based on paper outline.
+             group == "CWM" & exp_num %in% c("M6","M7", "M9")|
+             group == "RM" & exp_num %in% c("M7","M9")|
+             group == "FRM" & exp_num %in% c("M9"))
+#Total number of experiments in post-CDI subset
+post_CDI_total_exp <- post_CDI_mice %>%
+  filter(!duplicated(exp_num)) %>% #Remove duplicate experiment numbers
+  tally() #Total number of unique experiments
 ```
@@
 1-day PEG treatment post *C.
-difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C.
-difficile* 630 spores on day 0.
-15% PEG 3350 was administered in the drinking water for 1 day.
-Mice feces were collected daily through the end of the experiment (30 days post-infection for Clind.
-+ 1-day PEG group, 15 days post-infection for all others).
-CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths.
+difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** A.
+Setup off the experimental time line for experiments with post-CDI PEG treated mice.
+There were a total of 4 different treatment groups.
+All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with 10^3-5^ *C.
+difficile* 630 spores.
+1.
+Did not receive any additional treatment.
+2.
+Immediately after *C.
+difficile* challenge, mice received 15% PEG 3350 in the drinking water for 1 day.
+3-4.
+3-days after challenge, mice received 1-day PEG treatment and then received either a fecal microbiota transplant (3) or PBS (4) solution by oral gavage.
+Mice were followed through 15-30 days post-infection (only the post-CDI 1-day PEG group was followed through 30 days post-infection).
 B.
@@
 difficile* stool measured over time via serial dilutions.
-C.
-Median and individual Shannon Diversity measured over time.
-D.
-Median and individual richness measured over time.
+The black line represents the limit of detection for the first serial dilution.
+C-D.
+Shannon diversity (C) and richness (D) in stool communities over time.
 B-D.
-Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction for testing multiple times points.
-Black line represents the limit of detection for the first serial dilution.
-Background shading indicates PEG treatment period for applicable groups.
-Opacity of points reflects time point.
-The data presented are from a total of *** separate experiments.
+Each symbol represents a stool sample from an individual mouse with the lines representing the median value for each treatment group.
+The transparancy of the symbol corresponds to the day post-infection.
+Asterisks indicate time points with significant differences between groups by a Kruskall-Wallis test with a Benjamini-Hochberg correction for testing multiple times points.
+Colored rectangles indicates the 1-day PEG treatment period for applicable groups.
+The data presented are from a total of `r post_CDI_total_exp` separate experiments.
 \newpage
@@
 difficile* challenge mice that also receive an FMT only some bacterial genera were restored.** A.
-PCoA of all groups + FMT plotted over time.
-Opacity of points reflects time point.
+PCoA of Bray-Curtis distances from stool samples collected over time as well as the FMT solution that was administered to one treatment group.
+Each circle represents an individual sample, the transparency of the circle corresponds to day post-infection as shown in Fig.
+6C-D.
 B.
-Relative abundance of genera significantly different over multiple time points post FMT or PBS treatment, plotted over time.
-Limit of detection is .1%.
-C.
-Relative abundances of top 6 significant genera ranked by number of days significant, plotted over time.
-Limit of detection is .1%.
+Median relative abundances of 2 genera that were significantly different over multiple time points in mice that were administered either FMT or PBS solution via gavage C.
+Median relative abundances of the top 6 out of `r post_group_multi_day` genera that were significant over multiple timepoints, plotted over time (Data Set S1, Sheet X).
+For B-C, colored rectangles indicates the 1-day PEG treatment period for applicable groups.
+Gray horizontal lines represent the limit of detection.
 
@@
 \includegraphics{figure_7.pdf}
-**Figure 7. Specific microbiota features associated with prolonged *C. difficile* colonization in PEG treated mice.** A.
+**Figure 7.
+Specific microbiota features associated with prolonged *C.
+difficile* colonization in PEG treated mice.** A.
+Top ten bacteria that contributed to the random forest model trained on five day post-infection community relative abundance data to predict wehther mice would still be colonized with *C.
+difficile* 10 days post-infection.
+The median (point) and interquantile range (lines) change in AUROC when the bacteria is left out of the model is shown.
+B.
+The median relative abundances at 5 days post-infection of the top ten bacteria that contributed to the random forest classification model.
+Color indicates whether the mice were still colonized with *C.
+difficile* 10 days post-infection and the black horizontal line represents the median relative abundance.
+Each symbol represents a stool sample from an individual mouse and the shape of the symbol indicates whether the PEG-treated mice recieved a 5-day (Fig.
+1-3), 1-day (Fig.
+4) or post-CDI PEG (Fig.
+5-6) treatment.
+C.
+The median relative abundances of the 5 genera with greater than 1% median relative abundance in the stool community over time.
+For B-C, the gray horizontal line represents the limit of detection.
 \newpage
@@
 \includegraphics{figure_8.pdf}
-**Figure 8. Schematic summarizing findings.** A.
+**Figure 8.
+Schematic summarizing findings.** A.
+The gut microbiota of our C57Bl/6 mice is restant to *C.
+difficle* but treatment with either the antibiotic, clindamycin, or the osmotic laxative, PEG 3350 renders the mice susceptible to *C.
+difficile* colonization.
+Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-infection.
+However, for mice that recieved a 5-day PEG treatment or a 1-day PEG treatment after *C.
+difficile* challenge recovery of colonization resistance is more uncertain because mice were still colonized with *C.
+difficile* 30 days post-infection in the case of several PEG treatments.
+We found increased *Porphyromonadaceae* and *Lachnospiraceae* were associated with recovery of colonization resistance, while increased *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C.
+difficle* colonization.
+B.Considering that most hospitals recommend not performing *C.
+difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs.
+further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.
+Further studies are needed to understand the impact of osmotic laxatives on *C.
+difficile* colonization resistance and clearance in human patients.
 \newpage
@@
 \includegraphics{figure_S1.pdf}
-**Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota dynamics post-infection.** A.
+**Figure S1.
+Microbiota dynamics post-infection in the 5-day PEG treatment plus 10-day recovery mice.** A.
+*C.
+difficile* CFU/g over time in the stool samples collected from 5-day PEG treatment plus 10-day recovery mice.
+Same data presented in Fig.
+1C, but the data for the other 3 treatment groups have been removed.
+B.
+Median relative abundances of 8 bacterial genera from day 0 post-infection onward from the 10-day recovery mice.
+We analyzed samples from day 0 and day 8 post-infection, which represented the the time points where mice were challenged with *C.
+difficile* and when the median relative *C.
+difficile* CFU stabilized for the group using the paired Wilcoxan signed-rank test, but no genera were signifcant after Benjamini-Hochberg correction.
 \newpage
@@
 \includegraphics{figure_S2.pdf}
-**Figure S2. Impact of PEG treatment on the mucosal microbiota is greatest 6 days post-infection**  
+**Figure S2. PEG treatment still has a large impact on the mucosal microbiota 6 days post-infection** A. The   
 \newpage
@@
 \includegraphics{figure_S4.pdf}
-**Figure S4. Specific OTUs associated with clearance that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A.
+**Figure S4. Specific OTUs associated with clearance by 10 days post-infection that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A.
 \newpage

### 2021-04-28 submission/manuscript.tex
Figure legend edits and additions

@@
 clindamycin, or PEG and clindamycin treatment all mice were challenged
-with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 1A). All
+with 10\textsuperscript{5} \emph{C. difficile} 630 spores (Fig. 1A). All
 treatments rendered mice susceptible to \emph{C. difficile}
@@
 groups at 6 days post-infection was \emph{Peptostreptococcaceae} (the
-genera with an OTU sequence that matches \emph{C. difficile}).
+genera that contains a sequence that matches \emph{C. difficile}).
 \emph{Peptostreptococcaceae} was primarily only present in the 5-day PEG
@@
 as \emph{Muribaculum intestinale} (16). We decided to check our
-\emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
-classified as \emph{Porphyromonadaceae} by the Ribosomal Database
-Project (RDP) database (31) and \emph{Porphyromonadaceae} was a top
-feature in the random forest model predicting prolonged \emph{C.
-difficile} colonization. We identified 4 OTUs that had at least 92\%
-identity to \emph{Muribaculum intestinale} and examined their abundance
-in mice that either cleared or were still colonized with \emph{C.
-difficile} at 10 days post-infection. While all of the OTUs, were
-decreased by PEG and clindamycin treatment, there was some recovery in
-the mice that cleared (Fig. S4A). We also examined other
-\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs since these
-were the 2 genera that were important to our classification model and
-contained multiple OTUs that were different at 5 days post-infection
-between mice that either cleared or remained colonized with \emph{C.
-difficile} by 10 days post-infection (Data Set S1, sheet X). While
-individual \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs
-tended to be more abundant in the mice that clear \emph{C. difficile}
-relative to the mice that exhibit prolonged colonization (Fig. S4B-C),
-there is no single OTU that fits the pattern we observed at the genus
-level (Fig. 7C), suggesting multiple \emph{Porphyromonadaceae} and
-\emph{Lachnospiraceae} OTUs influenced \emph{C. difficile} clearance.
-Overall, our results suggest that specific bacterial community
-differences explain the prolonged \emph{C. difficile} colonization we
-observed in 5-day PEG and post-CDI 1-day PEG treated mice.
+\emph{Porphyromonadaceae} operational taxonomic units (OTUs) because
+\emph{Muribaculum intestinale} is classified as
+\emph{Porphyromonadaceae} by the Ribosomal Database Project (RDP)
+database (31) and \emph{Porphyromonadaceae} was a top feature in the
+random forest model predicting prolonged \emph{C. difficile}
+colonization. We identified 4 OTUs that had at least 92\% identity to
+\emph{Muribaculum intestinale} and examined their abundance in mice that
+either cleared or were still colonized with \emph{C. difficile} at 10
+days post-infection. While all of the OTUs, were decreased by PEG and
+clindamycin treatment, there was some recovery in the mice that cleared
+(Fig. S4A). We also examined other \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} OTUs since these were the 2 genera that were
+important to our classification model and contained multiple OTUs that
+were different at 5 days post-infection between mice that either cleared
+or remained colonized with \emph{C. difficile} by 10 days post-infection
+(Data Set S1, sheet X). While individual \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} OTUs tended to be more abundant in the mice that
+clear \emph{C. difficile} relative to the mice that exhibit prolonged
+colonization (Fig. S4B-C), there is no single OTU that fits the pattern
+we observed at the genus level (Fig. 7C), suggesting multiple
+\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs influenced
+\emph{C. difficile} clearance. Overall, our results suggest that
+specific bacterial community differences explain the prolonged \emph{C.
+difficile} colonization we observed in 5-day PEG and post-CDI 1-day PEG
+treated mice.
 
@@
 become persistently colonized with \emph{C. difficile}.} A. Setup of the
-experimental timeline for experiments with 5-day PEG treated mice.
-Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
-15\% PEG 3350 was administered in the drinking water for five days. B.
-Weight change from baseline weight in groups after treatment with PEG
-and/or clindamycin, followed by \emph{C. difficile} challenge. C.
-\emph{C. difficile} CFU/gram stool measured over time (N = 16-59 mice
-per time point) via serial dilutions. The black line represents the
-limit of detection for the first serial dilution. CFU quantification
-data was not available for each mouse due to stool sampling difficulties
-(particularly the day the mice came off of the PEG treatment) or early
-deaths. For B-C, lines represent the median for each treatment group and
-circles represent samples from individual mice. Asterisks indicate time
-points where the weight change or CFU/g was significantly different
-between groups by the Kruskal-Wallis test with Benjamini-Hochberg
-correction for testing multiple time points. The data presented are from
-a total of 5 separate experiments. \newpage
+experimental time line for experiments with 5-day PEG treated mice
+consisting of 4 treatment groups. 1. Clindamycin was administered at 10
+mg/kg by intraperitoneal injection. 2. 15\% PEG 3350 was administered in
+the drinking water for five days. 3. 5-day PEG plus clindamycin
+treatment. 4. 5-day PEG plus 10-day recovery treatment. All treatment
+groups were then challenged with 10\textsuperscript{5} \emph{C.
+difficile} 630 spores. A subset of mice were euthanized on either 4 or 6
+days post-infection and tissues were collected for histopathology
+analysis, the remaining mice were followed through 20 or 30 days
+post-infection. B. Weight change from baseline weight in groups after
+treatment with PEG and/or clindamycin, followed by \emph{C. difficile}
+challenge. C. \emph{C. difficile} CFU/gram stool measured over time (N =
+16-59 mice per time point) via serial dilutions. The black line
+represents the limit of detection for the first serial dilution. CFU
+quantification data was not available for each mouse due to stool
+sampling difficulties (particularly the day the mice came off of the PEG
+treatment) or early deaths. For B-C, lines represent the median for each
+treatment group and circles represent samples from individual mice.
+Asterisks indicate time points where the weight change or CFU/g was
+significantly different between groups by the Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing multiple time points. The data
+presented are from a total of 5 separate experiments. \newpage
 
@@
 disrupts the stool microbiota for a longer amount of time compared to
-clindamycin-treated mice.} A. PCoA of Bray-Curtis distances from stool
-samples collected throughout the experiment. \newpage
+clindamycin-treated mice.} A. Principal Coordinate analysis (PCoA) of
+Bray-Curtis distances from stool samples collected throughout the
+experiment. For A and C, each circle represents a sample from an
+individual mouse and the transparency of the symbol corresponds to the
+day post-infection. B. Bray-Curtis distances of stool samples collected
+on either day 10 or 30 post-infection relative to the baseline sample
+collected for each mouse (before any drug treatments were administered).
+C. Shannon diversity in stool communities over time. The line indicates
+the median value for each treatment group. The colors of the symbols and
+lines represent the four treatment groups. D. 14 of the 33 genera
+affected by PEG treatment (Data Set S1, sheet X). The symbols represent
+the median relative abundance for a treatment group at either baseline
+(open circle) or 1-day post treatment (closed circle). Data from the
+5-day PEG and 5-day PEG plus 10-day recovery groups were analyzed by
+paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for
+testing all identified genera. The clindamycin and 5-day PEG plus
+clindamycin treatment groups are shown for comparison. E. 6 of the 24
+genera that were significantly different between the four treatment
+groups over multiple time points. Differences between treatment groups
+were identified by Kruskal-Wallis test with Benjamini-Hochberg
+correction for testing all identified genera. The gray vertical line (D)
+and horizontal vertical lines (E) indicate the limit of detection.
+\newpage
 
@@
 does not result in more severe CDIs, although mucosal microbiota is
-altered.} A. \newpage
+altered.} A. Shannon diversity in cecum communities over time. The line
+indicates the median value for each treatment group. The colors of the
+symbols and lines represent the four treatment groups. A similar pattern
+was observed with the proximal and distal colon communities (Data Set
+S1, sheet X-X). B. PCoA of Bray-Curtis distances from mucosal samples
+collected throughout the experiment. Circles, triangles, and squares
+indicate cecum, proximal colon, and distal colon communities,
+respectively. For A-B, transparency of the symbol corresponds to the day
+post-infection that the sample was collected. C. The median relative
+abundance of the 4 genera that were significantly different between the
+cecum communities of different treatment groups on day 6 and day 30
+(Data Set S1, sheet X). The gray vertical line indicate the limit of
+detection. D-E. The histopathology summary scores from cecum and colon
+H\&E stained slides. The summary score is the total score based on
+evaluation of edema, cellular infiltration, and inflammation. Each
+category is given a score ranging from 0-4, thus the maximum possible
+summary score is 12. The tissue for histology was collected at either 4
+(D) or 6 (E) days post-infection with the exception that one set of
+5-day PEG treated mock-challenged mice were collected on day 0
+post-infection (first set of open circles in D). Histology data were
+analyzed with the Kruskal-Wallis test followed by pairwise Wilcoxon
+comparisons with Benjamini-Hochberg correction. \newpage
 
@@
 renders mice susceptible to transient \emph{C. difficile} colonization.}
-A. Setup of the experimental timeline for the 1-day PEG treated subset
-of mice. B. CFU/gram stool measured over time (N = 12-18 mice per time
-point) via several dilutions. The black dotted line represents the limit
-of detection for the first serial dilution. Asterisks indicate time
-points where the CFU/gram was significantly different between groups
-using the Kruskall-Wallis test with a Benjamini-Hochberg correction for
-multiple time points. C. Principle Coordinate Analysis plot of the
-groups over time with the alpha representating the same time scale as in
-panel D (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} =
-0.19). D. Shannon diversity Index of the groups over time. Only days
-with samples from all groups are shown. E. Line plots of relative
-percent abundance of selected genera over time. Only days with samples
-from all groups shown. The gray line represents the limit of detection.
+A. Setup of the experimental time line for the 1-day PEG treated mice
+consisting of 3 treatment groups. 1. Clindamycin was administered at 10
+mg/kg by intraperitoneal injection. 2. 15\% PEG 3350 was administered in
+the drinking water for 1 day. 3. 1-day PEG plus 1-day recovery. The
+three treatment groups were then challenged with 10\textsuperscript{5}
+\emph{C. difficile} 630 spores. B. \emph{C. difficile} CFU/gram stool
+measured over time (N = 12-18 mice per time point) via several
+dilutions. The black dotted line represents the limit of detection for
+the first serial dilution. Asterisks indicate time points where the
+CFU/gram was significantly different between treatment groups by
+Kruskall-Wallis test with Benjamini-Hochberg correction for testing
+multiple time points. C. PCoA of Bray-Curtis distances from stool
+communities collected from the three treatment groups over time (day:
+R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} = 0.19). D.
+Shannon diversity in stool communities over time with colored lines
+representing the median value for each treatment group. For B-D, each
+symbol represents a sample from an individual mouse and symbol
+transparency corresponds to the day post-infection that the sample was
+collected. E. Median relative abundances per treatment group for 6 out
+of the 14 genera that were affected by treatment, but recovered close to
+baseline levels by 7 days post-infection (Fig. 3E, Data Set S1, Sheet
+X). Stool samples from either baseline and day 1 or baseline and day 7
+were analyzed by paired Wilcoxan signed-rank test with
+Benjamini-Hochberg correction for testing all identified genera. The
+gray horizontal line represents the limit of detection.
 
@@
 post \emph{C. difficile} challenge prolongs colonization regardless of
-whether an FMT is also administered.} All mice were administered 10
-mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
-\emph{C. difficile} 630 spores on day 0. 15\% PEG 3350 was administered
-in the drinking water for 1 day. Mice feces were collected daily through
-the end of the experiment (30 days post-infection for Clind. + 1-day PEG
-group, 15 days post-infection for all others). CFU quantification data
-was not available for each mouse due to stool sampling difficulties
-(particularly the day the mice came off of the PEG treatment) or early
-deaths. B. CFU/g of \emph{C. difficile} stool measured over time via
-serial dilutions. C. Median and individual Shannon Diversity measured
-over time. D. Median and individual richness measured over time. B-D.
-Asterisks indicate time points with significant differences between
-groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction
-for testing multiple times points. Black line represents the limit of
-detection for the first serial dilution. Background shading indicates
-PEG treatment period for applicable groups. Opacity of points reflects
-time point. The data presented are from a total of *** separate
+whether an FMT is also administered.} A. Setup off the experimental time
+line for experiments with post-CDI PEG treated mice. There were a total
+of 4 different treatment groups. All mice were administered 10 mg/kg
+clindamycin intraperitoneally (IP) 1 day before challenge with
+10\textsuperscript{3-5} \emph{C. difficile} 630 spores. 1. Did not
+receive any additional treatment. 2. Immediately after \emph{C.
+difficile} challenge, mice received 15\% PEG 3350 in the drinking water
+for 1 day. 3-4. 3-days after challenge, mice received 1-day PEG
+treatment and then received either a fecal microbiota transplant (3) or
+PBS (4) solution by oral gavage. Mice were followed through 15-30 days
+post-infection (only the post-CDI 1-day PEG group was followed through
+30 days post-infection). B. CFU/g of \emph{C. difficile} stool measured
+over time via serial dilutions. The black line represents the limit of
+detection for the first serial dilution. C-D. Shannon diversity (C) and
+richness (D) in stool communities over time. B-D. Each symbol represents
+a stool sample from an individual mouse with the lines representing the
+median value for each treatment group. The transparancy of the symbol
+corresponds to the day post-infection. Asterisks indicate time points
+with significant differences between groups by a Kruskall-Wallis test
+with a Benjamini-Hochberg correction for testing multiple times points.
+Colored rectangles indicates the 1-day PEG treatment period for
+applicable groups. The data presented are from a total of 3 separate
 experiments. \newpage
@@
 post \emph{C. difficile} challenge mice that also receive an FMT only
-some bacterial genera were restored.} A. PCoA of all groups + FMT
-plotted over time. Opacity of points reflects time point. B. Relative
-abundance of genera significantly different over multiple time points
-post FMT or PBS treatment, plotted over time. Limit of detection is
-.1\%. C. Relative abundances of top 6 significant genera ranked by
-number of days significant, plotted over time. Limit of detection is
-.1\%.
+some bacterial genera were restored.} A. PCoA of Bray-Curtis distances
+from stool samples collected over time as well as the FMT solution that
+was administered to one treatment group. Each circle represents an
+individual sample, the transparency of the circle corresponds to day
+post-infection as shown in Fig. 6C-D. B. Median relative abundances of 2
+genera that were significantly different over multiple time points in
+mice that were administered either FMT or PBS solution via gavage C.
+Median relative abundances of the top 6 out of 24 genera that were
+significant over multiple timepoints, plotted over time (Data Set S1,
+Sheet X). For B-C, colored rectangles indicates the 1-day PEG treatment
+period for applicable groups. Gray horizontal lines represent the limit
+of detection.
 
@@
 features associated with prolonged \emph{C. difficile} colonization in
-PEG treated mice.} A. \newpage
+PEG treated mice.} A. Top ten bacteria that contributed to the random
+forest model trained on five day post-infection community relative
+abundance data to predict wehther mice would still be colonized with
+\emph{C. difficile} 10 days post-infection. The median (point) and
+interquantile range (lines) change in AUROC when the bacteria is left
+out of the model is shown. B. The median relative abundances at 5 days
+post-infection of the top ten bacteria that contributed to the random
+forest classification model. Color indicates whether the mice were still
+colonized with \emph{C. difficile} 10 days post-infection and the black
+horizontal line represents the median relative abundance. Each symbol
+represents a stool sample from an individual mouse and the shape of the
+symbol indicates whether the PEG-treated mice recieved a 5-day (Fig.
+1-3), 1-day (Fig. 4) or post-CDI PEG (Fig. 5-6) treatment. C. The median
+relative abundances of the 5 genera with greater than 1\% median
+relative abundance in the stool community over time. For B-C, the gray
+horizontal line represents the limit of detection. \newpage
 
 \includegraphics{figure_8.pdf} \textbf{Figure 8. Schematic summarizing
-findings.} A. \newpage
+findings.} A. The gut microbiota of our C57Bl/6 mice is restant to
+\emph{C. difficle} but treatment with either the antibiotic,
+clindamycin, or the osmotic laxative, PEG 3350 renders the mice
+susceptible to \emph{C. difficile} colonization. Recovery of
+colonization resistance in clindamycin-treated mice is relatively
+straightforward and the mice clear \emph{C.difficile} within 10 days
+post-infection. However, for mice that recieved a 5-day PEG treatment or
+a 1-day PEG treatment after \emph{C. difficile} challenge recovery of
+colonization resistance is more uncertain because mice were still
+colonized with \emph{C. difficile} 30 days post-infection in the case of
+several PEG treatments. We found increased \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} were associated with recovery of colonization
+resistance, while increased \emph{Enterobacteriaceae} and
+\emph{Bacteroides} were associated with prolonged \emph{C. difficle}
+colonization. B.Considering that most hospitals recommend not performing
+\emph{C. difficile} testing on patients taking laxatives and laxatives
+are used when administering fecal microbiota transplants via colonoscopy
+to patients with recurrent CDIs. further studies are needed to evaluate
+if laxatives impact human microbiota colonization resistance. Further
+studies are needed to understand the impact of osmotic laxatives on
+\emph{C. difficile} colonization resistance and clearance in human
+patients. \newpage
 
-\includegraphics{figure_S1.pdf} \textbf{Figure S1. 5-day PEG treatment
-plus 10-day recovery mice microbiota dynamics post-infection.} A.
-\newpage
+\includegraphics{figure_S1.pdf} \textbf{Figure S1. Microbiota dynamics
+post-infection in the 5-day PEG treatment plus 10-day recovery mice.} A.
+\emph{C. difficile} CFU/g over time in the stool samples collected from
+5-day PEG treatment plus 10-day recovery mice. Same data presented in
+Fig. 1C, but the data for the other 3 treatment groups have been
+removed. B. Median relative abundances of 8 bacterial genera from day 0
+post-infection onward from the 10-day recovery mice. We analyzed samples
+from day 0 and day 8 post-infection, which represented the the time
+points where mice were challenged with \emph{C. difficile} and when the
+median relative \emph{C. difficile} CFU stabilized for the group using
+the paired Wilcoxan signed-rank test, but no genera were signifcant
+after Benjamini-Hochberg correction. \newpage
 
-\includegraphics{figure_S2.pdf} \textbf{Figure S2. Impact of PEG
-treatment on the mucosal microbiota is greatest 6 days post-infection}\\
+\includegraphics{figure_S2.pdf} \textbf{Figure S2. PEG treatment still
+has a large impact on the mucosal microbiota 6 days post-infection} A.
+The\\
 \newpage
@@
 \includegraphics{figure_S4.pdf} \textbf{Figure S4. Specific OTUs
-associated with clearance that are mostly absent in mice with prolonged
-\emph{C. difficile} colonization. Ex. \emph{Muribaculum intestinale}.}
-A. \newpage
+associated with clearance by 10 days post-infection that are mostly
+absent in mice with prolonged \emph{C. difficile} colonization. Ex.
+\emph{Muribaculum intestinale}.} A. \newpage
 

### 2021-04-30 code/5_days_PEG_16S.R
Fix typo subsetting distal colon tissues and update lineplots

@@ -350,7 +350,7 @@ genus_stools <- subset_stool(genus_subset)
 genus_tissues <- subset_tissue(genus_subset)
 genus_cecum <- subset_cecum(genus_subset)
 genus_proximal_colon <- subset_proximal_colon(genus_subset)
-genus_distal_colon <- subset_proximal_colon(genus_subset)
+genus_distal_colon <- subset_distal_colon(genus_subset)
 #The above subsets exclude mock challenged mice (group = WMN or CN)
 #Also create dataframes of diversity data that includes mock challenged mice (WMN and C), separated into stool and tissue samples
 genus_mock_stools <- subset_stool(add_mocks(genus_subset, agg_genus_data))
@@ -497,7 +497,7 @@ shared_sig_distal_colon_genus #0 genera, because 0 genera were different at day 
 
 #Examine genera that were shared between tissue compartments on day 6
 sig_shared_tissues <- intersect_all(sig_genus_cecum_day6, sig_genus_proximal_colon_day6, sig_genus_distal_colon_day6)
-#16 genera total including Peptostrep.
+#10 genera total including Peptostrep.
 
 #No genera were significant for proximal and distal colon compartments on day 30
 
@@ -638,7 +638,7 @@ line_plot_tissue_comp_time <- function(sample_df, genera, tissue_type, timepoint
     labs(title=NULL,
          x="Days Post-Infection",
          y="Relative abundance (%)")+
-    facet_wrap(~genus, nrow = 4, labeller = label_wrap_gen(width = 10))+
+    facet_wrap(~genus, nrow = 2, labeller = label_wrap_gen(width = 10))+
     theme_classic()+
     theme(strip.background = element_blank(), #get rid of box around facet_wrap labels
           plot.title = element_markdown(hjust = 0.5), #Have only the genera names show up as italics
@@ -656,13 +656,13 @@ top_d6_sig_stools <- kw_genus_stools %>%
 shared_d6_genera <- intersect_all(sig_shared_tissues, top_d6_sig_genus)
 #14 genera shared between both compartments (Drops Peptostreptococcaceae and Firmicutes Unclassified)
 
-#16 genera different in all compartments on day 6
+#10 genera different in all compartments on day 6
[…]

### 2021-04-30 submission/manuscript.Rmd
Add code for referring to the number of significant genera shared between tissue compartments

@@
 
-#To Do: Insert code for number of genera different in all 3 types of tissue samples
+#Function to find which significant otus/genera/families are shared
+intersect_all <- function(a,b,...){
+  Reduce(intersect, list(a,b,...))
+}
+#Number of genera different in all 3 types of tissue samples
+kw_genus_cecum <- read_tsv(here("data/process/5_days_PEG_genus_group_cecum.tsv")) %>% 
+  filter(p.value.adj < 0.05) %>% pull(genus)
+kw_genus_proximal_colon <- read_tsv(here("data/process/5_days_PEG_genus_group_proximal_colon.tsv")) %>% 
+  filter(p.value.adj < 0.05) %>% pull(genus)
+kw_genus_distal_colon <-  read_tsv(here("data/process/5_days_PEG_genus_group_distal_colon.tsv")) %>% 
+  filter(p.value.adj < 0.05) %>% pull(genus)
+# of shared genera with significantly different relative abundances between groups
+shared_tissue_genera <- length(intersect_all(kw_genus_cecum, kw_genus_proximal_colon, kw_genus_distal_colon))
 
@@
 3B, Data Set S1, Sheet X).
-We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with 16 genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
+We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with `r shared_tissue_genera` genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
 S3A, Data Set S1, Sheet X).

### 2021-04-30 code/5_days_PEG_16S.R
Fix cut off genus names

@@ -643,7 +643,7 @@ line_plot_tissue_comp_time <- function(sample_df, genera, tissue_type, timepoint
     theme(strip.background = element_blank(), #get rid of box around facet_wrap labels
           plot.title = element_markdown(hjust = 0.5), #Have only the genera names show up as italics
           text = element_text(size = 16),
-          strip.text = element_text(face = "italic", size = 12),
+          strip.text = element_text(face = "italic", size = 9.5),
           legend.position = "None")
 }
 #Genera that were significant in the stool samples on day 6

### 2021-04-30 code/figure_S2.R

@@ -1,6 +1,6 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_cecum_d6_16_genera.png")
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_cecum_d6_10_genera.png")
 b <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_peptostreptococcacea.png")
 b <- plot_grid(b, NULL, labels = NULL)
 

### 2021-04-30 code/5_days_PEG_16S.R
Update 10-day recovery plots to indicate individual mice

@@ -864,6 +864,42 @@ lp_stool_WMR <- genus_stools %>%
           legend.position = "None")
 save_plot(filename = "results/figures/5_days_PEG_genus_lineplot_stools_WMR.png", lp_stool_WMR, base_height = 4, base_width = 8.5)
 
+#Create lineplots of 10 genera
+#Alternative version with lines for individual mice 
+#Same scale used in code/5_days_PEG_16S.R
+color_mice <- color_mice <- c("5_M5",  "6_M5",  "7_M5",  "8_M5",  "9_M5",  "10_M5", "7_M6",  "9_M6",  "10_M6", "11_M6", "12_M6")
+color_mice_values <- c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", 
+                       "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a", "#9B870C")
+color_mice_labels <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10 , 11)
+
+lp_stool_WMR_indiv <- genus_stools %>% 
+  filter(group == "WMR") %>% #Only plot WMR mice
+  filter(genus %in% WMR_genus) %>% #Select only genera of interest
+  mutate(genus = fct_relevel(genus, WMR_genus)) %>% #Reorder genera to match order of genera of interest
+  filter(day %in% lp_stool_days) %>% #Select only timepoints of interest
+  mutate(day = as.integer(day)) %>% 
+  mutate(agg_rel_abund = agg_rel_abund + 1/2000) %>% # 2,000 is 2 times the subsampling parameter of 1000 
+  ggplot()+
+  geom_line(aes(x = day, y=agg_rel_abund, color=unique_mouse_id, group =unique_mouse_id), linetype = "solid", alpha = 0.8)+
+  scale_colour_manual(name="Mouse",
+                      values=color_mice_values,
+                      breaks=color_mice,
+                      labels=color_mice_labels)+
+  scale_x_continuous(limits = c(-1.5,31), breaks = c(0, 5, 10, 15, 20, 25, 30), labels = c(0, 5, 10, 15, 20, 25, 30))+
+  scale_y_continuous(trans = "log10", limits = c(1/10900, 1), breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100))+
+  geom_hline(yintercept=1/1000, color="gray")+ #Represents limit of detection
+  labs(title=NULL,
[…]

### 2021-04-30 code/5_days_PEG_cfu_weight.R

@@ -209,6 +209,41 @@ wmr_cfu_plot <-  ggplot(NULL) +
           panel.grid.minor.x = element_line(size = 0.4, color = "grey"))#Add gray lines to clearly separate symbols by days)
 save_plot(filename = "results/figures/5_days_PEG_cfu_WMR.png", wmr_cfu_plot, base_height = 4, base_width = 8.5, base_aspect_ratio = 2)
 
+#Make a second version where each unique mouse = a different colored line
+#Plot of just the cfu data for the WMR group (5-day PEG + 10-day recovery)
+wmr_cfu_mice <- cfudata %>% 
+  filter(group == "WMR") %>% 
+  filter(day %in% c("-1", "0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "15", "20", "25", "30")) %>% 
+  filter(!duplicated(unique_mouse_id)) %>% #Remove duplicate mouse ids
+  mutate(unique_mouse_id = factor(unique_mouse_id, levels = unique(as.factor(unique_mouse_id)))) %>% 
+  pull(unique_mouse_id)
+color_mice <- c("5_M5",  "6_M5",  "7_M5",  "8_M5",  "9_M5",  "10_M5", "7_M6",  "9_M6",  "10_M6", "11_M6", "12_M6")
+color_mice_values <- c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", 
+                       "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a", "#9B870C")
+color_mice_labels <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10 , 11)
+
+wmr_cfu_plot_indiv <-  wmr_cfu %>% 
+  ggplot()+
+  geom_line(mapping = aes(x = day, y = avg_cfu, group = unique_mouse_id, color = unique_mouse_id), alpha = 0.8, size = 1.5) +
+  scale_colour_manual(name="Mouse",
+                      values=color_mice_values,
+                      breaks=color_mice,
+                      labels=color_mice_labels)+
+  labs(x = "Days Post-Infection", y = "CFU/g Feces") +
+  scale_y_log10(labels=fancy_scientific, breaks = c(10, 100, 10^3, 10^4, 10^5, 10^6, 10^7, 10^8, 10^9, 10^10, 10^11, 10^12))+ #Scientific notation labels for y-axis
[…]

### 2021-04-30 code/figure_S1.R

@@ -1,8 +1,8 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
-a <- ggdraw() + draw_image("results/figures/5_days_PEG_cfu_WMR.png")
-b <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_stools_WMR.png")
+a <- ggdraw() + draw_image("results/figures/5_days_PEG_cfu_WMR_indiv.png")
+b <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_stools_WMR_indiv.png")
 
-plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S1.pdf", width=6.875, height=7)+
-  ggsave("submission/figure_S1.pdf", width=6.875, height=7)
+plot_grid(a, b, labels = c("A", "B"), label_size = 12, ncol=1, rel_heights = c(1, 2))+
+  ggsave("results/figures/figure_S1.pdf", width=6.875, height=9)+
+  ggsave("submission/figure_S1.pdf", width=6.875, height=9)

### 2021-04-30 code/blast_otus.R
BLAST lachnospiraceae and and porphyromonadaceae otus of interest and plot results

@@ -375,3 +375,107 @@ bacteriodes_1_and_2_otus <- otu_1_and_2_blast_results %>%
         legend.position = "none", #Remove legend
         axis.text.x = element_blank())
 save_plot("results/figures/otus_1_and_2_bacteriodes_blast_results.png", bacteriodes_1_and_2_otus, base_height =5, base_width = 6)
+
+#Investigate Lachnospiraceae and Porphyromonadaceae OTUs
+porphyr_otu_nums <- c("Otu0006", "Otu0007", "Otu0008", "Otu0010", "Otu0021", "Otu0025", "Otu0048", "Otu0122")
+lachno_otu_nums <- c("Otu0029", "Otu0033", "Otu0062", "Otu0073")
+
+porphyr_otus <- taxonomy %>%
+  filter(OTU %in% porphyr_otu_nums)
+
+lachno_otus <- taxonomy %>%
+  filter(OTU %in% lachno_otu_nums)
+
+porphyr_seq_all <- map_df(porphyr_otu_nums, function(porphyr_otu_nums){
+  porphyr_seq <- otu_seqs %>% 
+    filter(str_detect(name, porphyr_otu_nums))
+})
+
+lachno_seq_all <- map_df(lachno_otu_nums, function(lachno_otu_nums){
+  lachno_seq <- otu_seqs %>% 
+    filter(str_detect(name, lachno_otu_nums))
+})
+
+porphyr_seq_all %>% pull(sequence)
+lachno_seq_all %>% pull(sequence)
+
+#Load data from porphyromonadaceae Otus
+otu_0010_blast_results <- read_csv("data/process/blast/OTU_0010_HitTable.csv") %>%
+  mutate(otu = "Otu0010") %>% head(4)
+otu_0025_blast_results <- read_csv("data/process/blast/OTU_0025_HitTable.csv" ) %>%
+  mutate(otu = "Otu0025") %>% head(4)
+otu_0122_blast_results <- read_csv("data/process/blast/OTU_0122_HitTable.csv") %>%
+  mutate(otu = "Otu0122") %>% head(4)
+otu_0048_blast_results <- read_csv("data/process/blast/OTU_0048_HitTable.csv") %>%
+  mutate(otu = "Otu0048") %>% head(4)
+otu_8_blast_results <- read_csv("data/process/blast/OTU_0008_HitTable.csv") %>%
+  mutate(otu = "Otu0008") %>% head(4)
[…]

### 2021-04-30 code/5_days_PEG_16S.R
Note mouse that died

@@ -842,7 +842,8 @@ color_mice <- color_mice <- c("5_M5",  "6_M5",  "7_M5",  "8_M5",  "9_M5",  "10_M
 color_mice_values <- c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", 
                        "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a", "#9B870C")
 color_mice_labels <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10 , 11)
-
+#Note: 8_M6 and 11_M6 (not numbered and #10, respectively) died during the experiment. 8_M6 died before challenge
+# 11_M6 #10 was found dead by us on 6/3/19 which corresponded to 6 days post-infection
 lp_stool_WMR_indiv <- genus_stools %>% 
   filter(group == "WMR") %>% #Only plot WMR mice
   filter(genus %in% WMR_genus) %>% #Select only genera of interest

### 2021-04-30 submission/manuscript.Rmd
Edit results section

@@
 We used clindamycin-treated mice as a control because they consistently clear *C.
-difficile* within 10 days post-infection (dpi).
+difficile* within 10 days post-challenge (dpc).
 To examine how PEG treatment impacts clearance, we administered PEG for 1 day to clindamycin-treated, *C.
-difficile*-challenged mice either immediately following challenge or 3 dpi.
+difficile*-challenged mice either immediately following challenge or 3 dpc.
 We collected longitudinal stool samples to examine *C.
 difficile* levels in the stool via anaerobic culture and profiled the microbiota by 16S rRNA sequencing.
-PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpi.
+PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpc.
 Additionally, 5-day PEG treated mice remained susceptible to CDI 10-days post treatment.
 In contrast, 1-day PEG treated mice were transiently colonized, clearing *C.
-difficile* within 7 dpi.
+difficile* within 7 dpc.
 Although 5-day PEG-treated mice exhibited prolonged *C.
@@
 Additionally, administering PEG to mice after *C.
-difficile* challenge prolonged colonization up to 30 dpi in mice that received PEG immediately after challenge and 15 dpi in mice that received PEG 3 dpi.
+difficile* challenge prolonged colonization up to 30 dpc in mice that received PEG immediately after challenge and 15 dpc in mice that received PEG 3 dpc.
 When we examined microbiota composition across our different treatment groups, we found decreased richness in the PEG-treated mice that exhibited prolonged *C.
@@
 Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020].
-Stool consistency often serves as an approximation of intestinal motility.
+Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015].
 Our group has shown that when *C.
@@
 difficile* [@Olson2013; @Diebel2014].
-However, whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C.
+Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C.
 difficile* after challenge is unclear.
@@
 
-In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C.
+Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C.
 difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020].
@@
 difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C.
-difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs.
-Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in more severe CDI and sustained *C.
-difficile* colonization, and if PEG treatment post-CDI can promote *C.
+difficile* clearance.
+Addressing these questions will better inform how we think about laxatives in the context of CDIs.
+Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C.
+difficile* colonization and severe CDI and  than mice treated with clindamycin, and if PEG treatment after colonization can promote *C.
 difficile* clearance.
@@
 
-**5-day laxative treatment leads to prolonged *C.
-difficile* colonization in mice.** We compared mice treated with the osmotic laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C.
-difficile*, with mice typically clearing *C.
-difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020].
-All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig.
+**5-day laxative treatment led to prolonged *C.
+difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C.
+difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C.
+difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C.
+difficile*.
+We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C.
+difficile* colonization, with mice typically clearing *C.
+difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020].
+All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, 1.
+received no additional treatment, 2.
+was also treated with clindamycin, and 3.
+was allowed to recover for 10 days prior to challenge (Fig.
 1A).
-PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice recovered most of the weight five days after treatment (Fig.
+PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice relative to their baseline weights, with the greatest change in weight observed on the fifth day of PEG treatment.
+The mice recovered most of the lost weight by five days after treatment (Fig.
 1B).
-After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C.
+After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C.
 difficile* 630 spores (Fig.
@@
 However, PEG-treated mice remained colonized with *C.
-difficile* at a high level through thirty days post-infection (Fig.
+difficile* at a high level through thirty days post-challenge (Fig.
 1C).
 In contrast, the clindamycin-treated mice cleared *C.
-difficile* within ten days post-infection.
-Surprisingly, PEG-treated mice were still susceptible to *C.
-difficile* infection after 10-days of recovery from treatment, although *C.
-difficile* was not detectable in most of the group in the initial five days post-infection (Fig.
+difficile* within ten days post-challenge (Fig.
 1C).
-From 8 days post-infection onward, the median *C.
-difficile* CFU stabilized for the 10-day recovery group of PEG-treated mice and remained high through 30 days post-infection (Fig.
+Therefore, PEG treatment led to sustained colonization in contrast to clindamycin mice that naturally cleared *C.
+difficile* within ten days post-challenge.
+
+Notably, we also found PEG-treated mice were still susceptible to *C.
+difficile* colonization after a 10-day recovery period, although *C.
+difficile* was not detectable in most of the group in the initial five days post-challenge (Fig.
+1C, S1A).
+One mouse was found dead on the 6th day post-challenge, presumable due to *C.
+difficile* as the bacteria became detecable in that mouse on the 4th day post-challenge (Fig.
+S1A, mouse 10).
+From 8 days post-challenge onward, the density of *C.
+difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig.
 1C).
 Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C.
-difficile* colonization and PEG-treated mice remained susceptible up to ten days post-treatment.
+difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.
 
@@
 ```
-**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C.
+**5-day laxative treatment differentially disrupted the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C.
 difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota.
@@
 2A, Data Set S1, Sheet X).
-Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015].
-We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage.
-Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
-Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C.
+Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to coprophagy [@Nguyen2015].
+We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups.
+Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
+None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C.
 difficile* colonization in PEG-treated mice (Fig.
@@
 
-Next, we examined  alpha diversity by looking at Shannon diversity over time.
-Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig.
+Since none of the communities completely recovered in the follow-up period after treatments, we next profiled community diversity and composition.
+We examined the alpha diversity dynamics by calculating the communities' Shannon diversity.
+Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-challenge (Fig.
 2C).
-We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment.
+We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment.
 We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X).
-The majority of the bacterial relative abundances decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment, as *Bacteroides* actually decreased in clindamycin treated mice (Fig.
+The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased.
+The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig.
 2D).
-Finally, we examined the bacteria that differed across treatment groups over multiple time points.
-We found `r f_d_group_multi_day` genera were different over multiple time points out of the `r f_d_group_total_genera` genera that were different between treatment groups (Fig.
+Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points.
+Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig.
 2E, Data Set S1, Sheet X).
@@
 difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge.
-We decided to examine the bacteria that changed during the post-infection period when the group median *C.
-difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 days post-infection (Fig.
+We decided to examine if there were genera that changed during the post-challenge period when the group median *C.
+difficile* shifted from undetectable at 1 day post-challenge to detectable in the stool samples with the density stabilizing around 8 days post-challenge (Fig.
 S1A).
-Interestingly, we found *Erysipelotrichaceae*, *Enterobacteriaceae*, and *Akkermansia* were changing during the time period when *C.
-difficile* was becoming detectable in the stools (Fig.
-S1B), although none of the bacteria were significant after multiple hypothesis correction (Data Set S1, Sheet X).
+We found no bacteria were significantly different over the two time periods after multiple hypothesis correction (Data Set S1, Sheet X).
+However, there was also wide variation between individual mice regarding when *C.
+difficile* became detectable (Fig.
+S1A) as well as the relative abundances of bacterial genera in the communities (Fig.
+S1B).
+For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period and this corresponded to mouse 10, which died on the sixth day post-challenge and mouse 11, where *C.
+difficile* was present at a high density from the 4th day post-challenge onward (Fig.
+S1B).
 While we did not identify a clear signal to explain the delayed appearance of *C.
@@
 
+tissue_genera_max_p <- read_tsv(here("data/process/5_days_PEG_genus_group_cecum.tsv")) %>% 
+  add_row(read_tsv(here("data/process/5_days_PEG_genus_group_proximal_colon.tsv"))) %>% 
+  add_row(read_tsv(here("data/process/5_days_PEG_genus_group_distal_colon.tsv"))) %>% 
+  filter(p.value.adj < 0.05) %>% 
+  #Limit to genus found in all 3 types of tissues
+  filter(genus %in% intersect_all(kw_genus_cecum, kw_genus_proximal_colon, kw_genus_distal_colon)) %>%
+  select(p.value.adj) %>% max()
+
 ```
-**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
+**5-day laxative treatment did not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
 To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon.
-Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treatment (Fig.
+Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig.
 3A).
-Although alpha diversity continued to increase over time based on the communities from PEG-treated mice collected at 20 and 30 days post-infection (Fig.
+Alpha diversity continued to increase over time with the PEG-treated mice collected at 20 and 30 days post-challenge (Fig.
 3A, Data Set S1, Sheet X).
@@
 3B, Data Set S1, Sheet X).
-We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with `r shared_tissue_genera` genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
-S3A, Data Set S1, Sheet X).
-However, at 30 days post-infection only *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig.
+We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
+S2A, Data Set S1, Sheet X).
+Interestingly, *Peptostreptococcaceae* (the family with a sequence that matches *C.
+difficile*) was one of the genera that had a significant difference in relative abundance between treatment groups at 6 days post-challenge.
+This population was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig.
+S2B).
+By 30 days post-challenge, only the relative abundances of *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig.
 3C, Fig.
 2E, Data Set S1, Sheet X).
-Interestingly, one of the bacteria that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera that contains a sequence that matches *C.
-difficile*).
-*Peptostreptococcaceae* was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig.
-S2B).
 Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C.
@@
 
-Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C.
-difficile* at 4 days post-infection (Fig.
+Next, we examined the severity of *C.
+difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C.
+difficile* at 4 days post-challenge (Fig.
 3D), the time point typically examined in *C.
 difficile* 630 challenged mice [@Dieterle2020].
-We also looked at 6 days post-infection because that was when we started to see a large difference in *C.
-difficile* colonization levels between PEG- and clindamycin-treated mice (Fig.
+We also looked at 6 days post-challenge because that was when there was a large difference in *C.
+difficile* density between PEG- and clindamycin-treated mice (Fig.
 1C).
-Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is twelve (Fig.
+Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5-2.5 out of 12, Fig.
 3E).
 Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C.
-difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.
+difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.
 
 ***C.
-difficile* challenge does not have a synergistic disruptive effect on the microbiota of PEG-treated mice** Because *C.
+difficile* challenge did not have a synergistic disruptive effect on the microbiota of PEG-treated mice.** Because *C.
 difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice.
@@
 difficile* challenged mice (Fig.
-S2A-C).
+S3A-C).
 Similarly, there was no difference between the tissues of mock and *C.
 difficile* challenged mice (Fig.
-S2D-F).
+S3D-F).
 Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C.
@@
   filter(p.value.adj < 0.05)
-#Genera that were still different at 7 days post-infection  
+#Genera that were still different at 7 days post-challenge  
 pt_d7 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_BtoD7.tsv")) %>%
@@
 ```
-**1-day laxative treatment results in transient *C.
-difficile* colonization and minor microbiota disruption** Next, we decided to examine how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C.
+**1-day laxative treatment resulted in transient *C.
+difficile* colonization and minor microbiota disruption.** Next, we examined how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C.
 difficile*.
@@
 In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C.
-difficile* by 7 days post-infection (Fig.
+difficile* by 7 days post-challenge (Fig.
 3B).
-The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering close to baseline by 7 days post-infection (Fig.
+The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering by 7 days post-challenge (Fig.
 3C-D).
-We found `r recovered_genera` bacteria were impacted by treatment, but recovered close to baseline levels by 7 days post-infection including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig.
+We found the relative abundances of `r recovered_genera` genera were impacted by treatment, but recovered close to baseline levels by 7 days post-challenge including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig.
 3E, Data Set S1, Sheet X).
-These findings suggest the duration of treatment matters when considering the impact of the laxatives, with shorter treatments resulting in a transient loss of *C.
+These findings suggest the duration of the PEG treatment was relevant, with shorter treatments resulting in a transient loss of *C.
 difficile* colonization resistance.
@@
 ```
-**Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered** Since 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
+**Post-CDI laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
 difficile* spores from the intestine.
-We were also interested in exploring whether PEG might help with engraftment in the context of FMTs.
 To examine the impact of PEG treatment on *C.
-difficile* clearance and FMT treatment, we treated 4 groups of mice with clindamycin and then challenged all mice with *C.
+difficile* clearance, we treated 4 groups of mice with clindamycin and then challenged all mice with *C.
 difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig.
@@
 5B).
-Next we examined how post-CDI PEG treatment impacted the microbiota.
+Next we examined how post-CDI PEG treatment impacted microbiota diversity.
 Alpha diversity was lower in the PEG-treated mice with the exception of the PEG-treated mice that were administered an FMT (Fig 5C-D).
+
+We were also interested in exploring whether PEG might help with engraftment in the context of FMTs.
 The FMT appeared to partially restore Shannon diversity but not richness (Fig.
 5C-D).
-Similarly, we saw some overlap between the communities of mice that received FMT and the mice treated with only clindamycin after 5 days post-infection (Fig.
+Similarly, we saw some overlap between the communities of mice that received FMT and the mice treated with only clindamycin after 5 days post-challenge (Fig.
 6A).
@@
 difficile* colonization in the FMT treated mice.
-However, only *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice received either an FMT or PBS gavage (Fig.
+However, only the relative abundances of *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice received either an FMT or PBS gavage (Fig.
 6B), suggesting the FMT only restored a couple of genera.
-Overall, we found there were `r post_group_multi_day` bacteria that differed over multiple time points between treatment groups including increased *Akkermansia* and decreased *Ruminococcaceae*, *Clostridiales*, *Lachnospiraceae*, and *Oscillibacter* in mice that received PEG after *C.
-difficile* challenge (Fig.
+Overall, we found the relative abundances of `r post_group_multi_day` genera were different between treatment groups over multiple timepoints.
+For example, the relative abundance of *Akkermansia* was increased and the relative abundances of *Ruminococcaceae*, *Clostridiales*, *Lachnospiraceae*, and *Oscillibacter* were decreased in mice that received PEG after *C.
+difficile* challenge relative to clindamycin treated mice (Fig.
 6C).
 In sum, administering PEG actually prolonged *C.
-difficile* colonization, including in mice that received an FMT, which only restored a couple of bacterial genera.
+difficile* colonization, including in mice that received an FMT, which only restored 2 bacterial genera.
 
@@
 ```
-**Five-day post-infection community data can predict which mice that will have prolonged *C.
-difficile* colonization** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C.
+**Five-day post-challenge community data can predict which mice that will have prolonged *C.
+difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C.
 difficile* colonization.
-We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with input bacterial community data from 5 days post-infection to predict whether the mice were still colonized with *C.
-difficile* 10 days post-infection.
-We chose 5 days post-infection because that was the earliest time point where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after *C.
-difficile* challenge and then administered an FMT or PBS gavage.
+We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with bacterial community data from 5 days post-challenge to predict whether the mice were still colonized with *C.
+difficile* 10 days post-challenge.
+We chose 5 days post-challenge because that was the earliest time point where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after challenge.
 The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C.
@@
 We selected the top 10 bacteria contributing to our models performance (Fig.
-7A) and examined their relative abundance at 5 days post-infection, the time point used to predict *C.
+7A) and examined their relative abundance at 5 days post-challenge, the time point used to predict *C.
 difficile* colonization status on day 10 (Fig.
@@
 Next, we focused on the 5 genera that had a greater than 1 % relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time.
-We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
+We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, the relative abundance of *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
 difficile* (Fig.
@@
 
-Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of S24-7, also known as *Muribaculum intestinale* [@Tropini2018].
-We decided to check our *Porphyromonadaceae* operational taxonomic units (OTUs) because *Muribaculum intestinale* is classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020] and *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
+Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of *Muribaculum intestinale* [@Tropini2018]., which is classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020].
+*Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
 difficile* colonization.
-We identified 4 OTUs that had at least 92% identity to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C.
-difficile* at 10 days post-infection.
+We identified 4 OTUs that had 92-96% identity to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C.
+difficile* at 10 days post-challenge.
 While all of the OTUs, were decreased by PEG and clindamycin treatment, there was some recovery in the mice that cleared (Fig.
 S4A).
-We also examined other *Porphyromonadaceae* and *Lachnospiraceae* OTUs since these were the 2 genera that were important to our classification model and contained multiple OTUs that were different at 5 days post-infection between mice that either cleared or remained colonized with *C.
-difficile* by 10 days post-infection (Data Set S1, sheet X).
+We also examined other *Porphyromonadaceae* and *Lachnospiraceae* OTUs since these were the 2 genera that were important to our classification model and contained multiple OTUs that were different at 5 days post-challenge between mice that either cleared or remained colonized with *C.
+difficile* by 10 days post-challenge (Data Set S1, sheet X).
 While individual *Porphyromonadaceae* and *Lachnospiraceae* OTUs tended to be more abundant in the mice that clear *C.
@@
   filter(sample_type == "stool") %>% #Select only stool samples
-  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 8, 9, 10, 15, 20, 25, 30)) %>%  #select time points on plot
+  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20, 25, 30)) %>%  #select time points on plot
   filter(!is.na(avg_cfu)) %>% #Remove rows with NA values
@@
 
-#Figure out which experiments eneded at 4, 6, or 30 days post-infection
+#Figure out which experiments eneded at 4, 6, or 30 days post-challenge
 five_day_mice_samples <- five_day_mice %>% 
@@
   tally()
-#Figure out number of stool samples from 4,6, 10, 20 and 30 days post-infection
+#Figure out number of stool samples from 4,6, 10, 20 and 30 days post-challenge
 five_day_mice_samples <- five_day_mice %>% 
@@
 difficile* 630 spores.
-A subset of mice were euthanized on either 4 or 6 days post-infection and tissues were collected for histopathology analysis, the remaining mice were followed through 20 or 30 days post-infection.
+A subset of mice were euthanized on either 4 or 6 days post-challenge and tissues were collected for histopathology analysis, the remaining mice were followed through 20 or 30 days post-challenge.
 B.
@@
 Principal Coordinate analysis (PCoA) of Bray-Curtis distances from stool samples collected throughout the experiment.
-For A and C, each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-infection.
+For A and C, each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-challenge.
 B.
-Bray-Curtis distances of stool samples collected on either day 10 or 30 post-infection relative to the baseline sample collected for each mouse (before any drug treatments were administered).
+Bray-Curtis distances of stool samples collected on either day 10 or 30 post-challenge relative to the baseline sample collected for each mouse (before any drug treatments were administered).
 C.
@@
 Circles, triangles, and squares indicate cecum, proximal colon, and distal colon communities, respectively.
-For A-B, transparency of the symbol corresponds to the day post-infection that the sample was collected.
+For A-B, transparency of the symbol corresponds to the day post-challenge that the sample was collected.
 C.
@@
 Each category is given a score ranging from 0-4, thus the maximum possible summary score is 12.
-The tissue for histology was collected at either 4 (D) or 6 (E) days post-infection with the exception that one set of 5-day PEG treated mock-challenged mice were collected on day 0 post-infection (first set of open circles in D).
+The tissue for histology was collected at either 4 (D) or 6 (E) days post-challenge with the exception that one set of 5-day PEG treated mock-challenged mice were collected on day 0 post-challenge (first set of open circles in D).
 Histology data were analyzed with the Kruskal-Wallis test followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
@@
              group == "M1" & exp_num %in% c("M6")) %>%
+  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)) %>%  #select time points on plot
+  filter(!is.na(avg_cfu)) %>% #Remove rows with NA values
   group_by(day) %>%
@@
 Shannon diversity in stool communities over time with colored lines representing the median value for each treatment group.
-For B-D, each symbol represents a sample from an individual mouse and symbol transparency corresponds to the day post-infection that the sample was collected.
+For B-D, each symbol represents a sample from an individual mouse and symbol transparency corresponds to the day post-challenge that the sample was collected.
 E.
-Median relative abundances per treatment group for 6 out of the `r recovered_genera` genera that were affected by treatment, but recovered close to baseline levels by 7 days post-infection (Fig.
+Median relative abundances per treatment group for 6 out of the `r recovered_genera` genera that were affected by treatment, but recovered close to baseline levels by 7 days post-challenge (Fig.
 3E, Data Set S1, Sheet X).
@@
 3-days after challenge, mice received 1-day PEG treatment and then received either a fecal microbiota transplant (3) or PBS (4) solution by oral gavage.
-Mice were followed through 15-30 days post-infection (only the post-CDI 1-day PEG group was followed through 30 days post-infection).
+Mice were followed through 15-30 days post-challenge (only the post-CDI 1-day PEG group was followed through 30 days post-challenge).
 B.
@@
 Each symbol represents a stool sample from an individual mouse with the lines representing the median value for each treatment group.
-The transparancy of the symbol corresponds to the day post-infection.
+The transparancy of the symbol corresponds to the day post-challenge.
 Asterisks indicate time points with significant differences between groups by a Kruskall-Wallis test with a Benjamini-Hochberg correction for testing multiple times points.
@@
 PCoA of Bray-Curtis distances from stool samples collected over time as well as the FMT solution that was administered to one treatment group.
-Each circle represents an individual sample, the transparency of the circle corresponds to day post-infection as shown in Fig.
+Each circle represents an individual sample, the transparency of the circle corresponds to day post-challenge as shown in Fig.
 6C-D.
@@
 difficile* colonization in PEG treated mice.** A.
-Top ten bacteria that contributed to the random forest model trained on five day post-infection community relative abundance data to predict wehther mice would still be colonized with *C.
-difficile* 10 days post-infection.
+Top ten bacteria that contributed to the random forest model trained on five day post-challenge community relative abundance data to predict wehther mice would still be colonized with *C.
+difficile* 10 days post-challenge.
 The median (point) and interquantile range (lines) change in AUROC when the bacteria is left out of the model is shown.
 B.
-The median relative abundances at 5 days post-infection of the top ten bacteria that contributed to the random forest classification model.
+The median relative abundances at 5 days post-challenge of the top ten bacteria that contributed to the random forest classification model.
 Color indicates whether the mice were still colonized with *C.
-difficile* 10 days post-infection and the black horizontal line represents the median relative abundance.
+difficile* 10 days post-challenge and the black horizontal line represents the median relative abundance.
 Each symbol represents a stool sample from an individual mouse and the shape of the symbol indicates whether the PEG-treated mice recieved a 5-day (Fig.
@@
 difficile* colonization.
-Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-infection.
+Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-challenge.
 However, for mice that recieved a 5-day PEG treatment or a 1-day PEG treatment after *C.
 difficile* challenge recovery of colonization resistance is more uncertain because mice were still colonized with *C.
-difficile* 30 days post-infection in the case of several PEG treatments.
+difficile* 30 days post-challenge in the case of several PEG treatments.
 We found increased *Porphyromonadaceae* and *Lachnospiraceae* were associated with recovery of colonization resistance, while increased *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C.
@@
 **Figure S1.
-Microbiota dynamics post-infection in the 5-day PEG treatment plus 10-day recovery mice.** A.
+Microbiota dynamics post-challenge in the 5-day PEG treatment plus 10-day recovery mice.** A.
 *C.
@@
 B.
-Median relative abundances of 8 bacterial genera from day 0 post-infection onward from the 10-day recovery mice.
-We analyzed samples from day 0 and day 8 post-infection, which represented the the time points where mice were challenged with *C.
+Median relative abundances of 8 bacterial genera from day 0 post-challenge onward from the 10-day recovery mice.
+We analyzed samples from day 0 and day 8 post-challenge, which represented the the time points where mice were challenged with *C.
 difficile* and when the median relative *C.
@@
 \includegraphics{figure_S2.pdf}
-**Figure S2. PEG treatment still has a large impact on the mucosal microbiota 6 days post-infection** A. The   
+**Figure S2. PEG treatment still has a large impact on the mucosal microbiota 6 days post-challenge** A. The   
 \newpage
@@
 \includegraphics{figure_S4.pdf}
-**Figure S4. Specific OTUs associated with clearance by 10 days post-infection that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A.
+**Figure S4. Specific OTUs associated with clearance by 10 days post-challenge that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A.
 \newpage

### 2021-04-30 submission/manuscript.tex
Edit results section

@@
 spores. We used clindamycin-treated mice as a control because they
-consistently clear \emph{C. difficile} within 10 days post-infection
-(dpi). To examine how PEG treatment impacts clearance, we administered
+consistently clear \emph{C. difficile} within 10 days post-challenge
+(dpc). To examine how PEG treatment impacts clearance, we administered
 PEG for 1 day to clindamycin-treated, \emph{C. difficile}-challenged
-mice either immediately following challenge or 3 dpi. We collected
+mice either immediately following challenge or 3 dpc. We collected
 longitudinal stool samples to examine \emph{C. difficile} levels in the
@@
 susceptible to CDI and 5-day PEG-treated mice remain colonized for up to
-30 dpi. Additionally, 5-day PEG treated mice remained susceptible to CDI
+30 dpc. Additionally, 5-day PEG treated mice remained susceptible to CDI
 10-days post treatment. In contrast, 1-day PEG treated mice were
-transiently colonized, clearing \emph{C. difficile} within 7 dpi.
+transiently colonized, clearing \emph{C. difficile} within 7 dpc.
 Although 5-day PEG-treated mice exhibited prolonged \emph{C. difficile}
@@
 mice after \emph{C. difficile} challenge prolonged colonization up to 30
-dpi in mice that received PEG immediately after challenge and 15 dpi in
-mice that received PEG 3 dpi. When we examined microbiota composition
+dpc in mice that received PEG immediately after challenge and 15 dpc in
+mice that received PEG 3 dpc. When we examined microbiota composition
 across our different treatment groups, we found decreased richness in
@@
 microbiota composition and function (10, 11). Stool consistency often
-serves as an approximation of intestinal motility. Our group has shown
-that when \emph{C. difficile} negative controls are separated into two
-groups based on stool consistency, there are shared microbiota features
-such as lower alpha diversity in samples from CDI patients and control
-patients with diarrhea compared to control samples that were \emph{C.
-difficile} negative with non-diarrheal consistency (12). These results
-led to a hypothesis that bacterial communities from patients
+serves as an approximation of intestinal motility (10). Our group has
+shown that when \emph{C. difficile} negative controls are separated into
+two groups based on stool consistency, there are shared microbiota
+features such as lower alpha diversity in samples from CDI patients and
+control patients with diarrhea compared to control samples that were
+\emph{C. difficile} negative with non-diarrheal consistency (12). These
+results led to a hypothesis that bacterial communities from patients
 experiencing diarrhea are susceptible to developing CDIs.
@@
 difficile} infections by protecting intestinal epithelial cells from the
-toxins produced by \emph{C. difficile} (18, 19). However, whether
-laxative administration results in more severe CDIs in mice and how long
-mice remain colonized with \emph{C. difficile} after challenge is
-unclear.
+toxins produced by \emph{C. difficile} (18, 19). Whether laxative
+administration results in more severe CDIs in mice and how long mice
+remain colonized with \emph{C. difficile} after challenge is unclear.
 
@@
 
-In the past, our group has used C57BL6 mice to characterize how
-antibiotics including clindamycin disrupt the microbiota and influence
-\emph{C. difficile} susceptibility and clearance (24--26). Although, two
-groups have now shown PEG treatment alone renders mice susceptible to
-\emph{C. difficile} (15, 17), these studies have raised additional
-questions regarding the dynamics and severity of infection as well as
-the role of laxative treatment in \emph{C. difficile} clearance that
-should be addressed to better inform how we think about laxatives in the
-context of CDIs. Here, we used our C57BL/6 clindamycin model as a
-control group to characterize how long PEG-treated mice remain
-susceptible, whether PEG treatment results in more severe CDI and
-sustained \emph{C. difficile} colonization, and if PEG treatment
-post-CDI can promote \emph{C. difficile} clearance.
+Our group has used C57BL/6 mice to characterize how antibiotics disrupt
+the microbiota and influence \emph{C. difficile} susceptibility and
+clearance (24--26). Although, two groups have now shown PEG treatment
+alone renders mice susceptible to \emph{C. difficile} (15, 17), these
+studies have raised additional questions regarding the dynamics and
+severity of infection as well as the role of laxative treatment in
+\emph{C. difficile} clearance. Addressing these questions will better
+inform how we think about laxatives in the context of CDIs. Here, we
+characterized how long PEG-treated mice remain susceptible, whether PEG
+treatment results in more sustained \emph{C. difficile} colonization and
+severe CDI and than mice treated with clindamycin, and if PEG treatment
+after colonization can promote \emph{C. difficile} clearance.
 
@@
 
-\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
-colonization in mice.} We compared mice treated with the osmotic
-laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
-temporarily renders the mice susceptible to \emph{C. difficile}, with
+\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
+colonization in mice.} Building off of previous work that showed
+treating mice with the osmotic laxative, PEG 3350, rendered mice
+susceptible to \emph{C. difficile} colonization (15, 17), we decided to
+test how long \emph{C. difficile} colonization is sustained and how long
+PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
+three groups of mice treated with PEG 3350 to one group of mice treated
+with our standard 10 mg/kg clindamycin treatment, which temporarily
+renders the mice susceptible to \emph{C. difficile} colonization, with
 mice typically clearing \emph{C. difficile} within 10 days
-post-infection (9, 26). All PEG-treated mice were administered a 15\%
-PEG solution in the drinking water for 5-days, one group was also
-treated with clindamycin, and one group was allowed to recover for 10
-days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
-in all 3 groups of PEG-treated mice, with the greatest change in weight
-observed on the fifth day of PEG treatment and the mice recovered most
-of the weight five days after treatment (Fig. 1B). After either PEG,
-clindamycin, or PEG and clindamycin treatment all mice were challenged
-with 10\textsuperscript{5} \emph{C. difficile} 630 spores (Fig. 1A). All
-treatments rendered mice susceptible to \emph{C. difficile}
-colonization. However, PEG-treated mice remained colonized with \emph{C.
-difficile} at a high level through thirty days post-infection (Fig. 1C).
-In contrast, the clindamycin-treated mice cleared \emph{C. difficile}
-within ten days post-infection. Surprisingly, PEG-treated mice were
-still susceptible to \emph{C. difficile} infection after 10-days of
-recovery from treatment, although \emph{C. difficile} was not detectable
-in most of the group in the initial five days post-infection (Fig. 1C).
-From 8 days post-infection onward, the median \emph{C. difficile} CFU
-stabilized for the 10-day recovery group of PEG-treated mice and
-remained high through 30 days post-infection (Fig. 1C). Thus, osmotic
-laxative treatment alone was sufficient to render mice susceptible to
-prolonged \emph{C. difficile} colonization and PEG-treated mice remained
-susceptible up to ten days post-treatment.
-
-\textbf{5-day laxative treatment differentially disrupts the fecal
+post-challenge (9, 26). All three groups of PEG-treated mice were
+administered a 15\% PEG solution in the drinking water for 5-days, 1.
+received no additional treatment, 2. was also treated with clindamycin,
+and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A).
+PEG treatment resulted in weight loss in all 3 groups of PEG-treated
+mice relative to their baseline weights, with the greatest change in
+weight observed on the fifth day of PEG treatment. The mice recovered
+most of the lost weight by five days after treatment (Fig. 1B). After
+either the PEG, clindamycin, or PEG and clindamycin treatment all mice
+were challenged with 10\textsuperscript{5} \emph{C. difficile} 630
+spores (Fig. 1A). All treatments rendered mice susceptible to \emph{C.
+difficile} colonization. However, PEG-treated mice remained colonized
+with \emph{C. difficile} at a high level through thirty days
+post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice
+cleared \emph{C. difficile} within ten days post-challenge (Fig. 1C).
+Therefore, PEG treatment led to sustained colonization in contrast to
+clindamycin mice that naturally cleared \emph{C. difficile} within ten
+days post-challenge.
+
+Notably, we also found PEG-treated mice were still susceptible to
+\emph{C. difficile} colonization after a 10-day recovery period,
+although \emph{C. difficile} was not detectable in most of the group in
+the initial five days post-challenge (Fig. 1C, S1A). One mouse was found
+dead on the 6th day post-challenge, presumable due to \emph{C.
+difficile} as the bacteria became detecable in that mouse on the 4th day
+post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward,
+the density of \emph{C. difficile} stabilized in the 10-day recovery
+group and remained high through 20-30 days post-challenge (Fig. 1C).
+Thus, osmotic laxative treatment alone was sufficient to render mice
+susceptible to prolonged \emph{C. difficile} colonization and
+PEG-treated mice remained susceptible through ten days post-treatment.
+
+\textbf{5-day laxative treatment differentially disrupted the fecal
 microbiota compared to clindamycin treatment.} Since osmotic laxatives
@@
 well-documented phenomenon that mice housed in the same cages have
-similar microbial communities due to copraphagy (27). We tried to
+similar microbial communities due to coprophagy (27). We tried to
 minimize the impact of cage effects on our experiment by breaking up
-cagemates when assigning mice to treatment groups and primarily housing
-only two mice per cage. Importantly, although we conducted a total of 5
-separate experiments, the experiment number and it's interaction with
-treatment group was not one of the variables that significantly
-explained the observed variation in fecal communities (Data Set S1,
-Sheet X). Interestingly, none of the treatment groups recovered to their
-baseline community structure either 10 or 30 days post-infection
+cagemates when assigning mice to treatment groups. Importantly, although
+we conducted a total of 5 separate experiments, the experiment number
+and its interaction with treatment group was not one of the variables
+that significantly explained the observed variation in fecal communities
+(Data Set S1, Sheet X). None of the treatment groups recovered to their
+baseline community structure either 10 or 30 days post-challenge
 suggesting other community features besides recovery to baseline were
@@
 
-Next, we examined alpha diversity by looking at Shannon diversity over
-time. Although both clindamycin and PEG treatments decreased diversity,
-Shannon diversity was lower in the groups of mice that received PEG
-treatment compared to those that received clindamycin through thirty
-days post-infection (Fig. 2C). We next examined the bacterial genera
-that shifted after PEG treatment by comparing the baseline samples of
-mice treated with only PEG to samples from the same mice one day
-post-treatment. We found 18 genera that were altered by PEG treatment
-(Data Set S1, Sheet X). The majority of the bacterial relative
-abundances decreased after PEG, but \emph{Enterobacteriaceae} and
-\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
-unique to PEG treatment, as \emph{Bacteroides} actually decreased in
-clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria
-that differed across treatment groups over multiple time points. We
-found 24 genera were different over multiple time points out of the 33
-genera that were different between treatment groups (Fig. 2E, Data Set
-S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota
-that was maintained over time and was distinct from clindamycin
-treatment.
+Since none of the communities completely recovered in the follow-up
+period after treatments, we next profiled community diversity and
+composition. We examined the alpha diversity dynamics by calculating the
+communities' Shannon diversity. Although both clindamycin and PEG
+treatments decreased diversity, Shannon diversity was lower in the
+groups of mice that received PEG treatment compared to those that
+received clindamycin through thirty days post-challenge (Fig. 2C). We
+next examined the bacterial genera that shifted after PEG treatment by
+comparing the baseline samples of mice treated with only PEG to samples
+from the same mice one day post-PEG-treatment. We found 18 genera that
+were altered by PEG treatment (Data Set S1, Sheet X). The majority of
+the bacterial relative abundances decreased after the PEG treatment, but
+the relative abundance among members of the \emph{Enterobacteriaceae}
+and \emph{Bacteroides} increased. The increase in \emph{Bacteroides}
+relative abundance was unique to PEG treated mice, as the
+\emph{Bacteroides} relative abundance actually decreased in clindamycin
+treated mice (Fig. 2D). Finally, we identified the genera whose relative
+abundance differed across treatment groups over multiple time points. Of
+the 33 genera that were different between treatment groups, 24 genera
+were different over multiple time points (Fig. 2E, Data Set S1, Sheet
+X). Thus, PEG had a significant impact on the fecal microbiota that was
+maintained over time and was distinct from clindamycin treatment.
 
@@
 stools of the PEG-treated mice that were allowed to recover for 10 days
-prior to challenge. We decided to examine the bacteria that changed
-during the post-infection period when the group median \emph{C.
-difficile} shifted from undetectable at 1 day post-infection to
-detectable in the stool samples with the median stabilizing around 8
-days post-infection (Fig. S1A). Interestingly, we found
-\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
-\emph{Akkermansia} were changing during the time period when \emph{C.
-difficile} was becoming detectable in the stools (Fig. S1B), although
-none of the bacteria were significant after multiple hypothesis
-correction (Data Set S1, Sheet X). While we did not identify a clear
-signal to explain the delayed appearance of \emph{C. difficile} in the
-5-day PEG mice that were allowed to recover for 10 days prior to
-challenge, the delay is striking and could reflect changes in microbial
-activity or metabolites that were not examined in this study.
-
-\textbf{5-day laxative treatment does not promote more severe CDIs
+prior to challenge. We decided to examine if there were genera that
+changed during the post-challenge period when the group median \emph{C.
+difficile} shifted from undetectable at 1 day post-challenge to
+detectable in the stool samples with the density stabilizing around 8
+days post-challenge (Fig. S1A). We found no bacteria were significantly
+different over the two time periods after multiple hypothesis correction
+(Data Set S1, Sheet X). However, there was also wide variation between
+individual mice regarding when \emph{C. difficile} became detectable
+(Fig. S1A) as well as the relative abundances of bacterial genera in the
+communities (Fig. S1B). For example, two mice had a high relative
+abundance of \emph{Enterobacteriaceae} throughout the post-challenge
+period and this corresponded to mouse 10, which died on the sixth day
+post-challenge and mouse 11, where \emph{C. difficile} was present at a
+high density from the 4th day post-challenge onward (Fig. S1B). While we
+did not identify a clear signal to explain the delayed appearance of
+\emph{C. difficile} in the 5-day PEG mice that were allowed to recover
+for 10 days prior to challenge, the delay is striking and could reflect
+changes in microbial activity or metabolites that were not examined in
+this study.
+
+\textbf{5-day laxative treatment did not promote more severe CDIs
 despite altering the mucosal microbiota.} Given the findings from a
@@
 observed with the stool samples, alpha diversity was lower in the
-PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although
-alpha diversity continued to increase over time based on the communities
-from PEG-treated mice collected at 20 and 30 days post-infection (Fig.
-3A, Data Set S1, Sheet X). Group, time point, and their interactions
-with other variables (cage, experiment number, and sample type)
-explained the majority of the variation observed in mucosal communities
-(PERMANOVA combined R\textsuperscript{2} = 0.83, \emph{P} \textless{}
-0.05, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in
-the mucosal microbiota between treatment groups at 6 days post-infection
-with 16 genera that were significantly different in all three of the
-tissue types we collected (cecum, proximal colon, and distal colon; Fig.
-S3A, Data Set S1, Sheet X). However, at 30 days post-infection only
-\emph{Bacteroides}, \emph{Clostridiales}, \emph{Firmicutes}, and
-\emph{Ruminococcaceae} were different between treatment groups and only
-in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X).
-Interestingly, one of the bacteria that was different between treatment
-groups at 6 days post-infection was \emph{Peptostreptococcaceae} (the
-genera that contains a sequence that matches \emph{C. difficile}).
-\emph{Peptostreptococcaceae} was primarily only present in the 5-day PEG
-treatment group of mice and decreased in the proximal and distal colon
-tissues over time (Fig. S2B). Thus, PEG treatment had a significant
-impact on the mucosal microbiota and we detected \emph{C. difficile}
-sequences in the cecum, proximal colon, and distal colon tissue
-communities.
-
-Next, we examined CDI severity by evaluating cecum and colon
-histopathology (28) and found there was no difference in cecum and colon
-scores between clindamycin and PEG-treated mice that were challenged
-with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the time
-point typically examined in \emph{C. difficile} 630 challenged mice
-(29). We also looked at 6 days post-infection because that was when we
-started to see a large difference in \emph{C. difficile} colonization
-levels between PEG- and clindamycin-treated mice (Fig. 1C). Although,
-there was a slight difference in the colon between PEG and
-clindamycin-treated mice, there was no difference in the cecum and the
-overall score is still relatively low given that the max possible
-summary score is twelve (Fig. 3E). Therefore, although PEG treatment had
-a disruptive effect on the mucosal microbiota, the impact of \emph{C.
-difficile} 630 infection on the cecum and colon was similar between PEG
-and clindamycin treated mice.
-
-\textbf{\emph{C. difficile} challenge does not have a synergistic
-disruptive effect on the microbiota of PEG-treated mice} Because
+PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha
+diversity continued to increase over time with the PEG-treated mice
+collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet
+X). Group, time point, and their interactions with other variables
+(cage, experiment number, and sample type) explained the majority of the
+variation observed in mucosal communities (PERMANOVA combined
+R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
+Set S1, Sheet X). We saw the greatest difference in the relative
+abundance of the mucosal microbiota between treatment groups
+(clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days
+post-challenge with 10 genera that were significantly different
+(\emph{P} `r format.pval(tissue\_genera\_max\_p, eps = 0.05)) in all
+three of the tissue types we collected (cecum, proximal colon, and
+distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly,
+\emph{Peptostreptococcaceae} (the family with a sequence that matches
+\emph{C. difficile}) was one of the genera that had a significant
+difference in relative abundance between treatment groups at 6 days
+post-challenge. This population was primarily only present in the 5-day
+PEG treatment group of mice and decreased in the proximal and distal
+colon tissues over time (Fig. S2B). By 30 days post-challenge, only the
+relative abundances of \emph{Bacteroides}, \emph{Clostridiales},
+\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
+treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data
+Set S1, Sheet X). Thus, PEG treatment had a significant impact on the
+mucosal microbiota and we detected \emph{C. difficile} sequences in the
+cecum, proximal colon, and distal colon tissue communities.
+
+Next, we examined the severity of \emph{C. difficile} challenge by
+evaluating cecum and colon H\&E stained histopathology (28) and found
+there was no difference in cecum and colon scores between clindamycin
+and PEG-treated mice that were challenged with \emph{C. difficile} at 4
+days post-challenge (Fig. 3D), the time point typically examined in
+\emph{C. difficile} 630 challenged mice (29). We also looked at 6 days
+post-challenge because that was when there was a large difference in
+\emph{C. difficile} density between PEG- and clindamycin-treated mice
+(Fig. 1C). Although, there was a slight difference in the colon between
+PEG and clindamycin-treated mice, there was not a signifant difference
+in the cecum and the overall score was relatively low (1.5-2.5 out of
+12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect
+on the mucosal microbiota, the impact of \emph{C. difficile} 630
+challenge on the cecum and colon was similar between PEG and clindamycin
+treated mice.
+
+\textbf{\emph{C. difficile} challenge did not have a synergistic
+disruptive effect on the microbiota of PEG-treated mice.} Because
 \emph{C. difficile} itself can have an impact on the microbiota (30), we
@@
 revealed similar bacterial disruptions as the \emph{C. difficile}
-challenged mice (Fig. S2A-C). Similarly, there was no difference between
+challenged mice (Fig. S3A-C). Similarly, there was no difference between
 the tissues of mock and \emph{C. difficile} challenged mice (Fig.
-S2D-F). Thus, most of the microbiota alterations we observed in the
+S3D-F). Thus, most of the microbiota alterations we observed in the
 PEG-treated mice were a result of the laxative and not an interaction
@@
 
-\textbf{1-day laxative treatment results in transient \emph{C.
-difficile} colonization and minor microbiota disruption} Next, we
-decided to examine how a shorter osmotic laxative perturbation would
-impact the microbiome and susceptibility to \emph{C. difficile}. We
-administered either a 1-day PEG treatment, a 1-day PEG treatment with a
-1-day recovery period, or clindamycin to mice before challenging them
-with \emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
+\textbf{1-day laxative treatment resulted in transient \emph{C.
+difficile} colonization and minor microbiota disruption.} Next, we
+examined how a shorter osmotic laxative perturbation would impact the
+microbiome and susceptibility to \emph{C. difficile}. We administered
+either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day
+recovery period, or clindamycin to mice before challenging them with
+\emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
 mice, the 1-day PEG treated mice were only transiently colonized and
-cleared \emph{C. difficile} by 7 days post-infection (Fig. 3B). The
+cleared \emph{C. difficile} by 7 days post-challenge (Fig. 3B). The
 stool communities of PEG-treated mice were also only transiently
-disrupted, with Shannon diversity recovering close to baseline by 7 days
-post-infection (Fig. 3C-D). We found 14 bacteria were impacted by
-treatment, but recovered close to baseline levels by 7 days
-post-infection including \emph{Enterobacteriaceae},
+disrupted, with Shannon diversity recovering by 7 days post-challenge
+(Fig. 3C-D). We found the relative abundances of 14 genera were impacted
+by treatment, but recovered close to baseline levels by 7 days
+post-challenge including \emph{Enterobacteriaceae},
 \emph{Clostridiales}, \emph{Porpyromonadaceae}, and
 \emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
-suggest the duration of treatment matters when considering the impact of
-the laxatives, with shorter treatments resulting in a transient loss of
-\emph{C. difficile} colonization resistance.
-
-\textbf{Post-CDI laxative treatment disrupts clearance in
-clindamycin-treated mice regardless of whether an FMT is also
-administered} Since 1-day PEG treatment resulted in a more mild
+suggest the duration of the PEG treatment was relevant, with shorter
+treatments resulting in a transient loss of \emph{C. difficile}
+colonization resistance.
+
+\textbf{Post-CDI laxative treatment disrupted clearance in
+clindamycin-treated mice regardless of whether an FMT was also
+administered.} Since a 1-day PEG treatment resulted in a more mild
 microbiota perturbation, we decided to use the 1-day treatment to
 examine the hypothesis that PEG helps to flush \emph{C. difficile}
-spores from the intestine. We were also interested in exploring whether
-PEG might help with engraftment in the context of FMTs. To examine the
-impact of PEG treatment on \emph{C. difficile} clearance and FMT
-treatment, we treated 4 groups of mice with clindamycin and then
-challenged all mice with \emph{C. difficile} before administering the
-following treatments: no additional treatment, 1-day PEG immediately
-after challenge, and 1-day PEG treatment 3 days after challenge followed
-by either administration of an FMT or PBS solution by oral gavage (Fig.
-5A). Contrary to our hypothesis, all groups of mice that received PEG
-exhibited prolonged \emph{C. difficile} colonization (Fig. 5B). Next we
-examined how post-CDI PEG treatment impacted the microbiota. Alpha
-diversity was lower in the PEG-treated mice with the exception of the
-PEG-treated mice that were administered an FMT (Fig 5C-D). The FMT
-appeared to partially restore Shannon diversity but not richness (Fig.
-5C-D). Similarly, we saw some overlap between the communities of mice
-that received FMT and the mice treated with only clindamycin after 5
-days post-infection (Fig. 6A). The increase in Shannon diversity
-suggests that the FMT did have an impact on the microbiota, despite
-seeing prolonged \emph{C. difficile} colonization in the FMT treated
-mice. However, only \emph{Bacteroidales} and \emph{Porphyromonadaceae}
+spores from the intestine. To examine the impact of PEG treatment on
+\emph{C. difficile} clearance, we treated 4 groups of mice with
+clindamycin and then challenged all mice with \emph{C. difficile} before
+administering the following treatments: no additional treatment, 1-day
+PEG immediately after challenge, and 1-day PEG treatment 3 days after
+challenge followed by either administration of an FMT or PBS solution by
+oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice
+that received PEG exhibited prolonged \emph{C. difficile} colonization
+(Fig. 5B). Next we examined how post-CDI PEG treatment impacted
+microbiota diversity. Alpha diversity was lower in the PEG-treated mice
+with the exception of the PEG-treated mice that were administered an FMT
+(Fig 5C-D).
+
+We were also interested in exploring whether PEG might help with
+engraftment in the context of FMTs. The FMT appeared to partially
+restore Shannon diversity but not richness (Fig. 5C-D). Similarly, we
+saw some overlap between the communities of mice that received FMT and
+the mice treated with only clindamycin after 5 days post-challenge (Fig.
+6A). The increase in Shannon diversity suggests that the FMT did have an
+impact on the microbiota, despite seeing prolonged \emph{C. difficile}
+colonization in the FMT treated mice. However, only the relative
+abundances of \emph{Bacteroidales} and \emph{Porphyromonadaceae}
 consistently differed between the mice received either an FMT or PBS
 gavage (Fig. 6B), suggesting the FMT only restored a couple of genera.
-Overall, we found there were 24 bacteria that differed over multiple
-time points between treatment groups including increased
-\emph{Akkermansia} and decreased \emph{Ruminococcaceae},
-\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Oscillibacter}
-in mice that received PEG after \emph{C. difficile} challenge (Fig. 6C).
-In sum, administering PEG actually prolonged \emph{C. difficile}
-colonization, including in mice that received an FMT, which only
-restored a couple of bacterial genera.
-
-\textbf{Five-day post-infection community data can predict which mice
-that will have prolonged \emph{C. difficile} colonization} After
+Overall, we found the relative abundances of 24 genera were different
+between treatment groups over multiple timepoints. For example, the
+relative abundance of \emph{Akkermansia} was increased and the relative
+abundances of \emph{Ruminococcaceae}, \emph{Clostridiales},
+\emph{Lachnospiraceae}, and \emph{Oscillibacter} were decreased in mice
+that received PEG after \emph{C. difficile} challenge relative to
+clindamycin treated mice (Fig. 6C). In sum, administering PEG actually
+prolonged \emph{C. difficile} colonization, including in mice that
+received an FMT, which only restored 2 bacterial genera.
+
+\textbf{Five-day post-challenge community data can predict which mice
+that will have prolonged \emph{C. difficile} colonization.} After
 identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day
@@
 machine learning models (random forest, logistic regression, and support
-vector machine) with input bacterial community data from 5 days
-post-infection to predict whether the mice were still colonized with
-\emph{C. difficile} 10 days post-infection. We chose 5 days
-post-infection because that was the earliest time point where we would
-see a treatment effect in the mice that were given 1-day PEG treatment
-three days after \emph{C. difficile} challenge and then administered an
-FMT or PBS gavage. The random forest model had the highest performance
-(AUROC = 0.90, Data Set S1, Sheet X), so we next performed permutation
-importance to examine the bacteria that were the top contributors to the
-random forest model predicting prolonged \emph{C. difficile
-}colonization. We selected the top 10 bacteria contributing to our
-models performance (Fig. 7A) and examined their relative abundance at 5
-days post-infection, the time point used to predict \emph{C. difficile}
+vector machine) with bacterial community data from 5 days post-challenge
+to predict whether the mice were still colonized with \emph{C.
+difficile} 10 days post-challenge. We chose 5 days post-challenge
+because that was the earliest time point where we would see a treatment
+effect in the mice that were given 1-day PEG treatment three days after
+challenge. The random forest model had the highest performance (AUROC =
+0.90, Data Set S1, Sheet X), so we next performed permutation importance
+to examine the bacteria that were the top contributors to the random
+forest model predicting prolonged \emph{C. difficile }colonization. We
+selected the top 10 bacteria contributing to our models performance
+(Fig. 7A) and examined their relative abundance at 5 days
+post-challenge, the time point used to predict \emph{C. difficile}
 colonization status on day 10 (Fig. 7B). Next, we focused on the 5
@@
 time. We found \emph{Enterobacteriaceae} and \emph{Bacteroides} tended
-to have a higher relative abundance, \emph{Akkermansia} was initially
-decreased and then increased, and \emph{Porphyromonadaceae} and
-\emph{Lachnospiraceae} had a lower relative abundance in the mice with
-prolonged colonization compared to the mice that cleared \emph{C.
-difficile} (Fig. 7C). Together these results suggest a combination of
-low and high abundance bacterial genera influence the prolonged
-colonization observed in 5-day PEG and post-CDI PEG treated mice.
+to have a higher relative abundance, the relative abundance of
+\emph{Akkermansia} was initially decreased and then increased, and
+\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} had a lower
+relative abundance in the mice with prolonged colonization compared to
+the mice that cleared \emph{C. difficile} (Fig. 7C). Together these
+results suggest a combination of low and high abundance bacterial genera
+influence the prolonged colonization observed in 5-day PEG and post-CDI
+PEG treated mice.
 
 Previous work examining the impact of PEG on the murine microbiota found
-that PEG treatment resulted in the permanent loss of S24-7, also known
-as \emph{Muribaculum intestinale} (16). We decided to check our
-\emph{Porphyromonadaceae} operational taxonomic units (OTUs) because
-\emph{Muribaculum intestinale} is classified as
-\emph{Porphyromonadaceae} by the Ribosomal Database Project (RDP)
-database (31) and \emph{Porphyromonadaceae} was a top feature in the
-random forest model predicting prolonged \emph{C. difficile}
-colonization. We identified 4 OTUs that had at least 92\% identity to
-\emph{Muribaculum intestinale} and examined their abundance in mice that
-either cleared or were still colonized with \emph{C. difficile} at 10
-days post-infection. While all of the OTUs, were decreased by PEG and
-clindamycin treatment, there was some recovery in the mice that cleared
-(Fig. S4A). We also examined other \emph{Porphyromonadaceae} and
-\emph{Lachnospiraceae} OTUs since these were the 2 genera that were
-important to our classification model and contained multiple OTUs that
-were different at 5 days post-infection between mice that either cleared
-or remained colonized with \emph{C. difficile} by 10 days post-infection
-(Data Set S1, sheet X). While individual \emph{Porphyromonadaceae} and
-\emph{Lachnospiraceae} OTUs tended to be more abundant in the mice that
-clear \emph{C. difficile} relative to the mice that exhibit prolonged
-colonization (Fig. S4B-C), there is no single OTU that fits the pattern
-we observed at the genus level (Fig. 7C), suggesting multiple
-\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs influenced
-\emph{C. difficile} clearance. Overall, our results suggest that
-specific bacterial community differences explain the prolonged \emph{C.
-difficile} colonization we observed in 5-day PEG and post-CDI 1-day PEG
-treated mice.
+that PEG treatment resulted in the permanent loss of \emph{Muribaculum
+intestinale} (16)., which is classified as \emph{Porphyromonadaceae} by
+the Ribosomal Database Project (RDP) database (31).
+\emph{Porphyromonadaceae} was a top feature in the random forest model
+predicting prolonged \emph{C. difficile} colonization. We identified 4
+OTUs that had 92-96\% identity to \emph{Muribaculum intestinale} and
+examined their abundance in mice that either cleared or were still
+colonized with \emph{C. difficile} at 10 days post-challenge. While all
+of the OTUs, were decreased by PEG and clindamycin treatment, there was
+some recovery in the mice that cleared (Fig. S4A). We also examined
+other \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs since
+these were the 2 genera that were important to our classification model
+and contained multiple OTUs that were different at 5 days post-challenge
+between mice that either cleared or remained colonized with \emph{C.
+difficile} by 10 days post-challenge (Data Set S1, sheet X). While
+individual \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs
+tended to be more abundant in the mice that clear \emph{C. difficile}
+relative to the mice that exhibit prolonged colonization (Fig. S4B-C),
+there is no single OTU that fits the pattern we observed at the genus
+level (Fig. 7C), suggesting multiple \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} OTUs influenced \emph{C. difficile} clearance.
+Overall, our results suggest that specific bacterial community
+differences explain the prolonged \emph{C. difficile} colonization we
+observed in 5-day PEG and post-CDI 1-day PEG treated mice.
 
@@
 difficile} 630 spores. A subset of mice were euthanized on either 4 or 6
-days post-infection and tissues were collected for histopathology
+days post-challenge and tissues were collected for histopathology
 analysis, the remaining mice were followed through 20 or 30 days
-post-infection. B. Weight change from baseline weight in groups after
+post-challenge. B. Weight change from baseline weight in groups after
 treatment with PEG and/or clindamycin, followed by \emph{C. difficile}
 challenge. C. \emph{C. difficile} CFU/gram stool measured over time (N =
-16-59 mice per time point) via serial dilutions. The black line
+10-59 mice per time point) via serial dilutions. The black line
 represents the limit of detection for the first serial dilution. CFU
@@
 individual mouse and the transparency of the symbol corresponds to the
-day post-infection. B. Bray-Curtis distances of stool samples collected
-on either day 10 or 30 post-infection relative to the baseline sample
+day post-challenge. B. Bray-Curtis distances of stool samples collected
+on either day 10 or 30 post-challenge relative to the baseline sample
 collected for each mouse (before any drug treatments were administered).
@@
 respectively. For A-B, transparency of the symbol corresponds to the day
-post-infection that the sample was collected. C. The median relative
+post-challenge that the sample was collected. C. The median relative
 abundance of the 4 genera that were significantly different between the
@@
 summary score is 12. The tissue for histology was collected at either 4
-(D) or 6 (E) days post-infection with the exception that one set of
+(D) or 6 (E) days post-challenge with the exception that one set of
 5-day PEG treated mock-challenged mice were collected on day 0
-post-infection (first set of open circles in D). Histology data were
+post-challenge (first set of open circles in D). Histology data were
 analyzed with the Kruskal-Wallis test followed by pairwise Wilcoxon
@@
 symbol represents a sample from an individual mouse and symbol
-transparency corresponds to the day post-infection that the sample was
+transparency corresponds to the day post-challenge that the sample was
 collected. E. Median relative abundances per treatment group for 6 out
 of the 14 genera that were affected by treatment, but recovered close to
-baseline levels by 7 days post-infection (Fig. 3E, Data Set S1, Sheet
+baseline levels by 7 days post-challenge (Fig. 3E, Data Set S1, Sheet
 X). Stool samples from either baseline and day 1 or baseline and day 7
@@
 PBS (4) solution by oral gavage. Mice were followed through 15-30 days
-post-infection (only the post-CDI 1-day PEG group was followed through
-30 days post-infection). B. CFU/g of \emph{C. difficile} stool measured
+post-challenge (only the post-CDI 1-day PEG group was followed through
+30 days post-challenge). B. CFU/g of \emph{C. difficile} stool measured
 over time via serial dilutions. The black line represents the limit of
@@
 median value for each treatment group. The transparancy of the symbol
-corresponds to the day post-infection. Asterisks indicate time points
+corresponds to the day post-challenge. Asterisks indicate time points
 with significant differences between groups by a Kruskall-Wallis test
@@
 individual sample, the transparency of the circle corresponds to day
-post-infection as shown in Fig. 6C-D. B. Median relative abundances of 2
+post-challenge as shown in Fig. 6C-D. B. Median relative abundances of 2
 genera that were significantly different over multiple time points in
@@
 PEG treated mice.} A. Top ten bacteria that contributed to the random
-forest model trained on five day post-infection community relative
+forest model trained on five day post-challenge community relative
 abundance data to predict wehther mice would still be colonized with
-\emph{C. difficile} 10 days post-infection. The median (point) and
+\emph{C. difficile} 10 days post-challenge. The median (point) and
 interquantile range (lines) change in AUROC when the bacteria is left
 out of the model is shown. B. The median relative abundances at 5 days
-post-infection of the top ten bacteria that contributed to the random
+post-challenge of the top ten bacteria that contributed to the random
 forest classification model. Color indicates whether the mice were still
-colonized with \emph{C. difficile} 10 days post-infection and the black
+colonized with \emph{C. difficile} 10 days post-challenge and the black
 horizontal line represents the median relative abundance. Each symbol
@@
 straightforward and the mice clear \emph{C.difficile} within 10 days
-post-infection. However, for mice that recieved a 5-day PEG treatment or
+post-challenge. However, for mice that recieved a 5-day PEG treatment or
 a 1-day PEG treatment after \emph{C. difficile} challenge recovery of
 colonization resistance is more uncertain because mice were still
-colonized with \emph{C. difficile} 30 days post-infection in the case of
+colonized with \emph{C. difficile} 30 days post-challenge in the case of
 several PEG treatments. We found increased \emph{Porphyromonadaceae} and
@@
 \includegraphics{figure_S1.pdf} \textbf{Figure S1. Microbiota dynamics
-post-infection in the 5-day PEG treatment plus 10-day recovery mice.} A.
+post-challenge in the 5-day PEG treatment plus 10-day recovery mice.} A.
 \emph{C. difficile} CFU/g over time in the stool samples collected from
@@
 removed. B. Median relative abundances of 8 bacterial genera from day 0
-post-infection onward from the 10-day recovery mice. We analyzed samples
-from day 0 and day 8 post-infection, which represented the the time
+post-challenge onward from the 10-day recovery mice. We analyzed samples
+from day 0 and day 8 post-challenge, which represented the the time
 points where mice were challenged with \emph{C. difficile} and when the
@@
 \includegraphics{figure_S2.pdf} \textbf{Figure S2. PEG treatment still
-has a large impact on the mucosal microbiota 6 days post-infection} A.
+has a large impact on the mucosal microbiota 6 days post-challenge} A.
 The\\
@@
 \includegraphics{figure_S4.pdf} \textbf{Figure S4. Specific OTUs
-associated with clearance by 10 days post-infection that are mostly
+associated with clearance by 10 days post-challenge that are mostly
 absent in mice with prolonged \emph{C. difficile} colonization. Ex.

### 2021-04-30 code/5_days_PEG_16S.R
Additional figure edits

@@ -126,8 +126,34 @@ shannon_stools <- plot_shannon_overtime(diversity_stools) +
   scale_x_continuous(breaks = c(-15, -10, -5, -4, -2, -1:10, 15, 20, 30),
                      limits = c(-16,31),
                      minor_breaks = c(-15.5,-14.5, -10.5, -9.5, -5.5, -4.5, -3.5, -2.5, -1.5:10.5, 14.5, 15.5, 19.5, 20.5, 29.5, 30.5)) +
-  theme(legend.position = "bottom")
+  theme(legend.position = "none")
 save_plot(filename = "results/figures/5_days_PEG_shannon_stools.png", shannon_stools, base_height = 4, base_width = 9, base_aspect_ratio = 2)
+
+#Create 2nd version of shannon overtime for stool samples with no points
+shannon_stools_median <- diversity_stools %>%
+  group_by(group, day) %>%
+  mutate(median_shannon = median(shannon)) %>%
+  ggplot(x = day, y = shannon, colour = group)+
+  geom_line(mapping = aes(x = day, y = median_shannon, group = group, color = group), alpha = 0.6, size = 1) +
+  scale_colour_manual(name=NULL,
+                      values=color_scheme,
+                      breaks=color_groups,
+                      labels=color_labels) +
+  scale_y_continuous(limits = c(0,4.1), expand = c(0, .2))+ #expand argument gets rid of the extra space around the scale
+  scale_x_continuous(breaks = c(-15, -10, -5, -4, -2, -1:10, 15, 20, 30),
+                     limits = c(-16,31),
+                     minor_breaks = c(-15.5,-14.5, -10.5, -9.5, -5.5, -4.5, -3.5, -2.5, -1.5:10.5, 14.5, 15.5, 19.5, 20.5, 29.5, 30.5)) +
+  
+  theme_classic()+
+  labs(title=NULL,
+       x="Days post-challenge",
+       y="Shannon diversity index")+
+  annotate("text", y = y_position, x = x_annotation, label = label, size =7)+ #Add statistical annotations
+  theme(legend.position = "none",
+        text = element_text(size = 16), # Change font size for entire plot
+        axis.ticks.x = element_blank(),
+        panel.grid.minor.x = element_line(size = 0.4, color = "grey"))#Add gray lines to clearly separate symbols by days))
[…]

### 2021-04-30 code/figure_2.R

@@ -1,16 +1,18 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
 a <- ggdraw() + draw_image("results/figures/5_days_PEG_stool_PCoA.png", scale = 1)
+legend <- ggdraw() + draw_image("results/figures/5_days_PEG_pcoa_legend.png", scale = 1)
+legend<- plot_grid(legend, NULL, NULL, labels = NULL, label_size = 12, nrow = 1, rel_heights = c(1.5, 1), rel_widths = c(1, 0.3))
 b <- ggdraw() + draw_image("results/figures/5_days_PEG_bc_combined_baseline.png", scale = 1)
-c <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools.png", scale = 1)
+c <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_median.png", scale = 1)
 d <- ggdraw() + draw_image("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell_v2.png")
 e <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_stools.png", scale =0.95)
 
-top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow = 1, rel_widths = c(1, 1.5))
+top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow = 1, rel_widths = c(1, 1.2))
 middle_panel <- plot_grid(c, labels = c("C"), label_size = 12, ncol = 1)
 b_right_panel <- plot_grid(e, NULL, labels = c("E", ""), label_size = 12, ncol = 1)
 bottom_panel <- plot_grid(d, b_right_panel, labels = c("D", ""), label_size = 12, nrow = 1, rel_heights = c(1.2, 1))
-plot_grid(top_panel, middle_panel, bottom_panel, ncol = 1, rel_heights = c(1, 1, 2)) +
+plot_grid(top_panel, legend, middle_panel, bottom_panel, ncol = 1, rel_heights = c(1, 0.2, 1, 2)) +
   ggsave("results/figures/figure_2.pdf", width = 6.875, height = 9)+
   ggsave("submission/figure_2.pdf", width = 6.875, height = 9)
 

### 2021-04-30 code/relative_dist_to_baseline.R

@@ -101,6 +101,34 @@ save_plot(filename = "results/figures/5_days_PEG_bc_d30_baseline.png", d30_basel
 plot_grid(d10_baseline, d30_baseline, nrow = 1)+
   ggsave("results/figures/5_days_PEG_bc_combined_baseline.png", height = 3, width = 5)
 
+#Create faceted version for day 10 and day 30 relative to baseline
+facet_labels <- c("Day 10", "Day 30") #Create descriptive labels for facets
+names(facet_labels) <- c("10", "30") #values that correspond to group, which is the variable we're faceting by
+d10_30_facet_plot <- five_d_stools %>% 
+  filter(row_day %in% c("10", "30"),
+         col_day %in% c("baseline")) %>% 
+  group_by(group, row_day) %>% 
+  mutate(median = median(distances)) %>% 
+  ungroup() %>% 
+  ggplot(aes(x = group, y = distances, color = group)) +
+  geom_errorbar(aes(ymax=median, ymin=median), color = "gray50", size = 1, show.legend = FALSE)+ #Add lines to indicate median 
+  geom_jitter(size=2) +
+  scale_colour_manual(name=NULL,
+                      values=color_scheme,
+                      breaks=color_groups,
+                      labels=color_labels)+ 
+  labs(title = NULL, x = NULL, y = "Bray-Curtis distance \n relative to baseline")+
+  facet_wrap(~row_day, labeller = labeller(row_day = facet_labels)) + #Make sure you specify facet_labels before running function
+  theme(plot.title = element_text(hjust = 0.5)) +#Center plot title
+  theme_classic()+
+  lims(y = c(0, 1))+
+  theme(legend.position = "none",
+        strip.background = element_blank(), #get rid of box around facet_wrap labels
+        text = element_text(size = 14),# Change font size for entire plot
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank())+ #Remove x axis ticks
+  ggsave("results/figures/5_days_PEG_bc_combined_baseline.png", height = 3, width = 5)
+
 #Format 1-day PEG stool samples distance matrix----
 one_d_stools <- format_dist_df("data/process/1_day_PEG/peg3350.opti_mcc.braycurtis.0.03.lt.ave.dist") %>% 
   #Rename day values to baseline based on the group 
[…]

### 2021-04-30 code/5_days_PEG_16S.R
Add PCoA alpha legend

@@ -338,15 +338,17 @@ save_plot(filename = paste0("results/figures/5_days_PEG_tissues_PCoA.png"), pcoa
 
 #Create stand alone legend
 pcoa_shape_legend <- pcoa_5_day_PEG_tissues  %>%
-  ggplot(aes(x = axis1, y = axis2, shape = sample_type))+
+  ggplot(aes(x = axis1, y = axis2, shape = sample_type, alpha = day))+
   scale_shape_manual(name = NULL, 
                      values=sample_shape_scheme,
                      breaks=sample_shape_groups,
                      labels=sample_shape_labels)+
   geom_point()+ theme_classic()+
-  theme(legend.position = "bottom")
+  theme(legend.position = "bottom",
+        legend.box="vertical",
+        legend.margin=margin())
 pcoa_shape_legend<- get_legend(pcoa_shape_legend)
-save_plot("results/figures/5_days_PEG_pcoa_legend_tissues.png", pcoa_shape_legend, base_height = .25, base_width = 3)
+save_plot("results/figures/5_days_PEG_pcoa_legend_tissues.png", pcoa_shape_legend, base_height = .7, base_width = 3)
 
 #PCoA faceted over time
 pcoa_plot_time <- plot_pcoa_t(pcoa_5_day_PEG_tissues)+

### 2021-04-30 code/1_day_PEG_16S.R
Add PCoA alpha scale, adjust composite figure

@@ -231,14 +231,15 @@ save_plot(filename = paste0("results/figures/1_Day_PEG_PCoA.png"), pcoa_subset_p
 
 #Create stand alone legend
 group_legend <- pcoa_1_day_PEG  %>%
-  ggplot(aes(x = axis1, y = axis2, color = group))+
+  ggplot(aes(x = axis1, y = axis2, alpha = day))+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_groups,
                       labels=color_labels)+
-  geom_point()+ theme_classic() 
+  geom_point()+ theme_classic() +
+  theme(legend.position = "right")
 group_legend <- get_legend(group_legend)
-save_plot("results/figures/1_day_PEG_pcoa_legend.png", group_legend, base_height = .8, base_width = 2.2)
+save_plot("results/figures/1_day_PEG_pcoa_legend.png", group_legend, base_height = 2.2, base_width = .5)
 
 pcoa_animated <- plot_pcoa(pcoa_1_day_PEG)+
   labs(x = paste("PCoA 1 (", axis1, "%)", sep = ""), #Annotations for each axis from loadings file

### 2021-04-30 code/figure_4.R

@@ -2,10 +2,14 @@ source("code/utilities.R")
 
 a <- ggdraw() + draw_image("results/figures/1_Day_PEG_Schematic.png", scale = 1.45, x = .09)
 b <- ggdraw() + draw_image("results/figures/1_day_PEG_cfu.png", scale = 1.00)
-c <- ggdraw() + draw_image("results/figures/1_Day_PEG_PCoA.png", scale = 1.25)
+c <- ggdraw() + draw_image("results/figures/1_Day_PEG_PCoA.png", scale = 1)
+legend <- ggdraw() + draw_image("results/figures/1_day_PEG_pcoa_legend.png", scale = 1)
+c <- plot_grid(c, legend, nrow = 1, rel_widths = c(1, .1), scale = 1.18)
 d <- ggdraw() + draw_image("results/figures/1_Day_PEG_shannon.png")
 e <- ggdraw() + draw_image("results/figures/1_Day_PEG_genus_6_v2_baselinetoD1_lineplot.png")
-top_panel <- plot_grid(a, b, c, d, labels = c("A", "B", "C", "D"), label_size = 12, ncol = 2, rel_widths = c(.60, 1, 1,.60), rel_heights = c(1, 1))
+top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow = 1, rel_widths = c(.60, 1), rel_heights = c(1, 1))
+middle_panel <- plot_grid(c, d, labels = c("C", "D"), label_size = 12, nrow = 1, rel_widths = c(.8, 1), rel_heights = c(1, 1))
+top_panel <- plot_grid(top_panel, middle_panel, ncol = 1)
 bottom_panel <- plot_grid(e, labels = c("E"), label_size = 12)
 plot_grid(top_panel, bottom_panel, nrow = 2, rel_heights = c(1.1, .8)) +
   ggsave("results/figures/figure_4.pdf", width = 6.875, height = 6)+

### 2021-04-30 code/post_CDI_PEG_16S.R
Fix shaded rectangle annotation

@@ -70,9 +70,10 @@ shannon_post_cdi_peg_overtime_stool <- diversity_stools %>%
                      limits = c(-2,31), #removes day -15 here
                      minor_breaks = c(-1.5:10.5, 14.5, 15.5, 19.5, 20.5, 24.5, 25.5, 29.5, 30.5)) +
   theme(legend.position = "none")+ #Removing legend to save separately
+  scale_y_continuous(limits = c(0,4.1), expand = c(0, 0))+ #expand argument gets rid of the extra space around the scale
   annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = Inf, fill = "#88419d", alpha = .15)+ #shade to indicate PEG treatment in Clind + 1-day PEG group
   annotate("rect", xmin = 3, xmax = 4, ymin = 0, ymax = 2, fill = "#225ea8", alpha = .15)+ #shade to indicate PEG treatment in Clind + 3-day recovery + 1-day PEG + FMT/PBS
-  annotate("rect", xmin = 3, xmax = 4, ymin = 2, ymax = Inf, fill = "#f768a1", alpha = .15)
+  annotate("rect", xmin = 3, xmax = 4, ymin = 2, ymax = Inf, fill = "#f768a1", alpha = .15)+
 save_plot("results/figures/post_CDI_PEG_shannon_stool.png", shannon_post_cdi_peg_overtime_stool, base_height = 4, base_width = 8.5, base_aspect_ratio = 2) #Save full Shannon over time plot without legend
 
 #Extract Shannon Legend for all plots

### 2021-04-30 code/figure_6.R
Add alpha legend to PCoA, adjust figure

@@ -1,7 +1,7 @@
 source("code/utilities.R") #Loads libraries, reads in metadata, functions
 
 #Vert version of Fig 4 16S
-a <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png", scale = 1.2)
+a <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png", scale = 1)
 b <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend_vert.png")
 c <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_fmt.png")
 lineplot <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_stools.png")

### 2021-04-30 code/post_CDI_PEG_16S.R

@@ -167,7 +167,8 @@ axis2_stool <- pcoa_axes_post_cdi_PEG_stool %>% filter(axis == 2) %>% pull(loadi
 
 pcoa_subset_plot_stool <- plot_pcoa(pcoa_post_cdi_peg_stool)+
   labs(x = paste("PCoA 1 (", axis1_stool, "%)", sep = ""), #Annotations for each axis from loadings file
-       y = paste("PCoA 2 (", axis2_stool,"%)", sep = ""))
+       y = paste("PCoA 2 (", axis2_stool,"%)", sep = ""))+
+  theme(legend.position = "none")
   #annotate("text", x = -.6, y = .41, label = paste(str_to_title(top_R2_title)), size = 3.2) +
   #annotate("text", x = -.45, y = .41, label = paste(str_to_title(sec_R2_title)), size = 3.2) +
   #annotate("text", x = -.6, y = .375, label = paste("italic(R)^2: ", top_R2, sep = "" ), parse = TRUE, size = 3.2) +
@@ -178,25 +179,26 @@ save_plot(filename = paste0("results/figures/post_CDI_PEG_stool_pcoa.png"), pcoa
 
 #Create stand alone legend
 group_legend <- pcoa_post_cdi_peg_stool  %>%
-  ggplot(aes(x = axis1, y = axis2, color = group))+
+  ggplot(aes(x = axis1, y = axis2, color = group, alpha = day))+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_groups,
                       labels=color_labels)+
   geom_point()+ theme_classic()+
-  guides(color = guide_legend(ncol = 2))
+  guides(color = guide_legend(ncol = 2))+ 
+  theme(legend.position = "bottom")
 group_legend <- get_legend(group_legend)
-save_plot("results/figures/post_CDI_PEG_pcoa_legend.png", group_legend, base_height = .8, base_width = 6)
+save_plot("results/figures/post_CDI_PEG_pcoa_legend.png", group_legend, base_height = .8, base_width = 9)
 #Create Standalone legend, vertical
 group_legend_vert <- pcoa_post_cdi_peg_stool  %>%
-  ggplot(aes(x = axis1, y = axis2, color = group))+
+  ggplot(aes(x = axis1, y = axis2, color = group, alpha = day))+
   scale_colour_manual(name=NULL,
                       values=color_scheme,
                       breaks=color_groups,
[…]

### 2021-05-05 submission/manuscript.Rmd
Add treatment legend to figure

@@
 ## Introduction
-
 Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because they disrupt microbiota colonization resistance [@Britton2014].
@@
 ## Results
-
 **5-day laxative treatment led to prolonged *C.
@@
 ## Discussion
-
-* Summary of major findings (Fig. 8A)
+By following osmotic laxative treated mice over time, we found 5-day and 1-day PEG treatments exhibited either prolonged or transient *C.
+difficile* colonization after challenge.
+The differences in *C.
+difficile* colonization dynamics between the 5- and 1-day PEG treatments were associated with differences in how much the treatments disrupted the microbiota.
+We also found the intestinal communties of 5-day PEG treated mice had not regained colonization resistance after a 10-day recovery period.
+Additionally, treating mice with PEG after *C.
+difficile* challenge prolongs colonization instead of enhancing clearance, as suggested by the literature.
+Overall, our results demonstrated that treating mice with an osmotic laxative rendered mice susceptible to *C.
+difficile* colonization and duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after *C.
+difficile* challenge.
+Note (remove figure 8?)
 
@@
 * Histopathology [@Theriot2011]
+
+\newpage
+
+## References
 
@@
 \includegraphics{figure_S2.pdf}
-**Figure S2. PEG treatment still has a large impact on the mucosal microbiota 6 days post-challenge** A. The   
+**Figure S2.
+PEG treatment still has a large impact on the mucosal microbiota 6 days post-challenge** A.
+The relative abundances of the 10 bacterial genera that were significantly different between treatment groups at 6 days post-infection in cecum tissue (the relative abundances of the 10 genera were also significantly different in the proximal and distal colon tissues, Data Sheet S1, Sheet X).
+Each symbol represents a tissue sample from an individual mouse, the horizonal line represents the median relative abundance for each treatment group.
+B.
+The relative abunance of *Peptostreptococacceae* in the three types of tissue sample communities over time.
+For A-B, the gray horizontal lines represent the limit of detection.
 \newpage
@@
 \includegraphics{figure_S3.pdf}
-**Figure S3. *C. difficile* challenge does not enhance the disruptive effect of PEG on the microbiota.** A.
+**Figure S3.
+*C.
+difficile* challenge does not enhance the disruptive effect of PEG on the microbiota.** A, D.
+PCoAs of the Bray-Curtis distances from the stool (A) and tissue (D) communities  from mock- and *C.
+difficile*-challenged treatment groups.
+Each symbol represents a sample from an individual mouse with open and closed circles representing mock and *C.
+difficile*-challenged mice, respectively.
+B, E.
+Median Shannon diversity in stool (B) and tissue (E) communities collected over time.
+C, F.
+The median relative abundances of genera that were significant between treatment groups displayed in either stool (Fig.
+2E) or cecum tissue (Fig.
+3C) communities for stool (C) and tissue (F) commmunites from mock- and *C.
+difficile*-challenged mice.
+For B-F, the dashed and solid lines represent the median value for mock and *C.
+difficile*-challenged mice, respectively.
+For E-F, tissues from mock-challenged clindamycin treated mice were only collected 4 days post-challenge so there is no dashed line for this group.
 \newpage
@@
 \includegraphics{figure_S4.pdf}
-**Figure S4. Specific OTUs associated with clearance by 10 days post-challenge that are mostly absent in mice with prolonged *C. difficile* colonization. Ex. *Muribaculum intestinale*.** A.
+**Figure S4. Specific OTUs associated with clearance by 10 days post-challenge that are mostly absent in mice with prolonged *C. difficile* colonization.** Remove this figure?
 \newpage
-
-## References

### 2021-05-05 code/figure_6.R

@@ -3,10 +3,11 @@ source("code/utilities.R") #Loads libraries, reads in metadata, functions
 #Vert version of Fig 4 16S
 a <- ggdraw() + draw_image("results/figures/post_CDI_PEG_stool_pcoa.png", scale = 1)
 b <- ggdraw() + draw_image("results/figures/post_CDI_PEG_pcoa_legend_vert.png")
+legend <- ggdraw() + draw_image("results/figures/post_CDI_PEG_treatment_legend.png") #Treatment legend for rectangles
 c <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_fmt.png")
 lineplot <- ggdraw() + draw_image("results/figures/post_CDI_PEG_genus_lineplot_stools.png")
 pcoa1 <- plot_grid(a ,b, labels = c("A", ""), label_size = 12, nrow = 1, rel_heights = c(1,.5), rel_widths = c(1,.5))
 
-plot_grid(pcoa1, c, lineplot, labels = c("", "B", "C"), label_size = 12, nrow = 3)+
-  ggsave("results/figures/figure_6.pdf", width = 6.3, height = 9)+
-  ggsave("submission/figure_6.pdf", width = 6.3, height = 9)
+plot_grid(pcoa1, c, legend, lineplot, labels = c("", "B", "", "C"), label_size = 12, nrow = 4, rel_heights = c(1, 1, .2, 1))+
+  ggsave("results/figures/figure_6.pdf", width = 6, height = 9)+
+  ggsave("submission/figure_6.pdf", width = 6, height = 9)

### 2021-05-06 code/figure_8.R
Remove panel B

@@ -3,5 +3,5 @@ source("code/utilities.R") #Loads libraries, reads in metadata, functions
 a <- ggdraw() + draw_image("results/figures/summary_schematic.png")
 
 plot_grid(a, labels = NULL, label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_8.pdf", width=6.875, height=4.5)+
-  ggsave("submission/figure_8.pdf", width=6.875, height=4.5)
+  ggsave("results/figures/figure_8.pdf", width=5, height=4.5)+
+  ggsave("submission/figure_8.pdf", width=5, height=4.5)

### 2021-05-06 submission/manuscript.Rmd
Add to discussion

@@
 difficile* colonization dynamics between the 5- and 1-day PEG treatments were associated with differences in how much the treatments disrupted the microbiota.
-We also found the intestinal communties of 5-day PEG treated mice had not regained colonization resistance after a 10-day recovery period.
-Additionally, treating mice with PEG after *C.
-difficile* challenge prolongs colonization instead of enhancing clearance, as suggested by the literature.
-Overall, our results demonstrated that treating mice with an osmotic laxative rendered mice susceptible to *C.
-difficile* colonization and duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after *C.
+The intestinal communities of 5-day PEG treated mice had not regained colonization resistance after a 10-day recovery period.
+In contrast to the hypothesize suggested by the literature, treating mice with PEG after *C.
+difficile* challenge prolonged colonization.
+We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C.
+difficile* colonization (Fig.
+8).
+Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C.
+difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after *C.
 difficile* challenge.
-Note (remove figure 8?)
+Note (remove figure 8B, check 8A...specify 5-day vs 1-day or just  in figure legend)
 
-* Discussion of prolonged persistence. *C. difficile* sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent papers.
+Some of the lower abundance bacteria could also play important roles in clearing C.
+difficile.
+For example, the relative abundance of Oscillibacter was increased in mice that cleared *C.
+difficile* compared to the mice with prolonged colonization.
+Oscillibacter __ can produce the SCFA valerate, which has been shown to inhibit *C.
+difficile* growth based on ___ data.
+Similarly, Acetatifactor can produce __, another SCFA with ___ role in the context of CDIs.
+Thus bacteria with potential protective functions might have played a role in promoting *C.
+difficile* clearance.
 
-* Discuss why we might not have observed more severe histology in PEG mice relative to clindamycin-treated mice
+There was an interesting pattern of increased Enterobacteriaceae, Bacteroides, and C. difficile in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy. 
+Akkermansia is also a mucin degrader with potentially beneficial or detetrimental roles depending on context in other diseases ().
+In our study the relative abundance of *Akkermansia* shifted over time between groups.
+In the stool it was initially increased in mice that cleared *C.
+difficile*, but shifted after 5-days post-challenge so that it was increased in mice that had prolonged colonization.
+In the context of CDIs, some studies suggest a protective role, while others suggest detrimental.
+
+* Discussion of prolonged persistence.
+*C.
+difficile* sequences detected in tissue samples.
+Association with mucin-degrading bacteria suggested by recent papers.
+Akkermansia, Bacteroides are known mucin degraders and could be compete.
+
+Bacteroides and Enterobacteriaceae synergy in the context of CRC (Dejea paper?)
+
+* Protective bacteria missing in PEG-treated mice
+
+Despite identifying an altered compositional profile that included higher relative abundance of the *C.
+difficile* sequence in the mucosal tissues of mice treated with 5-day PEG compared to the clindamycin group, we did not see a difference in histopathology scores between the groups.
+One reason there was no difference could be the *C.
+difficile* strain used, *C.
+difficile* 630 is mild in mice compared to VPI (ref).
+Part of our hypothesis for why we might see increased histopathology scores is because PEG was previously shown to disrupt the mucus layer in mice.
+However, a recent study demonstrated that broad spectrum antibiotics (clinda included) can also disrupt the host mucus layer in mice?
+ref.
+Birchenough.
+Future research is needed to tease out the interplay between drugs that influence the mucus layer and different strains of *C.
+difficile* in the context of CDIs.
+
+Discuss why we might not have observed more severe histology in PEG mice relative to clindamycin-treated mice
   + Antibiotics may also impact mucus layer
@@
 
-* Protective bacteria missing in PEG-treated mice
-
-* Discuss what these findings might mean for human patients (Fig. 8B)
+* Discuss what these findings might mean for human patients 
   + What's known regarding laxatives and susceptibility to CDIs

### 2021-05-06 submission/manuscript.Rmd
Remove OTU analysis based on Pat's feedback

@@
 
-Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of *Muribaculum intestinale* [@Tropini2018]., which is classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020].
-*Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C.
-difficile* colonization.
-We identified 4 OTUs that had 92-96% identity to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C.
-difficile* at 10 days post-challenge.
-While all of the OTUs, were decreased by PEG and clindamycin treatment, there was some recovery in the mice that cleared (Fig.
-S4A).
-We also examined other *Porphyromonadaceae* and *Lachnospiraceae* OTUs since these were the 2 genera that were important to our classification model and contained multiple OTUs that were different at 5 days post-challenge between mice that either cleared or remained colonized with *C.
-difficile* by 10 days post-challenge (Data Set S1, sheet X).
-While individual *Porphyromonadaceae* and *Lachnospiraceae* OTUs tended to be more abundant in the mice that clear *C.
-difficile* relative to the mice that exhibit prolonged colonization (Fig.
-S4B-C), there is no single OTU that fits the pattern we observed at the genus level (Fig.
-7C), suggesting multiple *Porphyromonadaceae* and *Lachnospiraceae* OTUs influenced *C.
-difficile* clearance.
-Overall, our results suggest that specific bacterial community differences explain the prolonged *C.
-difficile* colonization we observed in 5-day PEG and post-CDI 1-day PEG treated mice.
-
 ## Discussion
@@
 \newpage
-
-\includegraphics{figure_S4.pdf}
-**Figure S4. Specific OTUs associated with clearance by 10 days post-challenge that are mostly absent in mice with prolonged *C. difficile* colonization.** Remove this figure?
-\newpage

### 2021-05-06 code/figure_S4.R

@@ -1,9 +1,0 @@
-source("code/utilities.R") #Loads libraries, reads in metadata, functions
-
-a <- ggdraw() + draw_image("results/figures/ml_abund_otu_muribaculum_lineplot.png")
-b <- ggdraw() + draw_image("results/figures/ml_abund_otu_porphyromonadaceae_lineplot.png")
-c <- ggdraw() + draw_image("results/figures/ml_abund_otu_lachnospiraceae_lineplot.png")
-
-plot_grid(a, b, c, labels = c("A", "B", "C"), label_size = 12, ncol=1)+
-  ggsave("results/figures/figure_S4.pdf", width=6.875, height=6.5)+
-  ggsave("submission/figure_S4.pdf", width=6.875, height=6.5)

### 2021-05-06 code/ml_results.R

@@ -304,263 +304,4 @@ topabund_5_model_taxa_line_plot <- agg_genus_data %>%
         strip.text = element_markdown(size = 12),
         legend.position = "None")
 save_plot(filename = paste0("results/figures/ml_abund_5_genus_lineplot.png"), topabund_5_model_taxa_line_plot, base_height = 3, base_width = 10)  
-
-#Examine Porphyromondaceae OTUs that are similar to Muribaculum intestinale----
-#See code/blast_otus.R for how blast search was performed and results
-muribac_otus <- c("Porphyromonadaceae (OTU 6)", "Porphyromonadaceae (OTU 7)", "Porphyromonadaceae (OTU 8)", "Porphyromonadaceae (OTU 21)")
-muribac_otu_names <- c("*Porphyromonadaceae* (OTU 6)", "*Porphyromonadaceae* (OTU 7)", "*Porphyromonadaceae* (OTU 8)", "*Porphyromonadaceae* (OTU 21)")
-#Line plots instead of area plot to convey how the most abundant genera that were top contributors to machine learning models were changing over time
-muribac_otus_line_plot <- agg_otu_data %>% 
-  #Solve different baseline timepoints across groups by specifying baseline for each group
-  mutate(day = case_when(group == "C" & day %in% c("-15", "-11", "-1") ~ "B",
-                         group == "WM" & day == "-5" ~ "B",
-                         group == "WMC" & day == "-5" ~ "B",
-                         group == "WMR" & day == "-15" ~ "B",
-                         group == "CWM" & day == "-1" ~ "B",
-                         group == "RM" & day == "-1" ~ "B",
-                         group == "FRM" & day == "-1" ~ "B",
-                         group == "M1" & day == "-11" ~ "B",
-                         group == "1RM1" & day == "-2" ~ "B",
-                         TRUE ~ day)) %>% 
-  filter(day %in% c("B", "0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "15")) %>%  #Use baseline through day 15 timepoints
-  mutate(day = fct_relevel(day, "B", "0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "15")) %>% 
-  filter(clearance_status_d10 %in% c("colonized", "cleared")) %>% #Remove samples we don't have clearance status d10 data for
-  filter(otu %in% muribac_otus) %>%
[…]

### 2021-05-06 submission/manuscript.Rmd
Add PEG treatment durations

@@
 ## Discussion
-By following osmotic laxative treated mice over time, we found 5-day and 1-day PEG treatments exhibited either prolonged or transient *C.
-difficile* colonization after challenge.
+By following osmotic laxative treated mice over time, we found 5-day PEG treatment before challenge resulted in prolonged *C.
+difficile* colonization upon challenge, while a 1-day PEG treatment resulted in transient colonization.
 The differences in *C.
-difficile* colonization dynamics between the 5- and 1-day PEG treatments were associated with differences in how much the treatments disrupted the microbiota.
-The intestinal communities of 5-day PEG treated mice had not regained colonization resistance after a 10-day recovery period.
-In contrast to the hypothesize suggested by the literature, treating mice with PEG after *C.
+difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota.
+Additionally, the intestinal communities of 5-day PEG treated mice had not regained *C.
+difficile* colonization resistance after a 10-day recovery period.
+In contrast to the hypothesis suggested by the literature, treating mice with PEG after *C.
 difficile* challenge prolonged colonization.
@@
 8).
-Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C.
+Overall, our results demonstrated that an osmotic laxative treatment alone rendered mice susceptible to *C.
 difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after *C.
-difficile* challenge.
-Note (remove figure 8B, check 8A...specify 5-day vs 1-day or just  in figure legend)
+difficile* challenge.)
 

### 2021-05-14 code/community_type_analysis.R
Remove exporatory analysis

@@ -1,304 +1,0 @@
-source("code/utilities.R") #Loads libraries, reads in metadata, functions
-
-#Visualize get.communitytype analysis results----
-#Read in dmm fit data to evaluate community type fit with the laplace value depending on the number of community types
-dmm_fit <- read_tsv("data/process/peg3350.subsample.genus.genus.dmm.mix.fit")
-
-laplace_plot <- dmm_fit %>% 
-  ggplot()+
-  geom_line(aes(x = K, y = Laplace))+
-  theme_classic()
-#Save results
-save_plot(filename = "exploratory/notebook/motility_community_type_laplace.png", laplace_plot)
-
-#Read in the rest of the output files from mothur's get.communitytype function:
-#Read in data that indicate the bacteria (at the genus level) membership of the 15 different  community types
-community_otus <- read_tsv("data/process/peg3350.subsample.genus.genus.dmm.mix.summary") %>% 
-  rename("otu" = "OTU") #rename to match otu column in taxonomy
-community_parameters <- read_tsv("data/process/peg3350.subsample.genus.genus.dmm.mix.parameters")
-sample_community_membership <- read_tsv("data/process/peg3350.subsample.genus.genus.dmm.15.mix.posterior")
-sample_best_community_fit <- read_tsv("data/process/peg3350.subsample.genus.genus.dmm.mix.design", col_names=c("unique_label", "best_fitting_community")) %>% 
-  left_join(metadata, by = "unique_label") %>%  #Join to metadata
-  mutate(community=str_replace(best_fitting_community,"Partition_(\\d*)","\\1")) %>% #Add a column with just the community number
-  mutate(community= as.numeric(community)) #Transform community variable type from character to numeric
-sample_community_rel_abund <- read_tsv("data/process/peg3350.subsample.genus.genus.dmm.15.mix.relabund")
-
-## Import genus-level taxonomy into data frame and clean up names
-taxonomy <- read_tsv(file="data/process/peg3350.genus.taxonomy") %>%
-  rename_all(tolower) %>% #remove uppercase from column names
-  rename(genus = taxonomy) %>% #Rename taxonomy to genus
-  # Clean up genus names  
-  mutate(genus=str_replace_all(genus, c('Bacteria_unclassified' = 'Unclassified',
-                                        "Clostridium_" = "Clostridium ", #Remove underscores after Clostridium
-                                        "_" = " ", #Removes all other underscores
-                                        "unclassified" = "unclassified")))
-
[…]

### 2021-05-14 code/metalonda_analysis.R

@@ -1,142 +1,0 @@
-#To install MetaLonDA
-#install.packages("MetaLonDA")
-#may need to change CRANmirror: chooseCRANmirror() Make a selection by typing in the number of desired option
-#Also need to install the following packages that are dependencies:
-#‘metagenomeSeq’, ‘DESeq2’, ‘edgeR’
-#Not available for R version 4.0.2
-#metagenomeSeq suggests installing Bioconductor packages with BiocManager
-#To install BiocManager:
-#if (!requireNamespace("BiocManager", quietly = TRUE))
-#  install.packages("BiocManager")
-#BiocManager::install(version = "3.12")
-#Once BiocManager installed:
-#library(BiocManager) #Only need to install packages
-#To install additional packages required for MetaLonDA with BioConductor
-#BiocManager::install(c("metagenomeSeq", "DESeq2", "edgeR"))
-#Will also ask you to update older versions of packages already installed. For now, select n for update none
-library(MetaLonDA)
-
-#Load read counts of 8 features from 100 samples
-#Samples are from 2 groups, 5 subjects per group, and 10 time points per subject
-#data("metalonda_test_data")
-#Create Group, Time & ID annotation vectors
-# n.group = 2
-# n.sample = 5
-# n.timepoints = 10
-# Group = factor(c(rep("A", n.sample*n.timepoints), rep("B", n.sample*n.timepoints)))
-# Time = rep(rep(1:n.timepoints, times = n.sample), 2)
-# ID = factor(rep(1:(2*n.sample), each = n.timepoints))
-
-#Define the prediction timepoints
-# points = seq(1, 10, length.out = 100)
-# output.metalonda.f5 = metalonda(Count = metalonda_test_data[5,], Time = Time, Group = Group, ID = ID,
-#                                 n.perm = 20, fit.method = "nbinomial", points = points,
-#                                 text = rownames(metalonda_test_data)[5], parall = FALSE, pvalue.threshold = 0.5,
-#                                 adjust.method = "BH", time.unit = "hours", ylabel = "Read Counts", col = c("chartreuse", "blue4"))
-# #Note for test example 20 permutations were used, for real analysis. Recommend at least 1000 permutations
-#
-# ## Identify significant time intervals for all features:
-# output.metalonda.all = metalondaAll(Count = metalonda_test_data, Time = Time, Group = Group,
[…]

### 2021-05-14 code/spiec-easi_analysis.R

@@ -1,169 +1,0 @@
-source("code/utilities.R") #Loads libraries, reads in metadata, functions
-
-library(devtools)
-#install_github("zdk123/SpiecEasi") how to install SpiecEasi
-#Initially: got the following error:
-#clang: error: linker command failed with exit code 1 (use -v to see invocation)
-#make: *** [SpiecEasi.so] Error 1
-#ERROR: compilation failed for package ‘SpiecEasi’
-#* removing ‘/Library/Frameworks/R.framework/Versions/4.0/Resources/library/SpiecEasi’
-#Error: Failed to install 'SpiecEasi' from GitHub: 
-#Solution:
-#Similar error encountered by others with potential solutions: https://github.com/zdk123/SpiecEasi/issues/138
-#Had to install fortran: https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos/
-#Alternative option: install package with conda: https://anaconda.org/bioconda/r-spieceasi 
-library(SpiecEasi)
-library(Matrix)
-library(igraph)
-library(ggnet) #To install: devtools::install_github("briatte/ggnet")
-library(GGally)
-library(network)
-library(sna)
-source('code/functions/ggnet2.R') #Modified by Nick in his 2020 preprint paper
-#Nick's implemnetation of Spiec-Easi: https://github.com/SchlossLab/Lesniak_Clearance_XXXX_2020/blob/master/code/build_fig6.R
-
-set.seed(19760620) #Same seed used for mothur analysis
-seed <- 19760620 
-
-metadata <- metadata %>%
-  mutate(day = as.integer(day))  #Day variable (transformed to integer to get rid of decimals on PCoA animation
-
-#Analysis of project data at the genus level
-genus_data <- read_tsv("data/process/peg3350.subsample.genus.shared") %>%
-  select(-label, -numOtus) %>%
-  rename(unique_label = Group) %>% 
-  left_join(select(metadata, group, unique_label, day, avg_cfu), by = "unique_label") #Join to metadata
-
-#Read in taxonomy labels at the genus level
-## Import genus-level taxonomy into data frame and clean up names
-network_labels <- read_tsv(file="data/process/peg3350.genus.taxonomy") %>%
-  rename_all(tolower) %>% #remove uppercase from column names
-  rename(genus = taxonomy) %>% #Rename taxonomy to genus
-  # Clean up genus names  
[…]

### 2021-05-14 code/slurm/communitytype.sh
Rename and reorganize for clarity

@@ -1,41 +1,0 @@
-#!/bin/bash
-
-###############################
-#                             #
-#  1) Job Submission Options  #
-#                             #
-###############################
-
-# Name
-#SBATCH --job-name=communitytype
-
-# Resources
-# For MPI, increase ntasks-per-node
-# For multithreading, increase cpus-per-task
-#SBATCH --nodes=1
-#SBATCH --ntasks-per-node=1
-#SBATCH --cpus-per-task=10
-#SBATCH --mem=2GB
-#SBATCH --time=24:00:00
-
-# Account
-#SBATCH --account=pschloss1
-#SBATCH --partition=standard
-
-# Logs
-#SBATCH --mail-user=tomkoset@umich.edu
-#SBATCH --mail-type=BEGIN,END,FAIL
-#SBATCH --output=%x-%j.out
-
-# Environment
-#SBATCH --export=ALL
-
-# To run mothur: conda activate mothur_v1.43 before submitting job.
-
-#####################
-#                   #
-#  2) Job Commands  #
-#                   #
-#####################
-
-mothur code/community_type.batch

### 2021-05-14 code/slurm/metalonda.sh

@@ -1,41 +1,0 @@
-#!/bin/bash
-
-###############################
-#                             #
-#  1) Job Submission Options  #
-#                             #
-###############################
-
-# Name
-#SBATCH --job-name=metalonda_rm_wm
-
-# Resources
-# For MPI, increase ntasks-per-node
-# For multithreading, increase cpus-per-task
-#SBATCH --nodes=1
-#SBATCH --ntasks-per-node=1
-#SBATCH --cpus-per-task=20
-#SBATCH --mem=120GB
-#SBATCH --time=72:00:00
-
-# Account
-#SBATCH --account=pschloss1
-#SBATCH --partition=standard
-
-# Logs
-#SBATCH --mail-user=tomkoset@umich.edu
-#SBATCH --mail-type=BEGIN,END,FAIL
-#SBATCH --output=%x-%j.out
-
-# Environment
-#SBATCH --export=ALL
-
-# To run metalonda: conda activate rstats before submitting job.
-
-#####################
-#                   #
-#  2) Job Commands  #
-#                   #
-#####################
-
-Rscript code/metalonda_analysis.R

### 2021-05-14 code/supplemental_data_set_S1.R
Start putting together data set S1

@@ -1,0 +1,27 @@
+source("code/utilities.R")
+
+#Generate single excel workbook containing sheets for Data Set S1
+
+#Sheet 1: 5-day PEG stool PERMANOVA results----
+permanova_5_d_stools <- read_tsv("data/process/5_days_PEG_permanova_stools.tsv")
+
+
+#Combine all the above tables as sheets---
+sheets_combined <- list("Sheet 1" = permanova_5_d_stools, 
+                    "Sheet 2" = _____,
+                    "Sheet 3" = _____,
+                    "Sheet 4"  = _____, 
+                    "Sheet 5" = _____, 
+                    "Sheet 6" = _____,
+                    "Sheet 7" = _____,
+                    "Sheet 8" = _____, 
+                    "Sheet 9" = _____,
+                    "Sheet 10" = _____,
+                    "Sheet 11" = _____,
+                    "Sheet 12" = _____, 
+                    "Sheet 13" = _____,
+                    "Sheet 14" = _____,
+                    "Sheet 15" = _____)
+
+#Write sheets to  Data Set S1 excel file----
+write_xlsx(sheets_combined, "submission/Data Set S1.xlsx", format_headers = TRUE)

### 2021-05-14 submission/manuscript.Rmd
Add abstract, methods, discussion

@@
 ## Abstract
-(Modify depending on target journal, currently abstract submitted to World Microbe Forum) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome.
+Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota.
 However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear.
 To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C.
-difficile* 630 spores.
+difficile* 630.
 We used clindamycin-treated mice as a control because they consistently clear *C.
 difficile* within 10 days post-challenge (dpc).
-To examine how PEG treatment impacts clearance, we administered PEG for 1 day to clindamycin-treated, *C.
-difficile*-challenged mice either immediately following challenge or 3 dpc.
-We collected longitudinal stool samples to examine *C.
-difficile* levels in the stool via anaerobic culture and profiled the microbiota by 16S rRNA sequencing.
-PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpc.
-Additionally, 5-day PEG treated mice remained susceptible to CDI 10-days post treatment.
+PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remain colonized for up to 30 dpc.
 In contrast, 1-day PEG treated mice were transiently colonized, clearing *C.
 difficile* within 7 dpc.
-Although 5-day PEG-treated mice exhibited prolonged *C.
-difficile* colonization, we saw no difference in histological inflammation between PEG- and clindamycin-treated mice.
-Additionally, administering PEG to mice after *C.
-difficile* challenge prolonged colonization up to 30 dpc in mice that received PEG immediately after challenge and 15 dpc in mice that received PEG 3 dpc.
-When we examined microbiota composition across our different treatment groups, we found decreased richness in the PEG-treated mice that exhibited prolonged *C.
-difficile* colonization.
-Importantly, there were increased Bacteroides and Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in most of the PEG-treated mice with prolonged *C.
-difficile* colonization.
-Our findings suggest the osmotic laxative PEG 3350 alters the mouse microbiota and disrupts colonization resistance to *C.
-difficile*, as well as clearance in mice with a CDI.
+To examine how PEG treatment impacts clearance, we administered a 1-day PEG treatment to clindamycin-treated, *C.
+difficile*-challenged mice.
+Administering PEG to mice after *C.
+difficile* challenge prolonged colonization up to 30 dpc.
+When we trained a random forest model with community data from 5 dpc, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90).
+Five of the top ten bacterial features important for predicting prolonged colonization had high relative abundances in the community.
+Examining the dynamics of these bacterial during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C.
+difficile* colonization in PEG-treated mice.
+
+## Importance
+Diarrheal samples induced by medications such as laxatives are typically rejected for *Clostridiodes difficile* testing.
+However, there are some microbiota similarities between diarrheal and *C.
+difficile* infection (CDI) communities such as lower diversity compared to healthy controls, which lead us to hypothesize that diarrhea may be an indicator of *C.
+difficile* risk.
+We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C.
+difficile* by administering a laxative to mice either before or after *C.
+difficile* challenge Our findings suggest the osmotic laxative disrupts colonization resistance to *C.
+difficile*, as well as clearance in mice already colonized with *C.
+difficile*.
 Considering that most hospitals recommend not performing *C.
-difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.
+difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.
 
@@
 However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020].
-Although, other medications (proton pump inhibitors, osmotic laxatives, antimotility agents, and opioids) have been implicated as risk or protective factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].
+Although, other medications (proton pump inhibitors, osmotic laxatives, antimotility agents, and opioids) have been implicated as risk or protective factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiota is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].
 
@@
 Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C.
-difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020].
-The mucus barrier is thought to mediate protection from *C.
-difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C.
+difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020].
+The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C.
 difficile* [@Olson2013; @Diebel2014].
@@
 Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C.
-difficile* colonization and severe CDI and  than mice treated with clindamycin, and if PEG treatment after colonization can promote *C.
+difficile* colonization and severe CDI than mice treated with clindamycin, and if PEG treatment after challenge can promote *C.
 difficile* clearance.
@@
 difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020].
-All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, 1.
+All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days: 1.
 received no additional treatment, 2.
@@
 1C, S1A).
-One mouse was found dead on the 6th day post-challenge, presumable due to *C.
-difficile* as the bacteria became detecable in that mouse on the 4th day post-challenge (Fig.
+One mouse was found dead on the 6th day post-challenge, presumably due to *C.
+difficile* as the bacteria became detectable in stool samples from that mouse on the 4th day post-challenge (Fig.
 S1A, mouse 10).
@@
 Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to coprophagy [@Nguyen2015].
-We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups.
-Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X).
+We tried to minimize the impact of cage effects on our experiment by breaking up littermates when assigning mice to treatment groups.
+Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group did no significantly explain the observed variation in fecal communities (Data Set S1, Sheet X).
 None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C.
@@
 S1A).
-We found no bacteria were significantly different over the two time periods after multiple hypothesis correction (Data Set S1, Sheet X).
+We found no bacteria with relative abundances that were significantly different over the two time points after multiple hypothesis correction (Data Set S1, Sheet X).
 However, there was also wide variation between individual mice regarding when *C.
@@
 3A).
-Alpha diversity continued to increase over time with the PEG-treated mice collected at 20 and 30 days post-challenge (Fig.
+Alpha diversity continued to increase over time based on the tissues from PEG-treated mice collected at 20 and 30 days post-challenge (Fig.
 3A, Data Set S1, Sheet X).
@@
 3B, Data Set S1, Sheet X).
-We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
+We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)`) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
 S2A, Data Set S1, Sheet X).
@@
 
-Next, we examined the severity of *C.
-difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C.
+Because there were differences in the mucosal microbiota including detectable *C.
+difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C.
+difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011].
+However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C.
 difficile* at 4 days post-challenge (Fig.
@@
 ```
-**Post-CDI laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
+**Post-challenge laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
 difficile* spores from the intestine.
@@
 5B).
-Next we examined how post-CDI PEG treatment impacted microbiota diversity.
-Alpha diversity was lower in the PEG-treated mice with the exception of the PEG-treated mice that were administered an FMT (Fig 5C-D).
 
@@
   group_by(method) %>%
-  summarize(median = median(AUC)) %>%
+  summarize(median = median(AUC)) %>% #Median AUC for the 100 seeds of each method
   arrange(desc(median))
-#Highest AUC value
+#Highest median AUC value out of the 3 methods (random forest) run.
 max_auc <- auc_results %>% select(median) %>% max()
-
-#To Do: Insert minimum Percent identity as variable name once blast results to Muribaculum have been combined into 1 file
 
@@
 **Five-day post-challenge community data can predict which mice that will have prolonged *C.
-difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C.
+difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C.
 difficile* colonization.
@@
 difficile* 10 days post-challenge.
-We chose 5 days post-challenge because that was the earliest time point where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after challenge.
-The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C.
+We chose 5 days post-challenge because that was the earliest time point where we would see a treatment effect in the post-challenge 1-day PEG experiments.
+The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C.
 difficile *colonization.
-We selected the top 10 bacteria contributing to our models performance (Fig.
+We selected the top 10 bacteria contributing to our model's performance (Fig.
 7A) and examined their relative abundance at 5 days post-challenge, the time point used to predict *C.
@@
 7C).
-Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-CDI PEG treated mice.
+Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-challenge 1-day PEG treated mice.
 
 ## Discussion
+While the disruptive effect of antibiotics on *C.
+difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear.
 By following osmotic laxative treated mice over time, we found 5-day PEG treatment before challenge resulted in prolonged *C.
-difficile* colonization upon challenge, while a 1-day PEG treatment resulted in transient colonization.
+difficile* colonization, while a 1-day PEG treatment resulted in transient colonization.
 The differences in *C.
 difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota.
-Additionally, the intestinal communities of 5-day PEG treated mice had not regained *C.
-difficile* colonization resistance after a 10-day recovery period.
-In contrast to the hypothesis suggested by the literature, treating mice with PEG after *C.
-difficile* challenge prolonged colonization.
+Additionally, the intestinal communities of 5-day PEG treated mice had not regained colonization resistance after a 10-day recovery period.
+Although in contrast to the other 5-day PEG treatment groups, *C.
+difficile* was not immediately detectable in the stools of most of the mice in the 10-day recovery group.
+We also examined the impact of PEG treatment after *C.
+difficile* challenge and in opposition to the hypothesis suggested by the literature, found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment.
 We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C.
@@
 8).
-Overall, our results demonstrated that an osmotic laxative treatment alone rendered mice susceptible to *C.
-difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after *C.
-difficile* challenge.)
+Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C.
+difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after challenge.
 
-Some of the lower abundance bacteria could also play important roles in clearing C.
-difficile.
-For example, the relative abundance of Oscillibacter was increased in mice that cleared *C.
-difficile* compared to the mice with prolonged colonization.
-Oscillibacter __ can produce the SCFA valerate, which has been shown to inhibit *C.
-difficile* growth based on ___ data.
-Similarly, Acetatifactor can produce __, another SCFA with ___ role in the context of CDIs.
-Thus bacteria with potential protective functions might have played a role in promoting *C.
-difficile* clearance.
+In addition to altering composition, laxative treatment may alter microbiota-produced metabolites.
+A previous study demonstrated that a 5-day treatment of 10% PEG depleted acetate and butyrate and increased succinate compared to untreated mice [@Ferreyra2014].
+While we did not perform metabolomic analysis, we did see bacteria known to produce benefical metabolites were depleted in mice that cleared *C.
+difficile* compared to mice with prolonged colonization.
+For example, *Oscillibacter valericigenes* can produce the SCFA valerate [@Iino2007], and separate studies demonstrated valerate is depleted after clindamycin treatment and inhibited *C.
+difficile* growth based on *in vitro* data and an experiment with C57BL/6 mice [@Jump2014; @McDonald2018].
+Similarly, *Acetatifactor* can produce acetate and butyrate [@Pfeiffer2012], SCFAs that are decreased in mice with prolonged *C.
+difficile* infection after antibiotic treatment [@Lawley2012].
+Thus protective bacteria and their metabolites could be depleted by osmotive laxative treatment depending on the timing and duration of treatment.
 
-There was an interesting pattern of increased Enterobacteriaceae, Bacteroides, and C. difficile in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy. 
-Akkermansia is also a mucin degrader with potentially beneficial or detetrimental roles depending on context in other diseases ().
+One possible explanation for the prolonged *C.
+difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment.
+In fact, it has been hypothesized that *C.
+difficile* biofilms may serve as reservoirs for recurrent infections [@Frost2021] and *C.
+difficile* biofilms were recently identified in patients and aggregates with *Fusobacterium nucleatum*, another species capable of forming biofilms [@Engevik2021].
+There was an interesting pattern of increased *Enterobacteriaceae*, *Bacteroides*, and *C.
+difficile* in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy.
+*Bacteroides* has the potential to degrade mucus and the osmotic laxative may have allowed *Bacteroides* to colonize the mucosal niche by degrading mucin glycans with glycosyl hydrolases that are absent in *C.
+difficile* [@Engevik2020].
+*Bacteroides* persistent in the mucosal tissue might also have helped *Enterobacteriaceae* to colonize the region, as a synergy between mucus-degrading *B.
+fragilis* and *E.
+coli* has previously been described [@Dejea2018].
+A separate study demonstrated *C.
+difficile* was present in the outer mucus layer and associated with *Enterobacteriaceae* and *Bacteroidaceae* using fluorescent in situ hybridization (FISH) staining [@Semenyuk2015].
+However, *B.
+fragilis* prevented CDI morbidity in a mouse model and inhibited *C.
+difficile* adherence *in vitro* [@Deng2018].
+In coculture experiments *B.
+longum* decreased *C.
+difficile* biofilm formation while *B.
+thetaiotamicron* enhanced biofilm formation [@Normington2021].
+Therefore, whether *Bacterodes* is detrimental or beneficial in the context of *C.
+difficile* infection or colonization is still unclear.
+
+*Akkermansia* is also a mucin degrader with potentially beneficial or detrimental roles depending on context in other diseases [@Cirstea2018; @BorgesCanha2015].
 In our study the relative abundance of *Akkermansia* shifted over time between groups.
@@
 difficile*, but shifted after 5-days post-challenge so that it was increased in mice that had prolonged colonization.
-In the context of CDIs, some studies suggest a protective role, while others suggest detrimental.
-
-* Discussion of prolonged persistence.
-*C.
-difficile* sequences detected in tissue samples.
-Association with mucin-degrading bacteria suggested by recent papers.
-Akkermansia, Bacteroides are known mucin degraders and could be compete.
-
-Bacteroides and Enterobacteriaceae synergy in the context of CRC (Dejea paper?)
-
-* Protective bacteria missing in PEG-treated mice
+In the context of CDIs, some studies suggest a protective role [@Pereira2020; @Rodriguez2016], while others suggest detrimental [@Sangster2016; @Vakili2020a; @Vakili2020b].
+Because the relative abundance of *Akkermansia* was dynamic in our study so it is unclear whether *Akkermansia* helps with clearance of *C.
+difficile* or allows it to persist.
+A better understanding how *C.
+difficile* interacts with the mucosal microbiota may lead to insights into CDIs, asymptomatic *C.
+difficile* colonization, and resistance.
 
@@
 difficile* strain used, *C.
-difficile* 630 is mild in mice compared to VPI (ref).
-Part of our hypothesis for why we might see increased histopathology scores is because PEG was previously shown to disrupt the mucus layer in mice.
-However, a recent study demonstrated that broad spectrum antibiotics (clinda included) can also disrupt the host mucus layer in mice?
-ref.
-Birchenough.
-Future research is needed to tease out the interplay between drugs that influence the mucus layer and different strains of *C.
+difficile* 630 results in mild histopathology summary scores in mice compared to VPI 10463 despite both strains producing toxin in mice [@Theriot2011].
+Part of our hypothesis for why there could have been increased histopathology scores in PEG-treated mice was because PEG was previously shown to disrupt the mucus layer in mice.
+However, recent studies demonstrated that broad spectrum antibiotics can also disrupt the host mucosal barrier in mice [@Kester2020; @Bergstrom2020].
+Future research is needed to tease out the interplay between medications that influence the mucus layer and different strains of *C.
 difficile* in the context of CDIs.
 
-Discuss why we might not have observed more severe histology in PEG mice relative to clindamycin-treated mice
-  + Antibiotics may also impact mucus layer
-  + Strain of bacteria used
-
-* Discuss what these findings might mean for human patients 
+It is more difficult what are findings mean in the context of *C.
+difficile* colonization resistance mean for human patients.
+Considering that most hospitals recommend not performing *C.
+difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs.
+further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.
+Further studies are needed to understand the impact of osmotic laxatives on *C.
+difficile* colonization resistance and clearance in human patients.
 + What's known regarding laxatives and susceptibility to CDIs
   + Clinical trial of PEG, results never posted [@Dieterle2018]
-  + Relevance to human FMTs? Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.
 
-## Conclusions
+Relevance to human FMTs?
+Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.
+However, results from the FMT National Registry where 85% of FMTs were delivered by colonoscopy demonstrate FMTs are highly effective treatment for recurrent CDIs with 90% achieving resolution in the 1 month follow-up window [@Kelly2021].
+
+We have demonstrated that osmotic laxative treatment alone has a substantial impact on the microbiota and rendered mice susceptible to prolonged *C.
+difficile* colonization in contrast to clindamycin-treated mice.
+The duration and timing of the laxative treatment impacted the duration of *C.
+difficile* colonization, with only 5-day PEG and post-challenge 1-day PEG treatments prolonging colonization compared to clindamycin treated mice.
+Further studies are warranted to ascertain whether laxatives have a similar impact on *C.
+difficile* colonization resistance on the human microbiota.
 
@@
 ## Materials and Methods 
-* Histopathology [@Theriot2011]
+**Animals.** All experiments were approved by the University of Michigan Animal Care and Use Committee IACUC (protocol numbers PRO00006983 and PRO00008975).
+All mice were C57Bl/6 and part of the Schloss lab colony which was established in 2010 with mice donated from Vincent Young's lab colony (established with mice purchased from The Jackson Laboratory in 2002).
+We used 7-19 week old female mice for all experiments, which allowed us to break up littermates prior to starting an experiment and distribute them as evenly as possible across treatment groups in order to minimize microbiota differences between experiment groups prior to starting treatments with medications.
+During the experiment, mice were housed at a density of 2-3 mice per cage, with the majority of cages limited to two mice.
+
+**Drug treatments.** For PEG treament groups, fifteen percent PEG 3350 (Miralax) was administered in the drinking water for either 5 or 1-day periods depending on the experiment.
+PEG solution was prepared fresh every 2 days in distilled water and administered to the mice in water bottles.
+Clindamycin treatment groups received distilled water in water bottles during the PEG-treatment periods, with the water being changed at the same frequency.
+For clindamycin treatment, groups of mice received 10 mg/kg clindamycin (Sigma-Aldrich) via intraperitoneal injection.
+All PEG treatment groups received a sham intraperitoneal injection containing filter sterilized saline.
+
+***C.
+difficile* challenge model.** Mice were challenged with 25 microliters off *C.
+difficile* 630 spores at 10^5^ concentration, except for 1 experiment (Fig.
+5A) where the concentration was 10^3^.
+All mock challenged mice received 25 ul vehicle solution (Ultrapure water).
+A Dymax stepper pipette was used to administer the same challenge dose to mice via oral gavage.
+Mice were weighed daily throughout the experiment and stool was collected for quantifying *C.
+difficile* CFU and 16S rRNA gene sequencing.
+There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment.
+The fecal microbiota transplant was prepared with stool samples collected from the mice in the experiment prior to the start of any treatments.
+The stool samples were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS and glycerol was added to make a 15% glycerol solution.
+The solution was then aliquoted into tubes and stored at -80&deg;C until the day of the gavage.
+An aliquot of both the FMT and PBS solutions were also set aside in the -80&deg;C for 16S rRNA gene sequencing.
+The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute.
+The supernatant was then transferred to a separate tube to prevent the gavage needle from clogging with debris during gavage.
+The PBS solution that was administered to the other group was also 15% glycerol.
+Each mouse was administered 100 microliters of either the FMT or PBS solution via gavage.
+When we refer to mice that cleared *C.
+difficile*, we mean that no *C.
+difficile* was detected in the first serial dilution (limit of detection: 100 CFU).
+In some experiments, we collected tissues for 16SrRNA gene sequencing, histopathology, or both.
+For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C.
+For histopathology, cecum and colon tissues were placed into separate casettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.
+
+***C.
+difficile* quantification.** Stool samples from mice were transferred to an anaerobic chamber and serially diluted in reduced PBS.
+Serial dilutions were plated onto taurocholate-cycloserine-cefoxitin-fructose agar (TCCFA) plates plates and counted after 24 hours of incubation at 37&deg;C.
+Stool samples collected from the mice on day 0 post-challenge were also plated onto TCCFA plates to ensure mice were not already colonized with *C.
+difficile* prior to challenge.
+
+**16S rRNA gene sequencing.** Stool samples that were stored in the -80&deg;C were placed into 96-well plates for DNA extractions and library preparation.
+DNA was extracted using the DNeasy Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
+For library preparation, each plate had a mock community control (ZymoBIOMICS microbial community DNA standards) and a negative control (water).
+The V4 region of the 16S rRNA gene was amplified with the AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as previously described [@Kozich2013].
+The PCR amplicons were normalized (SequalPrep normalizatin plate kit from Thermo Fisher Scientific), pooled and quantified (KAPA library quantification kit from KAPA Biosystems), and sequenced with the MiSeq system (Illumina).
+
+**16S rRNA gene sequence analysis.** All sequences were processed with mothur (v.
+1.43) using a previously published protocol [@Schloss2009; @Kozich2013].
+Paired sequencing reads were combined and  aligned with the SILVA (v.
+132) reference database [@Quast2012] and taxonomy was assigned with a modified version (v.
+16) of the Ribosomal Database Project (v.
+11.5) [@Cole2013].
+The error rate for are sequencing data was 0.0559% based on the 17 mock communities we ran with the samples.
+Samples were rarefied to 1,000 sequences, 1,000 times for alpha and beta diversity analyses in order to account for uneven sequencing across samples.
+All but 3 out of 17 water controls had less than 1000 sequences.
+PCoAs were generated based on Bray-Curtis Index distance matrices.
+Permutational multivariate analysis of variance (PERMANOVA) tests were performed on mothur-generated Bray-Curtis distance matrices with the adonis function from the vegan R package [@Vegan2018].
+
+**Histopathology.**  H&E stained sections of cecum and colon tissues collected at either 0, 4, or 6 days post-challenge were coded to be scored in a blinded manner by a board-certified veterinary pathologist (ILB).
+Slides were evaluated using a scoring system developed for mouse models of *C.
+difficile* infection [@Theriot2011].
+Each slide was evaluated for edema, cellular infiltration, and inflammation and given a score ranging from 0-4.
+The summary score was calculated by combining the scores from the 3 categories and ranged from 0-12.
+
+**Classification model training and evaluation.** We used the mikropml package to train and evaluate models to predict *C.
+difficile* colonization status 10 days post-challenge where mice were categorized as either cleared or colonized [@mikropml; @Topcuoglu2021].
+We removed the *C.
+difficile* genus relative abundance data prior to training the model.
+Input community relative abundance data at the genus level from 5 days post-challenge was used to generate random forest, logistic regression, and  support vector machine classification models to predict *C.
+difficile* colonization status 10 days post-challenge.
+To accommodate the small number of samples in our data set we used 50% training and 50% testing splits with repeated 2-fold cross-validation of the training data for hyperparamter tuning.
+Permutation importance was performed as described previously [@Topcuoglu2020] using mikropml [@mikropml; @Topcuoglu2021] with the random forest model because it had the highest AUROC value.
+
+**Statistical analysis.** R (v.
+`r paste(version$major, ".", version$minor, sep = "")`) and the tidyverse package (v.
+`r paste(packageVersion("tidyverse")[1])`) were used for statistical analysis [@r_citation_2020; @Tidyverse2019].
+Kruskal-Wallis tests with Bejamini-Hochberg correction for testing multiple time points were used to analyze differences in *C.
+difficile* CFU, mouse weight change, and alpha diversity between treatment groups.
+Paired Wilcoxon rank signed rank tests were used to identify genera impacted by treatments on matched pairs of samples from 2 time points.
+Bacterial relative abundances that varied between treatment groups at the genus level were identified with the Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs, followed by pairwise Wilcoxon comparisons with Benjamini-Hocherg correction.
+
+**Code availability.** Code for data analysis and generating this paper with accompanying figures is available at https://github.com/SchlossLab/Tomkovich_PEG3350_XXXX_2021.
+
+**Data availability.** The 16S rRNA sequencing data have been deposited in the National Center for Biotechnology Information Sequence Read Archive (BioProject Accession no. PRJNA727293).
 
@@
 ## References
+
+<div id="refs"></div>
 
@@
 *C.
-difficile* CFU/gram stool measured over time (N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per time point) via serial dilutions.
+difficile* CFU/gram stool measured over time via serial dilutions(N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per time point).
 The black line represents the limit of detection for the first serial dilution.
@@
 Principal Coordinate analysis (PCoA) of Bray-Curtis distances from stool samples collected throughout the experiment.
-For A and C, each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-challenge.
+Each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-challenge.
 B.
@@
 The line indicates the median value for each treatment group.
-The colors of the symbols and lines represent the four treatment groups.
+For B-C, the symboles represent samples from individual mice and the line indicates the median value for each treatment group.
 D.
@@
 The symbols represent the median relative abundance for a treatment group at either baseline (open circle) or 1-day post treatment (closed circle).
-Data from the 5-day PEG and 5-day PEG plus 10-day recovery groups were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera.
-The clindamycin and 5-day PEG plus clindamycin treatment groups are shown for comparison.
+Relative abundance data from paired baseline and 1-day post treatment stool sampes from the 5-day PEG and 5-day PEG plus 10-day recovery groups were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera.
+The clindamycin and 5-day PEG plus clindamycin treatment groups are shown on the plot for comparison.
 E.
-6 of the `r f_d_group_multi_day` genera that were significantly different between the four treatment groups over multiple time points.
+6 of the `r f_d_group_multi_day` genera that were significantly different between the treatment groups over multiple time points, the 5-day PEG plus clindamycin treatment group was only followed through 6-days post-challenge.
 Differences between treatment groups were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified genera.
@@
 The line indicates the median value for each treatment group.
-The colors of the symbols and lines represent the four treatment groups.
+The colors of the symbols and lines represent individual and median relative abundance values for four treatment groups.
 A similar pattern was observed with the proximal and distal colon communities (Data Set S1, sheet X-X).
@@
 PCoA of Bray-Curtis distances from mucosal samples collected throughout the experiment.
-Circles, triangles, and squares indicate cecum, proximal colon, and distal colon communities, respectively.
-For A-B, transparency of the symbol corresponds to the day post-challenge that the sample was collected.
+Circles, triangles, and squares indicate the cecum, proximal colon, and distal colon communities, respectively.
+Transparency of the symbol corresponds to the day post-challenge that the sample was collected.
 C.
-The median relative abundance of the 4 genera that were significantly different between the cecum communities of different treatment groups on day 6 and day 30  (Data Set S1, sheet X).
-The gray vertical line indicate the limit of detection.
+The median relative abundance of the 4 genera that were significantly different between the cecum communities of different treatment groups on day 6 and day 30 post-challenge (Data Set S1, sheet X).
+The gray vertical lines indicate the limit of detection.
 D-E.
-The histopathology summary scores from cecum and colon H&E stained slides.
-The summary score is the total score based on evaluation of edema, cellular infiltration, and inflammation.
+The histopathology summary scores from cecum and colon H&E stained tissue sections.
+The summary score is the total score based on evaluation of edema, cellular infiltration, and inflammation in either the cecum or colon tissue.
 Each category is given a score ranging from 0-4, thus the maximum possible summary score is 12.
-The tissue for histology was collected at either 4 (D) or 6 (E) days post-challenge with the exception that one set of 5-day PEG treated mock-challenged mice were collected on day 0 post-challenge (first set of open circles in D).
+The tissue for histology was collected at either 4 (D) or 6 (E) days post-challenge with the exception that one set of 5-day PEG treated mock-challenged mice were collected on day 0 post-challenge (first set of open purple circles in D).
 Histology data were analyzed with the Kruskal-Wallis test followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
@@
   count()
-#To Do: this code is not specific for CFU values or the time points shown on panel B
+
+#Figure out min and max number of stool samples plated to quantify C. difficile CFU per day post-challenge
 one_day_sample_min <- min(one_day_samples_per_day$n)
@@
 *C.
-difficile* CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per time point) via several dilutions.
-The black dotted line represents the limit of detection for the first serial dilution.
-Asterisks indicate time points where the CFU/gram was significantly different between treatment groups by Kruskall-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
+difficile* CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per time point) by serial dilutions.
+The black dashedd horizontal line represents the limit of detection for the first serial dilution.
+For B and D, asterisks indicate time points where there was a significant difference between treatment groups by Kruskall-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
+For B-D, each symbol represents a sample from an individual mouse and lines indicate the median value for each treatment group.
 C.
-PCoA of Bray-Curtis distances from stool communities collected from the three treatment groups over time (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`).
+PCoA of Bray-Curtis distances from stool communities collected over time (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`).
+Symbol transparency represents the day post-challenge of the experiment.
+For C-E, the B on the day legend or days post-challenge X axis stands for baseline and represents the sample that was collected prior to any drug treatments.
 D.
-Shannon diversity in stool communities over time with colored lines representing the median value for each treatment group.
-For B-D, each symbol represents a sample from an individual mouse and symbol transparency corresponds to the day post-challenge that the sample was collected.
+Shannon diversity in stool communities over time.
 E.
@@
 3E, Data Set S1, Sheet X).
-Stool samples from either baseline and day 1 or baseline and day 7 were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera.
-The gray horizontal line represents the limit of detection.
+Paired stool sample relative abundance values either baseline and day 1 or baseline and day 7 were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera.
+Only genera that were different between baseline and 1-day post-challenge, but not baseline and 7-days post-challenge.
+The gray horizontal lines represents the limit of detection.
 
@@
 difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** A.
-Setup off the experimental time line for experiments with post-CDI PEG treated mice.
+Setup of the experimental time line for experiments with post-challenge PEG treated mice.
 There were a total of 4 different treatment groups.
@@
 1.
-Did not receive any additional treatment.
+Received no additional treatment (Clindamycin).
 2.
@@
 3-4.
-3-days after challenge, mice received 1-day PEG treatment and then received either a fecal microbiota transplant (3) or PBS (4) solution by oral gavage.
+3-days after challenge, mice received 1-day PEG treatment and then received either 100 microliters a fecal microbiota transplant (3) or PBS (4) solution by oral gavage.
 Mice were followed through 15-30 days post-challenge (only the post-CDI 1-day PEG group was followed through 30 days post-challenge).
@@
 Each symbol represents a stool sample from an individual mouse with the lines representing the median value for each treatment group.
-The transparancy of the symbol corresponds to the day post-challenge.
 Asterisks indicate time points with significant differences between groups by a Kruskall-Wallis test with a Benjamini-Hochberg correction for testing multiple times points.
@@
 difficile* challenge mice that also receive an FMT only some bacterial genera were restored.** A.
-PCoA of Bray-Curtis distances from stool samples collected over time as well as the FMT solution that was administered to one treatment group.
-Each circle represents an individual sample, the transparency of the circle corresponds to day post-challenge as shown in Fig.
-6C-D.
+PCoA of Bray-Curtis distances from stool samples collected over time as well as the FMT solution that was administered to one of the treatment groups.
+Each circle represents an individual sample, the transparency of the circle corresponds to day post-challenge.
 B.
@@
 Gray horizontal lines represent the limit of detection.
+Differences between treatment groups were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified genera.
+For pairwise comparisons of the groups (B), we performed pairwise Wilcoxon comparisons with Benjamini-Hochberg correction for testing all combinations of group pairs.
 
@@
 difficile* colonization in PEG treated mice.** A.
-Top ten bacteria that contributed to the random forest model trained on five day post-challenge community relative abundance data to predict wehther mice would still be colonized with *C.
+Top ten bacteria that contributed to the random forest model trained on five day post-challenge community relative abundance data to predict whether mice would still be colonized with *C.
 difficile* 10 days post-challenge.
-The median (point) and interquantile range (lines) change in AUROC when the bacteria is left out of the model is shown.
+The median (point) and interquartile range (lines) change in AUROC when the bacteria is left out of the model by permutation feature importance analysis.
 B.
-The median relative abundances at 5 days post-challenge of the top ten bacteria that contributed to the random forest classification model.
+The median relative abundances of the top ten bacteria that contributed to the random forest classification model at 5 days post-challenge .
 Color indicates whether the mice were still colonized with *C.
-difficile* 10 days post-challenge and the black horizontal line represents the median relative abundance.
-Each symbol represents a stool sample from an individual mouse and the shape of the symbol indicates whether the PEG-treated mice recieved a 5-day (Fig.
+difficile* 10 days post-challenge and the black horizontal line represents the median relative abundance for the two categories.
+Each symbol represents a stool sample from an individual mouse and the shape of the symbol indicates whether the PEG-treated mice received a 5-day (Fig.
 1-3), 1-day (Fig.
-4) or post-CDI PEG (Fig.
+4) or post-challenge PEG (Fig.
 5-6) treatment.
@@
 The median relative abundances of the 5 genera with greater than 1% median relative abundance in the stool community over time.
-For B-C, the gray horizontal line represents the limit of detection.
+For B-C, the gray horizontal lines represents the limit of detection.
 \newpage
@@
 \includegraphics{figure_8.pdf}
+
 **Figure 8.
-Schematic summarizing findings.** A.
-The gut microbiota of our C57Bl/6 mice is restant to *C.
-difficle* but treatment with either the antibiotic, clindamycin, or the osmotic laxative, PEG 3350 renders the mice susceptible to *C.
+Schematic summarizing findings.** The gut microbiota of our C57Bl/6 mice is resistant to *C.
+difficle* but treatment with either the antibiotic, clindamycin, or the osmotic laxative, PEG 3350, renders the mice susceptible to *C.
 difficile* colonization.
 Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-challenge.
-However, for mice that recieved a 5-day PEG treatment or a 1-day PEG treatment after *C.
-difficile* challenge recovery of colonization resistance is more uncertain because mice were still colonized with *C.
-difficile* 30 days post-challenge in the case of several PEG treatments.
-We found increased *Porphyromonadaceae* and *Lachnospiraceae* were associated with recovery of colonization resistance, while increased *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C.
+However, for mice that received either a 5-day PEG treatment pre-*C.
+difficile* challenge or a 1-day PEG treatment post-challenge recovery of colonization resistance was delayed because most mice were still colonized with *C.
+difficile* 10 days post-challenge.
+We found increased relative abundances of *Porphyromonadaceae* and *Lachnospiraceae* were associated with recovery of colonization resistance, while increased relative abundances of *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C.
 difficle* colonization.
-B.Considering that most hospitals recommend not performing *C.
-difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs.
-further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.
-Further studies are needed to understand the impact of osmotic laxatives on *C.
-difficile* colonization resistance and clearance in human patients.
 \newpage
@@
 *C.
-difficile* CFU/g over time in the stool samples collected from 5-day PEG treatment plus 10-day recovery mice.
+difficile* CFU/g over time in the stool samples collected from 5-day PEG treated mice that were allowed to recover for 10 days prior to challenge.
 Same data presented in Fig.
-1C, but the data for the other 3 treatment groups have been removed.
+1C, but the data for the other 3 treatment groups have been removed and each line represents the CFU over time for an individual mouse.
+Mouse 10 was found dead 6 days post-challenge.
 B.
-Median relative abundances of 8 bacterial genera from day 0 post-challenge onward from the 10-day recovery mice.
+Relative abundances of eight bacterial genera from day 0 post-challenge onward in each off the 10-day recovery mice.
 We analyzed samples from day 0 and day 8 post-challenge, which represented the the time points where mice were challenged with *C.
 difficile* and when the median relative *C.
-difficile* CFU stabilized for the group using the paired Wilcoxan signed-rank test, but no genera were signifcant after Benjamini-Hochberg correction.
+difficile* CFU stabilized for the group using the paired Wilcoxan signed-rank test, but no genera were significantly different after Benjamini-Hochberg correction.
 \newpage
@@
 PEG treatment still has a large impact on the mucosal microbiota 6 days post-challenge** A.
-The relative abundances of the 10 bacterial genera that were significantly different between treatment groups at 6 days post-infection in cecum tissue (the relative abundances of the 10 genera were also significantly different in the proximal and distal colon tissues, Data Sheet S1, Sheet X).
-Each symbol represents a tissue sample from an individual mouse, the horizonal line represents the median relative abundance for each treatment group.
+The relative abundances of the 10 bacterial genera that were significantly different between treatment groups at 6 days post-infection in the cecum tissue (the relative abundances of the 10 genera were also significantly different in the proximal and distal colon tissues, Data Sheet S1, Sheet X).
+Each symbol represents a tissue sample from an individual mouse, the black horizontal lines represents the median relative abundances for each treatment group.
 B.
-The relative abunance of *Peptostreptococacceae* in the three types of tissue sample communities over time.
+The relative abundance of *Peptostreptococacceae* in the three types of tissue sample communities over time.
 For A-B, the gray horizontal lines represent the limit of detection.
@@
 C, F.
-The median relative abundances of genera that were significant between treatment groups displayed in either stool (Fig.
+The median relative abundances of genera that were significantly different between the *C.
+difficile* challenged treatment groups in either the stool (Fig.
 2E) or cecum tissue (Fig.
-3C) communities for stool (C) and tissue (F) commmunites from mock- and *C.
+3C) communities in the stool (C) and tissue (F) communities from mock- and *C.
 difficile*-challenged mice.

### 2021-05-14 submission/manuscript.tex
Add abstract, methods, discussion

@@
 
-(Modify depending on target journal, currently abstract submitted to
-World Microbe Forum) Antibiotics are a major risk factor for
-\emph{Clostridioides difficile} infections (CDIs) because of their
-impact on the intestinal microbiome. However, non-antibiotic medications
-such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350,
-also alter the microbiota, but whether PEG impacts CDI susceptibility
-and clearance is unclear. To examine how PEG impacts susceptibility, we
-treated C57Bl/6 mice with 5-day and 1-day doses of 15\% PEG in the
-drinking water and then challenged the mice with \emph{C. difficile} 630
-spores. We used clindamycin-treated mice as a control because they
-consistently clear \emph{C. difficile} within 10 days post-challenge
-(dpc). To examine how PEG treatment impacts clearance, we administered
-PEG for 1 day to clindamycin-treated, \emph{C. difficile}-challenged
-mice either immediately following challenge or 3 dpc. We collected
-longitudinal stool samples to examine \emph{C. difficile} levels in the
-stool via anaerobic culture and profiled the microbiota by 16S rRNA
-sequencing. PEG treatment alone was sufficient to render mice
-susceptible to CDI and 5-day PEG-treated mice remain colonized for up to
-30 dpc. Additionally, 5-day PEG treated mice remained susceptible to CDI
-10-days post treatment. In contrast, 1-day PEG treated mice were
-transiently colonized, clearing \emph{C. difficile} within 7 dpc.
-Although 5-day PEG-treated mice exhibited prolonged \emph{C. difficile}
-colonization, we saw no difference in histological inflammation between
-PEG- and clindamycin-treated mice. Additionally, administering PEG to
-mice after \emph{C. difficile} challenge prolonged colonization up to 30
-dpc in mice that received PEG immediately after challenge and 15 dpc in
-mice that received PEG 3 dpc. When we examined microbiota composition
-across our different treatment groups, we found decreased richness in
-the PEG-treated mice that exhibited prolonged \emph{C. difficile}
-colonization. Importantly, there were increased Bacteroides and
-Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in
-most of the PEG-treated mice with prolonged \emph{C. difficile}
-colonization. Our findings suggest the osmotic laxative PEG 3350 alters
-the mouse microbiota and disrupts colonization resistance to \emph{C.
-difficile}, as well as clearance in mice with a CDI. Considering that
-most hospitals recommend not performing \emph{C. difficile} testing on
+Antibiotics are a major risk factor for \emph{Clostridioides difficile}
+infections (CDIs) because of their impact on the microbiota. However,
+non-antibiotic medications such as the ubiquitous osmotic laxative
+polyethylene glycol (PEG) 3350, also alter the microbiota, but whether
+PEG impacts CDI susceptibility and clearance is unclear. To examine how
+PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day
+doses of 15\% PEG in the drinking water and then challenged the mice
+with \emph{C. difficile} 630. We used clindamycin-treated mice as a
+control because they consistently clear \emph{C. difficile} within 10
+days post-challenge (dpc). PEG treatment alone was sufficient to render
+mice susceptible and 5-day PEG-treated mice remain colonized for up to
+30 dpc. In contrast, 1-day PEG treated mice were transiently colonized,
+clearing \emph{C. difficile} within 7 dpc. To examine how PEG treatment
+impacts clearance, we administered a 1-day PEG treatment to
+clindamycin-treated, \emph{C. difficile}-challenged mice. Administering
+PEG to mice after \emph{C. difficile} challenge prolonged colonization
+up to 30 dpc. When we trained a random forest model with community data
+from 5 dpc, we were able to predict which mice would exhibit prolonged
+colonization (AUROC = 0.90). Five of the top ten bacterial features
+important for predicting prolonged colonization had high relative
+abundances in the community. Examining the dynamics of these bacterial
+during the post-challenge period revealed patterns in the relative
+abundances of \emph{Bacteroides}, \emph{Enterobacteriaceae},
+\emph{Porphyromonadaceae}, \emph{Lachnospiraceae}, and
+\emph{Akkermansia} that were associated with prolonged \emph{C.
+difficile} colonization in PEG-treated mice.
+
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
+
+Diarrheal samples induced by medications such as laxatives are typically
+rejected for \emph{Clostridiodes difficile} testing. However, there are
+some microbiota similarities between diarrheal and \emph{C. difficile}
+infection (CDI) communities such as lower diversity compared to healthy
+controls, which lead us to hypothesize that diarrhea may be an indicator
+of \emph{C. difficile} risk. We explored how osmotic laxatives disrupt
+the microbiota's colonization resistance to \emph{C. difficile} by
+administering a laxative to mice either before or after \emph{C.
+difficile} challenge Our findings suggest the osmotic laxative disrupts
+colonization resistance to \emph{C. difficile}, as well as clearance in
+mice already colonized with \emph{C. difficile}. Considering that most
+hospitals recommend not performing \emph{C. difficile} testing on
 patients taking laxatives and laxatives are used when administering
 fecal microbiota transplants via colonoscopy to patients with recurrent
-CDIs, further studies are needed to evaluate if laxatives impact human
-microbiota colonization resistance.
+CDIs, further studies are needed to evaluate if laxatives impact
+microbiota colonization resistance in humans.
 
@@
 for CDIs through epidemiological studies, whether the association is due
-to their impact on the microbiome is still unclear (5--9).
+to their impact on the microbiota is still unclear (5--9).
 
@@
 reduced acetate and butyrate production, altered the mucus barrier, and
-rendered the mice susceptible to \emph{C. difficile} infection (14--17).
-The mucus barrier is thought to mediate protection from \emph{C.
-difficile} infections by protecting intestinal epithelial cells from the
-toxins produced by \emph{C. difficile} (18, 19). Whether laxative
-administration results in more severe CDIs in mice and how long mice
-remain colonized with \emph{C. difficile} after challenge is unclear.
+rendered the mice susceptible to \emph{C. difficile} colonization
+(14--17). The mucus barrier is thought to mediate protection from CDIs
+by protecting intestinal epithelial cells from the toxins produced by
+\emph{C. difficile} (18, 19). Whether laxative administration results in
+more severe CDIs in mice and how long mice remain colonized with
+\emph{C. difficile} after challenge is unclear.
 
@@
 treatment results in more sustained \emph{C. difficile} colonization and
-severe CDI and than mice treated with clindamycin, and if PEG treatment
-after colonization can promote \emph{C. difficile} clearance.
+severe CDI than mice treated with clindamycin, and if PEG treatment
+after challenge can promote \emph{C. difficile} clearance.
 
@@
 post-challenge (9, 26). All three groups of PEG-treated mice were
-administered a 15\% PEG solution in the drinking water for 5-days, 1.
+administered a 15\% PEG solution in the drinking water for 5-days: 1.
 received no additional treatment, 2. was also treated with clindamycin,
@@
 the initial five days post-challenge (Fig. 1C, S1A). One mouse was found
-dead on the 6th day post-challenge, presumable due to \emph{C.
-difficile} as the bacteria became detecable in that mouse on the 4th day
-post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward,
-the density of \emph{C. difficile} stabilized in the 10-day recovery
-group and remained high through 20-30 days post-challenge (Fig. 1C).
-Thus, osmotic laxative treatment alone was sufficient to render mice
-susceptible to prolonged \emph{C. difficile} colonization and
-PEG-treated mice remained susceptible through ten days post-treatment.
+dead on the 6th day post-challenge, presumably due to \emph{C.
+difficile} as the bacteria became detectable in stool samples from that
+mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days
+post-challenge onward, the density of \emph{C. difficile} stabilized in
+the 10-day recovery group and remained high through 20-30 days
+post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was
+sufficient to render mice susceptible to prolonged \emph{C. difficile}
+colonization and PEG-treated mice remained susceptible through ten days
+post-treatment.
 
@@
 minimize the impact of cage effects on our experiment by breaking up
-cagemates when assigning mice to treatment groups. Importantly, although
-we conducted a total of 5 separate experiments, the experiment number
-and its interaction with treatment group was not one of the variables
-that significantly explained the observed variation in fecal communities
-(Data Set S1, Sheet X). None of the treatment groups recovered to their
-baseline community structure either 10 or 30 days post-challenge
-suggesting other community features besides recovery to baseline were
-responsible for the prolonged \emph{C. difficile} colonization in
-PEG-treated mice (Fig. 2B).
+littermates when assigning mice to treatment groups. Importantly,
+although we conducted a total of 5 separate experiments, the experiment
+number and its interaction with treatment group did no significantly
+explain the observed variation in fecal communities (Data Set S1, Sheet
+X). None of the treatment groups recovered to their baseline community
+structure either 10 or 30 days post-challenge suggesting other community
+features besides recovery to baseline were responsible for the prolonged
+\emph{C. difficile} colonization in PEG-treated mice (Fig. 2B).
 
@@
 detectable in the stool samples with the density stabilizing around 8
-days post-challenge (Fig. S1A). We found no bacteria were significantly
-different over the two time periods after multiple hypothesis correction
-(Data Set S1, Sheet X). However, there was also wide variation between
-individual mice regarding when \emph{C. difficile} became detectable
-(Fig. S1A) as well as the relative abundances of bacterial genera in the
-communities (Fig. S1B). For example, two mice had a high relative
-abundance of \emph{Enterobacteriaceae} throughout the post-challenge
-period and this corresponded to mouse 10, which died on the sixth day
-post-challenge and mouse 11, where \emph{C. difficile} was present at a
-high density from the 4th day post-challenge onward (Fig. S1B). While we
-did not identify a clear signal to explain the delayed appearance of
-\emph{C. difficile} in the 5-day PEG mice that were allowed to recover
-for 10 days prior to challenge, the delay is striking and could reflect
-changes in microbial activity or metabolites that were not examined in
-this study.
+days post-challenge (Fig. S1A). We found no bacteria with relative
+abundances that were significantly different over the two time points
+after multiple hypothesis correction (Data Set S1, Sheet X). However,
+there was also wide variation between individual mice regarding when
+\emph{C. difficile} became detectable (Fig. S1A) as well as the relative
+abundances of bacterial genera in the communities (Fig. S1B). For
+example, two mice had a high relative abundance of
+\emph{Enterobacteriaceae} throughout the post-challenge period and this
+corresponded to mouse 10, which died on the sixth day post-challenge and
+mouse 11, where \emph{C. difficile} was present at a high density from
+the 4th day post-challenge onward (Fig. S1B). While we did not identify
+a clear signal to explain the delayed appearance of \emph{C. difficile}
+in the 5-day PEG mice that were allowed to recover for 10 days prior to
+challenge, the delay is striking and could reflect changes in microbial
+activity or metabolites that were not examined in this study.
 
@@
 PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha
-diversity continued to increase over time with the PEG-treated mice
-collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet
-X). Group, time point, and their interactions with other variables
-(cage, experiment number, and sample type) explained the majority of the
-variation observed in mucosal communities (PERMANOVA combined
-R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
-Set S1, Sheet X). We saw the greatest difference in the relative
-abundance of the mucosal microbiota between treatment groups
+diversity continued to increase over time based on the tissues from
+PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A,
+Data Set S1, Sheet X). Group, time point, and their interactions with
+other variables (cage, experiment number, and sample type) explained the
+majority of the variation observed in mucosal communities (PERMANOVA
+combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
+3B, Data Set S1, Sheet X). We saw the greatest difference in the
+relative abundance of the mucosal microbiota between treatment groups
 (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days
 post-challenge with 10 genera that were significantly different
-(\emph{P} `r format.pval(tissue\_genera\_max\_p, eps = 0.05)) in all
-three of the tissue types we collected (cecum, proximal colon, and
-distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly,
-\emph{Peptostreptococcaceae} (the family with a sequence that matches
-\emph{C. difficile}) was one of the genera that had a significant
-difference in relative abundance between treatment groups at 6 days
-post-challenge. This population was primarily only present in the 5-day
-PEG treatment group of mice and decreased in the proximal and distal
-colon tissues over time (Fig. S2B). By 30 days post-challenge, only the
-relative abundances of \emph{Bacteroides}, \emph{Clostridiales},
-\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
-treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data
-Set S1, Sheet X). Thus, PEG treatment had a significant impact on the
-mucosal microbiota and we detected \emph{C. difficile} sequences in the
-cecum, proximal colon, and distal colon tissue communities.
-
-Next, we examined the severity of \emph{C. difficile} challenge by
-evaluating cecum and colon H\&E stained histopathology (28) and found
-there was no difference in cecum and colon scores between clindamycin
-and PEG-treated mice that were challenged with \emph{C. difficile} at 4
-days post-challenge (Fig. 3D), the time point typically examined in
-\emph{C. difficile} 630 challenged mice (29). We also looked at 6 days
+(\emph{P} \textless{} 0.05) in all three of the tissue types we
+collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set
+S1, Sheet X). Interestingly, \emph{Peptostreptococcaceae} (the family
+with a sequence that matches \emph{C. difficile}) was one of the genera
+that had a significant difference in relative abundance between
+treatment groups at 6 days post-challenge. This population was primarily
+only present in the 5-day PEG treatment group of mice and decreased in
+the proximal and distal colon tissues over time (Fig. S2B). By 30 days
+post-challenge, only the relative abundances of \emph{Bacteroides},
+\emph{Clostridiales}, \emph{Firmicutes}, and \emph{Ruminococcaceae} were
+different between treatment groups and only in the cecum tissues (Fig.
+3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a
+significant impact on the mucosal microbiota and we detected \emph{C.
+difficile} sequences in the cecum, proximal colon, and distal colon
+tissue communities.
+
+Because there were differences in the mucosal microbiota including
+detectable \emph{C. difficile} sequences in tissues from PEG-treated
+mice relative to mice treated with clindamycin, we next examined the
+severity of \emph{C. difficile} challenge by evaluating cecum and colon
+H\&E stained histopathology (28). However, we found there was no
+difference in cecum and colon scores between clindamycin and PEG-treated
+mice that were challenged with \emph{C. difficile} at 4 days
+post-challenge (Fig. 3D), the time point typically examined in \emph{C.
+difficile} 630 challenged mice (29). We also looked at 6 days
 post-challenge because that was when there was a large difference in
@@
 
-\textbf{Post-CDI laxative treatment disrupted clearance in
+\textbf{Post-challenge laxative treatment disrupted clearance in
 clindamycin-treated mice regardless of whether an FMT was also
@@
 that received PEG exhibited prolonged \emph{C. difficile} colonization
-(Fig. 5B). Next we examined how post-CDI PEG treatment impacted
-microbiota diversity. Alpha diversity was lower in the PEG-treated mice
-with the exception of the PEG-treated mice that were administered an FMT
-(Fig 5C-D).
+(Fig. 5B).
 
@@
 that will have prolonged \emph{C. difficile} colonization.} After
-identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day
-PEG treatments, we decided to examine the bacteria that influenced
+identifying bacteria associated with the 5-day, 1-day and post-challenge
+1-day PEG treatments, we decided to examine the bacteria that influenced
 prolonged \emph{C. difficile} colonization. We trained 3 types of
@@
 because that was the earliest time point where we would see a treatment
-effect in the mice that were given 1-day PEG treatment three days after
-challenge. The random forest model had the highest performance (AUROC =
-0.90, Data Set S1, Sheet X), so we next performed permutation importance
-to examine the bacteria that were the top contributors to the random
-forest model predicting prolonged \emph{C. difficile }colonization. We
-selected the top 10 bacteria contributing to our models performance
-(Fig. 7A) and examined their relative abundance at 5 days
-post-challenge, the time point used to predict \emph{C. difficile}
-colonization status on day 10 (Fig. 7B). Next, we focused on the 5
-genera that had a greater than 1 \% relative abundance in either the
-cleared or colonized mice and examined how the bacteria changed over
-time. We found \emph{Enterobacteriaceae} and \emph{Bacteroides} tended
-to have a higher relative abundance, the relative abundance of
-\emph{Akkermansia} was initially decreased and then increased, and
-\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} had a lower
-relative abundance in the mice with prolonged colonization compared to
-the mice that cleared \emph{C. difficile} (Fig. 7C). Together these
-results suggest a combination of low and high abundance bacterial genera
-influence the prolonged colonization observed in 5-day PEG and post-CDI
-PEG treated mice.
-
-Previous work examining the impact of PEG on the murine microbiota found
-that PEG treatment resulted in the permanent loss of \emph{Muribaculum
-intestinale} (16)., which is classified as \emph{Porphyromonadaceae} by
-the Ribosomal Database Project (RDP) database (31).
-\emph{Porphyromonadaceae} was a top feature in the random forest model
-predicting prolonged \emph{C. difficile} colonization. We identified 4
-OTUs that had 92-96\% identity to \emph{Muribaculum intestinale} and
-examined their abundance in mice that either cleared or were still
-colonized with \emph{C. difficile} at 10 days post-challenge. While all
-of the OTUs, were decreased by PEG and clindamycin treatment, there was
-some recovery in the mice that cleared (Fig. S4A). We also examined
-other \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs since
-these were the 2 genera that were important to our classification model
-and contained multiple OTUs that were different at 5 days post-challenge
-between mice that either cleared or remained colonized with \emph{C.
-difficile} by 10 days post-challenge (Data Set S1, sheet X). While
-individual \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs
-tended to be more abundant in the mice that clear \emph{C. difficile}
-relative to the mice that exhibit prolonged colonization (Fig. S4B-C),
-there is no single OTU that fits the pattern we observed at the genus
-level (Fig. 7C), suggesting multiple \emph{Porphyromonadaceae} and
-\emph{Lachnospiraceae} OTUs influenced \emph{C. difficile} clearance.
-Overall, our results suggest that specific bacterial community
-differences explain the prolonged \emph{C. difficile} colonization we
-observed in 5-day PEG and post-CDI 1-day PEG treated mice.
+effect in the post-challenge 1-day PEG experiments. The random forest
+model had the highest performance (median AUROC = 0.90, Data Set S1,
+Sheet X), so we next performed permutation importance to examine the
+bacteria that were the top contributors to the random forest model
+predicting prolonged \emph{C. difficile }colonization. We selected the
+top 10 bacteria contributing to our model's performance (Fig. 7A) and
+examined their relative abundance at 5 days post-challenge, the time
+point used to predict \emph{C. difficile} colonization status on day 10
+(Fig. 7B). Next, we focused on the 5 genera that had a greater than 1 \%
+relative abundance in either the cleared or colonized mice and examined
+how the bacteria changed over time. We found \emph{Enterobacteriaceae}
+and \emph{Bacteroides} tended to have a higher relative abundance, the
+relative abundance of \emph{Akkermansia} was initially decreased and
+then increased, and \emph{Porphyromonadaceae} and \emph{Lachnospiraceae}
+had a lower relative abundance in the mice with prolonged colonization
+compared to the mice that cleared \emph{C. difficile} (Fig. 7C).
+Together these results suggest a combination of low and high abundance
+bacterial genera influence the prolonged colonization observed in 5-day
+PEG and post-challenge 1-day PEG treated mice.
 
@@
 
-\begin{itemize}
-\item
-  Summary of major findings (Fig. 8A)
-\item
-  Discussion of prolonged persistence. \emph{C. difficile} sequences
-  detected in tissue samples. Association with mucin-degrading bacteria
-  suggested by recent papers.
-\item
-  Discuss why we might not have observed more severe histology in PEG
-  mice relative to clindamycin-treated mice
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Antibiotics may also impact mucus layer
-  \item
-    Strain of bacteria used
-  \end{itemize}
-\item
-  Protective bacteria missing in PEG-treated mice
-\item
-  Discuss what these findings might mean for human patients (Fig. 8B)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    What's known regarding laxatives and susceptibility to CDIs
-  \item
-    Clinical trial of PEG, results never posted (32)
-  \item
-    Relevance to human FMTs? Unclear what the best administration route
-    is because there have been no studies designed to evaluate the best
-    administration route for FMTs.
-  \end{itemize}
-\end{itemize}
-
-\hypertarget{conclusions}{%
-\subsection{Conclusions}\label{conclusions}}
+While the disruptive effect of antibiotics on \emph{C. difficile}
+colonization resistance is well established, the extent to which other
+drugs such as laxatives disrupt colonization resistance was unclear. By
+following osmotic laxative treated mice over time, we found 5-day PEG
+treatment before challenge resulted in prolonged \emph{C. difficile}
+colonization, while a 1-day PEG treatment resulted in transient
+colonization. The differences in \emph{C. difficile} colonization
+dynamics between the 5- and 1-day PEG treated mice were associated with
+differences in how much the treatments disrupted the microbiota.
+Additionally, the intestinal communities of 5-day PEG treated mice had
+not regained colonization resistance after a 10-day recovery period.
+Although in contrast to the other 5-day PEG treatment groups, \emph{C.
+difficile} was not immediately detectable in the stools of most of the
+mice in the 10-day recovery group. We also examined the impact of PEG
+treatment after \emph{C. difficile} challenge and in opposition to the
+hypothesis suggested by the literature, found that PEG treatment
+prolonged colonization relative to mice that only recieved clindamycin
+treatment. We identified patterns in the relative abundances of
+\emph{Bacteroides}, \emph{Enterobacteriaceae}, \emph{Akkermansia},
+\emph{Porphyromonadaceae}, and \emph{Lachnospiraceae} that were
+associated with prolonged \emph{C. difficile} colonization (Fig. 8).
+Overall, our results demonstrated that osmotic laxative treatment alone
+rendered mice susceptible to \emph{C. difficile} colonization and the
+duration of colonization depended on the length of PEG treatment and
+whether treatment was administered before or after challenge.
+
+In addition to altering composition, laxative treatment may alter
+microbiota-produced metabolites. A previous study demonstrated that a
+5-day treatment of 10\% PEG depleted acetate and butyrate and increased
+succinate compared to untreated mice (15). While we did not perform
+metabolomic analysis, we did see bacteria known to produce benefical
+metabolites were depleted in mice that cleared \emph{C. difficile}
+compared to mice with prolonged colonization. For example,
+\emph{Oscillibacter valericigenes} can produce the SCFA valerate (31),
+and separate studies demonstrated valerate is depleted after clindamycin
+treatment and inhibited \emph{C. difficile} growth based on \emph{in
+vitro} data and an experiment with C57BL/6 mice (32, 33). Similarly,
+\emph{Acetatifactor} can produce acetate and butyrate (34), SCFAs that
+are decreased in mice with prolonged \emph{C. difficile} infection after
+antibiotic treatment (35). Thus protective bacteria and their
+metabolites could be depleted by osmotive laxative treatment depending
+on the timing and duration of treatment.
+
+One possible explanation for the prolonged \emph{C. difficile}
+colonization in 5-day PEG treated mice, might be due to the bacteria's
+persistence in the mucosal compartment. In fact, it has been
+hypothesized that \emph{C. difficile} biofilms may serve as reservoirs
+for recurrent infections (36) and \emph{C. difficile} biofilms were
+recently identified in patients and aggregates with \emph{Fusobacterium
+nucleatum}, another species capable of forming biofilms (37). There was
+an interesting pattern of increased \emph{Enterobacteriaceae},
+\emph{Bacteroides}, and \emph{C. difficile} in both the stool and
+mucosal communities of PEG-treated mice suggesting a potential synergy.
+\emph{Bacteroides} has the potential to degrade mucus and the osmotic
+laxative may have allowed \emph{Bacteroides} to colonize the mucosal
+niche by degrading mucin glycans with glycosyl hydrolases that are
+absent in \emph{C. difficile} (38). \emph{Bacteroides} persistent in the
+mucosal tissue might also have helped \emph{Enterobacteriaceae} to
+colonize the region, as a synergy between mucus-degrading \emph{B.
+fragilis} and \emph{E. coli} has previously been described (39). A
+separate study demonstrated \emph{C. difficile} was present in the outer
+mucus layer and associated with \emph{Enterobacteriaceae} and
+\emph{Bacteroidaceae} using fluorescent in situ hybridization (FISH)
+staining (40). However, \emph{B. fragilis} prevented CDI morbidity in a
+mouse model and inhibited \emph{C. difficile} adherence \emph{in vitro}
+(41). In coculture experiments \emph{B. longum} decreased \emph{C.
+difficile} biofilm formation while \emph{B. thetaiotamicron} enhanced
+biofilm formation (42). Therefore, whether \emph{Bacterodes} is
+detrimental or beneficial in the context of \emph{C. difficile}
+infection or colonization is still unclear.
+
+\emph{Akkermansia} is also a mucin degrader with potentially beneficial
+or detrimental roles depending on context in other diseases (43, 44). In
+our study the relative abundance of \emph{Akkermansia} shifted over time
+between groups. In the stool it was initially increased in mice that
+cleared \emph{C. difficile}, but shifted after 5-days post-challenge so
+that it was increased in mice that had prolonged colonization. In the
+context of CDIs, some studies suggest a protective role (45, 46), while
+others suggest detrimental (47--49). Because the relative abundance of
+\emph{Akkermansia} was dynamic in our study so it is unclear whether
+\emph{Akkermansia} helps with clearance of \emph{C. difficile} or allows
+it to persist. A better understanding how \emph{C. difficile} interacts
+with the mucosal microbiota may lead to insights into CDIs, asymptomatic
+\emph{C. difficile} colonization, and resistance.
+
+Despite identifying an altered compositional profile that included
+higher relative abundance of the \emph{C. difficile} sequence in the
+mucosal tissues of mice treated with 5-day PEG compared to the
+clindamycin group, we did not see a difference in histopathology scores
+between the groups. One reason there was no difference could be the
+\emph{C. difficile} strain used, \emph{C. difficile} 630 results in mild
+histopathology summary scores in mice compared to VPI 10463 despite both
+strains producing toxin in mice (50). Part of our hypothesis for why
+there could have been increased histopathology scores in PEG-treated
+mice was because PEG was previously shown to disrupt the mucus layer in
+mice. However, recent studies demonstrated that broad spectrum
+antibiotics can also disrupt the host mucosal barrier in mice (51, 52).
+Future research is needed to tease out the interplay between medications
+that influence the mucus layer and different strains of \emph{C.
+difficile} in the context of CDIs.
+
+It is more difficult what are findings mean in the context of \emph{C.
+difficile} colonization resistance mean for human patients. Considering
+that most hospitals recommend not performing \emph{C. difficile} testing
+on patients taking laxatives and laxatives are used when administering
+fecal microbiota transplants via colonoscopy to patients with recurrent
+CDIs. further studies are needed to evaluate if laxatives impact human
+microbiota colonization resistance. Further studies are needed to
+understand the impact of osmotic laxatives on \emph{C. difficile}
+colonization resistance and clearance in human patients. + What's known
+regarding laxatives and susceptibility to CDIs + Clinical trial of PEG,
+results never posted (53)
+
+Relevance to human FMTs? Unclear what the best administration route is
+because there have been no studies designed to evaluate the best
+administration route for FMTs. However, results from the FMT National
+Registry where 85\% of FMTs were delivered by colonoscopy demonstrate
+FMTs are highly effective treatment for recurrent CDIs with 90\%
+achieving resolution in the 1 month follow-up window (54).
+
+We have demonstrated that osmotic laxative treatment alone has a
+substantial impact on the microbiota and rendered mice susceptible to
+prolonged \emph{C. difficile} colonization in contrast to
+clindamycin-treated mice. The duration and timing of the laxative
+treatment impacted the duration of \emph{C. difficile} colonization,
+with only 5-day PEG and post-challenge 1-day PEG treatments prolonging
+colonization compared to clindamycin treated mice. Further studies are
+warranted to ascertain whether laxatives have a similar impact on
+\emph{C. difficile} colonization resistance on the human microbiota.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Histopathology (33)
-\end{itemize}
+\textbf{Animals.} All experiments were approved by the University of
+Michigan Animal Care and Use Committee IACUC (protocol numbers
+PRO00006983 and PRO00008975). All mice were C57Bl/6 and part of the
+Schloss lab colony which was established in 2010 with mice donated from
+Vincent Young's lab colony (established with mice purchased from The
+Jackson Laboratory in 2002). We used 7-19 week old female mice for all
+experiments, which allowed us to break up littermates prior to starting
+an experiment and distribute them as evenly as possible across treatment
+groups in order to minimize microbiota differences between experiment
+groups prior to starting treatments with medications. During the
+experiment, mice were housed at a density of 2-3 mice per cage, with the
+majority of cages limited to two mice.
+
+\textbf{Drug treatments.} For PEG treament groups, fifteen percent PEG
+3350 (Miralax) was administered in the drinking water for either 5 or
+1-day periods depending on the experiment. PEG solution was prepared
+fresh every 2 days in distilled water and administered to the mice in
+water bottles. Clindamycin treatment groups received distilled water in
+water bottles during the PEG-treatment periods, with the water being
+changed at the same frequency. For clindamycin treatment, groups of mice
+received 10 mg/kg clindamycin (Sigma-Aldrich) via intraperitoneal
+injection. All PEG treatment groups received a sham intraperitoneal
+injection containing filter sterilized saline.
+
+\textbf{\emph{C. difficile} challenge model.} Mice were challenged with
+25 microliters off \emph{C. difficile} 630 spores at
+10\textsuperscript{5} concentration, except for 1 experiment (Fig. 5A)
+where the concentration was 10\textsuperscript{3}. All mock challenged
+mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper
+pipette was used to administer the same challenge dose to mice via oral
+gavage. Mice were weighed daily throughout the experiment and stool was
+collected for quantifying \emph{C. difficile} CFU and 16S rRNA gene
+sequencing. There were two groups of mice that received either a PBS or
+fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal
+microbiota transplant was prepared with stool samples collected from the
+mice in the experiment prior to the start of any treatments. The stool
+samples were transferred to an anaerobic chamber and diluted 1:10 in
+reduced PBS and glycerol was added to make a 15\% glycerol solution. The
+solution was then aliquoted into tubes and stored at -80°C until the day
+of the gavage. An aliquot of both the FMT and PBS solutions were also
+set aside in the -80°C for 16S rRNA gene sequencing. The day of the
+gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute.
+The supernatant was then transferred to a separate tube to prevent the
+gavage needle from clogging with debris during gavage. The PBS solution
+that was administered to the other group was also 15\% glycerol. Each
+mouse was administered 100 microliters of either the FMT or PBS solution
+via gavage. When we refer to mice that cleared \emph{C. difficile}, we
+mean that no \emph{C. difficile} was detected in the first serial
+dilution (limit of detection: 100 CFU). In some experiments, we
+collected tissues for 16SrRNA gene sequencing, histopathology, or both.
+For 16S rRNA gene sequencing, we collected small snips of cecum,
+proximal colon, and distal colon tissues in microcentrifuge tubes, snap
+froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
+and colon tissues were placed into separate casettes, fixed, and then
+submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,
+embedding, and hematoxylin and eosin (H\&E) staining.
+
+\textbf{\emph{C. difficile} quantification.} Stool samples from mice
+were transferred to an anaerobic chamber and serially diluted in reduced
+PBS. Serial dilutions were plated onto
+taurocholate-cycloserine-cefoxitin-fructose agar (TCCFA) plates plates
+and counted after 24 hours of incubation at 37°C. Stool samples
+collected from the mice on day 0 post-challenge were also plated onto
+TCCFA plates to ensure mice were not already colonized with \emph{C.
+difficile} prior to challenge.
+
+\textbf{16S rRNA gene sequencing.} Stool samples that were stored in the
+-80°C were placed into 96-well plates for DNA extractions and library
+preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit
+(Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
+For library preparation, each plate had a mock community control
+(ZymoBIOMICS microbial community DNA standards) and a negative control
+(water). The V4 region of the 16S rRNA gene was amplified with the
+AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom
+barcoded primers, as previously described (55). The PCR amplicons were
+normalized (SequalPrep normalizatin plate kit from Thermo Fisher
+Scientific), pooled and quantified (KAPA library quantification kit from
+KAPA Biosystems), and sequenced with the MiSeq system (Illumina).
+
+\textbf{16S rRNA gene sequence analysis.} All sequences were processed
+with mothur (v. 1.43) using a previously published protocol (55, 56).
+Paired sequencing reads were combined and aligned with the SILVA (v.
+132) reference database (57) and taxonomy was assigned with a modified
+version (v. 16) of the Ribosomal Database Project (v. 11.5) (58). The
+error rate for are sequencing data was 0.0559\% based on the 17 mock
+communities we ran with the samples. Samples were rarefied to 1,000
+sequences, 1,000 times for alpha and beta diversity analyses in order to
+account for uneven sequencing across samples. All but 3 out of 17 water
+controls had less than 1000 sequences. PCoAs were generated based on
+Bray-Curtis Index distance matrices. Permutational multivariate analysis
+of variance (PERMANOVA) tests were performed on mothur-generated
+Bray-Curtis distance matrices with the adonis function from the vegan R
+package (59).
+
+\textbf{Histopathology.} H\&E stained sections of cecum and colon
+tissues collected at either 0, 4, or 6 days post-challenge were coded to
+be scored in a blinded manner by a board-certified veterinary
+pathologist (ILB). Slides were evaluated using a scoring system
+developed for mouse models of \emph{C. difficile} infection (50). Each
+slide was evaluated for edema, cellular infiltration, and inflammation
+and given a score ranging from 0-4. The summary score was calculated by
+combining the scores from the 3 categories and ranged from 0-12.
+
+\textbf{Classification model training and evaluation.} We used the
+mikropml package to train and evaluate models to predict \emph{C.
+difficile} colonization status 10 days post-challenge where mice were
+categorized as either cleared or colonized (60, 61). We removed the
+\emph{C. difficile} genus relative abundance data prior to training the
+model. Input community relative abundance data at the genus level from 5
+days post-challenge was used to generate random forest, logistic
+regression, and support vector machine classification models to predict
+\emph{C. difficile} colonization status 10 days post-challenge. To
+accommodate the small number of samples in our data set we used 50\%
+training and 50\% testing splits with repeated 2-fold cross-validation
+of the training data for hyperparamter tuning. Permutation importance
+was performed as described previously (62) using mikropml (60, 61) with
+the random forest model because it had the highest AUROC value.
+
+\textbf{Statistical analysis.} R (v. 4.0.2) and the tidyverse package
+(v. 1.3.0) were used for statistical analysis (63, 64). Kruskal-Wallis
+tests with Bejamini-Hochberg correction for testing multiple time points
+were used to analyze differences in \emph{C. difficile} CFU, mouse
+weight change, and alpha diversity between treatment groups. Paired
+Wilcoxon rank signed rank tests were used to identify genera impacted by
+treatments on matched pairs of samples from 2 time points. Bacterial
+relative abundances that varied between treatment groups at the genus
+level were identified with the Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing all identified OTUs, followed
+by pairwise Wilcoxon comparisons with Benjamini-Hocherg correction.
+
+\textbf{Code availability.} Code for data analysis and generating this
+paper with accompanying figures is available at
+\url{https://github.com/SchlossLab/Tomkovich_PEG3350_XXXX_2021}.
+
+\textbf{Data availability.} The 16S rRNA sequencing data have been
+deposited in the National Center for Biotechnology Information Sequence
+Read Archive (BioProject Accession no. PRJNA727293).
+
+\newpage
+
+\hypertarget{references}{%
+\subsection{References}\label{references}}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Britton2014}{}%
+1. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
+microbiota in resistance to colonization by clostridium difficile.
+Gastroenterology \textbf{146}:1547--1553.
+doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
+
+\leavevmode\hypertarget{ref-Maier2018}{}%
+2. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
+\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
+\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
+\textbf{Mori H}, \textbf{Patil KR}, \textbf{Bork P}, \textbf{Typas A}.
+2018. Extensive impact of non-antibiotic drugs on human gut bacteria.
+Nature \textbf{555}:623--628.
+doi:\href{https://doi.org/10.1038/nature25979}{10.1038/nature25979}.
+
+\leavevmode\hypertarget{ref-LeBastard2017}{}%
+3. \textbf{Bastard QL}, \textbf{Al-Ghalith GA}, \textbf{Grégoire M},
+\textbf{Chapelet G}, \textbf{Javaudin F}, \textbf{Dailly E},
+\textbf{Batard E}, \textbf{Knights D}, \textbf{Montassier E}. 2017.
+Systematic review: Human gut dysbiosis induced by non-antibiotic
+prescription medications. Alimentary Pharmacology \& Therapeutics
+\textbf{47}:332--345.
+doi:\href{https://doi.org/10.1111/apt.14451}{10.1111/apt.14451}.
+
+\leavevmode\hypertarget{ref-VichVila2020}{}%
+4. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
+T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
+\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
+\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
+\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
+composition and metabolic function of the gut microbiota. Nature
+Communications \textbf{11}.
+doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
+
+\leavevmode\hypertarget{ref-Oh2018}{}%
+5. \textbf{Oh J}, \textbf{Makar M}, \textbf{Fusco C}, \textbf{McCaffrey
+R}, \textbf{Rao K}, \textbf{Ryan EE}, \textbf{Washer L}, \textbf{West
+LR}, \textbf{Young VB}, \textbf{Guttag J}, \textbf{Hooper DC},
+\textbf{Shenoy ES}, \textbf{Wiens J}. 2018. A generalizable, data-driven
+approach to predict daily risk of clostridium difficile infection at two
+large academic health centers. Infection Control \& Hospital
+Epidemiology \textbf{39}:425--433.
+doi:\href{https://doi.org/10.1017/ice.2018.16}{10.1017/ice.2018.16}.
+
+\leavevmode\hypertarget{ref-Mora2012}{}%
+6. \textbf{Mora AL}, \textbf{Salazar M}, \textbf{Pablo-Caeiro J},
+\textbf{Frost CP}, \textbf{Yadav Y}, \textbf{DuPont HL}, \textbf{Garey
+KW}. 2012. Moderate to high use of opioid analgesics are associated with
+an increased risk of clostridium difficile infection. The American
+Journal of the Medical Sciences \textbf{343}:277--280.
+doi:\href{https://doi.org/10.1097/maj.0b013e31822f42eb}{10.1097/maj.0b013e31822f42eb}.
+
+\leavevmode\hypertarget{ref-Nehra2018}{}%
+7. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
+\textbf{Nehra V}. 2018. Proton pump inhibitors: Review of emerging
+concerns. Mayo Clinic Proceedings \textbf{93}:240--246.
+doi:\href{https://doi.org/10.1016/j.mayocp.2017.10.022}{10.1016/j.mayocp.2017.10.022}.
+
+\leavevmode\hypertarget{ref-Krishna2013}{}%
+8. \textbf{Krishna SG}, \textbf{Zhao W}, \textbf{Apewokin SK},
+\textbf{Krishna K}, \textbf{Chepyala P}, \textbf{Anaissie EJ}. 2013.
+Risk factors, preemptive therapy, and antiperistaltic agents
+forClostridium difficileinfection in cancer patients. Transplant
+Infectious Disease n/a--n/a.
+doi:\href{https://doi.org/10.1111/tid.12112}{10.1111/tid.12112}.
+
+\leavevmode\hypertarget{ref-Tomkovich2019}{}%
+9. \textbf{Tomkovich S}, \textbf{Lesniak NA}, \textbf{Li Y},
+\textbf{Bishop L}, \textbf{Fitzgerald MJ}, \textbf{Schloss PD}. 2019.
+The proton pump inhibitor omeprazole does not promote
+\emph{Clostridioides difficile} colonization in a murine model. mSphere
+\textbf{4}.
+doi:\href{https://doi.org/10.1128/msphere.00693-19}{10.1128/msphere.00693-19}.
+
+\leavevmode\hypertarget{ref-Vandeputte2015}{}%
+10. \textbf{Vandeputte D}, \textbf{Falony G}, \textbf{Vieira-Silva S},
+\textbf{Tito RY}, \textbf{Joossens M}, \textbf{Raes J}. 2015. Stool
+consistency is strongly associated with gut microbiota richness and
+composition, enterotypes and bacterial growth rates. Gut
+\textbf{65}:57--62.
+doi:\href{https://doi.org/10.1136/gutjnl-2015-309618}{10.1136/gutjnl-2015-309618}.
+
+\leavevmode\hypertarget{ref-VujkovicCvijin2020}{}%
+11. \textbf{Vujkovic-Cvijin I}, \textbf{Sklar J}, \textbf{Jiang L},
+\textbf{Natarajan L}, \textbf{Knight R}, \textbf{Belkaid Y}. 2020. Host
+variables confound gut microbiota studies of human disease. Nature
+\textbf{587}:448--454.
+doi:\href{https://doi.org/10.1038/s41586-020-2881-9}{10.1038/s41586-020-2881-9}.
+
+\leavevmode\hypertarget{ref-Schubert2014}{}%
+12. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
+\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
+distinguish patients with clostridium difficile infection and non-c.
+Difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
+doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
+
+\leavevmode\hypertarget{ref-Nagata2019}{}%
+13. \textbf{Nagata N}, \textbf{Tohya M}, \textbf{Fukuda S}, \textbf{Suda
+W}, \textbf{Nishijima S}, \textbf{Takeuchi F}, \textbf{Ohsugi M},
+\textbf{Tsujimoto T}, \textbf{Nakamura T}, \textbf{Shimomura A},
+\textbf{Yanagisawa N}, \textbf{Hisada Y}, \textbf{Watanabe K},
+\textbf{Imbe K}, \textbf{Akiyama J}, \textbf{Mizokami M},
+\textbf{Miyoshi-Akiyama T}, \textbf{Uemura N}, \textbf{Hattori M}. 2019.
+Effects of bowel preparation on the human gut microbiome and metabolome.
+Scientific Reports \textbf{9}.
+doi:\href{https://doi.org/10.1038/s41598-019-40182-9}{10.1038/s41598-019-40182-9}.
+
+\leavevmode\hypertarget{ref-Kashyap2013}{}%
+14. \textbf{Kashyap PC}, \textbf{Marcobal A}, \textbf{Ursell LK},
+\textbf{Larauche M}, \textbf{Duboc H}, \textbf{Earle KA},
+\textbf{Sonnenburg ED}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
+\textbf{Million M}, \textbf{Tache Y}, \textbf{Pasricha PJ},
+\textbf{Knight R}, \textbf{Farrugia G}, \textbf{Sonnenburg JL}. 2013.
+Complex interactions among diet, gastrointestinal transit, and gut
+microbiota in humanized mice. Gastroenterology \textbf{144}:967--977.
+doi:\href{https://doi.org/10.1053/j.gastro.2013.01.047}{10.1053/j.gastro.2013.01.047}.
+
+\leavevmode\hypertarget{ref-Ferreyra2014}{}%
+15. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
+\textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
+Gut microbiota-produced succinate promotes c.~difficile infection after
+antibiotic treatment or motility disturbance. Cell Host \& Microbe
+\textbf{16}:770--777.
+doi:\href{https://doi.org/10.1016/j.chom.2014.11.003}{10.1016/j.chom.2014.11.003}.
+
+\leavevmode\hypertarget{ref-Tropini2018}{}%
+16. \textbf{Tropini C}, \textbf{Moss EL}, \textbf{Merrill BD},
+\textbf{Ng KM}, \textbf{Higginbottom SK}, \textbf{Casavant EP},
+\textbf{Gonzalez CG}, \textbf{Fremin B}, \textbf{Bouley DM},
+\textbf{Elias JE}, \textbf{Bhatt AS}, \textbf{Huang KC},
+\textbf{Sonnenburg JL}. 2018. Transient osmotic perturbation causes
+long-term alteration to the gut microbiota. Cell
+\textbf{173}:1742--1754.e17.
+doi:\href{https://doi.org/10.1016/j.cell.2018.05.008}{10.1016/j.cell.2018.05.008}.
+
+\leavevmode\hypertarget{ref-VanInsberghe2020}{}%
+17. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
+\textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
+Diarrhoeal events can trigger long-term clostridium difficile
+colonization with recurrent blooms. Nature Microbiology
+\textbf{5}:642--650.
+doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.
+
+\leavevmode\hypertarget{ref-Olson2013}{}%
+18. \textbf{Olson A}, \textbf{Diebel LN}, \textbf{Liberati DM}. 2013.
+Effect of host defenses on clostridium difficile toxininduced intestinal
+barrier injury. Journal of Trauma and Acute Care Surgery
+\textbf{74}:983--990.
+doi:\href{https://doi.org/10.1097/ta.0b013e3182858477}{10.1097/ta.0b013e3182858477}.
+
+\leavevmode\hypertarget{ref-Diebel2014}{}%
+19. \textbf{Diebel LN}, \textbf{Liberati DM}. 2014. Reinforcement of the
+intestinal mucus layer protects against clostridium difficile intestinal
+injury in~vitro. Journal of the American College of Surgeons
+\textbf{219}:460--468.
+doi:\href{https://doi.org/10.1016/j.jamcollsurg.2014.05.005}{10.1016/j.jamcollsurg.2014.05.005}.
+
+\leavevmode\hypertarget{ref-vanNood2013}{}%
+20. \textbf{Nood E van}, \textbf{Vrieze A}, \textbf{Nieuwdorp M},
+\textbf{Fuentes S}, \textbf{Zoetendal EG}, \textbf{Vos WM de},
+\textbf{Visser CE}, \textbf{Kuijper EJ}, \textbf{Bartelsman JFWM},
+\textbf{Tijssen JGP}, \textbf{Speelman P}, \textbf{Dijkgraaf MGW},
+\textbf{Keller JJ}. 2013. Duodenal infusion of donor feces for
+RecurrentClostridium difficile. New England Journal of Medicine
+\textbf{368}:407--415.
+doi:\href{https://doi.org/10.1056/nejmoa1205037}{10.1056/nejmoa1205037}.
+
+\leavevmode\hypertarget{ref-Razik2017}{}%
+21. \textbf{Razik R}, \textbf{Osman M}, \textbf{Lieberman A},
+\textbf{Allegretti JR}, \textbf{Kassam Z}. 2017. Faecal microbiota
+transplantation for clostridium difficile infection: A multicentre study
+of non-responders. Medical Journal of Australia \textbf{207}:159--160.
+doi:\href{https://doi.org/10.5694/mja16.01452}{10.5694/mja16.01452}.
+
+\leavevmode\hypertarget{ref-Postigo2012}{}%
+22. \textbf{Postigo R}, \textbf{Kim JH}. 2012. Colonoscopic versus
+nasogastric fecal transplantation for the treatment of clostridium
+difficile infection: A review and pooled analysis. Infection
+\textbf{40}:643--648.
+doi:\href{https://doi.org/10.1007/s15010-012-0307-9}{10.1007/s15010-012-0307-9}.
+
+\leavevmode\hypertarget{ref-Liacouras1996}{}%
+23. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
+irrigation as an adjunct to the treatment of chronic, relapsing
+clostridium difficile colitis. Journal of Clinical Gastroenterology
+\textbf{22}:186--189.
+doi:\href{https://doi.org/10.1097/00004836-199604000-00007}{10.1097/00004836-199604000-00007}.
+
+\leavevmode\hypertarget{ref-Schubert2015}{}%
+24. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+Antibiotic-induced alterations of the murine gut microbiota and
+subsequent effects on colonization resistance against \emph{Clostridium
+difficile}. mBio \textbf{6}.
+doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+
+\leavevmode\hypertarget{ref-Jenior2017}{}%
+25. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
+alternative nutrient niches during infection across distinct murine gut
+microbiomes. mSystems \textbf{2}.
+doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
+
+\leavevmode\hypertarget{ref-Tomkovich2020}{}%
+26. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
+antibiotic perturbation influence clostridioides difficile clearance in
+mice. mSphere \textbf{5}.
+doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
+
+\leavevmode\hypertarget{ref-Nguyen2015}{}%
+27. \textbf{Nguyen TLA}, \textbf{Vieira-Silva S}, \textbf{Liston A},
+\textbf{Raes J}. 2015. How informative is the mouse for human gut
+microbiota research? Disease Models \& Mechanisms \textbf{8}:1--16.
+doi:\href{https://doi.org/10.1242/dmm.017400}{10.1242/dmm.017400}.
+
+\leavevmode\hypertarget{ref-Reeves2011}{}%
+28. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
+interplay between microbiome dynamics and pathogen dynamics in a murine
+model of \emph{Clostridium difficile} infection \textbf{2}:145--158.
+doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.
+
+\leavevmode\hypertarget{ref-Dieterle2020}{}%
+29. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
+\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
+\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
+K}. 2020. Systemic inflammatory mediators are effective biomarkers for
+predicting adverse outcomes in clostridioides difficile infection. mBio
+\textbf{11}.
+doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
+
+\leavevmode\hypertarget{ref-Jenior2018}{}%
+30. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+\textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
+structure and metabolism of distinct cecal microbiomes during initial
+infection to promote sustained colonization. mSphere \textbf{3}.
+doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
+
+\leavevmode\hypertarget{ref-Iino2007}{}%
+31. \textbf{Iino T}, \textbf{Mori K}, \textbf{Tanaka K}, \textbf{Suzuki
+K-i}, \textbf{Harayama S}. 2007. Oscillibacter valericigenes gen. Nov.,
+sp. Nov., a valerate-producing anaerobic bacterium isolated from the
+alimentary canal of a japanese corbicula clam. International Journal of
+Systematic and Evolutionary Microbiology \textbf{57}:1840--1845.
+doi:\href{https://doi.org/10.1099/ijs.0.64717-0}{10.1099/ijs.0.64717-0}.
+
+\leavevmode\hypertarget{ref-Jump2014}{}%
+32. \textbf{Jump RLP}, \textbf{Polinkovsky A}, \textbf{Hurless K},
+\textbf{Sitzlar B}, \textbf{Eckart K}, \textbf{Tomas M},
+\textbf{Deshpande A}, \textbf{Nerandzic MM}, \textbf{Donskey CJ}. 2014.
+Metabolomics analysis identifies intestinal microbiota-derived
+biomarkers of colonization resistance in clindamycin-treated mice. PLoS
+ONE \textbf{9}:e101267.
+doi:\href{https://doi.org/10.1371/journal.pone.0101267}{10.1371/journal.pone.0101267}.
+
+\leavevmode\hypertarget{ref-McDonald2018}{}%
+33. \textbf{McDonald JAK}, \textbf{Mullish BH}, \textbf{Pechlivanis A},
+\textbf{Liu Z}, \textbf{Brignardello J}, \textbf{Kao D}, \textbf{Holmes
+E}, \textbf{Li JV}, \textbf{Clarke TB}, \textbf{Thursz MR},
+\textbf{Marchesi JR}. 2018. Inhibiting growth of clostridioides
+difficile by restoring valerate, produced by the intestinal microbiota.
+Gastroenterology \textbf{155}:1495--1507.e15.
+doi:\href{https://doi.org/10.1053/j.gastro.2018.07.014}{10.1053/j.gastro.2018.07.014}.
+
+\leavevmode\hypertarget{ref-Pfeiffer2012}{}%
+34. \textbf{Pfeiffer N}, \textbf{Desmarchelier C}, \textbf{Blaut M},
+\textbf{Daniel H}, \textbf{Haller D}, \textbf{Clavel T}. 2012.
+Acetatifactor muris gen. Nov., sp. Nov., a novel bacterium isolated from
+the intestine of an obese mouse. Archives of Microbiology
+\textbf{194}:901--907.
+doi:\href{https://doi.org/10.1007/s00203-012-0822-1}{10.1007/s00203-012-0822-1}.
+
+\leavevmode\hypertarget{ref-Lawley2012}{}%
+35. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
+\textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
+\textbf{Goulding D}, \textbf{Rad R}, \textbf{Schreiber F},
+\textbf{Brandt C}, \textbf{Deakin LJ}, \textbf{Pickard DJ},
+\textbf{Duncan SH}, \textbf{Flint HJ}, \textbf{Clark TG},
+\textbf{Parkhill J}, \textbf{Dougan G}. 2012. Targeted restoration of
+the intestinal microbiota with a simple, defined bacteriotherapy
+resolves relapsing \emph{Clostridium difficile} disease in mice. PLoS
+Pathogens \textbf{8}:e1002995.
+doi:\href{https://doi.org/10.1371/journal.ppat.1002995}{10.1371/journal.ppat.1002995}.
+
+\leavevmode\hypertarget{ref-Frost2021}{}%
+36. \textbf{Frost LR}, \textbf{Cheng JKJ}, \textbf{Unnikrishnan M}.
+2021. Clostridioides difficile biofilms: A mechanism of persistence in
+the gut? PLOS Pathogens \textbf{17}:e1009348.
+doi:\href{https://doi.org/10.1371/journal.ppat.1009348}{10.1371/journal.ppat.1009348}.
+
+\leavevmode\hypertarget{ref-Engevik2021}{}%
+37. \textbf{Engevik MA}, \textbf{Danhof HA}, \textbf{Auchtung J},
+\textbf{Endres BT}, \textbf{Ruan W}, \textbf{Bassères E},
+\textbf{Engevik AC}, \textbf{Wu Q}, \textbf{Nicholson M}, \textbf{Luna
+RA}, \textbf{Garey KW}, \textbf{Crawford SE}, \textbf{Estes MK},
+\textbf{Lux R}, \textbf{Yacyshyn MB}, \textbf{Yacyshyn B},
+\textbf{Savidge T}, \textbf{Britton RA}, \textbf{Versalovic J}. 2021.
+Fusobacterium nucleatum adheres to clostridioides difficile via the RadD
+adhesin to enhance biofilm formation in intestinal mucus.
+Gastroenterology \textbf{160}:1301--1314.e8.
+doi:\href{https://doi.org/10.1053/j.gastro.2020.11.034}{10.1053/j.gastro.2020.11.034}.
+
+\leavevmode\hypertarget{ref-Engevik2020}{}%
+38. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
+\textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
+\textbf{Luna RA}, \textbf{Hyser JM}, \textbf{Spinler JK},
+\textbf{Versalovic J}. 2020. Mucin-degrading microbes release
+monosaccharides that chemoattract clostridioides difficile and
+facilitate colonization of the human intestinal mucus layer. ACS
+Infectious Diseases \textbf{7}:1126--1142.
+doi:\href{https://doi.org/10.1021/acsinfecdis.0c00634}{10.1021/acsinfecdis.0c00634}.
+
+\leavevmode\hypertarget{ref-Dejea2018}{}%
+39. \textbf{Dejea CM}, \textbf{Fathi P}, \textbf{Craig JM},
+\textbf{Boleij A}, \textbf{Taddese R}, \textbf{Geis AL}, \textbf{Wu X},
+\textbf{Shields CED}, \textbf{Hechenbleikner EM}, \textbf{Huso DL},
+\textbf{Anders RA}, \textbf{Giardiello FM}, \textbf{Wick EC},
+\textbf{Wang H}, \textbf{Wu S}, \textbf{Pardoll DM}, \textbf{Housseau
+F}, \textbf{Sears CL}. 2018. Patients with familial adenomatous
+polyposis harbor colonic biofilms containing tumorigenic bacteria.
+Science \textbf{359}:592--597.
+doi:\href{https://doi.org/10.1126/science.aah3648}{10.1126/science.aah3648}.
+
+\leavevmode\hypertarget{ref-Semenyuk2015}{}%
+40. \textbf{Semenyuk EG}, \textbf{Poroyko VA}, \textbf{Johnston PF},
+\textbf{Jones SE}, \textbf{Knight KL}, \textbf{Gerding DN},
+\textbf{Driks A}. 2015. Analysis of bacterial communities during
+clostridium difficile infection in the mouse. Infection and Immunity
+\textbf{83}:4383--4391.
+doi:\href{https://doi.org/10.1128/iai.00145-15}{10.1128/iai.00145-15}.
+
+\leavevmode\hypertarget{ref-Deng2018}{}%
+41. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
+\textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
+\textbf{Fan H}, \textbf{Zhao X}, \textbf{Zhi F}. 2018. Bacteroides
+fragilis prevents clostridium difficile infection in a mouse model by
+restoring gut barrier and microbiome regulation. Frontiers in
+Microbiology \textbf{9}.
+doi:\href{https://doi.org/10.3389/fmicb.2018.02976}{10.3389/fmicb.2018.02976}.
+
+\leavevmode\hypertarget{ref-Normington2021}{}%
+42. \textbf{Normington C}, \textbf{Moura IB}, \textbf{Bryant JA},
+\textbf{Ewin DJ}, \textbf{Clark EV}, \textbf{Kettle MJ}, \textbf{Harris
+HC}, \textbf{Spittal W}, \textbf{Davis G}, \textbf{Henn MR},
+\textbf{Ford CB}, \textbf{Wilcox MH}, \textbf{Buckley AM}. 2021.
+Biofilms harbour clostridioides difficile, serving as a reservoir for
+recurrent infection. npj Biofilms and Microbiomes \textbf{7}.
+doi:\href{https://doi.org/10.1038/s41522-021-00184-w}{10.1038/s41522-021-00184-w}.
+
+\leavevmode\hypertarget{ref-Cirstea2018}{}%
+43. \textbf{Cirstea M}, \textbf{Radisavljevic N}, \textbf{Finlay BB}.
+2018. Good bug, bad bug: Breaking through microbial stereotypes. Cell
+Host \& Microbe \textbf{23}:10--13.
+doi:\href{https://doi.org/10.1016/j.chom.2017.12.008}{10.1016/j.chom.2017.12.008}.
+
+\leavevmode\hypertarget{ref-BorgesCanha2015}{}%
+44. \textbf{Canha MB}. 2015. Role of colonic microbiota in colorectal
+carcinogenesis: A systematic review. Revista Española de Enfermedades
+Digestivas \textbf{107}.
+doi:\href{https://doi.org/10.17235/reed.2015.3830/2015}{10.17235/reed.2015.3830/2015}.
+
+\leavevmode\hypertarget{ref-Pereira2020}{}%
+45. \textbf{Pereira FC}, \textbf{Wasmund K}, \textbf{Cobankovic I},
+\textbf{Jehmlich N}, \textbf{Herbold CW}, \textbf{Lee KS},
+\textbf{Sziranyi B}, \textbf{Vesely C}, \textbf{Decker T},
+\textbf{Stocker R}, \textbf{Warth B}, \textbf{Bergen M von},
+\textbf{Wagner M}, \textbf{Berry D}. 2020. Rational design of a
+microbial consortium of mucosal sugar utilizers reduces clostridiodes
+difficile colonization. Nature Communications \textbf{11}.
+doi:\href{https://doi.org/10.1038/s41467-020-18928-1}{10.1038/s41467-020-18928-1}.
+
+\leavevmode\hypertarget{ref-Rodriguez2016}{}%
+46. \textbf{Rodriguez C}, \textbf{Taminiau B}, \textbf{Korsak N},
+\textbf{Avesani V}, \textbf{Broeck JV}, \textbf{Brach P}, \textbf{Delmée
+M}, \textbf{Daube G}. 2016. Longitudinal survey of clostridium difficile
+presence and gut microbiota composition in a belgian nursing home. BMC
+Microbiology \textbf{16}.
+doi:\href{https://doi.org/10.1186/s12866-016-0848-7}{10.1186/s12866-016-0848-7}.
+
+\leavevmode\hypertarget{ref-Sangster2016}{}%
+47. \textbf{Sangster W}, \textbf{Hegarty JP}, \textbf{Schieffer KM},
+\textbf{Wright JR}, \textbf{Hackman J}, \textbf{Toole DR},
+\textbf{Lamendella R}, \textbf{Stewart DB}. 2016. Bacterial and fungal
+microbiota changes distinguish c. Difficile infection from other forms
+of diarrhea: Results of a prospective inpatient study. Frontiers in
+Microbiology \textbf{7}.
+doi:\href{https://doi.org/10.3389/fmicb.2016.00789}{10.3389/fmicb.2016.00789}.
+
+\leavevmode\hypertarget{ref-Vakili2020a}{}%
+48. \textbf{Vakili B}, \textbf{Fateh A}, \textbf{Aghdaei HA},
+\textbf{Sotoodehnejadnematalahi F}, \textbf{Siadat SD}. 2020. Intestinal
+microbiota in elderly inpatients with clostridioides difficile
+infection. Infection and Drug Resistance \textbf{Volume 13}:2723--2731.
+doi:\href{https://doi.org/10.2147/idr.s262019}{10.2147/idr.s262019}.
+
+\leavevmode\hypertarget{ref-Vakili2020b}{}%
+49. \textbf{Vakili B}, \textbf{Fateh A}, \textbf{Aghdaei HA},
+\textbf{Sotoodehnejadnematalahi F}, \textbf{Siadat SD}. 2020.
+Characterization of gut microbiota in hospitalized patients with
+clostridioides difficile infection. Current Microbiology
+\textbf{77}:1673--1680.
+doi:\href{https://doi.org/10.1007/s00284-020-01980-x}{10.1007/s00284-020-01980-x}.
+
+\leavevmode\hypertarget{ref-Theriot2011}{}%
+50. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
+\textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
+Cefoperazone-treated mice as an experimental platform to assess
+differential virulence ofClostridium difficilestrains. Gut Microbes
+\textbf{2}:326--334.
+doi:\href{https://doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}.
+
+\leavevmode\hypertarget{ref-Kester2020}{}%
+51. \textbf{Kester JC}, \textbf{Brubaker DK}, \textbf{Velazquez J},
+\textbf{Wright C}, \textbf{Lauffenburger DA}, \textbf{Griffith LG}.
+2020. Clostridioides difficile-associated antibiotics alter human
+mucosal barrier functions by microbiome-independent mechanisms.
+Antimicrobial Agents and Chemotherapy \textbf{64}.
+doi:\href{https://doi.org/10.1128/aac.01404-19}{10.1128/aac.01404-19}.
+
+\leavevmode\hypertarget{ref-Bergstrom2020}{}%
+52. \textbf{Bergstrom K}, \textbf{Shan X}, \textbf{Casero D},
+\textbf{Batushansky A}, \textbf{Lagishetty V}, \textbf{Jacobs JP},
+\textbf{Hoover C}, \textbf{Kondo Y}, \textbf{Shao B}, \textbf{Gao L},
+\textbf{Zandberg W}, \textbf{Noyovitz B}, \textbf{McDaniel JM},
+\textbf{Gibson DL}, \textbf{Pakpour S}, \textbf{Kazemian N},
+\textbf{McGee S}, \textbf{Houchen CW}, \textbf{Rao CV}, \textbf{Griffin
+TM}, \textbf{Sonnenburg JL}, \textbf{McEver RP}, \textbf{Braun J},
+\textbf{Xia L}. 2020. Proximal colonderived o-glycosylated mucus
+encapsulates and modulates the microbiota. Science
+\textbf{370}:467--472.
+doi:\href{https://doi.org/10.1126/science.aay7367}{10.1126/science.aay7367}.
+
+\leavevmode\hypertarget{ref-Dieterle2018}{}%
+53. \textbf{Dieterle MG}, \textbf{Rao K}, \textbf{Young VB}. 2018. Novel
+therapies and preventative strategies for primary and
+recurrentClostridium difficileinfections. Annals of the New York Academy
+of Sciences \textbf{1435}:110--138.
+doi:\href{https://doi.org/10.1111/nyas.13958}{10.1111/nyas.13958}.
+
+\leavevmode\hypertarget{ref-Kelly2021}{}%
+54. \textbf{Kelly CR}, \textbf{Yen EF}, \textbf{Grinspan AM},
+\textbf{Kahn SA}, \textbf{Atreja A}, \textbf{Lewis JD}, \textbf{Moore
+TA}, \textbf{Rubin DT}, \textbf{Kim AM}, \textbf{Serra S},
+\textbf{Nersesova Y}, \textbf{Fredell L}, \textbf{Hunsicker D},
+\textbf{McDonald D}, \textbf{Knight R}, \textbf{Allegretti JR},
+\textbf{Pekow J}, \textbf{Absah I}, \textbf{Hsu R}, \textbf{Vincent J},
+\textbf{Khanna S}, \textbf{Tangen L}, \textbf{Crawford CV},
+\textbf{Mattar MC}, \textbf{Chen LA}, \textbf{Fischer M},
+\textbf{Arsenescu RI}, \textbf{Feuerstadt P}, \textbf{Goldstein J},
+\textbf{Kerman D}, \textbf{Ehrlich AC}, \textbf{Wu GD}, \textbf{Laine
+L}. 2021. Fecal microbiota transplantation is highly effective in
+real-world practice: Initial results from the FMT national registry.
+Gastroenterology \textbf{160}:183--192.e3.
+doi:\href{https://doi.org/10.1053/j.gastro.2020.09.038}{10.1053/j.gastro.2020.09.038}.
+
+\leavevmode\hypertarget{ref-Kozich2013}{}%
+55. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\leavevmode\hypertarget{ref-Schloss2009}{}%
+56. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
+\leavevmode\hypertarget{ref-Quast2012}{}%
+57. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
+\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
+J}, \textbf{Glöckner FO}. 2012. The SILVA ribosomal RNA gene database
+project: Improved data processing and web-based tools. Nucleic Acids
+Research \textbf{41}:D590--D596.
+doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.
+
+\leavevmode\hypertarget{ref-Cole2013}{}%
+58. \textbf{Cole JR}, \textbf{Wang Q}, \textbf{Fish JA}, \textbf{Chai
+B}, \textbf{McGarrell DM}, \textbf{Sun Y}, \textbf{Brown CT},
+\textbf{Porras-Alfaro A}, \textbf{Kuske CR}, \textbf{Tiedje JM}. 2013.
+Ribosomal database project: Data and tools for high throughput rRNA
+analysis. Nucleic Acids Research \textbf{42}:D633--D642.
+doi:\href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
+
+\leavevmode\hypertarget{ref-Vegan2018}{}%
+59. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
+\textbf{Kindt R}, \textbf{Legendre P}, \textbf{McGlinn D},
+\textbf{Minchin PR}, \textbf{O'Hara RB}, \textbf{Simpson GL},
+\textbf{Solymos P}, \textbf{Stevens MHH}, \textbf{Szoecs E},
+\textbf{Wagner H}. 2018. Vegan: Community ecology package.
+
+\leavevmode\hypertarget{ref-mikropml}{}%
+60. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool KL},
+\textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020.
+mikRopML: User-friendly r package for robust machine learning pipelines.
+
+\leavevmode\hypertarget{ref-Topcuoglu2021}{}%
+61. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
+\textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
+Mikropml: User-friendly r package for supervised machine learning
+pipelines. Journal of Open Source Software \textbf{6}:3073.
+doi:\href{https://doi.org/10.21105/joss.03073}{10.21105/joss.03073}.
+
+\leavevmode\hypertarget{ref-Topcuoglu2020}{}%
+62. \textbf{Topçuoğlu BD}, \textbf{Lesniak NA}, \textbf{Ruffin MT},
+\textbf{Wiens J}, \textbf{Schloss PD}. 2020. A framework for effective
+application of machine learning to microbiome-based classification
+problems. mBio \textbf{11}.
+doi:\href{https://doi.org/10.1128/mbio.00434-20}{10.1128/mbio.00434-20}.
+
+\leavevmode\hypertarget{ref-r_citation_2020}{}%
+63. \textbf{R Core Team}. 2020. R: A language and environment for
+statistical computing. R Foundation for Statistical Computing, Vienna,
+Austria.
+
+\leavevmode\hypertarget{ref-Tidyverse2019}{}%
+64. \textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan J},
+\textbf{Chang W}, \textbf{McGowan LD}, \textbf{François R},
+\textbf{Grolemund G}, \textbf{Hayes A}, \textbf{Henry L}, \textbf{Hester
+J}, \textbf{Kuhn M}, \textbf{Pedersen TL}, \textbf{Miller E},
+\textbf{Bache SM}, \textbf{Müller K}, \textbf{Ooms J}, \textbf{Robinson
+D}, \textbf{Seidel DP}, \textbf{Spinu V}, \textbf{Takahashi K},
+\textbf{Vaughan D}, \textbf{Wilke C}, \textbf{Woo K}, \textbf{Yutani H}.
+2019. Welcome to the tidyverse. Journal of Open Source Software
+\textbf{4}:1686.
+doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}.
 
@@
 treatment with PEG and/or clindamycin, followed by \emph{C. difficile}
-challenge. C. \emph{C. difficile} CFU/gram stool measured over time (N =
-10-59 mice per time point) via serial dilutions. The black line
+challenge. C. \emph{C. difficile} CFU/gram stool measured over time via
+serial dilutions(N = 10-59 mice per time point). The black line
 represents the limit of detection for the first serial dilution. CFU
@@
 Bray-Curtis distances from stool samples collected throughout the
-experiment. For A and C, each circle represents a sample from an
-individual mouse and the transparency of the symbol corresponds to the
-day post-challenge. B. Bray-Curtis distances of stool samples collected
-on either day 10 or 30 post-challenge relative to the baseline sample
-collected for each mouse (before any drug treatments were administered).
-C. Shannon diversity in stool communities over time. The line indicates
-the median value for each treatment group. The colors of the symbols and
-lines represent the four treatment groups. D. 14 of the 33 genera
-affected by PEG treatment (Data Set S1, sheet X). The symbols represent
-the median relative abundance for a treatment group at either baseline
-(open circle) or 1-day post treatment (closed circle). Data from the
-5-day PEG and 5-day PEG plus 10-day recovery groups were analyzed by
-paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for
-testing all identified genera. The clindamycin and 5-day PEG plus
-clindamycin treatment groups are shown for comparison. E. 6 of the 24
-genera that were significantly different between the four treatment
-groups over multiple time points. Differences between treatment groups
-were identified by Kruskal-Wallis test with Benjamini-Hochberg
-correction for testing all identified genera. The gray vertical line (D)
-and horizontal vertical lines (E) indicate the limit of detection.
-\newpage
+experiment. Each circle represents a sample from an individual mouse and
+the transparency of the symbol corresponds to the day post-challenge. B.
+Bray-Curtis distances of stool samples collected on either day 10 or 30
+post-challenge relative to the baseline sample collected for each mouse
+(before any drug treatments were administered). C. Shannon diversity in
+stool communities over time. The line indicates the median value for
+each treatment group. For B-C, the symboles represent samples from
+individual mice and the line indicates the median value for each
+treatment group. D. 14 of the 33 genera affected by PEG treatment (Data
+Set S1, sheet X). The symbols represent the median relative abundance
+for a treatment group at either baseline (open circle) or 1-day post
+treatment (closed circle). Relative abundance data from paired baseline
+and 1-day post treatment stool sampes from the 5-day PEG and 5-day PEG
+plus 10-day recovery groups were analyzed by paired Wilcoxan signed-rank
+test with Benjamini-Hochberg correction for testing all identified
+genera. The clindamycin and 5-day PEG plus clindamycin treatment groups
+are shown on the plot for comparison. E. 6 of the 24 genera that were
+significantly different between the treatment groups over multiple time
+points, the 5-day PEG plus clindamycin treatment group was only followed
+through 6-days post-challenge. Differences between treatment groups were
+identified by Kruskal-Wallis test with Benjamini-Hochberg correction for
+testing all identified genera. The gray vertical line (D) and horizontal
+vertical lines (E) indicate the limit of detection. \newpage
 
@@
 indicates the median value for each treatment group. The colors of the
-symbols and lines represent the four treatment groups. A similar pattern
-was observed with the proximal and distal colon communities (Data Set
-S1, sheet X-X). B. PCoA of Bray-Curtis distances from mucosal samples
-collected throughout the experiment. Circles, triangles, and squares
-indicate cecum, proximal colon, and distal colon communities,
-respectively. For A-B, transparency of the symbol corresponds to the day
-post-challenge that the sample was collected. C. The median relative
-abundance of the 4 genera that were significantly different between the
-cecum communities of different treatment groups on day 6 and day 30
-(Data Set S1, sheet X). The gray vertical line indicate the limit of
-detection. D-E. The histopathology summary scores from cecum and colon
-H\&E stained slides. The summary score is the total score based on
-evaluation of edema, cellular infiltration, and inflammation. Each
-category is given a score ranging from 0-4, thus the maximum possible
-summary score is 12. The tissue for histology was collected at either 4
-(D) or 6 (E) days post-challenge with the exception that one set of
-5-day PEG treated mock-challenged mice were collected on day 0
-post-challenge (first set of open circles in D). Histology data were
-analyzed with the Kruskal-Wallis test followed by pairwise Wilcoxon
-comparisons with Benjamini-Hochberg correction. \newpage
+symbols and lines represent individual and median relative abundance
+values for four treatment groups. A similar pattern was observed with
+the proximal and distal colon communities (Data Set S1, sheet X-X). B.
+PCoA of Bray-Curtis distances from mucosal samples collected throughout
+the experiment. Circles, triangles, and squares indicate the cecum,
+proximal colon, and distal colon communities, respectively. Transparency
+of the symbol corresponds to the day post-challenge that the sample was
+collected. C. The median relative abundance of the 4 genera that were
+significantly different between the cecum communities of different
+treatment groups on day 6 and day 30 post-challenge (Data Set S1, sheet
+X). The gray vertical lines indicate the limit of detection. D-E. The
+histopathology summary scores from cecum and colon H\&E stained tissue
+sections. The summary score is the total score based on evaluation of
+edema, cellular infiltration, and inflammation in either the cecum or
+colon tissue. Each category is given a score ranging from 0-4, thus the
+maximum possible summary score is 12. The tissue for histology was
+collected at either 4 (D) or 6 (E) days post-challenge with the
+exception that one set of 5-day PEG treated mock-challenged mice were
+collected on day 0 post-challenge (first set of open purple circles in
+D). Histology data were analyzed with the Kruskal-Wallis test followed
+by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
+\newpage
 
@@
 \emph{C. difficile} 630 spores. B. \emph{C. difficile} CFU/gram stool
-measured over time (N = 12-18 mice per time point) via several
-dilutions. The black dotted line represents the limit of detection for
-the first serial dilution. Asterisks indicate time points where the
-CFU/gram was significantly different between treatment groups by
+measured over time (N = 12-18 mice per time point) by serial dilutions.
+The black dashedd horizontal line represents the limit of detection for
+the first serial dilution. For B and D, asterisks indicate time points
+where there was a significant difference between treatment groups by
 Kruskall-Wallis test with Benjamini-Hochberg correction for testing
-multiple time points. C. PCoA of Bray-Curtis distances from stool
-communities collected from the three treatment groups over time (day:
-R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} = 0.19). D.
-Shannon diversity in stool communities over time with colored lines
-representing the median value for each treatment group. For B-D, each
-symbol represents a sample from an individual mouse and symbol
-transparency corresponds to the day post-challenge that the sample was
-collected. E. Median relative abundances per treatment group for 6 out
+multiple time points. For B-D, each symbol represents a sample from an
+individual mouse and lines indicate the median value for each treatment
+group. C. PCoA of Bray-Curtis distances from stool communities collected
+over time (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2}
+= 0.19). Symbol transparency represents the day post-challenge of the
+experiment. For C-E, the B on the day legend or days post-challenge X
+axis stands for baseline and represents the sample that was collected
+prior to any drug treatments. D. Shannon diversity in stool communities
+over time. E. Median relative abundances per treatment group for 6 out
 of the 14 genera that were affected by treatment, but recovered close to
 baseline levels by 7 days post-challenge (Fig. 3E, Data Set S1, Sheet
-X). Stool samples from either baseline and day 1 or baseline and day 7
-were analyzed by paired Wilcoxan signed-rank test with
-Benjamini-Hochberg correction for testing all identified genera. The
-gray horizontal line represents the limit of detection.
+X). Paired stool sample relative abundance values either baseline and
+day 1 or baseline and day 7 were analyzed by paired Wilcoxan signed-rank
+test with Benjamini-Hochberg correction for testing all identified
+genera. Only genera that were different between baseline and 1-day
+post-challenge, but not baseline and 7-days post-challenge. The gray
+horizontal lines represents the limit of detection.
 
@@
 post \emph{C. difficile} challenge prolongs colonization regardless of
-whether an FMT is also administered.} A. Setup off the experimental time
-line for experiments with post-CDI PEG treated mice. There were a total
-of 4 different treatment groups. All mice were administered 10 mg/kg
-clindamycin intraperitoneally (IP) 1 day before challenge with
-10\textsuperscript{3-5} \emph{C. difficile} 630 spores. 1. Did not
-receive any additional treatment. 2. Immediately after \emph{C.
+whether an FMT is also administered.} A. Setup of the experimental time
+line for experiments with post-challenge PEG treated mice. There were a
+total of 4 different treatment groups. All mice were administered 10
+mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
+10\textsuperscript{3-5} \emph{C. difficile} 630 spores. 1. Received no
+additional treatment (Clindamycin). 2. Immediately after \emph{C.
 difficile} challenge, mice received 15\% PEG 3350 in the drinking water
 for 1 day. 3-4. 3-days after challenge, mice received 1-day PEG
-treatment and then received either a fecal microbiota transplant (3) or
-PBS (4) solution by oral gavage. Mice were followed through 15-30 days
-post-challenge (only the post-CDI 1-day PEG group was followed through
-30 days post-challenge). B. CFU/g of \emph{C. difficile} stool measured
-over time via serial dilutions. The black line represents the limit of
-detection for the first serial dilution. C-D. Shannon diversity (C) and
-richness (D) in stool communities over time. B-D. Each symbol represents
-a stool sample from an individual mouse with the lines representing the
-median value for each treatment group. The transparancy of the symbol
-corresponds to the day post-challenge. Asterisks indicate time points
-with significant differences between groups by a Kruskall-Wallis test
-with a Benjamini-Hochberg correction for testing multiple times points.
-Colored rectangles indicates the 1-day PEG treatment period for
-applicable groups. The data presented are from a total of 3 separate
-experiments. \newpage
+treatment and then received either 100 microliters a fecal microbiota
+transplant (3) or PBS (4) solution by oral gavage. Mice were followed
+through 15-30 days post-challenge (only the post-CDI 1-day PEG group was
+followed through 30 days post-challenge). B. CFU/g of \emph{C.
+difficile} stool measured over time via serial dilutions. The black line
+represents the limit of detection for the first serial dilution. C-D.
+Shannon diversity (C) and richness (D) in stool communities over time.
+B-D. Each symbol represents a stool sample from an individual mouse with
+the lines representing the median value for each treatment group.
+Asterisks indicate time points with significant differences between
+groups by a Kruskall-Wallis test with a Benjamini-Hochberg correction
+for testing multiple times points. Colored rectangles indicates the
+1-day PEG treatment period for applicable groups. The data presented are
+from a total of 3 separate experiments. \newpage
 
@@
 from stool samples collected over time as well as the FMT solution that
-was administered to one treatment group. Each circle represents an
-individual sample, the transparency of the circle corresponds to day
-post-challenge as shown in Fig. 6C-D. B. Median relative abundances of 2
-genera that were significantly different over multiple time points in
-mice that were administered either FMT or PBS solution via gavage C.
-Median relative abundances of the top 6 out of 24 genera that were
-significant over multiple timepoints, plotted over time (Data Set S1,
-Sheet X). For B-C, colored rectangles indicates the 1-day PEG treatment
-period for applicable groups. Gray horizontal lines represent the limit
-of detection.
+was administered to one of the treatment groups. Each circle represents
+an individual sample, the transparency of the circle corresponds to day
+post-challenge. B. Median relative abundances of 2 genera that were
+significantly different over multiple time points in mice that were
+administered either FMT or PBS solution via gavage C. Median relative
+abundances of the top 6 out of 24 genera that were significant over
+multiple timepoints, plotted over time (Data Set S1, Sheet X). For B-C,
+colored rectangles indicates the 1-day PEG treatment period for
+applicable groups. Gray horizontal lines represent the limit of
+detection. Differences between treatment groups were identified by
+Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
+identified genera. For pairwise comparisons of the groups (B), we
+performed pairwise Wilcoxon comparisons with Benjamini-Hochberg
+correction for testing all combinations of group pairs.
 
@@
 forest model trained on five day post-challenge community relative
-abundance data to predict wehther mice would still be colonized with
+abundance data to predict whether mice would still be colonized with
 \emph{C. difficile} 10 days post-challenge. The median (point) and
-interquantile range (lines) change in AUROC when the bacteria is left
-out of the model is shown. B. The median relative abundances at 5 days
-post-challenge of the top ten bacteria that contributed to the random
-forest classification model. Color indicates whether the mice were still
-colonized with \emph{C. difficile} 10 days post-challenge and the black
-horizontal line represents the median relative abundance. Each symbol
-represents a stool sample from an individual mouse and the shape of the
-symbol indicates whether the PEG-treated mice recieved a 5-day (Fig.
-1-3), 1-day (Fig. 4) or post-CDI PEG (Fig. 5-6) treatment. C. The median
+interquartile range (lines) change in AUROC when the bacteria is left
+out of the model by permutation feature importance analysis. B. The
+median relative abundances of the top ten bacteria that contributed to
+the random forest classification model at 5 days post-challenge . Color
+indicates whether the mice were still colonized with \emph{C. difficile}
+10 days post-challenge and the black horizontal line represents the
+median relative abundance for the two categories. Each symbol represents
+a stool sample from an individual mouse and the shape of the symbol
+indicates whether the PEG-treated mice received a 5-day (Fig. 1-3),
+1-day (Fig. 4) or post-challenge PEG (Fig. 5-6) treatment. C. The median
 relative abundances of the 5 genera with greater than 1\% median
 relative abundance in the stool community over time. For B-C, the gray
-horizontal line represents the limit of detection. \newpage
-
-\includegraphics{figure_8.pdf} \textbf{Figure 8. Schematic summarizing
-findings.} A. The gut microbiota of our C57Bl/6 mice is restant to
-\emph{C. difficle} but treatment with either the antibiotic,
-clindamycin, or the osmotic laxative, PEG 3350 renders the mice
-susceptible to \emph{C. difficile} colonization. Recovery of
-colonization resistance in clindamycin-treated mice is relatively
-straightforward and the mice clear \emph{C.difficile} within 10 days
-post-challenge. However, for mice that recieved a 5-day PEG treatment or
-a 1-day PEG treatment after \emph{C. difficile} challenge recovery of
-colonization resistance is more uncertain because mice were still
-colonized with \emph{C. difficile} 30 days post-challenge in the case of
-several PEG treatments. We found increased \emph{Porphyromonadaceae} and
-\emph{Lachnospiraceae} were associated with recovery of colonization
-resistance, while increased \emph{Enterobacteriaceae} and
-\emph{Bacteroides} were associated with prolonged \emph{C. difficle}
-colonization. B.Considering that most hospitals recommend not performing
-\emph{C. difficile} testing on patients taking laxatives and laxatives
-are used when administering fecal microbiota transplants via colonoscopy
-to patients with recurrent CDIs. further studies are needed to evaluate
-if laxatives impact human microbiota colonization resistance. Further
-studies are needed to understand the impact of osmotic laxatives on
-\emph{C. difficile} colonization resistance and clearance in human
-patients. \newpage
+horizontal lines represents the limit of detection. \newpage
+
+\includegraphics{figure_8.pdf}
+
+\textbf{Figure 8. Schematic summarizing findings.} The gut microbiota of
+our C57Bl/6 mice is resistant to \emph{C. difficle} but treatment with
+either the antibiotic, clindamycin, or the osmotic laxative, PEG 3350,
+renders the mice susceptible to \emph{C. difficile} colonization.
+Recovery of colonization resistance in clindamycin-treated mice is
+relatively straightforward and the mice clear \emph{C.difficile} within
+10 days post-challenge. However, for mice that received either a 5-day
+PEG treatment pre-\emph{C. difficile} challenge or a 1-day PEG treatment
+post-challenge recovery of colonization resistance was delayed because
+most mice were still colonized with \emph{C. difficile} 10 days
+post-challenge. We found increased relative abundances of
+\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} were associated
+with recovery of colonization resistance, while increased relative
+abundances of \emph{Enterobacteriaceae} and \emph{Bacteroides} were
+associated with prolonged \emph{C. difficle} colonization. \newpage
 
@@
 \emph{C. difficile} CFU/g over time in the stool samples collected from
-5-day PEG treatment plus 10-day recovery mice. Same data presented in
-Fig. 1C, but the data for the other 3 treatment groups have been
-removed. B. Median relative abundances of 8 bacterial genera from day 0
-post-challenge onward from the 10-day recovery mice. We analyzed samples
-from day 0 and day 8 post-challenge, which represented the the time
-points where mice were challenged with \emph{C. difficile} and when the
-median relative \emph{C. difficile} CFU stabilized for the group using
-the paired Wilcoxan signed-rank test, but no genera were signifcant
-after Benjamini-Hochberg correction. \newpage
+5-day PEG treated mice that were allowed to recover for 10 days prior to
+challenge. Same data presented in Fig. 1C, but the data for the other 3
+treatment groups have been removed and each line represents the CFU over
+time for an individual mouse. Mouse 10 was found dead 6 days
+post-challenge. B. Relative abundances of eight bacterial genera from
+day 0 post-challenge onward in each off the 10-day recovery mice. We
+analyzed samples from day 0 and day 8 post-challenge, which represented
+the the time points where mice were challenged with \emph{C. difficile}
+and when the median relative \emph{C. difficile} CFU stabilized for the
+group using the paired Wilcoxan signed-rank test, but no genera were
+significantly different after Benjamini-Hochberg correction. \newpage
 
@@
 has a large impact on the mucosal microbiota 6 days post-challenge} A.
-The\\
-\newpage
+The relative abundances of the 10 bacterial genera that were
+significantly different between treatment groups at 6 days
+post-infection in the cecum tissue (the relative abundances of the 10
+genera were also significantly different in the proximal and distal
+colon tissues, Data Sheet S1, Sheet X). Each symbol represents a tissue
+sample from an individual mouse, the black horizontal lines represents
+the median relative abundances for each treatment group. B. The relative
+abundance of \emph{Peptostreptococacceae} in the three types of tissue
+sample communities over time. For A-B, the gray horizontal lines
+represent the limit of detection. \newpage
 
@@
 challenge does not enhance the disruptive effect of PEG on the
-microbiota.} A. \newpage
-
-\includegraphics{figure_S4.pdf} \textbf{Figure S4. Specific OTUs
-associated with clearance by 10 days post-challenge that are mostly
-absent in mice with prolonged \emph{C. difficile} colonization. Ex.
-\emph{Muribaculum intestinale}.} A. \newpage
-
-\hypertarget{references}{%
-\subsection*{References}\label{references}}
-\addcontentsline{toc}{subsection}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Britton2014}{}%
-1. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
-microbiota in resistance to colonization by clostridium difficile.
-Gastroenterology \textbf{146}:1547--1553.
-doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
-
-\leavevmode\hypertarget{ref-Maier2018}{}%
-2. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
-\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
-\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
-\textbf{Mori H}, \textbf{Patil KR}, \textbf{Bork P}, \textbf{Typas A}.
-2018. Extensive impact of non-antibiotic drugs on human gut bacteria.
-Nature \textbf{555}:623--628.
-doi:\href{https://doi.org/10.1038/nature25979}{10.1038/nature25979}.
-
-\leavevmode\hypertarget{ref-LeBastard2017}{}%
-3. \textbf{Bastard QL}, \textbf{Al-Ghalith GA}, \textbf{Grégoire M},
-\textbf{Chapelet G}, \textbf{Javaudin F}, \textbf{Dailly E},
-\textbf{Batard E}, \textbf{Knights D}, \textbf{Montassier E}. 2017.
-Systematic review: Human gut dysbiosis induced by non-antibiotic
-prescription medications. Alimentary Pharmacology \& Therapeutics
-\textbf{47}:332--345.
-doi:\href{https://doi.org/10.1111/apt.14451}{10.1111/apt.14451}.
-
-\leavevmode\hypertarget{ref-VichVila2020}{}%
-4. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
-T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
-\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
-\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
-\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
-composition and metabolic function of the gut microbiota. Nature
-Communications \textbf{11}.
-doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
-
-\leavevmode\hypertarget{ref-Oh2018}{}%
-5. \textbf{Oh J}, \textbf{Makar M}, \textbf{Fusco C}, \textbf{McCaffrey
-R}, \textbf{Rao K}, \textbf{Ryan EE}, \textbf{Washer L}, \textbf{West
-LR}, \textbf{Young VB}, \textbf{Guttag J}, \textbf{Hooper DC},
-\textbf{Shenoy ES}, \textbf{Wiens J}. 2018. A generalizable, data-driven
-approach to predict daily risk ofClostridium difficileInfection at two
-large academic health centers. Infection Control \& Hospital
-Epidemiology \textbf{39}:425--433.
-doi:\href{https://doi.org/10.1017/ice.2018.16}{10.1017/ice.2018.16}.
-
-\leavevmode\hypertarget{ref-Mora2012}{}%
-6. \textbf{Mora AL}, \textbf{Salazar M}, \textbf{Pablo-Caeiro J},
-\textbf{Frost CP}, \textbf{Yadav Y}, \textbf{DuPont HL}, \textbf{Garey
-KW}. 2012. Moderate to high use of opioid analgesics are associated with
-an increased risk of clostridium difficile infection. The American
-Journal of the Medical Sciences \textbf{343}:277--280.
-doi:\href{https://doi.org/10.1097/maj.0b013e31822f42eb}{10.1097/maj.0b013e31822f42eb}.
-
-\leavevmode\hypertarget{ref-Nehra2018}{}%
-7. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
-\textbf{Nehra V}. 2018. Proton pump inhibitors: Review of emerging
-concerns. Mayo Clinic Proceedings \textbf{93}:240--246.
-doi:\href{https://doi.org/10.1016/j.mayocp.2017.10.022}{10.1016/j.mayocp.2017.10.022}.
-
-\leavevmode\hypertarget{ref-Krishna2013}{}%
-8. \textbf{Krishna SG}, \textbf{Zhao W}, \textbf{Apewokin SK},
-\textbf{Krishna K}, \textbf{Chepyala P}, \textbf{Anaissie EJ}. 2013.
-Risk factors, preemptive therapy, and antiperistaltic agents
-forClostridium difficileinfection in cancer patients. Transplant
-Infectious Disease n/a--n/a.
-doi:\href{https://doi.org/10.1111/tid.12112}{10.1111/tid.12112}.
-
-\leavevmode\hypertarget{ref-Tomkovich2019}{}%
-9. \textbf{Tomkovich S}, \textbf{Lesniak NA}, \textbf{Li Y},
-\textbf{Bishop L}, \textbf{Fitzgerald MJ}, \textbf{Schloss PD}. 2019.
-The proton pump inhibitor omeprazole does not promote
-\emph{Clostridioides difficile} colonization in a murine model. mSphere
-\textbf{4}.
-doi:\href{https://doi.org/10.1128/msphere.00693-19}{10.1128/msphere.00693-19}.
-
-\leavevmode\hypertarget{ref-Vandeputte2015}{}%
-10. \textbf{Vandeputte D}, \textbf{Falony G}, \textbf{Vieira-Silva S},
-\textbf{Tito RY}, \textbf{Joossens M}, \textbf{Raes J}. 2015. Stool
-consistency is strongly associated with gut microbiota richness and
-composition, enterotypes and bacterial growth rates. Gut
-\textbf{65}:57--62.
-doi:\href{https://doi.org/10.1136/gutjnl-2015-309618}{10.1136/gutjnl-2015-309618}.
-
-\leavevmode\hypertarget{ref-VujkovicCvijin2020}{}%
-11. \textbf{Vujkovic-Cvijin I}, \textbf{Sklar J}, \textbf{Jiang L},
-\textbf{Natarajan L}, \textbf{Knight R}, \textbf{Belkaid Y}. 2020. Host
-variables confound gut microbiota studies of human disease. Nature
-\textbf{587}:448--454.
-doi:\href{https://doi.org/10.1038/s41586-020-2881-9}{10.1038/s41586-020-2881-9}.
-
-\leavevmode\hypertarget{ref-Schubert2014}{}%
-12. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
-\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
-distinguish patients with clostridium difficile infection and non-c.
-Difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
-doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
-
-\leavevmode\hypertarget{ref-Nagata2019}{}%
-13. \textbf{Nagata N}, \textbf{Tohya M}, \textbf{Fukuda S}, \textbf{Suda
-W}, \textbf{Nishijima S}, \textbf{Takeuchi F}, \textbf{Ohsugi M},
-\textbf{Tsujimoto T}, \textbf{Nakamura T}, \textbf{Shimomura A},
-\textbf{Yanagisawa N}, \textbf{Hisada Y}, \textbf{Watanabe K},
-\textbf{Imbe K}, \textbf{Akiyama J}, \textbf{Mizokami M},
-\textbf{Miyoshi-Akiyama T}, \textbf{Uemura N}, \textbf{Hattori M}. 2019.
-Effects of bowel preparation on the human gut microbiome and metabolome.
-Scientific Reports \textbf{9}.
-doi:\href{https://doi.org/10.1038/s41598-019-40182-9}{10.1038/s41598-019-40182-9}.
-
-\leavevmode\hypertarget{ref-Kashyap2013}{}%
-14. \textbf{Kashyap PC}, \textbf{Marcobal A}, \textbf{Ursell LK},
-\textbf{Larauche M}, \textbf{Duboc H}, \textbf{Earle KA},
-\textbf{Sonnenburg ED}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
-\textbf{Million M}, \textbf{Tache Y}, \textbf{Pasricha PJ},
-\textbf{Knight R}, \textbf{Farrugia G}, \textbf{Sonnenburg JL}. 2013.
-Complex interactions among diet, gastrointestinal transit, and gut
-microbiota in humanized mice. Gastroenterology \textbf{144}:967--977.
-doi:\href{https://doi.org/10.1053/j.gastro.2013.01.047}{10.1053/j.gastro.2013.01.047}.
-
-\leavevmode\hypertarget{ref-Ferreyra2014}{}%
-15. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
-\textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
-Gut microbiota-produced succinate promotes c.~difficile infection after
-antibiotic treatment or motility disturbance. Cell Host \& Microbe
-\textbf{16}:770--777.
-doi:\href{https://doi.org/10.1016/j.chom.2014.11.003}{10.1016/j.chom.2014.11.003}.
-
-\leavevmode\hypertarget{ref-Tropini2018}{}%
-16. \textbf{Tropini C}, \textbf{Moss EL}, \textbf{Merrill BD},
-\textbf{Ng KM}, \textbf{Higginbottom SK}, \textbf{Casavant EP},
-\textbf{Gonzalez CG}, \textbf{Fremin B}, \textbf{Bouley DM},
-\textbf{Elias JE}, \textbf{Bhatt AS}, \textbf{Huang KC},
-\textbf{Sonnenburg JL}. 2018. Transient osmotic perturbation causes
-long-term alteration to the gut microbiota. Cell
-\textbf{173}:1742--1754.e17.
-doi:\href{https://doi.org/10.1016/j.cell.2018.05.008}{10.1016/j.cell.2018.05.008}.
-
-\leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-17. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
-\textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
-Diarrhoeal events can trigger long-term clostridium difficile
-colonization with recurrent blooms. Nature Microbiology
-\textbf{5}:642--650.
-doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.
-
-\leavevmode\hypertarget{ref-Olson2013}{}%
-18. \textbf{Olson A}, \textbf{Diebel LN}, \textbf{Liberati DM}. 2013.
-Effect of host defenses on clostridium difficile toxininduced intestinal
-barrier injury. Journal of Trauma and Acute Care Surgery
-\textbf{74}:983--990.
-doi:\href{https://doi.org/10.1097/ta.0b013e3182858477}{10.1097/ta.0b013e3182858477}.
-
-\leavevmode\hypertarget{ref-Diebel2014}{}%
-19. \textbf{Diebel LN}, \textbf{Liberati DM}. 2014. Reinforcement of the
-intestinal mucus layer protects against clostridium difficile intestinal
-injury in~vitro. Journal of the American College of Surgeons
-\textbf{219}:460--468.
-doi:\href{https://doi.org/10.1016/j.jamcollsurg.2014.05.005}{10.1016/j.jamcollsurg.2014.05.005}.
-
-\leavevmode\hypertarget{ref-vanNood2013}{}%
-20. \textbf{Nood E van}, \textbf{Vrieze A}, \textbf{Nieuwdorp M},
-\textbf{Fuentes S}, \textbf{Zoetendal EG}, \textbf{Vos WM de},
-\textbf{Visser CE}, \textbf{Kuijper EJ}, \textbf{Bartelsman JFWM},
-\textbf{Tijssen JGP}, \textbf{Speelman P}, \textbf{Dijkgraaf MGW},
-\textbf{Keller JJ}. 2013. Duodenal infusion of donor feces for
-RecurrentClostridium difficile. New England Journal of Medicine
-\textbf{368}:407--415.
-doi:\href{https://doi.org/10.1056/nejmoa1205037}{10.1056/nejmoa1205037}.
-
-\leavevmode\hypertarget{ref-Razik2017}{}%
-21. \textbf{Razik R}, \textbf{Osman M}, \textbf{Lieberman A},
-\textbf{Allegretti JR}, \textbf{Kassam Z}. 2017. Faecal microbiota
-transplantation for clostridium difficile infection: A multicentre study
-of non-responders. Medical Journal of Australia \textbf{207}:159--160.
-doi:\href{https://doi.org/10.5694/mja16.01452}{10.5694/mja16.01452}.
-
-\leavevmode\hypertarget{ref-Postigo2012}{}%
-22. \textbf{Postigo R}, \textbf{Kim JH}. 2012. Colonoscopic versus
-nasogastric fecal transplantation for the treatment of clostridium
-difficile infection: A review and pooled analysis. Infection
-\textbf{40}:643--648.
-doi:\href{https://doi.org/10.1007/s15010-012-0307-9}{10.1007/s15010-012-0307-9}.
-
-\leavevmode\hypertarget{ref-Liacouras1996}{}%
-23. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
-irrigation as an adjunct to the treatment of chronic, relapsing
-clostridium difficile colitis. Journal of Clinical Gastroenterology
-\textbf{22}:186--189.
-doi:\href{https://doi.org/10.1097/00004836-199604000-00007}{10.1097/00004836-199604000-00007}.
-
-\leavevmode\hypertarget{ref-Schubert2015}{}%
-24. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against \emph{Clostridium
-difficile}. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
-
-\leavevmode\hypertarget{ref-Jenior2017}{}%
-25. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
-alternative nutrient niches during infection across distinct murine gut
-microbiomes. mSystems \textbf{2}.
-doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
-
-\leavevmode\hypertarget{ref-Tomkovich2020}{}%
-26. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
-\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
-antibiotic perturbation influence clostridioides difficile clearance in
-mice. mSphere \textbf{5}.
-doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
-
-\leavevmode\hypertarget{ref-Nguyen2015}{}%
-27. \textbf{Nguyen TLA}, \textbf{Vieira-Silva S}, \textbf{Liston A},
-\textbf{Raes J}. 2015. How informative is the mouse for human gut
-microbiota research? Disease Models \& Mechanisms \textbf{8}:1--16.
-doi:\href{https://doi.org/10.1242/dmm.017400}{10.1242/dmm.017400}.
-
-\leavevmode\hypertarget{ref-Reeves2011}{}%
-28. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
-\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
-interplay between microbiome dynamics and pathogen dynamics in a murine
-model of \emph{Clostridium difficile} infection \textbf{2}:145--158.
-doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.
-
-\leavevmode\hypertarget{ref-Dieterle2020}{}%
-29. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
-\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
-\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
-\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
-K}. 2020. Systemic inflammatory mediators are effective biomarkers for
-predicting adverse outcomes in clostridioides difficile infection. mBio
-\textbf{11}.
-doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
-
-\leavevmode\hypertarget{ref-Jenior2018}{}%
-30. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
-structure and metabolism of distinct cecal microbiomes during initial
-infection to promote sustained colonization. mSphere \textbf{3}.
-doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
-
-\leavevmode\hypertarget{ref-Vornhagen2020}{}%
-31. \textbf{Vornhagen J}, \textbf{Bassis CM}, \textbf{Ramakrishnan S},
-\textbf{Hein R}, \textbf{Mason S}, \textbf{Bergman Y}, \textbf{Sunshine
-N}, \textbf{Fan Y}, \textbf{Timp W}, \textbf{Schatz MC}, \textbf{Young
-VB}, \textbf{Simner PJ}, \textbf{Bachman MA}. 2020. A plasmid locus
-associated with klebsiella clinical infections encodes a
-microbiome-dependent gut fitness factor.
-doi:\href{https://doi.org/10.1101/2020.02.26.963322}{10.1101/2020.02.26.963322}.
-
-\leavevmode\hypertarget{ref-Dieterle2018}{}%
-32. \textbf{Dieterle MG}, \textbf{Rao K}, \textbf{Young VB}. 2018. Novel
-therapies and preventative strategies for primary and
-recurrentClostridium difficileinfections. Annals of the New York Academy
-of Sciences \textbf{1435}:110--138.
-doi:\href{https://doi.org/10.1111/nyas.13958}{10.1111/nyas.13958}.
-
-\leavevmode\hypertarget{ref-Theriot2011}{}%
-33. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
-\textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
-Cefoperazone-treated mice as an experimental platform to assess
-differential virulence ofClostridium difficilestrains. Gut Microbes
-\textbf{2}:326--334.
-doi:\href{https://doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}.
+microbiota.} A, D. PCoAs of the Bray-Curtis distances from the stool (A)
+and tissue (D) communities from mock- and \emph{C. difficile}-challenged
+treatment groups. Each symbol represents a sample from an individual
+mouse with open and closed circles representing mock and \emph{C.
+difficile}-challenged mice, respectively. B, E. Median Shannon diversity
+in stool (B) and tissue (E) communities collected over time. C, F. The
+median relative abundances of genera that were significantly different
+between the \emph{C. difficile} challenged treatment groups in either
+the stool (Fig. 2E) or cecum tissue (Fig. 3C) communities in the stool
+(C) and tissue (F) communities from mock- and \emph{C.
+difficile}-challenged mice. For B-F, the dashed and solid lines
+represent the median value for mock and \emph{C. difficile}-challenged
+mice, respectively. For E-F, tissues from mock-challenged clindamycin
+treated mice were only collected 4 days post-challenge so there is no
+dashed line for this group. \newpage
 

### 2021-05-28 submission/manuscript.Rmd
Finish draft plus edits

@@
 When we trained a random forest model with community data from 5 dpc, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90).
-Five of the top ten bacterial features important for predicting prolonged colonization had high relative abundances in the community.
 Examining the dynamics of these bacterial during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C.
@@
 However, there are some microbiota similarities between diarrheal and *C.
-difficile* infection (CDI) communities such as lower diversity compared to healthy controls, which lead us to hypothesize that diarrhea may be an indicator of *C.
+difficile* infection (CDI) communities such as lower diversity compared to samples from healthy patients, which lead us to hypothesize that diarrhea may be an indicator of *C.
 difficile* risk.
@@
 difficile* by administering a laxative to mice either before or after *C.
-difficile* challenge Our findings suggest the osmotic laxative disrupts colonization resistance to *C.
+difficile* challenge.
+Our findings suggest the osmotic laxative disrupts colonization resistance to *C.
 difficile*, as well as clearance in mice already colonized with *C.
@@
 Our group has shown that when *C.
-difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C.
-difficile* negative with non-diarrheal consistency [@Schubert2014].
-These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.
-
-Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019].
+difficile* negative samples from patients are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and *C.
+difficile* negative patients with diarrhea compared to samples that were *C.
+difficile* negative with non-diarrheal stool consistency [@Schubert2014].
+These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.
+
+Osmotic laxatives can lead to diarrhea depending on the dose administered and temporarily disrupt the human intestinal microbiota [@Nagata2019].
 The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies.
@@
 difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020].
-Although, two groups have now shown PEG treatment alone renders mice susceptible to *C.
+Although two groups have now shown PEG treatment alone renders mice susceptible to *C.
 difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C.
 difficile* clearance.
-Addressing these questions will better inform how we think about laxatives in the context of CDIs.
 Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C.
@@
 difficile* clearance.
+Addressing these questions will better inform how we think about laxatives and diarrhea in the context of CDIs.
 
@@
 difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020].
-All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days: 1.
-received no additional treatment, 2.
-was also treated with clindamycin, and 3.
-was allowed to recover for 10 days prior to challenge (Fig.
+All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days.
+The first group received no additional treatment.
+The second group was also treated with clindamycin.
+A third group was allowed to recover for 10 days prior to challenge (Fig.
 1A).
-PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice relative to their baseline weights, with the greatest change in weight observed on the fifth day of PEG treatment.
+The PEG treatment resulted in weight loss for the 3 groups of mice, with the greatest change in weight observed on the fifth day of the PEG treatment.
 The mice recovered most of the lost weight by five days after treatment (Fig.
@@
 difficile* colonization.
-However, PEG-treated mice remained colonized with *C.
+In contrast to the mice that only received clindamycin, PEG-treated mice remained colonized with *C.
 difficile* at a high level through thirty days post-challenge (Fig.
 1C).
-In contrast, the clindamycin-treated mice cleared *C.
+The clindamycin-treated mice cleared *C.
 difficile* within ten days post-challenge (Fig.
 1C).
-Therefore, PEG treatment led to sustained colonization in contrast to clindamycin mice that naturally cleared *C.
-difficile* within ten days post-challenge.
-
-Notably, we also found PEG-treated mice were still susceptible to *C.
+It was noteworthy that PEG-treated mice were still susceptible to *C.
 difficile* colonization after a 10-day recovery period, although *C.
@@
 One mouse was found dead on the 6th day post-challenge, presumably due to *C.
-difficile* as the bacteria became detectable in stool samples from that mouse on the 4th day post-challenge (Fig.
+difficile* as the bacterium became detectable in stool samples from that mouse on the 4th day post-challenge (Fig.
 S1A, mouse 10).
@@
   pull(effects)
+#R2 for time (day)
+r2_day <- f_d_sign_contrib %>% 
+  filter(effects == "day") %>% 
+  pull(R2)
+#R2 for group
+r2_group <- f_d_sign_contrib %>% 
+  filter(effects == "group") %>% 
+  pull(R2)
 #Sum of R2 values for variables that significantly explained variation in stool samples
@@
 We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups.
-We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig.
-2A, Data Set S1, Sheet X).
-Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to coprophagy [@Nguyen2015].
-We tried to minimize the impact of cage effects on our experiment by breaking up littermates when assigning mice to treatment groups.
-Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group did no significantly explain the observed variation in fecal communities (Data Set S1, Sheet X).
+We found time (R^2^ = `r r2_day`) and treatment group (R^2^ = `r r2_group`) explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig.
+2A, Data Set S1, sheet 1).
 None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C.
@@
 Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-challenge (Fig.
-2C).
+2C; Data Set S1, sheet 2).
 We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment.
-We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X).
+We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, sheet 3).
 The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased.
@@
 Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig.
-2E, Data Set S1, Sheet X).
+2E, Data Set S1, sheet 4).
 Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.
 
-Interestingly, *C.
-difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge.
-We decided to examine if there were genera that changed during the post-challenge period when the group median *C.
+Because *C.
+difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge, we decided to examine if there were genera that changed during the post-challenge period.
+We compared the communities from when *C.
 difficile* shifted from undetectable at 1 day post-challenge to detectable in the stool samples with the density stabilizing around 8 days post-challenge (Fig.
 S1A).
-We found no bacteria with relative abundances that were significantly different over the two time points after multiple hypothesis correction (Data Set S1, Sheet X).
+We found no genera with relative abundances that were significantly different over the two time points (Data Set S1, sheet 5).
 However, there was also wide variation between individual mice regarding when *C.
@@
 S1B).
-For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period and this corresponded to mouse 10, which died on the sixth day post-challenge and mouse 11, where *C.
+For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period.
+One mouse died on the sixth day post-challenge and in the other *C.
 difficile* was present at a high density from the 4th day post-challenge onward (Fig.
@@
 While we did not identify a clear signal to explain the delayed appearance of *C.
-difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.
+difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay was striking and could reflect changes in microbial activity or metabolites that were not examined in this study.
 
@@
   pull(effects)
+#R2 for time (day)
+r2_day_t <- f_d_sign_contrib_t %>% 
+  filter(effects == "day") %>% 
+  pull(R2)
+#R2 for group
+r2_group_t <- f_d_sign_contrib_t %>% 
+  filter(effects == "group") %>% 
+  pull(R2)
 #Sum of R2 values for variables that significantly explained variation in tissue samples
@@
 **5-day laxative treatment did not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity.
-To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon.
-Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig.
+To evaluate the mucosal microbiota, we sequenced communities associated with tissues collected from the cecum, proximal colon, and distal colon.
+Similar to what was observed with the stool samples, the alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig.
+3A, Data Set S1, sheet 6).
+The alpha diversity of the tissue-associated community increased in PEG-treated mice collected at 20 and 30 days post-challenge (Fig.
 3A).
-Alpha diversity continued to increase over time based on the tissues from PEG-treated mice collected at 20 and 30 days post-challenge (Fig.
-3A, Data Set S1, Sheet X).
-Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig.
-3B, Data Set S1, Sheet X).
+Group (R^2^ = `r r2_group_t`), time point (R^2^ = `r r2_day_t`), and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig.
+3B, Data Set S1, sheet 7).
 We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)`) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig.
-S2A, Data Set S1, Sheet X).
+S2A, Data Set S1, sheet 8, 9, and 10).
 Interestingly, *Peptostreptococcaceae* (the family with a sequence that matches *C.
@@
 3C, Fig.
-2E, Data Set S1, Sheet X).
+2E, Data Set S1, sheet 8).
 Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C.
@@
 difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C.
-difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011].
+difficile* challenge by evaluating cecum and colon histopathology [@Reeves2011].
 However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C.
@@
 1C).
-Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5-2.5 out of 12, Fig.
+Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig.
 3E).
@@
 difficile* challenge did not have a synergistic disruptive effect on the microbiota of PEG-treated mice.** Because *C.
-difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice.
+difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged mice treated with clindamycin or PEG.
 Examining the stools of the mock-challenged mice revealed similar bacterial disruptions as the *C.
@@
 S3A-C).
-Similarly, there was no difference between the tissues of mock and *C.
+Similarly, there was no difference between the communities found in the tissues of mock and *C.
 difficile* challenged mice (Fig.
@@
 3A).
-In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C.
+In contrast to the 5-day PEG treated mice, the 1-day PEG groups were only transiently colonized and cleared *C.
 difficile* by 7 days post-challenge (Fig.
 3B).
-The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering by 7 days post-challenge (Fig.
-3C-D).
+The stool communities of the 1-day PEG treatment groups were also only transiently disrupted, with Shannon diversity recovering by 7 days post-challenge (Fig.
+3C-D, Data Set S1, sheets 11 and 12).
 We found the relative abundances of `r recovered_genera` genera were impacted by treatment, but recovered close to baseline levels by 7 days post-challenge including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig.
-3E, Data Set S1, Sheet X).
+3E, Data Set S1, sheet 13 and 14).
 These findings suggest the duration of the PEG treatment was relevant, with shorter treatments resulting in a transient loss of *C.
@@
 ```
-**Post-challenge laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
+**Post-challenge laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild perturbation of the microbiota, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C.
 difficile* spores from the intestine.
+This hypothesis is proposed in the discussion section of FMT studies where bowel prep is part of the preparation undergone by patients receiving FMTs via colonoscopy [@vanNood2013; @Razik2017; @Postigo2012; @Liacouras1996].
 To examine the impact of PEG treatment on *C.
@@
 5A).
-Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C.
+Contrary to the hypothesis, all groups of mice that received PEG exhibited prolonged *C.
 difficile* colonization (Fig.
@@
 We were also interested in exploring whether PEG might help with engraftment in the context of FMTs.
+An FMT was prepared under anaerobic conditions using stool collected from the same group of mice pre-treatment representing the baseline community.
 The FMT appeared to partially restore Shannon diversity but not richness (Fig.
-5C-D).
+5C-D, Data Set S1, sheets 15 and 16).
 Similarly, we saw some overlap between the communities of mice that received FMT and the mice treated with only clindamycin after 5 days post-challenge (Fig.
-6A).
+6A, Data Set S1, sheet 17).
 The increase in Shannon diversity suggests that the FMT did have an impact on the microbiota, despite seeing prolonged *C.
 difficile* colonization in the FMT treated mice.
-However, only the relative abundances of *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice received either an FMT or PBS gavage (Fig.
-6B), suggesting the FMT only restored a couple of genera.
-Overall, we found the relative abundances of `r post_group_multi_day` genera were different between treatment groups over multiple timepoints.
+However, only the relative abundances of *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice that received either an FMT or PBS gavage (Fig.
+6B).
+Overall, we found the relative abundances of `r post_group_multi_day` genera were different between treatment groups over multiple time points (Data Set S1, sheet 18).
 For example, the relative abundance of *Akkermansia* was increased and the relative abundances of *Ruminococcaceae*, *Clostridiales*, *Lachnospiraceae*, and *Oscillibacter* were decreased in mice that received PEG after *C.
@@
 ```
-**Five-day post-challenge community data can predict which mice that will have prolonged *C.
-difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C.
+**Five-day post-challenge community data can predict which mice will have prolonged *C.
+difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG treatments, we examined the bacteria that influenced prolonged *C.
 difficile* colonization.
-We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with bacterial community data from 5 days post-challenge to predict whether the mice were still colonized with *C.
+We trained 3 machine learning models (random forest, logistic regression, and support vector machine) with bacterial community data from 5 days post-challenge to predict whether the mice were still colonized with *C.
 difficile* 10 days post-challenge.
-We chose 5 days post-challenge because that was the earliest time point where we would see a treatment effect in the post-challenge 1-day PEG experiments.
-The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C.
+We chose 5 days post-challenge because that was the earliest time point where we saw a treatment effect in the post-challenge 1-day PEG experiments.
+The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, sheet 19), suggesting the 5-day post challenge microbiota is an excellent predictor of prolonged *C.
+difficile* colonization.
+Next, we performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C.
 difficile *colonization.
-We selected the top 10 bacteria contributing to our model's performance (Fig.
+We selected 10 genera that contributed the most to our model's performance (Fig.
 7A) and examined their relative abundance at 5 days post-challenge, the time point used to predict *C.
@@
 7B).
-Next, we focused on the 5 genera that had a greater than 1 % relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time.
+Next, we focused on the 5 genera that had a greater than 1% relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time.
 We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, the relative abundance of *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
@@
 difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear.
-By following osmotic laxative treated mice over time, we found 5-day PEG treatment before challenge resulted in prolonged *C.
+By following osmotic laxative treated mice over time, we found that a 5-day PEG treatment before challenge resulted in prolonged *C.
 difficile* colonization, while a 1-day PEG treatment resulted in transient colonization.
@@
 difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota.
-Additionally, the intestinal communities of 5-day PEG treated mice had not regained colonization resistance after a 10-day recovery period.
-Although in contrast to the other 5-day PEG treatment groups, *C.
+Additionally, the intestinal communities of 5-day PEG treated mice did not regain colonization resistance after a 10-day recovery period.
+In contrast to the other 5-day PEG treatment groups, *C.
 difficile* was not immediately detectable in the stools of most of the mice in the 10-day recovery group.
 We also examined the impact of PEG treatment after *C.
-difficile* challenge and in opposition to the hypothesis suggested by the literature, found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment.
+difficile* challenge.
+In opposition to the hypothesis suggested by the literature, we found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment.
 We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C.
@@
 For example, *Oscillibacter valericigenes* can produce the SCFA valerate [@Iino2007], and separate studies demonstrated valerate is depleted after clindamycin treatment and inhibited *C.
-difficile* growth based on *in vitro* data and an experiment with C57BL/6 mice [@Jump2014; @McDonald2018].
+difficile* growth *in vitro* and in C57BL/6 mice [@Jump2014; @McDonald2018].
 Similarly, *Acetatifactor* can produce acetate and butyrate [@Pfeiffer2012], SCFAs that are decreased in mice with prolonged *C.
@@
 difficile* biofilms may serve as reservoirs for recurrent infections [@Frost2021] and *C.
-difficile* biofilms were recently identified in patients and aggregates with *Fusobacterium nucleatum*, another species capable of forming biofilms [@Engevik2021].
+difficile* biofilms in the mucus layer were recently identified in patients as aggregates with *Fusobacterium nucleatum* [@Engevik2021].
 There was an interesting pattern of increased *Enterobacteriaceae*, *Bacteroides*, and *C.
@@
 difficile* was present in the outer mucus layer and associated with *Enterobacteriaceae* and *Bacteroidaceae* using fluorescent in situ hybridization (FISH) staining [@Semenyuk2015].
-However, *B.
+However, protective roles for *Bacteroides* have also been demonstrated.
+For example, *B.
 fragilis* prevented CDI morbidity in a mouse model and inhibited *C.
@@
 difficile* biofilm formation while *B.
-thetaiotamicron* enhanced biofilm formation [@Normington2021].
-Therefore, whether *Bacterodes* is detrimental or beneficial in the context of *C.
-difficile* infection or colonization is still unclear.
+thetaiotamicron* enhanced biofilm formation [@Normington2021] and *B.
+dorei* reduced *C.
+difficile* growth in a 9-species community *in vitro* [@Hassall2021].
+Therefore, whether *Bacteroides* is detrimental or beneficial in the context of *C.
+difficile* infection or colonization is still unclear, but the niche and interactions with other bacteria may contribute.
 
 *Akkermansia* is also a mucin degrader with potentially beneficial or detrimental roles depending on context in other diseases [@Cirstea2018; @BorgesCanha2015].
-In our study the relative abundance of *Akkermansia* shifted over time between groups.
+In our study the relative abundance of *Akkermansia* shifted over time between groups of mice that either cleared *C.
+difficile* or had prolonged colonization.
 In the stool it was initially increased in mice that cleared *C.
 difficile*, but shifted after 5-days post-challenge so that it was increased in mice that had prolonged colonization.
-In the context of CDIs, some studies suggest a protective role [@Pereira2020; @Rodriguez2016], while others suggest detrimental [@Sangster2016; @Vakili2020a; @Vakili2020b].
-Because the relative abundance of *Akkermansia* was dynamic in our study so it is unclear whether *Akkermansia* helps with clearance of *C.
+In the context of CDIs, some studies suggest a protective role [@Pereira2020; @Rodriguez2016], while others suggest a detrimental role because *Akkermansia* was positively correlated with *C.
+difficile* [@Sangster2016; @Vakili2020a; @Vakili2020b; @Li2019].
+Because the relative abundance of *Akkermansia* was dynamic in our study, it is unclear whether *Akkermansia* helps with clearance of *C.
 difficile* or allows it to persist.
@@
 difficile* interacts with the mucosal microbiota may lead to insights into CDIs, asymptomatic *C.
-difficile* colonization, and resistance.
+difficile* carriage, and colonizatiion resistance.
 
@@
 
-It is more difficult what are findings mean in the context of *C.
-difficile* colonization resistance mean for human patients.
-Considering that most hospitals recommend not performing *C.
-difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs.
-further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.
+It is more difficult to interpret what are findings mean in the context of *C.
+difficile* colonization resistance in human patients.
+The main difficulty being that most hospitals recommend not performing *C.
+difficile* testing if the patient is currently taking a laxative.
+This recommendation is in accordance with the Infectious Diseases Society of America and Society for Healthcare Epidemiology of America guidelines [@McDonald2018b].
+The rationale behind the recommendation is that patients taking laxatives may be asymptomatically colonized with *C.
+difficile*, resulting in unnecessary antibiotic treatment [@Ahmad2017; @Bilinskaya2018; @Cook2020].
+Furthermore, some studies identified laxatives as a risk factor for developing CDIs or recurrent CDIs [@Appaneal2019; @Dubberke2011; @Obritsch2010] and a recent study found the proportion of severe CDIs was similar between patients taking and not taking laxatives [@White2019].
+However, there have also been some studies that suggest laxatives are not a risk factor for developing CDIs [@Poirier2019; @Huang2014].
+Although, it is unclear whether laxatives impact CDI susceptibility in human paitents, it is clear that laxatives also have a transient impact on the human microbiota [@Nagata2019; @Gorkiewicz2013; @Shobar2016; @Drago2016; @Jalanka2014].
+Further studies to examine the relationship between laxatives, *C.
+difficile* colonization, and CDIs are warranted.
+
+Considering laxatives are also used to prepare patients when administering fecal microbiota transplants via colonoscopy to treat recurrent CDIs, it will be important to determine if osmotic laxatives impact *C.
+difficile* clearance in the human intestinal tract.
+It is still unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs [@Kampouri2021].
+Nevertheless, results from the FMT National Registry where 85% of FMTs were delivered by colonoscopy demonstrate FMTs are highly effective treatment for recurrent CDIs with 90% achieving resolution in the 1 month follow-up window [@Kelly2021].
+A surprising number of studies continue to hypothesize that PEG or bowel preparation can clear *C.
+difficile* spores and toxins despite the paucity of supporting evidence [@Liacouras1996; @vanNood2013; @Razik2017; @Postigo2012].
+There was even a clinical trial (NCT01630096) designed to examine whether administering PEG 3350 (NuLYTELY) prior to antibiotic treatment reduced disease severity that started recruitment in 2012  [@Dieterle2018], but no results have been posted to date.
+Here we sought to evaluate the impact of treating *C.
+difficile* colonized mice with PEG (with or without FMT) and found clearance was delayed.
 Further studies are needed to understand the impact of osmotic laxatives on *C.
-difficile* colonization resistance and clearance in human patients.
-+ What's known regarding laxatives and susceptibility to CDIs
-  + Clinical trial of PEG, results never posted [@Dieterle2018]
-
-Relevance to human FMTs?
-Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.
-However, results from the FMT National Registry where 85% of FMTs were delivered by colonoscopy demonstrate FMTs are highly effective treatment for recurrent CDIs with 90% achieving resolution in the 1 month follow-up window [@Kelly2021].
+difficile* colonization resistance and clearance in human patients receiving FMTs.
 
@@
 We thank members of the Schloss lab for feedback on planning the experiments and data presentation.
-We thank Andrew Henry for help with media preparation and bacterial culture and the Microbiology and Immunology department's postdoc association writing group members for their feedback on manuscript drafts.
+We thank Andrew Henry for help with media preparation and bacterial culture and the Microbiology and Immunology department's postdoctoral association writing group members for their feedback on manuscript drafts.
 We also thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the institutional support for our mouse experiments.
@@
 This work was supported by the National Institutes of Health (U01AI124255).
-ST was supported by the Michigan Institute for Clincial and Health Research Postdoctoral Translation Scholars Program (UL1TR002240 from the National Center for Advancing Translational Sciences).
+ST was supported by the Michigan Institute for Clinical and Health Research Postdoctoral Translation Scholars Program (UL1TR002240 from the National Center for Advancing Translational Sciences).
 
@@
 All mice were C57Bl/6 and part of the Schloss lab colony which was established in 2010 with mice donated from Vincent Young's lab colony (established with mice purchased from The Jackson Laboratory in 2002).
-We used 7-19 week old female mice for all experiments, which allowed us to break up littermates prior to starting an experiment and distribute them as evenly as possible across treatment groups in order to minimize microbiota differences between experiment groups prior to starting treatments with medications.
+We used 7-19 week old female mice for all experiments.
+This allowed us to break up littermates and distribute them as evenly as possible across treatment groups in order to minimize microbiota differences prior to starting treatments with medications.
 During the experiment, mice were housed at a density of 2-3 mice per cage, with the majority of cages limited to two mice.
@@
 difficile* challenge model.** Mice were challenged with 25 microliters off *C.
-difficile* 630 spores at 10^5^ concentration, except for 1 experiment (Fig.
-5A) where the concentration was 10^3^.
+difficile* 630 spores at 10^5^ concentration, except for 1 experiment where the concentration was 10^3^ (Fig.
+5A).
 All mock challenged mice received 25 ul vehicle solution (Ultrapure water).
@@
 For B-C, lines represent the median for each treatment group and circles represent samples from individual mice.
-Asterisks indicate time points where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
+Asterisks indicate time points where the weight change or CFU/g was significantly different (*P* < 0.05) between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
 The data presented are from a total of `r five_day_total_exp` separate experiments.
@@
 Each circle represents a sample from an individual mouse and the transparency of the symbol corresponds to the day post-challenge.
+See Data Set S1, sheet 1 for PERMANOVA results.
 B.
 Bray-Curtis distances of stool samples collected on either day 10 or 30 post-challenge relative to the baseline sample collected for each mouse (before any drug treatments were administered).
+The symbols represent samples from individual mice and the line indicates the median value for each treatment group.
 C.
 Shannon diversity in stool communities over time.
-The line indicates the median value for each treatment group.
-For B-C, the symboles represent samples from individual mice and the line indicates the median value for each treatment group.
+The line indicates the median value for each treatment group (Data Set S1, sheet 2).
 D.
-14 of the `r f_d_group_total_genera` genera affected by PEG treatment  (Data Set S1, sheet X).
+14 of the `r f_d_group_total_genera` genera affected by PEG treatment  (Data Set S1, sheet 3).
 The symbols represent the median relative abundance for a treatment group at either baseline (open circle) or 1-day post treatment (closed circle).
@@
 E.
-6 of the `r f_d_group_multi_day` genera that were significantly different between the treatment groups over multiple time points, the 5-day PEG plus clindamycin treatment group was only followed through 6-days post-challenge.
-Differences between treatment groups were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified genera.
+6 of the `r f_d_group_multi_day` genera that were significantly different between the treatment groups over multiple time points (see Data Set S1, sheet 4 for complete list).
+The 5-day PEG plus clindamycin treatment group was only followed through 6-days post-challenge.
+Differences between treatment groups were identified by Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified genera (\*\, *P* < 0.05).
 The gray vertical line (D) and horizontal vertical lines (E) indicate the limit of detection.
@@
 Shannon diversity in cecum communities over time.
-The line indicates the median value for each treatment group.
-The colors of the symbols and lines represent individual and median relative abundance values for four treatment groups.
-A similar pattern was observed with the proximal and distal colon communities (Data Set S1, sheet X-X).
+The colors of the symbols and lines represent individual and median relative abundance values for four treatment groups (Data Set S1, sheet 6).
 B.
@@
 Transparency of the symbol corresponds to the day post-challenge that the sample was collected.
+See Data Set S1, sheet 7 for PERMANOVA results.
 C.
-The median relative abundance of the 4 genera that were significantly different between the cecum communities of different treatment groups on day 6 and day 30 post-challenge (Data Set S1, sheet X).
+The median relative abundance of the 4 genera that were significantly different between the cecum communities of different treatment groups on day 6 and day 30 post-challenge (Data Set S1, sheet 8).
 The gray vertical lines indicate the limit of detection.
@@
 Histology data were analyzed with the Kruskal-Wallis test followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
+\*\, *P* < 0.05.
 \newpage
@@
 difficile* CFU/gram stool measured over time (N = `r one_day_sample_min`-`r one_day_sample_max` mice per time point) by serial dilutions.
-The black dashedd horizontal line represents the limit of detection for the first serial dilution.
-For B and D, asterisks indicate time points where there was a significant difference between treatment groups by Kruskall-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
+The black dashed horizontal line represents the limit of detection for the first serial dilution.
+For B and D, asterisks indicate time points where there was a significant difference (*P* < 0.05) between treatment groups by Kruskall-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
 For B-D, each symbol represents a sample from an individual mouse and lines indicate the median value for each treatment group.
 C.
-PCoA of Bray-Curtis distances from stool communities collected over time (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`).
+PCoA of Bray-Curtis distances from stool communities collected over time (`r one_day_top_contrib_name`: R^2^ = `r top_R2`; `r one_day_sec_contrib_name`: R^2^ = `r sec_R2`, Data Set S1, sheet 11).
 Symbol transparency represents the day post-challenge of the experiment.
-For C-E, the B on the day legend or days post-challenge X axis stands for baseline and represents the sample that was collected prior to any drug treatments.
+For C-E, the B on the day legend or days post-challenge X-axis stands for baseline and represents the sample that was collected prior to any drug treatments.
 D.
-Shannon diversity in stool communities over time.
+Shannon diversity in stool communities over time (Data Set S1, sheet 12).
 E.
 Median relative abundances per treatment group for 6 out of the `r recovered_genera` genera that were affected by treatment, but recovered close to baseline levels by 7 days post-challenge (Fig.
-3E, Data Set S1, Sheet X).
+3E, Data Set S1, sheets 13 and 14).
 Paired stool sample relative abundance values either baseline and day 1 or baseline and day 7 were analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing all identified genera.
-Only genera that were different between baseline and 1-day post-challenge, but not baseline and 7-days post-challenge.
+Only genera that were different between baseline and 1-day post-challenge, but not baseline and 7-days post-challenge are shown.
 The gray horizontal lines represents the limit of detection.
@@
 C-D.
-Shannon diversity (C) and richness (D) in stool communities over time.
+Shannon diversity (C) and richness (D) in stool communities over time (Data Set S1, sheets 15 and 16).
 B-D.
 Each symbol represents a stool sample from an individual mouse with the lines representing the median value for each treatment group.
-Asterisks indicate time points with significant differences between groups by a Kruskall-Wallis test with a Benjamini-Hochberg correction for testing multiple times points.
+Asterisks indicate time points with significant differences (*P* < 0.05) between groups by the Kruskall-Wallis test with a Benjamini-Hochberg correction for testing multiple times points.
 Colored rectangles indicates the 1-day PEG treatment period for applicable groups.
@@
 Each circle represents an individual sample, the transparency of the circle corresponds to day post-challenge.
+See Data Set S1, sheet 17 for PERMANOVA results.
 B.
 Median relative abundances of 2 genera that were significantly different over multiple time points in mice that were administered either FMT or PBS solution via gavage C.
-Median relative abundances of the top 6 out of `r post_group_multi_day` genera that were significant over multiple timepoints, plotted over time (Data Set S1, Sheet X).
+Median relative abundances of the top 6 out of `r post_group_multi_day` genera that were significant over multiple time points, plotted over time (see Data Set S1, sheet 18 for complete list).
 For B-C, colored rectangles indicates the 1-day PEG treatment period for applicable groups.
@@
 difficile* colonization in PEG treated mice.** A.
-Top ten bacteria that contributed to the random forest model trained on five day post-challenge community relative abundance data to predict whether mice would still be colonized with *C.
+Top ten bacteria that contributed to the random forest model trained on 5-day post-challenge community relative abundance data, predicting whether mice would still be colonized with *C.
 difficile* 10 days post-challenge.
-The median (point) and interquartile range (lines) change in AUROC when the bacteria is left out of the model by permutation feature importance analysis.
+The median (point) and interquartile range (lines) change in AUROC when the bacteria were left out of the model by permutation feature importance analysis.
 B.
 The median relative abundances of the top ten bacteria that contributed to the random forest classification model at 5 days post-challenge .
-Color indicates whether the mice were still colonized with *C.
+Red indicates the mice were still colonized with *C.
+difficile* while blue indicates mice that cleared *C.
 difficile* 10 days post-challenge and the black horizontal line represents the median relative abundance for the two categories.
@@
 Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-challenge.
-However, for mice that received either a 5-day PEG treatment pre-*C.
+However, for mice that received either a 5-day PEG treatment prior to *C.
 difficile* challenge or a 1-day PEG treatment post-challenge recovery of colonization resistance was delayed because most mice were still colonized with *C.
@@
 B.
-Relative abundances of eight bacterial genera from day 0 post-challenge onward in each off the 10-day recovery mice.
-We analyzed samples from day 0 and day 8 post-challenge, which represented the the time points where mice were challenged with *C.
+Relative abundances of eight bacterial genera from day 0 post-challenge onward in each of the 10-day recovery mice.
+We analyzed samples from day 0 and day 8 post-challenge, which represented the time points where mice were challenged with *C.
 difficile* and when the median relative *C.
-difficile* CFU stabilized for the group using the paired Wilcoxan signed-rank test, but no genera were significantly different after Benjamini-Hochberg correction.
+difficile* CFU stabilized for the group using the paired Wilcoxan signed-rank test, but no genera were significantly different after Benjamini-Hochberg correction (Data Set S1, sheet 5).
 \newpage
@@
 PEG treatment still has a large impact on the mucosal microbiota 6 days post-challenge** A.
-The relative abundances of the 10 bacterial genera that were significantly different between treatment groups at 6 days post-infection in the cecum tissue (the relative abundances of the 10 genera were also significantly different in the proximal and distal colon tissues, Data Sheet S1, Sheet X).
+The relative abundances of the 10 bacterial genera that were significantly different between treatment groups at 6 days post-infection in the cecum tissue (the relative abundances of the 10 genera were also significantly different in the proximal and distal colon tissues, Data Set S1, sheets 8, 9, and 10).
 Each symbol represents a tissue sample from an individual mouse, the black horizontal lines represents the median relative abundances for each treatment group.
@@
 \newpage
+
+## Data Set S1
+**Data Set S1, Sheets 1-19. Excel workbook with 19 sheets.**
+
+**Data Set S1, Sheet 1. PERMANOVA results for the stool communities from mice in the 5-day PEG subset.**
+
+**Data Set S1, Sheet 2. Shannon diversity analysis for the stool communities from mice in the 5-day PEG subset.**
+
+**Data Set S1, Sheet 3. Genera with relative abundances impacted by PEG treatment based on stool communities of 5-day PEG treated mice.**
+
+**Data Set S1, Sheet 4. Genera with relative abundances that vary between treatment groups in the stool communities from mice in the 5-day PEG subset.**
+
+**Data Set S1, Sheet 5.
+Statistical analysis results for genera with relative abundances that varied in stool communities in the 5-day PEG plus 10-day recovery mice between the day 1 and day 8 time points.**
+
+**Data Set S1, Sheet 6. Shannon diversity analysis for the cecum communities from mice in the 5-day PEG experiments.**
+
+**Data Set S1, Sheet 7. PERMANOVA results for the tissue communities from mice in the 5-day PEG subset.**
+
+**Data Set S1, Sheet 8. Genera with relative abundances that vary between treatment groups in the cecum communities from mice in the 5-day PEG esubset.**
+
+**Data Set S1, Sheet 9. Genera with relative abundances that vary between treatment groups in the proximal colon communities from mice in the 5-day PEG subset.**
+
+**Data Set S1, Sheet 10. Genera with relative abundances that vary between treatment groups in the distal colon communities from mice in the set of 5-day PEG subset.**
+
+**Data Set S1, Sheet 11. PERMANOVA results for the stool communities from mice in the set of 1-day PEG subset.**
+
+**Data Set S1, Sheet 12. Shannon diversity analysis for the stool communities from mice in the 1-day PEG experiments.**
+
+**Data Set S1, Sheet 13. Genera with different relative abundances between the baseline and day 1 time points in the 1-day PEG subset.**
+
+**Data Set S1, Sheet 14. Genera with different relative abundances between the baseline and day 7 time points in the 1-day PEG subset..**
+
+**Data Set S1, Sheet 15. Shannon diversity analysis for the stool communities from mice in the post-challenge PEG experiments.**
+
+**Data Set S1, Sheet 16. Richness analysis for the stool communities from mice in the post-challenge PEG experiments.**
+
+**Data Set S1, Sheet 17. PERMANOVA results for the stool communities from mice in the post-challenge PEG subset.**
+
+**Data Set S1, Sheet 18. Genera with relative abundances that vary between treatment groups in the stool communities from mice in the post-challenge PEG subset.**
+
+**Data Set S1, Sheet 19. AUROC results for the 100 different seeds from each of the 3 models tested.**

### 2021-05-28 submission/manuscript.tex
Finish draft plus edits

@@
 from 5 dpc, we were able to predict which mice would exhibit prolonged
-colonization (AUROC = 0.90). Five of the top ten bacterial features
-important for predicting prolonged colonization had high relative
-abundances in the community. Examining the dynamics of these bacterial
+colonization (AUROC = 0.90). Examining the dynamics of these bacterial
 during the post-challenge period revealed patterns in the relative
@@
 some microbiota similarities between diarrheal and \emph{C. difficile}
-infection (CDI) communities such as lower diversity compared to healthy
-controls, which lead us to hypothesize that diarrhea may be an indicator
-of \emph{C. difficile} risk. We explored how osmotic laxatives disrupt
-the microbiota's colonization resistance to \emph{C. difficile} by
-administering a laxative to mice either before or after \emph{C.
-difficile} challenge Our findings suggest the osmotic laxative disrupts
-colonization resistance to \emph{C. difficile}, as well as clearance in
-mice already colonized with \emph{C. difficile}. Considering that most
-hospitals recommend not performing \emph{C. difficile} testing on
-patients taking laxatives and laxatives are used when administering
-fecal microbiota transplants via colonoscopy to patients with recurrent
-CDIs, further studies are needed to evaluate if laxatives impact
-microbiota colonization resistance in humans.
+infection (CDI) communities such as lower diversity compared to samples
+from healthy patients, which lead us to hypothesize that diarrhea may be
+an indicator of \emph{C. difficile} risk. We explored how osmotic
+laxatives disrupt the microbiota's colonization resistance to \emph{C.
+difficile} by administering a laxative to mice either before or after
+\emph{C. difficile} challenge. Our findings suggest the osmotic laxative
+disrupts colonization resistance to \emph{C. difficile}, as well as
+clearance in mice already colonized with \emph{C. difficile}.
+Considering that most hospitals recommend not performing \emph{C.
+difficile} testing on patients taking laxatives and laxatives are used
+when administering fecal microbiota transplants via colonoscopy to
+patients with recurrent CDIs, further studies are needed to evaluate if
+laxatives impact microbiota colonization resistance in humans.
 
@@
 serves as an approximation of intestinal motility (10). Our group has
-shown that when \emph{C. difficile} negative controls are separated into
-two groups based on stool consistency, there are shared microbiota
-features such as lower alpha diversity in samples from CDI patients and
-control patients with diarrhea compared to control samples that were
-\emph{C. difficile} negative with non-diarrheal consistency (12). These
-results led to a hypothesis that bacterial communities from patients
-experiencing diarrhea are susceptible to developing CDIs.
-
-Osmotic laxatives can lead to diarrhea depending on the administered
-dose and temporarily disrupt the human intestinal microbiota (13). The
-ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
-Miralax, Nulytely, and Golytely and is also commonly used as bowel
-preparation for colonoscopies. Interestingly, previous studies have
-shown that treating mice with PEG alone altered microbiota composition,
-reduced acetate and butyrate production, altered the mucus barrier, and
-rendered the mice susceptible to \emph{C. difficile} colonization
-(14--17). The mucus barrier is thought to mediate protection from CDIs
-by protecting intestinal epithelial cells from the toxins produced by
-\emph{C. difficile} (18, 19). Whether laxative administration results in
-more severe CDIs in mice and how long mice remain colonized with
-\emph{C. difficile} after challenge is unclear.
+shown that when \emph{C. difficile} negative samples from patients are
+separated into two groups based on stool consistency, there are shared
+microbiota features such as lower alpha diversity in samples from CDI
+patients and \emph{C. difficile} negative patients with diarrhea
+compared to samples that were \emph{C. difficile} negative with
+non-diarrheal stool consistency (12). These results led to the
+hypothesis that bacterial communities from patients experiencing
+diarrhea are susceptible to developing CDIs, regardless of how they
+developed diarrhea.
+
+Osmotic laxatives can lead to diarrhea depending on the dose
+administered and temporarily disrupt the human intestinal microbiota
+(13). The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is
+found in Miralax, Nulytely, and Golytely and is also commonly used as
+bowel preparation for colonoscopies. Interestingly, previous studies
+have shown that treating mice with PEG alone altered microbiota
+composition, reduced acetate and butyrate production, altered the mucus
+barrier, and rendered the mice susceptible to \emph{C. difficile}
+colonization (14--17). The mucus barrier is thought to mediate
+protection from CDIs by protecting intestinal epithelial cells from the
+toxins produced by \emph{C. difficile} (18, 19). Whether laxative
+administration results in more severe CDIs in mice and how long mice
+remain colonized with \emph{C. difficile} after challenge is unclear.
 
@@
 the microbiota and influence \emph{C. difficile} susceptibility and
-clearance (24--26). Although, two groups have now shown PEG treatment
+clearance (24--26). Although two groups have now shown PEG treatment
 alone renders mice susceptible to \emph{C. difficile} (15, 17), these
@@
 severity of infection as well as the role of laxative treatment in
-\emph{C. difficile} clearance. Addressing these questions will better
-inform how we think about laxatives in the context of CDIs. Here, we
-characterized how long PEG-treated mice remain susceptible, whether PEG
-treatment results in more sustained \emph{C. difficile} colonization and
-severe CDI than mice treated with clindamycin, and if PEG treatment
-after challenge can promote \emph{C. difficile} clearance.
+\emph{C. difficile} clearance. Here, we characterized how long
+PEG-treated mice remain susceptible, whether PEG treatment results in
+more sustained \emph{C. difficile} colonization and severe CDI than mice
+treated with clindamycin, and if PEG treatment after challenge can
+promote \emph{C. difficile} clearance. Addressing these questions will
+better inform how we think about laxatives and diarrhea in the context
+of CDIs.
 
@@
 post-challenge (9, 26). All three groups of PEG-treated mice were
-administered a 15\% PEG solution in the drinking water for 5-days: 1.
-received no additional treatment, 2. was also treated with clindamycin,
-and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A).
-PEG treatment resulted in weight loss in all 3 groups of PEG-treated
-mice relative to their baseline weights, with the greatest change in
-weight observed on the fifth day of PEG treatment. The mice recovered
-most of the lost weight by five days after treatment (Fig. 1B). After
-either the PEG, clindamycin, or PEG and clindamycin treatment all mice
-were challenged with 10\textsuperscript{5} \emph{C. difficile} 630
-spores (Fig. 1A). All treatments rendered mice susceptible to \emph{C.
-difficile} colonization. However, PEG-treated mice remained colonized
-with \emph{C. difficile} at a high level through thirty days
-post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice
-cleared \emph{C. difficile} within ten days post-challenge (Fig. 1C).
-Therefore, PEG treatment led to sustained colonization in contrast to
-clindamycin mice that naturally cleared \emph{C. difficile} within ten
-days post-challenge.
-
-Notably, we also found PEG-treated mice were still susceptible to
-\emph{C. difficile} colonization after a 10-day recovery period,
-although \emph{C. difficile} was not detectable in most of the group in
-the initial five days post-challenge (Fig. 1C, S1A). One mouse was found
-dead on the 6th day post-challenge, presumably due to \emph{C.
-difficile} as the bacteria became detectable in stool samples from that
-mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days
-post-challenge onward, the density of \emph{C. difficile} stabilized in
-the 10-day recovery group and remained high through 20-30 days
-post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was
-sufficient to render mice susceptible to prolonged \emph{C. difficile}
-colonization and PEG-treated mice remained susceptible through ten days
-post-treatment.
+administered a 15\% PEG solution in the drinking water for 5-days. The
+first group received no additional treatment. The second group was also
+treated with clindamycin. A third group was allowed to recover for 10
+days prior to challenge (Fig. 1A). The PEG treatment resulted in weight
+loss for the 3 groups of mice, with the greatest change in weight
+observed on the fifth day of the PEG treatment. The mice recovered most
+of the lost weight by five days after treatment (Fig. 1B). After either
+the PEG, clindamycin, or PEG and clindamycin treatment all mice were
+challenged with 10\textsuperscript{5} \emph{C. difficile} 630 spores
+(Fig. 1A). All treatments rendered mice susceptible to \emph{C.
+difficile} colonization. In contrast to the mice that only received
+clindamycin, PEG-treated mice remained colonized with \emph{C.
+difficile} at a high level through thirty days post-challenge (Fig. 1C).
+The clindamycin-treated mice cleared \emph{C. difficile} within ten days
+post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were
+still susceptible to \emph{C. difficile} colonization after a 10-day
+recovery period, although \emph{C. difficile} was not detectable in most
+of the group in the initial five days post-challenge (Fig. 1C, S1A). One
+mouse was found dead on the 6th day post-challenge, presumably due to
+\emph{C. difficile} as the bacterium became detectable in stool samples
+from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From
+8 days post-challenge onward, the density of \emph{C. difficile}
+stabilized in the 10-day recovery group and remained high through 20-30
+days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone
+was sufficient to render mice susceptible to prolonged \emph{C.
+difficile} colonization and PEG-treated mice remained susceptible
+through ten days post-treatment.
 
@@
 V4 region of the 16S rRNA gene to compare changes across treatment
-groups. We found time and treatment group explained half of the observed
-variation between fecal communities with most of the remaining variation
-explained by interactions between treatment group and other experimental
-variables including time, cage effects, and sequencing preparation plate
-(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
-0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
-well-documented phenomenon that mice housed in the same cages have
-similar microbial communities due to coprophagy (27). We tried to
-minimize the impact of cage effects on our experiment by breaking up
-littermates when assigning mice to treatment groups. Importantly,
-although we conducted a total of 5 separate experiments, the experiment
-number and its interaction with treatment group did no significantly
-explain the observed variation in fecal communities (Data Set S1, Sheet
-X). None of the treatment groups recovered to their baseline community
-structure either 10 or 30 days post-challenge suggesting other community
-features besides recovery to baseline were responsible for the prolonged
-\emph{C. difficile} colonization in PEG-treated mice (Fig. 2B).
+groups. We found time (R\textsuperscript{2} = 0.29) and treatment group
+(R\textsuperscript{2} = 0.21) explained half of the observed variation
+between fecal communities with most of the remaining variation explained
+by interactions between treatment group and other experimental variables
+including time, cage, and sequencing preparation plate (PERMANOVA
+combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
+2A, Data Set S1, sheet 1). None of the treatment groups recovered to
+their baseline community structure either 10 or 30 days post-challenge
+suggesting other community features besides recovery to baseline were
+responsible for the prolonged \emph{C. difficile} colonization in
+PEG-treated mice (Fig. 2B).
 
@@
 groups of mice that received PEG treatment compared to those that
-received clindamycin through thirty days post-challenge (Fig. 2C). We
-next examined the bacterial genera that shifted after PEG treatment by
-comparing the baseline samples of mice treated with only PEG to samples
-from the same mice one day post-PEG-treatment. We found 18 genera that
-were altered by PEG treatment (Data Set S1, Sheet X). The majority of
-the bacterial relative abundances decreased after the PEG treatment, but
-the relative abundance among members of the \emph{Enterobacteriaceae}
-and \emph{Bacteroides} increased. The increase in \emph{Bacteroides}
-relative abundance was unique to PEG treated mice, as the
-\emph{Bacteroides} relative abundance actually decreased in clindamycin
-treated mice (Fig. 2D). Finally, we identified the genera whose relative
-abundance differed across treatment groups over multiple time points. Of
-the 33 genera that were different between treatment groups, 24 genera
-were different over multiple time points (Fig. 2E, Data Set S1, Sheet
-X). Thus, PEG had a significant impact on the fecal microbiota that was
-maintained over time and was distinct from clindamycin treatment.
-
-Interestingly, \emph{C. difficile} was not immediately detectable in the
-stools of the PEG-treated mice that were allowed to recover for 10 days
-prior to challenge. We decided to examine if there were genera that
-changed during the post-challenge period when the group median \emph{C.
-difficile} shifted from undetectable at 1 day post-challenge to
+received clindamycin through thirty days post-challenge (Fig. 2C; Data
+Set S1, sheet 2). We next examined the bacterial genera that shifted
+after PEG treatment by comparing the baseline samples of mice treated
+with only PEG to samples from the same mice one day post-PEG-treatment.
+We found 18 genera that were altered by PEG treatment (Data Set S1,
+sheet 3). The majority of the bacterial relative abundances decreased
+after the PEG treatment, but the relative abundance among members of the
+\emph{Enterobacteriaceae} and \emph{Bacteroides} increased. The increase
+in \emph{Bacteroides} relative abundance was unique to PEG treated mice,
+as the \emph{Bacteroides} relative abundance actually decreased in
+clindamycin treated mice (Fig. 2D). Finally, we identified the genera
+whose relative abundance differed across treatment groups over multiple
+time points. Of the 33 genera that were different between treatment
+groups, 24 genera were different over multiple time points (Fig. 2E,
+Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal
+microbiota that was maintained over time and was distinct from
+clindamycin treatment.
+
+Because \emph{C. difficile} was not immediately detectable in the stools
+of the PEG-treated mice that were allowed to recover for 10 days prior
+to challenge, we decided to examine if there were genera that changed
+during the post-challenge period. We compared the communities from when
+\emph{C. difficile} shifted from undetectable at 1 day post-challenge to
 detectable in the stool samples with the density stabilizing around 8
-days post-challenge (Fig. S1A). We found no bacteria with relative
+days post-challenge (Fig. S1A). We found no genera with relative
 abundances that were significantly different over the two time points
-after multiple hypothesis correction (Data Set S1, Sheet X). However,
-there was also wide variation between individual mice regarding when
-\emph{C. difficile} became detectable (Fig. S1A) as well as the relative
-abundances of bacterial genera in the communities (Fig. S1B). For
-example, two mice had a high relative abundance of
-\emph{Enterobacteriaceae} throughout the post-challenge period and this
-corresponded to mouse 10, which died on the sixth day post-challenge and
-mouse 11, where \emph{C. difficile} was present at a high density from
-the 4th day post-challenge onward (Fig. S1B). While we did not identify
-a clear signal to explain the delayed appearance of \emph{C. difficile}
-in the 5-day PEG mice that were allowed to recover for 10 days prior to
-challenge, the delay is striking and could reflect changes in microbial
+(Data Set S1, sheet 5). However, there was also wide variation between
+individual mice regarding when \emph{C. difficile} became detectable
+(Fig. S1A) as well as the relative abundances of bacterial genera in the
+communities (Fig. S1B). For example, two mice had a high relative
+abundance of \emph{Enterobacteriaceae} throughout the post-challenge
+period. One mouse died on the sixth day post-challenge and in the other
+\emph{C. difficile} was present at a high density from the 4th day
+post-challenge onward (Fig. S1B). While we did not identify a clear
+signal to explain the delayed appearance of \emph{C. difficile} in the
+5-day PEG mice that were allowed to recover for 10 days prior to
+challenge, the delay was striking and could reflect changes in microbial
 activity or metabolites that were not examined in this study.
@@
 impact of PEG treatment on the mucosal microbiota and CDI severity. To
-evaluate the mucosal microbiota, we sequenced snips of tissue collected
-from the cecum, proximal colon, and distal colon. Similar to what was
-observed with the stool samples, alpha diversity was lower in the
-PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha
-diversity continued to increase over time based on the tissues from
-PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A,
-Data Set S1, Sheet X). Group, time point, and their interactions with
-other variables (cage, experiment number, and sample type) explained the
-majority of the variation observed in mucosal communities (PERMANOVA
-combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
-3B, Data Set S1, Sheet X). We saw the greatest difference in the
-relative abundance of the mucosal microbiota between treatment groups
-(clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days
-post-challenge with 10 genera that were significantly different
+evaluate the mucosal microbiota, we sequenced communities associated
+with tissues collected from the cecum, proximal colon, and distal colon.
+Similar to what was observed with the stool samples, the alpha diversity
+was lower in the PEG-treated mice compared to clindamycin treated mice
+(Fig. 3A, Data Set S1, sheet 6). The alpha diversity of the
+tissue-associated community increased in PEG-treated mice collected at
+20 and 30 days post-challenge (Fig. 3A). Group (R\textsuperscript{2} =
+0.33), time point (R\textsuperscript{2} = 0.11), and their interactions
+with other variables (cage, experiment number, and sample type)
+explained the majority of the variation observed in mucosal communities
+(PERMANOVA combined R\textsuperscript{2} = 0.83, \emph{P} \textless{}
+0.05, Fig. 3B, Data Set S1, sheet 7). We saw the greatest difference in
+the relative abundance of the mucosal microbiota between treatment
+groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6
+days post-challenge with 10 genera that were significantly different
 (\emph{P} \textless{} 0.05) in all three of the tissue types we
 collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set
-S1, Sheet X). Interestingly, \emph{Peptostreptococcaceae} (the family
-with a sequence that matches \emph{C. difficile}) was one of the genera
-that had a significant difference in relative abundance between
-treatment groups at 6 days post-challenge. This population was primarily
-only present in the 5-day PEG treatment group of mice and decreased in
-the proximal and distal colon tissues over time (Fig. S2B). By 30 days
-post-challenge, only the relative abundances of \emph{Bacteroides},
-\emph{Clostridiales}, \emph{Firmicutes}, and \emph{Ruminococcaceae} were
-different between treatment groups and only in the cecum tissues (Fig.
-3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a
-significant impact on the mucosal microbiota and we detected \emph{C.
-difficile} sequences in the cecum, proximal colon, and distal colon
-tissue communities.
+S1, sheet 8, 9, and 10). Interestingly, \emph{Peptostreptococcaceae}
+(the family with a sequence that matches \emph{C. difficile}) was one of
+the genera that had a significant difference in relative abundance
+between treatment groups at 6 days post-challenge. This population was
+primarily only present in the 5-day PEG treatment group of mice and
+decreased in the proximal and distal colon tissues over time (Fig. S2B).
+By 30 days post-challenge, only the relative abundances of
+\emph{Bacteroides}, \emph{Clostridiales}, \emph{Firmicutes}, and
+\emph{Ruminococcaceae} were different between treatment groups and only
+in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, sheet 8). Thus, PEG
+treatment had a significant impact on the mucosal microbiota and we
+detected \emph{C. difficile} sequences in the cecum, proximal colon, and
+distal colon tissue communities.
 
@@
 severity of \emph{C. difficile} challenge by evaluating cecum and colon
-H\&E stained histopathology (28). However, we found there was no
-difference in cecum and colon scores between clindamycin and PEG-treated
-mice that were challenged with \emph{C. difficile} at 4 days
-post-challenge (Fig. 3D), the time point typically examined in \emph{C.
-difficile} 630 challenged mice (29). We also looked at 6 days
-post-challenge because that was when there was a large difference in
-\emph{C. difficile} density between PEG- and clindamycin-treated mice
-(Fig. 1C). Although, there was a slight difference in the colon between
-PEG and clindamycin-treated mice, there was not a signifant difference
-in the cecum and the overall score was relatively low (1.5-2.5 out of
-12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect
-on the mucosal microbiota, the impact of \emph{C. difficile} 630
-challenge on the cecum and colon was similar between PEG and clindamycin
-treated mice.
+histopathology (27). However, we found there was no difference in cecum
+and colon scores between clindamycin and PEG-treated mice that were
+challenged with \emph{C. difficile} at 4 days post-challenge (Fig. 3D),
+the time point typically examined in \emph{C. difficile} 630 challenged
+mice (28). We also looked at 6 days post-challenge because that was when
+there was a large difference in \emph{C. difficile} density between PEG-
+and clindamycin-treated mice (Fig. 1C). Although there was a slight
+difference in the histopathology score of the colon between PEG and
+clindamycin-treated mice, there was not a signifant difference in the
+cecum and the overall score was relatively low (1.5 to 2.5 out of 12,
+Fig. 3E). Therefore, although PEG treatment had a disruptive effect on
+the mucosal microbiota, the impact of \emph{C. difficile} 630 challenge
+on the cecum and colon was similar between PEG and clindamycin treated
+mice.
 
@@
 disruptive effect on the microbiota of PEG-treated mice.} Because
-\emph{C. difficile} itself can have an impact on the microbiota (30), we
-also sequenced the tissue and stools of mock-challenged clindamycin and
-5-day PEG treated mice. Examining the stools of the mock-challenged mice
-revealed similar bacterial disruptions as the \emph{C. difficile}
+\emph{C. difficile} itself can have an impact on the microbiota (29), we
+also sequenced the tissue and stools of mock-challenged mice treated
+with clindamycin or PEG. Examining the stools of the mock-challenged
+mice revealed similar bacterial disruptions as the \emph{C. difficile}
 challenged mice (Fig. S3A-C). Similarly, there was no difference between
-the tissues of mock and \emph{C. difficile} challenged mice (Fig.
-S3D-F). Thus, most of the microbiota alterations we observed in the
-PEG-treated mice were a result of the laxative and not an interaction
-between the laxative and \emph{C. difficile}.
+the communities found in the tissues of mock and \emph{C. difficile}
+challenged mice (Fig. S3D-F). Thus, most of the microbiota alterations
+we observed in the PEG-treated mice were a result of the laxative and
+not an interaction between the laxative and \emph{C. difficile}.
 
@@
 \emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
-mice, the 1-day PEG treated mice were only transiently colonized and
-cleared \emph{C. difficile} by 7 days post-challenge (Fig. 3B). The
-stool communities of PEG-treated mice were also only transiently
+mice, the 1-day PEG groups were only transiently colonized and cleared
+\emph{C. difficile} by 7 days post-challenge (Fig. 3B). The stool
+communities of the 1-day PEG treatment groups were also only transiently
 disrupted, with Shannon diversity recovering by 7 days post-challenge
-(Fig. 3C-D). We found the relative abundances of 14 genera were impacted
-by treatment, but recovered close to baseline levels by 7 days
-post-challenge including \emph{Enterobacteriaceae},
-\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
-\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
-suggest the duration of the PEG treatment was relevant, with shorter
-treatments resulting in a transient loss of \emph{C. difficile}
-colonization resistance.
+(Fig. 3C-D, Data Set S1, sheets 11 and 12). We found the relative
+abundances of 14 genera were impacted by treatment, but recovered close
+to baseline levels by 7 days post-challenge including
+\emph{Enterobacteriaceae}, \emph{Clostridiales},
+\emph{Porpyromonadaceae}, and \emph{Ruminococcaceae} (Fig. 3E, Data Set
+S1, sheet 13 and 14). These findings suggest the duration of the PEG
+treatment was relevant, with shorter treatments resulting in a transient
+loss of \emph{C. difficile} colonization resistance.
 
@@
 administered.} Since a 1-day PEG treatment resulted in a more mild
-microbiota perturbation, we decided to use the 1-day treatment to
+perturbation of the microbiota, we decided to use the 1-day treatment to
 examine the hypothesis that PEG helps to flush \emph{C. difficile}
-spores from the intestine. To examine the impact of PEG treatment on
-\emph{C. difficile} clearance, we treated 4 groups of mice with
-clindamycin and then challenged all mice with \emph{C. difficile} before
-administering the following treatments: no additional treatment, 1-day
-PEG immediately after challenge, and 1-day PEG treatment 3 days after
-challenge followed by either administration of an FMT or PBS solution by
-oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice
-that received PEG exhibited prolonged \emph{C. difficile} colonization
-(Fig. 5B).
+spores from the intestine. This hypothesis is proposed in the discussion
+section of FMT studies where bowel prep is part of the preparation
+undergone by patients receiving FMTs via colonoscopy (20--23). To
+examine the impact of PEG treatment on \emph{C. difficile} clearance, we
+treated 4 groups of mice with clindamycin and then challenged all mice
+with \emph{C. difficile} before administering the following treatments:
+no additional treatment, 1-day PEG immediately after challenge, and
+1-day PEG treatment 3 days after challenge followed by either
+administration of an FMT or PBS solution by oral gavage (Fig. 5A).
+Contrary to the hypothesis, all groups of mice that received PEG
+exhibited prolonged \emph{C. difficile} colonization (Fig. 5B).
 
 We were also interested in exploring whether PEG might help with
-engraftment in the context of FMTs. The FMT appeared to partially
-restore Shannon diversity but not richness (Fig. 5C-D). Similarly, we
-saw some overlap between the communities of mice that received FMT and
-the mice treated with only clindamycin after 5 days post-challenge (Fig.
-6A). The increase in Shannon diversity suggests that the FMT did have an
+engraftment in the context of FMTs. An FMT was prepared under anaerobic
+conditions using stool collected from the same group of mice
+pre-treatment representing the baseline community. The FMT appeared to
+partially restore Shannon diversity but not richness (Fig. 5C-D, Data
+Set S1, sheets 15 and 16). Similarly, we saw some overlap between the
+communities of mice that received FMT and the mice treated with only
+clindamycin after 5 days post-challenge (Fig. 6A, Data Set S1, sheet
+17). The increase in Shannon diversity suggests that the FMT did have an
 impact on the microbiota, despite seeing prolonged \emph{C. difficile}
@@
 abundances of \emph{Bacteroidales} and \emph{Porphyromonadaceae}
-consistently differed between the mice received either an FMT or PBS
-gavage (Fig. 6B), suggesting the FMT only restored a couple of genera.
-Overall, we found the relative abundances of 24 genera were different
-between treatment groups over multiple timepoints. For example, the
-relative abundance of \emph{Akkermansia} was increased and the relative
-abundances of \emph{Ruminococcaceae}, \emph{Clostridiales},
-\emph{Lachnospiraceae}, and \emph{Oscillibacter} were decreased in mice
-that received PEG after \emph{C. difficile} challenge relative to
-clindamycin treated mice (Fig. 6C). In sum, administering PEG actually
-prolonged \emph{C. difficile} colonization, including in mice that
-received an FMT, which only restored 2 bacterial genera.
+consistently differed between the mice that received either an FMT or
+PBS gavage (Fig. 6B). Overall, we found the relative abundances of 24
+genera were different between treatment groups over multiple time points
+(Data Set S1, sheet 18). For example, the relative abundance of
+\emph{Akkermansia} was increased and the relative abundances of
+\emph{Ruminococcaceae}, \emph{Clostridiales}, \emph{Lachnospiraceae},
+and \emph{Oscillibacter} were decreased in mice that received PEG after
+\emph{C. difficile} challenge relative to clindamycin treated mice (Fig.
+6C). In sum, administering PEG actually prolonged \emph{C. difficile}
+colonization, including in mice that received an FMT, which only
+restored 2 bacterial genera.
 
 \textbf{Five-day post-challenge community data can predict which mice
-that will have prolonged \emph{C. difficile} colonization.} After
-identifying bacteria associated with the 5-day, 1-day and post-challenge
-1-day PEG treatments, we decided to examine the bacteria that influenced
-prolonged \emph{C. difficile} colonization. We trained 3 types of
-machine learning models (random forest, logistic regression, and support
-vector machine) with bacterial community data from 5 days post-challenge
-to predict whether the mice were still colonized with \emph{C.
-difficile} 10 days post-challenge. We chose 5 days post-challenge
-because that was the earliest time point where we would see a treatment
-effect in the post-challenge 1-day PEG experiments. The random forest
-model had the highest performance (median AUROC = 0.90, Data Set S1,
-Sheet X), so we next performed permutation importance to examine the
-bacteria that were the top contributors to the random forest model
-predicting prolonged \emph{C. difficile }colonization. We selected the
-top 10 bacteria contributing to our model's performance (Fig. 7A) and
-examined their relative abundance at 5 days post-challenge, the time
+will have prolonged \emph{C. difficile} colonization.} After identifying
+bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG
+treatments, we examined the bacteria that influenced prolonged \emph{C.
+difficile} colonization. We trained 3 machine learning models (random
+forest, logistic regression, and support vector machine) with bacterial
+community data from 5 days post-challenge to predict whether the mice
+were still colonized with \emph{C. difficile} 10 days post-challenge. We
+chose 5 days post-challenge because that was the earliest time point
+where we saw a treatment effect in the post-challenge 1-day PEG
+experiments. The random forest model had the highest performance (median
+AUROC = 0.90, Data Set S1, sheet 19), suggesting the 5-day post
+challenge microbiota is an excellent predictor of prolonged \emph{C.
+difficile} colonization. Next, we performed permutation importance to
+examine the bacteria that were the top contributors to the random forest
+model predicting prolonged \emph{C. difficile }colonization. We selected
+10 genera that contributed the most to our model's performance (Fig. 7A)
+and examined their relative abundance at 5 days post-challenge, the time
 point used to predict \emph{C. difficile} colonization status on day 10
-(Fig. 7B). Next, we focused on the 5 genera that had a greater than 1 \%
+(Fig. 7B). Next, we focused on the 5 genera that had a greater than 1\%
 relative abundance in either the cleared or colonized mice and examined
@@
 drugs such as laxatives disrupt colonization resistance was unclear. By
-following osmotic laxative treated mice over time, we found 5-day PEG
-treatment before challenge resulted in prolonged \emph{C. difficile}
+following osmotic laxative treated mice over time, we found that a 5-day
+PEG treatment before challenge resulted in prolonged \emph{C. difficile}
 colonization, while a 1-day PEG treatment resulted in transient
@@
 differences in how much the treatments disrupted the microbiota.
-Additionally, the intestinal communities of 5-day PEG treated mice had
-not regained colonization resistance after a 10-day recovery period.
-Although in contrast to the other 5-day PEG treatment groups, \emph{C.
-difficile} was not immediately detectable in the stools of most of the
-mice in the 10-day recovery group. We also examined the impact of PEG
-treatment after \emph{C. difficile} challenge and in opposition to the
-hypothesis suggested by the literature, found that PEG treatment
-prolonged colonization relative to mice that only recieved clindamycin
-treatment. We identified patterns in the relative abundances of
-\emph{Bacteroides}, \emph{Enterobacteriaceae}, \emph{Akkermansia},
+Additionally, the intestinal communities of 5-day PEG treated mice did
+not regain colonization resistance after a 10-day recovery period. In
+contrast to the other 5-day PEG treatment groups, \emph{C. difficile}
+was not immediately detectable in the stools of most of the mice in the
+10-day recovery group. We also examined the impact of PEG treatment
+after \emph{C. difficile} challenge. In opposition to the hypothesis
+suggested by the literature, we found that PEG treatment prolonged
+colonization relative to mice that only recieved clindamycin treatment.
+We identified patterns in the relative abundances of \emph{Bacteroides},
+\emph{Enterobacteriaceae}, \emph{Akkermansia},
 \emph{Porphyromonadaceae}, and \emph{Lachnospiraceae} that were
@@
 compared to mice with prolonged colonization. For example,
-\emph{Oscillibacter valericigenes} can produce the SCFA valerate (31),
+\emph{Oscillibacter valericigenes} can produce the SCFA valerate (30),
 and separate studies demonstrated valerate is depleted after clindamycin
-treatment and inhibited \emph{C. difficile} growth based on \emph{in
-vitro} data and an experiment with C57BL/6 mice (32, 33). Similarly,
-\emph{Acetatifactor} can produce acetate and butyrate (34), SCFAs that
-are decreased in mice with prolonged \emph{C. difficile} infection after
-antibiotic treatment (35). Thus protective bacteria and their
-metabolites could be depleted by osmotive laxative treatment depending
-on the timing and duration of treatment.
+treatment and inhibited \emph{C. difficile} growth \emph{in vitro} and
+in C57BL/6 mice (31, 32). Similarly, \emph{Acetatifactor} can produce
+acetate and butyrate (33), SCFAs that are decreased in mice with
+prolonged \emph{C. difficile} infection after antibiotic treatment (34).
+Thus protective bacteria and their metabolites could be depleted by
+osmotive laxative treatment depending on the timing and duration of
+treatment.
 
@@
 hypothesized that \emph{C. difficile} biofilms may serve as reservoirs
-for recurrent infections (36) and \emph{C. difficile} biofilms were
-recently identified in patients and aggregates with \emph{Fusobacterium
-nucleatum}, another species capable of forming biofilms (37). There was
-an interesting pattern of increased \emph{Enterobacteriaceae},
-\emph{Bacteroides}, and \emph{C. difficile} in both the stool and
-mucosal communities of PEG-treated mice suggesting a potential synergy.
-\emph{Bacteroides} has the potential to degrade mucus and the osmotic
-laxative may have allowed \emph{Bacteroides} to colonize the mucosal
-niche by degrading mucin glycans with glycosyl hydrolases that are
-absent in \emph{C. difficile} (38). \emph{Bacteroides} persistent in the
-mucosal tissue might also have helped \emph{Enterobacteriaceae} to
-colonize the region, as a synergy between mucus-degrading \emph{B.
-fragilis} and \emph{E. coli} has previously been described (39). A
-separate study demonstrated \emph{C. difficile} was present in the outer
-mucus layer and associated with \emph{Enterobacteriaceae} and
-\emph{Bacteroidaceae} using fluorescent in situ hybridization (FISH)
-staining (40). However, \emph{B. fragilis} prevented CDI morbidity in a
-mouse model and inhibited \emph{C. difficile} adherence \emph{in vitro}
-(41). In coculture experiments \emph{B. longum} decreased \emph{C.
-difficile} biofilm formation while \emph{B. thetaiotamicron} enhanced
-biofilm formation (42). Therefore, whether \emph{Bacterodes} is
-detrimental or beneficial in the context of \emph{C. difficile}
-infection or colonization is still unclear.
+for recurrent infections (35) and \emph{C. difficile} biofilms in the
+mucus layer were recently identified in patients as aggregates with
+\emph{Fusobacterium nucleatum} (36). There was an interesting pattern of
+increased \emph{Enterobacteriaceae}, \emph{Bacteroides}, and \emph{C.
+difficile} in both the stool and mucosal communities of PEG-treated mice
+suggesting a potential synergy. \emph{Bacteroides} has the potential to
+degrade mucus and the osmotic laxative may have allowed
+\emph{Bacteroides} to colonize the mucosal niche by degrading mucin
+glycans with glycosyl hydrolases that are absent in \emph{C. difficile}
+(37). \emph{Bacteroides} persistent in the mucosal tissue might also
+have helped \emph{Enterobacteriaceae} to colonize the region, as a
+synergy between mucus-degrading \emph{B. fragilis} and \emph{E. coli}
+has previously been described (38). A separate study demonstrated
+\emph{C. difficile} was present in the outer mucus layer and associated
+with \emph{Enterobacteriaceae} and \emph{Bacteroidaceae} using
+fluorescent in situ hybridization (FISH) staining (39). However,
+protective roles for \emph{Bacteroides} have also been demonstrated. For
+example, \emph{B. fragilis} prevented CDI morbidity in a mouse model and
+inhibited \emph{C. difficile} adherence \emph{in vitro} (40). In
+coculture experiments \emph{B. longum} decreased \emph{C. difficile}
+biofilm formation while \emph{B. thetaiotamicron} enhanced biofilm
+formation (41) and \emph{B. dorei} reduced \emph{C. difficile} growth in
+a 9-species community \emph{in vitro} (42). Therefore, whether
+\emph{Bacteroides} is detrimental or beneficial in the context of
+\emph{C. difficile} infection or colonization is still unclear, but the
+niche and interactions with other bacteria may contribute.
 
@@
 our study the relative abundance of \emph{Akkermansia} shifted over time
-between groups. In the stool it was initially increased in mice that
-cleared \emph{C. difficile}, but shifted after 5-days post-challenge so
-that it was increased in mice that had prolonged colonization. In the
-context of CDIs, some studies suggest a protective role (45, 46), while
-others suggest detrimental (47--49). Because the relative abundance of
-\emph{Akkermansia} was dynamic in our study so it is unclear whether
-\emph{Akkermansia} helps with clearance of \emph{C. difficile} or allows
-it to persist. A better understanding how \emph{C. difficile} interacts
-with the mucosal microbiota may lead to insights into CDIs, asymptomatic
-\emph{C. difficile} colonization, and resistance.
+between groups of mice that either cleared \emph{C. difficile} or had
+prolonged colonization. In the stool it was initially increased in mice
+that cleared \emph{C. difficile}, but shifted after 5-days
+post-challenge so that it was increased in mice that had prolonged
+colonization. In the context of CDIs, some studies suggest a protective
+role (45, 46), while others suggest a detrimental role because
+\emph{Akkermansia} was positively correlated with \emph{C. difficile}
+(47--50). Because the relative abundance of \emph{Akkermansia} was
+dynamic in our study, it is unclear whether \emph{Akkermansia} helps
+with clearance of \emph{C. difficile} or allows it to persist. A better
+understanding how \emph{C. difficile} interacts with the mucosal
+microbiota may lead to insights into CDIs, asymptomatic \emph{C.
+difficile} carriage, and colonizatiion resistance.
 
@@
 histopathology summary scores in mice compared to VPI 10463 despite both
-strains producing toxin in mice (50). Part of our hypothesis for why
+strains producing toxin in mice (51). Part of our hypothesis for why
 there could have been increased histopathology scores in PEG-treated
@@
 mice. However, recent studies demonstrated that broad spectrum
-antibiotics can also disrupt the host mucosal barrier in mice (51, 52).
+antibiotics can also disrupt the host mucosal barrier in mice (52, 53).
 Future research is needed to tease out the interplay between medications
@@
 
-It is more difficult what are findings mean in the context of \emph{C.
-difficile} colonization resistance mean for human patients. Considering
-that most hospitals recommend not performing \emph{C. difficile} testing
-on patients taking laxatives and laxatives are used when administering
-fecal microbiota transplants via colonoscopy to patients with recurrent
-CDIs. further studies are needed to evaluate if laxatives impact human
-microbiota colonization resistance. Further studies are needed to
-understand the impact of osmotic laxatives on \emph{C. difficile}
-colonization resistance and clearance in human patients. + What's known
-regarding laxatives and susceptibility to CDIs + Clinical trial of PEG,
-results never posted (53)
-
-Relevance to human FMTs? Unclear what the best administration route is
-because there have been no studies designed to evaluate the best
-administration route for FMTs. However, results from the FMT National
-Registry where 85\% of FMTs were delivered by colonoscopy demonstrate
-FMTs are highly effective treatment for recurrent CDIs with 90\%
-achieving resolution in the 1 month follow-up window (54).
+It is more difficult to interpret what are findings mean in the context
+of \emph{C. difficile} colonization resistance in human patients. The
+main difficulty being that most hospitals recommend not performing
+\emph{C. difficile} testing if the patient is currently taking a
+laxative. This recommendation is in accordance with the Infectious
+Diseases Society of America and Society for Healthcare Epidemiology of
+America guidelines (54). The rationale behind the recommendation is that
+patients taking laxatives may be asymptomatically colonized with
+\emph{C. difficile}, resulting in unnecessary antibiotic treatment
+(55--57). Furthermore, some studies identified laxatives as a risk
+factor for developing CDIs or recurrent CDIs (58--60) and a recent study
+found the proportion of severe CDIs was similar between patients taking
+and not taking laxatives (61). However, there have also been some
+studies that suggest laxatives are not a risk factor for developing CDIs
+(62, 63). Although, it is unclear whether laxatives impact CDI
+susceptibility in human paitents, it is clear that laxatives also have a
+transient impact on the human microbiota (13, 64--67). Further studies
+to examine the relationship between laxatives, \emph{C. difficile}
+colonization, and CDIs are warranted.
+
+Considering laxatives are also used to prepare patients when
+administering fecal microbiota transplants via colonoscopy to treat
+recurrent CDIs, it will be important to determine if osmotic laxatives
+impact \emph{C. difficile} clearance in the human intestinal tract. It
+is still unclear what the best administration route is because there
+have been no studies designed to evaluate the best administration route
+for FMTs (68). Nevertheless, results from the FMT National Registry
+where 85\% of FMTs were delivered by colonoscopy demonstrate FMTs are
+highly effective treatment for recurrent CDIs with 90\% achieving
+resolution in the 1 month follow-up window (69). A surprising number of
+studies continue to hypothesize that PEG or bowel preparation can clear
+\emph{C. difficile} spores and toxins despite the paucity of supporting
+evidence (20--23). There was even a clinical trial (NCT01630096)
+designed to examine whether administering PEG 3350 (NuLYTELY) prior to
+antibiotic treatment reduced disease severity that started recruitment
+in 2012 (70), but no results have been posted to date. Here we sought to
+evaluate the impact of treating \emph{C. difficile} colonized mice with
+PEG (with or without FMT) and found clearance was delayed. Further
+studies are needed to understand the impact of osmotic laxatives on
+\emph{C. difficile} colonization resistance and clearance in human
+patients receiving FMTs.
 
@@
 media preparation and bacterial culture and the Microbiology and
-Immunology department's postdoc association writing group members for
-their feedback on manuscript drafts. We also thank the Unit for
+Immunology department's postdoctoral association writing group members
+for their feedback on manuscript drafts. We also thank the Unit for
 Laboratory Animal Medicine at the University of Michigan for maintaining
@@
 chamber. This work was supported by the National Institutes of Health
-(U01AI124255). ST was supported by the Michigan Institute for Clincial
+(U01AI124255). ST was supported by the Michigan Institute for Clinical
 and Health Research Postdoctoral Translation Scholars Program
@@
 Jackson Laboratory in 2002). We used 7-19 week old female mice for all
-experiments, which allowed us to break up littermates prior to starting
-an experiment and distribute them as evenly as possible across treatment
-groups in order to minimize microbiota differences between experiment
-groups prior to starting treatments with medications. During the
-experiment, mice were housed at a density of 2-3 mice per cage, with the
-majority of cages limited to two mice.
+experiments. This allowed us to break up littermates and distribute them
+as evenly as possible across treatment groups in order to minimize
+microbiota differences prior to starting treatments with medications.
+During the experiment, mice were housed at a density of 2-3 mice per
+cage, with the majority of cages limited to two mice.
 
@@
 25 microliters off \emph{C. difficile} 630 spores at
-10\textsuperscript{5} concentration, except for 1 experiment (Fig. 5A)
-where the concentration was 10\textsuperscript{3}. All mock challenged
+10\textsuperscript{5} concentration, except for 1 experiment where the
+concentration was 10\textsuperscript{3} (Fig. 5A). All mock challenged
 mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper
@@
 AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom
-barcoded primers, as previously described (55). The PCR amplicons were
+barcoded primers, as previously described (71). The PCR amplicons were
 normalized (SequalPrep normalizatin plate kit from Thermo Fisher
@@
 \textbf{16S rRNA gene sequence analysis.} All sequences were processed
-with mothur (v. 1.43) using a previously published protocol (55, 56).
+with mothur (v. 1.43) using a previously published protocol (71, 72).
 Paired sequencing reads were combined and aligned with the SILVA (v.
-132) reference database (57) and taxonomy was assigned with a modified
-version (v. 16) of the Ribosomal Database Project (v. 11.5) (58). The
+132) reference database (73) and taxonomy was assigned with a modified
+version (v. 16) of the Ribosomal Database Project (v. 11.5) (74). The
 error rate for are sequencing data was 0.0559\% based on the 17 mock
@@
 Bray-Curtis distance matrices with the adonis function from the vegan R
-package (59).
+package (75).
 
@@
 pathologist (ILB). Slides were evaluated using a scoring system
-developed for mouse models of \emph{C. difficile} infection (50). Each
+developed for mouse models of \emph{C. difficile} infection (51). Each
 slide was evaluated for edema, cellular infiltration, and inflammation
@@
 difficile} colonization status 10 days post-challenge where mice were
-categorized as either cleared or colonized (60, 61). We removed the
+categorized as either cleared or colonized (76, 77). We removed the
 \emph{C. difficile} genus relative abundance data prior to training the
@@
 of the training data for hyperparamter tuning. Permutation importance
-was performed as described previously (62) using mikropml (60, 61) with
+was performed as described previously (78) using mikropml (76, 77) with
 the random forest model because it had the highest AUROC value.
@@
 \textbf{Statistical analysis.} R (v. 4.0.2) and the tidyverse package
-(v. 1.3.0) were used for statistical analysis (63, 64). Kruskal-Wallis
+(v. 1.3.0) were used for statistical analysis (79, 80). Kruskal-Wallis
 tests with Bejamini-Hochberg correction for testing multiple time points
@@
 1. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
-microbiota in resistance to colonization by clostridium difficile.
-Gastroenterology \textbf{146}:1547--1553.
+microbiota in resistance to colonization by \emph{Clostridium
+difficile}. Gastroenterology \textbf{146}:1547--1553.
 doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
@@
 \textbf{Shenoy ES}, \textbf{Wiens J}. 2018. A generalizable, data-driven
-approach to predict daily risk of clostridium difficile infection at two
-large academic health centers. Infection Control \& Hospital
+approach to predict daily risk of \emph{Clostridium difficile} infection
+at two large academic health centers. Infection Control \& Hospital
 Epidemiology \textbf{39}:425--433.
@@
 KW}. 2012. Moderate to high use of opioid analgesics are associated with
-an increased risk of clostridium difficile infection. The American
-Journal of the Medical Sciences \textbf{343}:277--280.
+an increased risk of \emph{Clostridium difficile} infection. The
+American Journal of the Medical Sciences \textbf{343}:277--280.
 doi:\href{https://doi.org/10.1097/maj.0b013e31822f42eb}{10.1097/maj.0b013e31822f42eb}.
@@
 \textbf{Krishna K}, \textbf{Chepyala P}, \textbf{Anaissie EJ}. 2013.
-Risk factors, preemptive therapy, and antiperistaltic agents
-forClostridium difficileinfection in cancer patients. Transplant
+Risk factors, preemptive therapy, and antiperistaltic agents for
+\emph{Clostridium difficile} infection in cancer patients. Transplant
 Infectious Disease n/a--n/a.
@@
 \textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
-distinguish patients with clostridium difficile infection and non-c.
-Difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
+distinguish patients with \emph{Clostridium difficile} infection and
+non-\emph{C.~difficile}-associated diarrhea from healthy controls. mBio
+\textbf{5}.
 doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
@@
 \textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
-Gut microbiota-produced succinate promotes c.~difficile infection after
-antibiotic treatment or motility disturbance. Cell Host \& Microbe
+Gut microbiota-produced succinate promotes \emph{C.~difficile} infection
+after antibiotic treatment or motility disturbance. Cell Host \& Microbe
 \textbf{16}:770--777.
@@
 \textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
-Diarrhoeal events can trigger long-term clostridium difficile
+Diarrhoeal events can trigger long-term \emph{Clostridium difficile}
 colonization with recurrent blooms. Nature Microbiology
@@
 18. \textbf{Olson A}, \textbf{Diebel LN}, \textbf{Liberati DM}. 2013.
-Effect of host defenses on clostridium difficile toxininduced intestinal
-barrier injury. Journal of Trauma and Acute Care Surgery
+Effect of host defenses on \emph{Clostridium difficile} toxininduced
+intestinal barrier injury. Journal of Trauma and Acute Care Surgery
 \textbf{74}:983--990.
@@
 19. \textbf{Diebel LN}, \textbf{Liberati DM}. 2014. Reinforcement of the
-intestinal mucus layer protects against clostridium difficile intestinal
-injury in~vitro. Journal of the American College of Surgeons
+intestinal mucus layer protects against \emph{Clostridium difficile}
+intestinal injury in~vitro. Journal of the American College of Surgeons
 \textbf{219}:460--468.
@@
 \textbf{Tijssen JGP}, \textbf{Speelman P}, \textbf{Dijkgraaf MGW},
-\textbf{Keller JJ}. 2013. Duodenal infusion of donor feces for
-RecurrentClostridium difficile. New England Journal of Medicine
+\textbf{Keller JJ}. 2013. Duodenal infusion of donor feces for recurrent
+\emph{Clostridium difficile}. New England Journal of Medicine
 \textbf{368}:407--415.
@@
 \textbf{Allegretti JR}, \textbf{Kassam Z}. 2017. Faecal microbiota
-transplantation for clostridium difficile infection: A multicentre study
-of non-responders. Medical Journal of Australia \textbf{207}:159--160.
+transplantation for \emph{Clostridium difficile} infection: A
+multicentre study of non-responders. Medical Journal of Australia
+\textbf{207}:159--160.
 doi:\href{https://doi.org/10.5694/mja16.01452}{10.5694/mja16.01452}.
@@
 22. \textbf{Postigo R}, \textbf{Kim JH}. 2012. Colonoscopic versus
-nasogastric fecal transplantation for the treatment of clostridium
-difficile infection: A review and pooled analysis. Infection
+nasogastric fecal transplantation for the treatment of \emph{Clostridium
+difficile} infection: A review and pooled analysis. Infection
 \textbf{40}:643--648.
@@
 irrigation as an adjunct to the treatment of chronic, relapsing
-clostridium difficile colitis. Journal of Clinical Gastroenterology
-\textbf{22}:186--189.
+\emph{Clostridium difficile} colitis. Journal of Clinical
+Gastroenterology \textbf{22}:186--189.
 doi:\href{https://doi.org/10.1097/00004836-199604000-00007}{10.1097/00004836-199604000-00007}.
@@
 \textbf{Schloss PD}. 2020. The initial gut microbiota and response to
-antibiotic perturbation influence clostridioides difficile clearance in
-mice. mSphere \textbf{5}.
+antibiotic perturbation influence \emph{Clostridioides difficile}
+clearance in mice. mSphere \textbf{5}.
 doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
 
-\leavevmode\hypertarget{ref-Nguyen2015}{}%
-27. \textbf{Nguyen TLA}, \textbf{Vieira-Silva S}, \textbf{Liston A},
-\textbf{Raes J}. 2015. How informative is the mouse for human gut
-microbiota research? Disease Models \& Mechanisms \textbf{8}:1--16.
-doi:\href{https://doi.org/10.1242/dmm.017400}{10.1242/dmm.017400}.
-
 \leavevmode\hypertarget{ref-Reeves2011}{}%
-28. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+27. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
 \textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
@@
 \leavevmode\hypertarget{ref-Dieterle2020}{}%
-29. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+28. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
 \textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
@@
 K}. 2020. Systemic inflammatory mediators are effective biomarkers for
-predicting adverse outcomes in clostridioides difficile infection. mBio
-\textbf{11}.
+predicting adverse outcomes in \emph{Clostridioides difficile}
+infection. mBio \textbf{11}.
 doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
@@
 \leavevmode\hypertarget{ref-Jenior2018}{}%
-30. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+29. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
@@
 \leavevmode\hypertarget{ref-Iino2007}{}%
-31. \textbf{Iino T}, \textbf{Mori K}, \textbf{Tanaka K}, \textbf{Suzuki
-K-i}, \textbf{Harayama S}. 2007. Oscillibacter valericigenes gen. Nov.,
-sp. Nov., a valerate-producing anaerobic bacterium isolated from the
-alimentary canal of a japanese corbicula clam. International Journal of
-Systematic and Evolutionary Microbiology \textbf{57}:1840--1845.
+30. \textbf{Iino T}, \textbf{Mori K}, \textbf{Tanaka K}, \textbf{Suzuki
+K-i}, \textbf{Harayama S}. 2007. \emph{Oscillibacter valericigenes} gen.
+Nov., sp. Nov., a valerate-producing anaerobic bacterium isolated from
+the alimentary canal of a japanese corbicula clam. International Journal
+of Systematic and Evolutionary Microbiology \textbf{57}:1840--1845.
 doi:\href{https://doi.org/10.1099/ijs.0.64717-0}{10.1099/ijs.0.64717-0}.
@@
 \leavevmode\hypertarget{ref-Jump2014}{}%
-32. \textbf{Jump RLP}, \textbf{Polinkovsky A}, \textbf{Hurless K},
+31. \textbf{Jump RLP}, \textbf{Polinkovsky A}, \textbf{Hurless K},
 \textbf{Sitzlar B}, \textbf{Eckart K}, \textbf{Tomas M},
@@
 \leavevmode\hypertarget{ref-McDonald2018}{}%
-33. \textbf{McDonald JAK}, \textbf{Mullish BH}, \textbf{Pechlivanis A},
+32. \textbf{McDonald JAK}, \textbf{Mullish BH}, \textbf{Pechlivanis A},
 \textbf{Liu Z}, \textbf{Brignardello J}, \textbf{Kao D}, \textbf{Holmes
 E}, \textbf{Li JV}, \textbf{Clarke TB}, \textbf{Thursz MR},
-\textbf{Marchesi JR}. 2018. Inhibiting growth of clostridioides
-difficile by restoring valerate, produced by the intestinal microbiota.
+\textbf{Marchesi JR}. 2018. Inhibiting growth of \emph{Clostridioides
+difficile} by restoring valerate, produced by the intestinal microbiota.
 Gastroenterology \textbf{155}:1495--1507.e15.
@@
 \leavevmode\hypertarget{ref-Pfeiffer2012}{}%
-34. \textbf{Pfeiffer N}, \textbf{Desmarchelier C}, \textbf{Blaut M},
+33. \textbf{Pfeiffer N}, \textbf{Desmarchelier C}, \textbf{Blaut M},
 \textbf{Daniel H}, \textbf{Haller D}, \textbf{Clavel T}. 2012.
-Acetatifactor muris gen. Nov., sp. Nov., a novel bacterium isolated from
-the intestine of an obese mouse. Archives of Microbiology
+\emph{Acetatifactor muris} gen. Nov., sp. Nov., a novel bacterium
+isolated from the intestine of an obese mouse. Archives of Microbiology
 \textbf{194}:901--907.
@@
 \leavevmode\hypertarget{ref-Lawley2012}{}%
-35. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
+34. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
 \textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
@@
 \leavevmode\hypertarget{ref-Frost2021}{}%
-36. \textbf{Frost LR}, \textbf{Cheng JKJ}, \textbf{Unnikrishnan M}.
-2021. Clostridioides difficile biofilms: A mechanism of persistence in
-the gut? PLOS Pathogens \textbf{17}:e1009348.
+35. \textbf{Frost LR}, \textbf{Cheng JKJ}, \textbf{Unnikrishnan M}.
+2021. \emph{Clostridioides difficile} biofilms: A mechanism of
+persistence in the gut? PLOS Pathogens \textbf{17}:e1009348.
 doi:\href{https://doi.org/10.1371/journal.ppat.1009348}{10.1371/journal.ppat.1009348}.
@@
 \leavevmode\hypertarget{ref-Engevik2021}{}%
-37. \textbf{Engevik MA}, \textbf{Danhof HA}, \textbf{Auchtung J},
+36. \textbf{Engevik MA}, \textbf{Danhof HA}, \textbf{Auchtung J},
 \textbf{Endres BT}, \textbf{Ruan W}, \textbf{Bassères E},
@@
 \textbf{Savidge T}, \textbf{Britton RA}, \textbf{Versalovic J}. 2021.
-Fusobacterium nucleatum adheres to clostridioides difficile via the RadD
-adhesin to enhance biofilm formation in intestinal mucus.
+Fusobacterium nucleatum adheres to \emph{Clostridioides difficile} via
+the RadD adhesin to enhance biofilm formation in intestinal mucus.
 Gastroenterology \textbf{160}:1301--1314.e8.
@@
 \leavevmode\hypertarget{ref-Engevik2020}{}%
-38. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
+37. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
 \textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
@@
 \textbf{Versalovic J}. 2020. Mucin-degrading microbes release
-monosaccharides that chemoattract clostridioides difficile and
+monosaccharides that chemoattract \emph{Clostridioides difficile} and
 facilitate colonization of the human intestinal mucus layer. ACS
@@
 \leavevmode\hypertarget{ref-Dejea2018}{}%
-39. \textbf{Dejea CM}, \textbf{Fathi P}, \textbf{Craig JM},
+38. \textbf{Dejea CM}, \textbf{Fathi P}, \textbf{Craig JM},
 \textbf{Boleij A}, \textbf{Taddese R}, \textbf{Geis AL}, \textbf{Wu X},
@@
 \leavevmode\hypertarget{ref-Semenyuk2015}{}%
-40. \textbf{Semenyuk EG}, \textbf{Poroyko VA}, \textbf{Johnston PF},
+39. \textbf{Semenyuk EG}, \textbf{Poroyko VA}, \textbf{Johnston PF},
 \textbf{Jones SE}, \textbf{Knight KL}, \textbf{Gerding DN},
 \textbf{Driks A}. 2015. Analysis of bacterial communities during
-clostridium difficile infection in the mouse. Infection and Immunity
-\textbf{83}:4383--4391.
+\emph{Clostridium difficile} infection in the mouse. Infection and
+Immunity \textbf{83}:4383--4391.
 doi:\href{https://doi.org/10.1128/iai.00145-15}{10.1128/iai.00145-15}.
@@
 \leavevmode\hypertarget{ref-Deng2018}{}%
-41. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
+40. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
 \textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
 \textbf{Fan H}, \textbf{Zhao X}, \textbf{Zhi F}. 2018. Bacteroides
-fragilis prevents clostridium difficile infection in a mouse model by
-restoring gut barrier and microbiome regulation. Frontiers in
+fragilis prevents \emph{Clostridium difficile} infection in a mouse
+model by restoring gut barrier and microbiome regulation. Frontiers in
 Microbiology \textbf{9}.
@@
 \leavevmode\hypertarget{ref-Normington2021}{}%
-42. \textbf{Normington C}, \textbf{Moura IB}, \textbf{Bryant JA},
+41. \textbf{Normington C}, \textbf{Moura IB}, \textbf{Bryant JA},
 \textbf{Ewin DJ}, \textbf{Clark EV}, \textbf{Kettle MJ}, \textbf{Harris
@@
 \textbf{Ford CB}, \textbf{Wilcox MH}, \textbf{Buckley AM}. 2021.
-Biofilms harbour clostridioides difficile, serving as a reservoir for
-recurrent infection. npj Biofilms and Microbiomes \textbf{7}.
+Biofilms harbour \emph{Clostridioides difficile}, serving as a reservoir
+for recurrent infection. npj Biofilms and Microbiomes \textbf{7}.
 doi:\href{https://doi.org/10.1038/s41522-021-00184-w}{10.1038/s41522-021-00184-w}.
+
+\leavevmode\hypertarget{ref-Hassall2021}{}%
+42. \textbf{Hassall J}, \textbf{Cheng JKJ}, \textbf{Unnikrishnan M}.
+2021. Dissecting individual interactions between pathogenic and
+commensal bacteria within a multispecies gut microbial community.
+mSphere \textbf{6}.
+doi:\href{https://doi.org/10.1128/msphere.00013-21}{10.1128/msphere.00013-21}.
 
@@
 \textbf{Wagner M}, \textbf{Berry D}. 2020. Rational design of a
-microbial consortium of mucosal sugar utilizers reduces clostridiodes
-difficile colonization. Nature Communications \textbf{11}.
+microbial consortium of mucosal sugar utilizers reduces
+\emph{Clostridioides difficile} colonization. Nature Communications
+\textbf{11}.
 doi:\href{https://doi.org/10.1038/s41467-020-18928-1}{10.1038/s41467-020-18928-1}.
@@
 \textbf{Avesani V}, \textbf{Broeck JV}, \textbf{Brach P}, \textbf{Delmée
-M}, \textbf{Daube G}. 2016. Longitudinal survey of clostridium difficile
-presence and gut microbiota composition in a belgian nursing home. BMC
-Microbiology \textbf{16}.
+M}, \textbf{Daube G}. 2016. Longitudinal survey of \emph{Clostridium
+difficile} presence and gut microbiota composition in a belgian nursing
+home. BMC Microbiology \textbf{16}.
 doi:\href{https://doi.org/10.1186/s12866-016-0848-7}{10.1186/s12866-016-0848-7}.
@@
 \textbf{Lamendella R}, \textbf{Stewart DB}. 2016. Bacterial and fungal
-microbiota changes distinguish c. Difficile infection from other forms
-of diarrhea: Results of a prospective inpatient study. Frontiers in
-Microbiology \textbf{7}.
+microbiota changes distinguish \emph{C.~difficile} infection from other
+forms of diarrhea: Results of a prospective inpatient study. Frontiers
+in Microbiology \textbf{7}.
 doi:\href{https://doi.org/10.3389/fmicb.2016.00789}{10.3389/fmicb.2016.00789}.
@@
 \textbf{Sotoodehnejadnematalahi F}, \textbf{Siadat SD}. 2020. Intestinal
-microbiota in elderly inpatients with clostridioides difficile
+microbiota in elderly inpatients with \emph{Clostridioides difficile}
 infection. Infection and Drug Resistance \textbf{Volume 13}:2723--2731.
@@
 Characterization of gut microbiota in hospitalized patients with
-clostridioides difficile infection. Current Microbiology
+\emph{Clostridioides difficile} infection. Current Microbiology
 \textbf{77}:1673--1680.
@@
 
+\leavevmode\hypertarget{ref-Li2019}{}%
+50. \textbf{Li X}, \textbf{Chu Q}, \textbf{Huang Y}, \textbf{Xiao Y},
+\textbf{Song L}, \textbf{Zhu S}, \textbf{Kang Y}, \textbf{Lu S},
+\textbf{Xu J}, \textbf{Ren Z}. 2019. Consortium of probiotics attenuates
+colonization of \emph{Clostridioides difficile}. Frontiers in
+Microbiology \textbf{10}.
+doi:\href{https://doi.org/10.3389/fmicb.2019.02871}{10.3389/fmicb.2019.02871}.
+
 \leavevmode\hypertarget{ref-Theriot2011}{}%
-50. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
+51. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
 \textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
 Cefoperazone-treated mice as an experimental platform to assess
-differential virulence ofClostridium difficilestrains. Gut Microbes
-\textbf{2}:326--334.
+differential virulence of \emph{Clostridium difficile} strains. Gut
+Microbes \textbf{2}:326--334.
 doi:\href{https://doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}.
@@
 \leavevmode\hypertarget{ref-Kester2020}{}%
-51. \textbf{Kester JC}, \textbf{Brubaker DK}, \textbf{Velazquez J},
+52. \textbf{Kester JC}, \textbf{Brubaker DK}, \textbf{Velazquez J},
 \textbf{Wright C}, \textbf{Lauffenburger DA}, \textbf{Griffith LG}.
-2020. Clostridioides difficile-associated antibiotics alter human
+2020. \emph{Clostridioides difficile}-associated antibiotics alter human
 mucosal barrier functions by microbiome-independent mechanisms.
@@
 \leavevmode\hypertarget{ref-Bergstrom2020}{}%
-52. \textbf{Bergstrom K}, \textbf{Shan X}, \textbf{Casero D},
+53. \textbf{Bergstrom K}, \textbf{Shan X}, \textbf{Casero D},
 \textbf{Batushansky A}, \textbf{Lagishetty V}, \textbf{Jacobs JP},
@@
 
-\leavevmode\hypertarget{ref-Dieterle2018}{}%
-53. \textbf{Dieterle MG}, \textbf{Rao K}, \textbf{Young VB}. 2018. Novel
-therapies and preventative strategies for primary and
-recurrentClostridium difficileinfections. Annals of the New York Academy
-of Sciences \textbf{1435}:110--138.
-doi:\href{https://doi.org/10.1111/nyas.13958}{10.1111/nyas.13958}.
+\leavevmode\hypertarget{ref-McDonald2018b}{}%
+54. \textbf{McDonald LC}, \textbf{Gerding DN}, \textbf{Johnson S},
+\textbf{Bakken JS}, \textbf{Carroll KC}, \textbf{Coffin SE},
+\textbf{Dubberke ER}, \textbf{Garey KW}, \textbf{Gould CV},
+\textbf{Kelly C}, \textbf{Loo V}, \textbf{Sammons JS}, \textbf{Sandora
+TJ}, \textbf{Wilcox MH}. 2018. Clinical practice guidelines for
+\emph{Clostridium difficile} infection in adults and children: 2017
+update by the infectious diseases society of america (IDSA) and society
+for healthcare epidemiology of america (SHEA). Clinical Infectious
+Diseases \textbf{66}:e1--e48.
+doi:\href{https://doi.org/10.1093/cid/cix1085}{10.1093/cid/cix1085}.
+
+\leavevmode\hypertarget{ref-Ahmad2017}{}%
+55. \textbf{Ahmad SM}, \textbf{Blanco N}, \textbf{Dewart CM},
+\textbf{Dobosz A}, \textbf{Malani AN}. 2017. Laxative use in the setting
+of positive testing for \emph{Clostridium difficile} infection.
+Infection Control \& Hospital Epidemiology \textbf{38}:1513--1515.
+doi:\href{https://doi.org/10.1017/ice.2017.221}{10.1017/ice.2017.221}.
+
+\leavevmode\hypertarget{ref-Bilinskaya2018}{}%
+56. \textbf{Bilinskaya A}, \textbf{Goodlet KJ}, \textbf{Nailor MD}.
+2018. Evaluation of a best practice alert to reduce unnecessary
+\emph{Clostridium difficile} testing following receipt of a laxative.
+Diagnostic Microbiology and Infectious Disease \textbf{92}:50--55.
+doi:\href{https://doi.org/10.1016/j.diagmicrobio.2018.04.009}{10.1016/j.diagmicrobio.2018.04.009}.
+
+\leavevmode\hypertarget{ref-Cook2020}{}%
+57. \textbf{Cook PP}, \textbf{Nichols S}, \textbf{Coogan M},
+\textbf{Opera J}, \textbf{DeHart M}. 2020. Reduction in testing and
+change in testing algorithm associated with decrease in number of
+nosocomial \emph{Clostridioides} (\emph{Clostridium}) \emph{difficile}
+infections. American Journal of Infection Control
+\textbf{48}:1019--1022.
+doi:\href{https://doi.org/10.1016/j.ajic.2019.12.028}{10.1016/j.ajic.2019.12.028}.
+
+\leavevmode\hypertarget{ref-Appaneal2019}{}%
+58. \textbf{Appaneal HJ}, \textbf{Caffrey AR}, \textbf{Beganovic M},
+\textbf{Avramovic S}, \textbf{LaPlante KL}. 2019. Predictors of
+\emph{Clostridioides difficile} recurrence across a national cohort of
+veterans in outpatient, acute, and long-term care settings. American
+Journal of Health-System Pharmacy \textbf{76}:581--590.
+doi:\href{https://doi.org/10.1093/ajhp/zxz032}{10.1093/ajhp/zxz032}.
+
+\leavevmode\hypertarget{ref-Dubberke2011}{}%
+59. \textbf{Dubberke ER}, \textbf{Yan Y}, \textbf{Reske KA},
+\textbf{Butler AM}, \textbf{Doherty J}, \textbf{Pham V}, \textbf{Fraser
+VJ}. 2011. Development and validation of a \emph{Clostridium difficile}
+infection risk prediction model. Infection Control \& Hospital
+Epidemiology \textbf{32}:360--366.
+doi:\href{https://doi.org/10.1086/658944}{10.1086/658944}.
+
+\leavevmode\hypertarget{ref-Obritsch2010}{}%
+60. \textbf{Obritsch MD}, \textbf{Stroup JS}, \textbf{Carnahan RM},
+\textbf{Scheck DN}. 2010. \emph{Clostridium difficile}-associated
+diarrhea in a tertiary care medical center. Baylor University Medical
+Center Proceedings \textbf{23}:363--367.
+doi:\href{https://doi.org/10.1080/08998280.2010.11928654}{10.1080/08998280.2010.11928654}.
+
+\leavevmode\hypertarget{ref-White2019}{}%
+61. \textbf{White NC}, \textbf{Mendo-Lopez R}, \textbf{Papamichael K},
+\textbf{Cuddemi CA}, \textbf{Barrett C}, \textbf{Daugherty K},
+\textbf{Pollock N}, \textbf{Kelly CP}, \textbf{Alonso CD}. 2019.
+Laxative use does not preclude diagnosis or reduce disease severity in
+\emph{Clostridioides difficile} infection. Clinical Infectious Diseases
+\textbf{71}:1472--1478.
+doi:\href{https://doi.org/10.1093/cid/ciz978}{10.1093/cid/ciz978}.
+
+\leavevmode\hypertarget{ref-Poirier2019}{}%
+62. \textbf{Poirier D}, \textbf{Gervais P}, \textbf{Fuchs M},
+\textbf{Roussy J-F}, \textbf{Paquet-Bolduc B}, \textbf{Trottier S},
+\textbf{Longtin J}, \textbf{Loo VG}, \textbf{Longtin Y}. 2019.
+Predictors of \emph{Clostridioides difficile} infection among
+asymptomatic, colonized patients: A retrospective cohort study. Clinical
+Infectious Diseases \textbf{70}:2103--2210.
+doi:\href{https://doi.org/10.1093/cid/ciz626}{10.1093/cid/ciz626}.
+
+\leavevmode\hypertarget{ref-Huang2014}{}%
+63. \textbf{Huang H}, \textbf{Wu S}, \textbf{Chen R}, \textbf{Xu S},
+\textbf{Fang H}, \textbf{Weintraub A}, \textbf{Nord CE}. 2014. Risk
+factors of \emph{Clostridium difficile} infections among patients in a
+university hospital in shanghai, china. Anaerobe \textbf{30}:65--69.
+doi:\href{https://doi.org/10.1016/j.anaerobe.2014.08.015}{10.1016/j.anaerobe.2014.08.015}.
+
+\leavevmode\hypertarget{ref-Gorkiewicz2013}{}%
+64. \textbf{Gorkiewicz G}, \textbf{Thallinger GG}, \textbf{Trajanoski
+S}, \textbf{Lackner S}, \textbf{Stocker G}, \textbf{Hinterleitner T},
+\textbf{Gülly C}, \textbf{Högenauer C}. 2013. Alterations in the colonic
+microbiota in response to osmotic diarrhea. PLoS ONE \textbf{8}:e55817.
+doi:\href{https://doi.org/10.1371/journal.pone.0055817}{10.1371/journal.pone.0055817}.
+
+\leavevmode\hypertarget{ref-Shobar2016}{}%
+65. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
+\textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
+\textbf{Losurdo J}, \textbf{Engen PA}, \textbf{Sun Y}, \textbf{Koenig
+L}, \textbf{Mutlu EA}. 2016. The effects of bowel preparation on
+microbiota-related metrics differ in health and in inflammatory bowel
+disease and for the mucosal and luminal microbiota compartments.
+Clinical and Translational Gastroenterology \textbf{7}:e143.
+doi:\href{https://doi.org/10.1038/ctg.2015.54}{10.1038/ctg.2015.54}.
+
+\leavevmode\hypertarget{ref-Drago2016}{}%
+66. \textbf{Drago L}, \textbf{Toscano M}, \textbf{Grandi RD},
+\textbf{Casini V}, \textbf{Pace F}. 2016. Persisting changes of
+intestinal microbiota after bowel lavage and colonoscopy. European
+Journal of Gastroenterology \& Hepatology \textbf{28}:532--537.
+doi:\href{https://doi.org/10.1097/meg.0000000000000581}{10.1097/meg.0000000000000581}.
+
+\leavevmode\hypertarget{ref-Jalanka2014}{}%
+67. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
+\textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
+\textbf{Gowland P}, \textbf{Hoad C}, \textbf{Garsed K}, \textbf{Lam C},
+\textbf{Palva A}, \textbf{Spiller RC}, \textbf{Vos WM de}. 2014. Effects
+of bowel cleansing on the intestinal microbiota. Gut
+\textbf{64}:1562--1568.
+doi:\href{https://doi.org/10.1136/gutjnl-2014-307240}{10.1136/gutjnl-2014-307240}.
+
+\leavevmode\hypertarget{ref-Kampouri2021}{}%
+68. \textbf{Kampouri E}, \textbf{Croxatto A}, \textbf{Prod'hom G},
+\textbf{Guery B}. 2021. \emph{Clostridioides difficile} infection, still
+a long way to go. Journal of Clinical Medicine \textbf{10}:389.
+doi:\href{https://doi.org/10.3390/jcm10030389}{10.3390/jcm10030389}.
 
 \leavevmode\hypertarget{ref-Kelly2021}{}%
-54. \textbf{Kelly CR}, \textbf{Yen EF}, \textbf{Grinspan AM},
+69. \textbf{Kelly CR}, \textbf{Yen EF}, \textbf{Grinspan AM},
 \textbf{Kahn SA}, \textbf{Atreja A}, \textbf{Lewis JD}, \textbf{Moore
@@
 
+\leavevmode\hypertarget{ref-Dieterle2018}{}%
+70. \textbf{Dieterle MG}, \textbf{Rao K}, \textbf{Young VB}. 2018. Novel
+therapies and preventative strategies for primary and recurrent
+\emph{Clostridiuml difficile} infections. Annals of the New York Academy
+of Sciences \textbf{1435}:110--138.
+doi:\href{https://doi.org/10.1111/nyas.13958}{10.1111/nyas.13958}.
+
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-55. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+71. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-56. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+72. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \leavevmode\hypertarget{ref-Quast2012}{}%
-57. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
+73. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
 \textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
@@
 \leavevmode\hypertarget{ref-Cole2013}{}%
-58. \textbf{Cole JR}, \textbf{Wang Q}, \textbf{Fish JA}, \textbf{Chai
+74. \textbf{Cole JR}, \textbf{Wang Q}, \textbf{Fish JA}, \textbf{Chai
 B}, \textbf{McGarrell DM}, \textbf{Sun Y}, \textbf{Brown CT},
@@
 \leavevmode\hypertarget{ref-Vegan2018}{}%
-59. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
+75. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
 \textbf{Kindt R}, \textbf{Legendre P}, \textbf{McGlinn D},
@@
 \leavevmode\hypertarget{ref-mikropml}{}%
-60. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool KL},
+76. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool KL},
 \textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020.
@@
 \leavevmode\hypertarget{ref-Topcuoglu2021}{}%
-61. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
+77. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
 \textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
@@
 \leavevmode\hypertarget{ref-Topcuoglu2020}{}%
-62. \textbf{Topçuoğlu BD}, \textbf{Lesniak NA}, \textbf{Ruffin MT},
+78. \textbf{Topçuoğlu BD}, \textbf{Lesniak NA}, \textbf{Ruffin MT},
 \textbf{Wiens J}, \textbf{Schloss PD}. 2020. A framework for effective
@@
 \leavevmode\hypertarget{ref-r_citation_2020}{}%
-63. \textbf{R Core Team}. 2020. R: A language and environment for
+79. \textbf{R Core Team}. 2020. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \leavevmode\hypertarget{ref-Tidyverse2019}{}%
-64. \textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan J},
+80. \textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan J},
 \textbf{Chang W}, \textbf{McGowan LD}, \textbf{François R},
@@
 Asterisks indicate time points where the weight change or CFU/g was
-significantly different between groups by the Kruskal-Wallis test with
-Benjamini-Hochberg correction for testing multiple time points. The data
-presented are from a total of 5 separate experiments. \newpage
+significantly different (\emph{P} \textless{} 0.05) between groups by
+the Kruskal-Wallis test with Benjamini-Hochberg correction for testing
+multiple time points. The data presented are from a total of 5 separate
+experiments. \newpage
 
@@
 experiment. Each circle represents a sample from an individual mouse and
-the transparency of the symbol corresponds to the day post-challenge. B.
-Bray-Curtis distances of stool samples collected on either day 10 or 30
-post-challenge relative to the baseline sample collected for each mouse
-(before any drug treatments were administered). C. Shannon diversity in
-stool communities over time. The line indicates the median value for
-each treatment group. For B-C, the symboles represent samples from
+the transparency of the symbol corresponds to the day post-challenge.
+See Data Set S1, sheet 1 for PERMANOVA results. B. Bray-Curtis distances
+of stool samples collected on either day 10 or 30 post-challenge
+relative to the baseline sample collected for each mouse (before any
+drug treatments were administered). The symbols represent samples from
 individual mice and the line indicates the median value for each
-treatment group. D. 14 of the 33 genera affected by PEG treatment (Data
-Set S1, sheet X). The symbols represent the median relative abundance
-for a treatment group at either baseline (open circle) or 1-day post
-treatment (closed circle). Relative abundance data from paired baseline
-and 1-day post treatment stool sampes from the 5-day PEG and 5-day PEG
-plus 10-day recovery groups were analyzed by paired Wilcoxan signed-rank
+treatment group. C. Shannon diversity in stool communities over time.
+The line indicates the median value for each treatment group (Data Set
+S1, sheet 2). D. 14 of the 33 genera affected by PEG treatment (Data Set
+S1, sheet 3). The symbols represent the median relative abundance for a
+treatment group at either baseline (open circle) or 1-day post treatment
+(closed circle). Relative abundance data from paired baseline and 1-day
+post treatment stool sampes from the 5-day PEG and 5-day PEG plus 10-day
+recovery groups were analyzed by paired Wilcoxan signed-rank test with
+Benjamini-Hochberg correction for testing all identified genera. The
+clindamycin and 5-day PEG plus clindamycin treatment groups are shown on
+the plot for comparison. E. 6 of the 24 genera that were significantly
+different between the treatment groups over multiple time points (see
+Data Set S1, sheet 4 for complete list). The 5-day PEG plus clindamycin
+treatment group was only followed through 6-days post-challenge.
+Differences between treatment groups were identified by Kruskal-Wallis
 test with Benjamini-Hochberg correction for testing all identified
-genera. The clindamycin and 5-day PEG plus clindamycin treatment groups
-are shown on the plot for comparison. E. 6 of the 24 genera that were
-significantly different between the treatment groups over multiple time
-points, the 5-day PEG plus clindamycin treatment group was only followed
-through 6-days post-challenge. Differences between treatment groups were
-identified by Kruskal-Wallis test with Benjamini-Hochberg correction for
-testing all identified genera. The gray vertical line (D) and horizontal
-vertical lines (E) indicate the limit of detection. \newpage
+genera (*, \emph{P} \textless{} 0.05). The gray vertical line (D) and
+horizontal vertical lines (E) indicate the limit of detection. \newpage
 
@@
 does not result in more severe CDIs, although mucosal microbiota is
-altered.} A. Shannon diversity in cecum communities over time. The line
-indicates the median value for each treatment group. The colors of the
-symbols and lines represent individual and median relative abundance
-values for four treatment groups. A similar pattern was observed with
-the proximal and distal colon communities (Data Set S1, sheet X-X). B.
+altered.} A. Shannon diversity in cecum communities over time. The
+colors of the symbols and lines represent individual and median relative
+abundance values for four treatment groups (Data Set S1, sheet 6). B.
 PCoA of Bray-Curtis distances from mucosal samples collected throughout
@@
 of the symbol corresponds to the day post-challenge that the sample was
-collected. C. The median relative abundance of the 4 genera that were
-significantly different between the cecum communities of different
-treatment groups on day 6 and day 30 post-challenge (Data Set S1, sheet
-X). The gray vertical lines indicate the limit of detection. D-E. The
-histopathology summary scores from cecum and colon H\&E stained tissue
-sections. The summary score is the total score based on evaluation of
-edema, cellular infiltration, and inflammation in either the cecum or
-colon tissue. Each category is given a score ranging from 0-4, thus the
-maximum possible summary score is 12. The tissue for histology was
-collected at either 4 (D) or 6 (E) days post-challenge with the
-exception that one set of 5-day PEG treated mock-challenged mice were
-collected on day 0 post-challenge (first set of open purple circles in
-D). Histology data were analyzed with the Kruskal-Wallis test followed
-by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction.
-\newpage
+collected. See Data Set S1, sheet 7 for PERMANOVA results. C. The median
+relative abundance of the 4 genera that were significantly different
+between the cecum communities of different treatment groups on day 6 and
+day 30 post-challenge (Data Set S1, sheet 8). The gray vertical lines
+indicate the limit of detection. D-E. The histopathology summary scores
+from cecum and colon H\&E stained tissue sections. The summary score is
+the total score based on evaluation of edema, cellular infiltration, and
+inflammation in either the cecum or colon tissue. Each category is given
+a score ranging from 0-4, thus the maximum possible summary score is 12.
+The tissue for histology was collected at either 4 (D) or 6 (E) days
+post-challenge with the exception that one set of 5-day PEG treated
+mock-challenged mice were collected on day 0 post-challenge (first set
+of open purple circles in D). Histology data were analyzed with the
+Kruskal-Wallis test followed by pairwise Wilcoxon comparisons with
+Benjamini-Hochberg correction. *, \emph{P} \textless{} 0.05. \newpage
 
@@
 measured over time (N = 12-18 mice per time point) by serial dilutions.
-The black dashedd horizontal line represents the limit of detection for
+The black dashed horizontal line represents the limit of detection for
 the first serial dilution. For B and D, asterisks indicate time points
-where there was a significant difference between treatment groups by
-Kruskall-Wallis test with Benjamini-Hochberg correction for testing
-multiple time points. For B-D, each symbol represents a sample from an
-individual mouse and lines indicate the median value for each treatment
-group. C. PCoA of Bray-Curtis distances from stool communities collected
-over time (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2}
-= 0.19). Symbol transparency represents the day post-challenge of the
-experiment. For C-E, the B on the day legend or days post-challenge X
-axis stands for baseline and represents the sample that was collected
-prior to any drug treatments. D. Shannon diversity in stool communities
-over time. E. Median relative abundances per treatment group for 6 out
-of the 14 genera that were affected by treatment, but recovered close to
-baseline levels by 7 days post-challenge (Fig. 3E, Data Set S1, Sheet
-X). Paired stool sample relative abundance values either baseline and
-day 1 or baseline and day 7 were analyzed by paired Wilcoxan signed-rank
-test with Benjamini-Hochberg correction for testing all identified
-genera. Only genera that were different between baseline and 1-day
-post-challenge, but not baseline and 7-days post-challenge. The gray
-horizontal lines represents the limit of detection.
+where there was a significant difference (\emph{P} \textless{} 0.05)
+between treatment groups by Kruskall-Wallis test with Benjamini-Hochberg
+correction for testing multiple time points. For B-D, each symbol
+represents a sample from an individual mouse and lines indicate the
+median value for each treatment group. C. PCoA of Bray-Curtis distances
+from stool communities collected over time (day: R\textsuperscript{2} =
+0.43; group: R\textsuperscript{2} = 0.19, Data Set S1, sheet 11). Symbol
+transparency represents the day post-challenge of the experiment. For
+C-E, the B on the day legend or days post-challenge X-axis stands for
+baseline and represents the sample that was collected prior to any drug
+treatments. D. Shannon diversity in stool communities over time (Data
+Set S1, sheet 12). E. Median relative abundances per treatment group for
+6 out of the 14 genera that were affected by treatment, but recovered
+close to baseline levels by 7 days post-challenge (Fig. 3E, Data Set S1,
+sheets 13 and 14). Paired stool sample relative abundance values either
+baseline and day 1 or baseline and day 7 were analyzed by paired
+Wilcoxan signed-rank test with Benjamini-Hochberg correction for testing
+all identified genera. Only genera that were different between baseline
+and 1-day post-challenge, but not baseline and 7-days post-challenge are
+shown. The gray horizontal lines represents the limit of detection.
 
@@
 represents the limit of detection for the first serial dilution. C-D.
-Shannon diversity (C) and richness (D) in stool communities over time.
-B-D. Each symbol represents a stool sample from an individual mouse with
-the lines representing the median value for each treatment group.
-Asterisks indicate time points with significant differences between
-groups by a Kruskall-Wallis test with a Benjamini-Hochberg correction
-for testing multiple times points. Colored rectangles indicates the
-1-day PEG treatment period for applicable groups. The data presented are
-from a total of 3 separate experiments. \newpage
+Shannon diversity (C) and richness (D) in stool communities over time
+(Data Set S1, sheets 15 and 16). B-D. Each symbol represents a stool
+sample from an individual mouse with the lines representing the median
+value for each treatment group. Asterisks indicate time points with
+significant differences (\emph{P} \textless{} 0.05) between groups by
+the Kruskall-Wallis test with a Benjamini-Hochberg correction for
+testing multiple times points. Colored rectangles indicates the 1-day
+PEG treatment period for applicable groups. The data presented are from
+a total of 3 separate experiments. \newpage
 
@@
 an individual sample, the transparency of the circle corresponds to day
-post-challenge. B. Median relative abundances of 2 genera that were
-significantly different over multiple time points in mice that were
-administered either FMT or PBS solution via gavage C. Median relative
-abundances of the top 6 out of 24 genera that were significant over
-multiple timepoints, plotted over time (Data Set S1, Sheet X). For B-C,
+post-challenge. See Data Set S1, sheet 17 for PERMANOVA results. B.
+Median relative abundances of 2 genera that were significantly different
+over multiple time points in mice that were administered either FMT or
+PBS solution via gavage C. Median relative abundances of the top 6 out
+of 24 genera that were significant over multiple time points, plotted
+over time (see Data Set S1, sheet 18 for complete list). For B-C,
 colored rectangles indicates the 1-day PEG treatment period for
@@
 PEG treated mice.} A. Top ten bacteria that contributed to the random
-forest model trained on five day post-challenge community relative
-abundance data to predict whether mice would still be colonized with
+forest model trained on 5-day post-challenge community relative
+abundance data, predicting whether mice would still be colonized with
 \emph{C. difficile} 10 days post-challenge. The median (point) and
-interquartile range (lines) change in AUROC when the bacteria is left
+interquartile range (lines) change in AUROC when the bacteria were left
 out of the model by permutation feature importance analysis. B. The
 median relative abundances of the top ten bacteria that contributed to
-the random forest classification model at 5 days post-challenge . Color
-indicates whether the mice were still colonized with \emph{C. difficile}
-10 days post-challenge and the black horizontal line represents the
-median relative abundance for the two categories. Each symbol represents
-a stool sample from an individual mouse and the shape of the symbol
+the random forest classification model at 5 days post-challenge . Red
+indicates the mice were still colonized with \emph{C. difficile} while
+blue indicates mice that cleared \emph{C. difficile} 10 days
+post-challenge and the black horizontal line represents the median
+relative abundance for the two categories. Each symbol represents a
+stool sample from an individual mouse and the shape of the symbol
 indicates whether the PEG-treated mice received a 5-day (Fig. 1-3),
@@
 10 days post-challenge. However, for mice that received either a 5-day
-PEG treatment pre-\emph{C. difficile} challenge or a 1-day PEG treatment
-post-challenge recovery of colonization resistance was delayed because
-most mice were still colonized with \emph{C. difficile} 10 days
+PEG treatment prior to \emph{C. difficile} challenge or a 1-day PEG
+treatment post-challenge recovery of colonization resistance was delayed
+because most mice were still colonized with \emph{C. difficile} 10 days
 post-challenge. We found increased relative abundances of
@@
 post-challenge. B. Relative abundances of eight bacterial genera from
-day 0 post-challenge onward in each off the 10-day recovery mice. We
+day 0 post-challenge onward in each of the 10-day recovery mice. We
 analyzed samples from day 0 and day 8 post-challenge, which represented
-the the time points where mice were challenged with \emph{C. difficile}
-and when the median relative \emph{C. difficile} CFU stabilized for the
+the time points where mice were challenged with \emph{C. difficile} and
+when the median relative \emph{C. difficile} CFU stabilized for the
 group using the paired Wilcoxan signed-rank test, but no genera were
-significantly different after Benjamini-Hochberg correction. \newpage
+significantly different after Benjamini-Hochberg correction (Data Set
+S1, sheet 5). \newpage
 
@@
 genera were also significantly different in the proximal and distal
-colon tissues, Data Sheet S1, Sheet X). Each symbol represents a tissue
-sample from an individual mouse, the black horizontal lines represents
-the median relative abundances for each treatment group. B. The relative
-abundance of \emph{Peptostreptococacceae} in the three types of tissue
-sample communities over time. For A-B, the gray horizontal lines
-represent the limit of detection. \newpage
+colon tissues, Data Set S1, sheets 8, 9, and 10). Each symbol represents
+a tissue sample from an individual mouse, the black horizontal lines
+represents the median relative abundances for each treatment group. B.
+The relative abundance of \emph{Peptostreptococacceae} in the three
+types of tissue sample communities over time. For A-B, the gray
+horizontal lines represent the limit of detection. \newpage
 
@@
 
+\hypertarget{data-set-s1}{%
+\subsection{Data Set S1}\label{data-set-s1}}
+
+\textbf{Data Set S1, Sheets 1-19. Excel workbook with 19 sheets.}
+
+\textbf{Data Set S1, Sheet 1. PERMANOVA results for the stool
+communities from mice in the 5-day PEG subset.}
+
+\textbf{Data Set S1, Sheet 2. Shannon diversity analysis for the stool
+communities from mice in the 5-day PEG subset.}
+
+\textbf{Data Set S1, Sheet 3. Genera with relative abundances impacted
+by PEG treatment based on stool communities of 5-day PEG treated mice.}
+
+\textbf{Data Set S1, Sheet 4. Genera with relative abundances that vary
+between treatment groups in the stool communities from mice in the 5-day
+PEG subset.}
+
+\textbf{Data Set S1, Sheet 5. Statistical analysis results for genera
+with relative abundances that varied in stool communities in the 5-day
+PEG plus 10-day recovery mice between the day 1 and day 8 time points.}
+
+\textbf{Data Set S1, Sheet 6. Shannon diversity analysis for the cecum
+communities from mice in the 5-day PEG experiments.}
+
+\textbf{Data Set S1, Sheet 7. PERMANOVA results for the tissue
+communities from mice in the 5-day PEG subset.}
+
+\textbf{Data Set S1, Sheet 8. Genera with relative abundances that vary
+between treatment groups in the cecum communities from mice in the 5-day
+PEG esubset.}
+
+\textbf{Data Set S1, Sheet 9. Genera with relative abundances that vary
+between treatment groups in the proximal colon communities from mice in
+the 5-day PEG subset.}
+
+\textbf{Data Set S1, Sheet 10. Genera with relative abundances that vary
+between treatment groups in the distal colon communities from mice in
+the set of 5-day PEG subset.}
+
+\textbf{Data Set S1, Sheet 11. PERMANOVA results for the stool
+communities from mice in the set of 1-day PEG subset.}
+
+\textbf{Data Set S1, Sheet 12. Shannon diversity analysis for the stool
+communities from mice in the 1-day PEG experiments.}
+
+\textbf{Data Set S1, Sheet 13. Genera with different relative abundances
+between the baseline and day 1 time points in the 1-day PEG subset.}
+
+\textbf{Data Set S1, Sheet 14. Genera with different relative abundances
+between the baseline and day 7 time points in the 1-day PEG subset..}
+
+\textbf{Data Set S1, Sheet 15. Shannon diversity analysis for the stool
+communities from mice in the post-challenge PEG experiments.}
+
+\textbf{Data Set S1, Sheet 16. Richness analysis for the stool
+communities from mice in the post-challenge PEG experiments.}
+
+\textbf{Data Set S1, Sheet 17. PERMANOVA results for the stool
+communities from mice in the post-challenge PEG subset.}
+
+\textbf{Data Set S1, Sheet 18. Genera with relative abundances that vary
+between treatment groups in the stool communities from mice in the
+post-challenge PEG subset.}
+
+\textbf{Data Set S1, Sheet 19. AUROC results for the 100 different seeds
+from each of the 3 models tested.}
+
 \end{document}

### 2021-06-15 code/5_days_PEG_16S.R
Add legend for barbell plot

@@ -1207,6 +1207,21 @@ peg_impacted_genera_bb_plot <- peg_impacted_genera_bb %>%
         legend.position = "none") 
 save_plot(filename = paste0("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell_v2.png"), peg_impacted_genera_bb_plot, base_height = 7, base_width = 6)
 
+#Create legend for barbell plot
+barbell_peg_legend_plot <- tibble(x = 1:6, y = c("a", "b","a", "b","a", "b")) %>%  #Create fake dataframe 
+  ggplot(aes(x = x, y = y, shape = y))+
+  geom_point()+
+  #Create shape scale based on actual barbell plot above
+  scale_shape_manual(name = NULL,
+                     values = c(1, 16),
+                     breaks = c("a", "b"),
+                     labels = c("baseline", "1-day post PEG treatment"))+
+  guides(shape = guide_legend(ncol = 2))+ 
+  theme(legend.position = "bottom")+
+  theme_classic()
+barbell_peg_legend <- get_legend(barbell_peg_legend_plot)
+save_plot("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell_legend.png", barbell_peg_legend, base_height = .3, base_width = 2.8)
+
 #Examine genera of interest over time----
 genus_stools_t <- genus_stools %>% 
   mutate(day = as.integer(day))  #Day variable transformed to integer for over time plots

### 2021-06-15 code/figure_2.R

@@ -6,6 +6,8 @@ legend<- plot_grid(legend, NULL, NULL, labels = NULL, label_size = 12, nrow = 1,
 b <- ggdraw() + draw_image("results/figures/5_days_PEG_bc_combined_baseline.png", scale = 1)
 c <- ggdraw() + draw_image("results/figures/5_days_PEG_shannon_stools_median.png", scale = 1)
 d <- ggdraw() + draw_image("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell_v2.png")
+d_legend <- ggdraw() + draw_image("results/figures/5_days_PEG_genera_impacted_by_PEG_barbell_legend.png")
+d <- plot_grid(d, d_legend, labels = NULL, label_size = 12, ncol = 1, rel_heights = c(1, .08))
 e <- ggdraw() + draw_image("results/figures/5_days_PEG_genus_lineplot_stools.png", scale =0.95)
 
 top_panel <- plot_grid(a, b, labels = c("A", "B"), label_size = 12, nrow = 1, rel_widths = c(1, 1.2))

### 2021-06-15 submission/manuscript.Rmd
Address edits from Pat

@@
 title: ''
-csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
+csl: msphere.csl #Get themes at https://github.com/citation-style-language/styles
 fontsize: 11pt
@@
 Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota.
-However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear.
+However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350 also alter the microbiota.
+But whether PEG impacts CDI susceptibility and clearance is unclear.
 To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C.
@@
 We used clindamycin-treated mice as a control because they consistently clear *C.
-difficile* within 10 days post-challenge (dpc).
-PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remain colonized for up to 30 dpc.
+difficile* within 10 days post-challenge.
+PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remained colonized for up to 30 days post-challenge.
 In contrast, 1-day PEG treated mice were transiently colonized, clearing *C.
-difficile* within 7 dpc.
+difficile* within 7 days post-challenge.
 To examine how PEG treatment impacts clearance, we administered a 1-day PEG treatment to clindamycin-treated, *C.
@@
 Administering PEG to mice after *C.
-difficile* challenge prolonged colonization up to 30 dpc.
-When we trained a random forest model with community data from 5 dpc, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90).
-Examining the dynamics of these bacterial during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C.
+difficile* challenge prolonged colonization up to 30 days post-challenge.
+When we trained a random forest model with community data from 5 days post-challenge, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90).
+Examining the dynamics of these bacterial populations during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C.
 difficile* colonization in PEG-treated mice.
+Thus, the osmotic laxative, PEG, rendered mice susceptible to *C.
+difficile* colonization and hindered clearance.
 
 ## Importance
-Diarrheal samples induced by medications such as laxatives are typically rejected for *Clostridiodes difficile* testing.
-However, there are some microbiota similarities between diarrheal and *C.
-difficile* infection (CDI) communities such as lower diversity compared to samples from healthy patients, which lead us to hypothesize that diarrhea may be an indicator of *C.
+Diarrheal samples from patients taking laxatives are typically rejected for *Clostridiodes difficile* testing.
+However, there are similarities between the bacterial communities from people with diarrhea or *C.
+difficile* infections (CDI) including lower diversity compared to communities from healthy patients, which led us to hypothesize that diarrhea may be an indicator of *C.
 difficile* risk.
@@
 difficile* challenge.
-Our findings suggest the osmotic laxative disrupts colonization resistance to *C.
-difficile*, as well as clearance in mice already colonized with *C.
+Our findings suggest that osmotic laxatives disrupt colonization resistance to *C.
+difficile*, and prevent clearance among mice already colonized with *C.
 difficile*.
@@
 Our group has shown that when *C.
-difficile* negative samples from patients are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and *C.
+difficile* negative samples from patients are separated into two groups based on stool consistency, there are similar microbiota features between samples from CDI patients and *C.
 difficile* negative patients with diarrhea compared to samples that were *C.
 difficile* negative with non-diarrheal stool consistency [@Schubert2014].
+The similar community features between CDI patients and patients with diarrhea included low alpha diversity and only 6 bacterial taxa had higher relative abundances in communities from CDI patients.
 These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.
 
-Osmotic laxatives can lead to diarrhea depending on the dose administered and temporarily disrupt the human intestinal microbiota [@Nagata2019].
+depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019].
 The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies.
@@
 For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure.
-Many of the FMT studies to date rationalize the use of laxatives [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C.
+Many of the FMT studies to date rationalize the use of laxatives prior to the FMT [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C.
 difficile* spores and toxins from the intestine [@Liacouras1996].
@@
 difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020].
-Although two groups have now shown PEG treatment alone renders mice susceptible to *C.
+Although two groups have now shown that PEG treatment alone renders mice susceptible to *C.
 difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C.
@@
 Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C.
-difficile* colonization and severe CDI than mice treated with clindamycin, and if PEG treatment after challenge can promote *C.
+difficile* colonization and severe CDI than mice treated with clindamycin, and whether PEG treatment after challenge can promote *C.
 difficile* clearance.
@@
 difficile*.
-We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C.
+We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders mice susceptible to *C.
 difficile* colonization, with mice typically clearing *C.
@@
 
-Since none of the communities completely recovered in the follow-up period after treatments, we next profiled community diversity and composition.
+Because time and treatment group influenced most of the variation between communities, we next profiled community diversity and composition.
 We examined the alpha diversity dynamics by calculating the communities' Shannon diversity.
-Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-challenge (Fig.
+Although both clindamycin and PEG treatments decreased diversity, the Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin alone through thirty days post-challenge (Fig.
 2C; Data Set S1, sheet 2).
-We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment.
-We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, sheet 3).
+We next identified the bacterial genera whose relative abundances shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment.
+We found `r PEG_genera` genera whose relative abundances were altered by PEG treatment (Data Set S1, sheet 3).
 The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased.
@@
 Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C.
-difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.
+difficile* challenge on the cecum and colon was similar between PEG and clindamycin treated mice.
 
@@
 difficile* 10 days post-challenge.
-We chose 5 days post-challenge because that was the earliest time point where we saw a treatment effect in the post-challenge 1-day PEG experiments.
-The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, sheet 19), suggesting the 5-day post challenge microbiota is an excellent predictor of prolonged *C.
+We chose to predict the status based on communities 5 days post-challenge because that was the earliest time point where we saw a treatment effect in the post-challenge 1-day PEG experiments.
+The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, sheet 19) and indicated that the 5-day post challenge microbiota was an excellent predictor of prolonged *C.
 difficile* colonization.
-Next, we performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C.
+Next, we performed a permutation importance test to identify the bacteria that were the top contributors to the random forest model for predicting prolonged *C.
 difficile *colonization.
@@
 Next, we focused on the 5 genera that had a greater than 1% relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time.
-We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, the relative abundance of *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
+We found *Enterobacteriaceae* and *Bacteroides* tended to consistently have a higher relative abundance, the relative abundance of *Akkermansia* was initially low and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C.
 difficile* (Fig.
@@
 difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear.
-By following osmotic laxative treated mice over time, we found that a 5-day PEG treatment before challenge resulted in prolonged *C.
-difficile* colonization, while a 1-day PEG treatment resulted in transient colonization.
+By following mice treated with an osmotic laxative over time, we found that a 5-day PEG treatment before challenge resulted in prolonged *C.
+difficile* colonization, while a 1-day PEG treatment resulted in transient colonization without the use of antibiotics.
 The differences in *C.
@@
 
-Considering laxatives are also used to prepare patients when administering fecal microbiota transplants via colonoscopy to treat recurrent CDIs, it will be important to determine if osmotic laxatives impact *C.
+Considering laxatives are also used to prepare patients when administering fecal microbiota transplants via colonoscopy to treat recurrent CDIs, it will be important to determine whether osmotic laxatives impact *C.
 difficile* clearance in the human intestinal tract.
@@
 ***C.
-difficile* challenge model.** Mice were challenged with 25 microliters off *C.
+difficile* challenge model.** Mice were challenged with 25 microliters of *C.
 difficile* 630 spores at 10^5^ concentration, except for 1 experiment where the concentration was 10^3^ (Fig.
@@
 
-**16S rRNA gene sequencing.** Stool samples that were stored in the -80&deg;C were placed into 96-well plates for DNA extractions and library preparation.
+**16S rRNA gene sequencing.** Stool samples were stored at -80&deg;C and were placed into 96-well plates for DNA extractions and library preparation.
 DNA was extracted using the DNeasy Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf).
@@
 Paired sequencing reads were combined and  aligned with the SILVA (v.
-132) reference database [@Quast2012] and taxonomy was assigned with a modified version (v.
-16) of the Ribosomal Database Project (v.
-11.5) [@Cole2013].
+132) reference database [@Quast2012] and taxonomy was assigned with a modified version of the Ribosomal Database Project reference sequences (v.
+16) [@Cole2013].
 The error rate for are sequencing data was 0.0559% based on the 17 mock communities we ran with the samples.
 Samples were rarefied to 1,000 sequences, 1,000 times for alpha and beta diversity analyses in order to account for uneven sequencing across samples.
-All but 3 out of 17 water controls had less than 1000 sequences.
+All but 3 of the 17 water controls had fewer than 1000 sequences.
 PCoAs were generated based on Bray-Curtis Index distance matrices.
@@
 
-**Code availability.** Code for data analysis and generating this paper with accompanying figures is available at https://github.com/SchlossLab/Tomkovich_PEG3350_XXXX_2021.
+**Code availability.** Code for data analysis and generating this paper with accompanying figures is available at https://github.com/SchlossLab/Tomkovich_PEG3350_mSphere_2021.
 

### 2021-06-15 submission/manuscript.tex
Address edits from Pat

@@
 non-antibiotic medications such as the ubiquitous osmotic laxative
-polyethylene glycol (PEG) 3350, also alter the microbiota, but whether
+polyethylene glycol (PEG) 3350 also alter the microbiota. But whether
 PEG impacts CDI susceptibility and clearance is unclear. To examine how
@@
 control because they consistently clear \emph{C. difficile} within 10
-days post-challenge (dpc). PEG treatment alone was sufficient to render
-mice susceptible and 5-day PEG-treated mice remain colonized for up to
-30 dpc. In contrast, 1-day PEG treated mice were transiently colonized,
-clearing \emph{C. difficile} within 7 dpc. To examine how PEG treatment
-impacts clearance, we administered a 1-day PEG treatment to
-clindamycin-treated, \emph{C. difficile}-challenged mice. Administering
-PEG to mice after \emph{C. difficile} challenge prolonged colonization
-up to 30 dpc. When we trained a random forest model with community data
-from 5 dpc, we were able to predict which mice would exhibit prolonged
-colonization (AUROC = 0.90). Examining the dynamics of these bacterial
-during the post-challenge period revealed patterns in the relative
-abundances of \emph{Bacteroides}, \emph{Enterobacteriaceae},
-\emph{Porphyromonadaceae}, \emph{Lachnospiraceae}, and
-\emph{Akkermansia} that were associated with prolonged \emph{C.
-difficile} colonization in PEG-treated mice.
+days post-challenge. PEG treatment alone was sufficient to render mice
+susceptible and 5-day PEG-treated mice remained colonized for up to 30
+days post-challenge. In contrast, 1-day PEG treated mice were
+transiently colonized, clearing \emph{C. difficile} within 7 days
+post-challenge. To examine how PEG treatment impacts clearance, we
+administered a 1-day PEG treatment to clindamycin-treated, \emph{C.
+difficile}-challenged mice. Administering PEG to mice after \emph{C.
+difficile} challenge prolonged colonization up to 30 days
+post-challenge. When we trained a random forest model with community
+data from 5 days post-challenge, we were able to predict which mice
+would exhibit prolonged colonization (AUROC = 0.90). Examining the
+dynamics of these bacterial populations during the post-challenge period
+revealed patterns in the relative abundances of \emph{Bacteroides},
+\emph{Enterobacteriaceae}, \emph{Porphyromonadaceae},
+\emph{Lachnospiraceae}, and \emph{Akkermansia} that were associated with
+prolonged \emph{C. difficile} colonization in PEG-treated mice. Thus,
+the osmotic laxative, PEG, rendered mice susceptible to \emph{C.
+difficile} colonization and hindered clearance.
 
@@
 
-Diarrheal samples induced by medications such as laxatives are typically
-rejected for \emph{Clostridiodes difficile} testing. However, there are
-some microbiota similarities between diarrheal and \emph{C. difficile}
-infection (CDI) communities such as lower diversity compared to samples
-from healthy patients, which lead us to hypothesize that diarrhea may be
-an indicator of \emph{C. difficile} risk. We explored how osmotic
-laxatives disrupt the microbiota's colonization resistance to \emph{C.
-difficile} by administering a laxative to mice either before or after
-\emph{C. difficile} challenge. Our findings suggest the osmotic laxative
-disrupts colonization resistance to \emph{C. difficile}, as well as
-clearance in mice already colonized with \emph{C. difficile}.
-Considering that most hospitals recommend not performing \emph{C.
-difficile} testing on patients taking laxatives and laxatives are used
-when administering fecal microbiota transplants via colonoscopy to
-patients with recurrent CDIs, further studies are needed to evaluate if
-laxatives impact microbiota colonization resistance in humans.
+Diarrheal samples from patients taking laxatives are typically rejected
+for \emph{Clostridiodes difficile} testing. However, there are
+similarities between the bacterial communities from people with diarrhea
+or \emph{C. difficile} infections (CDI) including lower diversity
+compared to communities from healthy patients, which led us to
+hypothesize that diarrhea may be an indicator of \emph{C. difficile}
+risk. We explored how osmotic laxatives disrupt the microbiota's
+colonization resistance to \emph{C. difficile} by administering a
+laxative to mice either before or after \emph{C. difficile} challenge.
+Our findings suggest that osmotic laxatives disrupt colonization
+resistance to \emph{C. difficile}, and prevent clearance among mice
+already colonized with \emph{C. difficile}. Considering that most
+hospitals recommend not performing \emph{C. difficile} testing on
+patients taking laxatives and laxatives are used when administering
+fecal microbiota transplants via colonoscopy to patients with recurrent
+CDIs, further studies are needed to evaluate if laxatives impact
+microbiota colonization resistance in humans.
 
@@
 shown that when \emph{C. difficile} negative samples from patients are
-separated into two groups based on stool consistency, there are shared
-microbiota features such as lower alpha diversity in samples from CDI
-patients and \emph{C. difficile} negative patients with diarrhea
-compared to samples that were \emph{C. difficile} negative with
-non-diarrheal stool consistency (12). These results led to the
-hypothesis that bacterial communities from patients experiencing
-diarrhea are susceptible to developing CDIs, regardless of how they
-developed diarrhea.
-
-Osmotic laxatives can lead to diarrhea depending on the dose
-administered and temporarily disrupt the human intestinal microbiota
-(13). The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is
-found in Miralax, Nulytely, and Golytely and is also commonly used as
-bowel preparation for colonoscopies. Interestingly, previous studies
-have shown that treating mice with PEG alone altered microbiota
-composition, reduced acetate and butyrate production, altered the mucus
-barrier, and rendered the mice susceptible to \emph{C. difficile}
-colonization (14--17). The mucus barrier is thought to mediate
-protection from CDIs by protecting intestinal epithelial cells from the
-toxins produced by \emph{C. difficile} (18, 19). Whether laxative
-administration results in more severe CDIs in mice and how long mice
-remain colonized with \emph{C. difficile} after challenge is unclear.
+separated into two groups based on stool consistency, there are similar
+microbiota features between samples from CDI patients and \emph{C.
+difficile} negative patients with diarrhea compared to samples that were
+\emph{C. difficile} negative with non-diarrheal stool consistency (12).
+The similar community features between CDI patients and patients with
+diarrhea included low alpha diversity and only 6 bacterial taxa had
+higher relative abundances in communities from CDI patients. These
+results led to the hypothesis that bacterial communities from patients
+experiencing diarrhea are susceptible to developing CDIs, regardless of
+how they developed diarrhea.
+
+depending on the dose administered, osmotic laxatives can lead to
+diarrhea and temporarily disrupt the human intestinal microbiota (13).
+The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
+in Miralax, Nulytely, and Golytely and is also commonly used as bowel
+preparation for colonoscopies. Interestingly, previous studies have
+shown that treating mice with PEG alone altered microbiota composition,
+reduced acetate and butyrate production, altered the mucus barrier, and
+rendered the mice susceptible to \emph{C. difficile} colonization
+(14--17). The mucus barrier is thought to mediate protection from CDIs
+by protecting intestinal epithelial cells from the toxins produced by
+\emph{C. difficile} (18, 19). Whether laxative administration results in
+more severe CDIs in mice and how long mice remain colonized with
+\emph{C. difficile} after challenge is unclear.
 
@@
 to the procedure. Many of the FMT studies to date rationalize the use of
-laxatives (20--22) based on a 1996 case study with 2 pediatric patients
-where the authors suggested in the discussion that the laxative may help
-flush \emph{C. difficile} spores and toxins from the intestine (23).
+laxatives prior to the FMT (20--22) based on a 1996 case study with 2
+pediatric patients where the authors suggested in the discussion that
+the laxative may help flush \emph{C. difficile} spores and toxins from
+the intestine (23).
 
@@
 the microbiota and influence \emph{C. difficile} susceptibility and
-clearance (24--26). Although two groups have now shown PEG treatment
-alone renders mice susceptible to \emph{C. difficile} (15, 17), these
-studies have raised additional questions regarding the dynamics and
-severity of infection as well as the role of laxative treatment in
-\emph{C. difficile} clearance. Here, we characterized how long
-PEG-treated mice remain susceptible, whether PEG treatment results in
-more sustained \emph{C. difficile} colonization and severe CDI than mice
-treated with clindamycin, and if PEG treatment after challenge can
-promote \emph{C. difficile} clearance. Addressing these questions will
-better inform how we think about laxatives and diarrhea in the context
-of CDIs.
+clearance (24--26). Although two groups have now shown that PEG
+treatment alone renders mice susceptible to \emph{C. difficile} (15,
+17), these studies have raised additional questions regarding the
+dynamics and severity of infection as well as the role of laxative
+treatment in \emph{C. difficile} clearance. Here, we characterized how
+long PEG-treated mice remain susceptible, whether PEG treatment results
+in more sustained \emph{C. difficile} colonization and severe CDI than
+mice treated with clindamycin, and whether PEG treatment after challenge
+can promote \emph{C. difficile} clearance. Addressing these questions
+will better inform how we think about laxatives and diarrhea in the
+context of CDIs.
 
@@
 with our standard 10 mg/kg clindamycin treatment, which temporarily
-renders the mice susceptible to \emph{C. difficile} colonization, with
-mice typically clearing \emph{C. difficile} within 10 days
-post-challenge (9, 26). All three groups of PEG-treated mice were
-administered a 15\% PEG solution in the drinking water for 5-days. The
-first group received no additional treatment. The second group was also
-treated with clindamycin. A third group was allowed to recover for 10
-days prior to challenge (Fig. 1A). The PEG treatment resulted in weight
-loss for the 3 groups of mice, with the greatest change in weight
-observed on the fifth day of the PEG treatment. The mice recovered most
-of the lost weight by five days after treatment (Fig. 1B). After either
-the PEG, clindamycin, or PEG and clindamycin treatment all mice were
-challenged with 10\textsuperscript{5} \emph{C. difficile} 630 spores
-(Fig. 1A). All treatments rendered mice susceptible to \emph{C.
-difficile} colonization. In contrast to the mice that only received
-clindamycin, PEG-treated mice remained colonized with \emph{C.
-difficile} at a high level through thirty days post-challenge (Fig. 1C).
-The clindamycin-treated mice cleared \emph{C. difficile} within ten days
+renders mice susceptible to \emph{C. difficile} colonization, with mice
+typically clearing \emph{C. difficile} within 10 days post-challenge (9,
+26). All three groups of PEG-treated mice were administered a 15\% PEG
+solution in the drinking water for 5-days. The first group received no
+additional treatment. The second group was also treated with
+clindamycin. A third group was allowed to recover for 10 days prior to
+challenge (Fig. 1A). The PEG treatment resulted in weight loss for the 3
+groups of mice, with the greatest change in weight observed on the fifth
+day of the PEG treatment. The mice recovered most of the lost weight by
+five days after treatment (Fig. 1B). After either the PEG, clindamycin,
+or PEG and clindamycin treatment all mice were challenged with
+10\textsuperscript{5} \emph{C. difficile} 630 spores (Fig. 1A). All
+treatments rendered mice susceptible to \emph{C. difficile}
+colonization. In contrast to the mice that only received clindamycin,
+PEG-treated mice remained colonized with \emph{C. difficile} at a high
+level through thirty days post-challenge (Fig. 1C). The
+clindamycin-treated mice cleared \emph{C. difficile} within ten days
 post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were
@@
 
-Since none of the communities completely recovered in the follow-up
-period after treatments, we next profiled community diversity and
+Because time and treatment group influenced most of the variation
+between communities, we next profiled community diversity and
 composition. We examined the alpha diversity dynamics by calculating the
 communities' Shannon diversity. Although both clindamycin and PEG
-treatments decreased diversity, Shannon diversity was lower in the
+treatments decreased diversity, the Shannon diversity was lower in the
 groups of mice that received PEG treatment compared to those that
-received clindamycin through thirty days post-challenge (Fig. 2C; Data
-Set S1, sheet 2). We next examined the bacterial genera that shifted
-after PEG treatment by comparing the baseline samples of mice treated
-with only PEG to samples from the same mice one day post-PEG-treatment.
-We found 18 genera that were altered by PEG treatment (Data Set S1,
-sheet 3). The majority of the bacterial relative abundances decreased
-after the PEG treatment, but the relative abundance among members of the
+received clindamycin alone through thirty days post-challenge (Fig. 2C;
+Data Set S1, sheet 2). We next identified the bacterial genera whose
+relative abundances shifted after PEG treatment by comparing the
+baseline samples of mice treated with only PEG to samples from the same
+mice one day post-PEG-treatment. We found 18 genera whose relative
+abundances were altered by PEG treatment (Data Set S1, sheet 3). The
+majority of the bacterial relative abundances decreased after the PEG
+treatment, but the relative abundance among members of the
 \emph{Enterobacteriaceae} and \emph{Bacteroides} increased. The increase
@@
 Fig. 3E). Therefore, although PEG treatment had a disruptive effect on
-the mucosal microbiota, the impact of \emph{C. difficile} 630 challenge
-on the cecum and colon was similar between PEG and clindamycin treated
+the mucosal microbiota, the impact of \emph{C. difficile} challenge on
+the cecum and colon was similar between PEG and clindamycin treated
 mice.
@@
 were still colonized with \emph{C. difficile} 10 days post-challenge. We
-chose 5 days post-challenge because that was the earliest time point
-where we saw a treatment effect in the post-challenge 1-day PEG
-experiments. The random forest model had the highest performance (median
-AUROC = 0.90, Data Set S1, sheet 19), suggesting the 5-day post
-challenge microbiota is an excellent predictor of prolonged \emph{C.
-difficile} colonization. Next, we performed permutation importance to
-examine the bacteria that were the top contributors to the random forest
-model predicting prolonged \emph{C. difficile }colonization. We selected
-10 genera that contributed the most to our model's performance (Fig. 7A)
-and examined their relative abundance at 5 days post-challenge, the time
-point used to predict \emph{C. difficile} colonization status on day 10
-(Fig. 7B). Next, we focused on the 5 genera that had a greater than 1\%
-relative abundance in either the cleared or colonized mice and examined
-how the bacteria changed over time. We found \emph{Enterobacteriaceae}
-and \emph{Bacteroides} tended to have a higher relative abundance, the
-relative abundance of \emph{Akkermansia} was initially decreased and
-then increased, and \emph{Porphyromonadaceae} and \emph{Lachnospiraceae}
-had a lower relative abundance in the mice with prolonged colonization
-compared to the mice that cleared \emph{C. difficile} (Fig. 7C).
-Together these results suggest a combination of low and high abundance
-bacterial genera influence the prolonged colonization observed in 5-day
-PEG and post-challenge 1-day PEG treated mice.
+chose to predict the status based on communities 5 days post-challenge
+because that was the earliest time point where we saw a treatment effect
+in the post-challenge 1-day PEG experiments. The random forest model had
+the highest performance (median AUROC = 0.90, Data Set S1, sheet 19) and
+indicated that the 5-day post challenge microbiota was an excellent
+predictor of prolonged \emph{C. difficile} colonization. Next, we
+performed a permutation importance test to identify the bacteria that
+were the top contributors to the random forest model for predicting
+prolonged \emph{C. difficile }colonization. We selected 10 genera that
+contributed the most to our model's performance (Fig. 7A) and examined
+their relative abundance at 5 days post-challenge, the time point used
+to predict \emph{C. difficile} colonization status on day 10 (Fig. 7B).
+Next, we focused on the 5 genera that had a greater than 1\% relative
+abundance in either the cleared or colonized mice and examined how the
+bacteria changed over time. We found \emph{Enterobacteriaceae} and
+\emph{Bacteroides} tended to consistently have a higher relative
+abundance, the relative abundance of \emph{Akkermansia} was initially
+low and then increased, and \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} had a lower relative abundance in the mice with
+prolonged colonization compared to the mice that cleared \emph{C.
+difficile} (Fig. 7C). Together these results suggest a combination of
+low and high abundance bacterial genera influence the prolonged
+colonization observed in 5-day PEG and post-challenge 1-day PEG treated
+mice.
 
@@
 drugs such as laxatives disrupt colonization resistance was unclear. By
-following osmotic laxative treated mice over time, we found that a 5-day
-PEG treatment before challenge resulted in prolonged \emph{C. difficile}
-colonization, while a 1-day PEG treatment resulted in transient
-colonization. The differences in \emph{C. difficile} colonization
-dynamics between the 5- and 1-day PEG treated mice were associated with
-differences in how much the treatments disrupted the microbiota.
-Additionally, the intestinal communities of 5-day PEG treated mice did
-not regain colonization resistance after a 10-day recovery period. In
-contrast to the other 5-day PEG treatment groups, \emph{C. difficile}
-was not immediately detectable in the stools of most of the mice in the
-10-day recovery group. We also examined the impact of PEG treatment
-after \emph{C. difficile} challenge. In opposition to the hypothesis
-suggested by the literature, we found that PEG treatment prolonged
-colonization relative to mice that only recieved clindamycin treatment.
-We identified patterns in the relative abundances of \emph{Bacteroides},
+following mice treated with an osmotic laxative over time, we found that
+a 5-day PEG treatment before challenge resulted in prolonged \emph{C.
+difficile} colonization, while a 1-day PEG treatment resulted in
+transient colonization without the use of antibiotics. The differences
+in \emph{C. difficile} colonization dynamics between the 5- and 1-day
+PEG treated mice were associated with differences in how much the
+treatments disrupted the microbiota. Additionally, the intestinal
+communities of 5-day PEG treated mice did not regain colonization
+resistance after a 10-day recovery period. In contrast to the other
+5-day PEG treatment groups, \emph{C. difficile} was not immediately
+detectable in the stools of most of the mice in the 10-day recovery
+group. We also examined the impact of PEG treatment after \emph{C.
+difficile} challenge. In opposition to the hypothesis suggested by the
+literature, we found that PEG treatment prolonged colonization relative
+to mice that only recieved clindamycin treatment. We identified patterns
+in the relative abundances of \emph{Bacteroides},
 \emph{Enterobacteriaceae}, \emph{Akkermansia},
@@
 administering fecal microbiota transplants via colonoscopy to treat
-recurrent CDIs, it will be important to determine if osmotic laxatives
-impact \emph{C. difficile} clearance in the human intestinal tract. It
-is still unclear what the best administration route is because there
-have been no studies designed to evaluate the best administration route
-for FMTs (68). Nevertheless, results from the FMT National Registry
-where 85\% of FMTs were delivered by colonoscopy demonstrate FMTs are
-highly effective treatment for recurrent CDIs with 90\% achieving
-resolution in the 1 month follow-up window (69). A surprising number of
-studies continue to hypothesize that PEG or bowel preparation can clear
-\emph{C. difficile} spores and toxins despite the paucity of supporting
-evidence (20--23). There was even a clinical trial (NCT01630096)
-designed to examine whether administering PEG 3350 (NuLYTELY) prior to
-antibiotic treatment reduced disease severity that started recruitment
-in 2012 (70), but no results have been posted to date. Here we sought to
-evaluate the impact of treating \emph{C. difficile} colonized mice with
-PEG (with or without FMT) and found clearance was delayed. Further
-studies are needed to understand the impact of osmotic laxatives on
-\emph{C. difficile} colonization resistance and clearance in human
-patients receiving FMTs.
+recurrent CDIs, it will be important to determine whether osmotic
+laxatives impact \emph{C. difficile} clearance in the human intestinal
+tract. It is still unclear what the best administration route is because
+there have been no studies designed to evaluate the best administration
+route for FMTs (68). Nevertheless, results from the FMT National
+Registry where 85\% of FMTs were delivered by colonoscopy demonstrate
+FMTs are highly effective treatment for recurrent CDIs with 90\%
+achieving resolution in the 1 month follow-up window (69). A surprising
+number of studies continue to hypothesize that PEG or bowel preparation
+can clear \emph{C. difficile} spores and toxins despite the paucity of
+supporting evidence (20--23). There was even a clinical trial
+(NCT01630096) designed to examine whether administering PEG 3350
+(NuLYTELY) prior to antibiotic treatment reduced disease severity that
+started recruitment in 2012 (70), but no results have been posted to
+date. Here we sought to evaluate the impact of treating \emph{C.
+difficile} colonized mice with PEG (with or without FMT) and found
+clearance was delayed. Further studies are needed to understand the
+impact of osmotic laxatives on \emph{C. difficile} colonization
+resistance and clearance in human patients receiving FMTs.
 
@@
 \textbf{\emph{C. difficile} challenge model.} Mice were challenged with
-25 microliters off \emph{C. difficile} 630 spores at
+25 microliters of \emph{C. difficile} 630 spores at
 10\textsuperscript{5} concentration, except for 1 experiment where the
@@
 
-\textbf{16S rRNA gene sequencing.} Stool samples that were stored in the
--80°C were placed into 96-well plates for DNA extractions and library
+\textbf{16S rRNA gene sequencing.} Stool samples were stored at -80°C
+and were placed into 96-well plates for DNA extractions and library
 preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit
@@
 132) reference database (73) and taxonomy was assigned with a modified
-version (v. 16) of the Ribosomal Database Project (v. 11.5) (74). The
-error rate for are sequencing data was 0.0559\% based on the 17 mock
-communities we ran with the samples. Samples were rarefied to 1,000
-sequences, 1,000 times for alpha and beta diversity analyses in order to
-account for uneven sequencing across samples. All but 3 out of 17 water
-controls had less than 1000 sequences. PCoAs were generated based on
-Bray-Curtis Index distance matrices. Permutational multivariate analysis
-of variance (PERMANOVA) tests were performed on mothur-generated
-Bray-Curtis distance matrices with the adonis function from the vegan R
-package (75).
+version of the Ribosomal Database Project reference sequences (v. 16)
+(74). The error rate for are sequencing data was 0.0559\% based on the
+17 mock communities we ran with the samples. Samples were rarefied to
+1,000 sequences, 1,000 times for alpha and beta diversity analyses in
+order to account for uneven sequencing across samples. All but 3 of the
+17 water controls had fewer than 1000 sequences. PCoAs were generated
+based on Bray-Curtis Index distance matrices. Permutational multivariate
+analysis of variance (PERMANOVA) tests were performed on
+mothur-generated Bray-Curtis distance matrices with the adonis function
+from the vegan R package (75).
 
@@
 paper with accompanying figures is available at
-\url{https://github.com/SchlossLab/Tomkovich_PEG3350_XXXX_2021}.
+\url{https://github.com/SchlossLab/Tomkovich_PEG3350_mSphere_2021}.
 
@@
 \leavevmode\hypertarget{ref-Britton2014}{}%
-1. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
-microbiota in resistance to colonization by \emph{Clostridium
-difficile}. Gastroenterology \textbf{146}:1547--1553.
-doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
+1. Britton RA, Young VB. 2014. Role of the intestinal microbiota in
+resistance to colonization by \emph{Clostridium difficile}.
+Gastroenterology 146:1547--1553.
 
 \leavevmode\hypertarget{ref-Maier2018}{}%
-2. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
-\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
-\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
-\textbf{Mori H}, \textbf{Patil KR}, \textbf{Bork P}, \textbf{Typas A}.
+2. Maier L, Pruteanu M, Kuhn M, Zeller G, Telzerow A, Anderson EE,
+Brochado AR, Fernandez KC, Dose H, Mori H, Patil KR, Bork P, Typas A.
 2018. Extensive impact of non-antibiotic drugs on human gut bacteria.
-Nature \textbf{555}:623--628.
-doi:\href{https://doi.org/10.1038/nature25979}{10.1038/nature25979}.
+Nature 555:623--628.
 
 \leavevmode\hypertarget{ref-LeBastard2017}{}%
-3. \textbf{Bastard QL}, \textbf{Al-Ghalith GA}, \textbf{Grégoire M},
-\textbf{Chapelet G}, \textbf{Javaudin F}, \textbf{Dailly E},
-\textbf{Batard E}, \textbf{Knights D}, \textbf{Montassier E}. 2017.
-Systematic review: Human gut dysbiosis induced by non-antibiotic
-prescription medications. Alimentary Pharmacology \& Therapeutics
-\textbf{47}:332--345.
-doi:\href{https://doi.org/10.1111/apt.14451}{10.1111/apt.14451}.
+3. Bastard QL, Al-Ghalith GA, Grégoire M, Chapelet G, Javaudin F, Dailly
+E, Batard E, Knights D, Montassier E. 2017. Systematic review: Human gut
+dysbiosis induced by non-antibiotic prescription medications. Alimentary
+Pharmacology \& Therapeutics 47:332--345.
 
 \leavevmode\hypertarget{ref-VichVila2020}{}%
-4. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
-T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
-\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
-\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
-\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
+4. Vila AV, Collij V, Sanna S, Sinha T, Imhann F, Bourgonje AR, Mujagic
+Z, Jonkers DMAE, Masclee AAM, Fu J, Kurilshikov A, Wijmenga C,
+Zhernakova A, Weersma RK. 2020. Impact of commonly used drugs on the
 composition and metabolic function of the gut microbiota. Nature
-Communications \textbf{11}.
-doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
+Communications 11.
 
 \leavevmode\hypertarget{ref-Oh2018}{}%
-5. \textbf{Oh J}, \textbf{Makar M}, \textbf{Fusco C}, \textbf{McCaffrey
-R}, \textbf{Rao K}, \textbf{Ryan EE}, \textbf{Washer L}, \textbf{West
-LR}, \textbf{Young VB}, \textbf{Guttag J}, \textbf{Hooper DC},
-\textbf{Shenoy ES}, \textbf{Wiens J}. 2018. A generalizable, data-driven
-approach to predict daily risk of \emph{Clostridium difficile} infection
-at two large academic health centers. Infection Control \& Hospital
-Epidemiology \textbf{39}:425--433.
-doi:\href{https://doi.org/10.1017/ice.2018.16}{10.1017/ice.2018.16}.
+5. Oh J, Makar M, Fusco C, McCaffrey R, Rao K, Ryan EE, Washer L, West
+LR, Young VB, Guttag J, Hooper DC, Shenoy ES, Wiens J. 2018. A
+generalizable, data-driven approach to predict daily risk of
+\emph{Clostridium difficile} infection at two large academic health
+centers. Infection Control \& Hospital Epidemiology 39:425--433.
 
 \leavevmode\hypertarget{ref-Mora2012}{}%
-6. \textbf{Mora AL}, \textbf{Salazar M}, \textbf{Pablo-Caeiro J},
-\textbf{Frost CP}, \textbf{Yadav Y}, \textbf{DuPont HL}, \textbf{Garey
-KW}. 2012. Moderate to high use of opioid analgesics are associated with
-an increased risk of \emph{Clostridium difficile} infection. The
-American Journal of the Medical Sciences \textbf{343}:277--280.
-doi:\href{https://doi.org/10.1097/maj.0b013e31822f42eb}{10.1097/maj.0b013e31822f42eb}.
+6. Mora AL, Salazar M, Pablo-Caeiro J, Frost CP, Yadav Y, DuPont HL,
+Garey KW. 2012. Moderate to high use of opioid analgesics are associated
+with an increased risk of \emph{Clostridium difficile} infection. The
+American Journal of the Medical Sciences 343:277--280.
 
 \leavevmode\hypertarget{ref-Nehra2018}{}%
-7. \textbf{Nehra AK}, \textbf{Alexander JA}, \textbf{Loftus CG},
-\textbf{Nehra V}. 2018. Proton pump inhibitors: Review of emerging
-concerns. Mayo Clinic Proceedings \textbf{93}:240--246.
-doi:\href{https://doi.org/10.1016/j.mayocp.2017.10.022}{10.1016/j.mayocp.2017.10.022}.
+7. Nehra AK, Alexander JA, Loftus CG, Nehra V. 2018. Proton pump
+inhibitors: Review of emerging concerns. Mayo Clinic Proceedings
+93:240--246.
 
 \leavevmode\hypertarget{ref-Krishna2013}{}%
-8. \textbf{Krishna SG}, \textbf{Zhao W}, \textbf{Apewokin SK},
-\textbf{Krishna K}, \textbf{Chepyala P}, \textbf{Anaissie EJ}. 2013.
-Risk factors, preemptive therapy, and antiperistaltic agents for
+8. Krishna SG, Zhao W, Apewokin SK, Krishna K, Chepyala P, Anaissie EJ.
+2013. Risk factors, preemptive therapy, and antiperistaltic agents for
 \emph{Clostridium difficile} infection in cancer patients. Transplant
 Infectious Disease n/a--n/a.
-doi:\href{https://doi.org/10.1111/tid.12112}{10.1111/tid.12112}.
 
 \leavevmode\hypertarget{ref-Tomkovich2019}{}%
-9. \textbf{Tomkovich S}, \textbf{Lesniak NA}, \textbf{Li Y},
-\textbf{Bishop L}, \textbf{Fitzgerald MJ}, \textbf{Schloss PD}. 2019.
-The proton pump inhibitor omeprazole does not promote
+9. Tomkovich S, Lesniak NA, Li Y, Bishop L, Fitzgerald MJ, Schloss PD.
+2019. The proton pump inhibitor omeprazole does not promote
 \emph{Clostridioides difficile} colonization in a murine model. mSphere
-\textbf{4}.
-doi:\href{https://doi.org/10.1128/msphere.00693-19}{10.1128/msphere.00693-19}.
+4.
 
 \leavevmode\hypertarget{ref-Vandeputte2015}{}%
-10. \textbf{Vandeputte D}, \textbf{Falony G}, \textbf{Vieira-Silva S},
-\textbf{Tito RY}, \textbf{Joossens M}, \textbf{Raes J}. 2015. Stool
-consistency is strongly associated with gut microbiota richness and
-composition, enterotypes and bacterial growth rates. Gut
-\textbf{65}:57--62.
-doi:\href{https://doi.org/10.1136/gutjnl-2015-309618}{10.1136/gutjnl-2015-309618}.
+10. Vandeputte D, Falony G, Vieira-Silva S, Tito RY, Joossens M, Raes J.
+2015. Stool consistency is strongly associated with gut microbiota
+richness and composition, enterotypes and bacterial growth rates. Gut
+65:57--62.
 
 \leavevmode\hypertarget{ref-VujkovicCvijin2020}{}%
-11. \textbf{Vujkovic-Cvijin I}, \textbf{Sklar J}, \textbf{Jiang L},
-\textbf{Natarajan L}, \textbf{Knight R}, \textbf{Belkaid Y}. 2020. Host
-variables confound gut microbiota studies of human disease. Nature
-\textbf{587}:448--454.
-doi:\href{https://doi.org/10.1038/s41586-020-2881-9}{10.1038/s41586-020-2881-9}.
+11. Vujkovic-Cvijin I, Sklar J, Jiang L, Natarajan L, Knight R, Belkaid
+Y. 2020. Host variables confound gut microbiota studies of human
+disease. Nature 587:448--454.
 
 \leavevmode\hypertarget{ref-Schubert2014}{}%
-12. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
-\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
-distinguish patients with \emph{Clostridium difficile} infection and
+12. Schubert AM, Rogers MAM, Ring C, Mogle J, Petrosino JP, Young VB,
+Aronoff DM, Schloss PD. 2014. Microbiome data distinguish patients with
+\emph{Clostridium difficile} infection and
 non-\emph{C.~difficile}-associated diarrhea from healthy controls. mBio
-\textbf{5}.
-doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
+5.
 
 \leavevmode\hypertarget{ref-Nagata2019}{}%
-13. \textbf{Nagata N}, \textbf{Tohya M}, \textbf{Fukuda S}, \textbf{Suda
-W}, \textbf{Nishijima S}, \textbf{Takeuchi F}, \textbf{Ohsugi M},
-\textbf{Tsujimoto T}, \textbf{Nakamura T}, \textbf{Shimomura A},
-\textbf{Yanagisawa N}, \textbf{Hisada Y}, \textbf{Watanabe K},
-\textbf{Imbe K}, \textbf{Akiyama J}, \textbf{Mizokami M},
-\textbf{Miyoshi-Akiyama T}, \textbf{Uemura N}, \textbf{Hattori M}. 2019.
-Effects of bowel preparation on the human gut microbiome and metabolome.
-Scientific Reports \textbf{9}.
-doi:\href{https://doi.org/10.1038/s41598-019-40182-9}{10.1038/s41598-019-40182-9}.
+13. Nagata N, Tohya M, Fukuda S, Suda W, Nishijima S, Takeuchi F, Ohsugi
+M, Tsujimoto T, Nakamura T, Shimomura A, Yanagisawa N, Hisada Y,
+Watanabe K, Imbe K, Akiyama J, Mizokami M, Miyoshi-Akiyama T, Uemura N,
+Hattori M. 2019. Effects of bowel preparation on the human gut
+microbiome and metabolome. Scientific Reports 9.
 
 \leavevmode\hypertarget{ref-Kashyap2013}{}%
-14. \textbf{Kashyap PC}, \textbf{Marcobal A}, \textbf{Ursell LK},
-\textbf{Larauche M}, \textbf{Duboc H}, \textbf{Earle KA},
-\textbf{Sonnenburg ED}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
-\textbf{Million M}, \textbf{Tache Y}, \textbf{Pasricha PJ},
-\textbf{Knight R}, \textbf{Farrugia G}, \textbf{Sonnenburg JL}. 2013.
-Complex interactions among diet, gastrointestinal transit, and gut
-microbiota in humanized mice. Gastroenterology \textbf{144}:967--977.
-doi:\href{https://doi.org/10.1053/j.gastro.2013.01.047}{10.1053/j.gastro.2013.01.047}.
+14. Kashyap PC, Marcobal A, Ursell LK, Larauche M, Duboc H, Earle KA,
+Sonnenburg ED, Ferreyra JA, Higginbottom SK, Million M, Tache Y,
+Pasricha PJ, Knight R, Farrugia G, Sonnenburg JL. 2013. Complex
+interactions among diet, gastrointestinal transit, and gut microbiota in
+humanized mice. Gastroenterology 144:967--977.
 
 \leavevmode\hypertarget{ref-Ferreyra2014}{}%
-15. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
-\textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
-Gut microbiota-produced succinate promotes \emph{C.~difficile} infection
-after antibiotic treatment or motility disturbance. Cell Host \& Microbe
-\textbf{16}:770--777.
-doi:\href{https://doi.org/10.1016/j.chom.2014.11.003}{10.1016/j.chom.2014.11.003}.
+15. Ferreyra JA, Wu KJ, Hryckowian AJ, Bouley DM, Weimer BC, Sonnenburg
+JL. 2014. Gut microbiota-produced succinate promotes \emph{C.~difficile}
+infection after antibiotic treatment or motility disturbance. Cell Host
+\& Microbe 16:770--777.
 
 \leavevmode\hypertarget{ref-Tropini2018}{}%
-16. \textbf{Tropini C}, \textbf{Moss EL}, \textbf{Merrill BD},
-\textbf{Ng KM}, \textbf{Higginbottom SK}, \textbf{Casavant EP},
-\textbf{Gonzalez CG}, \textbf{Fremin B}, \textbf{Bouley DM},
-\textbf{Elias JE}, \textbf{Bhatt AS}, \textbf{Huang KC},
-\textbf{Sonnenburg JL}. 2018. Transient osmotic perturbation causes
-long-term alteration to the gut microbiota. Cell
-\textbf{173}:1742--1754.e17.
-doi:\href{https://doi.org/10.1016/j.cell.2018.05.008}{10.1016/j.cell.2018.05.008}.
+16. Tropini C, Moss EL, Merrill BD, Ng KM, Higginbottom SK, Casavant EP,
+Gonzalez CG, Fremin B, Bouley DM, Elias JE, Bhatt AS, Huang KC,
+Sonnenburg JL. 2018. Transient osmotic perturbation causes long-term
+alteration to the gut microbiota. Cell 173:1742--1754.e17.
 
 \leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-17. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
-\textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
-Diarrhoeal events can trigger long-term \emph{Clostridium difficile}
-colonization with recurrent blooms. Nature Microbiology
-\textbf{5}:642--650.
-doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.
+17. VanInsberghe D, Elsherbini JA, Varian B, Poutahidis T, Erdman S,
+Polz MF. 2020. Diarrhoeal events can trigger long-term \emph{Clostridium
+difficile} colonization with recurrent blooms. Nature Microbiology
+5:642--650.
 
 \leavevmode\hypertarget{ref-Olson2013}{}%
-18. \textbf{Olson A}, \textbf{Diebel LN}, \textbf{Liberati DM}. 2013.
-Effect of host defenses on \emph{Clostridium difficile} toxininduced
-intestinal barrier injury. Journal of Trauma and Acute Care Surgery
-\textbf{74}:983--990.
-doi:\href{https://doi.org/10.1097/ta.0b013e3182858477}{10.1097/ta.0b013e3182858477}.
+18. Olson A, Diebel LN, Liberati DM. 2013. Effect of host defenses on
+\emph{Clostridium difficile} toxininduced intestinal barrier injury.
+Journal of Trauma and Acute Care Surgery 74:983--990.
 
 \leavevmode\hypertarget{ref-Diebel2014}{}%
-19. \textbf{Diebel LN}, \textbf{Liberati DM}. 2014. Reinforcement of the
-intestinal mucus layer protects against \emph{Clostridium difficile}
-intestinal injury in~vitro. Journal of the American College of Surgeons
-\textbf{219}:460--468.
-doi:\href{https://doi.org/10.1016/j.jamcollsurg.2014.05.005}{10.1016/j.jamcollsurg.2014.05.005}.
+19. Diebel LN, Liberati DM. 2014. Reinforcement of the intestinal mucus
+layer protects against \emph{Clostridium difficile} intestinal injury
+in~vitro. Journal of the American College of Surgeons 219:460--468.
 
 \leavevmode\hypertarget{ref-vanNood2013}{}%
-20. \textbf{Nood E van}, \textbf{Vrieze A}, \textbf{Nieuwdorp M},
-\textbf{Fuentes S}, \textbf{Zoetendal EG}, \textbf{Vos WM de},
-\textbf{Visser CE}, \textbf{Kuijper EJ}, \textbf{Bartelsman JFWM},
-\textbf{Tijssen JGP}, \textbf{Speelman P}, \textbf{Dijkgraaf MGW},
-\textbf{Keller JJ}. 2013. Duodenal infusion of donor feces for recurrent
-\emph{Clostridium difficile}. New England Journal of Medicine
-\textbf{368}:407--415.
-doi:\href{https://doi.org/10.1056/nejmoa1205037}{10.1056/nejmoa1205037}.
+20. Nood E van, Vrieze A, Nieuwdorp M, Fuentes S, Zoetendal EG, Vos WM
+de, Visser CE, Kuijper EJ, Bartelsman JFWM, Tijssen JGP, Speelman P,
+Dijkgraaf MGW, Keller JJ. 2013. Duodenal infusion of donor feces for
+recurrent \emph{Clostridium difficile}. New England Journal of Medicine
+368:407--415.
 
 \leavevmode\hypertarget{ref-Razik2017}{}%
-21. \textbf{Razik R}, \textbf{Osman M}, \textbf{Lieberman A},
-\textbf{Allegretti JR}, \textbf{Kassam Z}. 2017. Faecal microbiota
-transplantation for \emph{Clostridium difficile} infection: A
+21. Razik R, Osman M, Lieberman A, Allegretti JR, Kassam Z. 2017. Faecal
+microbiota transplantation for \emph{Clostridium difficile} infection: A
 multicentre study of non-responders. Medical Journal of Australia
-\textbf{207}:159--160.
-doi:\href{https://doi.org/10.5694/mja16.01452}{10.5694/mja16.01452}.
+207:159--160.
 
 \leavevmode\hypertarget{ref-Postigo2012}{}%
-22. \textbf{Postigo R}, \textbf{Kim JH}. 2012. Colonoscopic versus
-nasogastric fecal transplantation for the treatment of \emph{Clostridium
-difficile} infection: A review and pooled analysis. Infection
-\textbf{40}:643--648.
-doi:\href{https://doi.org/10.1007/s15010-012-0307-9}{10.1007/s15010-012-0307-9}.
+22. Postigo R, Kim JH. 2012. Colonoscopic versus nasogastric fecal
+transplantation for the treatment of \emph{Clostridium difficile}
+infection: A review and pooled analysis. Infection 40:643--648.
 
 \leavevmode\hypertarget{ref-Liacouras1996}{}%
-23. \textbf{Liacouras CA}, \textbf{Piccoli DA}. 1996. Whole-bowel
-irrigation as an adjunct to the treatment of chronic, relapsing
-\emph{Clostridium difficile} colitis. Journal of Clinical
-Gastroenterology \textbf{22}:186--189.
-doi:\href{https://doi.org/10.1097/00004836-199604000-00007}{10.1097/00004836-199604000-00007}.
+23. Liacouras CA, Piccoli DA. 1996. Whole-bowel irrigation as an adjunct
+to the treatment of chronic, relapsing \emph{Clostridium difficile}
+colitis. Journal of Clinical Gastroenterology 22:186--189.
 
 \leavevmode\hypertarget{ref-Schubert2015}{}%
-24. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against \emph{Clostridium
-difficile}. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+24. Schubert AM, Sinani H, Schloss PD. 2015. Antibiotic-induced
+alterations of the murine gut microbiota and subsequent effects on
+colonization resistance against \emph{Clostridium difficile}. mBio 6.
 
 \leavevmode\hypertarget{ref-Jenior2017}{}%
-25. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
-alternative nutrient niches during infection across distinct murine gut
-microbiomes. mSystems \textbf{2}.
-doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
+25. Jenior ML, Leslie JL, Young VB, Schloss PD. 2017. \emph{Clostridium
+difficile} colonizes alternative nutrient niches during infection across
+distinct murine gut microbiomes. mSystems 2.
 
 \leavevmode\hypertarget{ref-Tomkovich2020}{}%
-26. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
-\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
-antibiotic perturbation influence \emph{Clostridioides difficile}
-clearance in mice. mSphere \textbf{5}.
-doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
+26. Tomkovich S, Stough JMA, Bishop L, Schloss PD. 2020. The initial gut
+microbiota and response to antibiotic perturbation influence
+\emph{Clostridioides difficile} clearance in mice. mSphere 5.
 
 \leavevmode\hypertarget{ref-Reeves2011}{}%
-27. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
-\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
-interplay between microbiome dynamics and pathogen dynamics in a murine
-model of \emph{Clostridium difficile} infection \textbf{2}:145--158.
-doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.
+27. Reeves AE, Theriot CM, Bergin IL, Huffnagle GB, Schloss PD, Young
+VB. 2011. The interplay between microbiome dynamics and pathogen
+dynamics in a murine model of \emph{Clostridium difficile} infection
+2:145--158.
 
 \leavevmode\hypertarget{ref-Dieterle2020}{}%
-28. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
-\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
-\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
-\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
-K}. 2020. Systemic inflammatory mediators are effective biomarkers for
-predicting adverse outcomes in \emph{Clostridioides difficile}
-infection. mBio \textbf{11}.
-doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
+28. Dieterle MG, Putler R, Perry DA, Menon A, Abernathy-Close L, Perlman
+NS, Penkevich A, Standke A, Keidan M, Vendrov KC, Bergin IL, Young VB,
+Rao K. 2020. Systemic inflammatory mediators are effective biomarkers
+for predicting adverse outcomes in \emph{Clostridioides difficile}
+infection. mBio 11.
 
 \leavevmode\hypertarget{ref-Jenior2018}{}%
-29. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
-structure and metabolism of distinct cecal microbiomes during initial
-infection to promote sustained colonization. mSphere \textbf{3}.
-doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
+29. Jenior ML, Leslie JL, Young VB, Schloss PD. 2018. \emph{Clostridium
+difficile} alters the structure and metabolism of distinct cecal
+microbiomes during initial infection to promote sustained colonization.
+mSphere 3.
 
 \leavevmode\hypertarget{ref-Iino2007}{}%
-30. \textbf{Iino T}, \textbf{Mori K}, \textbf{Tanaka K}, \textbf{Suzuki
-K-i}, \textbf{Harayama S}. 2007. \emph{Oscillibacter valericigenes} gen.
-Nov., sp. Nov., a valerate-producing anaerobic bacterium isolated from
-the alimentary canal of a japanese corbicula clam. International Journal
-of Systematic and Evolutionary Microbiology \textbf{57}:1840--1845.
-doi:\href{https://doi.org/10.1099/ijs.0.64717-0}{10.1099/ijs.0.64717-0}.
+30. Iino T, Mori K, Tanaka K, Suzuki K-i, Harayama S. 2007.
+\emph{Oscillibacter valericigenes} gen. Nov., sp. Nov., a
+valerate-producing anaerobic bacterium isolated from the alimentary
+canal of a japanese corbicula clam. International Journal of Systematic
+and Evolutionary Microbiology 57:1840--1845.
 
 \leavevmode\hypertarget{ref-Jump2014}{}%
-31. \textbf{Jump RLP}, \textbf{Polinkovsky A}, \textbf{Hurless K},
-\textbf{Sitzlar B}, \textbf{Eckart K}, \textbf{Tomas M},
-\textbf{Deshpande A}, \textbf{Nerandzic MM}, \textbf{Donskey CJ}. 2014.
-Metabolomics analysis identifies intestinal microbiota-derived
-biomarkers of colonization resistance in clindamycin-treated mice. PLoS
-ONE \textbf{9}:e101267.
-doi:\href{https://doi.org/10.1371/journal.pone.0101267}{10.1371/journal.pone.0101267}.
+31. Jump RLP, Polinkovsky A, Hurless K, Sitzlar B, Eckart K, Tomas M,
+Deshpande A, Nerandzic MM, Donskey CJ. 2014. Metabolomics analysis
+identifies intestinal microbiota-derived biomarkers of colonization
+resistance in clindamycin-treated mice. PLoS ONE 9:e101267.
 
 \leavevmode\hypertarget{ref-McDonald2018}{}%
-32. \textbf{McDonald JAK}, \textbf{Mullish BH}, \textbf{Pechlivanis A},
-\textbf{Liu Z}, \textbf{Brignardello J}, \textbf{Kao D}, \textbf{Holmes
-E}, \textbf{Li JV}, \textbf{Clarke TB}, \textbf{Thursz MR},
-\textbf{Marchesi JR}. 2018. Inhibiting growth of \emph{Clostridioides
-difficile} by restoring valerate, produced by the intestinal microbiota.
-Gastroenterology \textbf{155}:1495--1507.e15.
-doi:\href{https://doi.org/10.1053/j.gastro.2018.07.014}{10.1053/j.gastro.2018.07.014}.
+32. McDonald JAK, Mullish BH, Pechlivanis A, Liu Z, Brignardello J, Kao
+D, Holmes E, Li JV, Clarke TB, Thursz MR, Marchesi JR. 2018. Inhibiting
+growth of \emph{Clostridioides difficile} by restoring valerate,
+produced by the intestinal microbiota. Gastroenterology
+155:1495--1507.e15.
 
 \leavevmode\hypertarget{ref-Pfeiffer2012}{}%
-33. \textbf{Pfeiffer N}, \textbf{Desmarchelier C}, \textbf{Blaut M},
-\textbf{Daniel H}, \textbf{Haller D}, \textbf{Clavel T}. 2012.
-\emph{Acetatifactor muris} gen. Nov., sp. Nov., a novel bacterium
+33. Pfeiffer N, Desmarchelier C, Blaut M, Daniel H, Haller D, Clavel T.
+2012. \emph{Acetatifactor muris} gen. Nov., sp. Nov., a novel bacterium
 isolated from the intestine of an obese mouse. Archives of Microbiology
-\textbf{194}:901--907.
-doi:\href{https://doi.org/10.1007/s00203-012-0822-1}{10.1007/s00203-012-0822-1}.
+194:901--907.
 
 \leavevmode\hypertarget{ref-Lawley2012}{}%
-34. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
-\textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
-\textbf{Goulding D}, \textbf{Rad R}, \textbf{Schreiber F},
-\textbf{Brandt C}, \textbf{Deakin LJ}, \textbf{Pickard DJ},
-\textbf{Duncan SH}, \textbf{Flint HJ}, \textbf{Clark TG},
-\textbf{Parkhill J}, \textbf{Dougan G}. 2012. Targeted restoration of
-the intestinal microbiota with a simple, defined bacteriotherapy
+34. Lawley TD, Clare S, Walker AW, Stares MD, Connor TR, Raisen C,
+Goulding D, Rad R, Schreiber F, Brandt C, Deakin LJ, Pickard DJ, Duncan
+SH, Flint HJ, Clark TG, Parkhill J, Dougan G. 2012. Targeted restoration
+of the intestinal microbiota with a simple, defined bacteriotherapy
 resolves relapsing \emph{Clostridium difficile} disease in mice. PLoS
-Pathogens \textbf{8}:e1002995.
-doi:\href{https://doi.org/10.1371/journal.ppat.1002995}{10.1371/journal.ppat.1002995}.
+Pathogens 8:e1002995.
 
 \leavevmode\hypertarget{ref-Frost2021}{}%
-35. \textbf{Frost LR}, \textbf{Cheng JKJ}, \textbf{Unnikrishnan M}.
-2021. \emph{Clostridioides difficile} biofilms: A mechanism of
-persistence in the gut? PLOS Pathogens \textbf{17}:e1009348.
-doi:\href{https://doi.org/10.1371/journal.ppat.1009348}{10.1371/journal.ppat.1009348}.
+35. Frost LR, Cheng JKJ, Unnikrishnan M. 2021. \emph{Clostridioides
+difficile} biofilms: A mechanism of persistence in the gut? PLOS
+Pathogens 17:e1009348.
 
 \leavevmode\hypertarget{ref-Engevik2021}{}%
-36. \textbf{Engevik MA}, \textbf{Danhof HA}, \textbf{Auchtung J},
-\textbf{Endres BT}, \textbf{Ruan W}, \textbf{Bassères E},
-\textbf{Engevik AC}, \textbf{Wu Q}, \textbf{Nicholson M}, \textbf{Luna
-RA}, \textbf{Garey KW}, \textbf{Crawford SE}, \textbf{Estes MK},
-\textbf{Lux R}, \textbf{Yacyshyn MB}, \textbf{Yacyshyn B},
-\textbf{Savidge T}, \textbf{Britton RA}, \textbf{Versalovic J}. 2021.
-Fusobacterium nucleatum adheres to \emph{Clostridioides difficile} via
-the RadD adhesin to enhance biofilm formation in intestinal mucus.
-Gastroenterology \textbf{160}:1301--1314.e8.
-doi:\href{https://doi.org/10.1053/j.gastro.2020.11.034}{10.1053/j.gastro.2020.11.034}.
+36. Engevik MA, Danhof HA, Auchtung J, Endres BT, Ruan W, Bassères E,
+Engevik AC, Wu Q, Nicholson M, Luna RA, Garey KW, Crawford SE, Estes MK,
+Lux R, Yacyshyn MB, Yacyshyn B, Savidge T, Britton RA, Versalovic J.
+2021. Fusobacterium nucleatum adheres to \emph{Clostridioides difficile}
+via the RadD adhesin to enhance biofilm formation in intestinal mucus.
+Gastroenterology 160:1301--1314.e8.
 
 \leavevmode\hypertarget{ref-Engevik2020}{}%
-37. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
-\textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
-\textbf{Luna RA}, \textbf{Hyser JM}, \textbf{Spinler JK},
-\textbf{Versalovic J}. 2020. Mucin-degrading microbes release
-monosaccharides that chemoattract \emph{Clostridioides difficile} and
-facilitate colonization of the human intestinal mucus layer. ACS
-Infectious Diseases \textbf{7}:1126--1142.
-doi:\href{https://doi.org/10.1021/acsinfecdis.0c00634}{10.1021/acsinfecdis.0c00634}.
+37. Engevik MA, Engevik AC, Engevik KA, Auchtung JM, Chang-Graham AL,
+Ruan W, Luna RA, Hyser JM, Spinler JK, Versalovic J. 2020.
+Mucin-degrading microbes release monosaccharides that chemoattract
+\emph{Clostridioides difficile} and facilitate colonization of the human
+intestinal mucus layer. ACS Infectious Diseases 7:1126--1142.
 
 \leavevmode\hypertarget{ref-Dejea2018}{}%
-38. \textbf{Dejea CM}, \textbf{Fathi P}, \textbf{Craig JM},
-\textbf{Boleij A}, \textbf{Taddese R}, \textbf{Geis AL}, \textbf{Wu X},
-\textbf{Shields CED}, \textbf{Hechenbleikner EM}, \textbf{Huso DL},
-\textbf{Anders RA}, \textbf{Giardiello FM}, \textbf{Wick EC},
-\textbf{Wang H}, \textbf{Wu S}, \textbf{Pardoll DM}, \textbf{Housseau
-F}, \textbf{Sears CL}. 2018. Patients with familial adenomatous
-polyposis harbor colonic biofilms containing tumorigenic bacteria.
-Science \textbf{359}:592--597.
-doi:\href{https://doi.org/10.1126/science.aah3648}{10.1126/science.aah3648}.
+38. Dejea CM, Fathi P, Craig JM, Boleij A, Taddese R, Geis AL, Wu X,
+Shields CED, Hechenbleikner EM, Huso DL, Anders RA, Giardiello FM, Wick
+EC, Wang H, Wu S, Pardoll DM, Housseau F, Sears CL. 2018. Patients with
+familial adenomatous polyposis harbor colonic biofilms containing
+tumorigenic bacteria. Science 359:592--597.
 
 \leavevmode\hypertarget{ref-Semenyuk2015}{}%
-39. \textbf{Semenyuk EG}, \textbf{Poroyko VA}, \textbf{Johnston PF},
-\textbf{Jones SE}, \textbf{Knight KL}, \textbf{Gerding DN},
-\textbf{Driks A}. 2015. Analysis of bacterial communities during
+39. Semenyuk EG, Poroyko VA, Johnston PF, Jones SE, Knight KL, Gerding
+DN, Driks A. 2015. Analysis of bacterial communities during
 \emph{Clostridium difficile} infection in the mouse. Infection and
-Immunity \textbf{83}:4383--4391.
-doi:\href{https://doi.org/10.1128/iai.00145-15}{10.1128/iai.00145-15}.
+Immunity 83:4383--4391.
 
 \leavevmode\hypertarget{ref-Deng2018}{}%
-40. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
-\textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
-\textbf{Fan H}, \textbf{Zhao X}, \textbf{Zhi F}. 2018. Bacteroides
-fragilis prevents \emph{Clostridium difficile} infection in a mouse
-model by restoring gut barrier and microbiome regulation. Frontiers in
-Microbiology \textbf{9}.
-doi:\href{https://doi.org/10.3389/fmicb.2018.02976}{10.3389/fmicb.2018.02976}.
+40. Deng H, Yang S, Zhang Y, Qian K, Zhang Z, Liu Y, Wang Y, Bai Y, Fan
+H, Zhao X, Zhi F. 2018. Bacteroides fragilis prevents \emph{Clostridium
+difficile} infection in a mouse model by restoring gut barrier and
+microbiome regulation. Frontiers in Microbiology 9.
 
 \leavevmode\hypertarget{ref-Normington2021}{}%
-41. \textbf{Normington C}, \textbf{Moura IB}, \textbf{Bryant JA},
-\textbf{Ewin DJ}, \textbf{Clark EV}, \textbf{Kettle MJ}, \textbf{Harris
-HC}, \textbf{Spittal W}, \textbf{Davis G}, \textbf{Henn MR},
-\textbf{Ford CB}, \textbf{Wilcox MH}, \textbf{Buckley AM}. 2021.
-Biofilms harbour \emph{Clostridioides difficile}, serving as a reservoir
-for recurrent infection. npj Biofilms and Microbiomes \textbf{7}.
-doi:\href{https://doi.org/10.1038/s41522-021-00184-w}{10.1038/s41522-021-00184-w}.
+41. Normington C, Moura IB, Bryant JA, Ewin DJ, Clark EV, Kettle MJ,
+Harris HC, Spittal W, Davis G, Henn MR, Ford CB, Wilcox MH, Buckley AM.
+2021. Biofilms harbour \emph{Clostridioides difficile}, serving as a
+reservoir for recurrent infection. npj Biofilms and Microbiomes 7.
 
 \leavevmode\hypertarget{ref-Hassall2021}{}%
-42. \textbf{Hassall J}, \textbf{Cheng JKJ}, \textbf{Unnikrishnan M}.
-2021. Dissecting individual interactions between pathogenic and
-commensal bacteria within a multispecies gut microbial community.
-mSphere \textbf{6}.
-doi:\href{https://doi.org/10.1128/msphere.00013-21}{10.1128/msphere.00013-21}.
+42. Hassall J, Cheng JKJ, Unnikrishnan M. 2021. Dissecting individual
+interactions between pathogenic and commensal bacteria within a
+multispecies gut microbial community. mSphere 6.
 
 \leavevmode\hypertarget{ref-Cirstea2018}{}%
-43. \textbf{Cirstea M}, \textbf{Radisavljevic N}, \textbf{Finlay BB}.
-2018. Good bug, bad bug: Breaking through microbial stereotypes. Cell
-Host \& Microbe \textbf{23}:10--13.
-doi:\href{https://doi.org/10.1016/j.chom.2017.12.008}{10.1016/j.chom.2017.12.008}.
+43. Cirstea M, Radisavljevic N, Finlay BB. 2018. Good bug, bad bug:
+Breaking through microbial stereotypes. Cell Host \& Microbe 23:10--13.
 
 \leavevmode\hypertarget{ref-BorgesCanha2015}{}%
-44. \textbf{Canha MB}. 2015. Role of colonic microbiota in colorectal
+44. Canha MB. 2015. Role of colonic microbiota in colorectal
 carcinogenesis: A systematic review. Revista Española de Enfermedades
-Digestivas \textbf{107}.
-doi:\href{https://doi.org/10.17235/reed.2015.3830/2015}{10.17235/reed.2015.3830/2015}.
+Digestivas 107.
 
 \leavevmode\hypertarget{ref-Pereira2020}{}%
-45. \textbf{Pereira FC}, \textbf{Wasmund K}, \textbf{Cobankovic I},
-\textbf{Jehmlich N}, \textbf{Herbold CW}, \textbf{Lee KS},
-\textbf{Sziranyi B}, \textbf{Vesely C}, \textbf{Decker T},
-\textbf{Stocker R}, \textbf{Warth B}, \textbf{Bergen M von},
-\textbf{Wagner M}, \textbf{Berry D}. 2020. Rational design of a
-microbial consortium of mucosal sugar utilizers reduces
-\emph{Clostridioides difficile} colonization. Nature Communications
-\textbf{11}.
-doi:\href{https://doi.org/10.1038/s41467-020-18928-1}{10.1038/s41467-020-18928-1}.
+45. Pereira FC, Wasmund K, Cobankovic I, Jehmlich N, Herbold CW, Lee KS,
+Sziranyi B, Vesely C, Decker T, Stocker R, Warth B, Bergen M von, Wagner
+M, Berry D. 2020. Rational design of a microbial consortium of mucosal
+sugar utilizers reduces \emph{Clostridioides difficile} colonization.
+Nature Communications 11.
 
 \leavevmode\hypertarget{ref-Rodriguez2016}{}%
-46. \textbf{Rodriguez C}, \textbf{Taminiau B}, \textbf{Korsak N},
-\textbf{Avesani V}, \textbf{Broeck JV}, \textbf{Brach P}, \textbf{Delmée
-M}, \textbf{Daube G}. 2016. Longitudinal survey of \emph{Clostridium
+46. Rodriguez C, Taminiau B, Korsak N, Avesani V, Broeck JV, Brach P,
+Delmée M, Daube G. 2016. Longitudinal survey of \emph{Clostridium
 difficile} presence and gut microbiota composition in a belgian nursing
-home. BMC Microbiology \textbf{16}.
-doi:\href{https://doi.org/10.1186/s12866-016-0848-7}{10.1186/s12866-016-0848-7}.
+home. BMC Microbiology 16.
 
 \leavevmode\hypertarget{ref-Sangster2016}{}%
-47. \textbf{Sangster W}, \textbf{Hegarty JP}, \textbf{Schieffer KM},
-\textbf{Wright JR}, \textbf{Hackman J}, \textbf{Toole DR},
-\textbf{Lamendella R}, \textbf{Stewart DB}. 2016. Bacterial and fungal
-microbiota changes distinguish \emph{C.~difficile} infection from other
-forms of diarrhea: Results of a prospective inpatient study. Frontiers
-in Microbiology \textbf{7}.
-doi:\href{https://doi.org/10.3389/fmicb.2016.00789}{10.3389/fmicb.2016.00789}.
+47. Sangster W, Hegarty JP, Schieffer KM, Wright JR, Hackman J, Toole
+DR, Lamendella R, Stewart DB. 2016. Bacterial and fungal microbiota
+changes distinguish \emph{C.~difficile} infection from other forms of
+diarrhea: Results of a prospective inpatient study. Frontiers in
+Microbiology 7.
 
 \leavevmode\hypertarget{ref-Vakili2020a}{}%
-48. \textbf{Vakili B}, \textbf{Fateh A}, \textbf{Aghdaei HA},
-\textbf{Sotoodehnejadnematalahi F}, \textbf{Siadat SD}. 2020. Intestinal
-microbiota in elderly inpatients with \emph{Clostridioides difficile}
-infection. Infection and Drug Resistance \textbf{Volume 13}:2723--2731.
-doi:\href{https://doi.org/10.2147/idr.s262019}{10.2147/idr.s262019}.
+48. Vakili B, Fateh A, Aghdaei HA, Sotoodehnejadnematalahi F, Siadat SD.
+2020. Intestinal microbiota in elderly inpatients with
+\emph{Clostridioides difficile} infection. Infection and Drug Resistance
+Volume 13:2723--2731.
 
 \leavevmode\hypertarget{ref-Vakili2020b}{}%
-49. \textbf{Vakili B}, \textbf{Fateh A}, \textbf{Aghdaei HA},
-\textbf{Sotoodehnejadnematalahi F}, \textbf{Siadat SD}. 2020.
-Characterization of gut microbiota in hospitalized patients with
+49. Vakili B, Fateh A, Aghdaei HA, Sotoodehnejadnematalahi F, Siadat SD.
+2020. Characterization of gut microbiota in hospitalized patients with
 \emph{Clostridioides difficile} infection. Current Microbiology
-\textbf{77}:1673--1680.
-doi:\href{https://doi.org/10.1007/s00284-020-01980-x}{10.1007/s00284-020-01980-x}.
+77:1673--1680.
 
 \leavevmode\hypertarget{ref-Li2019}{}%
-50. \textbf{Li X}, \textbf{Chu Q}, \textbf{Huang Y}, \textbf{Xiao Y},
-\textbf{Song L}, \textbf{Zhu S}, \textbf{Kang Y}, \textbf{Lu S},
-\textbf{Xu J}, \textbf{Ren Z}. 2019. Consortium of probiotics attenuates
-colonization of \emph{Clostridioides difficile}. Frontiers in
-Microbiology \textbf{10}.
-doi:\href{https://doi.org/10.3389/fmicb.2019.02871}{10.3389/fmicb.2019.02871}.
+50. Li X, Chu Q, Huang Y, Xiao Y, Song L, Zhu S, Kang Y, Lu S, Xu J, Ren
+Z. 2019. Consortium of probiotics attenuates colonization of
+\emph{Clostridioides difficile}. Frontiers in Microbiology 10.
 
 \leavevmode\hypertarget{ref-Theriot2011}{}%
-51. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
-\textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
-Cefoperazone-treated mice as an experimental platform to assess
-differential virulence of \emph{Clostridium difficile} strains. Gut
-Microbes \textbf{2}:326--334.
-doi:\href{https://doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}.
+51. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
+VB. 2011. Cefoperazone-treated mice as an experimental platform to
+assess differential virulence of \emph{Clostridium difficile} strains.
+Gut Microbes 2:326--334.
 
 \leavevmode\hypertarget{ref-Kester2020}{}%
-52. \textbf{Kester JC}, \textbf{Brubaker DK}, \textbf{Velazquez J},
-\textbf{Wright C}, \textbf{Lauffenburger DA}, \textbf{Griffith LG}.
-2020. \emph{Clostridioides difficile}-associated antibiotics alter human
-mucosal barrier functions by microbiome-independent mechanisms.
-Antimicrobial Agents and Chemotherapy \textbf{64}.
-doi:\href{https://doi.org/10.1128/aac.01404-19}{10.1128/aac.01404-19}.
+52. Kester JC, Brubaker DK, Velazquez J, Wright C, Lauffenburger DA,
+Griffith LG. 2020. \emph{Clostridioides difficile}-associated
+antibiotics alter human mucosal barrier functions by
+microbiome-independent mechanisms. Antimicrobial Agents and Chemotherapy
+64.
 
 \leavevmode\hypertarget{ref-Bergstrom2020}{}%
-53. \textbf{Bergstrom K}, \textbf{Shan X}, \textbf{Casero D},
-\textbf{Batushansky A}, \textbf{Lagishetty V}, \textbf{Jacobs JP},
-\textbf{Hoover C}, \textbf{Kondo Y}, \textbf{Shao B}, \textbf{Gao L},
-\textbf{Zandberg W}, \textbf{Noyovitz B}, \textbf{McDaniel JM},
-\textbf{Gibson DL}, \textbf{Pakpour S}, \textbf{Kazemian N},
-\textbf{McGee S}, \textbf{Houchen CW}, \textbf{Rao CV}, \textbf{Griffin
-TM}, \textbf{Sonnenburg JL}, \textbf{McEver RP}, \textbf{Braun J},
-\textbf{Xia L}. 2020. Proximal colonderived o-glycosylated mucus
-encapsulates and modulates the microbiota. Science
-\textbf{370}:467--472.
-doi:\href{https://doi.org/10.1126/science.aay7367}{10.1126/science.aay7367}.
+53. Bergstrom K, Shan X, Casero D, Batushansky A, Lagishetty V, Jacobs
+JP, Hoover C, Kondo Y, Shao B, Gao L, Zandberg W, Noyovitz B, McDaniel
+JM, Gibson DL, Pakpour S, Kazemian N, McGee S, Houchen CW, Rao CV,
+Griffin TM, Sonnenburg JL, McEver RP, Braun J, Xia L. 2020. Proximal
+colonderived o-glycosylated mucus encapsulates and modulates the
+microbiota. Science 370:467--472.
 
 \leavevmode\hypertarget{ref-McDonald2018b}{}%
-54. \textbf{McDonald LC}, \textbf{Gerding DN}, \textbf{Johnson S},
-\textbf{Bakken JS}, \textbf{Carroll KC}, \textbf{Coffin SE},
-\textbf{Dubberke ER}, \textbf{Garey KW}, \textbf{Gould CV},
-\textbf{Kelly C}, \textbf{Loo V}, \textbf{Sammons JS}, \textbf{Sandora
-TJ}, \textbf{Wilcox MH}. 2018. Clinical practice guidelines for
-\emph{Clostridium difficile} infection in adults and children: 2017
-update by the infectious diseases society of america (IDSA) and society
-for healthcare epidemiology of america (SHEA). Clinical Infectious
-Diseases \textbf{66}:e1--e48.
-doi:\href{https://doi.org/10.1093/cid/cix1085}{10.1093/cid/cix1085}.
+54. McDonald LC, Gerding DN, Johnson S, Bakken JS, Carroll KC, Coffin
+SE, Dubberke ER, Garey KW, Gould CV, Kelly C, Loo V, Sammons JS, Sandora
+TJ, Wilcox MH. 2018. Clinical practice guidelines for \emph{Clostridium
+difficile} infection in adults and children: 2017 update by the
+infectious diseases society of america (IDSA) and society for healthcare
+epidemiology of america (SHEA). Clinical Infectious Diseases 66:e1--e48.
 
 \leavevmode\hypertarget{ref-Ahmad2017}{}%
-55. \textbf{Ahmad SM}, \textbf{Blanco N}, \textbf{Dewart CM},
-\textbf{Dobosz A}, \textbf{Malani AN}. 2017. Laxative use in the setting
-of positive testing for \emph{Clostridium difficile} infection.
-Infection Control \& Hospital Epidemiology \textbf{38}:1513--1515.
-doi:\href{https://doi.org/10.1017/ice.2017.221}{10.1017/ice.2017.221}.
+55. Ahmad SM, Blanco N, Dewart CM, Dobosz A, Malani AN. 2017. Laxative
+use in the setting of positive testing for \emph{Clostridium difficile}
+infection. Infection Control \& Hospital Epidemiology 38:1513--1515.
 
 \leavevmode\hypertarget{ref-Bilinskaya2018}{}%
-56. \textbf{Bilinskaya A}, \textbf{Goodlet KJ}, \textbf{Nailor MD}.
-2018. Evaluation of a best practice alert to reduce unnecessary
-\emph{Clostridium difficile} testing following receipt of a laxative.
-Diagnostic Microbiology and Infectious Disease \textbf{92}:50--55.
-doi:\href{https://doi.org/10.1016/j.diagmicrobio.2018.04.009}{10.1016/j.diagmicrobio.2018.04.009}.
+56. Bilinskaya A, Goodlet KJ, Nailor MD. 2018. Evaluation of a best
+practice alert to reduce unnecessary \emph{Clostridium difficile}
+testing following receipt of a laxative. Diagnostic Microbiology and
+Infectious Disease 92:50--55.
 
 \leavevmode\hypertarget{ref-Cook2020}{}%
-57. \textbf{Cook PP}, \textbf{Nichols S}, \textbf{Coogan M},
-\textbf{Opera J}, \textbf{DeHart M}. 2020. Reduction in testing and
-change in testing algorithm associated with decrease in number of
-nosocomial \emph{Clostridioides} (\emph{Clostridium}) \emph{difficile}
-infections. American Journal of Infection Control
-\textbf{48}:1019--1022.
-doi:\href{https://doi.org/10.1016/j.ajic.2019.12.028}{10.1016/j.ajic.2019.12.028}.
+57. Cook PP, Nichols S, Coogan M, Opera J, DeHart M. 2020. Reduction in
+testing and change in testing algorithm associated with decrease in
+number of nosocomial \emph{Clostridioides} (\emph{Clostridium})
+\emph{difficile} infections. American Journal of Infection Control
+48:1019--1022.
 
 \leavevmode\hypertarget{ref-Appaneal2019}{}%
-58. \textbf{Appaneal HJ}, \textbf{Caffrey AR}, \textbf{Beganovic M},
-\textbf{Avramovic S}, \textbf{LaPlante KL}. 2019. Predictors of
-\emph{Clostridioides difficile} recurrence across a national cohort of
-veterans in outpatient, acute, and long-term care settings. American
-Journal of Health-System Pharmacy \textbf{76}:581--590.
-doi:\href{https://doi.org/10.1093/ajhp/zxz032}{10.1093/ajhp/zxz032}.
+58. Appaneal HJ, Caffrey AR, Beganovic M, Avramovic S, LaPlante KL.
+2019. Predictors of \emph{Clostridioides difficile} recurrence across a
+national cohort of veterans in outpatient, acute, and long-term care
+settings. American Journal of Health-System Pharmacy 76:581--590.
 
 \leavevmode\hypertarget{ref-Dubberke2011}{}%
-59. \textbf{Dubberke ER}, \textbf{Yan Y}, \textbf{Reske KA},
-\textbf{Butler AM}, \textbf{Doherty J}, \textbf{Pham V}, \textbf{Fraser
-VJ}. 2011. Development and validation of a \emph{Clostridium difficile}
+59. Dubberke ER, Yan Y, Reske KA, Butler AM, Doherty J, Pham V, Fraser
+VJ. 2011. Development and validation of a \emph{Clostridium difficile}
 infection risk prediction model. Infection Control \& Hospital
-Epidemiology \textbf{32}:360--366.
-doi:\href{https://doi.org/10.1086/658944}{10.1086/658944}.
+Epidemiology 32:360--366.
 
 \leavevmode\hypertarget{ref-Obritsch2010}{}%
-60. \textbf{Obritsch MD}, \textbf{Stroup JS}, \textbf{Carnahan RM},
-\textbf{Scheck DN}. 2010. \emph{Clostridium difficile}-associated
-diarrhea in a tertiary care medical center. Baylor University Medical
-Center Proceedings \textbf{23}:363--367.
-doi:\href{https://doi.org/10.1080/08998280.2010.11928654}{10.1080/08998280.2010.11928654}.
+60. Obritsch MD, Stroup JS, Carnahan RM, Scheck DN. 2010.
+\emph{Clostridium difficile}-associated diarrhea in a tertiary care
+medical center. Baylor University Medical Center Proceedings
+23:363--367.
 
 \leavevmode\hypertarget{ref-White2019}{}%
-61. \textbf{White NC}, \textbf{Mendo-Lopez R}, \textbf{Papamichael K},
-\textbf{Cuddemi CA}, \textbf{Barrett C}, \textbf{Daugherty K},
-\textbf{Pollock N}, \textbf{Kelly CP}, \textbf{Alonso CD}. 2019.
-Laxative use does not preclude diagnosis or reduce disease severity in
-\emph{Clostridioides difficile} infection. Clinical Infectious Diseases
-\textbf{71}:1472--1478.
-doi:\href{https://doi.org/10.1093/cid/ciz978}{10.1093/cid/ciz978}.
+61. White NC, Mendo-Lopez R, Papamichael K, Cuddemi CA, Barrett C,
+Daugherty K, Pollock N, Kelly CP, Alonso CD. 2019. Laxative use does not
+preclude diagnosis or reduce disease severity in \emph{Clostridioides
+difficile} infection. Clinical Infectious Diseases 71:1472--1478.
 
 \leavevmode\hypertarget{ref-Poirier2019}{}%
-62. \textbf{Poirier D}, \textbf{Gervais P}, \textbf{Fuchs M},
-\textbf{Roussy J-F}, \textbf{Paquet-Bolduc B}, \textbf{Trottier S},
-\textbf{Longtin J}, \textbf{Loo VG}, \textbf{Longtin Y}. 2019.
-Predictors of \emph{Clostridioides difficile} infection among
-asymptomatic, colonized patients: A retrospective cohort study. Clinical
-Infectious Diseases \textbf{70}:2103--2210.
-doi:\href{https://doi.org/10.1093/cid/ciz626}{10.1093/cid/ciz626}.
+62. Poirier D, Gervais P, Fuchs M, Roussy J-F, Paquet-Bolduc B, Trottier
+S, Longtin J, Loo VG, Longtin Y. 2019. Predictors of
+\emph{Clostridioides difficile} infection among asymptomatic, colonized
+patients: A retrospective cohort study. Clinical Infectious Diseases
+70:2103--2210.
 
 \leavevmode\hypertarget{ref-Huang2014}{}%
-63. \textbf{Huang H}, \textbf{Wu S}, \textbf{Chen R}, \textbf{Xu S},
-\textbf{Fang H}, \textbf{Weintraub A}, \textbf{Nord CE}. 2014. Risk
-factors of \emph{Clostridium difficile} infections among patients in a
-university hospital in shanghai, china. Anaerobe \textbf{30}:65--69.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2014.08.015}{10.1016/j.anaerobe.2014.08.015}.
+63. Huang H, Wu S, Chen R, Xu S, Fang H, Weintraub A, Nord CE. 2014.
+Risk factors of \emph{Clostridium difficile} infections among patients
+in a university hospital in shanghai, china. Anaerobe 30:65--69.
 
 \leavevmode\hypertarget{ref-Gorkiewicz2013}{}%
-64. \textbf{Gorkiewicz G}, \textbf{Thallinger GG}, \textbf{Trajanoski
-S}, \textbf{Lackner S}, \textbf{Stocker G}, \textbf{Hinterleitner T},
-\textbf{Gülly C}, \textbf{Högenauer C}. 2013. Alterations in the colonic
-microbiota in response to osmotic diarrhea. PLoS ONE \textbf{8}:e55817.
-doi:\href{https://doi.org/10.1371/journal.pone.0055817}{10.1371/journal.pone.0055817}.
+64. Gorkiewicz G, Thallinger GG, Trajanoski S, Lackner S, Stocker G,
+Hinterleitner T, Gülly C, Högenauer C. 2013. Alterations in the colonic
+microbiota in response to osmotic diarrhea. PLoS ONE 8:e55817.
 
 \leavevmode\hypertarget{ref-Shobar2016}{}%
-65. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
-\textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
-\textbf{Losurdo J}, \textbf{Engen PA}, \textbf{Sun Y}, \textbf{Koenig
-L}, \textbf{Mutlu EA}. 2016. The effects of bowel preparation on
-microbiota-related metrics differ in health and in inflammatory bowel
-disease and for the mucosal and luminal microbiota compartments.
-Clinical and Translational Gastroenterology \textbf{7}:e143.
-doi:\href{https://doi.org/10.1038/ctg.2015.54}{10.1038/ctg.2015.54}.
+65. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
+JE, Losurdo J, Engen PA, Sun Y, Koenig L, Mutlu EA. 2016. The effects of
+bowel preparation on microbiota-related metrics differ in health and in
+inflammatory bowel disease and for the mucosal and luminal microbiota
+compartments. Clinical and Translational Gastroenterology 7:e143.
 
 \leavevmode\hypertarget{ref-Drago2016}{}%
-66. \textbf{Drago L}, \textbf{Toscano M}, \textbf{Grandi RD},
-\textbf{Casini V}, \textbf{Pace F}. 2016. Persisting changes of
-intestinal microbiota after bowel lavage and colonoscopy. European
-Journal of Gastroenterology \& Hepatology \textbf{28}:532--537.
-doi:\href{https://doi.org/10.1097/meg.0000000000000581}{10.1097/meg.0000000000000581}.
+66. Drago L, Toscano M, Grandi RD, Casini V, Pace F. 2016. Persisting
+changes of intestinal microbiota after bowel lavage and colonoscopy.
+European Journal of Gastroenterology \& Hepatology 28:532--537.
 
 \leavevmode\hypertarget{ref-Jalanka2014}{}%
-67. \textbf{Jalanka J}, \textbf{Salonen A}, \textbf{Salojärvi J},
-\textbf{Ritari J}, \textbf{Immonen O}, \textbf{Marciani L},
-\textbf{Gowland P}, \textbf{Hoad C}, \textbf{Garsed K}, \textbf{Lam C},
-\textbf{Palva A}, \textbf{Spiller RC}, \textbf{Vos WM de}. 2014. Effects
-of bowel cleansing on the intestinal microbiota. Gut
-\textbf{64}:1562--1568.
-doi:\href{https://doi.org/10.1136/gutjnl-2014-307240}{10.1136/gutjnl-2014-307240}.
+67. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
+Gowland P, Hoad C, Garsed K, Lam C, Palva A, Spiller RC, Vos WM de.
+2014. Effects of bowel cleansing on the intestinal microbiota. Gut
+64:1562--1568.
 
 \leavevmode\hypertarget{ref-Kampouri2021}{}%
-68. \textbf{Kampouri E}, \textbf{Croxatto A}, \textbf{Prod'hom G},
-\textbf{Guery B}. 2021. \emph{Clostridioides difficile} infection, still
-a long way to go. Journal of Clinical Medicine \textbf{10}:389.
-doi:\href{https://doi.org/10.3390/jcm10030389}{10.3390/jcm10030389}.
+68. Kampouri E, Croxatto A, Prod'hom G, Guery B. 2021.
+\emph{Clostridioides difficile} infection, still a long way to go.
+Journal of Clinical Medicine 10:389.
 
 \leavevmode\hypertarget{ref-Kelly2021}{}%
-69. \textbf{Kelly CR}, \textbf{Yen EF}, \textbf{Grinspan AM},
-\textbf{Kahn SA}, \textbf{Atreja A}, \textbf{Lewis JD}, \textbf{Moore
-TA}, \textbf{Rubin DT}, \textbf{Kim AM}, \textbf{Serra S},
-\textbf{Nersesova Y}, \textbf{Fredell L}, \textbf{Hunsicker D},
-\textbf{McDonald D}, \textbf{Knight R}, \textbf{Allegretti JR},
-\textbf{Pekow J}, \textbf{Absah I}, \textbf{Hsu R}, \textbf{Vincent J},
-\textbf{Khanna S}, \textbf{Tangen L}, \textbf{Crawford CV},
-\textbf{Mattar MC}, \textbf{Chen LA}, \textbf{Fischer M},
-\textbf{Arsenescu RI}, \textbf{Feuerstadt P}, \textbf{Goldstein J},
-\textbf{Kerman D}, \textbf{Ehrlich AC}, \textbf{Wu GD}, \textbf{Laine
-L}. 2021. Fecal microbiota transplantation is highly effective in
+69. Kelly CR, Yen EF, Grinspan AM, Kahn SA, Atreja A, Lewis JD, Moore
+TA, Rubin DT, Kim AM, Serra S, Nersesova Y, Fredell L, Hunsicker D,
+McDonald D, Knight R, Allegretti JR, Pekow J, Absah I, Hsu R, Vincent J,
+Khanna S, Tangen L, Crawford CV, Mattar MC, Chen LA, Fischer M,
+Arsenescu RI, Feuerstadt P, Goldstein J, Kerman D, Ehrlich AC, Wu GD,
+Laine L. 2021. Fecal microbiota transplantation is highly effective in
 real-world practice: Initial results from the FMT national registry.
-Gastroenterology \textbf{160}:183--192.e3.
-doi:\href{https://doi.org/10.1053/j.gastro.2020.09.038}{10.1053/j.gastro.2020.09.038}.
+Gastroenterology 160:183--192.e3.
 
 \leavevmode\hypertarget{ref-Dieterle2018}{}%
-70. \textbf{Dieterle MG}, \textbf{Rao K}, \textbf{Young VB}. 2018. Novel
-therapies and preventative strategies for primary and recurrent
-\emph{Clostridiuml difficile} infections. Annals of the New York Academy
-of Sciences \textbf{1435}:110--138.
-doi:\href{https://doi.org/10.1111/nyas.13958}{10.1111/nyas.13958}.
+70. Dieterle MG, Rao K, Young VB. 2018. Novel therapies and preventative
+strategies for primary and recurrent \emph{Clostridiuml difficile}
+infections. Annals of the New York Academy of Sciences 1435:110--138.
 
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-71. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq illumina sequencing platform.
-Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+71. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+Development of a dual-index sequencing strategy and curation pipeline
+for analyzing amplicon sequence data on the MiSeq illumina sequencing
+platform. Applied and Environmental Microbiology 79:5112--5120.
 
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-72. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-source, platform-independent,
-community-supported software for describing and comparing microbial
-communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+72. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
+Thallinger GG, Horn DJV, Weber CF. 2009. Introducing mothur:
+Open-source, platform-independent, community-supported software for
+describing and comparing microbial communities. Applied and
+Environmental Microbiology 75:7537--7541.
 
 \leavevmode\hypertarget{ref-Quast2012}{}%
-73. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
-\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
-J}, \textbf{Glöckner FO}. 2012. The SILVA ribosomal RNA gene database
-project: Improved data processing and web-based tools. Nucleic Acids
-Research \textbf{41}:D590--D596.
-doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.
+73. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
+J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
+Improved data processing and web-based tools. Nucleic Acids Research
+41:D590--D596.
 
 \leavevmode\hypertarget{ref-Cole2013}{}%
-74. \textbf{Cole JR}, \textbf{Wang Q}, \textbf{Fish JA}, \textbf{Chai
-B}, \textbf{McGarrell DM}, \textbf{Sun Y}, \textbf{Brown CT},
-\textbf{Porras-Alfaro A}, \textbf{Kuske CR}, \textbf{Tiedje JM}. 2013.
-Ribosomal database project: Data and tools for high throughput rRNA
-analysis. Nucleic Acids Research \textbf{42}:D633--D642.
-doi:\href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
+74. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
+Porras-Alfaro A, Kuske CR, Tiedje JM. 2013. Ribosomal database project:
+Data and tools for high throughput rRNA analysis. Nucleic Acids Research
+42:D633--D642.
 
 \leavevmode\hypertarget{ref-Vegan2018}{}%
-75. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
-\textbf{Kindt R}, \textbf{Legendre P}, \textbf{McGlinn D},
-\textbf{Minchin PR}, \textbf{O'Hara RB}, \textbf{Simpson GL},
-\textbf{Solymos P}, \textbf{Stevens MHH}, \textbf{Szoecs E},
-\textbf{Wagner H}. 2018. Vegan: Community ecology package.
+75. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
+Wagner H. 2018. Vegan: Community ecology package.
 
 \leavevmode\hypertarget{ref-mikropml}{}%
-76. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool KL},
-\textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020.
-mikRopML: User-friendly r package for robust machine learning pipelines.
+76. Topçuoğlu B, Lapp Z, Sovacool KL, Snitkin E, Wiens J, Schloss PD.
+2020. mikRopML: User-friendly r package for robust machine learning
+pipelines.
 
 \leavevmode\hypertarget{ref-Topcuoglu2021}{}%
-77. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
-\textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
-Mikropml: User-friendly r package for supervised machine learning
-pipelines. Journal of Open Source Software \textbf{6}:3073.
-doi:\href{https://doi.org/10.21105/joss.03073}{10.21105/joss.03073}.
+77. Topçuoğlu B, Lapp Z, Sovacool K, Snitkin E, Wiens J, Schloss P.
+2021. Mikropml: User-friendly r package for supervised machine learning
+pipelines. Journal of Open Source Software 6:3073.
 
 \leavevmode\hypertarget{ref-Topcuoglu2020}{}%
-78. \textbf{Topçuoğlu BD}, \textbf{Lesniak NA}, \textbf{Ruffin MT},
-\textbf{Wiens J}, \textbf{Schloss PD}. 2020. A framework for effective
-application of machine learning to microbiome-based classification
-problems. mBio \textbf{11}.
-doi:\href{https://doi.org/10.1128/mbio.00434-20}{10.1128/mbio.00434-20}.
+78. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
+framework for effective application of machine learning to
+microbiome-based classification problems. mBio 11.
 
 \leavevmode\hypertarget{ref-r_citation_2020}{}%
-79. \textbf{R Core Team}. 2020. R: A language and environment for
-statistical computing. R Foundation for Statistical Computing, Vienna,
-Austria.
+79. R Core Team. 2020. R: A language and environment for statistical
+computing. R Foundation for Statistical Computing, Vienna, Austria.
 
 \leavevmode\hypertarget{ref-Tidyverse2019}{}%
-80. \textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan J},
-\textbf{Chang W}, \textbf{McGowan LD}, \textbf{François R},
-\textbf{Grolemund G}, \textbf{Hayes A}, \textbf{Henry L}, \textbf{Hester
-J}, \textbf{Kuhn M}, \textbf{Pedersen TL}, \textbf{Miller E},
-\textbf{Bache SM}, \textbf{Müller K}, \textbf{Ooms J}, \textbf{Robinson
-D}, \textbf{Seidel DP}, \textbf{Spinu V}, \textbf{Takahashi K},
-\textbf{Vaughan D}, \textbf{Wilke C}, \textbf{Woo K}, \textbf{Yutani H}.
-2019. Welcome to the tidyverse. Journal of Open Source Software
-\textbf{4}:1686.
-doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}.
+80. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
+Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
+Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K,
+Vaughan D, Wilke C, Woo K, Yutani H. 2019. Welcome to the tidyverse.
+Journal of Open Source Software 4:1686.
 

### 2021-06-17 submission/manuscript.Rmd
Additional edits

@@
 However, there are similarities between the bacterial communities from people with diarrhea or *C.
-difficile* infections (CDI) including lower diversity compared to communities from healthy patients, which led us to hypothesize that diarrhea may be an indicator of *C.
-difficile* risk.
+difficile* infections (CDI) including lower diversity compared to communities from healthy patients.
+This observation led us to hypothesize that diarrhea may be an indicator of *C.
+difficile* susceptibility.
 We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C.
@@
 Our group has shown that when *C.
-difficile* negative samples from patients are separated into two groups based on stool consistency, there are similar microbiota features between samples from CDI patients and *C.
-difficile* negative patients with diarrhea compared to samples that were *C.
-difficile* negative with non-diarrheal stool consistency [@Schubert2014].
-The similar community features between CDI patients and patients with diarrhea included low alpha diversity and only 6 bacterial taxa had higher relative abundances in communities from CDI patients.
+difficile* negative samples from patients were separated into two groups based on stool consistency, there were similar microbiota features between samples from CDI patients and *C.
+difficile* negative patients with diarrhea compared to non-diarrheal samples that were *C.
+difficile* negative [@Schubert2014].
+The similar community features between CDI patients and patients with diarrhea included low alpha diversity and there were also only 6 bacterial taxa had higher relative abundances in communities from CDI patients.
 These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.
 
-depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019].
+Depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019].
 The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies.
@@
 
-Because time and treatment group influenced most of the variation between communities, we next profiled community diversity and composition.
+Because time and treatment group influenced most of the variation between communities, we next explored whether there were differences in community diversity and composition between treatment groups.
 We examined the alpha diversity dynamics by calculating the communities' Shannon diversity.
@@
 difficile* [@Engevik2020].
-*Bacteroides* persistent in the mucosal tissue might also have helped *Enterobacteriaceae* to colonize the region, as a synergy between mucus-degrading *B.
+*Bacteroides* persistence in the mucosal tissue might also have helped *Enterobacteriaceae* to colonize the region, as a synergy between mucus-degrading *B.
 fragilis* and *E.
@@
 However, recent studies demonstrated that broad spectrum antibiotics can also disrupt the host mucosal barrier in mice [@Kester2020; @Bergstrom2020].
-Future research is needed to tease out the interplay between medications that influence the mucus layer and different strains of *C.
+Further research is needed to tease out the interplay between medications that influence the mucus layer and different strains of *C.
 difficile* in the context of CDIs.
@@
 Although, it is unclear whether laxatives impact CDI susceptibility in human paitents, it is clear that laxatives also have a transient impact on the human microbiota [@Nagata2019; @Gorkiewicz2013; @Shobar2016; @Drago2016; @Jalanka2014].
-Further studies to examine the relationship between laxatives, *C.
+Additional studies to examine the relationship between laxatives, *C.
 difficile* colonization, and CDIs are warranted.
@@
 ***C.
-difficile* challenge model.** Mice were challenged with 25 microliters of *C.
-difficile* 630 spores at 10^5^ concentration, except for 1 experiment where the concentration was 10^3^ (Fig.
+difficile* challenge model.** Mice were challenged with 25 microliters containing 10^5^ *C.
+difficile* 630 spores, except for 1 experiment where the concentration was 10^3^ (Fig.
 5A).
@@
 For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C.
-For histopathology, cecum and colon tissues were placed into separate casettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.
+For histopathology, cecum and colon tissues were placed into separate cassettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.
 

### 2021-06-17 submission/manuscript.tex
Additional edits

@@
 or \emph{C. difficile} infections (CDI) including lower diversity
-compared to communities from healthy patients, which led us to
-hypothesize that diarrhea may be an indicator of \emph{C. difficile}
-risk. We explored how osmotic laxatives disrupt the microbiota's
-colonization resistance to \emph{C. difficile} by administering a
-laxative to mice either before or after \emph{C. difficile} challenge.
-Our findings suggest that osmotic laxatives disrupt colonization
-resistance to \emph{C. difficile}, and prevent clearance among mice
-already colonized with \emph{C. difficile}. Considering that most
-hospitals recommend not performing \emph{C. difficile} testing on
-patients taking laxatives and laxatives are used when administering
-fecal microbiota transplants via colonoscopy to patients with recurrent
-CDIs, further studies are needed to evaluate if laxatives impact
-microbiota colonization resistance in humans.
+compared to communities from healthy patients. This observation led us
+to hypothesize that diarrhea may be an indicator of \emph{C. difficile}
+susceptibility. We explored how osmotic laxatives disrupt the
+microbiota's colonization resistance to \emph{C. difficile} by
+administering a laxative to mice either before or after \emph{C.
+difficile} challenge. Our findings suggest that osmotic laxatives
+disrupt colonization resistance to \emph{C. difficile}, and prevent
+clearance among mice already colonized with \emph{C. difficile}.
+Considering that most hospitals recommend not performing \emph{C.
+difficile} testing on patients taking laxatives and laxatives are used
+when administering fecal microbiota transplants via colonoscopy to
+patients with recurrent CDIs, further studies are needed to evaluate if
+laxatives impact microbiota colonization resistance in humans.
 
@@
 serves as an approximation of intestinal motility (10). Our group has
-shown that when \emph{C. difficile} negative samples from patients are
-separated into two groups based on stool consistency, there are similar
+shown that when \emph{C. difficile} negative samples from patients were
+separated into two groups based on stool consistency, there were similar
 microbiota features between samples from CDI patients and \emph{C.
-difficile} negative patients with diarrhea compared to samples that were
-\emph{C. difficile} negative with non-diarrheal stool consistency (12).
-The similar community features between CDI patients and patients with
-diarrhea included low alpha diversity and only 6 bacterial taxa had
-higher relative abundances in communities from CDI patients. These
+difficile} negative patients with diarrhea compared to non-diarrheal
+samples that were \emph{C. difficile} negative (12). The similar
+community features between CDI patients and patients with diarrhea
+included low alpha diversity and there were also only 6 bacterial taxa
+had higher relative abundances in communities from CDI patients. These
 results led to the hypothesis that bacterial communities from patients
@@
 
-depending on the dose administered, osmotic laxatives can lead to
+Depending on the dose administered, osmotic laxatives can lead to
 diarrhea and temporarily disrupt the human intestinal microbiota (13).
@@
 Because time and treatment group influenced most of the variation
-between communities, we next profiled community diversity and
-composition. We examined the alpha diversity dynamics by calculating the
-communities' Shannon diversity. Although both clindamycin and PEG
-treatments decreased diversity, the Shannon diversity was lower in the
-groups of mice that received PEG treatment compared to those that
-received clindamycin alone through thirty days post-challenge (Fig. 2C;
-Data Set S1, sheet 2). We next identified the bacterial genera whose
-relative abundances shifted after PEG treatment by comparing the
-baseline samples of mice treated with only PEG to samples from the same
-mice one day post-PEG-treatment. We found 18 genera whose relative
-abundances were altered by PEG treatment (Data Set S1, sheet 3). The
-majority of the bacterial relative abundances decreased after the PEG
-treatment, but the relative abundance among members of the
-\emph{Enterobacteriaceae} and \emph{Bacteroides} increased. The increase
-in \emph{Bacteroides} relative abundance was unique to PEG treated mice,
-as the \emph{Bacteroides} relative abundance actually decreased in
-clindamycin treated mice (Fig. 2D). Finally, we identified the genera
-whose relative abundance differed across treatment groups over multiple
-time points. Of the 33 genera that were different between treatment
-groups, 24 genera were different over multiple time points (Fig. 2E,
-Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal
-microbiota that was maintained over time and was distinct from
-clindamycin treatment.
+between communities, we next explored whether there were differences in
+community diversity and composition between treatment groups. We
+examined the alpha diversity dynamics by calculating the communities'
+Shannon diversity. Although both clindamycin and PEG treatments
+decreased diversity, the Shannon diversity was lower in the groups of
+mice that received PEG treatment compared to those that received
+clindamycin alone through thirty days post-challenge (Fig. 2C; Data Set
+S1, sheet 2). We next identified the bacterial genera whose relative
+abundances shifted after PEG treatment by comparing the baseline samples
+of mice treated with only PEG to samples from the same mice one day
+post-PEG-treatment. We found 18 genera whose relative abundances were
+altered by PEG treatment (Data Set S1, sheet 3). The majority of the
+bacterial relative abundances decreased after the PEG treatment, but the
+relative abundance among members of the \emph{Enterobacteriaceae} and
+\emph{Bacteroides} increased. The increase in \emph{Bacteroides}
+relative abundance was unique to PEG treated mice, as the
+\emph{Bacteroides} relative abundance actually decreased in clindamycin
+treated mice (Fig. 2D). Finally, we identified the genera whose relative
+abundance differed across treatment groups over multiple time points. Of
+the 33 genera that were different between treatment groups, 24 genera
+were different over multiple time points (Fig. 2E, Data Set S1, sheet
+4). Thus, PEG had a significant impact on the fecal microbiota that was
+maintained over time and was distinct from clindamycin treatment.
 
@@
 glycans with glycosyl hydrolases that are absent in \emph{C. difficile}
-(37). \emph{Bacteroides} persistent in the mucosal tissue might also
+(37). \emph{Bacteroides} persistence in the mucosal tissue might also
 have helped \emph{Enterobacteriaceae} to colonize the region, as a
@@
 antibiotics can also disrupt the host mucosal barrier in mice (52, 53).
-Future research is needed to tease out the interplay between medications
-that influence the mucus layer and different strains of \emph{C.
-difficile} in the context of CDIs.
+Further research is needed to tease out the interplay between
+medications that influence the mucus layer and different strains of
+\emph{C. difficile} in the context of CDIs.
 
@@
 susceptibility in human paitents, it is clear that laxatives also have a
-transient impact on the human microbiota (13, 64--67). Further studies
-to examine the relationship between laxatives, \emph{C. difficile}
-colonization, and CDIs are warranted.
+transient impact on the human microbiota (13, 64--67). Additional
+studies to examine the relationship between laxatives, \emph{C.
+difficile} colonization, and CDIs are warranted.
 
@@
 \textbf{\emph{C. difficile} challenge model.} Mice were challenged with
-25 microliters of \emph{C. difficile} 630 spores at
-10\textsuperscript{5} concentration, except for 1 experiment where the
-concentration was 10\textsuperscript{3} (Fig. 5A). All mock challenged
-mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper
-pipette was used to administer the same challenge dose to mice via oral
-gavage. Mice were weighed daily throughout the experiment and stool was
-collected for quantifying \emph{C. difficile} CFU and 16S rRNA gene
-sequencing. There were two groups of mice that received either a PBS or
-fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal
-microbiota transplant was prepared with stool samples collected from the
-mice in the experiment prior to the start of any treatments. The stool
-samples were transferred to an anaerobic chamber and diluted 1:10 in
-reduced PBS and glycerol was added to make a 15\% glycerol solution. The
-solution was then aliquoted into tubes and stored at -80°C until the day
-of the gavage. An aliquot of both the FMT and PBS solutions were also
-set aside in the -80°C for 16S rRNA gene sequencing. The day of the
-gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute.
-The supernatant was then transferred to a separate tube to prevent the
+25 microliters containing 10\textsuperscript{5} \emph{C. difficile} 630
+spores, except for 1 experiment where the concentration was
+10\textsuperscript{3} (Fig. 5A). All mock challenged mice received 25 ul
+vehicle solution (Ultrapure water). A Dymax stepper pipette was used to
+administer the same challenge dose to mice via oral gavage. Mice were
+weighed daily throughout the experiment and stool was collected for
+quantifying \emph{C. difficile} CFU and 16S rRNA gene sequencing. There
+were two groups of mice that received either a PBS or fecal microbiota
+transplant (FMT) gavage post-PEG treatment. The fecal microbiota
+transplant was prepared with stool samples collected from the mice in
+the experiment prior to the start of any treatments. The stool samples
+were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS
+and glycerol was added to make a 15\% glycerol solution. The solution
+was then aliquoted into tubes and stored at -80°C until the day of the
+gavage. An aliquot of both the FMT and PBS solutions were also set aside
+in the -80°C for 16S rRNA gene sequencing. The day of the gavage,
+aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The
+supernatant was then transferred to a separate tube to prevent the
 gavage needle from clogging with debris during gavage. The PBS solution
@@
 froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
-and colon tissues were placed into separate casettes, fixed, and then
+and colon tissues were placed into separate cassettes, fixed, and then
 submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,

### 2021-06-27 code/figure_5.R
Add numbers to treatment groups

@@ -21,7 +21,7 @@ PEG_treatment_legend <- ggplot() +
 save_plot("results/figures/post_CDI_PEG_treatment_legend.png", PEG_treatment_legend, base_height = 1, base_width = 9)
  
 #Figure with Schematic, CFU over time, and shannon over time
-a <- ggdraw() + draw_image("results/pictures/post_CDI_PEG_schematic.png")
+a <- ggdraw() + draw_image("results/pictures/post_CDI_PEG_exp_schematic.png")
 b <- ggdraw() + draw_image("results/figures/post_CDI_PEG_cfu.png")
 c <- ggdraw() + draw_image("results/figures/post_CDI_PEG_shannon_stool.png")
 d <- ggdraw() + draw_image("results/figures/post_CDI_PEG_richness_overtime_stool.png")

### 2021-06-27 submission/manuscript.Rmd
Incorporate edits from coauthors

@@
 However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350 also alter the microbiota.
+Clinicians also hypothesize that PEG helps clear *C.
+difficile*.
 But whether PEG impacts CDI susceptibility and clearance is unclear.
@@
 Considering that most hospitals recommend not performing *C.
-difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.
+difficile* testing on patients taking laxatives and laxatives are prescribed prior to administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.
 
@@
 difficile* negative [@Schubert2014].
-The similar community features between CDI patients and patients with diarrhea included low alpha diversity and there were also only 6 bacterial taxa had higher relative abundances in communities from CDI patients.
+The similar community features between CDI patients and patients with diarrhea included low alpha diversity and only 6 bacterial taxa with higher relative abundances in communities from CDI patients.
 These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.
@@
 
-Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization resistance.
+Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT), where a healthy microbiota is administered to the patient to restore colonization resistance.
 For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure.
@@
 2A, Data Set S1, sheet 1).
-None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C.
+None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge, suggesting other community features besides recovery to baseline were responsible for the prolonged *C.
 difficile* colonization in PEG-treated mice (Fig.
@@
 For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period.
-One mouse died on the sixth day post-challenge and in the other *C.
+One mouse died on the sixth day post-challenge and in the other, *C.
 difficile* was present at a high density from the 4th day post-challenge onward (Fig.
@@
 
-Because there were differences in the mucosal microbiota including detectable *C.
+Because there were differences in the mucosal microbiota, including detectable *C.
 difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C.
@@
 The differences in *C.
-difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota.
+difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in the degree to which treatments disrupted the microbiota.
 Additionally, the intestinal communities of 5-day PEG treated mice did not regain colonization resistance after a 10-day recovery period.
@@
 A previous study demonstrated that a 5-day treatment of 10% PEG depleted acetate and butyrate and increased succinate compared to untreated mice [@Ferreyra2014].
-While we did not perform metabolomic analysis, we did see bacteria known to produce benefical metabolites were depleted in mice that cleared *C.
+While we did not perform metabolomic analysis, we did see bacteria known to produce beneficial metabolites were depleted in mice that cleared *C.
 difficile* compared to mice with prolonged colonization.
@@
 difficile* adherence *in vitro* [@Deng2018].
-In coculture experiments *B.
+In coculture experiments, *B.
 longum* decreased *C.
@@
 *Akkermansia* is also a mucin degrader with potentially beneficial or detrimental roles depending on context in other diseases [@Cirstea2018; @BorgesCanha2015].
-In our study the relative abundance of *Akkermansia* shifted over time between groups of mice that either cleared *C.
+In our study, the relative abundance of *Akkermansia* shifted over time between groups of mice that either cleared *C.
 difficile* or had prolonged colonization.

### 2021-06-27 submission/manuscript.tex
Incorporate edits from coauthors

@@
 non-antibiotic medications such as the ubiquitous osmotic laxative
-polyethylene glycol (PEG) 3350 also alter the microbiota. But whether
+polyethylene glycol (PEG) 3350 also alter the microbiota. Clinicians
+also hypothesize that PEG helps clear \emph{C. difficile}. But whether
 PEG impacts CDI susceptibility and clearance is unclear. To examine how
@@
 Considering that most hospitals recommend not performing \emph{C.
-difficile} testing on patients taking laxatives and laxatives are used
-when administering fecal microbiota transplants via colonoscopy to
-patients with recurrent CDIs, further studies are needed to evaluate if
-laxatives impact microbiota colonization resistance in humans.
+difficile} testing on patients taking laxatives and laxatives are
+prescribed prior to administering fecal microbiota transplants via
+colonoscopy to patients with recurrent CDIs, further studies are needed
+to evaluate if laxatives impact microbiota colonization resistance in
+humans.
 
@@
 community features between CDI patients and patients with diarrhea
-included low alpha diversity and there were also only 6 bacterial taxa
-had higher relative abundances in communities from CDI patients. These
-results led to the hypothesis that bacterial communities from patients
-experiencing diarrhea are susceptible to developing CDIs, regardless of
-how they developed diarrhea.
+included low alpha diversity and only 6 bacterial taxa with higher
+relative abundances in communities from CDI patients. These results led
+to the hypothesis that bacterial communities from patients experiencing
+diarrhea are susceptible to developing CDIs, regardless of how they
+developed diarrhea.
 
@@
 Beyond susceptibility, PEG is also relevant in the context of treating
-recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
+recurrent CDIs via fecal microbiota transplant (FMT), where a healthy
 microbiota is administered to the patient to restore colonization
@@
 2A, Data Set S1, sheet 1). None of the treatment groups recovered to
-their baseline community structure either 10 or 30 days post-challenge
+their baseline community structure either 10 or 30 days post-challenge,
 suggesting other community features besides recovery to baseline were
@@
 abundance of \emph{Enterobacteriaceae} throughout the post-challenge
-period. One mouse died on the sixth day post-challenge and in the other
+period. One mouse died on the sixth day post-challenge and in the other,
 \emph{C. difficile} was present at a high density from the 4th day
@@
 
-Because there were differences in the mucosal microbiota including
+Because there were differences in the mucosal microbiota, including
 detectable \emph{C. difficile} sequences in tissues from PEG-treated
@@
 in \emph{C. difficile} colonization dynamics between the 5- and 1-day
-PEG treated mice were associated with differences in how much the
+PEG treated mice were associated with differences in the degree to which
 treatments disrupted the microbiota. Additionally, the intestinal
@@
 succinate compared to untreated mice (15). While we did not perform
-metabolomic analysis, we did see bacteria known to produce benefical
+metabolomic analysis, we did see bacteria known to produce beneficial
 metabolites were depleted in mice that cleared \emph{C. difficile}
@@
 inhibited \emph{C. difficile} adherence \emph{in vitro} (40). In
-coculture experiments \emph{B. longum} decreased \emph{C. difficile}
+coculture experiments, \emph{B. longum} decreased \emph{C. difficile}
 biofilm formation while \emph{B. thetaiotamicron} enhanced biofilm
@@
 or detrimental roles depending on context in other diseases (43, 44). In
-our study the relative abundance of \emph{Akkermansia} shifted over time
-between groups of mice that either cleared \emph{C. difficile} or had
-prolonged colonization. In the stool it was initially increased in mice
-that cleared \emph{C. difficile}, but shifted after 5-days
+our study, the relative abundance of \emph{Akkermansia} shifted over
+time between groups of mice that either cleared \emph{C. difficile} or
+had prolonged colonization. In the stool it was initially increased in
+mice that cleared \emph{C. difficile}, but shifted after 5-days
 post-challenge so that it was increased in mice that had prolonged

### 2021-08-30 submission/response_to_reviewers.md
Add edits and response to reviewers for revision

---
output:
  pdf_document: default
  html_document: default
header-includes:
    - \usepackage{setspace}\doublespacing  
---
### Reviewer #1 (Comments for the Author):

**In this manuscript, Tomkovich et al. examined experiments to show how PEG impacts mice susceptibility to C. difficile. They found PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remained colonized for up to 30 days post-challenge. In contrast, 1-day PEG treated mice were transiently colonized, clearing C. difficile within 7 days post-challenge. They concluded that osmotic laxatives disrupt colonization resistance to C. difficile, and prevent clearance among mice already colonized with C. difficile.**

**The study is very interesting and the data is strong to support their findings. I only have several minor comments following:**

**Line 47-49 Please clarify and make it more clear to led to your hypothesis.**

> We have revised to:
"These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea. For example, laxatives may disrupt colonization resistance to *C. difficile*."

**Line 720 Please clear Asterisks mean. Authors noticed it as the difference between groups, it's not clear which two groups (inter-group at the same time point or same group compared to day 0) were compared and showed a significant difference, please clear it for Fig. 1B and 1C.**

> We have rephrased to the following:
"Asterisks indicate the specific time points on the x-axis where the weight change or CFU/g was significantly different (*P* < 0.05) between the 3-4 groups indicated on the plot by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points."

**Did the author detect PEG concentration from stools along with experiments? Or any references can be referred to PEG metabolization and residence time in the gut?**

> We did not measure PEG concentrations as previous studies report PEG is nonabsorbable and not metabolized by bacteria (Hammer et al. JCI 1989; Di Palma et al. The American Journal of Gastroenterology 2002).

**Line 148 Did author detect the toxin concentration from fecal samples? It's interesting that more C. difficile CFU was detected and high weight loss in PEG treated group, while no significant difference in the gut.**

> We did not measure toxin concentration because we observed no histological differences between PEG and clindamycin treated mice. We agree it is an interesting observation, but suspect this may be partially due to the strain we used (*C. difficile* 630), and bring up this possibility in the discussion (lines 315-320).

**Line 176: Is it possible to include the histopathology figures in Fig. 3D?**

> *C. difficile* 630 typically causes mild histopathology and pictures are included in Theriot et al. Gut Microbes 2011 which is referred to in the methods section (lines 441-442). Because of figure size constraints and since histopathology was mild for all groups, we decided not to include histopatholgy pictures.

**Fig. 3D Label symbols of Clind. Without infection and Clind, After 5-day, 5-day PEG without infection and 5-day PEG are the same. Please correct.**

> The open and closed circles differentiate mice that were infected with *C. difficile* versus mice that were not. However, there were 2 sets of open circles that were the same color for the After 5-day PEG (Day 0) and 5-day PEG without infection groups, so we have added an additional shape to differentiate the Day 0 group in the Fig. 3D legend. Additionally, we have combined the scales to make the differences between groups more clear (Fig. 3D-E).

**Line 276-277 Could you notice which result Figure (Maybe Fig. 3C) you are discussed to help the reader to follow?**

> We have added a reference to Fig. 7B in the sentence.

**Line 284-285 Spores persistence in cell should be also noticed in this part as it can also contribute to C. difficile persistence. Ref: Castro-Córdova P, Mora-Uribe P, Reyes-Ramírez R, et al. Nature communications, 2021.**

> We have modified the sentence to the following:
"One possible explanation for the prolonged *C. difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment either within host cells (Castro-Córdova et al. Nature Communications 2021) or together with other bacteria."

**Line 256-358 So many parts for discussion, maybe can concentrate and minimize paragraphs.**

> The manuscript is within mSphere's 5000 word limit (excluding the Materials and Methods, References, and figure legends) for research articles. Given all the available literature on the microbiota, laxatives, and *C. difficile*, we wanted place our findings in context, which resulted in a longer discussion.

**Line 390 Stool was collected as fresh one or obtained from cages? How did you process your stool samples? Plating every day for CFU check or stock -80 {degree sign}C and did CFU check at the end of the experiment?**

> We have added the following sentences to clarify our stool processing methods:
"Fresh stool samples were collected from each mouse and split into two separate tubes. One tube was transferred to an anaerobic chamber the same day the sample was collected to quantify *C. difficile*, while the other tube was snap frozen in liquid nitrogen and stored in the -80&deg;C for 16S rRNA sequencing."

**Specific comments:**
**Over-referred in this manuscript (80 references were included).**

> There is no limit on the number of references for mSphere research articles.

**Typos**
**Line 4 Alter-alters**
**Line 59 Challenge-the challenge**
**Line 181 Signifant---Significant**
**Line 267 received**
**Line 312 Colonization**

> We thank the reviewer for identifying these typos and have corrected all except for line 4. Alter was used because we are stating multiple medications alter the microbiota and just giving PEG as a specific example.

### Reviewer #2 (Comments for the Author):

**As expected from this group, this is a well conducted study that is timely relevant for CDI. In this work authors address an intriguing question related with how laxative treatment affects the susceptibility of the host to C. difficile infections. Authors provide a comprehensive and very detailed overview of the impact of the osmotic laxative, PEG, to shifts in fecal and mucosal microbiota, and how PEG impacted the susceptibility of mice to infection and subsequent clearance. This work is of great public relevance and complements several other studies published on this topic.**

**I only have minor comments for the authors.**

**Specific comments:**
**Line 324: not clear please revise**

> We have revised the sentence to the following:
"The extent to which laxatives disrupt *C. difficile* colonization resistance in human patients is unclear based on the current literature."

**Figure 5: is there any reason why FMT was not used alone as a control in this experiment? There is data on how FMT impacts the microbiota composition in Figure 6, but not in the C. difficile infection experiment. Could the authors clarify this?**

> We did not use FMT alone in this experiment because clindamycin treated mice typically clear *C. difficile* within 10 days without any additional treatments such as an FMT.

**Figure 8: Please revise the narrative of the figure legend for consistencies in language.**

> We have modified the legend to:
"The gut microbiota of our lab's C57Bl/6 mice is resistant to *C. difficle* but treatment with either clindamycin or the osmotic laxative, PEG 3350, renders the mice susceptible to *C. difficile* colonization. Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-challenge. However, recovery of colonization resistance was delayed for mice that received either a 5-day PEG pre-treatment or a 1-day PEG post-*C. difficile* challenge treatment. We found increased relative abundances of *Porphyromonadaceae* and *Lachnospiraceae* were associated with *C. difficle* clearance, while increased relative abundances of *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C. difficle* colonization."

### 2021-08-30 submission/manuscript.Rmd
Add edits and response to reviewers for revision

@@
 These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.
+For example, laxatives may disrupt colonization resistance to *C.
+difficile*.
 
@@
 Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C.
-difficile* after challenge is unclear.
+difficile* after the challenge is unclear.
 
@@
 1C).
-Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig.
+Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a significant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig.
 3E).
@@
 difficile* challenge.
-In opposition to the hypothesis suggested by the literature, we found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment.
+In opposition to the hypothesis suggested by the literature, we found that PEG treatment prolonged colonization relative to mice that only received clindamycin treatment.
 We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C.
@@
 While we did not perform metabolomic analysis, we did see bacteria known to produce beneficial metabolites were depleted in mice that cleared *C.
-difficile* compared to mice with prolonged colonization.
+difficile* compared to mice with prolonged colonization (Fig.
+7B).
 For example, *Oscillibacter valericigenes* can produce the SCFA valerate [@Iino2007], and separate studies demonstrated valerate is depleted after clindamycin treatment and inhibited *C.
@@
 One possible explanation for the prolonged *C.
-difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment.
+difficile* colonization in 5-day PEG treated mice, might be due to the bacteria's persistence in the mucosal compartment either within host cells [@CastroCrdova2021] or together with other bacteria.
 In fact, it has been hypothesized that *C.
@@
 difficile* interacts with the mucosal microbiota may lead to insights into CDIs, asymptomatic *C.
-difficile* carriage, and colonizatiion resistance.
+difficile* carriage, and colonization resistance.
 
@@
 
-It is more difficult to interpret what are findings mean in the context of *C.
-difficile* colonization resistance in human patients.
+The extent to which laxatives disrupt *C.
+difficile* colonization resistance in human patients is unclear based on the current literature.
 The main difficulty being that most hospitals recommend not performing *C.
@@
 difficile* CFU and 16S rRNA gene sequencing.
+Fresh stool samples were collected from each mouse and split into two separate tubes.
+One tube was transferred to an anaerobic chamber the same day the sample was collected to quantify *C.
+difficile*, while the other tube was snap frozen in liquid nitrogen and stored in the -80&deg;C for 16S rRNA sequencing.
 There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment.
@@
 ```
-
-\includegraphics{figure_1.pdf}
 
@@
 For B-C, lines represent the median for each treatment group and circles represent samples from individual mice.
-Asterisks indicate time points where the weight change or CFU/g was significantly different (*P* < 0.05) between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
+Asterisks indicate the specific time points on the x-axis where the weight change or CFU/g was significantly different (*P* < 0.05) between the 3-4 groups indicated on the plot by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple time points.
 The data presented are from a total of `r five_day_total_exp` separate experiments.
@@
 
-\includegraphics{figure_2.pdf}
 **Figure 2.
@@
 
-\includegraphics{figure_3.pdf}
 **Figure 3.
@@
 ```
-\includegraphics{figure_4.pdf}
 **Figure 4.
@@
 
-\includegraphics{figure_5.pdf}
 **Figure 5.
@@
 
-\includegraphics{figure_6.pdf}
 **Figure 6.
@@
 
-\includegraphics{figure_7.pdf}
 **Figure 7.
@@
 
-\includegraphics{figure_8.pdf}
-
 **Figure 8.
-Schematic summarizing findings.** The gut microbiota of our C57Bl/6 mice is resistant to *C.
-difficle* but treatment with either the antibiotic, clindamycin, or the osmotic laxative, PEG 3350, renders the mice susceptible to *C.
+Schematic summarizing findings.** The gut microbiota of our lab's C57Bl/6 mice is resistant to *C.
+difficle* but treatment with either clindamycin or the osmotic laxative, PEG 3350, renders the mice susceptible to *C.
 difficile* colonization.
 Recovery of colonization resistance in clindamycin-treated mice is relatively straightforward and the mice clear *C.difficile* within 10 days post-challenge.
-However, for mice that received either a 5-day PEG treatment prior to *C.
-difficile* challenge or a 1-day PEG treatment post-challenge recovery of colonization resistance was delayed because most mice were still colonized with *C.
-difficile* 10 days post-challenge.
-We found increased relative abundances of *Porphyromonadaceae* and *Lachnospiraceae* were associated with recovery of colonization resistance, while increased relative abundances of *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C.
+However, recovery of colonization resistance was delayed for mice that received either a 5-day PEG pre-treatment or a 1-day PEG post-*C.
+difficile* challenge treatment.
+We found increased relative abundances of *Porphyromonadaceae* and *Lachnospiraceae* were associated with *C.
+difficle* clearance, while increased relative abundances of *Enterobacteriaceae* and *Bacteroides* were associated with prolonged *C.
 difficle* colonization.
@@
 
-\includegraphics{figure_S1.pdf}
 **Figure S1.
@@
 
-\includegraphics{figure_S2.pdf}
 **Figure S2.
@@
 
-\includegraphics{figure_S3.pdf}
 **Figure S3.

### 2021-08-30 submission/manuscript.tex
Add edits and response to reviewers for revision

@@
 diarrhea are susceptible to developing CDIs, regardless of how they
-developed diarrhea.
+developed diarrhea. For example, laxatives may disrupt colonization
+resistance to \emph{C. difficile}.
 
@@
 more severe CDIs in mice and how long mice remain colonized with
-\emph{C. difficile} after challenge is unclear.
+\emph{C. difficile} after the challenge is unclear.
 
@@
 difference in the histopathology score of the colon between PEG and
-clindamycin-treated mice, there was not a signifant difference in the
+clindamycin-treated mice, there was not a significant difference in the
 cecum and the overall score was relatively low (1.5 to 2.5 out of 12,
@@
 literature, we found that PEG treatment prolonged colonization relative
-to mice that only recieved clindamycin treatment. We identified patterns
+to mice that only received clindamycin treatment. We identified patterns
 in the relative abundances of \emph{Bacteroides},
@@
 metabolites were depleted in mice that cleared \emph{C. difficile}
-compared to mice with prolonged colonization. For example,
+compared to mice with prolonged colonization (Fig. 7B). For example,
 \emph{Oscillibacter valericigenes} can produce the SCFA valerate (30),
@@
 colonization in 5-day PEG treated mice, might be due to the bacteria's
-persistence in the mucosal compartment. In fact, it has been
-hypothesized that \emph{C. difficile} biofilms may serve as reservoirs
-for recurrent infections (35) and \emph{C. difficile} biofilms in the
-mucus layer were recently identified in patients as aggregates with
-\emph{Fusobacterium nucleatum} (36). There was an interesting pattern of
-increased \emph{Enterobacteriaceae}, \emph{Bacteroides}, and \emph{C.
-difficile} in both the stool and mucosal communities of PEG-treated mice
-suggesting a potential synergy. \emph{Bacteroides} has the potential to
-degrade mucus and the osmotic laxative may have allowed
-\emph{Bacteroides} to colonize the mucosal niche by degrading mucin
-glycans with glycosyl hydrolases that are absent in \emph{C. difficile}
-(37). \emph{Bacteroides} persistence in the mucosal tissue might also
-have helped \emph{Enterobacteriaceae} to colonize the region, as a
-synergy between mucus-degrading \emph{B. fragilis} and \emph{E. coli}
-has previously been described (38). A separate study demonstrated
-\emph{C. difficile} was present in the outer mucus layer and associated
-with \emph{Enterobacteriaceae} and \emph{Bacteroidaceae} using
-fluorescent in situ hybridization (FISH) staining (39). However,
-protective roles for \emph{Bacteroides} have also been demonstrated. For
-example, \emph{B. fragilis} prevented CDI morbidity in a mouse model and
-inhibited \emph{C. difficile} adherence \emph{in vitro} (40). In
-coculture experiments, \emph{B. longum} decreased \emph{C. difficile}
-biofilm formation while \emph{B. thetaiotamicron} enhanced biofilm
-formation (41) and \emph{B. dorei} reduced \emph{C. difficile} growth in
-a 9-species community \emph{in vitro} (42). Therefore, whether
+persistence in the mucosal compartment either within host cells (35) or
+together with other bacteria. In fact, it has been hypothesized that
+\emph{C. difficile} biofilms may serve as reservoirs for recurrent
+infections (36) and \emph{C. difficile} biofilms in the mucus layer were
+recently identified in patients as aggregates with \emph{Fusobacterium
+nucleatum} (37). There was an interesting pattern of increased
+\emph{Enterobacteriaceae}, \emph{Bacteroides}, and \emph{C. difficile}
+in both the stool and mucosal communities of PEG-treated mice suggesting
+a potential synergy. \emph{Bacteroides} has the potential to degrade
+mucus and the osmotic laxative may have allowed \emph{Bacteroides} to
+colonize the mucosal niche by degrading mucin glycans with glycosyl
+hydrolases that are absent in \emph{C. difficile} (38).
+\emph{Bacteroides} persistence in the mucosal tissue might also have
+helped \emph{Enterobacteriaceae} to colonize the region, as a synergy
+between mucus-degrading \emph{B. fragilis} and \emph{E. coli} has
+previously been described (39). A separate study demonstrated \emph{C.
+difficile} was present in the outer mucus layer and associated with
+\emph{Enterobacteriaceae} and \emph{Bacteroidaceae} using fluorescent in
+situ hybridization (FISH) staining (40). However, protective roles for
+\emph{Bacteroides} have also been demonstrated. For example, \emph{B.
+fragilis} prevented CDI morbidity in a mouse model and inhibited
+\emph{C. difficile} adherence \emph{in vitro} (41). In coculture
+experiments, \emph{B. longum} decreased \emph{C. difficile} biofilm
+formation while \emph{B. thetaiotamicron} enhanced biofilm formation
+(42) and \emph{B. dorei} reduced \emph{C. difficile} growth in a
+9-species community \emph{in vitro} (43). Therefore, whether
 \emph{Bacteroides} is detrimental or beneficial in the context of
@@
 \emph{Akkermansia} is also a mucin degrader with potentially beneficial
-or detrimental roles depending on context in other diseases (43, 44). In
+or detrimental roles depending on context in other diseases (44, 45). In
 our study, the relative abundance of \emph{Akkermansia} shifted over
@@
 colonization. In the context of CDIs, some studies suggest a protective
-role (45, 46), while others suggest a detrimental role because
+role (46, 47), while others suggest a detrimental role because
 \emph{Akkermansia} was positively correlated with \emph{C. difficile}
-(47--50). Because the relative abundance of \emph{Akkermansia} was
+(48--51). Because the relative abundance of \emph{Akkermansia} was
 dynamic in our study, it is unclear whether \emph{Akkermansia} helps
@@
 microbiota may lead to insights into CDIs, asymptomatic \emph{C.
-difficile} carriage, and colonizatiion resistance.
+difficile} carriage, and colonization resistance.
 
@@
 histopathology summary scores in mice compared to VPI 10463 despite both
-strains producing toxin in mice (51). Part of our hypothesis for why
+strains producing toxin in mice (52). Part of our hypothesis for why
 there could have been increased histopathology scores in PEG-treated
@@
 mice. However, recent studies demonstrated that broad spectrum
-antibiotics can also disrupt the host mucosal barrier in mice (52, 53).
+antibiotics can also disrupt the host mucosal barrier in mice (53, 54).
 Further research is needed to tease out the interplay between
@@
 
-It is more difficult to interpret what are findings mean in the context
-of \emph{C. difficile} colonization resistance in human patients. The
-main difficulty being that most hospitals recommend not performing
+The extent to which laxatives disrupt \emph{C. difficile} colonization
+resistance in human patients is unclear based on the current literature.
+The main difficulty being that most hospitals recommend not performing
 \emph{C. difficile} testing if the patient is currently taking a
@@
 Diseases Society of America and Society for Healthcare Epidemiology of
-America guidelines (54). The rationale behind the recommendation is that
+America guidelines (55). The rationale behind the recommendation is that
 patients taking laxatives may be asymptomatically colonized with
 \emph{C. difficile}, resulting in unnecessary antibiotic treatment
-(55--57). Furthermore, some studies identified laxatives as a risk
-factor for developing CDIs or recurrent CDIs (58--60) and a recent study
+(56--58). Furthermore, some studies identified laxatives as a risk
+factor for developing CDIs or recurrent CDIs (59--61) and a recent study
 found the proportion of severe CDIs was similar between patients taking
-and not taking laxatives (61). However, there have also been some
+and not taking laxatives (62). However, there have also been some
 studies that suggest laxatives are not a risk factor for developing CDIs
-(62, 63). Although, it is unclear whether laxatives impact CDI
+(63, 64). Although, it is unclear whether laxatives impact CDI
 susceptibility in human paitents, it is clear that laxatives also have a
-transient impact on the human microbiota (13, 64--67). Additional
+transient impact on the human microbiota (13, 65--68). Additional
 studies to examine the relationship between laxatives, \emph{C.
@@
 there have been no studies designed to evaluate the best administration
-route for FMTs (68). Nevertheless, results from the FMT National
+route for FMTs (69). Nevertheless, results from the FMT National
 Registry where 85\% of FMTs were delivered by colonoscopy demonstrate
 FMTs are highly effective treatment for recurrent CDIs with 90\%
-achieving resolution in the 1 month follow-up window (69). A surprising
+achieving resolution in the 1 month follow-up window (70). A surprising
 number of studies continue to hypothesize that PEG or bowel preparation
@@
 (NuLYTELY) prior to antibiotic treatment reduced disease severity that
-started recruitment in 2012 (70), but no results have been posted to
+started recruitment in 2012 (71), but no results have been posted to
 date. Here we sought to evaluate the impact of treating \emph{C.
@@
 weighed daily throughout the experiment and stool was collected for
-quantifying \emph{C. difficile} CFU and 16S rRNA gene sequencing. There
-were two groups of mice that received either a PBS or fecal microbiota
-transplant (FMT) gavage post-PEG treatment. The fecal microbiota
-transplant was prepared with stool samples collected from the mice in
-the experiment prior to the start of any treatments. The stool samples
-were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS
-and glycerol was added to make a 15\% glycerol solution. The solution
-was then aliquoted into tubes and stored at -80°C until the day of the
-gavage. An aliquot of both the FMT and PBS solutions were also set aside
-in the -80°C for 16S rRNA gene sequencing. The day of the gavage,
-aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The
-supernatant was then transferred to a separate tube to prevent the
-gavage needle from clogging with debris during gavage. The PBS solution
-that was administered to the other group was also 15\% glycerol. Each
-mouse was administered 100 microliters of either the FMT or PBS solution
-via gavage. When we refer to mice that cleared \emph{C. difficile}, we
-mean that no \emph{C. difficile} was detected in the first serial
-dilution (limit of detection: 100 CFU). In some experiments, we
-collected tissues for 16SrRNA gene sequencing, histopathology, or both.
-For 16S rRNA gene sequencing, we collected small snips of cecum,
-proximal colon, and distal colon tissues in microcentrifuge tubes, snap
-froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
-and colon tissues were placed into separate cassettes, fixed, and then
-submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,
-embedding, and hematoxylin and eosin (H\&E) staining.
+quantifying \emph{C. difficile} CFU and 16S rRNA gene sequencing. Fresh
+stool samples were collected from each mouse and split into two separate
+tubes. One tube was transferred to an anaerobic chamber the same day the
+sample was collected to quantify \emph{C. difficile}, while the other
+tube was snap frozen in liquid nitrogen and stored in the -80°C for 16S
+rRNA sequencing. There were two groups of mice that received either a
+PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The
+fecal microbiota transplant was prepared with stool samples collected
+from the mice in the experiment prior to the start of any treatments.
+The stool samples were transferred to an anaerobic chamber and diluted
+1:10 in reduced PBS and glycerol was added to make a 15\% glycerol
+solution. The solution was then aliquoted into tubes and stored at -80°C
+until the day of the gavage. An aliquot of both the FMT and PBS
+solutions were also set aside in the -80°C for 16S rRNA gene sequencing.
+The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM
+for 1 minute. The supernatant was then transferred to a separate tube to
+prevent the gavage needle from clogging with debris during gavage. The
+PBS solution that was administered to the other group was also 15\%
+glycerol. Each mouse was administered 100 microliters of either the FMT
+or PBS solution via gavage. When we refer to mice that cleared \emph{C.
+difficile}, we mean that no \emph{C. difficile} was detected in the
+first serial dilution (limit of detection: 100 CFU). In some
+experiments, we collected tissues for 16SrRNA gene sequencing,
+histopathology, or both. For 16S rRNA gene sequencing, we collected
+small snips of cecum, proximal colon, and distal colon tissues in
+microcentrifuge tubes, snap froze in liquid nitrogen, and stored at
+-80°C. For histopathology, cecum and colon tissues were placed into
+separate cassettes, fixed, and then submitted to McClinchey Histology
+Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and
+eosin (H\&E) staining.
 
@@
 AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom
-barcoded primers, as previously described (71). The PCR amplicons were
+barcoded primers, as previously described (72). The PCR amplicons were
 normalized (SequalPrep normalizatin plate kit from Thermo Fisher
@@
 \textbf{16S rRNA gene sequence analysis.} All sequences were processed
-with mothur (v. 1.43) using a previously published protocol (71, 72).
+with mothur (v. 1.43) using a previously published protocol (72, 73).
 Paired sequencing reads were combined and aligned with the SILVA (v.
-132) reference database (73) and taxonomy was assigned with a modified
+132) reference database (74) and taxonomy was assigned with a modified
 version of the Ribosomal Database Project reference sequences (v. 16)
-(74). The error rate for are sequencing data was 0.0559\% based on the
+(75). The error rate for are sequencing data was 0.0559\% based on the
 17 mock communities we ran with the samples. Samples were rarefied to
@@
 mothur-generated Bray-Curtis distance matrices with the adonis function
-from the vegan R package (75).
+from the vegan R package (76).
 
@@
 pathologist (ILB). Slides were evaluated using a scoring system
-developed for mouse models of \emph{C. difficile} infection (51). Each
+developed for mouse models of \emph{C. difficile} infection (52). Each
 slide was evaluated for edema, cellular infiltration, and inflammation
@@
 difficile} colonization status 10 days post-challenge where mice were
-categorized as either cleared or colonized (76, 77). We removed the
+categorized as either cleared or colonized (77, 78). We removed the
 \emph{C. difficile} genus relative abundance data prior to training the
@@
 of the training data for hyperparamter tuning. Permutation importance
-was performed as described previously (78) using mikropml (76, 77) with
+was performed as described previously (79) using mikropml (77, 78) with
 the random forest model because it had the highest AUROC value.
@@
 \textbf{Statistical analysis.} R (v. 4.0.2) and the tidyverse package
-(v. 1.3.0) were used for statistical analysis (79, 80). Kruskal-Wallis
+(v. 1.3.0) were used for statistical analysis (80, 81). Kruskal-Wallis
 tests with Bejamini-Hochberg correction for testing multiple time points
@@
 
+\leavevmode\hypertarget{ref-CastroCrdova2021}{}%
+35. Castro-Cordova P, Mora-Uribe P, Reyes-Ramirez R, Cofre-Araneda G,
+Orozco-Aguilar J, Brito-Silva C, Mendoza-Leon MJ, Kuehne SA, Minton NP,
+Pizarro-Guajardo M, Paredes-Sabja D. 2021. Entry of spores into
+intestinal epithelial cells contributes to recurrence of
+\emph{Clostridioides difficile} infection. Nature Communications 12.
+
 \leavevmode\hypertarget{ref-Frost2021}{}%
-35. Frost LR, Cheng JKJ, Unnikrishnan M. 2021. \emph{Clostridioides
+36. Frost LR, Cheng JKJ, Unnikrishnan M. 2021. \emph{Clostridioides
 difficile} biofilms: A mechanism of persistence in the gut? PLOS
@@
 \leavevmode\hypertarget{ref-Engevik2021}{}%
-36. Engevik MA, Danhof HA, Auchtung J, Endres BT, Ruan W, Bassères E,
+37. Engevik MA, Danhof HA, Auchtung J, Endres BT, Ruan W, Bassères E,
 Engevik AC, Wu Q, Nicholson M, Luna RA, Garey KW, Crawford SE, Estes MK,
@@
 \leavevmode\hypertarget{ref-Engevik2020}{}%
-37. Engevik MA, Engevik AC, Engevik KA, Auchtung JM, Chang-Graham AL,
+38. Engevik MA, Engevik AC, Engevik KA, Auchtung JM, Chang-Graham AL,
 Ruan W, Luna RA, Hyser JM, Spinler JK, Versalovic J. 2020.
@@
 \leavevmode\hypertarget{ref-Dejea2018}{}%
-38. Dejea CM, Fathi P, Craig JM, Boleij A, Taddese R, Geis AL, Wu X,
+39. Dejea CM, Fathi P, Craig JM, Boleij A, Taddese R, Geis AL, Wu X,
 Shields CED, Hechenbleikner EM, Huso DL, Anders RA, Giardiello FM, Wick
@@
 \leavevmode\hypertarget{ref-Semenyuk2015}{}%
-39. Semenyuk EG, Poroyko VA, Johnston PF, Jones SE, Knight KL, Gerding
+40. Semenyuk EG, Poroyko VA, Johnston PF, Jones SE, Knight KL, Gerding
 DN, Driks A. 2015. Analysis of bacterial communities during
@@
 \leavevmode\hypertarget{ref-Deng2018}{}%
-40. Deng H, Yang S, Zhang Y, Qian K, Zhang Z, Liu Y, Wang Y, Bai Y, Fan
+41. Deng H, Yang S, Zhang Y, Qian K, Zhang Z, Liu Y, Wang Y, Bai Y, Fan
 H, Zhao X, Zhi F. 2018. Bacteroides fragilis prevents \emph{Clostridium
@@
 \leavevmode\hypertarget{ref-Normington2021}{}%
-41. Normington C, Moura IB, Bryant JA, Ewin DJ, Clark EV, Kettle MJ,
+42. Normington C, Moura IB, Bryant JA, Ewin DJ, Clark EV, Kettle MJ,
 Harris HC, Spittal W, Davis G, Henn MR, Ford CB, Wilcox MH, Buckley AM.
@@
 \leavevmode\hypertarget{ref-Hassall2021}{}%
-42. Hassall J, Cheng JKJ, Unnikrishnan M. 2021. Dissecting individual
+43. Hassall J, Cheng JKJ, Unnikrishnan M. 2021. Dissecting individual
 interactions between pathogenic and commensal bacteria within a
@@
 \leavevmode\hypertarget{ref-Cirstea2018}{}%
-43. Cirstea M, Radisavljevic N, Finlay BB. 2018. Good bug, bad bug:
+44. Cirstea M, Radisavljevic N, Finlay BB. 2018. Good bug, bad bug:
 Breaking through microbial stereotypes. Cell Host \& Microbe 23:10--13.
@@
 \leavevmode\hypertarget{ref-BorgesCanha2015}{}%
-44. Canha MB. 2015. Role of colonic microbiota in colorectal
+45. Canha MB. 2015. Role of colonic microbiota in colorectal
 carcinogenesis: A systematic review. Revista Española de Enfermedades
@@
 \leavevmode\hypertarget{ref-Pereira2020}{}%
-45. Pereira FC, Wasmund K, Cobankovic I, Jehmlich N, Herbold CW, Lee KS,
+46. Pereira FC, Wasmund K, Cobankovic I, Jehmlich N, Herbold CW, Lee KS,
 Sziranyi B, Vesely C, Decker T, Stocker R, Warth B, Bergen M von, Wagner
@@
 \leavevmode\hypertarget{ref-Rodriguez2016}{}%
-46. Rodriguez C, Taminiau B, Korsak N, Avesani V, Broeck JV, Brach P,
+47. Rodriguez C, Taminiau B, Korsak N, Avesani V, Broeck JV, Brach P,
 Delmée M, Daube G. 2016. Longitudinal survey of \emph{Clostridium
@@
 \leavevmode\hypertarget{ref-Sangster2016}{}%
-47. Sangster W, Hegarty JP, Schieffer KM, Wright JR, Hackman J, Toole
+48. Sangster W, Hegarty JP, Schieffer KM, Wright JR, Hackman J, Toole
 DR, Lamendella R, Stewart DB. 2016. Bacterial and fungal microbiota
@@
 \leavevmode\hypertarget{ref-Vakili2020a}{}%
-48. Vakili B, Fateh A, Aghdaei HA, Sotoodehnejadnematalahi F, Siadat SD.
+49. Vakili B, Fateh A, Aghdaei HA, Sotoodehnejadnematalahi F, Siadat SD.
 2020. Intestinal microbiota in elderly inpatients with
@@
 \leavevmode\hypertarget{ref-Vakili2020b}{}%
-49. Vakili B, Fateh A, Aghdaei HA, Sotoodehnejadnematalahi F, Siadat SD.
+50. Vakili B, Fateh A, Aghdaei HA, Sotoodehnejadnematalahi F, Siadat SD.
 2020. Characterization of gut microbiota in hospitalized patients with
@@
 \leavevmode\hypertarget{ref-Li2019}{}%
-50. Li X, Chu Q, Huang Y, Xiao Y, Song L, Zhu S, Kang Y, Lu S, Xu J, Ren
+51. Li X, Chu Q, Huang Y, Xiao Y, Song L, Zhu S, Kang Y, Lu S, Xu J, Ren
 Z. 2019. Consortium of probiotics attenuates colonization of
@@
 \leavevmode\hypertarget{ref-Theriot2011}{}%
-51. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
+52. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
 VB. 2011. Cefoperazone-treated mice as an experimental platform to
@@
 \leavevmode\hypertarget{ref-Kester2020}{}%
-52. Kester JC, Brubaker DK, Velazquez J, Wright C, Lauffenburger DA,
+53. Kester JC, Brubaker DK, Velazquez J, Wright C, Lauffenburger DA,
 Griffith LG. 2020. \emph{Clostridioides difficile}-associated
@@
 \leavevmode\hypertarget{ref-Bergstrom2020}{}%
-53. Bergstrom K, Shan X, Casero D, Batushansky A, Lagishetty V, Jacobs
+54. Bergstrom K, Shan X, Casero D, Batushansky A, Lagishetty V, Jacobs
 JP, Hoover C, Kondo Y, Shao B, Gao L, Zandberg W, Noyovitz B, McDaniel
@@
 \leavevmode\hypertarget{ref-McDonald2018b}{}%
-54. McDonald LC, Gerding DN, Johnson S, Bakken JS, Carroll KC, Coffin
+55. McDonald LC, Gerding DN, Johnson S, Bakken JS, Carroll KC, Coffin
 SE, Dubberke ER, Garey KW, Gould CV, Kelly C, Loo V, Sammons JS, Sandora
@@
 \leavevmode\hypertarget{ref-Ahmad2017}{}%
-55. Ahmad SM, Blanco N, Dewart CM, Dobosz A, Malani AN. 2017. Laxative
+56. Ahmad SM, Blanco N, Dewart CM, Dobosz A, Malani AN. 2017. Laxative
 use in the setting of positive testing for \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Bilinskaya2018}{}%
-56. Bilinskaya A, Goodlet KJ, Nailor MD. 2018. Evaluation of a best
+57. Bilinskaya A, Goodlet KJ, Nailor MD. 2018. Evaluation of a best
 practice alert to reduce unnecessary \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Cook2020}{}%
-57. Cook PP, Nichols S, Coogan M, Opera J, DeHart M. 2020. Reduction in
+58. Cook PP, Nichols S, Coogan M, Opera J, DeHart M. 2020. Reduction in
 testing and change in testing algorithm associated with decrease in
@@
 \leavevmode\hypertarget{ref-Appaneal2019}{}%
-58. Appaneal HJ, Caffrey AR, Beganovic M, Avramovic S, LaPlante KL.
+59. Appaneal HJ, Caffrey AR, Beganovic M, Avramovic S, LaPlante KL.
 2019. Predictors of \emph{Clostridioides difficile} recurrence across a
@@
 \leavevmode\hypertarget{ref-Dubberke2011}{}%
-59. Dubberke ER, Yan Y, Reske KA, Butler AM, Doherty J, Pham V, Fraser
+60. Dubberke ER, Yan Y, Reske KA, Butler AM, Doherty J, Pham V, Fraser
 VJ. 2011. Development and validation of a \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Obritsch2010}{}%
-60. Obritsch MD, Stroup JS, Carnahan RM, Scheck DN. 2010.
+61. Obritsch MD, Stroup JS, Carnahan RM, Scheck DN. 2010.
 \emph{Clostridium difficile}-associated diarrhea in a tertiary care
@@
 \leavevmode\hypertarget{ref-White2019}{}%
-61. White NC, Mendo-Lopez R, Papamichael K, Cuddemi CA, Barrett C,
+62. White NC, Mendo-Lopez R, Papamichael K, Cuddemi CA, Barrett C,
 Daugherty K, Pollock N, Kelly CP, Alonso CD. 2019. Laxative use does not
@@
 \leavevmode\hypertarget{ref-Poirier2019}{}%
-62. Poirier D, Gervais P, Fuchs M, Roussy J-F, Paquet-Bolduc B, Trottier
+63. Poirier D, Gervais P, Fuchs M, Roussy J-F, Paquet-Bolduc B, Trottier
 S, Longtin J, Loo VG, Longtin Y. 2019. Predictors of
@@
 \leavevmode\hypertarget{ref-Huang2014}{}%
-63. Huang H, Wu S, Chen R, Xu S, Fang H, Weintraub A, Nord CE. 2014.
+64. Huang H, Wu S, Chen R, Xu S, Fang H, Weintraub A, Nord CE. 2014.
 Risk factors of \emph{Clostridium difficile} infections among patients
@@
 \leavevmode\hypertarget{ref-Gorkiewicz2013}{}%
-64. Gorkiewicz G, Thallinger GG, Trajanoski S, Lackner S, Stocker G,
+65. Gorkiewicz G, Thallinger GG, Trajanoski S, Lackner S, Stocker G,
 Hinterleitner T, Gülly C, Högenauer C. 2013. Alterations in the colonic
@@
 \leavevmode\hypertarget{ref-Shobar2016}{}%
-65. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
+66. Shobar RM, Velineni S, Keshavarzian A, Swanson G, DeMeo MT, Melson
 JE, Losurdo J, Engen PA, Sun Y, Koenig L, Mutlu EA. 2016. The effects of
@@
 \leavevmode\hypertarget{ref-Drago2016}{}%
-66. Drago L, Toscano M, Grandi RD, Casini V, Pace F. 2016. Persisting
+67. Drago L, Toscano M, Grandi RD, Casini V, Pace F. 2016. Persisting
 changes of intestinal microbiota after bowel lavage and colonoscopy.
@@
 \leavevmode\hypertarget{ref-Jalanka2014}{}%
-67. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
+68. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
 Gowland P, Hoad C, Garsed K, Lam C, Palva A, Spiller RC, Vos WM de.
@@
 \leavevmode\hypertarget{ref-Kampouri2021}{}%
-68. Kampouri E, Croxatto A, Prod'hom G, Guery B. 2021.
+69. Kampouri E, Croxatto A, Prod'hom G, Guery B. 2021.
 \emph{Clostridioides difficile} infection, still a long way to go.
@@
 \leavevmode\hypertarget{ref-Kelly2021}{}%
-69. Kelly CR, Yen EF, Grinspan AM, Kahn SA, Atreja A, Lewis JD, Moore
+70. Kelly CR, Yen EF, Grinspan AM, Kahn SA, Atreja A, Lewis JD, Moore
 TA, Rubin DT, Kim AM, Serra S, Nersesova Y, Fredell L, Hunsicker D,
@@
 \leavevmode\hypertarget{ref-Dieterle2018}{}%
-70. Dieterle MG, Rao K, Young VB. 2018. Novel therapies and preventative
+71. Dieterle MG, Rao K, Young VB. 2018. Novel therapies and preventative
 strategies for primary and recurrent \emph{Clostridiuml difficile}
@@
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-71. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+72. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-72. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+73. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
@@
 \leavevmode\hypertarget{ref-Quast2012}{}%
-73. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
+74. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
 J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
@@
 \leavevmode\hypertarget{ref-Cole2013}{}%
-74. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
+75. Cole JR, Wang Q, Fish JA, Chai B, McGarrell DM, Sun Y, Brown CT,
 Porras-Alfaro A, Kuske CR, Tiedje JM. 2013. Ribosomal database project:
@@
 \leavevmode\hypertarget{ref-Vegan2018}{}%
-75. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+76. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 Minchin PR, O'Hara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
@@
 \leavevmode\hypertarget{ref-mikropml}{}%
-76. Topçuoğlu B, Lapp Z, Sovacool KL, Snitkin E, Wiens J, Schloss PD.
+77. Topçuoğlu B, Lapp Z, Sovacool KL, Snitkin E, Wiens J, Schloss PD.
 2020. mikRopML: User-friendly r package for robust machine learning
@@
 \leavevmode\hypertarget{ref-Topcuoglu2021}{}%
-77. Topçuoğlu B, Lapp Z, Sovacool K, Snitkin E, Wiens J, Schloss P.
+78. Topçuoğlu B, Lapp Z, Sovacool K, Snitkin E, Wiens J, Schloss P.
 2021. Mikropml: User-friendly r package for supervised machine learning
@@
 \leavevmode\hypertarget{ref-Topcuoglu2020}{}%
-78. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
+79. Topçuoğlu BD, Lesniak NA, Ruffin MT, Wiens J, Schloss PD. 2020. A
 framework for effective application of machine learning to
@@
 \leavevmode\hypertarget{ref-r_citation_2020}{}%
-79. R Core Team. 2020. R: A language and environment for statistical
+80. R Core Team. 2020. R: A language and environment for statistical
 computing. R Foundation for Statistical Computing, Vienna, Austria.
@@
 \leavevmode\hypertarget{ref-Tidyverse2019}{}%
-80. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
+81. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
 Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
@@
 \newpage
-
-\includegraphics{figure_1.pdf}
 
@@
 treatment group and circles represent samples from individual mice.
-Asterisks indicate time points where the weight change or CFU/g was
-significantly different (\emph{P} \textless{} 0.05) between groups by
-the Kruskal-Wallis test with Benjamini-Hochberg correction for testing
-multiple time points. The data presented are from a total of 5 separate
-experiments. \newpage
-
-\includegraphics{figure_2.pdf} \textbf{Figure 2. 5-day PEG treatment
-disrupts the stool microbiota for a longer amount of time compared to
-clindamycin-treated mice.} A. Principal Coordinate analysis (PCoA) of
-Bray-Curtis distances from stool samples collected throughout the
-experiment. Each circle represents a sample from an individual mouse and
-the transparency of the symbol corresponds to the day post-challenge.
-See Data Set S1, sheet 1 for PERMANOVA results. B. Bray-Curtis distances
-of stool samples collected on either day 10 or 30 post-challenge
-relative to the baseline sample collected for each mouse (before any
-drug treatments were administered). The symbols represent samples from
-individual mice and the line indicates the median value for each
-treatment group. C. Shannon diversity in stool communities over time.
-The line indicates the median value for each treatment group (Data Set
-S1, sheet 2). D. 14 of the 33 genera affected by PEG treatment (Data Set
-S1, sheet 3). The symbols represent the median relative abundance for a
-treatment group at either baseline (open circle) or 1-day post treatment
-(closed circle). Relative abundance data from paired baseline and 1-day
-post treatment stool sampes from the 5-day PEG and 5-day PEG plus 10-day
-recovery groups were analyzed by paired Wilcoxan signed-rank test with
-Benjamini-Hochberg correction for testing all identified genera. The
-clindamycin and 5-day PEG plus clindamycin treatment groups are shown on
-the plot for comparison. E. 6 of the 24 genera that were significantly
-different between the treatment groups over multiple time points (see
-Data Set S1, sheet 4 for complete list). The 5-day PEG plus clindamycin
-treatment group was only followed through 6-days post-challenge.
-Differences between treatment groups were identified by Kruskal-Wallis
-test with Benjamini-Hochberg correction for testing all identified
-genera (*, \emph{P} \textless{} 0.05). The gray vertical line (D) and
-horizontal vertical lines (E) indicate the limit of detection. \newpage
-
-\includegraphics{figure_3.pdf} \textbf{Figure 3. 5-day PEG treatment
-does not result in more severe CDIs, although mucosal microbiota is
-altered.} A. Shannon diversity in cecum communities over time. The
-colors of the symbols and lines represent individual and median relative
-abundance values for four treatment groups (Data Set S1, sheet 6). B.
-PCoA of Bray-Curtis distances from mucosal samples collected throughout
-the experiment. Circles, triangles, and squares indicate the cecum,
-proximal colon, and distal colon communities, respectively. Transparency
-of the symbol corresponds to the day post-challenge that the sample was
-collected. See Data Set S1, sheet 7 for PERMANOVA results. C. The median
-relative abundance of the 4 genera that were significantly different
-between the cecum communities of different treatment groups on day 6 and
-day 30 post-challenge (Data Set S1, sheet 8). The gray vertical lines
-indicate the limit of detection. D-E. The histopathology summary scores
-from cecum and colon H\&E stained tissue sections. The summary score is
-the total score based on evaluation of edema, cellular infiltration, and
+Asterisks indicate the specific time points on the x-axis where the
+weight change or CFU/g was significantly different (\emph{P} \textless{}
+0.05) between the 3-4 groups indicated on the plot by the Kruskal-Wallis
+test with Benjamini-Hochberg correction for testing multiple time
+points. The data presented are from a total of 5 separate experiments.
+\newpage
+
+\textbf{Figure 2. 5-day PEG treatment disrupts the stool microbiota for
+a longer amount of time compared to clindamycin-treated mice.} A.
+Principal Coordinate analysis (PCoA) of Bray-Curtis distances from stool
+samples collected throughout the experiment. Each circle represents a
+sample from an individual mouse and the transparency of the symbol
+corresponds to the day post-challenge. See Data Set S1, sheet 1 for
+PERMANOVA results. B. Bray-Curtis distances of stool samples collected
+on either day 10 or 30 post-challenge relative to the baseline sample
+collected for each mouse (before any drug treatments were administered).
+The symbols represent samples from individual mice and the line
+indicates the median value for each treatment group. C. Shannon
+diversity in stool communities over time. The line indicates the median
+value for each treatment group (Data Set S1, sheet 2). D. 14 of the 33
+genera affected by PEG treatment (Data Set S1, sheet 3). The symbols
+represent the median relative abundance for a treatment group at either
+baseline (open circle) or 1-day post treatment (closed circle). Relative
+abundance data from paired baseline and 1-day post treatment stool
+sampes from the 5-day PEG and 5-day PEG plus 10-day recovery groups were
+analyzed by paired Wilcoxan signed-rank test with Benjamini-Hochberg
+correction for testing all identified genera. The clindamycin and 5-day
+PEG plus clindamycin treatment groups are shown on the plot for
+comparison. E. 6 of the 24 genera that were significantly different
+between the treatment groups over multiple time points (see Data Set S1,
+sheet 4 for complete list). The 5-day PEG plus clindamycin treatment
+group was only followed through 6-days post-challenge. Differences
+between treatment groups were identified by Kruskal-Wallis test with
+Benjamini-Hochberg correction for testing all identified genera (*,
+\emph{P} \textless{} 0.05). The gray vertical line (D) and horizontal
+vertical lines (E) indicate the limit of detection. \newpage
+
+\textbf{Figure 3. 5-day PEG treatment does not result in more severe
+CDIs, although mucosal microbiota is altered.} A. Shannon diversity in
+cecum communities over time. The colors of the symbols and lines
+represent individual and median relative abundance values for four
+treatment groups (Data Set S1, sheet 6). B. PCoA of Bray-Curtis
+distances from mucosal samples collected throughout the experiment.
+Circles, triangles, and squares indicate the cecum, proximal colon, and
+distal colon communities, respectively. Transparency of the symbol
+corresponds to the day post-challenge that the sample was collected. See
+Data Set S1, sheet 7 for PERMANOVA results. C. The median relative
+abundance of the 4 genera that were significantly different between the
+cecum communities of different treatment groups on day 6 and day 30
+post-challenge (Data Set S1, sheet 8). The gray vertical lines indicate
+the limit of detection. D-E. The histopathology summary scores from
+cecum and colon H\&E stained tissue sections. The summary score is the
+total score based on evaluation of edema, cellular infiltration, and
 inflammation in either the cecum or colon tissue. Each category is given
@@
 
-\includegraphics{figure_4.pdf} \textbf{Figure 4. 1-day PEG treatment
-renders mice susceptible to transient \emph{C. difficile} colonization.}
-A. Setup of the experimental time line for the 1-day PEG treated mice
-consisting of 3 treatment groups. 1. Clindamycin was administered at 10
-mg/kg by intraperitoneal injection. 2. 15\% PEG 3350 was administered in
-the drinking water for 1 day. 3. 1-day PEG plus 1-day recovery. The
-three treatment groups were then challenged with 10\textsuperscript{5}
+\textbf{Figure 4. 1-day PEG treatment renders mice susceptible to
+transient \emph{C. difficile} colonization.} A. Setup of the
+experimental time line for the 1-day PEG treated mice consisting of 3
+treatment groups. 1. Clindamycin was administered at 10 mg/kg by
+intraperitoneal injection. 2. 15\% PEG 3350 was administered in the
+drinking water for 1 day. 3. 1-day PEG plus 1-day recovery. The three
+treatment groups were then challenged with 10\textsuperscript{5}
 \emph{C. difficile} 630 spores. B. \emph{C. difficile} CFU/gram stool
@@
 
-\includegraphics{figure_5.pdf} \textbf{Figure 5. 1-day PEG treatment
-post \emph{C. difficile} challenge prolongs colonization regardless of
-whether an FMT is also administered.} A. Setup of the experimental time
-line for experiments with post-challenge PEG treated mice. There were a
-total of 4 different treatment groups. All mice were administered 10
-mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
+\textbf{Figure 5. 1-day PEG treatment post \emph{C. difficile} challenge
+prolongs colonization regardless of whether an FMT is also
+administered.} A. Setup of the experimental time line for experiments
+with post-challenge PEG treated mice. There were a total of 4 different
+treatment groups. All mice were administered 10 mg/kg clindamycin
+intraperitoneally (IP) 1 day before challenge with
 10\textsuperscript{3-5} \emph{C. difficile} 630 spores. 1. Received no
@@
 
-\includegraphics{figure_6.pdf} \textbf{Figure 6. For 1-day PEG treatment
-post \emph{C. difficile} challenge mice that also receive an FMT only
-some bacterial genera were restored.} A. PCoA of Bray-Curtis distances
-from stool samples collected over time as well as the FMT solution that
-was administered to one of the treatment groups. Each circle represents
-an individual sample, the transparency of the circle corresponds to day
-post-challenge. See Data Set S1, sheet 17 for PERMANOVA results. B.
-Median relative abundances of 2 genera that were significantly different
-over multiple time points in mice that were administered either FMT or
-PBS solution via gavage C. Median relative abundances of the top 6 out
-of 24 genera that were significant over multiple time points, plotted
-over time (see Data Set S1, sheet 18 for complete list). For B-C,
-colored rectangles indicates the 1-day PEG treatment period for
-applicable groups. Gray horizontal lines represent the limit of
-detection. Differences between treatment groups were identified by
-Kruskal-Wallis test with Benjamini-Hochberg correction for testing all
-identified genera. For pairwise comparisons of the groups (B), we
-performed pairwise Wilcoxon comparisons with Benjamini-Hochberg
-correction for testing all combinations of group pairs.
+\textbf{Figure 6. For 1-day PEG treatment post \emph{C. difficile}
+challenge mice that also receive an FMT only some bacterial genera were
+restored.} A. PCoA of Bray-Curtis distances from stool samples collected
+over time as well as the FMT solution that was administered to one of
+the treatment groups. Each circle represents an individual sample, the
+transparency of the circle corresponds to day post-challenge. See Data
+Set S1, sheet 17 for PERMANOVA results. B. Median relative abundances of
+2 genera that were significantly different over multiple time points in
+mice that were administered either FMT or PBS solution via gavage C.
+Median relative abundances of the top 6 out of 24 genera that were
+significant over multiple time points, plotted over time (see Data Set
+S1, sheet 18 for complete list). For B-C, colored rectangles indicates
+the 1-day PEG treatment period for applicable groups. Gray horizontal
+lines represent the limit of detection. Differences between treatment
+groups were identified by Kruskal-Wallis test with Benjamini-Hochberg
+correction for testing all identified genera. For pairwise comparisons
+of the groups (B), we performed pairwise Wilcoxon comparisons with
+Benjamini-Hochberg correction for testing all combinations of group
+pairs.
 
@@
 
-\includegraphics{figure_7.pdf} \textbf{Figure 7. Specific microbiota
-features associated with prolonged \emph{C. difficile} colonization in
-PEG treated mice.} A. Top ten bacteria that contributed to the random
-forest model trained on 5-day post-challenge community relative
-abundance data, predicting whether mice would still be colonized with
-\emph{C. difficile} 10 days post-challenge. The median (point) and
-interquartile range (lines) change in AUROC when the bacteria were left
-out of the model by permutation feature importance analysis. B. The
-median relative abundances of the top ten bacteria that contributed to
-the random forest classification model at 5 days post-challenge . Red
-indicates the mice were still colonized with \emph{C. difficile} while
-blue indicates mice that cleared \emph{C. difficile} 10 days
-post-challenge and the black horizontal line represents the median
-relative abundance for the two categories. Each symbol represents a
-stool sample from an individual mouse and the shape of the symbol
-indicates whether the PEG-treated mice received a 5-day (Fig. 1-3),
-1-day (Fig. 4) or post-challenge PEG (Fig. 5-6) treatment. C. The median
-relative abundances of the 5 genera with greater than 1\% median
-relative abundance in the stool community over time. For B-C, the gray
-horizontal lines represents the limit of detection. \newpage
-
-\includegraphics{figure_8.pdf}
+\textbf{Figure 7. Specific microbiota features associated with prolonged
+\emph{C. difficile} colonization in PEG treated mice.} A. Top ten
+bacteria that contributed to the random forest model trained on 5-day
+post-challenge community relative abundance data, predicting whether
+mice would still be colonized with \emph{C. difficile} 10 days
+post-challenge. The median (point) and interquartile range (lines)
+change in AUROC when the bacteria were left out of the model by
+permutation feature importance analysis. B. The median relative
+abundances of the top ten bacteria that contributed to the random forest
+classification model at 5 days post-challenge . Red indicates the mice
+were still colonized with \emph{C. difficile} while blue indicates mice
+that cleared \emph{C. difficile} 10 days post-challenge and the black
+horizontal line represents the median relative abundance for the two
+categories. Each symbol represents a stool sample from an individual
+mouse and the shape of the symbol indicates whether the PEG-treated mice
+received a 5-day (Fig. 1-3), 1-day (Fig. 4) or post-challenge PEG (Fig.
+5-6) treatment. C. The median relative abundances of the 5 genera with
+greater than 1\% median relative abundance in the stool community over
+time. For B-C, the gray horizontal lines represents the limit of
+detection. \newpage
 
 \textbf{Figure 8. Schematic summarizing findings.} The gut microbiota of
-our C57Bl/6 mice is resistant to \emph{C. difficle} but treatment with
-either the antibiotic, clindamycin, or the osmotic laxative, PEG 3350,
-renders the mice susceptible to \emph{C. difficile} colonization.
-Recovery of colonization resistance in clindamycin-treated mice is
-relatively straightforward and the mice clear \emph{C.difficile} within
-10 days post-challenge. However, for mice that received either a 5-day
-PEG treatment prior to \emph{C. difficile} challenge or a 1-day PEG
-treatment post-challenge recovery of colonization resistance was delayed
-because most mice were still colonized with \emph{C. difficile} 10 days
-post-challenge. We found increased relative abundances of
-\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} were associated
-with recovery of colonization resistance, while increased relative
-abundances of \emph{Enterobacteriaceae} and \emph{Bacteroides} were
-associated with prolonged \emph{C. difficle} colonization. \newpage
-
-\includegraphics{figure_S1.pdf} \textbf{Figure S1. Microbiota dynamics
-post-challenge in the 5-day PEG treatment plus 10-day recovery mice.} A.
-\emph{C. difficile} CFU/g over time in the stool samples collected from
-5-day PEG treated mice that were allowed to recover for 10 days prior to
-challenge. Same data presented in Fig. 1C, but the data for the other 3
-treatment groups have been removed and each line represents the CFU over
-time for an individual mouse. Mouse 10 was found dead 6 days
-post-challenge. B. Relative abundances of eight bacterial genera from
-day 0 post-challenge onward in each of the 10-day recovery mice. We
-analyzed samples from day 0 and day 8 post-challenge, which represented
-the time points where mice were challenged with \emph{C. difficile} and
-when the median relative \emph{C. difficile} CFU stabilized for the
-group using the paired Wilcoxan signed-rank test, but no genera were
-significantly different after Benjamini-Hochberg correction (Data Set
-S1, sheet 5). \newpage
-
-\includegraphics{figure_S2.pdf} \textbf{Figure S2. PEG treatment still
-has a large impact on the mucosal microbiota 6 days post-challenge} A.
-The relative abundances of the 10 bacterial genera that were
-significantly different between treatment groups at 6 days
-post-infection in the cecum tissue (the relative abundances of the 10
-genera were also significantly different in the proximal and distal
-colon tissues, Data Set S1, sheets 8, 9, and 10). Each symbol represents
-a tissue sample from an individual mouse, the black horizontal lines
-represents the median relative abundances for each treatment group. B.
-The relative abundance of \emph{Peptostreptococacceae} in the three
-types of tissue sample communities over time. For A-B, the gray
-horizontal lines represent the limit of detection. \newpage
-
-\includegraphics{figure_S3.pdf} \textbf{Figure S3. \emph{C. difficile}
-challenge does not enhance the disruptive effect of PEG on the
-microbiota.} A, D. PCoAs of the Bray-Curtis distances from the stool (A)
-and tissue (D) communities from mock- and \emph{C. difficile}-challenged
-treatment groups. Each symbol represents a sample from an individual
-mouse with open and closed circles representing mock and \emph{C.
-difficile}-challenged mice, respectively. B, E. Median Shannon diversity
-in stool (B) and tissue (E) communities collected over time. C, F. The
-median relative abundances of genera that were significantly different
-between the \emph{C. difficile} challenged treatment groups in either
-the stool (Fig. 2E) or cecum tissue (Fig. 3C) communities in the stool
-(C) and tissue (F) communities from mock- and \emph{C.
-difficile}-challenged mice. For B-F, the dashed and solid lines
-represent the median value for mock and \emph{C. difficile}-challenged
-mice, respectively. For E-F, tissues from mock-challenged clindamycin
-treated mice were only collected 4 days post-challenge so there is no
-dashed line for this group. \newpage
+our lab's C57Bl/6 mice is resistant to \emph{C. difficle} but treatment
+with either clindamycin or the osmotic laxative, PEG 3350, renders the
+mice susceptible to \emph{C. difficile} colonization. Recovery of
+colonization resistance in clindamycin-treated mice is relatively
+straightforward and the mice clear \emph{C.difficile} within 10 days
+post-challenge. However, recovery of colonization resistance was delayed
+for mice that received either a 5-day PEG pre-treatment or a 1-day PEG
+post-\emph{C. difficile} challenge treatment. We found increased
+relative abundances of \emph{Porphyromonadaceae} and
+\emph{Lachnospiraceae} were associated with \emph{C. difficle}
+clearance, while increased relative abundances of
+\emph{Enterobacteriaceae} and \emph{Bacteroides} were associated with
+prolonged \emph{C. difficle} colonization. \newpage
+
+\textbf{Figure S1. Microbiota dynamics post-challenge in the 5-day PEG
+treatment plus 10-day recovery mice.} A. \emph{C. difficile} CFU/g over
+time in the stool samples collected from 5-day PEG treated mice that
+were allowed to recover for 10 days prior to challenge. Same data
+presented in Fig. 1C, but the data for the other 3 treatment groups have
+been removed and each line represents the CFU over time for an
+individual mouse. Mouse 10 was found dead 6 days post-challenge. B.
+Relative abundances of eight bacterial genera from day 0 post-challenge
+onward in each of the 10-day recovery mice. We analyzed samples from day
+0 and day 8 post-challenge, which represented the time points where mice
+were challenged with \emph{C. difficile} and when the median relative
+\emph{C. difficile} CFU stabilized for the group using the paired
+Wilcoxan signed-rank test, but no genera were significantly different
+after Benjamini-Hochberg correction (Data Set S1, sheet 5). \newpage
+
+\textbf{Figure S2. PEG treatment still has a large impact on the mucosal
+microbiota 6 days post-challenge} A. The relative abundances of the 10
+bacterial genera that were significantly different between treatment
+groups at 6 days post-infection in the cecum tissue (the relative
+abundances of the 10 genera were also significantly different in the
+proximal and distal colon tissues, Data Set S1, sheets 8, 9, and 10).
+Each symbol represents a tissue sample from an individual mouse, the
+black horizontal lines represents the median relative abundances for
+each treatment group. B. The relative abundance of
+\emph{Peptostreptococacceae} in the three types of tissue sample
+communities over time. For A-B, the gray horizontal lines represent the
+limit of detection. \newpage
+
+\textbf{Figure S3. \emph{C. difficile} challenge does not enhance the
+disruptive effect of PEG on the microbiota.} A, D. PCoAs of the
+Bray-Curtis distances from the stool (A) and tissue (D) communities from
+mock- and \emph{C. difficile}-challenged treatment groups. Each symbol
+represents a sample from an individual mouse with open and closed
+circles representing mock and \emph{C. difficile}-challenged mice,
+respectively. B, E. Median Shannon diversity in stool (B) and tissue (E)
+communities collected over time. C, F. The median relative abundances of
+genera that were significantly different between the \emph{C. difficile}
+challenged treatment groups in either the stool (Fig. 2E) or cecum
+tissue (Fig. 3C) communities in the stool (C) and tissue (F) communities
+from mock- and \emph{C. difficile}-challenged mice. For B-F, the dashed
+and solid lines represent the median value for mock and \emph{C.
+difficile}-challenged mice, respectively. For E-F, tissues from
+mock-challenged clindamycin treated mice were only collected 4 days
+post-challenge so there is no dashed line for this group. \newpage
 
